### 2016-05-04 rigid_bodies.tex
Original commit

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
\usepackage[pdftex]{graphicx}
\usepackage{epstopdf}
\usepackage{caption}
\usepackage{subfig}
\usepackage[]{units}
\usepackage[]{natbib}
\usepackage[]{threeparttable}
\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
\newcommand{\tr}[1]{#1^T}                               % transposition
\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative

\begin{document}

\title{A Simplified Approach for Molecular Dynamics with Rigid Bodies}

\author{Charlles R. A. Abreu}
\email{abreu@eq.ufrj.br}
\affiliation{Chemical Engineering Department, Escola de Quimica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}

\date{\today}

\maketitle

\section{Introduction}
The advances in hardware along with the improvements in molecular dynamics algorithms and software has allowed to extend both the time and length scales attainable in atomistic simulations. This progress has resulted in an ambitious attempt to elucidate complex physical phenomena, such as the mechanism underlying molecular motors, the pathways for nanoparticle assembly, the interfacial phenomena, just to name a few. Because of its complexity and being computationally demanding, this challenging task requires strategies to further reduce the computational time.

One of the strategies commonly used to speed up the simulations is the suppression of the uninteresting degrees of freedom, as might be the case of the intramolecular vibrations. The rigid body molecular dynamics appears as the more suitable approach for rigid molecules. In contrast to the widely used iterative methods,\cite{Ryckaert1977,Andersen1983} one is able to make an implicit treatment of the holonomic constraints, which is computationally more efficient. In addition, and most important, it is possible to obtain a reversible and symplectic integrator in the microcanonical ensemble. The property of being symplectic implies the conservation of volume in phase space. This feature, together with time reversibility are mandatory for numerical integrators used in hybrid monte carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic reversible integrator for the rotational motion in terms of the unit quaternion. This approach has been successfully used in pure molecular dynamics as well as in HMC methods to simulate a wide range of systems including the ubiquitous case of water \cite{Sakamaki2011,Reinhardt2012,Palmer2014,Gonzales2014}, ice water \cite{Geiger2014}, hydrates \cite{Tribello2009,Gorman2012}, diffusing gases in metal-organic frameworks \citep{Ghoufi2010}. It is worth noting that it also can be used for systems designed as a collection of interconnected rigid bodies, which is a coarse-graining strategy widely used for proteins \citep{Terada2003}, molecular machines \citep{Akimov2008,Konyukhov2010}, nanoparticles \cite{Knorowski2012,Patra2013}, etc.

In this paper, we present a simplified version for the equations of motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as the corresponding numerical integrators to perform molecular dynamics. The contribution for the NVE case relies on the factorization of the rotation matrix, which allows a Hamiltonian formalism that avoids the use of the extended phase space methodology, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002}. In the NVT case, both the translational and rotational degrees of freedom (DOF) of all particles are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.\cite{Kamberaj2005}, which uses two independent thermostats, one for the rotational and another one for the translational DOF. It should be mentioned here that there are several numerical integrators\cite{Omelyan2007,Omelyan2008,vanZon2008} for the equations of motion for rigid bodies in the NVE ensemble which are more accurate than the symplectic Verlet-type integrators. However, since the application of that methods is limited to Hamiltonian systems, it's unclear whether their benefits will extend to the numerical integrators in the NVT ensemble. This, together with its great applicability, are the main reasons why our approach is based on the Verlet-type integrators.

We have also derived an expression for calculating the pressure of a rigid body system whose constituents atoms interact via a pairwise potential and are subjected to periodic boundary conditions. That expression depends on the resultant interaction forces between rigid body pairs. This implies that there is no need to evaluate the unknown forces of constraint, which one must take into account if the pressure is calculated in terms of the interatomic forces. This paper is organized as follows: In Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism. As a preamble to the integration schemes, we show an alternative way of interpreting the action of the Liouville operators on the phase space variables. In Sec.~\ref{sec:moleculardynamics} we present the factorization of the Liouville operator for both the NVE and NVT ensembles. Following this, in Sec.~\ref{sec:pressure} we show the pressure expression developed here and we compare the result with that obtained from other method. In Sec.~\ref{sec:performance} we discuss the performance of the numerical integrators according to the conservation of the total energy. For the NVT ensemble, we also present the results from a test that checks if the potential and kinetic energies are actually sampled according to the canonical distribution. Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations. Only ordinary matrix/vector operations are present in the formulation.

\section{Mechanics of Rigid Body Motion}
\label{sec:mechanics}
In this section we review one particular factorization of the rotation matrix which allows to obtain a mathematical relation between the rate of change of the four-dimensional quaternion and the angular velocity vector, while automatically satisfying the restriction imposed on the quaternion. That relation was already introduced by Kneller \textit{et al}\cite{Kneller1994} in the context of Lagrangian dynamics for rigid bodies, but here we use it in the derivation of the Hamiltonian equations of motion.

\subsection{Kinematics}

Consider a rigid body composed of $n_p$ individual particles, each particle $j$ with mass $M_j$ and located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
\[
\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
\]
where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body. The position of each particle $j$ in a space-fixed reference frame whose origin coincides with the body's center of mass is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$. In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
\[
{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{array}{ccc}
{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
-{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
-{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
\end{array} \right],
\]

For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal. Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
\[
{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
I_1 &   0 &   0 \\
  0 & I_2 &   0 \\
  0 &   0 & I_3 \\
\end{array} \right],
\]
where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are the corresponding eigenvectors. Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}. Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$. Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix. If we differentiate both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix. This product is related to $\vt \omega$, the angular velocity in the body-fixed frame, by\cite{Haug1989}
\begin{equation}
\label{eq:relation_A_omega}
\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
\end{equation}
where $\mt S(\cdot)$ is the skew-symmetric operator defined as
\begin{equation}
\label{eq:operator_S}
\mt S(\vt x) = \left[ \begin{array}{ccc}
 0   & -x_3 &  x_2 \\
 x_3 &  0   & -x_1 \\
-x_2 &  x_1 &  0
\end{array}\right].
\end{equation}

The operator  $\mt S(\cdot)$ can be used to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.

\subsection{Euler Parameters and Unit Quaternions}

It is possible to represent the orientation of a rigid body by Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$ that satisfies\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
\end{equation}

Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $R^4$. Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix. Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{widetext}
\begin{equation}
\label{eq:A_from_q}
\mt A = \left[ \begin{array}{ccc}
q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3) & 2(q_1 q_3 - q_0 q_2) \\
2(q_1 q_2 - q_0 q_3) & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
2(q_1 q_3 + q_0 q_2) & 2(q_2 q_3 - q_0 q_1) & q_0^2 - q_1^2 - q_2^2 + q_3^2  
\end{array} \right].
\end{equation}
\end{widetext}

The rate of change of the Euler parameters is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0.
\end{equation}

This equation implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$. Let us now introduce two new matrices that also depend on the Euler parameters, which are
\begin{subequations}
\label{eq:def_B_and_C}
\begin{align}
\mt B &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 & -q_3 &  q_2 \\
 q_3 &  q_0 & -q_1 \\
-q_2 &  q_1 &  q_0
\end{array}
\right] \text{ and} \label{eq:def_B} \\
\mt C &= \left[
\begin{array}{rrrr}
-q_1 & -q_2 & -q_3 \\
 q_0 &  q_3 & -q_2 \\
-q_3 &  q_0 &  q_1 \\
 q_2 & -q_1 &  q_0
\end{array}
\right].
\end{align}
\end{subequations}

Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited in Appendix \ref{sec:auxiliary_math} with our notation. The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}

This is a simple and convenient factorization, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters. Such relation is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

As explained in the appendix, premultiplication with either $\mt B$ or $\mt C$ transforms a vector into a quaternion that lies in the hyperplane orthogonal to $\vt q$. Therefore, the equation above satisfies Eq.~\ref{eq:diff_qTq} identically and guarantees that the norm of $\vt q$ remains unchanged. Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.

\subsection{Hamiltonian Dynamics}
\label{sec:hamiltonian}
As usual, the dynamics of a rigid body can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. Several authors have obtained Hamiltonian equations of motion for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.

We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively. Since $K = 1/2 m \tr{\dot{\vt r}} \dot{\vt r} + 1/2 \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
\[
\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
\]

To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the quaternion momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By performing the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
\begin{equation}
\label{eq:conj_momentum}
\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
\end{equation}

Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$. Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
\begin{equation}
\label{eq:H_with_B}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi + U(\vt r, \vt q).
\end{equation}

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} The equation above is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only. However, if we define
\[
\mt \Omega = \left[
\begin{array}{rrrr}
-\pi_1 & -\pi_2 & -\pi_3 \\
 \pi_0 & -\pi_3 &  \pi_2 \\
 \pi_3 &  \pi_0 & -\pi_1 \\
-\pi_2 &  \pi_1 &  \pi_0
\end{array}
\right],
\]
it follows that $\tr{\mt B}{\vt \pi} = -\tr{\mt \Omega}{\vt q}$ and, therefore,
\begin{equation}
\label{eq:H_with_Omega}
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
\end{equation}

Now this is convenient for differentiation with respect to $\vt q$.  Hence, after carrying out the differentiations as described in Appendix \ref{sec:Diff_Rules} and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following system of differential equations to describe the dynamics of a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = -\diff{U}{\vt r} \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega - \diff{U}{\vt q}
\end{align}
\end{subequations}

For a rigid body composed of individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body. To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}. As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.

Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$ and according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} {\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\mt B_k$ is the corresponding permutation matrix among those presented in Appendix \ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]

In order to rewrite Eq.~\ref{eq:EDO_system}, the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ are opened using Eq.~\ref{eq:product_B_vector}, leading to the final form for the equations of motion:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \vt F \\
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k {\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k {\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

\subsection{Symplectic Numerical Integration}  
\label{symplectic}
An exact solution for the system of equations represented by Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, it would describe a reversible, measure-preserving, and conservative motion with respect to the Hamiltonian $\mathcal{H}$. Nevertheless, a general analytical solution does not exist. This is so, above all, because $U(\vt r, \vt q)$ is arbitrary, but even in the simplest case of free motion (i.e., when $U$ is constant), no closed-form solution is known due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}. For a given trajectory in phase-space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i L f,
\]
where $i L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{i L t}f_0$, where $f_0$ is the initial condition for $f$. As $f$ can be a phase-space coordinate, premultiplication with the exponential operator $e^{i L t}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} normally represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}. Let us analyze three of them, which will be useful later on. In what follows, subscript $0$ always denotes ``initial values''. We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$. The exact solution for this case is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0
\end{align*}

Clearly, this corresponds to a uniform linear translation. Since the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected. Of course, the solution above is exactly the effect of the propagator $e^{i L_t t}$, where the Liouville operator $i L_t$ is, simply,
\[
i L_t = \frac{\tr{\vt p}}{m}\diff{}{\vt r}.
\]

In the second hypothetical case, referred to as the acceleration case, we consider a Hamiltonian $\mathcal{H}_a = U(\vt r, \vt q)$. The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
\begin{align*}
&{\vt r}(t) = {\vt r}_0 \\
&{\vt p}(t) = {\vt p}_0 + t \vt F_0 \\
&{\vt q}(t) = {\vt q}_0 \\
&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C_0 \vt \tau_0
\end{align*}

Note that $\vt r$ and $\vt q$ remain unaltered, thus keeping $\vt F$, $\mt C$, and $\vt \tau$ constant as well (hence the subscript $0$ assigned to them). This solution clearly conserves $\mathcal{H}_a$, and marks the effect of a propagator $e^{i L_a t}$, obtained from the operator
\[
i L_a = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
\]

Finally, the third hypothetical case consists of a Hamiltonian $\mathcal{H}_k = (1/2) I_k \omega_k^2$. Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation. The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = (1/2) \omega_k {\mt B}_k \vt q$, and $\dot{\vt \pi} = (1/2) \omega_k {\mt B}_k \vt \pi$. A special feature of these equations makes them analytically solvable. It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well. This decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form. The following analytical solution takes place:\cite{Miller2002}
\begin{align}
&{\vt r}(t) = {\vt r}_0 \\
&{\vt p}(t) = {\vt p}_0 \\
&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \mt B_k \vt q_0 \\
&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \mt B_k \vt \pi_0
\end{align}

These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{i L_k t}$ derived from the operator
\[
i L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{{\mt B}_k} \diff{}{\vt \pi} \right).
\]

Now note that the Liouville operator corresponding to the original case can be written as
\begin{align}
\label{eq:full_operator}
i L_{NVE} = i L_t + i L_a + \sum_{k=1}^3 i L_k,
\end{align}

The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic geometry of Eq.~\ref{eq:EDO_system_alternative}. Such formula is
\[
e^{i L_A t + i L_B t} \approx \left( e^{i L_B \frac{\Delta t}{2}} e^{i L_A \Delta t} e^{i L_B \frac{\Delta t}{2}} \right)^N + \mathcal{O}(\Delta t^2),
\]
where $\Delta t = t/N$ is the time step size and $N$ is the number of steps. In the next section we show the factorization schemes we considered for generating the dynamics in both the NVE and NVT ensembles.

\section{Molecular Dynamics}
\label{sec:moleculardynamics}

\subsection{Microcanonical Ensemble}

The splitting of the Liouville operator (Eq.~\ref{eq:full_operator}) we employ in this paper is similar to that proposed by Miller \textit{et al}.\cite{Miller2002}, but extended to include translation and linear acceleration. It is

\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{i L_{NVE} \Delta t} = e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L_a \frac{\Delta t}{2}},
\end{equation}

where $L_r = \sum_{k=1}^3 L_k$ is the Liouville operator for a complete rotation and, $i L_a $, $i L_k $ as well as $i L_t $ retain its previous definitions. Recall that each exponential operator in Eq.~\ref{eq:trotter_splitting_NVE} will act having the result of the preceding operator (the one at its right hand side) as initial condition. Following Ref.~\onlinecite{Miller2002} once again, the splitting employed for the rotational part is

\begin{equation}
\label{eq:splitting_rot}
e^{i L_r \Delta t} = \left( e^{i L_3 \frac{\delta t}{2}} e^{i L_2 \frac{\delta t}{2}} e^{i L_1 \delta t} e^{i L_2 \frac{\delta t}{2}} e^{i L_3 \frac{\delta t}{2}} \right)^n
\end{equation}

where $\delta t = \nicefrac{\Delta t}{n}$. This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step to improve accuracy. An alternative to a further factorization of Eq.~\ref{eq:splitting_rot} with higher values of $n$ could be the use of a higher-order scheme, as the one due to Yoshida \cite{Yoshida1990} and Suzuki\cite{Suzuki1991a,Suzuki1991b}, commonly used in Nos\'{e}-Hoover chain dynamics. 
The factorization of the operator is expressed as follows

\begin{equation}
\label{eq:ys_rot}
e^{i L_{r} \Delta t} = \prod_{j=1}^{n_{sy}} e^{i L_{r} \omega_j \Delta t}
\end{equation}

where $\omega_j$ are employed to weight the time step and $n_{sy}$ determines the order of the method. Note that we are not improving the order of precision for the integration of the translational degrees of freedom, which would be computationally more demanding. Nevertheless, we would like to investigate   how the performance of the integrator as a whole is affected by using a higher-order scheme for the rotational part or the operator. For that purpose, we have considered $n = 1$ with $n_{ys} = 1 $ (zero-order) and $n_{sy} = 7$ (sixth-order).

\subsection{Canonical Ensemble}
\label{sec:canonical}
The linear momentum equation for a rigid body considering a Nos\'{e}-Hoover thermostat is
\begin{align*}
\dot{\vt p} &= \vt F^d = {\vt F} - \frac{p_{\eta}}{Q} {\vt p} = \sum_{j=1}^{n_p} {\vt F}_j - \frac{p_{\eta}}{Q} \sum_{j=1}^{n_p} {\vt p}_j \\
&= \sum_{j=1}^{n_p} \left( {\vt F}_j - \frac{p_{\eta}}{Q} {\vt p}_j \right)
\end{align*}

Thus, the thermostat affects each particle $j$ individually. It is then intuitive that the resultant torque is affected by the thermostat according to
\begin{align*}
\vt \tau^d &= \sum_{j=1}^{n_p} {\vt \delta_j} \times \left( {\vt F_j}  - \frac{p_{\eta}}{Q} {\vt p}_j \right) = \\
&= \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j} - \frac{p_{\eta}}{Q} \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt p}_j = \\
&= \vt \tau - \frac{p_{\eta}}{Q} \vt L,
\end{align*}
where $\vt L$ is the angular momentum of the rigid body expressed in the space-fixed frame of reference. Thus, $\vt L = \tr{\mt A}{\mt I}{\vt \omega}$. Resorting to Eqs.~\ref{eq:factorization_of_A} and \ref{eq:conj_momentum}, we observe that
\[
\vt L = \tr{\mt C}{\mt B}{\mt I}{\vt \omega} = \frac{1}{2} \tr{\mt C}{\vt \pi}
\]
Therefore, replacing $\vt \tau$ by $\vt \tau^d$ in the quaternion momentum equation, we have its thermostated version as
\[
\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau - \frac{p_{\eta}}{Q} \vt \pi.
\]
Finally, the Nos\'{e}-Hoover chain thermostat applied to a system of $N$ rigid bodies becomes:

\begin{align*}
&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} & i = 1,\cdots,N \\
&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i & i = 1,\cdots,N \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i & i = 1,\cdots,N \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q} \vt \pi_i & i = 1,\cdots,N \\
&\dot{\eta}_j = \frac{p_{\eta_j}}{Q_j} & j = 1,\cdots,M \\
&{\dot p}_{\eta_j} = G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} & j = 1,\cdots,M-1 \\
&{\dot p}_{\eta_M} = G_M
\end{align*}
where
\begin{align*}
&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - 6 N k_B T \\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_B T &j = 2,\cdots,M \\
&Q_1 = 6 N k_B T t_d^2 \\
&Q_j = k_B T t_d^2 &j = 2,\cdots,M
\end{align*}
These equations have the conserved energy $\mathcal{H}^\prime$ given by
\begin{align*}
\mathcal{H}^\prime = \mathcal{H}(\vt r, \vt p, \vt q, \vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_k} + 6Nk_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j
\end{align*}

In these equations, $M$ is the number of thermostats in the chain, $T$ is the set-point temperature, $k_B$ is Boltzmann's constant, $6N$ is the number of degrees of freedom, and $t_d$ is the characteristic time period of the thermostat chain.  $\mathcal{H}$ was defined in section~\ref{sec:hamiltonian}.

As was done in the Hamiltonian case, the evolution of the extended phase space vector $ f(\vt r, \vt p, \vt q, \vt \pi, \eta_1,.. \eta_M, p_{\eta_1},..,p_{\eta_M}) $ is given by the relation $f(t) = e^{i L t}f_0$ in the context of the generalized Liouville theorem. This implies that we can employ the Trotter theorem to devise a strategy for integrating the equations of motion.

We consider first the usual separation of the non-Hamiltonian component of the equations of motion from the Hamiltonian component. Then, the Liouville operator can be written as follows
\[
i L = i L_{NHC} + i L_{NVE}
\]
where $i L_{NVE}$ has been previously defined (Eq.~\ref{eq:full_operator}).
It follows that the operator $ i L_{NHC} $ is given by

\begin{equation}
\begin{split}
& i L_{NHC} = -\sum_{i=1}^{N} \frac{p_{\eta_1}}{Q_1} \vt p_i \diff{}{\vt p_i} - \sum_{i=1}^{N} \frac{p_{\eta_1}}{Q_1} \vt \pi_i \diff{}{\vt \pi_i} \\
& + \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}} {Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right]\\
\end{split}
\end{equation}
\\
with $p_{\eta_{M+1}}/Q_{M+1} = 0$
\\

The part of the operator given by 
\begin{align*}
&\left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}} \\
\end{align*}

can be evaluated analytically, as pointed out by Martyna \textit{et al.}\cite{Martyna1996}, using the following equation:

\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} \leftarrow p_{\eta_j} + \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \phi\left( \frac{p_{\eta_{j+1}}}{Q_{j+1}} \Delta t \right) \Delta t \\
\end{split}
\end{equation}
\\
where $\phi(x) = 1-e^{-x}/x = \sum_{n=0}^\infty (-1)^n x^n/(n+1)!$. The function $\phi(x)$ can be evaluated by

\begin{equation}
\label{eq:phi}
\phi(x) = \begin{cases}
\frac{1-e^{-x}}{x} & \text{if} \; |x| > 10^{-4} \\
1 - \frac{x}{2}\left[1 - \frac{x}{3}\left(1 - \frac{x}{4}\right) \right] & \text{if} \; |x| \leq 10^{-4}
\end{cases}
\end{equation}

The NHC operator is factorized as follows
\\

\begin{widetext}
\begin{equation}
\begin{split}
e^{iL_{NHC} \frac{\Delta t}{2}} =  &\prod_{j=M}^{1} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
&  \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{2} \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right]  \prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt p_i \diff{}{\vt p_i}\right] \prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt \pi_i \diff{}{\vt \pi_i}\right]  \\
& \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]  \\
\end{split}
\end{equation}
\end{widetext}

The exponential factors in the second line of the equation above are evaluated according to

\begin{equation}
\begin{split}
&\vt \pi_i \leftarrow \vt \pi_i  exp\left[-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right]\\
&\vt p_i \leftarrow \vt p_i  exp\left[-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right]\\
&\eta_j  \leftarrow \eta_j + \frac{p_{\eta_j}}{ Q_j} \frac{\Delta t}{2} 
\end{split}
\end{equation}

Retaining the NVE factorization scheme, Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as

\begin{equation}
\label{eq:trotter_splitting_NHC}
e^{i L \Delta t} =
e^{i L_{NHC} \frac{\Delta t}{2}} e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t}  e^{i L_a \frac{\Delta t}{2}} e^{i L_{NHC} \frac{\Delta t}{2}},
\end{equation}
It is worth noting that analogous analytical solutions exists also for the following equations
\begin{align*}
&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i  \\
&\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q_1} \vt \pi_i  \\
\end{align*}
The solutions for $\vt p_i$ and $\vt \pi_i $  are then

\begin{equation}
\label{eq:p}
\begin{split}
{\vt p}_i \leftarrow {\vt p}_i + \left({\vt F}_i - \frac{p_{\eta_1}}{Q_1} {\vt p}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t \\
\end{split}
\end{equation}

\begin{equation}
\label{eq:pi}
\begin{split}
{\vt \pi}_i \leftarrow {\vt \pi}_i + \left(2 \tr{\vt \tau} \tr{\mt C} - \frac{p_{\eta_1}}{Q_1} {\vt \pi}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t \\
\end{split}
\end{equation}

with $\phi(x)$ given by Eq.~\ref{eq:phi}.
\\

Therefore, one of the aims is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when also Eqs.~\ref{eq:p} and~\ref{eq:pi} are considered. For that, let as define the modified operators ($\ast$) $ i L_a $ and $ i L_{NHC} $ as follows:

\begin{equation}
\begin{split}
i L^\ast_a = &i L_a -\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt p_i\diff{}{\vt p_i} - \sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt \pi_i  \diff{}{\vt \pi_i} =\\
             &\sum_{i=1}^{N}(\tr{\vt F_i} - \frac{p_{\eta_1}}{Q_1}\vt p_i) \diff{}{\vt p_i} + \sum_{i=1}^{N}(2 \tr{\vt \tau_i} \tr{\mt C_i} -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i)  \diff{}{\vt \pi_i}.
\end{split}
\end{equation}

\begin{equation}
\begin{split}
i L^\ast_{NHC} = \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right] 
\end{split}
\end{equation}

Finally, we define the following factorization scheme:
\begin{widetext}
\begin{equation}
\begin{split}
e^{iL^\ast_{NHC} \frac{\Delta t}{2}} =  &\prod_{j=M}^{1} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
& \times\prod_{j=1}^{M} exp\left[-\frac{\Delta t}{2}  \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right]   \\
& \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]  \\
\end{split}
\end{equation}
\end{widetext}

\begin{equation}
\label{eq:trotter_splitting_*}
\begin{split}
e^{i L \Delta t} = &e^{i L^\ast_{NHC} \frac{\Delta t}{2}} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L^\ast_{NHC} \frac{\Delta t}{2}},
\end{split}
\end{equation}

\subsection{Numerical Stability}
\label{sec:performance}
In this section, we resume some performance aspects of the integration schemes when applied to a molecular system in the liquid state. Given its practical importance, we simulated 356 TIP3P\cite{Price2004} water molecules at a density of 1 $g/cm^3$. The intermolecular interactions were truncated at 9 {\AA}. The damped shift force (DSF) method\cite{Fennell2006} was used to treat the coulombic interactions while  the "shifted-force potential"\citep{Allen1989} method was applied for the 12-6 Lennard-Jones (LJ) interactions.

We used Eq.~\ref{eq:performance} to evaluate the stability of the numerical solvers. It measures the  the average deviation of the conserved quantity, i.e. total energy, from its initial value $E_0$ after a certain number of time steps $N_{steps}$. Both the instantaneous fluctuations of energy, given by the term inside the summation, and the long-time conservation of energy, $D E$, are monitored.

\begin{equation}
\label{eq:performance}
D E =  \frac{1}{N_{steps}} \sum_{k=1}^{N_{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|
\end{equation}

Recall that the method in the NVE ensemble is a mathematical simplification of the equations of motion, so the trajectory must coincide, out of round-off errors, with that generated using the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm introduced by Miller~\textit{et al.}\cite{Miller2002}. In Fig.~\ref{fig:miller1fs} we plot for both methods the instantaneous fluctuations of energy with $n_{ys} = 1 $ and a time step of 1 fs, where it can be seen that the behaviour is essentially the same. The simulation with the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm was conducted using the LAMMPS\cite{Plimpton1995} software package, which does not have the shifted-force potential implemented, so no shifting was applied to the LJ interactions in this case.

\begin{center}
  \includegraphics{miller1fs}
  \captionof{figure}{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and Ref.\citenum{Miller2002} (dotted line). $n : 1 $, $n_{ys}:1$, $\Delta t : 1 fs$}
    \label{fig:miller1fs}
\end{center}

In Table~\ref{table:nve} we compare the energy drift when using $n_{ys}:1$ and $n_{ys}:7$ in the rotational operator, Eq.~\ref{eq:ys_rot}. As expected, for a small time step of 1 fs, that further factorization is worthless. As the time step is increased, more accurate trajectories are achived with $n_{ys} = 7 $. Nevertheless, the improvement is only marginal and all trajectories are stables exhibiting a small long-time energy drift. Thus, we proceed to analyze the NVT dynamics considering $n_{ys} = 1$.

\begin{table}[h]
\setlength{\tabcolsep}{7pt}
\caption{Average deviation of the conserved quantity ($DE$) from liquid water simulation with NVE integration with $n_{ys}:1$ and $n_{ys}:7$  }
\centering % centering table
\begin{tabular}{| c  c  c |}  
\hline
&  \multicolumn{2}{c|}{ $D E$ }\\
\cline{2-3}
$\Delta t$ &$n_{ys}:1$ &$n_{ys}:7$ \\
\hline % inserts single-line
 1fs & 0.000055 & 0.000074\\
 2fs & 0.00048 & 0.00031 \\
 3fs & 0.0012 & 0.0011\\
 \hline
\end{tabular}
\label{table:nve}
\end{table}

It is not the purpose of this work to achieve an optimal factorization of the NVT schemes, but rather to show that a unique chain thermostat, while providing the proper sampling of the NVT ensemble, is capable of maintaining the average temperature at its set point. In order to use a constant value for the smaller time step in the NHC and NHC* operators, we implemented a multiple time-scale scheme, like the RESPA methodology\cite{Tuckerman1992}. Recalling that in the NVT dynamics the conserved quantity is the extended energy $\mathcal{H}^\prime$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the temperatures obtained in each case.

\begin{table}[h]
\setlength{\tabcolsep}{7pt}
\begin{threeparttable}
\caption{Average deviation of the conserved quantity ($DE$) and average temperatures from liquid water simulation with NVT integration \tnote{a}\tnote{b}} 
\label{table:denvt}

\centering % centering table
\begin{tabular}{|c c c c c |}  
\hline
& \multicolumn{2}{c}{Eq.~$\ref{eq:trotter_splitting_NHC}$} &\multicolumn{2}{c|}{Eq.~$\ref{eq:trotter_splitting_*}$} \\
\cline{2-5}
$\Delta t$ &$D E$ &$T/K$ &$D E$ &$T/K$\\
\hline % inserts single-line
 1fs  & 0.000043 & 298.16  & 0.000052  & 298.15 \\

 2fs  & 0.00013 & 298.23  & 0.00017 & 298.06  \\

 3fs  & 0.00089 & 298.17  & 0.0013 & 298.12  \\

 \hline
\end{tabular}
\begin{tablenotes}
\item[a] standard deviation of the mean temperature for all cases is $\sigma = 0.037$
\item[b] smaller time step in the NHC and NHC* operatores is 0.25 fs 
\end{tablenotes}
\end{threeparttable}
\end{table}

It can be seen that both integrators represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*} systematically achieve the temperature set point, regardless of the time step size. In addition, the energy drift is, at its worst, around $1 \%$.

\subsection{Sampling Validation}

We present the results from a simple method that serves for checking if an algorithm generates the correct ensemble distribution. The procedure, introduced by Shirts ~\cite{Shirts2013}, relies on a quantitative statistical analysis of fitting and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too. We report the results of the last approach as it doesn't depend on binning. However, since it is instructive to have a graphical insight, we also show one of the plots of the linear fitting.

Recall that in the NVT ensemble the probability of observing an energy E is given by:
\[
P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right)
\]
where $ \beta $ is the Boltzmann's constant, $ \Omega(E) $ is the density of states which does not depend on $ \beta $ , and $ Q(\beta) $  is the canonical partition function, that is linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$

Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get:

\begin{equation}
\label{eq:probability_ratio}
\frac{P(E/\beta_2)}{P(E/\beta_1)} = exp\left([\beta_2 A_2 - \beta_1 A_1] - [\beta_2 - \beta_1]E\right)
\end{equation}

Taking the logarithm of Eq.~\ref{eq:probability_ratio}, we obtain:
\begin{equation}
\label{eq:ln_probability_ratio}
ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = [\beta_2 A_2 - \beta_1 A_1] - [\beta_2 - \beta_1]E
\end{equation}

Eq.~\ref{eq:ln_probability_ratio} is the basis for the statistical analysis in the ensemble validation, which determines how many standard deviations the calculated slope $ \beta_2 - \beta_1 $ is from the specified one. Note that it is independent of the unknown free energies. For a guidance on how to choose the temperature gap the reader is referred to the aforementioned paper. Given that the potential and kinetic energies are separable, the analysis can be made separately:

\begin{equation}
\label{eq:probability_ratio_pot}
\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  exp\left([-\beta_2 - \beta_1]E_{pot}\right)
\end{equation}

\begin{equation}
\label{eq:probability_ratio_kin}
\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  exp\left([-\beta_2 - \beta_1]E_{kin}\right)
\end{equation}

In Fig.~\ref{fig:checkensemble} we show the linear graph for the potential energies sampled using Eq.~\ref{eq:trotter_splitting_NHC}. It can be seen that the algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies. We arrived at the same conclusions for the other approach given by Eq.~\ref{eq:trotter_splitting_*}.

In table~\ref{table:ensemblevalidation} we examine the results of the maximum likelihood approach. It can be seen that the deviations from the true slopes are less than $1\sigma$. This means that the sampled values do not deviate from the correct distribution to a statistically noticeable level.

\begin{center}
\includegraphics{checkensemble}
\captionof{figure}{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols). Simulation of liquid water using Eq. \ref{eq:trotter_splitting_NHC} $\Delta t : 1 fs$. $T_1:296.01$, $T_2 = 304.98$ }
\label{fig:checkensemble}
\end{center}

\begin{widetext}
\begin{table}[h]
\setlength{\tabcolsep}{7pt}
\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach}
\label{table:ensemblevalidation}
\centering % centering table
\begin{tabular}{l c  c  c  c  c  c }  
\hline
& & & \multicolumn{2}{c}{true $\Delta T$ = 8.970 K}\\
\cline{4-7}
 &true $T_{low} = 296.01 K$ &true $T_{high} = 304.98 K$& \multicolumn{2}{c}{potential} & \multicolumn{2}{c}{kinetic}\\
\hline
Integrator &$T_{low}$  &$T_{high}$ &$\Delta T$ &$\sigma$ deviation &$\Delta T$&$\sigma$ deviation\\
\hline % inserts single-line
 Eq.~$\ref{eq:trotter_splitting_NHC}$   & 296.03 $\pm$ 0.0286 & 304.99 $\pm$ 0.0297 & 9.016 $\pm$ 0.117 & 0.39 & 8.932 $\pm$ 0.045 & 0.84 \\
  \\ [0.5ex]
 Eq.~$\ref{eq:trotter_splitting_*}$ & 296.03 $\pm$ 0.0286 & 304.99 $\pm$ 0.0297 & 8.956 $\pm$ 0.103 & 0.14 & 8.958 $\pm$ 0.053 & 0.23\\
  \hline
\end{tabular}
\end{table}
\end{widetext}

In the NVE case, each Liouvillian operator is derived from part of the Hamiltonian, then each factor in the propagator is symplectic and, as a consequence, preserves the phase space volume. Recall that the inverse statement is not necessarily true. Now it is widely recognized the importance of the symplectic property for numerical stability,\cite{Skeel1997} however it is restricted to hamiltonian systems. For non-Hamiltonian equations of motion, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which possess the properties of reversibility and measure-preserving. In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to more ergodic trajectory behaviour. The invariant-measure for the  Nos\'{e}-Hoover chain system is given by
\begin{align*}
\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ]
\end{align*}

It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*} do not preserve the invariant-measure. Nevertheless, those numerical schemes successfully achieve the set point temperature and the long-term energy conservation is acceptable.

Note that the term that premultiplies the extended phase space volume in the invariant-measure is a function of the variable $\eta$. Given that the time evolution of the physical variables along with the thermostat momentum does not depend on $\eta$ we conclude that in this case the measure-preserving condition for every individual operator can actually be relaxed.
\section{Pressure of a System of Rigid Bodies}
\label{sec:pressure}
We now derive an expression for calculating the pressure of a rigid body system under periodic boundary conditions. For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies. This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended.

The potential energy per unit cell of a system with $N$ distinct rigid bodies (indexed from $1$ to $N$) and periodic boundary conditions is
\begin{equation}
\label{eq:U_pairwise_pbc}
U({\vt r}^N, {\vt q}^N) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
\end{equation}
where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
\begin{equation}
\label{eq:u_rigid_body_pair}
u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
\end{equation}

In the equation above, $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$. The dependence of this potential on the types of these two atoms is omitted for simplicity. The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
\[
{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
\]
The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
\[
P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
\]
The first term in the expression above accounts for the ideal gas contribution with the corresponding subtraction of three DOF given the conservation of linear momentum. 
Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
\begin{equation}
\begin{split}
\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right. \\ \left. \diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
\end{split}
\end{equation}

Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$. Therefore,
\[
\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \diff{u}{{\vt r}_{ij}} \cdot {\vt r}_{ij}.
\]

We can now use Eq.~\ref{eq:u_rigid_body_pair} to obtain the term inside the double summation above. By the chain rule, we have
\[
\diff{u}{{\vt r}_{ij}} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \diff{\vt r}{\vt r_{ij}} \diff{\upsilon}{\vt r},
\]
where $\vt r = {{\vt R}_i}_k - {{\vt R}_j}_l$. Notice that, inside the double summation above, the first derivative is the identity matrix and the second one is $-{\vt f}_{ijkl}$, with this force being the action of the $l$-th particle of body $j$ over the $k$-th particle of body $i$. Therefore, the equation above results in $-\vt F_{ij}$, with this force being the resultant action of body $j$ over body $i$. Finally, we conclude that

\begin{equation}
\label{eq:pressure}
P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
\end{equation}

This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs. Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.

Now we compare the results using Eq.~\ref{eq:pressure} with those from the simulation performed with LAMMPS, which uses a numerical estimation of the constraint forces in the pressure calculation. The simulations were run in the NVE ensemble, for which neither shifting to the LJ forces nor analytical tail corrections were considered. Figure~\ref{fig:pressure} displays the corresponding pressure distributions.

%\begin{center}
%  \label{fig:pressure}
%  \includegraphics{pressure2}
%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars). Simulation of liquid water with NVE integration. $\Delta t : 1 fs$  }
%\end{center}

\begin{center}
\label{fig:pressure}
\includegraphics{pressure3}
\captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS (unfilled bars). Simulation of liquid water with NVE integration. $\Delta t : 1 fs$  }
\end{center}

\subsection{To do}

Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
\[
W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
\]

\appendix

\section{\label{sec:quat_from_A}Obtaining the quaternion components from a rotation matrix}

The method described here was introduced by Shepperd.\cite{Shepperd1978} As one can observe in Eq.~\ref{eq:A_from_q}, the three main diagonal components of $\mt A$ and the fact that $\lVert \vt q \lVert^2 = 1$ provide a linear system for the squares of the quaternion components whose solution is
\begin{align*}
4 q_0^2 &= 1 + tr(\mt A) \\
4 q_k^2 &= 1 + A_{k,k} - tr(\mt A)
\end{align*}
where $k = 1,2,3$ and $tr(\mt A)$ is the trace of matrix $\mt A$. Since both $\vt q$ and $- \vt q$ correspond to the same rotation matrix, the sign the largest component\cite{Shepperd1978} can be made positive and the others can be obtained from the off-diagonal entries of $\mt A$, which give
\begin{align*}
4 q_0 q_1 &= A_{2,3} - A_{3,2} \\
4 q_0 q_2 &= A_{3,1} - A_{1,3} \\
4 q_0 q_3 &= A_{1,2} - A_{2,1} \\
4 q_k q_l &= A_{k,l} + A_{l,k} \\
\end{align*}
where $k,l = 1,2,3$ with $k \neq l$.

\section{\label{sec:auxiliary_math}Algebraic relations involving unit quaternions}

Let us analyze the matrix $\mt B$ defined as in Eq.~\ref{eq:def_B_and_C}. Note that $\tr{\mt B}\vt q = \vt 0$, where $\vt 0$ is a null vector.\cite{Haug1989, Shuster1993, Dichmann1999} This means that all columns of $\mt B$ are orthogonal to $\vt q$. As these columns have unit norm and are also orthogonal to each other, they form an orthonormal basis for the hyperplane orthogonal to $\mt q$, which is a subspace of $\mathbb{R}^4$. Therefore, the general operation $\vt y = \mt B \vt x$ transforms a vector $\vt x \in \mathbb{R}^3$ into a quaternion $\vt y \in \mathbb{R}^4$, but restricted to the mentioned hyperplane. Other two identities that hold for $\mt B$ are\citep{Haug1989}
\begin{align*}
\tr{\mt B}\mt B &= \lVert \vt q \lVert ^2{\mt 1}_3 \text{ and} \\
\mt B\tr{\mt B} &= \lVert \vt q \lVert ^2{\mt 1}_4 - {\vt q}\tr{\vt q}.
\end{align*}

If $\lVert \vt q \lVert^2 = 1$, then the operation $\vt y = \mt B \vt x$ can always be reverted by $\tr{\mt B}$. This is not true for a general operation in the opposite direction, $\vt x = \tr{\mt B} \vt y$. Besides unit norm, it is necessary that $\tr{\vt q}\vt y = 0$ (i.e. orthogonality of $\vt y$ and $\vt q$) for the operation to be reversible by $\mt B$. In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999}

All the conclusions drawn thus far for matrix $\mt B$ are equally valid for matrix $\mt C$.

As the columns of $\mt B$ are permutations of the Euler parameters, one can write $\mt B = [\begin{array}{ccc}{\mt B}_1{\vt q} & {\mt B}_2{\vt q} & {\mt B}_3{\vt q}\end{array}]$, where $\mt B_1$, $\mt B_2$, and $\mt B_3$ are permutation matrices given by
\[
{\mt B}_1 = \left[ \begin{array}{rrrr}
 0 & -1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 &  1 \\
 0 &  0 & -1 &  0 \\
\end{array} \right], \;
{\mt B}_2 = \left[ \begin{array}{rrrr}
 0 &  0 & -1 &  0 \\
 0 &  0 &  0 & -1 \\
 1 &  0 &  0 &  0 \\
 0 &  1 &  0 &  0 \\
\end{array} \right],
\]\[
\text{and }{\mt B}_3 = \left[ \begin{array}{rrrr}
 0 &  0 &  0 & -1 \\
 0 &  0 &  1 &  0 \\
 0 & -1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{array} \right].
\]

All these permutation matrices are skew-symmetric (i.e. $\tr{\mt B_k} = -\mt B_k$) and also satisfy the relations\cite{Dichmann1999}
\begin{equation}
\label{eq:BB_products}
\begin{aligned}
\tr{\mt B_1}{\mt B_1} &= \tr{\mt B_2}{\mt B_2} = \tr{\mt B_3}{\mt B_3} = \mt 1_3 \\
\tr{\mt B_1}{\mt B_2} &= -\tr{\mt B_2}{\mt B_1} = -\tr{\mt B_3} \\
\tr{\mt B_2}{\mt B_3} &= -\tr{\mt B_3}{\mt B_2} = -\tr{\mt B_1} \\
\tr{\mt B_3}{\mt B_1} &= -\tr{\mt B_1}{\mt B_3} = -\tr{\mt B_2}
\end{aligned}
\end{equation}

The product of $\mt B$ with an arbitrary vector $\vt x$ results in
\begin{equation}
\label{eq:product_B_vector}
\mt B \vt x = \left[ \begin{array}{ccc} \mt B_1 \vt q & \mt B_2 \vt q & \mt B_3 \vt q \end{array}\right] \left[ \begin{array}{c} x_1 \\ x_2 \\ x_3 \end{array} \right] = \sum_{k=1}^3 x_k {\mt B}_k \vt q.
\end{equation}

Similarly, the product of $\tr{\mt B}$ with an arbitrary quaternion $\vt y$ results in
\begin{equation}
\label{eq:vector_entries}
\tr{\mt B}\vt y =
\left[\begin{array}{c}
\tr{\vt q}\tr{\mt B}_1 \\
\tr{\vt q}\tr{\mt B}_2 \\
\tr{\vt q}\tr{\mt B}_3 \\
\end{array}\right]\vt y = 
\left[\begin{array}{c}
\tr{\vt q}\tr{\mt B}_1\vt y \\
\tr{\vt q}\tr{\mt B}_2\vt y \\
\tr{\vt q}\tr{\mt B}_3\vt y \\
\end{array}\right] = 
\left[\begin{array}{c}
\tr{\vt y}{\mt B}_1\vt q \\
\tr{\vt y}{\mt B}_2\vt q \\
\tr{\vt y}{\mt B}_3\vt q \\
\end{array}\right].
\end{equation}

Let us now consider the product $\tr{\mt B}\dot{\mt B}$, where $\dot{\mt B}$ is the time derivative of $\mt B$. Because the permutation matrices are constant, we can express $\dot{\mt B} = [\begin{array}{ccc}{\mt B}_1\dot{\vt q} & {\mt B}_2\dot{\vt q} & {\mt B}_3\dot{\vt q}\end{array}]$. By carrying out the multiplication and applying the relations of Eq.~\ref{eq:BB_products}, we observe that
\[
\tr{\mt B}\dot{\mt B} = \left[
\begin{array}{ccc}
0 & -\tr{\vt q} \tr{\mt B_3}\dot{\vt q} & \tr{\vt q} \tr{\mt B_2}\dot{\vt q} \\
\tr{\vt q} \tr{\mt B_3}\dot{\vt q} & 0 & -\tr{\vt q} \tr{\mt B_1}\dot{\vt q} \\
-\tr{\vt q} \tr{\mt B_2}\dot{\vt q} & \tr{\vt q} \tr{\mt B_1}\dot{\vt q}  & 0
\end{array}
\right],
\]
where the diagonal entries are zero as a consequence of the orthogonality between $\vt q$ and $\dot{\vt q}$. By contrasting this result with Eq.~\ref{eq:operator_S}, we draw the conclusion that
\begin{equation}
\label{eq:relation_B_qdot}
\tr{\mt B}\dot{\mt B} = {\mt S}\left( \tr{\mt B}\dot{\vt q} \right).
\end{equation}

Matrix $\mt C$ and its time-derivative $\dot{\mt C}$ can also be expressed as $[\begin{array}{ccc}{\mt C}_1{\vt q} & {\mt C}_2{\vt q} & {\mt C}_3{\vt q}\end{array}]$ and $[\begin{array}{ccc}{\mt C}_1\dot{\vt q} & {\mt C}_2\dot{\vt q} & {\mt C}_3\dot{\vt q}\end{array}]$, respectively, where $\mt C_1$, $\mt C_2$, and $\mt C_3$ are
\[
{\mt C}_1 = \left[ \begin{array}{rrrr}
 0 & -1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 & -1 \\
 0 &  0 &  1 &  0 \\
\end{array} \right], \;
{\mt C}_2 = \left[ \begin{array}{rrrr}
 0 &  0 & -1 &  0 \\
 0 &  0 &  0 &  1 \\
 1 &  0 &  0 &  0 \\
 0 & -1 &  0 &  0 \\
\end{array} \right],
\]\[
\text{and }{\mt C}_3 = \left[ \begin{array}{rrrr}
 0 &  0 &  0 & -1 \\
 0 &  0 & -1 &  0 \\
 0 &  1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{array} \right].
\]

A special connection between the permutation matrices for $\mt B$ and $\mt C$ is the fact that $\tr{\mt B_i}{\mt C_j}$ is symmetric for any combination of $i$ and $j$. Thus, each entry of $\tr{\mt B}\dot{\mt C}$ is
\begin{align*}
\{\tr{\mt B}\dot{\mt C}\}_{ij} &= \tr{\vt q}\tr{\mt B}_i{\mt C}_j\dot{\vt q} = \tr{\vt q}\tr{\mt C}_j{\mt B}_i\dot{\vt q} = \{\tr{\mt C}\dot{\mt B}\}_{ji} = \\
&= \{\tr{\dot{\mt B}}\mt C\}_{ij}.
\end{align*}

This means that $\tr{\mt B}\dot{\mt C} = \tr{\dot{\mt B}}{\mt C}$. Proceeding further, we observe that
\begin{align*}
\mt A \tr{\dot{\mt A}} &= \tr{\mt B}\mt C \tr{(\tr{{\mt B}}\dot{\mt C} + \tr{\dot{\mt B}}{\mt C})} = \tr{\mt B}\mt C \tr{(2 \tr{\dot{\mt B}}{\mt C})} = \\
&= 2 \tr{\mt B}\dot{\mt B},
\end{align*}
where we have applied the relations $\mt C \tr{\mt C} = \mt 1_4 - \vt q\tr{\vt q}$ and $\tr{\vt B}\vt q = \vt 0$. Finally, resorting to Eqs.~\ref{eq:relation_A_omega} and \ref{eq:relation_B_qdot}, the result above converts to
\begin{equation}
\vt \omega = 2\tr{\mt B}\dot{\vt q}.
\end{equation}

Demonstration of Euler equation:

\begin{align*}
\diff{(\tr{\mt B}\vt \pi)}{t} &= \tr{\dot{\mt B}}\vt \pi + \tr{\mt B}\dot{\vt \pi} = \\
&=\tr{\dot{\mt B}}{\mt B}\tr{\mt B}{\vt \pi}  + \dfrac{1}{2} \tr{\mt B}\mt \Omega \vt \omega + 2 \tr{\mt B}\mt C \vt \tau = \\
&= -\frac{1}{2}{\mt S}(\vt \omega)\tr{\mt B}{\vt \pi} + \dfrac{1}{2} {\mt S}(\tr{\mt B}\mt \pi) \vt \omega + 2 \mt A \vt \tau = \\
&= - \vt \omega \times (\tr{\mt B}{\vt \pi}) + 2 {\vt \tau}^\prime
\end{align*}

As $\tr{\mt B}\vt \pi = 2 \mt I \vt \omega$, the result above becomes
\[
\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = {\vt \tau}^\prime
\]

\section{\label{sec:Diff_Rules}Differentiation rules with vectors}

Along the paper, the differentiation of scalar functions with respect to vectors are done directly using the following rules:
\[
\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a
\]
and
\[
\diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = 2 \mt A \vt x,
\]
where $\vt a$ is vector and $\mt A$ is a matrix, both independent of $\vt x$.

\section{\label{sec:Diff_PotEng}Potential energy differentiation}

Care must be taken when differentiating the potential energy $U$ with respect to $\vt q$ because the Euler parameters are not all independent. Schay demonstrated in Ref.~\cite{Schay1995} that, when differentiating a scalar function $f$ with respect to a vector variable $\vt x$, subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ is given by
\[
\diff{f}{\vt x} = (\vt 1 - \vt n \tr{\vt n})\left( \diff{f}{\vt x} \right)^\ast,
\]
where $\vt n$ is the unit vector proportional to $(\partial g/\partial \vt x)^\ast$, with the asterisk denoting a differentiation performed without considering the interdependence of the entries of $\vt x$. We note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\cite{Schay1995}, where the Jacobian convention is used. In the  case of Euler parameters,
\[
g(\vt q) = \tr{\vt q}\vt q - 1 \; \rightarrow \; \left(\diff{g}{\vt q}\right)^\ast = 2 \vt q.
\]

Hence, the unit four-dimensional vector proportional to $(\partial g/\partial \vt x)^\ast$ is exactly $\vt q$, which means that the constrained derivative of $U$ is given by
\begin{equation}
\label{eq:diff_projection}
\diff{U}{\vt q} = (\mt 1_4 - {\vt q}\tr{\vt q})\left( \diff{U}{\vt q} \right)^\ast.
\end{equation}

Therefore, the constrained derivative is the projection of the unconstrained one in the hyperplane orthogonal to $\vt q$. Recently, Nielsen and Krenk\cite{Nielsen2012} employed a Lagrange multiplier approach to demonstrate the need of carrying out such projection. Schay\cite{Schay1995, Schay1998} discussed on the equivalence of the two approaches.

By applying the chain rule, we have
\[
\left( \diff{U}{\vt q} \right)^\ast = \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast \diff{U}{\vt R_j} = - \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast {\vt F_j},
\]
where $\vt F_j = -\partial U/\partial \vt R_j$ is, by definition, the force applied on particle $j$. As $\vt R_j = \vt r + \tr{\mt A}\vt d_j$,
\[
\left( \diff{\vt R_j}{\vt q} \right)^\ast = \left( \diff{(\tr{\mt A}\vt d_j)}{\vt q} \right)^\ast.
\]

For simplicity, we will drop the index $j$ for the next few lines. Remember that $\tr{\mt A}\vt d = \tr{\mt C} \mt B\vt d$ and, according to Eq.~\ref{eq:product_B_vector},
\[
\mt B \vt d = \sum_{k=1}^3 d_k \mt B_k \vt q = \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q.
\]

An identity equivalent to Eq.~\ref{eq:vector_entries} holds for the product of $\tr{\mt C}$ with an arbitrary quaternion, so that we can write
\[
\tr{\mt A}{\vt d} =
\left[\begin{array}{c}
\tr{\vt q}\tr{\mt C}_1 \mt B \vt d \\
\tr{\vt q}\tr{\mt C}_2 \mt B \vt d \\
\tr{\vt q}\tr{\mt C}_3 \mt B \vt d \\
\end{array}\right] = 
\left[\begin{array}{c}
\tr{\vt q} \tr{\mt C}_1 \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q \\
\tr{\vt q} \tr{\mt C}_2 \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q \\
\tr{\vt q} \tr{\mt C}_3 \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q \\
\end{array}\right].
\]

As mentioned in Appendix \ref{sec:auxiliary_math}, $\tr{\mt C}_i \mt B_j = \tr{\mt B}_j \mt C_i$ for any pair $ij$, so that all the constant matrices above between $\tr{\vt q}$ and $\vt q$ are constant and symmetric, which facilitates differentiation. After carrying it out and replacing back $(\sum_{k=1}^3 d_k \mt B_k) \vt q$ by $\mt B \vt d$, we have
\[
\left( \diff{\vt R}{\vt q} \right)^\ast = 2 \left[\begin{array}{ccc}
\tr{\mt C}_1 \mt B \vt d & \tr{\mt C}_2 \mt B \vt d & \tr{\mt C}_3 \mt B \vt d
\end{array}\right]
\]

Considering that we deal exclusively with unit quaternions $\vt q$, so that $\mt C \tr{\mt C} = \mt 1_4 - \vt q \tr{\vt q}$ can be substituted in Eq.~\ref{eq:diff_projection}, it is useful to evaluate the product
\[
\tr{\mt C} \left( \diff{\vt R}{\vt q} \right)^\ast = 2 \left[\begin{array}{c}
\tr{\vt q}\tr{\mt C_1} \\
\tr{\vt q}\tr{\mt C_2} \\
\tr{\vt q}\tr{\mt C_3}
\end{array}\right] \left[\begin{array}{ccc}
\tr{\mt C}_1 \mt B \vt d & \tr{\mt C}_2 \mt B \vt d & \tr{\mt C}_3 \mt B \vt d
\end{array}\right]
\]

The permutation matrices $\mt C_1$, $\mt C_2$, and $\mt C_3$ have the following properties:
\begin{equation}
\begin{aligned}
\tr{\mt C_1}\tr{\mt C_1} &= \tr{\mt C_2}\tr{\mt C_2} = \tr{\mt C_3}\tr{\mt C_3} = -\mt 1_4 \\
\tr{\mt C_1}\tr{\mt C_2} &= -\tr{\mt C_2}\tr{\mt C_1} = -\tr{\mt C_3} \\
\tr{\mt C_2}\tr{\mt C_3} &= -\tr{\mt C_3}\tr{\mt C_2} = -\tr{\mt C_1} \\
\tr{\mt C_3}\tr{\mt C_1} &= -\tr{\mt C_1}\tr{\mt C_3} = -\tr{\mt C_2}
\end{aligned}
\end{equation}

Along with the fact that $\mt B\vt d$ and $\vt q$ are necessarily orthogonal to each other, the properties above lead to
\begin{align*}
\tr{\mt C} \left( \diff{\vt R}{\vt q} \right)^\ast &= 2 \left[\begin{array}{ccc}
0 & -\tr{\vt q}\tr{\mt C}_3 \mt B \vt d & \tr{\vt q}\tr{\mt C}_2 \mt B \vt d \\
\tr{\vt q}\tr{\mt C}_3 \mt B \vt d & 0 & -\tr{\vt q}\tr{\mt C}_1 \mt B \vt d \\
-\tr{\vt q}\tr{\mt C}_2 \mt B \vt d & \tr{\vt q}\tr{\mt C}_1 \mt B \vt d & 0
\end{array}\right] \\
&= 2 \mt S(\tr{\mt C} \mt B \vt d) = 2 \mt S(\tr{\mt A}\vt d) = 2 \mt S(\vt \delta),
\end{align*}
where $\vt \delta$ is equivalent to $\vt d$, but expressed in the space-fixed frame of reference. Finally, we obtain
\[
\diff{U}{\vt q} = - 2 \mt C \sum_{j=1}^{n_p} \mt S(\vt \delta_j) {\vt F_j}  = - 2 \mt C \sum_{j=1}^{n_p} \vt \delta_j \times {\vt F_j}.
\]

\bibliography{rigid_bodies}

\end{document}

### 2016-06-14 rigid_bodies.tex
Small changes in the introductory section

@@
 
-\title{A Simplified Approach for Molecular Dynamics with Rigid Bodies}
+\title{A Simplified Formulation for Molecular Dynamics with Rigid Bodies}
+
+\author{Ana J. Silveira}
+\email{asilveira@plapiqui.edu.ar}
+\affiliation{Chemical Engineering Department, Escola de Qu\'imica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}
+\affiliation{Planta Piloto de Ingenier\'ia Qu\'imica, PLAPIQUI, Universidad Nacional del Sur, Camino La Carrindanga Km 7-CC: 717, Bah\'ia Blanca, Argentina}
 
@@
 \section{Introduction}
-The advances in hardware along with the improvements in molecular dynamics algorithms and software has allowed to extend both the time and length scales attainable in atomistic simulations.
-This progress has resulted in an ambitious attempt to elucidate complex physical phenomena, such as the mechanism underlying molecular motors, the pathways for nanoparticle assembly, the interfacial phenomena, just to name a few.
-Because of its complexity and being computationally demanding, this challenging task requires strategies to further reduce the computational time.
-
-One of the strategies commonly used to speed up the simulations is the suppression of the uninteresting degrees of freedom, as might be the case of the intramolecular vibrations.
-The rigid body molecular dynamics appears as the more suitable approach for rigid molecules.
-In contrast to the widely used iterative methods,\cite{Ryckaert1977,Andersen1983} one is able to make an implicit treatment of the holonomic constraints, which is computationally more efficient.
+
+Improvements in molecular dynamics algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and length scales attainable in their atomistic simulations.
+This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
+Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.
+
+One of the strategies commonly used to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
+Widely used iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieve
+
+the dynamics of rigid bodies is able to keep this type of constraint implicitly, and thus emerges as a suitable approach in these cases.
+
+one is able to make an implicit treatment of the holonomic constraints, which is computationally more efficient.
 In addition, and most important, it is possible to obtain a reversible and symplectic integrator in the microcanonical ensemble.

### 2016-06-24 rigid_bodies.tex
first modified version Ana

@@
 \newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
-
+\mathchardef\mhyphen="2D
+\usepackage{booktabs}
+\usepackage{multirow}
 \begin{document}
@@
 
-Improvements in molecular dynamics algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and length scales attainable in their atomistic simulations.
+Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and length scales attainable in their atomistic simulations.
 This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
@@
 In addition, and most important, it is possible to obtain a reversible and symplectic integrator in the microcanonical ensemble.
-The property of being symplectic implies the conservation of volume in phase space.
-This feature, together with time reversibility are mandatory for numerical integrators used in hybrid monte carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic reversible integrator for the rotational motion in terms of the unit quaternion.
-This approach has been successfully used in pure molecular dynamics as well as in HMC methods to simulate a wide range of systems including the ubiquitous case of water \cite{Sakamaki2011,Reinhardt2012,Palmer2014,Gonzales2014}, ice water \cite{Geiger2014}, hydrates \cite{Tribello2009,Gorman2012}, diffusing gases in metal-organic frameworks \citep{Ghoufi2010}.
+The property of being symplectic implies the preservation of phase space volume.
+This feature, together with time reversibility are mandatory for numerical integrators used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic reversible integrator for the rotational motion in terms of the unit quaternion.
+This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems including the ubiquitous case of water \cite{Sakamaki2011,Reinhardt2012,Palmer2014,Gonzales2014}, ice water \cite{Geiger2014}, hydrates \cite{Tribello2009,Gorman2012}, diffusing gases in metal-organic frameworks \citep{Ghoufi2010}, etc.
 It is worth noting that it also can be used for systems designed as a collection of interconnected rigid bodies, which is a coarse-graining strategy widely used for proteins \citep{Terada2003}, molecular machines \citep{Akimov2008,Konyukhov2010}, nanoparticles \cite{Knorowski2012,Patra2013}, etc.
 
-In this paper, we present a simplified version for the equations of motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as the corresponding numerical integrators to perform molecular dynamics.
-The contribution for the NVE case relies on the factorization of the rotation matrix, which allows a Hamiltonian formalism that avoids the use of the extended phase space methodology, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002}.
-In the NVT case, both the translational and rotational degrees of freedom (DOF) of all particles are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat.
+In this paper, we present a simplified version for the equations of motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as the corresponding numerical integrators to perform MD.
+The contribution to the NVE case relies on a factorization of the rotation matrix, which allows a Hamiltonian formalism that avoids the use of the extended phase space methodology, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002}.
+In the NVT case, both the translational and rotational degrees of freedom (DOF) are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat.
 This differs from the approach of Kamberaj \textit{et al}.\cite{Kamberaj2005}, which uses two independent thermostats, one for the rotational and another one for the translational DOF.
-It should be mentioned here that there are several numerical integrators\cite{Omelyan2007,Omelyan2008,vanZon2008} for the equations of motion for rigid bodies in the NVE ensemble which are more accurate than the symplectic Verlet-type integrators.
-However, since the application of that methods is limited to Hamiltonian systems, it's unclear whether their benefits will extend to the numerical integrators in the NVT ensemble.
-This, together with its great applicability, are the main reasons why our approach is based on the Verlet-type integrators.
-
-We have also derived an expression for calculating the pressure of a rigid body system whose constituents atoms interact via a pairwise potential and are subjected to periodic boundary conditions.
-That expression depends on the resultant interaction forces between rigid body pairs.
+We have also simplified the calculation of the pressure for a rigid body system, whose constituents atoms interact via a pairwise potential and are subjected to periodic boundary conditions (PBC).
+The expression we derive depends on the resultant interaction forces between rigid body pairs.
 This implies that there is no need to evaluate the unknown forces of constraint, which one must take into account if the pressure is calculated in terms of the interatomic forces.
+
 This paper is organized as follows: In Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism.
-As a preamble to the integration schemes, we show an alternative way of interpreting the action of the Liouville operators on the phase space variables.
-In Sec.~\ref{sec:moleculardynamics} we present the factorization of the Liouville operator for both the NVE and NVT ensembles.
-Following this, in Sec.~\ref{sec:pressure} we show the pressure expression developed here and we compare the result with that obtained from other method.
-In Sec.~\ref{sec:performance} we discuss the performance of the numerical integrators according to the conservation of the total energy.
-For the NVT ensemble, we also present the results from a test that checks if the potential and kinetic energies are actually sampled according to the canonical distribution.
+In Sec.~\ref{sec:moleculardynamics} we present the numerical integrators for both the NVE and NVT ensembles, whose performance we discuss in terms of the conservation of the total energy.
+Additionally, for the NVE case we present some results related to the partition of kinetic energy and for the NVT case, we also analyze the statistical distributions of the sampled energy.
+Following this, in Sec.~\ref{sec:pressure} we show the equation for the pressure developed in this work.
 Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations.
 Only ordinary matrix/vector operations are present in the formulation.
+Some mathematical details are given in the Appendix.
 
@@
 \label{sec:mechanics}
-In this section we review one particular factorization of the rotation matrix which allows to obtain a mathematical relation between the rate of change of the four-dimensional quaternion and the angular velocity vector, while automatically satisfying the restriction imposed on the quaternion.
-That relation was already introduced by Kneller \textit{et al}\cite{Kneller1994} in the context of Lagrangian dynamics for rigid bodies, but here we use it in the derivation of the Hamiltonian equations of motion.
+
+In this section we first review certain aspects of the kinematics or rigid bodies we consider relevant to this study.
+In Sec~\ref{sec:eulerparameters}, we introduce one particular factorization of the rotation matrix from which we obtain a mathematical relation between the rate of change of the four-dimensional quaternion and the angular velocity vector, while automatically satisfying the restriction imposed on the quaternion.
 
@@
 For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal.
+The corresponding axes are known as the principal inertia axes and the formulation of the equations of motion in this frame of reference is notably simplified.
 Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
@@
 \end{equation}
-
 The operator  $\mt S(\cdot)$ can be used to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.
 
-\subsection{Euler Parameters and Unit Quaternions}
-
-It is possible to represent the orientation of a rigid body by Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$ that satisfies\cite{Goldstein2002}
+%\subsection{Euler Parameters and Unit Quaternions}
+\subsection{The Rate of change of Euler Parameters}
+\label{sec:eulerparameters}
+
+Euler's rotation theorem establishes that an arbitrary rotation can be described using three independent parameters.
+One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in terms of those angles.
+Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$.
+The mathematical relationship between the Euler angles and the quaternion places the restriction over $\boldsymbol{q}$:\cite{Goldstein2002}
 \begin{equation}
@@
 \end{equation}
-
 Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $R^4$.
@@
 \end{equation}
-
 This equation implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.
+
 Let us now introduce two new matrices that also depend on the Euler parameters, which are
@@
 \end{subequations}
-
 Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited in Appendix \ref{sec:auxiliary_math} with our notation.
@@
 \end{equation}
-
 This is a simple and convenient factorization, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters.
@@
 \label{sec:hamiltonian}
+In the previous subsection we obtained the relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}), which is used in this part to develop the Hamiltonian formalism of the equations of motion.
 As usual, the dynamics of a rigid body can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively.
@@
 \end{equation}
-
 Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$.
@@
 
-The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} The equation above is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only.
+The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only.
 However, if we define
@@
 \end{equation}
-
 Now this is convenient for differentiation with respect to $\vt q$.
@@
 
-For a rigid body composed of individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
+For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
 \[
@@
 An exact solution for the system of equations represented by Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
-Hence, it would describe a reversible, measure-preserving, and conservative motion with respect to the Hamiltonian $\mathcal{H}$.
+Hence, it would describe a conservative dynamics with respect to the Hamiltonian $\mathcal{H}$, in which phase space volume element would be preserved.
 Nevertheless, a general analytical solution does not exist.
 This is so, above all, because $U(\vt r, \vt q)$ is arbitrary, but even in the simplest case of free motion (i.e., when $U$ is constant), no closed-form solution is known due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}.
+Therefore, it becomes necessary to introduce the Liouville operators formalism, which allows us to derive symplectic numeric integrators rigorously.
+
 For a given trajectory in phase-space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
@@
 Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}.
-Let us analyze three of them, which will be useful later on.
+Let us analyze three of them, which will be useful later on to derive a complete numerical solution.
 In what follows, subscript $0$ always denotes ``initial values''.
@@
 \end{align*}
-
 Clearly, this corresponds to a uniform linear translation.
@@
 \end{align*}
-
 Note that $\vt r$ and $\vt q$ remain unaltered, thus keeping $\vt F$, $\mt C$, and $\vt \tau$ constant as well (hence the subscript $0$ assigned to them).
@@
 It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well.
-This decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form.
+As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form.
 The following analytical solution takes place:\cite{Miller2002}
@@
 \end{align}
-
 These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{i L_k t}$ derived from the operator
@@
 
-Now note that the Liouville operator corresponding to the original case can be written as
+Now note that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
 \begin{align}
@@
 \end{equation}
-
 where $L_r = \sum_{k=1}^3 L_k$ is the Liouville operator for a complete rotation and, $i L_a $, $i L_k $ as well as $i L_t $ retain its previous definitions.
 Recall that each exponential operator in Eq.~\ref{eq:trotter_splitting_NVE} will act having the result of the preceding operator (the one at its right hand side) as initial condition.
-Following Ref.~\onlinecite{Miller2002} once again, the splitting employed for the rotational part is
+Following Ref.~\onlinecite{Miller2002} once again, the splitting employed for the rotational part is\cite{Tuckerman1992}
 
@@
 \label{eq:splitting_rot}
-e^{i L_r \Delta t} = \left( e^{i L_3 \frac{\delta t}{2}} e^{i L_2 \frac{\delta t}{2}} e^{i L_1 \delta t} e^{i L_2 \frac{\delta t}{2}} e^{i L_3 \frac{\delta t}{2}} \right)^n
-\end{equation}
-
+e^{i L_r \Delta t} = \left( e^{i L_3 \frac{\delta t}{2}} e^{i L_2 \frac{\delta t}{2}} e^{i L_1 \delta t} e^{i L_2 \frac{\delta t}{2}} e^{i L_3 \frac{\delta t}{2}} \right)^n,
+\end{equation}
 where $\delta t = \nicefrac{\Delta t}{n}$.
 This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step to improve accuracy.
-An alternative to a further factorization of Eq.~\ref{eq:splitting_rot} with higher values of $n$ could be the use of a higher-order scheme, as the one due to Yoshida \cite{Yoshida1990} and Suzuki\cite{Suzuki1991a,Suzuki1991b}, commonly used in Nos\'{e}-Hoover chain dynamics.
-The factorization of the operator is expressed as follows
+An alternative to a further factorization of Eq.~\ref{eq:splitting_rot} with higher values of $n$ could be the use of a higher-order scheme, as the one due to Yoshida \cite{Yoshida1990} and Suzuki\cite{Suzuki1991a,Suzuki1991b}, given by
 
@@
 \label{eq:ys_rot}
-e^{i L_{r} \Delta t} = \prod_{j=1}^{n_{sy}} e^{i L_{r} \omega_j \Delta t}
-\end{equation}
-
-where $\omega_j$ are employed to weight the time step and $n_{sy}$ determines the order of the method.
-Note that we are not improving the order of precision for the integration of the translational degrees of freedom, which would be computationally more demanding.
-Nevertheless, we would like to investigate   how the performance of the integrator as a whole is affected by using a higher-order scheme for the rotational part or the operator.
-For that purpose, we have considered $n = 1$ with $n_{ys} = 1 $ (zero-order) and $n_{sy} = 7$ (sixth-order).
+e^{i L_{\mathrm{r\mhyphen ys}} \Delta t} = \prod_{j=1}^{n_{sy}} e^{i L_{r} \omega_j \Delta t},
+\end{equation}
+where $\omega_j$ are used to weight the time step and $n_{ys}$ determines the order of this method.
+Note that we are not improving the order of precision for the integration of the translational DOF, which is computationally more demanding.
+Nevertheless, we would like to investigate how the performance of the integrator as a whole is affected by using a higher-order scheme for the rotational operator.
+For that purpose, in this paper we consider $n = 1$ with $n_{ys} = 1 $ (zero-order) and $n_{sy} = 7$ (sixth-order).
+In this case, the weights are:
+\begin{align*}
+&\omega_1 = \omega_7 = 0.784513610477560 \\
+&\omega_2 = \omega_6 = 0.235573213359357 \\
+&\omega_3 = \omega_5 = -1.17767998417887 \\
+&\omega_4 = 1 - \omega_1 -\omega_2 - \omega_3 - \omega_5 - \omega_6 - \omega_7
+\end{align*}
+
+It should be mentioned here that one can find in literature several numerical integrators\cite{Omelyan2007,Omelyan2008,vanZon2008} for NVE rigid body dynamics which are more accurate than the symplectic Verlet-type integrators.
+However, since the application of those methods is limited to Hamiltonian systems, it's unclear whether their benefits will extend to the numerical integrators in the NVT ensemble.
+This, together with its great applicability, are the main reasons why our approach is based on the Verlet-type integrators.
 
@@
 \dot{\vt p} &= \vt F^d = {\vt F} - \frac{p_{\eta}}{Q} {\vt p} = \sum_{j=1}^{n_p} {\vt F}_j - \frac{p_{\eta}}{Q} \sum_{j=1}^{n_p} {\vt p}_j \\
-&= \sum_{j=1}^{n_p} \left( {\vt F}_j - \frac{p_{\eta}}{Q} {\vt p}_j \right)
-\end{align*}
-
-Thus, the thermostat affects each particle $j$ individually. It is then intuitive that the resultant torque is affected by the thermostat according to
+&= \sum_{j=1}^{n_p} \left( {\vt F}_j - \frac{p_{\eta}}{Q} {\vt p}_j \right).
+\end{align*}
+Thus, the thermostat affects each particle $j$ individually. Analogously, the resultant torque is affected by the thermostat according to
 \begin{align*}
@@
 \[
-\vt L = \tr{\mt C}{\mt B}{\mt I}{\vt \omega} = \frac{1}{2} \tr{\mt C}{\vt \pi}
+\vt L = \tr{\mt C}{\mt B}{\mt I}{\vt \omega} = \frac{1}{2} \tr{\mt C}{\vt \pi}.
 \]
@@
 Finally, the Nos\'{e}-Hoover chain thermostat applied to a system of $N$ rigid bodies becomes:
-
-\begin{align*}
-&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} & i = 1,\cdots,N \\
-&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i & i = 1,\cdots,N \\
-&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i & i = 1,\cdots,N \\
-&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q} \vt \pi_i & i = 1,\cdots,N \\
+\begin{subequations}
+\label{eq:nhc_system}
+\begin{align}
+&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
+&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i  \\
+&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i  \\
+&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q} \vt \pi_i \\
 &\dot{\eta}_j = \frac{p_{\eta_j}}{Q_j} & j = 1,\cdots,M \\
-&{\dot p}_{\eta_j} = G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} & j = 1,\cdots,M-1 \\
+&{\dot p}_{\eta_j} = G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} & j = 1,\cdots,M-1 \label{eq:p_eta_j} \\
 &{\dot p}_{\eta_M} = G_M
-\end{align*}
+\end{align}
+\end{subequations}
 where
@@
 &G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - 6 N k_B T \\
-&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_B T &j = 2,\cdots,M \\
+&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T &j = 2,\cdots,M \\
 &Q_1 = 6 N k_B T t_d^2 \\
@@
 \end{align*}
-These equations have the conserved energy $\mathcal{H}^\prime$ given by
-\begin{align*}
-\mathcal{H}^\prime = \mathcal{H}(\vt r, \vt p, \vt q, \vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_k} + 6Nk_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j
-\end{align*}
-
-In these equations, $M$ is the number of thermostats in the chain, $T$ is the set-point temperature, $k_B$ is Boltzmann's constant, $6N$ is the number of degrees of freedom, and $t_d$ is the characteristic time period of the thermostat chain.
-$\mathcal{H}$ was defined in section~\ref{sec:hamiltonian}.
+In these equations, $i$ indexes the number of rigid bodies from $1$ to $N$, $M$ is the number of thermostats in the chain, $T$ is the specified temperature, $k_b$ is Boltzmann's constant, $6N$ is the number of DOF, and $t_d$ is the characteristic time period of the thermostat chain.
+The conserved quantity in this case is the extended energy $\mathcal{H}^\prime$ given by
+\begin{align*}
+\mathcal{H}^\prime = \mathcal{H} + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_k} + 6Nk_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j,
+\end{align*}
+with $\mathcal{H}$ defined in Eq.~\ref{eq:H_with_Omega}.
 
 As was done in the Hamiltonian case, the evolution of the extended phase space vector $ f(\vt r, \vt p, \vt q, \vt \pi, \eta_1,..
-\eta_M, p_{\eta_1},..,p_{\eta_M}) $ is given by the relation $f(t) = e^{i L t}f_0$ in the context of the generalized Liouville theorem.
+\eta_M, p_{\eta_1},..,p_{\eta_M}) $ is given by the relation $f(t) = e^{i L t}f_0$ in the context of the generalized Liouville theorem\cite{Tuckerman_1999}.
 This implies that we can employ the Trotter theorem to devise a strategy for integrating the equations of motion.
@@
 \[
-i L = i L_{NHC} + i L_{NVE}
-\]
-where $i L_{NVE}$ has been previously defined (Eq.~\ref{eq:full_operator}).
-It follows that the operator $ i L_{NHC} $ is given by
+i L = i L_{NHC} + i L_{NVE},
+\]
+where $i L_{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $ i L_{NHC} $ is given by
 
@@
 & i L_{NHC} = -\sum_{i=1}^{N} \frac{p_{\eta_1}}{Q_1} \vt p_i \diff{}{\vt p_i} - \sum_{i=1}^{N} \frac{p_{\eta_1}}{Q_1} \vt \pi_i \diff{}{\vt \pi_i} \\
-& + \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}} {Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right]\\
+& + \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}} {Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right],\\
 \end{split}
 \end{equation}
-\\
-with $p_{\eta_{M+1}}/Q_{M+1} = 0$
-\\
-
-The part of the operator given by 
-\begin{align*}
-&\left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}} \\
-\end{align*}
-
-can be evaluated analytically, as pointed out by Martyna \textit{et al.}\cite{Martyna1996}, using the following equation:
-
+with $p_{\eta_{M+1}}/Q_{M+1} = 0$.
+
+As pointed out by Martyna~\textit{et al.}~\cite{Martyna1996}, the part of the operator given by 
+\begin{align*}
+&\left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}},
+\end{align*}
+can be evaluated analytically using the following equation:
 \begin{equation}
@@
 \end{equation}
-\\
-where $\phi(x) = 1-e^{-x}/x = \sum_{n=0}^\infty (-1)^n x^n/(n+1)!$. The function $\phi(x)$ can be evaluated by
+where $\phi(x) = 1-e^{-x}/x = \sum_{n=0}^\infty (-1)^n x^n/(n+1)!$. For small values of $|x|$, the function $\phi(x)$ can be evaluated by
 
@@
 
-The NHC operator is factorized as follows
-\\
+Considering the above analytical solution, the NHC operator is factorized as follows
 
@@
 e^{iL_{NHC} \frac{\Delta t}{2}} =  &\prod_{j=M}^{1} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-&  \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{2} \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right]  \prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt p_i \diff{}{\vt p_i}\right] \prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt \pi_i \diff{}{\vt \pi_i}\right]  \\
-& \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]  \\
+&  \times \prod_{j=1}^{M} exp\left(\frac{\Delta t}{2} \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)  \prod_{i=1}^{N} exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt p_i^T \diff{}{\vt p_i}\right) \prod_{i=1}^{N} exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T \diff{}{\vt \pi_i}\right)  \\
+& \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
 \end{split}
@@
 \end{widetext}
-
-The exponential factors in the second line of the equation above are evaluated according to
-
-\begin{equation}
-\begin{split}
-&\vt \pi_i \leftarrow \vt \pi_i  exp\left[-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right]\\
-&\vt p_i \leftarrow \vt p_i  exp\left[-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right]\\
-&\eta_j  \leftarrow \eta_j + \frac{p_{\eta_j}}{ Q_j} \frac{\Delta t}{2} 
-\end{split}
-\end{equation}
-
-Retaining the NVE factorization scheme, Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as
+The remaining factors that appear in the second line of the equation above are evaluated according to\cite{Martyna1996}
+\begin{equation}
+\vt \pi_i \leftarrow \vt \pi_i \, exp\left(-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right)
+\end{equation}
+\begin{equation}
+\vt p_i \leftarrow \vt p_i \, exp\left(-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right)
+\end{equation}
+\begin{equation}
+\eta_j  \leftarrow \eta_j + \frac{p_{\eta_j}}{ Q_j} \frac{\Delta t}{2} 
+\end{equation}
+
+Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as
 
@@
 e^{i L \Delta t} =
-e^{i L_{NHC} \frac{\Delta t}{2}} e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t}  e^{i L_a \frac{\Delta t}{2}} e^{i L_{NHC} \frac{\Delta t}{2}},
-\end{equation}
-It is worth noting that analogous analytical solutions exists also for the following equations
-\begin{align*}
-&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i  \\
-&\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q_1} \vt \pi_i  \\
-\end{align*}
-The solutions for $\vt p_i$ and $\vt \pi_i $  are then
-
+e^{i L_{NHC} \frac{\Delta t}{2}} e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t}  e^{i L_a \frac{\Delta t}{2}} e^{i L_{NHC} \frac{\Delta t}{2}}.
+\end{equation}
+
+It is worth noting that analytical solutions exist also for the equations
+\begin{align*}
+&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i \, \text{ and} \\ 
+&\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q_1} \vt \pi_i,
+\end{align*}
+which are of identical form to Eq.~\ref{eq:p_eta_j}, so they can be evaluated using
 \begin{equation}
@@
 \begin{split}
-{\vt p}_i \leftarrow {\vt p}_i + \left({\vt F}_i - \frac{p_{\eta_1}}{Q_1} {\vt p}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t \\
+{\vt p}_i \leftarrow {\vt p}_i + \left({\vt F}_i - \frac{p_{\eta_1}}{Q_1} {\vt p}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t \, \text{ and}\\
 \end{split}
@@
 \begin{split}
-{\vt \pi}_i \leftarrow {\vt \pi}_i + \left(2 \tr{\vt \tau} \tr{\mt C} - \frac{p_{\eta_1}}{Q_1} {\vt \pi}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t \\
+{\vt \pi}_i \leftarrow {\vt \pi}_i + \left(2 \tr{\vt \tau} \tr{\mt C} - \frac{p_{\eta_1}}{Q_1} {\vt \pi}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t,
 \end{split}
 \end{equation}
-
-with $\phi(x)$ given by Eq.~\ref{eq:phi}.
-\\
-
-Therefore, one of the aims is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when also Eqs.~\ref{eq:p} and~\ref{eq:pi} are considered.
-For that, let as define the modified operators ($\ast$) $ i L_a $ and $ i L_{NHC} $ as follows:
-
+respectively, with $\phi(x)$ given by Eq.~\ref{eq:phi}.
+Therefore, one of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when also Eqs.~\ref{eq:p} and~\ref{eq:pi} are considered.
+For that, let as define the modified propagators $ e^{i L^\ast_a} $ and $ e^{i L^\ast_{NHC}} $ as follows:
+%\begin{equation}
+%\begin{split}
+%&i L^\ast_a = i L_a -\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1} \vt p_i^T \diff{}{\vt p_i} - %\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt \pi_i^T  \diff{}{\vt \pi_i} =\\
+%&\sum_{i=1}^{N} \left(\tr{\vt F_i} - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} + %\sum_{i=1}^{N} \left(2 \tr{\vt \tau_i} \tr{\mt C_i} -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T\right)  \diff{}{\vt %\pi_i}
+%\end{split}
+%\end{equation}
+%\begin{equation}
+%\begin{split}
+%i L^\ast_{NHC} = \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} %\frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right] 
+%\end{split}
+%\end{equation}
 \begin{equation}
 \begin{split}
-i L^\ast_a = &i L_a -\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt p_i\diff{}{\vt p_i} - \sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt \pi_i  \diff{}{\vt \pi_i} =\\
-             &\sum_{i=1}^{N}(\tr{\vt F_i} - \frac{p_{\eta_1}}{Q_1}\vt p_i) \diff{}{\vt p_i} + \sum_{i=1}^{N}(2 \tr{\vt \tau_i} \tr{\mt C_i} -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i)  \diff{}{\vt \pi_i}.
+e^{iL^\ast_a \frac{\Delta t}{2}} = &\prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \left( \vt F_i^T - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} \right] \\
+&\times \prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \left(2 \vt \tau_i^T \mt C_i^T -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T \right)  \diff{}{\vt \pi_i}   \right]
 \end{split}
@@
 
-\begin{equation}
-\begin{split}
-i L^\ast_{NHC} = \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right] 
-\end{split}
-\end{equation}
-
-Finally, we define the following factorization scheme:
-\begin{widetext}
 \begin{equation}
@@
 e^{iL^\ast_{NHC} \frac{\Delta t}{2}} = &\prod_{j=M}^{1} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-& \times\prod_{j=1}^{M} exp\left[-\frac{\Delta t}{2}  \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right]   \\
-& \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]  \\
+&\times\prod_{j=1}^{M} exp\left(-\frac{\Delta t}{2}  \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)   \\
+&\times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
 \end{split}
 \end{equation}
-\end{widetext}
-
+Finally, the complete propagator scheme is
 \begin{equation}
@@
 \begin{split}
-e^{i L \Delta t} = &e^{i L^\ast_{NHC} \frac{\Delta t}{2}} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L^\ast_{NHC} \frac{\Delta t}{2}},
+e^{i L \Delta t} = &e^{i L^\ast_{NHC} \frac{\Delta t}{2}} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L^\ast_{NHC} \frac{\Delta t}{2}}.
 \end{split}
 \end{equation}
+
+In the NVE case, each Liouvillian operator is derived from part of the Hamiltonian, then each factor in the propagator is symplectic and, as a consequence, preserves the phase space volume.
+Recall that the inverse statement is not necessarily true.
+Now it is widely recognized the importance of the symplectic property for numerical stability,\cite{Skeel1997} however it is restricted to Hamiltonian systems.
+For non-Hamiltonian equations of motion the phase space volume is no longer conserved.
+Nevertheless, there exists an invariant measure on the phase space, and for the the system of equations~\ref{eq:nhc_system} it is given by
+\begin{align*}
+\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ].
+\end{align*}
+
+In this sense, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which are both time  reversible an measure-preserving.
+In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to more ergodic trajectory behavior.
+It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*} do not preserve the invariant-measure.
+Nevertheless, as will be shown below, those numerical schemes successfully achieve the specified temperature and the long-term energy conservation is acceptable.
+Note that the term that premultiplies the extended phase space volume in the invariant-measure above is a function of the variable $\eta$.
+Given that the time evolution of the physical variables along with the thermostat momentum does not depend on $\eta$ we conclude that in this case the measure-preserving condition for every individual operator can actually be relaxed.
+
+\section{Numerical Results}
+
+In this section, we analyze some performance aspects of the integration schemes when applied to a molecular system in the liquid state.
+Given its practical importance, we simulated 356 TIP3P\cite{Price2004} water molecules at a density of 999.56 $kg/m^3$.
+The intermolecular interactions were truncated at 9 {\AA}.
+The shifted-force (SF) method\cite{Allen1989,Toxvaerd_2011} was used to treat both the coulombic and the 12-6 Lennard-Jones (LJ) interactions.
 
@@
 \label{sec:performance}
-In this section, we resume some performance aspects of the integration schemes when applied to a molecular system in the liquid state.
-Given its practical importance, we simulated 356 TIP3P\cite{Price2004} water molecules at a density of 1 $g/cm^3$.
-The intermolecular interactions were truncated at 9 {\AA}.
-The damped shift force (DSF) method\cite{Fennell2006} was used to treat the coulombic interactions while  the "shifted-force potential"\citep{Allen1989} method was applied for the 12-6 Lennard-Jones (LJ) interactions.
-
-We used Eq.~\ref{eq:performance} to evaluate the stability of the numerical solvers.
-It measures the  the average deviation of the conserved quantity, i.e.
+
+In order to evaluate the stability of the numerical solvers, we use a measure given by
+\begin{equation}
+\label{eq:performance}
+D E =  \frac{1}{N_{steps}} \sum_{k=1}^{N_{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|,
+\end{equation}
+from which it is possible to calculate the average deviation of the conserved quantity, i.e.
 total energy, from its initial value $E_0$ after a certain number of time steps $N_{steps}$.
@@
 
-\begin{equation}
-\label{eq:performance}
-D E =  \frac{1}{N_{steps}} \sum_{k=1}^{N_{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|
-\end{equation}
-
-Recall that the method in the NVE ensemble is a mathematical simplification of the equations of motion, so the trajectory must coincide, out of round-off errors, with that generated using the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm introduced by Miller~\textit{et al.}\cite{Miller2002}.
-In Fig.~\ref{fig:miller1fs} we plot for both methods the instantaneous fluctuations of energy with $n_{ys} = 1 $ and a time step of 1 fs, where it can be seen that the behaviour is essentially the same.
-The simulation with the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm was conducted using the LAMMPS\cite{Plimpton1995} software package, which does not have the shifted-force potential implemented, so no shifting was applied to the LJ interactions in this case.
-
-\begin{center}
-  \includegraphics{miller1fs}
-  \captionof{figure}{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and Ref.\citenum{Miller2002} (dotted line).
-$n : 1 $, $n_{ys}:1$, $\Delta t : 1 fs$}
+Recall that the method introduced here for the NVE dynamics is a mathematical reformulation of the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm introduced by Miller~\textit{et al.}\cite{Miller2002}, so the trajectories for both methods must coincide out of round-off errors.
+In Fig.~\ref{fig:miller1fs} we plot the instantaneous fluctuations of energy with $n_{ys} = 1 $ and a time step of 1 fs, where it can be seen that, in fact, the fluctuations for both methods are of similar magnitude.
+The simulation with the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm was conducted using the LAMMPS\cite{Plimpton1995} software package.
+For the purpose of comparison we applied the LJ model without shifting the forces, as it is not implemented in LAMMPS.
+
+\begin{figure}[!h]
+\centering
+\includegraphics{millerourmd}
+\captionof{figure}{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and Ref.\citenum{Miller2002} (dotted line) with $n = 1 $, $n_{ys} = 1$, $\Delta t = 1 fs$, $E_0 = -2799.87 \, kcal/mol $.}
 \label{fig:miller1fs}
-\end{center}
-
-In Table~\ref{table:nve} we compare the energy drift when using $n_{ys}:1$ and $n_{ys}:7$ in the rotational operator, Eq.~\ref{eq:ys_rot}.
+\end{figure}
+
+In Table~\ref{table:nve} we compare the energy drift when using $n_{ys} = 1$ and $n_{ys} = 7$ in the rotational operator, Eq.~\ref{eq:ys_rot}.
 As expected, for a small time step of 1 fs, that further factorization is worthless.
-As the time step is increased, more accurate trajectories are achived with $n_{ys} = 7 $.
+As the time step is increased, more accurate trajectories are achieved with $n_{ys} = 7 $.
 Nevertheless, the improvement is only marginal and all trajectories are stables exhibiting a small long-time energy drift.
@@
 \setlength{\tabcolsep}{7pt}
-\caption{Average deviation of the conserved quantity ($DE$) from liquid water simulation with NVE integration with $n_{ys}:1$ and $n_{ys}:7$  }
+\caption{Average deviation of the conserved quantity ($DE$) from liquid water NVE simulations with $n_{ys}=1$ and $n_{ys}=7$  }
 \centering % centering table
@@
 \cline{2-3}
-$\Delta t$ &$n_{ys}:1$ &$n_{ys}:7$ \\
+$\Delta t/fs$ &$n_{ys}=1$ &$n_{ys}=7$ \\
 \hline % inserts single-line
- 1fs & 0.000055 & 0.000074\\
- 2fs & 0.00048 & 0.00031 \\
- 3fs & 0.0012 & 0.0011\\
+ 1 & 0.000055 & 0.000074\\
+ 2 & 0.00048 & 0.00031 \\
+ 3 & 0.0012 & 0.0011\\
  \hline
@@
 
-It is not the purpose of this work to achieve an optimal factorization of the NVT schemes, but rather to show that a unique chain thermostat, while providing the proper sampling of the NVT ensemble, is capable of maintaining the average temperature at its set point.
-In order to use a constant value for the smaller time step in the NHC and NHC* operators, we implemented a multiple time-scale scheme, like the RESPA methodology\cite{Tuckerman1992}.
-Recalling that in the NVT dynamics the conserved quantity is the extended energy $\mathcal{H}^\prime$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the temperatures obtained in each case.
+It is not the purpose of this work to achieve an optimal factorization of the NVT schemes, but rather to show that a unique chain thermostat, while providing the ``proper'' sampling of the NVT ensemble, is capable of maintaining the average temperature at its specified value.
+In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\ref{eq:splitting_rot}.
+Recalling that in the NVT dynamics the conserved quantity is the extended energy $\mathcal{H}^\prime$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the average temperatures obtained in each case.
+It can be seen that both integrators, represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*}, systematically achieve the specified temperature, regardless of the time step size.
+In addition, the energy drift is, at its worst, around $1 \%$.
 
@@
 \cline{2-5}
-$\Delta t$ &$D E$ &$T/K$ &$D E$ &$T/K$\\
+$\Delta t/fs$ &$D E$ &$T/K$ &$D E$ &$T/K$\\
 \hline % inserts single-line
- 1fs  & 0.000043 & 298.16  & 0.000052  & 298.15 \\
-
- 2fs  & 0.00013 & 298.23  & 0.00017 & 298.06  \\
-
- 3fs  & 0.00089 & 298.17  & 0.0013 & 298.12  \\
+ 1  & 0.000043 & 298.16  & 0.000052  & 298.15 \\
+
+ 2  & 0.00013 & 298.23  & 0.00017 & 298.06  \\
+
+ 3  & 0.00089 & 298.17  & 0.0013 & 298.12  \\
 
@@
 \begin{tablenotes}
-\item[a] standard deviation of the mean temperature for all cases is $\sigma = 0.037$
-\item[b] smaller time step in the NHC and NHC* operatores is 0.25 fs 
+\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$
+\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs 
 \end{tablenotes}
@@
 
-It can be seen that both integrators represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*} systematically achieve the temperature set point, regardless of the time step size.
-In addition, the energy drift is, at its worst, around $1 \%$.
+\subsection{Partition of Energy}
+\label{sec:energypartition}
+
+Checking the consistency between the equipartition of energy and the simulated energies serves as another approach for investigating the precision of a simulation.
+In fact, as stated by Eastwood~\textit{et al.}~\cite{Eastwood_2010}, it has been useful to improve methodologies related to the calculation of electrostatic\cite{Levitt_1988,Guenot_1992,Arnold_1994} and dispersive\cite{Sagui_1999} interactions, as well as to detect errors in thermostats\cite{Harvey_1998,Mor_2008} and barostats\cite{Feller_1995}.
+The truncation error associated to the numerical schemes is a potential source of error, and is the focus of the work of Eastwood~\textit{et al.}~\cite{Eastwood_2010}, which investigates a solvated protein and a temperature difference of approximately 5.5 $K$ is observed between the protein and the solvent using a time step of 2 fs.
+
+In this paper, we analyze the partition of energy between the translational and rotational DOF, which are related to the temperature by the equipartition theorem as follows
+\begin{align}
+\label{eq:tras_equipartition}
+2 \langle E_{kin}^{trans} \rangle = (3N - 3) k_b T
+\end{align}
+
+\begin{align}
+\label{eq:rot_equipartition}
+2 \langle E_{kin}^{rot} \rangle = 3N k_b T
+\end{align} 
+In these equations, $N$ is the number of rigid bodies.
+In NVE MD simulations, conservation of total linear momentum, $ \boldsymbol{P}= cte $, results in the loss of three translational DOF : $ 3N-3 $.
+For a cubic box with PBC there is no rotational symmetry ~\cite{Frenkel_2013,Kuzkin_2014}, with which the total angular momentum is not conserved, giving a number of rotational DOF of $3N$.
+We calculate the average energy per DOF $ \langle E_{cin} \rangle_{DOF} = \langle E_{cin} \rangle /N_{DOF}$ and then we calculate both the translational and rotational energies from equipartition as follows
+
+\begin{equation}
+\langle E_{kin}^{trans}\rangle =  \langle E_{kin} \rangle_{DOF} \, (3N-3)
+\end{equation}
+
+\begin{equation}
+\langle E_{kin}^{rot} \rangle =  \langle E_{kin} \rangle_{DOF} \, (3N)
+\end{equation}
+
+In the tables below we report the translational and rotational energies using Eqs.~\ref{eq:tras_equipartition} and~\ref{eq:rot_equipartition}, together with those obtained in the NVE simulations.
+In this case, we simulated 903 molecules of liquid water in order to examine the behavior for a cutoff distance of $13$ {\AA}, which is a more conservative criterion as it provides better numerical stability.
+In Table~\ref{table:water_partition_LJ-SF} we observe that for a small time step of 0.5 fs the energies are in good correspondence, which is lost when the time step is increased.
+This result is confirmed also in Table~\ref{table:water_partition_13A_LJ-SF} for the cutoff distance of 13 {\AA}.
+This allows us to prove the influence of the truncation error on in this effect, in accordance with what Eastwood~\textit{et al.}~\cite{Eastwood_2010} point out.
+For significant deviations, the main implication is that Eqs.~\ref{eq:tras_equipartition} and~\ref{eq:rot_equipartition} are no longer suitable for the temperature calculation, which becomes an undefined property.
+In the aforementioned paper, the authors introduce an alternative approach for calculating the temperature which is based on the generalized equipartition theorem applied to the Shadow Hamiltonian, which will be investigated in future work.
+
+%\begin{table}[h]
+%\centering % centering table
+%\begin{threeparttable}
+%\caption{Partition of kinetic energy from an NVE simulation of 903 water molecules with $r_c = 9$ \AA~ and without shifting of LJ model\tnote{a}} %($5415 grados de libertad$)}
+%\begin{tabular}{| l c  c  c  c  c  c |}  
+%\toprule
+%\hline
+% & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
+%  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
+%\cline{4-7}
+%$\Delta t/fs $  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
+%\hline % inserts single-line
+% 0.5 & 305.04 & 1641.21  & 820.15  & 821.06  & 820.55 & 820.66  \\
+%  1  & 304.88  & 1640.36  & 819.73  & 820.63  & 820.56 & 819.80  \\
+%  2  & 304.04  & 1635.87  & 817.48  & 818.39  & 820.44 & 815.43 \\
+%  3  & 302.78  & 1629.10  & 814.10  & 815.00  & 819.85 & 809.25 \\
+%  \hline
+%\end{tabular}
+%\begin{tablenotes}
+%\item[a] Energies are in $kcal/mol$.
+%\end{tablenotes}
+%\label{table:water_partition_2}
+%\end{threeparttable}
+%\end{adjustbox}
+%\end{table}
+
+\begin{table}[h]
+\centering % centering table
+\begin{threeparttable}
+\caption{Partition of kinetic energy from NVE simulation of 903 water molecules with $r_c = 9 $ {\AA} and SF-LJ\tnote{a}} %($5415 grados de libertad$)}
+%\begin{adjustbox}{width=1\textwidth}
+\begin{tabular}{| l c  c  c  c  c  c |}  
+\hline
+ & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
+  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
+\cline{4-7}
+$\Delta t/fs$  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
+\hline % inserts single-line
+ 0.5 & 305.20  & 1642.11  & 820.60  & 821.06  & 820.60 & 821.51  \\
+  1  & 304.92  & 1640.57  & 819.83  & 820.74  & 820.63 & 819.94  \\
+  2  & 304.20  & 1636.74  & 817.92  & 818.82  & 820.56 & 816.18 \\
+  3  & 302.97  & 1630.07  & 814.59  & 815.48  & 820.47 & 809.60  \\
+  \hline
+ \end{tabular}
+ \begin{tablenotes}
+\item[a] Energies are in $kcal/mol$.
+\end{tablenotes}
+\label{table:water_partition_LJ-SF}
+\end{threeparttable}
+%\end{adjustbox}
+\end{table}
+
+\begin{table}[h]
+\centering % centering table
+\begin{threeparttable}
+\caption{Partition of kinetic energy from NVE simulation of 903 water molecules with $r_c = 13$ {\AA} and SF-LJ\tnote{a}} %($5415 grados de libertad$)}
+%\begin{adjustbox}{width=1\textwidth}
+\begin{tabular}{ |l c  c  c  c  c  c |}  
+\hline
+ & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
+  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
+\cline{4-7}
+$\Delta t/fs$  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
+\hline % inserts single-line
+ 0.5 & 307.17 & 1652.71  & 825.90  & 826.81  & 825.96 & 826.75  \\
+  1 & 306.98  & 1651.68  & 825.38  & 826.30  & 826.09 & 825.59  \\
+  2 & 306.04  & 1646.60  & 822.85  & 823.75 & 825.57 & 821.03 \\
+  3 & 304.79 & 1639.89  & 819.49  & 820.40  & 825.63 & 814.26  \\
+  \hline
+ \end{tabular}
+ \begin{tablenotes}
+\item[a] Energies are in $kcal/mol$.
+\end{tablenotes}
+ \label{table:water_partition_13A_LJ-SF}
+\end{threeparttable}
+%\end{adjustbox}
+\end{table}
 
 \subsection{Sampling Validation}
-
-We present the results from a simple method that serves for checking if an algorithm generates the correct ensemble distribution.
-The procedure, introduced by Shirts ~\cite{Shirts2013}, relies on a quantitative statistical analysis of fitting and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
-We report the results of the last approach as it doesn't depend on binning.
-However, since it is instructive to have a graphical insight, we also show one of the plots of the linear fitting.
-
-Recall that in the NVT ensemble the probability of observing an energy E is given by:
-\[
-P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right)
-\]
-where $ \beta $ is the Boltzmann's constant, $ \Omega(E) $ is the density of states which does not depend on $ \beta $ , and $ Q(\beta) $  is the canonical partition function, that is linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$
-
-Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get:
-
+\label{sec:samplingvalidation}
+In this subsection we present the results from a simple method that serves for checking the consistency between the sampled energy distributions and the theoretical ones.
+The procedure, introduced and implemented by Shirts ~\cite{Shirts2013}, relies on a quantitative statistical analysis of fitting and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
+This last approach does not involve any binning of the energies, hence no discretization error is present.
+We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
+
+Recall that in the NVT ensemble the probability of observing an energy E is given by
+\[
+P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right),
+\]
+where $ \beta $ is the Boltzmann's constant, $ \Omega(E) $ is the density of states which does not depend on $ \beta $ , and $ Q(\beta) $  is the canonical partition function, linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$.
+
+Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get
 \begin{equation}
 \label{eq:probability_ratio}
-\frac{P(E/\beta_2)}{P(E/\beta_1)} = exp\left([\beta_2 A_2 - \beta_1 A_1] - [\beta_2 - \beta_1]E\right)
-\end{equation}
-
-Taking the logarithm of Eq.~\ref{eq:probability_ratio}, we obtain:
+\frac{P(E/\beta_2)}{P(E/\beta_1)} = exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
+\end{equation}
+Taking the logarithm of Eq.~\ref{eq:probability_ratio}, we obtain
 \begin{equation}
 \label{eq:ln_probability_ratio}
-ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = [\beta_2 A_2 - \beta_1 A_1] - [\beta_2 - \beta_1]E
-\end{equation}
-
-Eq.~\ref{eq:ln_probability_ratio} is the basis for the statistical analysis in the ensemble validation, which determines how many standard deviations the calculated slope $ \beta_2 - \beta_1 $ is from the specified one.
+ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E.
+\end{equation}
+
+The Eq.~\ref{eq:ln_probability_ratio} is the basis for the statistical analysis in the ensemble validation, which determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
 Note that it is independent of the unknown free energies.
 For a guidance on how to choose the temperature gap the reader is referred to the aforementioned paper.
-Given that the potential and kinetic energies are separable, the analysis can be made separately:
+Given that the potential and kinetic energies are separable, the analysis can be made separately as follows
 
@@
 \label{eq:probability_ratio_pot}
-\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  exp\left([-\beta_2 - \beta_1]E_{pot}\right)
+\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  exp\left[(-\beta_2 - \beta_1)E_{pot}\right]
 \end{equation}
@@
 \label{eq:probability_ratio_kin}
-\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  exp\left([-\beta_2 - \beta_1]E_{kin}\right)
-\end{equation}
-
-In Fig.~\ref{fig:checkensemble} we show the linear graph for the potential energies sampled using Eq.~\ref{eq:trotter_splitting_NHC}.
-It can be seen that the algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-We arrived at the same conclusions for the other approach given by Eq.~\ref{eq:trotter_splitting_*}.
-
-In table~\ref{table:ensemblevalidation} we examine the results of the maximum likelihood approach.
-It can be seen that the deviations from the true slopes are less than $1\sigma$.
-This means that the sampled values do not deviate from the correct distribution to a statistically noticeable level.
+\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  exp\left[(-\beta_2 - \beta_1)E_{kin}\right]
+\end{equation}
+Furthermore,  it is possible to examine the consistency between the sampled kinetic energy and the corresponding Maxwell-Boltzmann distribution, which is given by the following expression:
+\begin{equation}
+\label{eq:mb}
+P(E_{cin}) = \frac{\beta}{\sqrt{N_{DOF}\pi}} \, exp \left[-\frac{\left(\beta E_{kin} - \frac{N_{DOF}}{2}\right)^2}{N_{DOF}} \right],
+\end{equation}
+where $N_{DOF}$ is the number of degrees of freedom.
+This last equations is a normal distribution characterized by an average of $ \langle E_{kin} \rangle = (N_{DOF}/2\beta) $ and variance $\sigma^2 = N_{DOF}/2\beta^2$, valid when equipartition of energy is satisfied.
+
+In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\ref{eq:trotter_splitting_NHC}.
+In figure legends we show the averages temperatures simulated.
+It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
+Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\ref{eq:trotter_splitting_*}.
 
@@
 \captionof{figure}{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
-Simulation of liquid water using Eq.
-\ref{eq:trotter_splitting_NHC} $\Delta t : 1 fs$.
-$T_1:296.01$, $T_2 = 304.98$ }
+Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t = 1$ fs, $\bar{T}_1=296.03 K$, $\bar{T}_2 = 304.99 K$.
+}
 \label{fig:checkensemble}
@@
 
-\begin{widetext}
-\begin{table}[h]
+\begin{center}
+\includegraphics{maxwell-botlzamm-paper}
+\captionof{figure}{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\ref{eq:mb}.
+Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t = 1$~fs, $\bar{T} = 304.99 K$.}
+\label{fig:mbdistribution}
+\end{center}
+
+Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
+It can be seen that the deviations from the true slope, $\Delta T= 8.970 K $, are less than $1\sigma$.
+This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
+
+\begingroup
+\squeezetable
+\begin{table}
 \setlength{\tabcolsep}{7pt}
-\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach}
+\begin{threeparttable}
+\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach \tnote{a} \tnote{b}}
 \label{table:ensemblevalidation}
 \centering % centering table
-\begin{tabular}{l c  c  c  c  c  c }  
+\begin{tabular}{|l c  c  c  c |}  
 \hline
-& & & \multicolumn{2}{c}{true $\Delta T$ = 8.970 K}\\
-\cline{4-7}
- &true $T_{low} = 296.01 K$ &true $T_{high} = 304.98 K$& \multicolumn{2}{c}{potential} & \multicolumn{2}{c}{kinetic}\\
+ & \multicolumn{4}{c|}{true $\Delta T$ = 8.970 K}\\
+\cline{2-5}
+  & \multicolumn{2}{c}{potential} & \multicolumn{2}{c|}{kinetic}\\
 \hline
-Integrator &$T_{low}$  &$T_{high}$ &$\Delta T$ &$\sigma$ deviation &$\Delta T$&$\sigma$ deviation\\
+Method  &$\Delta T/K$ &$\sigma$ dev &$\Delta T/K$&$\sigma$ dev\\
 \hline % inserts single-line 
- Eq.~$\ref{eq:trotter_splitting_NHC}$   & 296.03 $\pm$ 0.0286 & 304.99 $\pm$ 0.0297 & 9.016 $\pm$ 0.117 & 0.39 & 8.932 $\pm$ 0.045 & 0.84 \\
-  \\ [0.5ex]
- Eq.~$\ref{eq:trotter_splitting_*}$ & 296.03 $\pm$ 0.0286 & 304.99 $\pm$ 0.0297 & 8.956 $\pm$ 0.103 & 0.14 & 8.958 $\pm$ 0.053 & 0.23\\
+ Eq.~$\ref{eq:trotter_splitting_NHC}$   & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
+ Eq.~$\ref{eq:trotter_splitting_*}$   & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14\\
   \hline
 \end{tabular}
+\begin{tablenotes}
+\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.03$
+\item[b]$\bar{T}_1 = 296.03 K $  - $\bar{T}_2 = 304.99 K$
+\end{tablenotes}
+\end{threeparttable}
 \end{table}
-\end{widetext}
-
-In the NVE case, each Liouvillian operator is derived from part of the Hamiltonian, then each factor in the propagator is symplectic and, as a consequence, preserves the phase space volume.
-Recall that the inverse statement is not necessarily true.
-Now it is widely recognized the importance of the symplectic property for numerical stability,\cite{Skeel1997} however it is restricted to hamiltonian systems.
-For non-Hamiltonian equations of motion, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which possess the properties of reversibility and measure-preserving.
-In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to more ergodic trajectory behaviour.
-The invariant-measure for the  Nos\'{e}-Hoover chain system is given by
-\begin{align*}
-\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ]
-\end{align*}
-
-It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*} do not preserve the invariant-measure.
-Nevertheless, those numerical schemes successfully achieve the set point temperature and the long-term energy conservation is acceptable.
-
-Note that the term that premultiplies the extended phase space volume in the invariant-measure is a function of the variable $\eta$.
-Given that the time evolution of the physical variables along with the thermostat momentum does not depend on $\eta$ we conclude that in this case the measure-preserving condition for every individual operator can actually be relaxed.
+\endgroup
+
 \section{Pressure of a System of Rigid Bodies}
 \label{sec:pressure}
-We now derive an expression for calculating the pressure of a rigid body system under periodic boundary conditions.
+We now derive an expression for calculating the pressure of a rigid body system under PBC.
 For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies.
@@
 \]
+
 The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
@@
 \]
-The first term in the expression above accounts for the ideal gas contribution with the corresponding subtraction of three DOF given the conservation of linear momentum. 
+The first term in the expression above accounts for the ideal gas contribution.
 Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
@@
 \end{equation}
-
 Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
@@
 \end{equation}
-
 This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs.
@@
 
-Now we compare the results using Eq.~\ref{eq:pressure} with those from the simulation performed with LAMMPS, which uses a numerical estimation of the constraint forces in the pressure calculation.
+In Figure~\ref{fig:pressure} we compare the results using Eq.~\ref{eq:pressure} with those from a simulation performed with LAMMPS, which uses a numerical estimation of the constraint forces in the pressure calculation.
 The simulations were run in the NVE ensemble, for which neither shifting to the LJ forces nor analytical tail corrections were considered.
-Figure~\ref{fig:pressure} displays the corresponding pressure distributions.
 
@@
 \label{fig:pressure}
-\includegraphics{pressure3}
-\captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS (unfilled bars). Simulation of liquid water with NVE integration. $\Delta t : 1 fs$  }
+\includegraphics[width=0.5\textwidth,keepaspectratio]{FiguraAna}
+\captionof{figure}{Histogram of pressure values obtained using Eq.~\ref{eq:pressure} (filled bars), along with the results from a simulation using LAMMPS (unfilled bars).
+The average pressures are $\bar{P} = -0.60 $ bar and $\bar{P} = 2.14 $ bar, respectively.
+Simulation of liquid water with NVE integration, $\Delta t = 1$ fs, $n_{ys} = 1$}
 \end{center}

### 2016-06-28 rigid_bodies.tex
Tentativa de corrigir git pull
Mudanças na introdução

@@
 \usepackage{multirow}
+
 \begin{document}
@@
 
-One of the strategies commonly used to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
-Widely used iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieve
-
-the dynamics of rigid bodies is able to keep this type of constraint implicitly, and thus emerges as a suitable approach in these cases.
-
-one is able to make an implicit treatment of the holonomic constraints, which is computationally more efficient.
-In addition, and most important, it is possible to obtain a reversible and symplectic integrator in the microcanonical ensemble.
-The property of being symplectic implies the preservation of phase space volume.
-This feature, together with time reversibility are mandatory for numerical integrators used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic reversible integrator for the rotational motion in terms of the unit quaternion.
-This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems including the ubiquitous case of water \cite{Sakamaki2011,Reinhardt2012,Palmer2014,Gonzales2014}, ice water \cite{Geiger2014}, hydrates \cite{Tribello2009,Gorman2012}, diffusing gases in metal-organic frameworks \citep{Ghoufi2010}, etc.
-It is worth noting that it also can be used for systems designed as a collection of interconnected rigid bodies, which is a coarse-graining strategy widely used for proteins \citep{Terada2003}, molecular machines \citep{Akimov2008,Konyukhov2010}, nanoparticles \cite{Knorowski2012,Patra2013}, etc.
-
-In this paper, we present a simplified version for the equations of motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as the corresponding numerical integrators to perform MD.
-The contribution to the NVE case relies on a factorization of the rotation matrix, which allows a Hamiltonian formalism that avoids the use of the extended phase space methodology, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002}.
-In the NVT case, both the translational and rotational degrees of freedom (DOF) are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat.
-This differs from the approach of Kamberaj \textit{et al}.\cite{Kamberaj2005}, which uses two independent thermostats, one for the rotational and another one for the translational DOF.
-We have also simplified the calculation of the pressure for a rigid body system, whose constituents atoms interact via a pairwise potential and are subjected to periodic boundary conditions (PBC).
-The expression we derive depends on the resultant interaction forces between rigid body pairs.
-This implies that there is no need to evaluate the unknown forces of constraint, which one must take into account if the pressure is calculated in terms of the interatomic forces.
-
-This paper is organized as follows: In Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism.
+One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
+Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion.
+This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles.
+However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach.
+It is expected to be computationally more efficient because the constraints are treated implicitly.
+In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
+Being symplectic implies phase-space volume preservation and time reversibility, which are mandatory properties for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
+This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems containing rigid molecules, including liquid water, \cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc.
+It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
+
+In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as corresponding numerical integrators.
+The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat.
+This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two independent thermostat chains act separately on the rotational and the translational degrees of freedom.
+We have also simplified the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions (PBC).
+The expression we derive depends on the resultant interaction forces between body pairs.
+This avoids the evaluation of constraint forces, which are necessary if the pressure is calculated in terms of the interatomic forces.
+
+This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism.
 In Sec.~\ref{sec:moleculardynamics} we present the numerical integrators for both the NVE and NVT ensembles, whose performance we discuss in terms of the conservation of the total energy.

### 2016-06-29 rigid_bodies.tex
Changes in Section II - The title of Sec II.B was changed - Reference to Schay 1995 was added in the section - Derivatives dU/dr and dU/dq are now presented before the equation of motion - The equations of motion now have force and torque already substituted

@@
 In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
-Being symplectic implies phase-space volume preservation and time reversibility, which are mandatory properties for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this sense, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
+Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
 This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems containing rigid molecules, including liquid water, \cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc.
@@
 
-This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion followed by the Hamiltonian formalism.
-In Sec.~\ref{sec:moleculardynamics} we present the numerical integrators for both the NVE and NVT ensembles, whose performance we discuss in terms of the conservation of the total energy.
-Additionally, for the NVE case we present some results related to the partition of kinetic energy and for the NVT case, we also analyze the statistical distributions of the sampled energy.
-Following this, in Sec.~\ref{sec:pressure} we show the equation for the pressure developed in this work.
-Along this paper, we employ the word quaternion as a synonym for a vector in $R^4$ and avoid other interpretations.
+This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion of rigid bodies using a Hamiltonian formalism.
+In Sec.~\ref{sec:moleculardynamics} we devise numerical integrators for both the NVE and NVT ensembles, whose correctness we discuss in terms of conserved quantities.
+Additionally, for the NVE case we present some results related to kinetic energy partition and, for the NVT case, we also analyze the statistical distributions of the sampled energy.
+After that, in Sec.~\ref{sec:pressure} we present the equation developed for the pressure.
+
+Note that, throughout this paper, we employ the word ``quaternion'' as a synonym for a four-dimensional vector and avoid other interpretations.
 Only ordinary matrix/vector operations are present in the formulation.
@@
 
-In this section we first review certain aspects of the kinematics or rigid bodies we consider relevant to this study.
-In Sec~\ref{sec:eulerparameters}, we introduce one particular factorization of the rotation matrix from which we obtain a mathematical relation between the rate of change of the four-dimensional quaternion and the angular velocity vector, while automatically satisfying the restriction imposed on the quaternion.
+In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study.
+Unity quaternions are employed to represent the orientations of rigid bodies.
+Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector.
+An important feature of this relation is that it preserves the unit-norm constraint of the quaternion upon infinitesimal motion.
 
@@
 
-Consider a rigid body composed of $n_p$ individual particles, each particle $j$ with mass $M_j$ and located at a Cartesian coordinate $\vt R_j$. The center of mass of this body is located at
+Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$.
+The center of mass of this body is located at
 \[
@@
 where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body.
-The position of each particle $j$ in a space-fixed reference frame whose origin coincides with the body's center of mass is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$.
+The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$.
 In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
@@
 -{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
-\end{array} \right],
+\end{array} \right].
 \]
@@
 For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal.
-The corresponding axes are known as the principal inertia axes and the formulation of the equations of motion in this frame of reference is notably simplified.
+The corresponding axes are known as the principal axes and the formulation of the equations of motion in this frame of reference is notably simplified.
 Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
@@
 \]
-where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are the corresponding eigenvectors.
+where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors.
 Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}.
 Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$.
-Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix.
-If we differentiate both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix.
-This product is related to $\vt \omega$, the angular velocity in the body-fixed frame, by\cite{Haug1989}
+Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$ .
+By differentiating both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix.
+This product is related to $\vt \omega$, the angular velocity of the body in the body-fixed frame, by\cite{Haug1989}
 \begin{equation}
@@
 \end{equation}
-The operator  $\mt S(\cdot)$ can be used to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.
+
+The operator  $\mt S(\cdot)$ is able to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.
+This property is explored in the Appendix.
 
 %\subsection{Euler Parameters and Unit Quaternions}
-\subsection{The Rate of change of Euler Parameters}
+\subsection{The Rate of Change of Unit Quaternions}
 \label{sec:eulerparameters}
@@
 Euler's rotation theorem establishes that an arbitrary rotation can be described using three independent parameters.
-One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in terms of those angles.
+One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in such terms.
 Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$.
-The mathematical relationship between the Euler angles and the quaternion places the restriction over $\boldsymbol{q}$:\cite{Goldstein2002}
+The mathematical relation between the Euler angles and the Euler parameters requires that\cite{Goldstein2002}
 \begin{equation}
@@
 \end{equation}
-Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $R^4$.
+
+Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain.
 Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix.
@@
 
-The rate of change of the Euler parameters is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
+The rate of change of a quaternion is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
 \begin{equation}
 \label{eq:diff_qTq}
-\tr{\vt q}\dot{\vt q} = 0.
-\end{equation}
-This equation implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.
+\tr{\vt q}\dot{\vt q} = 0,
+\end{equation}
+which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.
 
@@
 \end{subequations}
-Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited in Appendix \ref{sec:auxiliary_math} with our notation.
+
+Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited with our particular notation in Appendix~\ref{sec:auxiliary_math}.
 The most remarkable feature is that
@@
 \end{equation}
-This is a simple and convenient factorization, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters.
+
+The equation above is a simple and convenient factorization of matrix $\mt A$, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters.
 As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters.
@@
 \label{sec:hamiltonian}
-In the previous subsection we obtained the relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}), which is used in this part to develop the Hamiltonian formalism of the equations of motion.
-As usual, the dynamics of a rigid body can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively.
+
+The relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body.
+As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively.
 The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$.
-Several authors have obtained Hamiltonian equations of motion for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.
+Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.
 
 We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
-Since $K = 1/2 m \tr{\dot{\vt r}} \dot{\vt r} + 1/2 \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
+Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
 \[
@@
 \end{equation}
-Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$.
+
+Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$ (premultiplication of $\mt I \vt \omega$ by $\mt B$).
 Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
@@
 \]
-it follows that $\tr{\mt B}{\vt \pi} = -\tr{\mt \Omega}{\vt q}$ and, therefore,
+it follows that $\tr{\mt B}{\vt \pi} = \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
 \begin{equation}
@@
 \end{equation}
-Now this is convenient for differentiation with respect to $\vt q$.
-Hence, after carrying out the differentiations as described in Appendix \ref{sec:Diff_Rules} and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following system of differential equations to describe the dynamics of a rigid body:
+
+For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
+\[
+\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
+\]
+where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body.
+To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}.
+The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}.
+As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
+\[
+\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
+\]
+where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame.
+
+Finally, after carrying out the needed differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
 \begin{subequations}
@@
 &\dot{\vt r} = \frac{1}{m} \vt p \\
-&\dot{\vt p} = -\diff{U}{\vt r} \\
+&\dot{\vt p} = \mt F \\
 &\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
-&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega - \diff{U}{\vt q}
+&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau
 \end{align}
@@
 
-For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
-\[
-\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
-\]
-where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body.
-To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}.
-As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
-\[
-\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
-\]
-where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.
-
 Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
-Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$ and according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
+Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
 \begin{equation}
 \label{eq:omega_entry}
-\omega_k = \frac{\tr{\vt \pi} {\mt B}_k \vt q}{2 I_k},
+\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
 \end{equation}

### 2016-07-04 rigid_bodies.tex
Update of section Molecular Dynamics

@@
 %\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
-\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
+\documentclass[aip,jcp,reprint,amsmath,amssymb,amsfont]{revtex4-1}
 \usepackage[pdftex]{graphicx}
@@
 \newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
-\newcommand{\tr}[1]{#1^T}                               % transposition
+\newcommand{\tr}[1]{#1^t}                               % transposition
 \newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
-\mathchardef\mhyphen="2D
+
+\raggedbottom
+
 \usepackage{booktabs}
@@
 \email{asilveira@plapiqui.edu.ar}
-\affiliation{Chemical Engineering Department, Escola de Qu\'imica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}
 \affiliation{Planta Piloto de Ingenier\'ia Qu\'imica, PLAPIQUI, Universidad Nacional del Sur, Camino La Carrindanga Km 7-CC: 717, Bah\'ia Blanca, Argentina}
@@
 \email{abreu@eq.ufrj.br}
-\affiliation{Chemical Engineering Department, Escola de Quimica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}
+\affiliation{Chemical Engineering Department, Escola de Qu\'imica, Universidade Federal do Rio de Janeiro, Rio de Janeiro, RJ 21941-909, Brazil}
 
 \date{\today}
+
+\begin{abstract}
+Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+\end{abstract}
 
@@
 Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
-This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide range of systems containing rigid molecules, including liquid water, \cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc.
+This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc.
 It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
@@
 Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$ .
-By differentiating both sides of this relation with respect to time, we conclude that $\mt A \tr{\dot {\mt A}} = - \dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix.
+By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix.
 This product is related to $\vt \omega$, the angular velocity of the body in the body-fixed frame, by\cite{Haug1989}
@@
 \end{equation}
-where $\mt S(\cdot)$ is the skew-symmetric operator defined as
-\begin{equation}
-\label{eq:operator_S}
-\mt S(\vt x) = \left[ \begin{array}{ccc}
- 0   & -x_3 &  x_2 \\
- x_3 &  0   & -x_1 \\
--x_2 &  x_1 &  0
-\end{array}\right].
-\end{equation}
-
-The operator  $\mt S(\cdot)$ is able to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.
-This property is explored in the Appendix.
-
-%\subsection{Euler Parameters and Unit Quaternions}
+where $\mt S(\cdot)$ is the skew-symmetric operator defined in Eq.~\ref{eq:operator_S}.
+
 \subsection{The Rate of Change of Unit Quaternions}
@@
 
-The relation between $\dot{\boldsymbol{q}}$ and $\boldsymbol{\omega}$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body.
+The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body.
 As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively.
 The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$.
-Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using Euler parameters as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.
+Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.
 
@@
 \]
-where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame.
-
-Finally, after carrying out the needed differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = (1/2) {\mt I}^{-1} \tr{\mt B} \vt \pi = (1/2) {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
+where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame. Note that the product $\mt C \vt \tau$ results in a quaternion orthogonal to $\vt q$.
+
+Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B} \vt \pi = \frac{1}{2} {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
 \begin{subequations}
@@
 \end{equation}
-where $\mt B_k$ is the corresponding permutation matrix among those presented in Appendix \ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
+where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
 \[
@@
 
-In order to rewrite Eq.~\ref{eq:EDO_system}, the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ are opened using Eq.~\ref{eq:product_B_vector}, leading to the final form for the equations of motion:
+In order to rewrite Eq.~\ref{eq:EDO_system}, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to a new form for the equations of motion:
 \begin{subequations}
@@
 &\dot{\vt p} = \vt F \\
-&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k {\mt B}_k \vt q \label{eq:edo_q} \\
-&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k {\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
+&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
+&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
 \end{align}
@@
 \label{symplectic}
-An exact solution for the system of equations represented by Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
-Hence, it would describe a conservative dynamics with respect to the Hamiltonian $\mathcal{H}$, in which phase space volume element would be preserved.
-Nevertheless, a general analytical solution does not exist.
-This is so, above all, because $U(\vt r, \vt q)$ is arbitrary, but even in the simplest case of free motion (i.e., when $U$ is constant), no closed-form solution is known due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}.
-Therefore, it becomes necessary to introduce the Liouville operators formalism, which allows us to derive symplectic numeric integrators rigorously.
-
-For a given trajectory in phase-space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
-\[
-\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i L f,
-\]
-where $i L$ is the Liouville operator.
-This equation has a formal solution expressed by $f(t) = e^{i L t}f_0$, where $f_0$ is the initial condition for $f$.
-As $f$ can be a phase-space coordinate, premultiplication with the exponential operator $e^{i L t}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} normally represents a transformation from the initial condition to the phase-space point occupied at time $t$.
+
+An exact solution of Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
+Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
+However, no closed-form solution is known, even in the simplest case of free motion (i.e., when $U$ is constant), due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}.
+Nevertheless, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010}
+
+For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
+\[
+\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
+\]
+where $i\!L$ is the Liouville operator.
+This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$.
+As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$.
 Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.
@@
 Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}.
-Let us analyze three of them, which will be useful later on to derive a complete numerical solution.
-In what follows, subscript $0$ always denotes ``initial values''.
+Let us analyze three of them, which will be useful later on to derive a complete approximate solution.
+In what follows, subscript $0$ denotes ``initial values''.
 We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$.
@@
 \end{align*}
-Clearly, this corresponds to a uniform linear translation.
-Since the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected.
-Of course, the solution above is exactly the effect of the propagator $e^{i L_t t}$, where the Liouville operator $i L_t$ is, simply,
+
+This corresponds to a free translation (hence the subscript $t$).
+As the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected.
+Of course, the solution above is the exact effect of the propagator $e^{t i\!L_t}$, where the Liouville operator $i\!L_t$ is, simply,
 \[
@@
 
-In the second hypothetical case, referred to as the acceleration case, we consider a Hamiltonian $\mathcal{H}_a = U(\vt r, \vt q)$.
+In the second hypothetical case, referred to as the boost case, we consider a Hamiltonian $\mathcal{H}_b = U(\vt r, \vt q)$.
 The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
@@
 \end{align*}
-Note that $\vt r$ and $\vt q$ remain unaltered, thus keeping $\vt F$, $\mt C$, and $\vt \tau$ constant as well (hence the subscript $0$ assigned to them).
-This solution clearly conserves $\mathcal{H}_a$, and marks the effect of a propagator $e^{i L_a t}$, obtained from the operator
-\[
-i L_a = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
+
+Note that $\vt r$ and $\vt q$ remain unaltered, thus keeping $\vt F$, $\mt C$, and $\vt \tau$ constant as well, which justifies the subscript $0$ assigned to them.
+This solution clearly conserves $\mathcal{H}_b$ and marks the effect of a propagator $e^{t i\!L_b}$, obtained from the operator
+\[
+i\!L_b = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
 \]
@@
 Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation.
-The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = (1/2) \omega_k {\mt B}_k \vt q$, and $\dot{\vt \pi} = (1/2) \omega_k {\mt B}_k \vt \pi$.
-A special feature of these equations makes them analytically solvable.
+The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt q$, and $\dot{\vt \pi} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt \pi$.
+A special feature of this system facilitates its solution.
 It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well.
 As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form.
-The following analytical solution takes place:\cite{Miller2002}
-\begin{align}
+The following analytical solution then takes place:\cite{Miller2002}
+\begin{align*}
 &{\vt r}(t) = {\vt r}_0 \\
 &{\vt p}(t) = {\vt p}_0 \\
-&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \mt B_k \vt q_0 \\
-&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \mt B_k \vt \pi_0
-\end{align}
-These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{i L_k t}$ derived from the operator
-\[
-i L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{{\mt B}_k} \diff{}{\vt \pi} \right).
+&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
+&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
+\end{align*}
+
+These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{t i\!L_k}$ derived from the operator
+\[
+i\!L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{\hat{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{\hat{\mt B}_k} \diff{}{\vt \pi} \right).
 \]
@@
 \label{eq:full_operator}
-i L_{NVE} = i L_t + i L_a + \sum_{k=1}^3 i L_k,
+i\!L = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k.
 \end{align}
 
-The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic geometry of Eq.~\ref{eq:EDO_system_alternative}.
-Such formula is
-\[
-e^{i L_A t + i L_B t} \approx \left( e^{i L_B \frac{\Delta t}{2}} e^{i L_A \Delta t} e^{i L_B \frac{\Delta t}{2}} \right)^N + \mathcal{O}(\Delta t^2),
+The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}.
+For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
+\[
+e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^N + \mathcal{O}(\Delta t^2),
 \]
 where $\Delta t = t/N$ is the time step size and $N$ is the number of steps.
-In the next section we show the factorization schemes we considered for generating the dynamics in both the NVE and NVT ensembles.
+We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side.
+Splitting schemes with higher-order approximations have been used for rigid body dynamics,\cite{Omelyan2007, Omelyan2008, vanZon2008} but they result in methods which are more complex and often require expensive force-gradient calculations.
+Moreover, it is unclear whether their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than NVE.
+
+By defining $i\!L_r = i\!L_1 + i\!L_2 + i\!L_3$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ commute, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
+\begin{equation}
+\label{eq:trotter_splitting_NVE}
+e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_t} \left[ \prod_{j=1}^n e^{{\delta t}_j i\!L_{r}} \right] e^{\frac{\Delta t}{2} i\!L_b},
+\end{equation}
+where $\sum_{j=1}^n {\delta t}_j = \Delta t$.
+This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
+Each complete rotation is split as\cite{Miller2002}
+\begin{equation}
+\label{eq:splitting_rot}
+e^{\delta t_j i\!L_r} = e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3}.
+\end{equation}
+
+Following Ref.~\onlinecite{Miller2002}, one could simply make $\delta t_j = \Delta t/n$ for all $j$.
+Alternatively, one could use a higher-order factorization scheme, such as the one due to Yoshida\cite{Yoshida1990} and Suzuki,\cite{Suzuki1991a, Suzuki1991b} in which $\delta t_j = \omega_j \Delta t$ and the weights $\omega_j$ are carefully chosen to determine the order of approximation.
+It is our intention in this paper to investigate how the performance of the integrator as a whole is affected by using such a higher-order scheme.
+For that, we will consider $n = 1$ with $\omega_1 = 1$ (zero-order) and $n = 7$ with $\omega_1 = \omega_7 = 0.784513610477560$, $\omega_2 = \omega_6 = 0.235573213359357$, $\omega_3 = \omega_5 = -1.17767998417887$, and $\omega_4 = 1 - \sum_{j \neq 4} \omega_j$ (sixth-order).
 
@@
 
+In this section, we present the factorization schemes we considered for generating molecular dynamics algorithms in both the NVE and NVT ensembles.
+
 \subsection{Microcanonical Ensemble}
-
-The splitting of the Liouville operator (Eq.~\ref{eq:full_operator}) we employ in this paper is similar to that proposed by Miller \textit{et al}.\cite{Miller2002}, but extended to include translation and linear acceleration.
-It is
-
-\begin{equation}
-\label{eq:trotter_splitting_NVE}
-e^{i L_{NVE} \Delta t} = e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L_a \frac{\Delta t}{2}},
-\end{equation}
-where $L_r = \sum_{k=1}^3 L_k$ is the Liouville operator for a complete rotation and, $i L_a $, $i L_k $ as well as $i L_t $ retain its previous definitions.
-Recall that each exponential operator in Eq.~\ref{eq:trotter_splitting_NVE} will act having the result of the preceding operator (the one at its right hand side) as initial condition.
-Following Ref.~\onlinecite{Miller2002} once again, the splitting employed for the rotational part is\cite{Tuckerman1992}
-
-\begin{equation}
-\label{eq:splitting_rot}
-e^{i L_r \Delta t} = \left( e^{i L_3 \frac{\delta t}{2}} e^{i L_2 \frac{\delta t}{2}} e^{i L_1 \delta t} e^{i L_2 \frac{\delta t}{2}} e^{i L_3 \frac{\delta t}{2}} \right)^n,
-\end{equation}
-where $\delta t = \nicefrac{\Delta t}{n}$.
-This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step to improve accuracy.
-An alternative to a further factorization of Eq.~\ref{eq:splitting_rot} with higher values of $n$ could be the use of a higher-order scheme, as the one due to Yoshida \cite{Yoshida1990} and Suzuki\cite{Suzuki1991a,Suzuki1991b}, given by
-
-\begin{equation}
-\label{eq:ys_rot}
-e^{i L_{\mathrm{r\mhyphen ys}} \Delta t} = \prod_{j=1}^{n_{sy}} e^{i L_{r} \omega_j \Delta t},
-\end{equation}
-where $\omega_j$ are used to weight the time step and $n_{ys}$ determines the order of this method.
-Note that we are not improving the order of precision for the integration of the translational DOF, which is computationally more demanding.
-Nevertheless, we would like to investigate how the performance of the integrator as a whole is affected by using a higher-order scheme for the rotational operator.
-For that purpose, in this paper we consider $n = 1$ with $n_{ys} = 1 $ (zero-order) and $n_{sy} = 7$ (sixth-order).
-In this case, the weights are:
-\begin{align*}
-&\omega_1 = \omega_7 = 0.784513610477560 \\
-&\omega_2 = \omega_6 = 0.235573213359357 \\
-&\omega_3 = \omega_5 = -1.17767998417887 \\
-&\omega_4 = 1 - \omega_1 -\omega_2 - \omega_3 - \omega_5 - \omega_6 - \omega_7
-\end{align*}
-
-It should be mentioned here that one can find in literature several numerical integrators\cite{Omelyan2007,Omelyan2008,vanZon2008} for NVE rigid body dynamics which are more accurate than the symplectic Verlet-type integrators.
-However, since the application of those methods is limited to Hamiltonian systems, it's unclear whether their benefits will extend to the numerical integrators in the NVT ensemble.
-This, together with its great applicability, are the main reasons why our approach is based on the Verlet-type integrators.
 
@@
 \label{sec:canonical}
-The linear momentum equation for a rigid body considering a Nos\'{e}-Hoover thermostat is
-\begin{align*}
-\dot{\vt p} &= \vt F^d = {\vt F} - \frac{p_{\eta}}{Q} {\vt p} = \sum_{j=1}^{n_p} {\vt F}_j - \frac{p_{\eta}}{Q} \sum_{j=1}^{n_p} {\vt p}_j \\
-&= \sum_{j=1}^{n_p} \left( {\vt F}_j - \frac{p_{\eta}}{Q} {\vt p}_j \right).
-\end{align*}
-Thus, the thermostat affects each particle $j$ individually. Analogously, the resultant torque is affected by the thermostat according to
-\begin{align*}
-\vt \tau^d &= \sum_{j=1}^{n_p} {\vt \delta_j} \times \left( {\vt F_j}  - \frac{p_{\eta}}{Q} {\vt p}_j \right) = \\
-&= \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j} - \frac{p_{\eta}}{Q} \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt p}_j = \\
-&= \vt \tau - \frac{p_{\eta}}{Q} \vt L,
-\end{align*}
-where $\vt L$ is the angular momentum of the rigid body expressed in the space-fixed frame of reference.
-Thus, $\vt L = \tr{\mt A}{\mt I}{\vt \omega}$.
-Resorting to Eqs.~\ref{eq:factorization_of_A} and \ref{eq:conj_momentum}, we observe that
-\[
-\vt L = \tr{\mt C}{\mt B}{\mt I}{\vt \omega} = \frac{1}{2} \tr{\mt C}{\vt \pi}.
-\]
-Therefore, replacing $\vt \tau$ by $\vt \tau^d$ in the quaternion momentum equation, we have its thermostated version as
-\[
-\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau - \frac{p_{\eta}}{Q} \vt \pi.
-\]
-Finally, the Nos\'{e}-Hoover chain thermostat applied to a system of $N$ rigid bodies becomes:
+
+Kamberaj \textit{et al.}\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble...........
+
+Understanding of the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space, which is the cornerstone for deterministic molecular dynamics at different ensembles.
+Sergi and Ferrario\cite{Sergi2001} described a simple way of systematically generating conservative trajectories in an arbitrary phase space.
+If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
+\begin{equation} \label{eq:eq_of_motion}
+\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
+\end{equation}
+will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric.
+This can be shown by using the chain rule to express the time-derivative of an arbitrary scalar function $f(\vt x)$, which yields
+\[
+\dot{f} = \tr{\left(\diff{f}{\vt x}\right)}\dot{\vt x} = \tr{\left(\diff{f}{\vt x}\right)} \boldsymbol{ \mathcal B } \diff{H}{\vt x}.
+\]
+
+If $f = H$, the equation above turns into a quadratic form, which is identically null if $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{\mathcal B}$, thus resulting in $\dot H = 0$.
+
+In what follows, index $i$ ranges for all $N$ particles and index $j$ ranges for all $M$ thermostats, unless otherwise stated.
+
+\begin{equation}
+\label{eq:H_nvt}
+\begin{split}
+H = U(\vt r, \vt q) + \sum_{i=1}^N \left( \frac{\tr{\vt p}_i {\vt p}_i}{2 m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8} \right) + \\
++ \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j
+\end{split}
+\end{equation}
+
+The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
+$\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$.
+However, the equations of motion for generalized momenta contain non-Hamiltonian terms.
+For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
+\[
+\begin{split}
+\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \qquad \text{and}\\
+\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
+\end{split}
+\]
+
+In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat.
+In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, plus a term regarding the action of a second thermostat.
+The result is
+\[
+{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
+\]
+
+Each other thermostat in the chain acts on the previous one, and is acted upon by the following one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
+\[
+{\dot p}_{\eta_j} = -\diff{H}{\eta_j} + p_{\eta_{j-1}} \diff{H}{p_{\eta_{j-1}}} - p_{\eta_j} \diff{H}{p_{\eta_{j+1}}}.
+\]
+
+Finally, the last thermostat just acts on 
+\[
+{\dot p}_{\eta_M} = -\diff{H}{\eta_M} + p_{\eta_{M-1}} \diff{H}{p_{\eta_{M-1}}}.
+\]
+
+For a three-dimensional system composed of non-collinear rigid bodies, the phase-space compressibility is given by\cite{Tuckerman2010}
+\[
+\kappa = -6N \diff{H}{p_{\eta_1}} - \sum_{j=2}^M \diff{H}{p_{\eta_j}} = -6N \dot{\eta}_1 - \sum_{j=2}^M \dot{\eta}_j.
+\]
+
+Thus, the metric determinant factor of the extended phase space is then given by $\sqrt{g} = e^{6N \eta_1 - \sum_{j=2}^M \eta_j}$.\cite{Tuckerman_1999, Tuckerman2001}
+
+Finally, the Nos\'{e}-Hoover chain thermostat applied to a system of $N$ rigid bodies becomes
 \begin{subequations}
@@
 &\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
-&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i  \\
+&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \\
 &\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \\
-&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q} \vt \pi_i \\
-&\dot{\eta}_j = \frac{p_{\eta_j}}{Q_j} & j = 1,\cdots,M \\
-&{\dot p}_{\eta_j} = G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} & j = 1,\cdots,M-1 \label{eq:p_eta_j} \\
-&{\dot p}_{\eta_M} = G_M
+&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \\
+&\dot{\eta}_j = \alpha_j \\
+&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}
 \end{align}
 \end{subequations}
-where
+where $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$.
 \begin{align*}
-&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - 6 N k_B T \\
-&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T &j = 2,\cdots,M \\
-&Q_1 = 6 N k_B T t_d^2 \\
-&Q_j = k_B T t_d^2 &j = 2,\cdots,M
+&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - 6N k_B T \\
+&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad (j > 1)
 \end{align*}
+
+$Q_1 = 6 N k_B T t_d^2$ and $Q_j = k_B T t_d^2$
+
 In these equations, $i$ indexes the number of rigid bodies from $1$ to $N$, $M$ is the number of thermostats in the chain, $T$ is the specified temperature, $k_b$ is Boltzmann's constant, $6N$ is the number of DOF, and $t_d$ is the characteristic time period of the thermostat chain.
@@
 
-As was done in the Hamiltonian case, the evolution of the extended phase space vector $ f(\vt r, \vt p, \vt q, \vt \pi, \eta_1,..
-\eta_M, p_{\eta_1},..,p_{\eta_M}) $ is given by the relation $f(t) = e^{i L t}f_0$ in the context of the generalized Liouville theorem\cite{Tuckerman_1999}.
-This implies that we can employ the Trotter theorem to devise a strategy for integrating the equations of motion.
-
-We consider first the usual separation of the non-Hamiltonian component of the equations of motion from the Hamiltonian component. Then, the Liouville operator can be written as follows
-\[
-i L = i L_{NHC} + i L_{NVE},
-\]
-where $i L_{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $ i L_{NHC} $ is given by
-
+\subsection{Integration Schemes for NVT Molecular Dynamics}
+
+As done in the Hamiltonian case, the evolution of an extended phase space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion.
+First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components.
+In this case, the Liouville operator can be written as
+\[
+i\!L = i\!L_\text{NHC} + i\!L_\text{NVE},
+\]
+where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $i\!L_\text{NHC}$ is given by
 \begin{equation}
 \begin{split}
-& i L_{NHC} = -\sum_{i=1}^{N} \frac{p_{\eta_1}}{Q_1} \vt p_i \diff{}{\vt p_i} - \sum_{i=1}^{N} \frac{p_{\eta_1}}{Q_1} \vt \pi_i \diff{}{\vt \pi_i} \\
-& + \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}} {Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right],\\
+i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) \\
+&+ \sum_{j=1}^M \left[ \alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
 \end{split}
 \end{equation}
-with $p_{\eta_{M+1}}/Q_{M+1} = 0$.
-
-As pointed out by Martyna~\textit{et al.}~\cite{Martyna1996}, the part of the operator given by 
-\begin{align*}
-&\left(G_j - p_{\eta_j} \frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}},
-\end{align*}
-can be evaluated analytically using the following equation:
+
+As pointed out by Martyna~\textit{et al}.,\cite{Martyna1996} the part of the operator given by $(G_j - \alpha_j p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}$ can be evaluated analytically using the following equation:
 \begin{equation}
@@
 \begin{split}
-&p_{\eta_j} \leftarrow p_{\eta_j} + \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \phi\left( \frac{p_{\eta_{j+1}}}{Q_{j+1}} \Delta t \right) \Delta t \\
+&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1} p_{\eta_j}^0 \right) \phi(\alpha_{j+1} t) t, \\
 \end{split}
 \end{equation}
-where $\phi(x) = 1-e^{-x}/x = \sum_{n=0}^\infty (-1)^n x^n/(n+1)!$. For small values of $|x|$, the function $\phi(x)$ can be evaluated by
-
-\begin{equation}
-\label{eq:phi}
-\phi(x) = \begin{cases}
-\frac{1-e^{-x}}{x} & \text{if} \; |x| > 10^{-4} \\
-1 - \frac{x}{2}\left[1 - \frac{x}{3}\left(1 - \frac{x}{4}\right) \right] & \text{if} \; |x| \leq 10^{-4}
-\end{cases}
-\end{equation}
-
-Considering the above analytical solution, the NHC operator is factorized as follows
-
+where $\phi(x) = (1-e^{-x})/x$.
+For small absolute values of $x$, this function can be evaluated by truncating its Taylor series form $\phi(x) = \sum_{n=0}^\infty \frac{(-1)^n x^n}{(n+1)!}$.
+Four terms result in a sufficiently precise value if $\|x\| \leq 10^{-4}$.
+Considering the analytical solution above, the NHC operator is factorized as
 \begin{widetext}
@@
 \begin{split}
-e^{iL_{NHC} \frac{\Delta t}{2}} =  &\prod_{j=M}^{1} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-&  \times \prod_{j=1}^{M} exp\left(\frac{\Delta t}{2} \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)  \prod_{i=1}^{N} exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt p_i^T \diff{}{\vt p_i}\right) \prod_{i=1}^{N} exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T \diff{}{\vt \pi_i}\right)  \\
-& \times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
+e^{i\!L_\text{NHC} \frac{\Delta t}{2}} =  &\prod_{j=M}^{1} \exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
+&  \times \prod_{j=1}^{M} \exp\left(\frac{\Delta t}{2} \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)  \prod_{i=1}^{N} \exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\tr{\vt p}_i \diff{}{\vt p_i}\right) \prod_{i=1}^{N} \exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\tr{\vt \pi}_i \diff{}{\vt \pi_i}\right)  \\
+& \times \prod_{j=1}^{M} \exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
 \end{split}
@@
 Given that the time evolution of the physical variables along with the thermostat momentum does not depend on $\eta$ we conclude that in this case the measure-preserving condition for every individual operator can actually be relaxed.
+
+\subsection{Pressure of a System of Rigid Bodies}
+\label{sec:pressure}
+
+We now derive an expression for calculating the pressure of a rigid body system under PBC.
+For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies.
+This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended.
+
+The potential energy per unit cell of a system with $N$ distinct rigid bodies (indexed from $1$ to $N$) and periodic boundary conditions is
+\begin{equation}
+\label{eq:U_pairwise_pbc}
+U({\vt r}^N, {\vt q}^N) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
+\end{equation}
+where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
+\begin{equation}
+\label{eq:u_rigid_body_pair}
+u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
+\end{equation}
+
+In the equation above, $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$.
+The dependence of this potential on the types of these two atoms is omitted for simplicity.
+The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
+\[
+{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
+\]
+
+The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
+\[
+P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
+\]
+The first term in the expression above accounts for the ideal gas contribution.
+Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
+\begin{equation}
+\begin{split}
+\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right.
+\\ \left.
+\diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
+\end{split}
+\end{equation}
+Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
+Therefore,
+\[
+\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \diff{u}{{\vt r}_{ij}} \cdot {\vt r}_{ij}.
+\]
+
+We can now use Eq.~\ref{eq:u_rigid_body_pair} to obtain the term inside the double summation above. By the chain rule, we have
+\[
+\diff{u}{{\vt r}_{ij}} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \diff{\vt r}{\vt r_{ij}} \diff{\upsilon}{\vt r},
+\]
+where $\vt r = {{\vt R}_i}_k - {{\vt R}_j}_l$.
+Notice that, inside the double summation above, the first derivative is the identity matrix and the second one is $-{\vt f}_{ijkl}$, with this force being the action of the $l$-th particle of body $j$ over the $k$-th particle of body $i$.
+Therefore, the equation above results in $-\vt F_{ij}$, with this force being the resultant action of body $j$ over body $i$.
+Finally, we conclude that
+
+\begin{equation}
+\label{eq:pressure}
+P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
+\end{equation}
+This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs.
+Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.
 
@@
 
-\section{Pressure of a System of Rigid Bodies}
-\label{sec:pressure}
-We now derive an expression for calculating the pressure of a rigid body system under PBC.
-For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies.
-This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended.
-
-The potential energy per unit cell of a system with $N$ distinct rigid bodies (indexed from $1$ to $N$) and periodic boundary conditions is
-\begin{equation}
-\label{eq:U_pairwise_pbc}
-U({\vt r}^N, {\vt q}^N) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
-\end{equation}
-where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
-\begin{equation}
-\label{eq:u_rigid_body_pair}
-u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
-\end{equation}
-
-In the equation above, $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$.
-The dependence of this potential on the types of these two atoms is omitted for simplicity.
-The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
-\[
-{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
-\]
-
-The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
-\[
-P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
-\]
-The first term in the expression above accounts for the ideal gas contribution.
-Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
-\begin{equation}
-\begin{split}
-\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right.
-\\ \left.
-\diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
-\end{split}
-\end{equation}
-Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
-Therefore,
-\[
-\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \diff{u}{{\vt r}_{ij}} \cdot {\vt r}_{ij}.
-\]
-
-We can now use Eq.~\ref{eq:u_rigid_body_pair} to obtain the term inside the double summation above. By the chain rule, we have
-\[
-\diff{u}{{\vt r}_{ij}} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \diff{\vt r}{\vt r_{ij}} \diff{\upsilon}{\vt r},
-\]
-where $\vt r = {{\vt R}_i}_k - {{\vt R}_j}_l$.
-Notice that, inside the double summation above, the first derivative is the identity matrix and the second one is $-{\vt f}_{ijkl}$, with this force being the action of the $l$-th particle of body $j$ over the $k$-th particle of body $i$.
-Therefore, the equation above results in $-\vt F_{ij}$, with this force being the resultant action of body $j$ over body $i$.
-Finally, we conclude that
-
-\begin{equation}
-\label{eq:pressure}
-P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
-\end{equation}
-This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs.
-Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.
+\subsection{Pressure Calculations}
 
@@
 
+\begin{equation}
+\label{eq:operator_S}
+\mt S(\vt x) = \left[ \begin{array}{ccc}
+0   & -x_3 &  x_2 \\
+x_3 &  0   & -x_1 \\
+-x_2 &  x_1 &  0
+\end{array}\right].
+\end{equation}
+
+The operator  $\mt S(\cdot)$ is able to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.
+This property is explored in the Appendix.
+
 Demonstration of Euler equation:

### 2016-07-05 rigid_bodies.tex
Update of splitting equations

@@
 \documentclass[aip,jcp,reprint,amsmath,amssymb,amsfont]{revtex4-1}
+
 \usepackage[pdftex]{graphicx}
@@
 \usepackage[]{threeparttable}
+\usepackage{booktabs}
+\usepackage{multirow}
+
 \newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}           % matrix symbol
@@
 \raggedbottom
-
-\usepackage{booktabs}
-\usepackage{multirow}
 
@@
 
+\[
+\begin{split}
+i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
+i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
+i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
+\end{split}
+\]
+
+The same actions, but simultaneously for all rigid bodies present in the system.
+
 \subsection{Canonical Ensemble}
@@
 
-As done in the Hamiltonian case, the evolution of an extended phase space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion.
+As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion.
 First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components.
@@
 \begin{split}
-&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1} p_{\eta_j}^0 \right) \phi(\alpha_{j+1} t) t, \\
+&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
 \end{split}
@@
 Considering the analytical solution above, the NHC operator is factorized as
-\begin{widetext}
 \begin{equation}
 \begin{split}
-e^{i\!L_\text{NHC} \frac{\Delta t}{2}} =  &\prod_{j=M}^{1} \exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-&  \times \prod_{j=1}^{M} \exp\left(\frac{\Delta t}{2} \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)  \prod_{i=1}^{N} \exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\tr{\vt p}_i \diff{}{\vt p_i}\right) \prod_{i=1}^{N} \exp\left(-\frac{\Delta t}{2} \frac{p_{\eta_1}}{Q_1}\tr{\vt \pi}_i \diff{}{\vt \pi_i}\right)  \\
-& \times \prod_{j=1}^{M} \exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left(G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
+&\times \prod_{j=1}^M \exp\left(\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
+&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \alpha_1 \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right)\right] \\
+&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
 \end{split}
 \end{equation}
-\end{widetext}
+
 The remaining factors that appear in the second line of the equation above are evaluated according to\cite{Martyna1996}
 \begin{equation}
-\vt \pi_i \leftarrow \vt \pi_i \, exp\left(-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right)
-\end{equation}
-\begin{equation}
-\vt p_i \leftarrow \vt p_i \, exp\left(-\frac{\Delta t}{2}\frac{p_{\eta_1}}{Q_1}\right)
-\end{equation}
-\begin{equation}
-\eta_j  \leftarrow \eta_j + \frac{p_{\eta_j}}{ Q_j} \frac{\Delta t}{2} 
+\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
+\end{equation}
+\begin{equation}
+\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
+\end{equation}
+\begin{equation}
+\eta_j = \eta_j^0 + \alpha_j^0 \frac{\Delta t}{2} 
 \end{equation}
@@
 \label{eq:trotter_splitting_NHC}
-e^{i L \Delta t} =
-e^{i L_{NHC} \frac{\Delta t}{2}} e^{i L_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t}  e^{i L_a \frac{\Delta t}{2}} e^{i L_{NHC} \frac{\Delta t}{2}}.
+e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
 \end{equation}
@@
 \begin{split}
-e^{iL^\ast_a \frac{\Delta t}{2}} = &\prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \left( \vt F_i^T - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} \right] \\
-&\times \prod_{i=1}^{N} exp\left[-\frac{\Delta t}{2} \left(2 \vt \tau_i^T \mt C_i^T -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T \right)  \diff{}{\vt \pi_i}   \right]
+e^{\frac{\Delta t}{2} i\!L_b^\ast} &= \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} \right] \\
+&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left(2 \tr{\vt \tau}_i \tr{\mt C}_i -  \alpha_1 \tr{\vt \pi}_i \right)  \diff{}{\vt \pi_i}   \right]
 \end{split}
@@
 \begin{split}
-e^{iL^\ast_{NHC} \frac{\Delta t}{2}} = &\prod_{j=M}^{1} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-&\times\prod_{j=1}^{M} exp\left(-\frac{\Delta t}{2}  \frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j}\right)   \\
-&\times \prod_{j=1}^{M} exp\left[\frac{\Delta t}{4} \left( G_j - \frac{p_{\eta_{j+1}}}{Q_{j+1}} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
+&\times \prod_{j=1}^M \exp\left(-\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
+&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
 \end{split}
@@
 \begin{split}
-e^{i L \Delta t} = &e^{i L^\ast_{NHC} \frac{\Delta t}{2}} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L_r \Delta t} e^{i L_t \Delta t} e^{i L^\ast_a \frac{\Delta t}{2}} e^{i L^\ast_{NHC} \frac{\Delta t}{2}}.
+e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast} e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast}.
 \end{split}
@@
 Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.
+
+\subsection{To do}
+
+Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
+\[
+W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
+\]
 
@@
 
-\subsection{To do}
-
-Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
-\[
-W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
-\]
-
 \appendix

### 2016-07-05 rigid_bodies.tex
Inclusion of microcanonical ensemble

@@
 \subsection{Symplectic Numerical Integration}  
-\label{symplectic}
+\label{sec:symplectic}
 
@@
 
-By defining $i\!L_r = i\!L_1 + i\!L_2 + i\!L_3$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ commute, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
+By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ commute, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
 \begin{equation}
 \label{eq:trotter_splitting_NVE}
-e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_t} \left[ \prod_{j=1}^n e^{{\delta t}_j i\!L_{r}} \right] e^{\frac{\Delta t}{2} i\!L_b},
+e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
+\end{equation}
+
+Update of forces and torques is required only after the free translation and free rotation have been carried out. Each complete rotation is split as\cite{Miller2002}
+\begin{equation}
+\label{eq:splitting_rot}
+e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3} \right),
 \end{equation}
@@
 This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
-Each complete rotation is split as\cite{Miller2002}
-\begin{equation}
-\label{eq:splitting_rot}
-e^{\delta t_j i\!L_r} = e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3}.
-\end{equation}
-
 Following Ref.~\onlinecite{Miller2002}, one could simply make $\delta t_j = \Delta t/n$ for all $j$.
@@
 \section{Molecular Dynamics}
-\label{sec:moleculardynamics}
-
-In this section, we present the factorization schemes we considered for generating molecular dynamics algorithms in both the NVE and NVT ensembles.
+\label{sec:molecular_dynamics}
+
+In this section, we present factorization schemes for generating molecular dynamics algorithms in the microcanonical (NVE) and canonical (NVT) ensembles.
+In both cases, a three-dimensional system is considered as consisting of $N$ interacting, non-collinear rigid bodies occupying a fixed volume $V$.
+An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$).
+Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their conjugate momenta.
+Of course, at least $2N$ integrals of motion are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.
 
@@
 
+Considering the system described above, with $\vt r^N$ and $\vt q^N$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
+\[
+\mathcal{H} = U(\vt r^N,\vt q^N) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
+\]
+
+As usual, sampling of a microcanonical ensemble can be carried out by a Hamiltonian dynamics in phase space, which will have $\mathcal{H} = E$ as an integral of motion.
+The dynamics of this system is governed by a Liouville operator expressed as in Eq.~\ref{eq:full_operator}, but now with individual operators defined as
 \[
 \begin{split}
-i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
-i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
-i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
+i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
+i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
+i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right).
 \end{split}
@@
 
-The same actions, but simultaneously for all rigid bodies present in the system.
+Therefore, a symplectic integrator can be devised using the same splitting scheme as described in Sec.~\ref{sec:symplectic}.
+The only difference is that the action of each propagator (boost, free translation, and free uniaxial rotation) must now occur simultaneously for all rigid bodies in the system.
+If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\vt P = \sum_{i=1}^N \vt p_i$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\vt G = \sum_{i=1}^N(\vt p_i t - m_i \vt r_i)$.
+If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.
 

### 2016-07-06 rigid_bodies.tex
Advancements in NVT integration schemes

@@
 %\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
-\documentclass[aip,jcp,reprint,amsmath,amssymb,amsfont]{revtex4-1}
+\documentclass[aip,jcp,reprint,amsmath,amssymb,raggedbottom]{revtex4-1}
 
@@
 \newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
-
-\raggedbottom
 
@@
 Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their conjugate momenta.
-Of course, at least $2N$ integrals of motion are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.
+Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.
 
@@
 Considering the system described above, with $\vt r^N$ and $\vt q^N$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
-\[
+\begin{equation}
+\label{eq:H_NVE}
 \mathcal{H} = U(\vt r^N,\vt q^N) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
-\]
+\end{equation}
 
@@
 The only difference is that the action of each propagator (boost, free translation, and free uniaxial rotation) must now occur simultaneously for all rigid bodies in the system.
-If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\vt P = \sum_{i=1}^N \vt p_i$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\vt G = \sum_{i=1}^N(\vt p_i t - m_i \vt r_i)$.
+If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$.
 If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.
@@
 
-Understanding of the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space, which is the cornerstone for deterministic molecular dynamics at different ensembles.
+Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space.
 Sergi and Ferrario\cite{Sergi2001} described a simple way of systematically generating conservative trajectories in an arbitrary phase space.
@@
 \end{equation}
-will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric.
-This can be shown by using the chain rule to express the time-derivative of an arbitrary scalar function $f(\vt x)$, which yields
-\[
-\dot{f} = \tr{\left(\diff{f}{\vt x}\right)}\dot{\vt x} = \tr{\left(\diff{f}{\vt x}\right)} \boldsymbol{ \mathcal B } \diff{H}{\vt x}.
-\]
-
-If $f = H$, the equation above turns into a quadratic form, which is identically null if $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{\mathcal B}$, thus resulting in $\dot H = 0$.
-
-In what follows, index $i$ ranges for all $N$ particles and index $j$ ranges for all $M$ thermostats, unless otherwise stated.
-
+will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric, that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$.
+We employ such technique to couple a unique Nos\'{e}-Hoover chain\cite{Martyna1992} with $M$ thermostats to a system with $N$ rigid bodies.
+For this, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
+The conserved ``extended Hamiltonian'' in this case is
 \begin{equation}
 \label{eq:H_nvt}
-\begin{split}
-H = U(\vt r, \vt q) + \sum_{i=1}^N \left( \frac{\tr{\vt p}_i {\vt p}_i}{2 m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8} \right) + \\
-+ \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j
-\end{split}
-\end{equation}
-
+H = \mathcal{H} + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j,
+\end{equation}
+where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE} and $L$ is a constant to be determined.
 The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
@@
 \begin{split}
-\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \qquad \text{and}\\
+\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
 \dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
@@
 
-In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat.
+In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the first thermostat.
 In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, plus a term regarding the action of a second thermostat.
@@
 
-Each other thermostat in the chain acts on the previous one, and is acted upon by the following one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
+Each additional thermostat will then act on the preceding one, and be acted upon by the succeeding one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
 \[
@@
 
-Finally, the last thermostat just acts on 
-\[
-{\dot p}_{\eta_M} = -\diff{H}{\eta_M} + p_{\eta_{M-1}} \diff{H}{p_{\eta_{M-1}}}.
-\]
-
-For a three-dimensional system composed of non-collinear rigid bodies, the phase-space compressibility is given by\cite{Tuckerman2010}
-\[
-\kappa = -6N \diff{H}{p_{\eta_1}} - \sum_{j=2}^M \diff{H}{p_{\eta_j}} = -6N \dot{\eta}_1 - \sum_{j=2}^M \dot{\eta}_j.
-\]
-
-Thus, the metric determinant factor of the extended phase space is then given by $\sqrt{g} = e^{6N \eta_1 - \sum_{j=2}^M \eta_j}$.\cite{Tuckerman_1999, Tuckerman2001}
-
-Finally, the Nos\'{e}-Hoover chain thermostat applied to a system of $N$ rigid bodies becomes
+Finally, as the last thermostat will just act on the penultimate one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
+Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'{e}-Hoover chain become
 \begin{subequations}
@@
 &\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \\
-&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \\
-&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \\
+&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
+&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
 &\dot{\eta}_j = \alpha_j \\
@@
 \end{subequations}
-where $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$.
+
+In the equations above, ${\vt \omega}_i = \frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
 \begin{align*}
-&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - 6N k_B T \\
-&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad (j > 1)
+&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_B T \quad \text{and}\\
+&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
 \end{align*}
 
-$Q_1 = 6 N k_B T t_d^2$ and $Q_j = k_B T t_d^2$
-
-In these equations, $i$ indexes the number of rigid bodies from $1$ to $N$, $M$ is the number of thermostats in the chain, $T$ is the specified temperature, $k_b$ is Boltzmann's constant, $6N$ is the number of DOF, and $t_d$ is the characteristic time period of the thermostat chain.
-The conserved quantity in this case is the extended energy $\mathcal{H}^\prime$ given by
-\begin{align*}
-\mathcal{H}^\prime = \mathcal{H} + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_k} + 6Nk_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j,
-\end{align*}
-with $\mathcal{H}$ defined in Eq.~\ref{eq:H_with_Omega}.
+As recommended in Ref.~\onlinecite{Martyna1992}, the inertial parameters can be calculated from a characteristic time scale $t_d$ of the system by $Q_1 = L k_B T t_d^2$ and, for $j > 1$, $Q_j = k_B T t_d^2$.
+Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
+For a system without external forces, the extended energy $H$ and the vector quantity $\vt P e^{\eta_1} = \vt K$ are integrals of motion, while the center-of-mass position $\vt R$ can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
+By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the extended phase space is $\sqrt{g} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.
 
@@
 First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components.
-In this case, the Liouville operator can be written as
-\[
-i\!L = i\!L_\text{NHC} + i\!L_\text{NVE},
-\]
-where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $i\!L_\text{NHC}$ is given by
+In this case, the Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}.
+It follows that the operator $i\!L_\text{NHC}$ is given by
 \begin{equation}
 \begin{split}
-i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) \\
-&+ \sum_{j=1}^M \left[ \alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
+i\!L_\text{NHC} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
+&+ \sum_{j=1}^M \alpha_j \diff{}{\eta_j} + \sum_{j=1}^M \left[ (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
 \end{split}
 \end{equation}
+
+If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\vt p}_i = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values.
+These equations can be solved altogether in analytical form, which results in
+\begin{subequations}
+\begin{align}
+&\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 t \right), \\
+&\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 t \right), \quad \text{and}\\
+&\eta_j = \eta_j^0 + \alpha_j^0 t,
+\end{align}
+\end{subequations}
+where the superscript $0$ stands for ``initial value''.
 
@@
 \begin{split}
-&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
+&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
 \end{split}
@@
 \begin{split}
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left(G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-&\times \prod_{j=1}^M \exp\left(\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
-&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \alpha_1 \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right)\right] \\
-&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}} \\
+&\times e^{\frac{\Delta t}{2} \left[-\alpha_1 \sum\limits_{i=1}^N \left(\tr{\vt p}_i \frac{\partial}{\partial \vt p_i} + \tr{\vt \pi}_i \frac{\partial}{\partial \vt \pi_i}\right) + \sum\limits_{j=1}^M \alpha_j \frac{\partial}{\partial \eta_j} \right]} \\
+&\times \prod_{j=1}^M e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}}.\\
 \end{split}
-\end{equation}
-
-The remaining factors that appear in the second line of the equation above are evaluated according to\cite{Martyna1996}
-\begin{equation}
-\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
-\end{equation}
-\begin{equation}
-\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
-\end{equation}
-\begin{equation}
-\eta_j = \eta_j^0 + \alpha_j^0 \frac{\Delta t}{2} 
 \end{equation}

### 2016-07-06 rigid_bodies.tex
Changes in the first integration scheme for NVT

@@
 Let us analyze three of them, which will be useful later on to derive a complete approximate solution.
-In what follows, subscript $0$ denotes ``initial values''.
+In what follows, an index $0$ stands for ``initial value''.
 We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$.
@@
 \[
-e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^N + \mathcal{O}(\Delta t^2),
-\]
-where $\Delta t = t/N$ is the time step size and $N$ is the number of steps.
+e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^\mathcal{N} + \mathcal{O}(\Delta t^2),
+\]
+where $\Delta t = t/\mathcal N$ is the time step size and $\mathcal N$ is the number of steps.
 We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side.
@@
 For a system without external forces, the extended energy $H$ and the vector quantity $\vt P e^{\eta_1} = \vt K$ are integrals of motion, while the center-of-mass position $\vt R$ can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
-By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the extended phase space is $\sqrt{g} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.
-
-\subsection{Integration Schemes for NVT Molecular Dynamics}
+By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.
+
+\subsection{Integration Schemes for the Canonical Ensemble}
 
@@
 First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components.
-In this case, the Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}.
-It follows that the operator $i\!L_\text{NHC}$ is given by
-\begin{equation}
+In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator}.
+Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as
+
+\begin{equation}
+\label{eq:trotter_splitting_NHC}
+e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
+\end{equation}
+
+It follows from Eq.~\ref{eq:nhc_system} that
+\begin{equation}
+\label{eq:iL_NHC}
 \begin{split}
@@
 \begin{align}
-&\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 t \right), \\
-&\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 t \right), \quad \text{and}\\
-&\eta_j = \eta_j^0 + \alpha_j^0 t,
+&\vt p_i = \vt p_i^0 e^{-\alpha_1^0 t}, \\
+&\vt \pi_i = \vt \pi_i^0 e^{-\alpha_1^0 t}, \quad \text{and}\\
+&\eta_j = \eta_j^0 + \alpha_j^0 t.
 \end{align}
 \end{subequations}
-where the superscript $0$ stands for ``initial value''.
-
-As pointed out by Martyna~\textit{et al}.,\cite{Martyna1996} the part of the operator given by $(G_j - \alpha_j p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}$ can be evaluated analytically using the following equation:
+
+On the other hand, if a single term of the last summation was the only term present in Eq.~\ref{eq:iL_NHC}, then $e^{t i\!L_\text{NHC}}$ would behave according to the corresponding equation $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$, with all variables other than $p_{\eta_j}$ kept unchanged.
+The analytical solution of this single equation is
 \begin{equation}
@@
 \begin{split}
-&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
+&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi\left(\alpha_{j+1}^0 t\right) t, \\
 \end{split}
@@
 where $\phi(x) = (1-e^{-x})/x$.
-For small absolute values of $x$, this function can be evaluated by truncating its Taylor series form $\phi(x) = \sum_{n=0}^\infty \frac{(-1)^n x^n}{(n+1)!}$.
-Four terms result in a sufficiently precise value if $\|x\| \leq 10^{-4}$.
-Considering the analytical solution above, the NHC operator is factorized as
+This is an alternative to the form found in Ref.~\onlinecite{Martyna1996}.
+We might note here that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^4 \frac{(-1)^n x^n}{(n+1)!}$ whenever  $|x| \leq 10^{-4}$.
+
+Considering the analytical solutions above, one can split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula recursively. The splitting we employ here is
 \begin{equation}
@@
 \end{split}
-\end{equation}
-
-Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as
-
-\begin{equation}
-\label{eq:trotter_splitting_NHC}
-e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
 \end{equation}

### 2016-07-07 rigid_bodies.tex
Almost finished the NVT integration section

@@
 
-\subsection{Integration Schemes for the Canonical Ensemble}
-
-As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion.
+\subsection{Integration of the Canonical Equations of Motion}
+
+As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion.
 First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components.
 In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator}.
-Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as
-
+Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as\cite{Martyna1996}
 \begin{equation}
@@
 
-If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\vt p}_i = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values.
-These equations can be solved altogether in analytical form, which results in
+If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\eta}_j = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values.
+This system of equations can be solved analytically, which results in
 \begin{subequations}
 \begin{align}
-&\vt p_i = \vt p_i^0 e^{-\alpha_1^0 t}, \\
-&\vt \pi_i = \vt \pi_i^0 e^{-\alpha_1^0 t}, \quad \text{and}\\
-&\eta_j = \eta_j^0 + \alpha_j^0 t.
+&\vt p_i(t) = \vt p_i^0 e^{-\alpha_1 t}, \\
+&\vt \pi_i(t) = \vt \pi_i^0 e^{-\alpha_1 t}, \quad \text{and}\\
+&\eta_j(t) = \eta_j^0 + \alpha_j t. \label{eq:solution_eta}
 \end{align}
@@
 
-On the other hand, if a single term of the last summation was the only term present in Eq.~\ref{eq:iL_NHC}, then $e^{t i\!L_\text{NHC}}$ would behave according to the corresponding equation $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$, with all variables other than $p_{\eta_j}$ kept unchanged.
+Otherwise, if a single term $(G_j - \alpha_{j+1} p_{\eta_j}) \partial/\partial p_{\eta_j}$ was the only term present in Eq.~\ref{eq:iL_NHC}, then $e^{t i\!L_\text{NHC}}$ would behave according to the equation $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$, with all variables other than $p_{\eta_j}$ kept unchanged.
 The analytical solution of this single equation is
 \begin{equation}
-\label{eq:p_eta}
+\label{eq:solution_p_eta}
 \begin{split}
-&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi\left(\alpha_{j+1}^0 t\right) t, \\
+&p_{\eta_j}(t) = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1} p_{\eta_j}^0 \right) \phi\left(\alpha_{j+1} t\right) t, \\
 \end{split}
@@
 where $\phi(x) = (1-e^{-x})/x$.
-This is an alternative to the form found in Ref.~\onlinecite{Martyna1996}.
-We might note here that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^4 \frac{(-1)^n x^n}{(n+1)!}$ whenever  $|x| \leq 10^{-4}$.
-
-Considering the analytical solutions above, one can split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula recursively. The splitting we employ here is
+We might note that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^3 {(-1)^n x^n}/{(n+1)!}$ whenever $|x| \leq 10^{-4}$.
+The solution above is equivalent to that found in Ref.~\onlinecite{Martyna1996}.
+
+Considering the analytical solutions described above, it is possible to split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula in a recursive way.
+The splitting we employ here is
+\begin{align*}
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = &\prod_{j=M}^1 \exp \left[ \frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right] \\
+\times \exp \Bigg[ -\frac{\Delta t}{2} &\alpha_1 \sum_{i=1}^N \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j} \Bigg] \\
+\times &\prod_{j=1}^M \exp \left[\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].\\
+\end{align*}
+
+It is worth noting that an analytical solution can also be derived for the system of equations $\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i$ for all $i \in [1,N]$ if all other variables are constant.
+These equations are equivalent in form to the one solved above, so that their solutions are
+\begin{subequations}
+\label{eq:solution_momenta}
+\begin{align}
+&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \quad \text{and} \label{eq:solution_p} \\
+&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t. \label{eq:solution_pi}
+\end{align}
+\end{subequations}
+
+By evaluating these functions simultaneously for all $i$, we accomplish the effect of a propagator $e^{t i\!L_b^\ast}$, where $i\!L_b^\ast$ is the Liouville operator given by
+\[
+i\!L_b^\ast = \sum_{i=1}^N \left[ \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} \right].
+\]
+
+This is a kind of boost that already incorporates the influence of the thermostat chain.
+One of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when the solution in Eq.~\ref{eq:solution_momenta} is considered.
+For that, we also define a modified Nos\'e-Hoover chain propagator $e^{i\!L^\ast_\text{NHC}}$, as follows:
 \begin{equation}
 \begin{split}
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}} \\
-&\times e^{\frac{\Delta t}{2} \left[-\alpha_1 \sum\limits_{i=1}^N \left(\tr{\vt p}_i \frac{\partial}{\partial \vt p_i} + \tr{\vt \pi}_i \frac{\partial}{\partial \vt \pi_i}\right) + \sum\limits_{j=1}^M \alpha_j \frac{\partial}{\partial \eta_j} \right]} \\
-&\times \prod_{j=1}^M e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}}.\\
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } = &\prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
+\times &\exp\left(-\frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j}\right) \\
+\times &\prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
 \end{split}
@@
 
-It is worth noting that analytical solutions exist also for the equations
-\begin{align*}
-&\dot{\vt p}_i = {\vt F}_i - \frac{p_{\eta_1}}{Q_1} \vt p_i \, \text{ and} \\ 
-&\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \frac{p_{\eta_1}}{Q_1} \vt \pi_i,
-\end{align*}
-which are of identical form to Eq.~\ref{eq:p_eta_j}, so they can be evaluated using
-\begin{equation}
-\label{eq:p}
-\begin{split}
-{\vt p}_i \leftarrow {\vt p}_i + \left({\vt F}_i - \frac{p_{\eta_1}}{Q_1} {\vt p}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t \, \text{ and}\\
-\end{split}
-\end{equation}
-
-\begin{equation}
-\label{eq:pi}
-\begin{split}
-{\vt \pi}_i \leftarrow {\vt \pi}_i + \left(2 \tr{\vt \tau} \tr{\mt C} - \frac{p_{\eta_1}}{Q_1} {\vt \pi}_i \right) \phi\left(\frac{p_{\eta_1}}{Q_1} \Delta t \right) \Delta t,
-\end{split}
-\end{equation}
-respectively, with $\phi(x)$ given by Eq.~\ref{eq:phi}.
-Therefore, one of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when also Eqs.~\ref{eq:p} and~\ref{eq:pi} are considered.
-For that, let as define the modified propagators $ e^{i L^\ast_a} $ and $ e^{i L^\ast_{NHC}} $ as follows:
-%\begin{equation}
-%\begin{split}
-%&i L^\ast_a = i L_a -\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1} \vt p_i^T \diff{}{\vt p_i} - %\sum_{i=1}^{N}\frac{p_{\eta_1}}{Q_1}\vt \pi_i^T  \diff{}{\vt \pi_i} =\\
-%&\sum_{i=1}^{N} \left(\tr{\vt F_i} - \frac{p_{\eta_1}}{Q_1}\vt p_i^T \right) \diff{}{\vt p_i} + %\sum_{i=1}^{N} \left(2 \tr{\vt \tau_i} \tr{\mt C_i} -  \frac{p_{\eta_1}}{Q_1}\vt \pi_i^T\right)  \diff{}{\vt %\pi_i}
-%\end{split}
-%\end{equation}
-%\begin{equation}
-%\begin{split}
-%i L^\ast_{NHC} = \sum_{j=1}^{M}\left[\frac{p_{\eta_j}}{Q_j}\diff{}{\eta_j} + \left(G_j - p_{\eta_j} %\frac{p_{\eta_{j+1}}}{Q_{j+1}}\right) \diff{}{p_{\eta_j}}\right] 
-%\end{split}
-%\end{equation}
-\begin{equation}
-\begin{split}
-e^{\frac{\Delta t}{2} i\!L_b^\ast} &= \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} \right] \\
-&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \left(2 \tr{\vt \tau}_i \tr{\mt C}_i -  \alpha_1 \tr{\vt \pi}_i \right)  \diff{}{\vt \pi_i}   \right]
-\end{split}
-\end{equation}
-
-\begin{equation}
-\begin{split}
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-&\times \prod_{j=1}^M \exp\left(-\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
-&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
-\end{split}
-\end{equation}
-Finally, the complete propagator scheme is
-\begin{equation}
-\label{eq:trotter_splitting_*}
+Of course, Eqs.~\ref{eq:solution_eta} and \ref{eq:solution_p_eta} are the ones used to evaluate the propagators above. Finally, the complete scheme is
+\begin{equation}
+\label{eq:modified_splitting}
 \begin{split}
@@
 
-In the NVE case, each Liouvillian operator is derived from part of the Hamiltonian, then each factor in the propagator is symplectic and, as a consequence, preserves the phase space volume.
+In the microcanonical case, each term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
+As a consequence, the whole propagator preserves the phase-space volume.
 Recall that the inverse statement is not necessarily true.
-Now it is widely recognized the importance of the symplectic property for numerical stability,\cite{Skeel1997} however it is restricted to Hamiltonian systems.
-For non-Hamiltonian equations of motion the phase space volume is no longer conserved.
-Nevertheless, there exists an invariant measure on the phase space, and for the the system of equations~\ref{eq:nhc_system} it is given by
+It is now widely recognized that the symplectic property is important for numerical stability,\cite{Skeel1997} although it is restricted to Hamiltonian systems.
+For non-Hamiltonian equations of motion, the phase space volume is no longer conserved.
+Nevertheless, there exists an invariant measure on the phase space, and for the system in Eq.~\ref{eq:nhc_system}, it is given by
 \begin{align*}
-\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ].
+%\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ].
+\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge {\vt r} \wedge {\vt p} \wedge {\vt q} \wedge {\vt \pi} \wedge {\vt \eta} \wedge {\vt p_\eta}
 \end{align*}
 
-In this sense, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which are both time  reversible an measure-preserving.
-In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to more ergodic trajectory behavior.
-It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*} do not preserve the invariant-measure.
-Nevertheless, as will be shown below, those numerical schemes successfully achieve the specified temperature and the long-term energy conservation is acceptable.
-Note that the term that premultiplies the extended phase space volume in the invariant-measure above is a function of the variable $\eta$.
-Given that the time evolution of the physical variables along with the thermostat momentum does not depend on $\eta$ we conclude that in this case the measure-preserving condition for every individual operator can actually be relaxed.
+In this sense, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which are both time-reversible and measure-preserving.
+In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to trajectories that are more ergodic.
+It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:modified_splitting} do not preserve the invariant measure.
+Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.
+Note that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
+Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
 
@@
 Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
-\begin{equation}
-\begin{split}
-\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right.
+\[
+%\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right.
 \\ \left.
 \diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
-\end{split}
-\end{equation}
+\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{{\vt r}_{ij}}{L} \diff{u}{{\vt r}_{ij}} + \diff{{\vt q}_i}{L} \diff{u}{{\vt q}_i} + \diff{{\vt q}_j}{L} \diff{u}{{\vt q}_j} \right).
+\]
+
 Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
@@
 \[
-\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \diff{u}{{\vt r}_{ij}} \cdot {\vt r}_{ij}.
+\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
 \]
@@
 \end{equation}
+
 This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs.
@@
 
-\subsection{To do}
-
-Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
+\textbf{To do}: Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
 \[

### 2016-07-08 rigid_bodies.tex
Some small changes

@@
 \subsection{Kinematics}
+\label{sec:kinematics}
 
@@
 Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
-For a system without external forces, the extended energy $H$ and the vector quantity $\vt P e^{\eta_1} = \vt K$ are integrals of motion, while the center-of-mass position $\vt R$ can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
+For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
 By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.
@@
 Considering the analytical solutions described above, it is possible to split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula in a recursive way.
-The splitting we employ here is
+The splitting we employ here, noting that the product signs run in reverse orders, is
 \begin{align*}
@@
 \begin{align*}
-%\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge \prod_{i=1}^{N}[\vt {r}_i \wedge\vt {p}_i\wedge \vt {q}_i\wedge \vt {\pi}_i ] \wedge \prod_{j=1}^{M}[ \eta_j\wedge p_{\eta_j} ].
-\overline{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge {\vt r} \wedge {\vt p} \wedge {\vt q} \wedge {\vt \pi} \wedge {\vt \eta} \wedge {\vt p_\eta}
+\bar{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge {\vt r} \wedge {\vt p} \wedge {\vt q} \wedge {\vt \pi} \wedge {\vt \eta} \wedge {\vt p_\eta}
 \end{align*}
@@
 
-We now derive an expression for calculating the pressure of a rigid body system under PBC.
+The usual way of computing pressure in molecular dynamics is based on.....
+
+We now derive an expression for calculating the pressure of a rigid body system under periodic boundary conditions.
 For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies.
 This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended.
-
-The potential energy per unit cell of a system with $N$ distinct rigid bodies (indexed from $1$ to $N$) and periodic boundary conditions is
+For a pairwise additive model, the potential energy of the $N$ rigid bodies whose centers of mass lie in a simulation box with periodic boundary conditions is
 \begin{equation}
 \label{eq:U_pairwise_pbc}
-U({\vt r}^N, {\vt q}^N) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
+U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
 \end{equation}
@@
 
-In the equation above, $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$.
+\begin{equation}
+\label{eq:u_rigid_body_pair}
+u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l\| \right).
+\end{equation}
+
+In the equation above, ${{\vt R}_i}_k$ is the position of the $k$-th particle of body $i$ and $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$.
 The dependence of this potential on the types of these two atoms is omitted for simplicity.
+From Sec.~\ref{sec:kinematics}, one can deduce that
+
 The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
@@
 \]
+
 The first term in the expression above accounts for the ideal gas contribution.
@@
 
-Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = {\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
+Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = \tr{\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
 Therefore,

### 2016-07-16 rigid_bodies.tex
Changes in the Pressure section

@@
 
-The usual way of computing pressure in molecular dynamics is based on.....
-
-We now derive an expression for calculating the pressure of a rigid body system under periodic boundary conditions.
-For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies.
-This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended.
-For a pairwise additive model, the potential energy of the $N$ rigid bodies whose centers of mass lie in a simulation box with periodic boundary conditions is
+The pressure of a system with constant $NVT$, for which $\sum_{i=1}^N{\vt F}_i = 0$, is calculated by\cite{Louwerse2006, Tuckerman2001}
+\[
+P = \frac{(N-1) k_B T}{V} - \left\langle \frac{dU}{dV} \right\rangle.
+\]
+
+For simplicity, we consider a system of rigid bodies under periodic boundary conditions whose potential energy function is pairwise additive with respect to individual atoms, so that the approach of Louwerse and Baerends\cite{Louwerse2006} can be applied (otherwise, one can employ the approach of Thompson \textit{et al.}\cite{Thompson2009}) Assuming the minimum image convention, the potential energy per unit box is given by
 \begin{equation}
 \label{eq:U_pairwise_pbc}
-U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
-\end{equation}
-where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
+U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^N \sum_{l=1}^{{n_p}_j} \upsilon(r_{ijkl}),
+\end{equation}
+where $\upsilon(r)$ is a pair interaction potential and $r_{ijkl}$ is the distance between the $l$-th atom of body $j$ and the nearest image of the $k$-th atom of body $i$.
+Such distance is the norm of a vector ${\vt r}_{ijkl} = \hat{\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_k}_i - {\mt A}({\vt q}_j){{\vt d}_l}_k$, where $\hat{\vt r}_{ij}$ points from the center of mass of body $j$ to that of the body which contains the mentioned atom image.
+We note that $U$ does not depend explicitly on $V$, which makes ${\partial U}/{\partial V} = 0$.
+In addition, we consider that an isotropic change in volume would only displace the centers of mass of the bodies, without any rotation.
+This makes\cite{Louwerse2006, Tuckerman2010}
+\[
+\frac{\partial \hat{\vt r}_{ij}}{\partial V} = \frac{\tr{\hat{\vt r}}_{ij}}{3V} \quad \text{and} \quad \frac{\partial {\vt q}_i}{\partial V} = \frac{\partial {\vt q}_j}{\partial V} = 0.
+\]
+
+Therefore, applying the chain rule to Eq.~\ref{eq:U_pairwise_pbc} yields
+\[
+\frac{dU}{dV} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^N \sum_{l=1}^{{n_p}_j} \tr{\hat{\vt r}}_{ij} \upsilon^\prime(r_{ijkl})\diff{r_{ijkl}}{\vt r_{ijkl}},
+\]
+where $\upsilon^\prime(r) = \partial \upsilon/\partial r$.
+
+: the one whose $k$-th particle is nearest to the $l$-th particle of body $j$.
+
+$\vt r_{ijkl} = \hat{\vt r}_{ij} + {\vt \delta_k}_i - {\vt \delta_l}_j$
+
+${\vt \delta_k}_i$ and ${\vt \delta_l}_j$ respectively depend on $\vt q_i$ and $\vt q_j$ as defined in Sec.~\ref{sec:kinematics}, while $\hat{\vt r}_{ij}$ is the vector pointing from the center of mass of body $j$ to that of an image of body $i$: the one whose $k$-th particle is nearest to the $l$-th particle of body $j$.
+
+defi
+
+atom $jl$ (the $l$-th particle of body $j$) to the nearest image of atom $ik$ (the $k$-th particle of body $i$).
+If $\hat{\vt r}_{ij}$ is the vector pointing from the center of mass of body $j$ to the center of mass of the body $i$ that co
+
+\begin{equation}
+\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^N \sum_{l=1}^{{n_p}_j} \upsilon(\|\vt r_{ijkl}\|),
+\end{equation}
+
+where ${\vt r}_{ijkl} = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l$
+
+where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ is the vector pointing from the center of mass of body $j$ to that of body $i$.
+Given that $U$ does not depend explicitly on $V$ and that the bodies would neither deform nor rotate by an isotropic change in volume, we have that $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$ and $\partial {\vt r}_{ij}/\partial V = \tr{\vt r}_{ij}/(3V)$.\cite{Louwerse2006} Therefore,
+\[
+\diff{U}{V} = \frac{1}{6V} \sum_{i=1}^N \sum_{\substack{j=1 \\ j\neq i}}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
+\]
+
+Now let us define ${\vt r}_{ijkl}$ as the vector pointing from the $k$-th atom of body $i$ to the $l$-th atom of body $j$, that is, ${\vt r}_{ijkl} = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l$.
+For a pair potential $\upsilon(r)$, whose dependence on the atom identities is omitted for simplicity, we have that
 \begin{equation}
 \label{eq:u_rigid_body_pair}
-u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
-\end{equation}
-
-\begin{equation}
-\label{eq:u_rigid_body_pair}
-u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l\| \right).
-\end{equation}
-
-In the equation above, ${{\vt R}_i}_k$ is the position of the $k$-th particle of body $i$ and $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$.
-The dependence of this potential on the types of these two atoms is omitted for simplicity.
-From Sec.~\ref{sec:kinematics}, one can deduce that
-
-The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
-\[
-{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
-\]
-
-The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
-\[
-P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
-\]
-
-The first term in the expression above accounts for the ideal gas contribution.
-Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
-\[
-%\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right.
-\\ \left.
-\diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
-\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{{\vt r}_{ij}}{L} \diff{u}{{\vt r}_{ij}} + \diff{{\vt q}_i}{L} \diff{u}{{\vt q}_i} + \diff{{\vt q}_j}{L} \diff{u}{{\vt q}_j} \right).
-\]
-
-Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = \tr{\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
-Therefore,
-\[
-\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
-\]
-
-We can now use Eq.~\ref{eq:u_rigid_body_pair} to obtain the term inside the double summation above. By the chain rule, we have
-\[
-\diff{u}{{\vt r}_{ij}} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \diff{\vt r}{\vt r_{ij}} \diff{\upsilon}{\vt r},
-\]
-where $\vt r = {{\vt R}_i}_k - {{\vt R}_j}_l$.
-Notice that, inside the double summation above, the first derivative is the identity matrix and the second one is $-{\vt f}_{ijkl}$, with this force being the action of the $l$-th particle of body $j$ over the $k$-th particle of body $i$.
+u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{\vt r}_{ijkl}\| \right).
+\end{equation}
+where $r_{ijkl} = \|{\vt r}_{ijkl}\|$. Then, by the chain rule,
+\[
+\diff{u}{{\vt r}_{ij}} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \diff{\vt r_{ijkl}}{\vt r_{ij}} \diff{\upsilon}{\vt r_{ijkl}}.
+\]
+
+Inside the double summation above, the first derivative is the identity matrix and the second one is the $-{\vt F}_{ijkl}$, with this force being the action of the $l$-th particle of body $j$ over the $k$-th particle of body $i$.
 Therefore, the equation above results in $-\vt F_{ij}$, with this force being the resultant action of body $j$ over body $i$.
 Finally, we conclude that
-
 \begin{equation}
 \label{eq:pressure}
-P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
+P = \frac{(N-1) k_B T}{V} - \frac{1}{6V} \left\langle \sum_{i=1}^N \sum_{\substack{j=1 \\ j\neq i}}^{\infty} {\vt r}_{ij} \cdot {\vt F}_{ij} \right\rangle.
+\end{equation}
+
+With minimum image convention:
+\begin{equation}
+\label{eq:pressure_minimum_image}
+P = \frac{(N-1) k_B T}{V} - \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^N {\vt r}_{ij}^\ast \cdot {\vt F}_{ij}^\ast \right\rangle.
 \end{equation}
@@
 W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
+\]
+
+\[
+W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^N \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot \left[ {\vt r}_{ijkl} - {\mt A}({\vt q}_i){{\vt d}_i}_k + {\mt A}({\vt q}_j){{\vt d}_j}_l \right].
+\]
+
+\[
+\begin{split}
+W = \sum_{i=1}^N \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ijkl} \\
+- \sum_{i=1}^N \sum_{k=1}^{{n_p}_i} \left( \sum_{j=i+1}^N \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \right) \cdot {\mt A}({\vt q}_i){{\vt d}_i}_k \\
++ \sum_{i=1}^N \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\mt A}({\vt q}_j){{\vt d}_j}_l.
+\end{split}
 \]

### 2016-07-19 rigid_bodies.tex
Finalization of Pressure Section and Changes in Results Section

--- 2016-07-08 rigid_bodies.tex
@@
 This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
-Following Ref.~\onlinecite{Miller2002}, one could simply make $\delta t_j = \Delta t/n$ for all $j$.
+Following Miller \textit{et al}.,\cite{Miller2002} one could simply make $\delta t_j = \Delta t/n$ for all $j$.
 Alternatively, one could use a higher-order factorization scheme, such as the one due to Yoshida\cite{Yoshida1990} and Suzuki,\cite{Suzuki1991a, Suzuki1991b} in which $\delta t_j = \omega_j \Delta t$ and the weights $\omega_j$ are carefully chosen to determine the order of approximation.
@@
 
-Considering the system described above, with $\vt r^N$ and $\vt q^N$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
+Considering the system described above, with $\vt r$ and $\vt q$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
 \begin{equation}
 \label{eq:H_NVE}
-\mathcal{H} = U(\vt r^N,\vt q^N) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
+\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
 \end{equation}
@@
 
-The usual way of computing pressure in molecular dynamics is based on.....
-
-We now derive an expression for calculating the pressure of a rigid body system under periodic boundary conditions.
-For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies.
-This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended.
-For a pairwise additive model, the potential energy of the $N$ rigid bodies whose centers of mass lie in a simulation box with periodic boundary conditions is
+The pressure of a system with constant NVT, for which $\sum_{i=1}^N{\vt F}_i = 0$, is calculated by\cite{Louwerse2006, Tuckerman2001}
+\[
+P = \frac{(N-1) k_B T + \left\langle W \right\rangle}{V},
+\]
+where $W$ is the internal virial of the system, defined as $W = - V \frac{dU}{dV}$.
+For simplicity, we consider a system of rigid bodies under periodic boundary conditions whose potential energy function is pairwise additive with respect to individual atoms, so that the approach of Louwerse and Baerends\cite{Louwerse2006} can be applied (otherwise, one can employ the approach of Thompson \textit{et al.}\cite{Thompson2009}).
+Due to the rigidity of the bodies, interactions amongst companion atoms (i.e., atoms belonging to the same body) are irrelevant.
+Assuming that the minimum image convention\cite{Allen1989} applies for entire bodies, meaning that the simulation box is large enough to preclude any atom from having their own images or images of its companion atoms within its cutoff radius, the potential energy per unit box is given by
 \begin{equation}
 \label{eq:U_pairwise_pbc}
-U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
-\end{equation}
-where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
-\begin{equation}
-\label{eq:u_rigid_body_pair}
-u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
-\end{equation}
-
-\begin{equation}
-\label{eq:u_rigid_body_pair}
-u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l\| \right).
-\end{equation}
-
-In the equation above, ${{\vt R}_i}_k$ is the position of the $k$-th particle of body $i$ and $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$.
-The dependence of this potential on the types of these two atoms is omitted for simplicity.
-From Sec.~\ref{sec:kinematics}, one can deduce that
-
-The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
-\[
-{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
-\]
-
-The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
-\[
-P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
-\]
-
-The first term in the expression above accounts for the ideal gas contribution.
-Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
-\[
-%\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right.
-\\ \left.
-\diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
-\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{{\vt r}_{ij}}{L} \diff{u}{{\vt r}_{ij}} + \diff{{\vt q}_i}{L} \diff{u}{{\vt q}_i} + \diff{{\vt q}_j}{L} \diff{u}{{\vt q}_j} \right).
-\]
-
-Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = \tr{\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$.
-Therefore,
-\[
-\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
-\]
-
-We can now use Eq.~\ref{eq:u_rigid_body_pair} to obtain the term inside the double summation above. By the chain rule, we have
-\[
-\diff{u}{{\vt r}_{ij}} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \diff{\vt r}{\vt r_{ij}} \diff{\upsilon}{\vt r},
-\]
-where $\vt r = {{\vt R}_i}_k - {{\vt R}_j}_l$.
-Notice that, inside the double summation above, the first derivative is the identity matrix and the second one is $-{\vt f}_{ijkl}$, with this force being the action of the $l$-th particle of body $j$ over the $k$-th particle of body $i$.
-Therefore, the equation above results in $-\vt F_{ij}$, with this force being the resultant action of body $j$ over body $i$.
-Finally, we conclude that
-
-\begin{equation}
-\label{eq:pressure}
-P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
-\end{equation}
-
-This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs.
-Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.
-
-\textbf{To do}: Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
-\[
-W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
+U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon(r_{ijkl}),
+\end{equation}
+where $\upsilon(r)$ is a pair interaction potential and $r_{ijkl}$ is the distance between the $k$-th atom of body $i$ to its nearest image of the $l$-th atom of body $j$.
+Such distance is the norm of a vector ${\vt r}_{ijkl} = \hat{\vt r}_{ij} + \tr{[{\mt A}({\vt q}_i)]}{{\vt d}_k}_i - \tr{[{\mt A}({\vt q}_j)]}{{\vt d}_l}_k$, where $\hat{\vt r}_{ij}$ is the minimum-image vector pointing from the center of mass of body $j$ to that of body $i$.
+We note that $U$ does not depend explicitly on $V$, which makes ${\partial U}/{\partial V} = 0$.
+If the atoms were free to move individually, an isotropic change in volume would displace all atoms and we would have ${\partial {\vt r}_{ijkl}}/{\partial V} = (3V)^{-1} \tr{\vt r}_{ijkl}$.\cite{Louwerse2006} In this event, the atomic internal virial would be computed by
+\begin{equation}
+\label{eq:virial_atoms}
+W_\text{a} = \frac{1}{3} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \tr{\vt r}_{ijkl} {\vt f}_{ijkl},
+\end{equation}
+where $\vt f_{ijkl} = -{\partial \upsilon}/{\partial \vt r_{ijkl}}$ is the minimum-image force exerted by the $l$-th atom of body $j$ on the $k$-th atom of body $i$.
+In the case of rigid bodies, considering that an isotropic change in volume would neither deform nor rotate them, but only promote center-of-mass displacements, we have that ${\partial \hat{\vt r}_{ij}}/{\partial V} = (3V)^{-1} \tr{\hat{\vt r}}_{ij}$ and ${\partial {\vt q}_i}/{\partial V} = {\partial {\vt q}_j}/{\partial V} = \tr{\vt 0}$.
+Therefore, by applying the chain rule to Eq.~\ref{eq:U_pairwise_pbc} and realizing that $\partial \vt r_{ijkl}/\partial \hat{\vt r}_{ij} = \vt 1_3$, we obtain
+\[
+\label{eq:dUdV_rigid_bodies}
+\frac{dU}{dV} = -\frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \tr{\hat{\vt r}}_{ij} \vt f_{ijkl}.
+\]
+
+There are two convenient ways of employing this result to compute the internal virial of a system of rigid bodies.
+On one hand, we can explore the fact that ${\hat{\vt r}}_{ij}$ does not depend on the indexes $k$ and $l$.
+On the other hand, we can replace $\hat{\vt r}_{ij} = {\vt r}_{ijkl} - {{\vt \delta}_k}_i + {{\vt \delta}_l}_k$ and make use of the fact that $\vt f_{ijkl} = -\vt f_{jilk}$.
+Finally, we have
+\begin{subequations}
+	\label{eq:virial_rigid_bodies}
+	\begin{align}
+	W &= \frac{1}{3} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \tr{\hat{\vt r}}_{ij} \vt F_{ij} \\
+	  &= W_\text{a} - \frac{1}{3} \sum_{i=1}^N \sum_{k=1}^{{n_p}_i} {\tr{\vt \delta}_k}_i {{\vt f}_k}_i,
+	\end{align}
+\end{subequations}
+where $W_\text{a}$ is the atomic virial given by Eq.~\ref{eq:virial_atoms}, while $\vt F_{ij}$ is the total action of body $j$ (and its images) over body $i$ and $\vt f_{ik}$ is the resultant force exerted on the $k$-th atom of body $i$, respectively given by
+\
+\[
+\vt F_{ij} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \vt f_{ijkl} \quad \text{and} \quad \vt f_{ik} = \sum_{\substack{j=1\\j \neq i}}^N \sum_{l=1}^{{n_p}_j} \vt f_{ijkl}.
 \]
@@
 In this section, we analyze some performance aspects of the integration schemes when applied to a molecular system in the liquid state.
-Given its practical importance, we simulated 356 TIP3P\cite{Price2004} water molecules at a density of 999.56 $kg/m^3$.
-The intermolecular interactions were truncated at 9 {\AA}.
-The shifted-force (SF) method\cite{Allen1989,Toxvaerd_2011} was used to treat both the coulombic and the 12-6 Lennard-Jones (LJ) interactions.
+For this, we simulated a system containing 356 TIP3P water molecules\cite{Price2004} at a density of 999.56 $kg/m^3$.
+The intermolecular interactions were truncated at 9 \AA.
+The shifted-force (SF) method\cite{Allen1989, Fennell2006, Toxvaerd_2011} was used to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
 
@@
 
-In order to evaluate the stability of the numerical solvers, we use a measure given by
+In order to evaluate the stability of the numerical integrators, we employ a measure given by\cite{Tuckerman2010}
 \begin{equation}
 \label{eq:performance}
-D E =  \frac{1}{N_{steps}} \sum_{k=1}^{N_{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|,
-\end{equation}
-from which it is possible to calculate the average deviation of the conserved quantity, i.e.
-total energy, from its initial value $E_0$ after a certain number of time steps $N_{steps}$.
-Both the instantaneous fluctuations of energy, given by the term inside the summation, and the long-time conservation of energy, $D E$, are monitored.
-
-Recall that the method introduced here for the NVE dynamics is a mathematical reformulation of the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm introduced by Miller~\textit{et al.}\cite{Miller2002}, so the trajectories for both methods must coincide out of round-off errors.
-In Fig.~\ref{fig:miller1fs} we plot the instantaneous fluctuations of energy with $n_{ys} = 1 $ and a time step of 1 fs, where it can be seen that, in fact, the fluctuations for both methods are of similar magnitude.
-The simulation with the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm was conducted using the LAMMPS\cite{Plimpton1995} software package.
-For the purpose of comparison we applied the LJ model without shifting the forces, as it is not implemented in LAMMPS.
-
-\begin{figure}[!h]
-\centering
+D\!E =  \frac{1}{N_\text{steps}} \sum_{k=1}^{N_\text{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|,
+\end{equation}
+from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $N_\text{steps}$.
+Both the instantaneous fluctuations, given by the term inside the summation, and the long-time conservation of energy, $D\!E$, are monitored.
+
+Recall that, when the splitting scheme of Eq.~\ref{eq:splitting_rot} is executed with $\delta t_j = \Delta t/n$ for all $j$, the method introduced here for the NVE dynamics is just a mathematical reformulation of the algorithm of Miller \textit{et al}.\cite{Miller2002} Therefore, short trajectories obtained with both methods should coincide up to round-off errors.
+In Fig.~\ref{fig:miller1fs}, we plot the instantaneous fluctuations of total energy obtained with $n = 1$ and a time step of 1~fs using both methods.
+The simulation with the algorithm of Ref.~\onlinecite{Miller2002} was performed using the LAMMPS software package.\cite{Plimpton1995} For the purpose of comparison, we did not carry out force shifting for the Lennard-Jones model, as this is not implemented in LAMMPS.
+It can be seen in Fig.~\ref{fig:miller1fs} that, in fact, the fluctuations of both methods are of similar magnitudes.
+
+\begin{figure}
 	\includegraphics{millerourmd}
-\captionof{figure}{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and Ref.\citenum{Miller2002} (dotted line) with $n = 1 $, $n_{ys} = 1$, $\Delta t = 1 fs$, $E_0 = -2799.87 \, kcal/mol $.}
+	\caption{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and the LAMMPS implementation of the algorithm of Ref.~\onlinecite{Miller2002} (dotted line) with $n = 1$, $\Delta t = 1$ fs, and $E_0 = -2799.87$ kcal/mol.}
 	\label{fig:miller1fs}
@@
 
-In Table~\ref{table:nve} we compare the energy drift when using $n_{ys} = 1$ and $n_{ys} = 7$ in the rotational operator, Eq.~\ref{eq:ys_rot}.
-As expected, for a small time step of 1 fs, that further factorization is worthless.
-As the time step is increased, more accurate trajectories are achieved with $n_{ys} = 7 $.
-Nevertheless, the improvement is only marginal and all trajectories are stables exhibiting a small long-time energy drift.
-Thus, we proceed to analyze the NVT dynamics considering $n_{ys} = 1$.
-
-\begin{table}[h]
-\setlength{\tabcolsep}{7pt}
-\caption{Average deviation of the conserved quantity ($DE$) from liquid water NVE simulations with $n_{ys}=1$ and $n_{ys}=7$  }
-\centering % centering table
-\begin{tabular}{| c  c  c |}  
+In Table~\ref{table:nve}, we compare the energy drifts observed when the rotational operator in Eq.~\ref{eq:splitting_rot} is split using either $n = 1$ or the sixth-order Yoshida-Suzuki factorization scheme ($n = 7$).
+As observed, for a small time steps of 1~fs, increasing the order of factorization does not provide any improvement.
+As the time step size increases, trajectories with higher accuracy are achieved with $n = 7$.
+Nevertheless, the improvement is only marginal and all trajectories are stable, exhibiting a small long-time energy drift.
+Thus, we proceed by considering $n = 1$ in the analysis of the NVT dynamics.
+
+\begin{table}
+	\caption{Average total-energy deviations for liquid water in NVE simulations for Eq.~\ref{eq:splitting_rot} with either $n = 1$ or $n = 7$ with Yoshida-Suzuki factorization.}
+	\label{table:nve}
+	\begin{ruledtabular}
+		\begin{tabular}{ccc}
+			 & \multicolumn{2}{c}{ $D\!E$ } \\
+			 \cline{2-3}
+			 $\Delta t$/fs & $n = 1$ & $n = 7$ \\
 			 \hline
-&  \multicolumn{2}{c|}{ $D E$ }\\
-\cline{2-3}
-$\Delta t/fs$ &$n_{ys}=1$ &$n_{ys}=7$ \\
-\hline % inserts single-line
- 1 & 0.000055 & 0.000074\\
- 2 & 0.00048 & 0.00031 \\
- 3 & 0.0012 & 0.0011\\
- \hline
+			 1.0 & 0.000055 & 0.000074 \\
+			 2.0 & 0.00048  & 0.00031  \\
+			 3.0 & 0.0012   & 0.0011   \\
 		\end{tabular}
-\label{table:nve}
+	\end{ruledtabular}
 \end{table}
 
-It is not the purpose of this work to achieve an optimal factorization of the NVT schemes, but rather to show that a unique chain thermostat, while providing the ``proper'' sampling of the NVT ensemble, is capable of maintaining the average temperature at its specified value.
-In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\ref{eq:splitting_rot}.
-Recalling that in the NVT dynamics the conserved quantity is the extended energy $\mathcal{H}^\prime$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the average temperatures obtained in each case.
-It can be seen that both integrators, represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*}, systematically achieve the specified temperature, regardless of the time step size.
+It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
+\textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\ref{eq:splitting_rot}}.
+Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the average temperatures obtained in each case.
+It can be seen that both integrators, represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:modified_splitting}, systematically achieve the specified temperature, regardless of the time step size.
 In addition, the energy drift is, at its worst, around $1 \%$.
 
-\begin{table}[h]
-\setlength{\tabcolsep}{7pt}
+\begin{table}
 	\begin{threeparttable}
@@
 		\label{table:denvt}
-
-\centering % centering table
-\begin{tabular}{|c c c c c |}  
+		\begin{ruledtabular}
+			\begin{tabular}{ccccc}
+				& \multicolumn{2}{c}{Eq.~\ref{eq:trotter_splitting_NHC}} & \multicolumn{2}{c}{Eq.~\ref{eq:modified_splitting}} \\
+				\cline{2-5}
+				$\Delta t$/fs & $D\!E$ & $T$/K & $D\!E$ & $T$/K \\
 				\hline
-& \multicolumn{2}{c}{Eq.~$\ref{eq:trotter_splitting_NHC}$} &\multicolumn{2}{c|}{Eq.~$\ref{eq:trotter_splitting_*}$} \\
-\cline{2-5}
-$\Delta t/fs$ &$D E$ &$T/K$ &$D E$ &$T/K$\\
-\hline % inserts single-line
- 1  & 0.000043 & 298.16  & 0.000052  & 298.15 \\
-
- 2  & 0.00013 & 298.23  & 0.00017 & 298.06  \\
-
- 3  & 0.00089 & 298.17  & 0.0013 & 298.12  \\
-
- \hline
+				1.0 & 0.000043 & 298.16  & 0.000052  & 298.15 \\
+				2.0 & 0.00013  & 298.23  & 0.00017   & 298.06 \\
+				3.0 & 0.00089  & 298.17  & 0.0013    & 298.12
 			\end{tabular}
+		\end{ruledtabular}
 		\begin{tablenotes}
-\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$
-\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs 
+			\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$.
+			\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs.
 		\end{tablenotes}
@@
 Checking the consistency between the equipartition of energy and the simulated energies serves as another approach for investigating the precision of a simulation.
-In fact, as stated by Eastwood~\textit{et al.}~\cite{Eastwood_2010}, it has been useful to improve methodologies related to the calculation of electrostatic\cite{Levitt_1988,Guenot_1992,Arnold_1994} and dispersive\cite{Sagui_1999} interactions, as well as to detect errors in thermostats\cite{Harvey_1998,Mor_2008} and barostats\cite{Feller_1995}.
-The truncation error associated to the numerical schemes is a potential source of error, and is the focus of the work of Eastwood~\textit{et al.}~\cite{Eastwood_2010}, which investigates a solvated protein and a temperature difference of approximately 5.5 $K$ is observed between the protein and the solvent using a time step of 2 fs.
-
-In this paper, we analyze the partition of energy between the translational and rotational DOF, which are related to the temperature by the equipartition theorem as follows
+In fact, as stated by Eastwood \textit{et al}.,\cite{Eastwood_2010} this kind of checking has been useful to improve methodologies related to the calculation of  electrostatic\cite{Levitt_1988, Guenot_1992, Arnold_1994} and dispersive\cite{Sagui_1999} interactions, as well as to detect errors in thermostats\cite{Harvey_1998, Mor_2008} and barostats.\cite{Feller_1995} The truncation error associated to the numerical schemes is a potential source of error, and is the focus of the work of Eastwood \textit{et al}.,\cite{Eastwood_2010} which investigates a solvated protein and a temperature difference of approximately 5.5~K is observed between the protein and the solvent using a time step of 2~fs.
+
+In this paper, we analyze the partition of energy between the translational and rotational degrees of freedom, which are related to the temperature by the equipartition theorem as follows
+\begin{subequations}
 \begin{align}
 \label{eq:tras_equipartition}
-2 \langle E_{kin}^{trans} \rangle = (3N - 3) k_b T
+2 \langle E_\text{trans} \rangle = (3N - 3) k_b T
 \end{align}
-
 \begin{align}
 \label{eq:rot_equipartition}
-2 \langle E_{kin}^{rot} \rangle = 3N k_b T
+2 \langle E_\text{rot} \rangle = 3N k_b T
 \end{align} 
+\end{subequations}
+
 In these equations, $N$ is the number of rigid bodies.
-In NVE MD simulations, conservation of total linear momentum, $ \boldsymbol{P}= cte $, results in the loss of three translational DOF : $ 3N-3 $.
-For a cubic box with PBC there is no rotational symmetry ~\cite{Frenkel_2013,Kuzkin_2014}, with which the total angular momentum is not conserved, giving a number of rotational DOF of $3N$.
-We calculate the average energy per DOF $ \langle E_{cin} \rangle_{DOF} = \langle E_{cin} \rangle /N_{DOF}$ and then we calculate both the translational and rotational energies from equipartition as follows
+In microcanonical MD simulations, conservation of total linear momentum $\boldsymbol{P}$ results in the loss of three translational degrees of freedom, which is the reason why the factor $3N-3$ appears in Eq.~\ref{eq:tras_equipartition}.
+For a cubic box with periodic boundary conditions, there is no rotational symmetry,\cite{Frenkel_2013, Kuzkin_2014} with which the total angular momentum is not conserved, giving a number of $3N$ rotational degrees of freedom.
+We calculate the average energy per DOF $\langle E_{cin} \rangle_{DOF} = \langle E_{cin} \rangle /N_{DOF}$ and then we calculate both the translational and rotational energies from equipartition as follows
 
@@
 
-\begin{table}[h]
-\centering % centering table
+\begin{table}
 	\begin{threeparttable}
-\caption{Partition of kinetic energy from NVE simulation of 903 water molecules with $r_c = 9 $ {\AA} and SF-LJ\tnote{a}} %($5415 grados de libertad$)}
-%\begin{adjustbox}{width=1\textwidth}
-\begin{tabular}{| l c  c  c  c  c  c |}  
+		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c = 9$ {\AA} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
+		\label{table:water_partition_LJ-SF}
+		\begin{ruledtabular}
+			\begin{tabular}{ccccccc}
+				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
+%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
+				\cline{4-7}
+				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
 				\hline
- & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
-  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
+				0.5 & 305.20  & 1642.11  & 820.60  & 821.06  & 820.60 & 821.51  \\
+				1.0 & 304.92  & 1640.57  & 819.83  & 820.74  & 820.63 & 819.94  \\
+				2.0 & 304.20  & 1636.74  & 817.92  & 818.82  & 820.56 & 816.18  \\
+				3.0 & 302.97  & 1630.07  & 814.59  & 815.48  & 820.47 & 809.60
+			\end{tabular}
+		\end{ruledtabular}
+		\begin{tablenotes}
+			\item[a] Energies are in kcal/mol.
+		\end{tablenotes}
+	\end{threeparttable}
+\end{table}
+
+\begin{table}
+	\begin{threeparttable}
+		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c = 13$ {\AA} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
+		\label{table:water_partition_13A_LJ-SF}
+		\begin{ruledtabular}
+			\begin{tabular}{ccccccc}
+				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
+%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
 				\cline{4-7}
-$\Delta t/fs$  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
-\hline % inserts single-line
- 0.5 & 305.20  & 1642.11  & 820.60  & 821.06  & 820.60 & 821.51  \\
-  1  & 304.92  & 1640.57  & 819.83  & 820.74  & 820.63 & 819.94  \\
-  2  & 304.20  & 1636.74  & 817.92  & 818.82  & 820.56 & 816.18 \\
-  3  & 302.97  & 1630.07  & 814.59  & 815.48  & 820.47 & 809.60  \\
+				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
 				\hline
+				0.5 & 307.17 & 1652.71  & 825.90  & 826.81  & 825.96 & 826.75  \\
+				1.0 & 306.98 & 1651.68  & 825.38  & 826.30  & 826.09 & 825.59  \\
+				2.0 & 306.04 & 1646.60  & 822.85  & 823.75  & 825.57 & 821.03  \\
+				3.0 & 304.79 & 1639.89  & 819.49  & 820.40  & 825.63 & 814.26
 			\end{tabular}
+		\end{ruledtabular}
 		\begin{tablenotes}
-\item[a] Energies are in $kcal/mol$.
+			\item[a] Energies are in kcal/mol.
 		\end{tablenotes}
-\label{table:water_partition_LJ-SF}
 	\end{threeparttable}
-%\end{adjustbox}
-\end{table}
-
-\begin{table}[h]
-\centering % centering table
-\begin{threeparttable}
-\caption{Partition of kinetic energy from NVE simulation of 903 water molecules with $r_c = 13$ {\AA} and SF-LJ\tnote{a}} %($5415 grados de libertad$)}
-%\begin{adjustbox}{width=1\textwidth}
-\begin{tabular}{ |l c  c  c  c  c  c |}  
-\hline
- & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
-  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
-\cline{4-7}
-$\Delta t/fs$  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
-\hline % inserts single-line
- 0.5 & 307.17 & 1652.71  & 825.90  & 826.81  & 825.96 & 826.75  \\
-  1 & 306.98  & 1651.68  & 825.38  & 826.30  & 826.09 & 825.59  \\
-  2 & 306.04  & 1646.60  & 822.85  & 823.75 & 825.57 & 821.03 \\
-  3 & 304.79 & 1639.89  & 819.49  & 820.40  & 825.63 & 814.26  \\
-  \hline
- \end{tabular}
- \begin{tablenotes}
-\item[a] Energies are in $kcal/mol$.
-\end{tablenotes}
- \label{table:water_partition_13A_LJ-SF}
-\end{threeparttable}
-%\end{adjustbox}
 \end{table}
@@
 \label{eq:probability_ratio}
-\frac{P(E/\beta_2)}{P(E/\beta_1)} = exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
+\frac{P(E/\beta_2)}{P(E/\beta_1)} = \exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
 \end{equation}
@@
 \label{eq:ln_probability_ratio}
-ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E.
+\ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E.
 \end{equation}
@@
 \label{eq:probability_ratio_pot}
-\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  exp\left[(-\beta_2 - \beta_1)E_{pot}\right]
+\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  \exp\left[(-\beta_2 - \beta_1)E_{pot}\right]
 \end{equation}
@@
 \label{eq:probability_ratio_kin}
-\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  exp\left[(-\beta_2 - \beta_1)E_{kin}\right]
+\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  \exp\left[(-\beta_2 - \beta_1)E_{kin}\right]
 \end{equation}
@@
 \label{eq:mb}
-P(E_{cin}) = \frac{\beta}{\sqrt{N_{DOF}\pi}} \, exp \left[-\frac{\left(\beta E_{kin} - \frac{N_{DOF}}{2}\right)^2}{N_{DOF}} \right],
+P(E_{cin}) = \frac{\beta}{\sqrt{N_{DOF}\pi}} \exp\left[-\frac{\left(\beta E_{kin} - \frac{N_{DOF}}{2}\right)^2}{N_{DOF}} \right],
 \end{equation}
@@
 It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\ref{eq:trotter_splitting_*}.
-
-\begin{center}
+Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\ref{eq:modified_splitting}.
+
+\begin{figure}
 	\includegraphics{checkensemble}
-\captionof{figure}{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
-Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t = 1$ fs, $\bar{T}_1=296.03 K$, $\bar{T}_2 = 304.99 K$.
-}
+	\caption{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
+Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
 	\label{fig:checkensemble}
-\end{center}
-
-\begin{center}
+\end{figure}
+
+\begin{figure}
 	\includegraphics{maxwell-botlzamm-paper}
-\captionof{figure}{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\ref{eq:mb}.
-Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t = 1$~fs, $\bar{T} = 304.99 K$.}
+	\caption{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\ref{eq:mb}.
+Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}$ = 304.99~K.}
 \label{fig:mbdistribution}
-\end{center}
+\end{figure}
 
 Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
-It can be seen that the deviations from the true slope, $\Delta T= 8.970 K $, are less than $1\sigma$.
+It can be seen that the deviations from the true slope, $\Delta T$ = 8.970~K, are less than $1\sigma$.
 This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
 
-\begingroup
-\squeezetable
+%\begingroup
+%\squeezetable
+%\begin{table}
+%\setlength{\tabcolsep}{7pt}
+%\begin{threeparttable}
+%\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach \tnote{a} \tnote{b}}
+%\label{table:ensemblevalidation_old}
+%\centering % centering table
+%\begin{tabular}{|l c  c  c  c |}  
+%\hline
+% & \multicolumn{4}{c|}{true $\Delta T$ = 8.970 K}\\
+%\cline{2-5}
+%  & \multicolumn{2}{c}{potential} & \multicolumn{2}{c|}{kinetic}\\
+%\hline
+%Method  &$\Delta T/K$ &$\sigma$ dev &$\Delta T/K$&$\sigma$ dev\\
+%\hline % inserts single-line 
+% Eq.~$\ref{eq:trotter_splitting_NHC}$   & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
+% Eq.~$\ref{eq:trotter_splitting_*}$   & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14\\
+%  \hline
+%\end{tabular}
+%\begin{tablenotes}
+%\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.03$
+%\item[b]$\bar{T}_1 = 296.03 K $  - $\bar{T}_2 = 304.99 K$
+%\end{tablenotes}
+%\end{threeparttable}
+%\end{table}
+%\endgroup
+
 \begin{table}
-\setlength{\tabcolsep}{7pt}
 	\begin{threeparttable}
@@
 		\label{table:ensemblevalidation}
-\centering % centering table
-\begin{tabular}{|l c  c  c  c |}  
+		\begin{ruledtabular}
+			\begin{tabular}{ccccc}
+				& \multicolumn{4}{c}{true $\Delta T$ = 8.970 K} \\
+				\cline{2-5}
+				& \multicolumn{2}{c}{potential} & \multicolumn{2}{c}{kinetic}\\
 				\hline
- & \multicolumn{4}{c|}{true $\Delta T$ = 8.970 K}\\
-\cline{2-5}
-  & \multicolumn{2}{c}{potential} & \multicolumn{2}{c|}{kinetic}\\
-\hline
-Method  &$\Delta T/K$ &$\sigma$ dev &$\Delta T/K$&$\sigma$ dev\\
+				Method  &$\Delta T/K$ & $\sigma$ dev & $\Delta T/K$ & $\sigma$ dev \\
 				\hline % inserts single-line 
- Eq.~$\ref{eq:trotter_splitting_NHC}$   & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
- Eq.~$\ref{eq:trotter_splitting_*}$   & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14\\
-  \hline
+				Eq.~\ref{eq:trotter_splitting_NHC} & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
+				Eq.~\ref{eq:modified_splitting}    & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14
 		\end{tabular}
+		\end{ruledtabular}
 		\begin{tablenotes}
-\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.03$
-\item[b]$\bar{T}_1 = 296.03 K $  - $\bar{T}_2 = 304.99 K$
+			\item[a] The standard deviation of the mean temperature for all cases is $\sigma$ = 0.03.
+			\item[b] $\bar{T}_1$ = 296.03 K and $\bar{T}_2$ = 304.99 K.
 		\end{tablenotes}
@@
 \end{table}
-\endgroup
 
@@
 
-In Figure~\ref{fig:pressure} we compare the results using Eq.~\ref{eq:pressure} with those from a simulation performed with LAMMPS, which uses a numerical estimation of the constraint forces in the pressure calculation.
+In Fig.~\ref{fig:pressure} we compare the results using Eq.~\ref{eq:virial_rigid_bodies} with those from a simulation performed with LAMMPS, which uses a numerical estimation of the constraint forces in the pressure calculation.
 The simulations were run in the NVE ensemble, for which neither shifting to the LJ forces nor analytical tail corrections were considered.
@@
 
-\begin{center}
+\begin{figure}
+	\includegraphics[width=0.5\textwidth,keepaspectratio]{FiguraAna}
+	\caption{Histogram of pressure values obtained using Eq.~\ref{eq:virial_rigid_bodies} (filled bars), along with the results from a simulation using LAMMPS (unfilled bars).
+The average pressures are $\bar{P} = -0.60$~bar and $\bar{P} = 2.14$~bar, respectively.
+Simulation of liquid water with NVE integration, $\Delta t = 1$~fs, and $n = 1$.}
 	\label{fig:pressure}
-\includegraphics[width=0.5\textwidth,keepaspectratio]{FiguraAna}
-\captionof{figure}{Histogram of pressure values obtained using Eq.~\ref{eq:pressure} (filled bars), along with the results from a simulation using LAMMPS (unfilled bars).
-The average pressures are $\bar{P} = -0.60 $ bar and $\bar{P} = 2.14 $ bar, respectively.
-Simulation of liquid water with NVE integration, $\Delta t = 1$ fs, $n_{ys} = 1$}
-\end{center}
+\end{figure}
+
+\section{Conclusion}
+
+\begin{acknowledgments}
+The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
+\end{acknowledgments}
 
@@
 
-\section{\label{sec:quat_from_A}Obtaining the quaternion components from a rotation matrix}
-
-The method described here was introduced by Shepperd.\cite{Shepperd1978} As one can observe in Eq.~\ref{eq:A_from_q}, the three main diagonal components of $\mt A$ and the fact that $\lVert \vt q \lVert^2 = 1$ provide a linear system for the squares of the quaternion components whose solution is
-\begin{align*}
-4 q_0^2 &= 1 + tr(\mt A) \\
-4 q_k^2 &= 1 + A_{k,k} - tr(\mt A)
-\end{align*}
-where $k = 1,2,3$ and $tr(\mt A)$ is the trace of matrix $\mt A$.
-Since both $\vt q$ and $- \vt q$ correspond to the same rotation matrix, the sign the largest component\cite{Shepperd1978} can be made positive and the others can be obtained from the off-diagonal entries of $\mt A$, which give
-\begin{align*}
-4 q_0 q_1 &= A_{2,3} - A_{3,2} \\
-4 q_0 q_2 &= A_{3,1} - A_{1,3} \\
-4 q_0 q_3 &= A_{1,2} - A_{2,1} \\
-4 q_k q_l &= A_{k,l} + A_{l,k} \\
-\end{align*}
-where $k,l = 1,2,3$ with $k \neq l$.
-
-\section{\label{sec:auxiliary_math}Algebraic relations involving unit quaternions}
+\section{Obtaining quaternion components from a rotation matrix}
+\label{sec:quat_from_A}
+
+The method described here was introduced by Shepperd.\cite{Shepperd1978} As one can observe in Eq.~\ref{eq:A_from_q}, the three main diagonal components of $\mt A$ and the fact that $\lVert \vt q \lVert^2 = 1$ provide a linear system for the squares of the quaternion components, whose solution is $4 q_0^2 = 1 + \text{tr}\mt A$ and $4 q_k^2 = 1 + A_{k,k} - \text{tr}\mt A$ for $k = 1, 2, 3$, where $\text{tr}\mt A$ is the trace of matrix $\mt A$.
+As both $\vt q$ and $- \vt q$ correspond to the same rotation matrix, the sign of the largest component\cite{Shepperd1978} can be made positive and the others can be obtained from the off-diagonal entries of $\mt A$, which give $4 q_0 q_1 = A_{2,3} - A_{3,2}$, $4 q_0 q_2 = A_{3,1} - A_{1,3}$, $4 q_0 q_3 = A_{1,2} - A_{2,1}$, and $4 q_k q_l = A_{k,l} + A_{l,k}$, where $k,l = 1,2,3$ with $k \neq l$.
+
+\section{Differentiation rules with vectors}
+\label{sec:Diff_Rules}
+
+Along the paper, the differentiation of scalar functions with respect to vectors are done directly using the rules
+\[
+\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a \quad \text{and} \quad \diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = 2 \mt A \vt x,
+\]
+where $\vt a$ is a constant vector and $\mt A$ is a constant matrix.
+
+\section{Algebraic relations involving unit quaternions}
+\label{sec:auxiliary_math}
 
@@
 \[
-{\mt B}_1 = \left[ \begin{array}{rrrr}
- 0 & -1 &  0 &  0 \\
+{\mt B}_1 = \left[ \begin{smallmatrix}
+ 0 & \text{-}1 &  0 &  0 \\
  1 &  0 &  0 &  0 \\
  0 &  0 &  0 &  1 \\
- 0 &  0 & -1 &  0 \\
-\end{array} \right], \;
-{\mt B}_2 = \left[ \begin{array}{rrrr}
- 0 &  0 & -1 &  0 \\
- 0 &  0 &  0 & -1 \\
+ 0 &  0 & \text{-}1 &  0 \\
+\end{smallmatrix} \right], \,
+{\mt B}_2 = \left[ \begin{smallmatrix}
+ 0 &  0 & \text{-}1 &  0 \\
+ 0 &  0 &  0 & \text{-}1 \\
  1 &  0 &  0 &  0 \\
  0 &  1 &  0 &  0 \\
-\end{array} \right],
-\]\[
-\text{and }{\mt B}_3 = \left[ \begin{array}{rrrr}
- 0 &  0 &  0 & -1 \\
+\end{smallmatrix} \right], \text{ and }
+{\mt B}_3 = \left[ \begin{smallmatrix}
+ 0 &  0 &  0 & \text{-}1 \\
  0 &  0 &  1 &  0 \\
- 0 & -1 &  0 &  0 \\
+ 0 & \text{-}1 &  0 &  0 \\
  1 &  0 &  0 &  0 \\
-\end{array} \right].
-\]
+\end{smallmatrix} \right].
+\]
+
+%\[
+%{\mt B}_1 = \left[\begin{smallmatrix}
+%0 & \text{-}1 &  0 &  0 \\
+%1 &  0 &  0 &  0 \\
+%0 &  0 &  0 &  1 \\
+%0 &  0 & \text{-}1 &  0 \\
+%\end{smallmatrix}, \;
+%{\mt B}_2 = \begin{bmatrix}
+%0 &  0 & \text{-}1 &  0 \\
+%0 &  0 &  0 & \text{-}1 \\
+%1 &  0 &  0 &  0 \\
+%0 &  1 &  0 &  0 \\
+%\end{bmatrix},
+%{\mt B}_3 = \begin{bmatrix}
+%0 &  0 &  0 & \text{-}1 \\
+%0 &  0 &  1 &  0 \\
+%0 & \text{-}1 &  0 &  0 \\
+%1 &  0 &  0 &  0 \\
+%\end{bmatrix}.
+%\]
 
@@
 \[
-{\mt C}_1 = \left[ \begin{array}{rrrr}
- 0 & -1 &  0 &  0 \\
+{\mt C}_1 = \left[ \begin{smallmatrix}
+ 0 & \text{-}1 &  0 &  0 \\
  1 &  0 &  0 &  0 \\
- 0 &  0 &  0 & -1 \\
+ 0 &  0 &  0 & \text{-}1 \\
  0 &  0 &  1 &  0 \\
-\end{array} \right], \;
-{\mt C}_2 = \left[ \begin{array}{rrrr}
- 0 &  0 & -1 &  0 \\
+\end{smallmatrix} \right], \,
+{\mt C}_2 = \left[ \begin{smallmatrix}
+ 0 &  0 & \text{-}1 &  0 \\
  0 &  0 &  0 &  1 \\
  1 &  0 &  0 &  0 \\
- 0 & -1 &  0 &  0 \\
-\end{array} \right],
-\]\[
-\text{and }{\mt C}_3 = \left[ \begin{array}{rrrr}
- 0 &  0 &  0 & -1 \\
- 0 &  0 & -1 &  0 \\
+ 0 & \text{-}1 &  0 &  0 \\
+\end{smallmatrix} \right], \text{ and }
+{\mt C}_3 = \left[ \begin{smallmatrix}
+ 0 &  0 &  0 & \text{-}1 \\
+ 0 &  0 & \text{-}1 &  0 \\
  0 &  1 &  0 &  0 \\
  1 &  0 &  0 &  0 \\
-\end{array} \right].
+\end{smallmatrix} \right].
 \]
@@
 
-\section{\label{sec:Diff_Rules}Differentiation rules with vectors}
-
-Along the paper, the differentiation of scalar functions with respect to vectors are done directly using the following rules:
-\[
-\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a
-\]
-and
-\[
-\diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = 2 \mt A \vt x,
-\]
-where $\vt a$ is vector and $\mt A$ is a matrix, both independent of $\vt x$.
-
-\section{\label{sec:Diff_PotEng}Potential energy differentiation}
+\section{Potential energy differentiation}
+\label{sec:Diff_PotEng}
 

### 2016-07-20 rigid_bodies.tex
Introduction: check

@@
 
-Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and length scales attainable in their atomistic simulations.
+Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and the length scales attainable in their atomistic simulations.
 This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
@@
 This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles.
-However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges a suitable approach.
+However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
 It is expected to be computationally more efficient because the constraints are treated implicitly.
 In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
-Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
-This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} etc.
-It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy often used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
+Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
+This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so on.
+It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
 
 In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as corresponding numerical integrators.
-The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat.
-This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two independent thermostat chains act separately on the rotational and the translational degrees of freedom.
-We have also simplified the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions (PBC).
-The expression we derive depends on the resultant interaction forces between body pairs.
-This avoids the evaluation of constraint forces, which are necessary if the pressure is calculated in terms of the interatomic forces.
-
-This paper is organized as follows: in Sec.~\ref{sec:mechanics} we present a mathematical reformulation of the equations of motion of rigid bodies using a Hamiltonian formalism.
-In Sec.~\ref{sec:moleculardynamics} we devise numerical integrators for both the NVE and NVT ensembles, whose correctness we discuss in terms of conserved quantities.
-Additionally, for the NVE case we present some results related to kinetic energy partition and, for the NVT case, we also analyze the statistical distributions of the sampled energy.
-After that, in Sec.~\ref{sec:pressure} we present the equation developed for the pressure.
-
-Note that, throughout this paper, we employ the word ``quaternion'' as a synonym for a four-dimensional vector and avoid other interpretations.
-Only ordinary matrix/vector operations are present in the formulation.
+The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat.
+This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom.
+We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
+For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and another one that corrects a pressure calculated in terms of interactions among free atoms.
+
+The paper is organized as follows.
+In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions.
+In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT ensembles, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
+Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure.
+Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
+
+Throughout this paper, we use the word ``quaternion'' as a synonym for a four-dimensional vector, avoiding other interpretations.
+Thus, only ordinary matrix/vector operations are present in the formulation.
 Some mathematical details are given in the Appendix.
@@
 \section{Numerical Results}
+\label{sec:numerical_results}
 
@@
 
-\section{Conclusion}
+\section{Concluding Remarks}
+\label{sec:conclusion}
 
@@
 orthogonality of $\vt y$ and $\vt q$) for the operation to be reversible by $\mt B$.
-In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999}
-
-All the conclusions drawn thus far for matrix $\mt B$ are equally valid for matrix $\mt C$.
-
-As the columns of $\mt B$ are permutations of the Euler parameters, one can write $\mt B = [\begin{array}{ccc}{\mt B}_1{\vt q} & {\mt B}_2{\vt q} & {\mt B}_3{\vt q}\end{array}]$, where $\mt B_1$, $\mt B_2$, and $\mt B_3$ are permutation matrices given by
+In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999} As the columns of $\mt B$ are permutations of the Euler parameters, one can write $\mt B = [\begin{array}{ccc}{\mt B}_1{\vt q} & {\mt B}_2{\vt q} & {\mt B}_3{\vt q}\end{array}]$, where $\mt B_1$, $\mt B_2$, and $\mt B_3$ are permutation matrices given by
 \[
@@
 
+All the conclusions drawn thus far for matrix $\mt B$ are equally valid for matrix $\mt C$.
+
 Matrix $\mt C$ and its time-derivative $\dot{\mt C}$ can also be expressed as $[\begin{array}{ccc}{\mt C}_1{\vt q} & {\mt C}_2{\vt q} & {\mt C}_3{\vt q}\end{array}]$ and $[\begin{array}{ccc}{\mt C}_1\dot{\vt q} & {\mt C}_2\dot{\vt q} & {\mt C}_3\dot{\vt q}\end{array}]$, respectively, where $\mt C_1$, $\mt C_2$, and $\mt C_3$ are

### 2016-07-23 rigid_bodies.tex
Appendices almost concluded

@@
 %\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
-\documentclass[aip,jcp,reprint,amsmath,amssymb,raggedbottom]{revtex4-1}
+\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
 
 \usepackage[pdftex]{graphicx}
-\usepackage{epstopdf}
-\usepackage{caption}
-\usepackage{subfig}
-\usepackage[]{units}
-\usepackage[]{natbib}
+%\usepackage{epstopdf}
+%\usepackage{caption}
+%\usepackage{subfig}
+%\usepackage[]{units}
+%\usepackage[]{natbib}
 \usepackage[]{threeparttable}
-\usepackage{booktabs}
-\usepackage{multirow}
+%\usepackage{booktabs}
+%\usepackage{multirow}
 
@@
 \newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}           % vector symbol
-\newcommand{\tr}[1]{#1^t}                               % transposition
+\newcommand{\tr}[1]{#1^\text{t}}                               % transposition
 \newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
+
+\newenvironment{smallarray}[1]
+{\null\,\vcenter\bgroup\scriptsize
+	\renewcommand{\arraystretch}{0.7}
+	\arraycolsep=.13885em
+	\hbox\bgroup$\array{@{}#1@{}}}
+{\endarray$\egroup\egroup\,\null}
 
@@
 In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study.
-Unity quaternions are employed to represent the orientations of rigid bodies.
+Unit quaternions are employed to represent the orientations of rigid bodies.
 Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector.
-An important feature of this relation is that it preserves the unit-norm constraint of the quaternion upon infinitesimal motion.
+An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal motion.
 
@@
 \[
-{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{array}{ccc}
+{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{smallarray}{ccc}
 {\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y      & -{\delta_j}_x {\delta_j}_z      \\
@@
 -{\delta_j}_z {\delta_j}_x      & -{\delta_j}_z {\delta_j}_y      & {\delta_j}_x^2 + {\delta_j}_y^2 \\
-\end{array} \right].
+\end{smallarray} \right].
 \]
@@
 For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal.
-The corresponding axes are known as the principal axes and the formulation of the equations of motion in this frame of reference is notably simplified.
+The formulation of the equations of motion in this frame of reference is notably simplified.
 Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
 \[
-{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{array}{ccc}
+{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{smallarray}{ccc}
 I_1 &   0 &   0 \\
@@
   0 &   0 & I_3 \\
-\end{array} \right],
+\end{smallarray} \right],
 \]
 where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors.
-Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found, for example, in Ref.~\onlinecite{Kopp2008}.
-Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$.
-Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$ .
-By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that $\mt A \tr{\dot{\mt A}}$ results in a skew-symmetric matrix.
-This product is related to $\vt \omega$, the angular velocity of the body in the body-fixed frame, by\cite{Haug1989}
-\begin{equation}
-\label{eq:relation_A_omega}
-\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A},
-\end{equation}
-where $\mt S(\cdot)$ is the skew-symmetric operator defined in Eq.~\ref{eq:operator_S}.
-
-\subsection{The Rate of Change of Unit Quaternions}
-\label{sec:eulerparameters}
-
-Euler's rotation theorem establishes that an arbitrary rotation can be described using three independent parameters.
+Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found in the literature.\cite{Kopp2008} Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$.
+Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$.
+By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric.
+The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).
+
+An arbitrary rotation can be described using three independent parameters.
 One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in such terms.
@@
 Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
-\begin{widetext}
 \begin{equation}
 \label{eq:A_from_q}
-\mt A = \left[ \begin{array}{ccc}
+\mt A = \left[ \begin{smallarray}{ccc}
 q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3)          & 2(q_1 q_3 - q_0 q_2) \\
@@
 2(q_1 q_3 + q_0 q_2)          & 2(q_2 q_3 - q_0 q_1)          & q_0^2 - q_1^2 - q_2^2 + q_3^2  
-\end{array} \right].
-\end{equation}
-\end{widetext}
-
-The rate of change of a quaternion is denoted by $\dot{\vt q}$. Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
+\end{smallarray} \right].
+\end{equation}
+
+Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
 \begin{equation}
@@
 \end{equation}
-which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.
-
-Let us now introduce two new matrices that also depend on the Euler parameters, which are
-\begin{subequations}
+which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, the rate of change $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.
+
+Let us now introduce two new matrices that also depend on the quaternion components, which are
+\begin{equation}
 \label{eq:def_B_and_C}
-\begin{align}
-\mt B &= \left[
-\begin{array}{rrrr}
+\mt B = \left[
+\begin{smallarray}{rrrr}
 -q_1 & -q_2 & -q_3 \\
@@
 -q_2 &  q_1 &  q_0
-\end{array}
-\right] \text{ and} \label{eq:def_B} \\
-\mt C &= \left[
-\begin{array}{rrrr}
+\end{smallarray}
+\right]
+\quad \text{and} \quad
+\mt C = \left[
+\begin{smallarray}{rrrr}
 -q_1 & -q_2 & -q_3 \\
@@
  q_2 & -q_1 &  q_0
-\end{array}
+\end{smallarray}
 \right].
-\end{align}
-\end{subequations}
+\end{equation}
 
@@
 The equation above is a simple and convenient factorization of matrix $\mt A$, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters.
-As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the Euler parameters.
-Such relation is
+As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the quaternion components, which is
 \begin{equation}
@@
 
-As explained in the appendix, premultiplication with either $\mt B$ or $\mt C$ transforms a vector into a quaternion that lies in the hyperplane orthogonal to $\vt q$.
-Therefore, the equation above satisfies Eq.~\ref{eq:diff_qTq} identically and guarantees that the norm of $\vt q$ remains unchanged.
+As explained in the mentioned appendix, premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$.
 Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.
+Therefore, one can define an intermediary frame of reference, fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body, whose connection to the body-fixed and the space-fixed frames are summarized as follows:
+\[
+\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
+\mathrel{\mathop{\rightleftarrows}^{\frac{1}{2}\mt B}_{\frac{1}{2}\tr{\mt B}}}
+\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
+\mathrel{\mathop{\rightleftarrows}^{2 \tr{\mt C}}_{2 \mt C}}
+\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
+\]
+
+Therefore, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply the angular velocity vector expressed in the quaternion-fixed frame.
+It also guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.
 
@@
 
-The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) obtained in the previous subsection is used here to develop a Hamiltonian formalism for the motion of rigid body.
+The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body.
 As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively.
@@
 \mt \Omega = \left[
-\begin{array}{rrrr}
+\begin{smallarray}{rrrr}
 -\pi_1 & -\pi_2 & -\pi_3 \\
@@
 -\pi_2 &  \pi_1 &  \pi_0
-\end{array}
+\end{smallarray}
 \right],
@@
 \]
-where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame. Note that the product $\mt C \vt \tau$ results in a quaternion orthogonal to $\vt q$.
-
-Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B} \vt \pi = \frac{1}{2} {\mt I}^{-1} \tr{\mt \Omega} \vt q$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
+where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame.
+Therefore, $-\partial U/\partial \vt q$ is simply the torque represented in the quaternion-fixed frame.
+Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
 \begin{subequations}
@@
 &\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
-&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau
+&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:euler_eq_quat}
 \end{align}
@@
 
-Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
+In Appendix \ref{eq:derivation_euler}, it is demonstrated that Eq.~\ref{eq:euler_eq_quat} is a quaternion-fixed frame representation of the well-known Euler equation.\cite{Goldstein2002} Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
 Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
@@
 \label{eq:performance}
-D\!E =  \frac{1}{N_\text{steps}} \sum_{k=1}^{N_\text{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|,
-\end{equation}
-from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $N_\text{steps}$.
+D\!E =  \frac{1}{\mathcal N} \sum_{t=1}^{\mathcal N} \left| \frac{E_t - E_0}{E_0} \right|,
+\end{equation}
+from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $\mathcal N$.
 Both the instantaneous fluctuations, given by the term inside the summation, and the long-time conservation of energy, $D\!E$, are monitored.
@@
 
-%\begin{table}[h]
-%\centering % centering table
-%\begin{threeparttable}
-%\caption{Partition of kinetic energy from an NVE simulation of 903 water molecules with $r_c = 9$ \AA~ and without shifting of LJ model\tnote{a}} %($5415 grados de libertad$)}
-%\begin{tabular}{| l c  c  c  c  c  c |}  
-%\toprule
-%\hline
-% & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
-%  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
-%\cline{4-7}
-%$\Delta t/fs $  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
-%\hline % inserts single-line
-% 0.5 & 305.04 & 1641.21  & 820.15  & 821.06  & 820.55 & 820.66  \\
-%  1  & 304.88  & 1640.36  & 819.73  & 820.63  & 820.56 & 819.80  \\
-%  2  & 304.04  & 1635.87  & 817.48  & 818.39  & 820.44 & 815.43 \\
-%  3  & 302.78  & 1629.10  & 814.10  & 815.00  & 819.85 & 809.25 \\
-%  \hline
-%\end{tabular}
-%\begin{tablenotes}
-%\item[a] Energies are in $kcal/mol$.
-%\end{tablenotes}
-%\label{table:water_partition_2}
-%\end{threeparttable}
-%\end{adjustbox}
-%\end{table}
-
 \begin{table}
 	\begin{threeparttable}
-		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c = 9$ {\AA} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
+		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c$ = 9~{\r A} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
 		\label{table:water_partition_LJ-SF}
@@
 	\begin{threeparttable}
-		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c = 13$ {\AA} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
+		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c$ = 13~{\r A} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
 		\label{table:water_partition_13A_LJ-SF}
@@
 \label{sec:samplingvalidation}
+
 In this subsection we present the results from a simple method that serves for checking the consistency between the sampled energy distributions and the theoretical ones.
@@
 This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
-
-%\begingroup
-%\squeezetable
-%\begin{table}
-%\setlength{\tabcolsep}{7pt}
-%\begin{threeparttable}
-%\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach \tnote{a} \tnote{b}}
-%\label{table:ensemblevalidation_old}
-%\centering % centering table
-%\begin{tabular}{|l c  c  c  c |}  
-%\hline
-% & \multicolumn{4}{c|}{true $\Delta T$ = 8.970 K}\\
-%\cline{2-5}
-%  & \multicolumn{2}{c}{potential} & \multicolumn{2}{c|}{kinetic}\\
-%\hline
-%Method  &$\Delta T/K$ &$\sigma$ dev &$\Delta T/K$&$\sigma$ dev\\
-%\hline % inserts single-line 
-% Eq.~$\ref{eq:trotter_splitting_NHC}$   & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
-% Eq.~$\ref{eq:trotter_splitting_*}$   & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14\\
-%  \hline
-%\end{tabular}
-%\begin{tablenotes}
-%\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.03$
-%\item[b]$\bar{T}_1 = 296.03 K $  - $\bar{T}_2 = 304.99 K$
-%\end{tablenotes}
-%\end{threeparttable}
-%\end{table}
-%\endgroup
 
@@
 			\item[a] The standard deviation of the mean temperature for all cases is $\sigma$ = 0.03.
-			\item[b] $\bar{T}_1$ = 296.03 K and $\bar{T}_2$ = 304.99 K.
+			\item[b] $\bar{T}_1$ = 296.03~K and $\bar{T}_2$ = 304.99~K.
 		\end{tablenotes}
@@
 The simulations were run in the NVE ensemble, for which neither shifting to the LJ forces nor analytical tail corrections were considered.
+
+Our purpose is....
+
+As one can see in Fig.~\ref{fig:pressure}....
 
@@
 \[
-\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a \quad \text{and} \quad \diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = 2 \mt A \vt x,
-\]
-where $\vt a$ is a constant vector and $\mt A$ is a constant matrix.
+\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a \quad \text{and} \quad \diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = (\mt A + \tr{\mt A})\vt x,
+\]
+where $\vt a$ is a constant vector and $\mt A$ is a constant matrix. If $\mt A$ is symmetric, then $\mt A + \tr{\mt A} = 2\mt A$.
 
@@
 
-Let us analyze the matrix $\mt B$ defined as in Eq.~\ref{eq:def_B_and_C}.
+We start by analyzing the matrix $\mt B$ as defined in Eq.~\ref{eq:def_B_and_C}.
 Note that $\tr{\mt B}\vt q = \vt 0$, where $\vt 0$ is a null vector.\cite{Haug1989, Shuster1993, Dichmann1999} This means that all columns of $\mt B$ are orthogonal to $\vt q$.
@@
 Other two identities that hold for $\mt B$ are\citep{Haug1989}
-\begin{align*}
-\tr{\mt B}\mt B &= \lVert \vt q \lVert ^2{\mt 1}_3 \text{ and} \\
-\mt B\tr{\mt B} &= \lVert \vt q \lVert ^2{\mt 1}_4 - {\vt q}\tr{\vt q}.
-\end{align*}
-
-If $\lVert \vt q \lVert^2 = 1$, then the operation $\vt y = \mt B \vt x$ can always be reverted by $\tr{\mt B}$.
+\begin{equation}
+\label{eq:basics_B}
+\begin{aligned}
+\tr{\mt B}\mt B &= (\tr{\vt q}{\vt q}) {\mt 1}_3 \text{ and} \\
+\mt B\tr{\mt B} &= (\tr{\vt q}{\vt q}) {\mt 1}_4 - {\vt q}\tr{\vt q}.
+\end{aligned}
+\end{equation}
+
+If $\tr{\vt q}{\vt q} = 1$, then the operation $\vt y = \mt B \vt x$ can always be reverted by $\tr{\mt B}$.
 This is not true for a general operation in the opposite direction, $\vt x = \tr{\mt B} \vt y$.
@@
 orthogonality of $\vt y$ and $\vt q$) for the operation to be reversible by $\mt B$.
-In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999} As the columns of $\mt B$ are permutations of the Euler parameters, one can write $\mt B = [\begin{array}{ccc}{\mt B}_1{\vt q} & {\mt B}_2{\vt q} & {\mt B}_3{\vt q}\end{array}]$, where $\mt B_1$, $\mt B_2$, and $\mt B_3$ are permutation matrices given by
-\[
-{\mt B}_1 = \left[ \begin{smallmatrix}
+In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999} As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = [\begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}]$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
+\[
+\hat{\mt B}_1 = \left[ \begin{smallarray}{cccc}
  0 & \text{-}1 &  0 &  0 \\
@@
  0 &  0 & \text{-}1 &  0 \\
-\end{smallmatrix} \right], \,
-{\mt B}_2 = \left[ \begin{smallmatrix}
+\end{smallarray} \right] \quad
+\hat{\mt B}_2 = \left[ \begin{smallarray}{cccc}
  0 &  0 & \text{-}1 &  0 \\
@@
  0 &  1 &  0 &  0 \\
-\end{smallmatrix} \right], \text{ and }
-{\mt B}_3 = \left[ \begin{smallmatrix}
+\end{smallarray} \right] \quad
+\hat{\mt B}_3 = \left[ \begin{smallarray}{cccc}
  0 &  0 &  0 & \text{-}1 \\
@@
  1 &  0 &  0 &  0 \\
-\end{smallmatrix} \right].
-\]
-
-%\[
-%{\mt B}_1 = \left[\begin{smallmatrix}
-%0 & \text{-}1 &  0 &  0 \\
-%1 &  0 &  0 &  0 \\
-%0 &  0 &  0 &  1 \\
-%0 &  0 & \text{-}1 &  0 \\
-%\end{smallmatrix}, \;
-%{\mt B}_2 = \begin{bmatrix}
-%0 &  0 & \text{-}1 &  0 \\
-%0 &  0 &  0 & \text{-}1 \\
-%1 &  0 &  0 &  0 \\
-%0 &  1 &  0 &  0 \\
-%\end{bmatrix},
-%{\mt B}_3 = \begin{bmatrix}
-%0 &  0 &  0 & \text{-}1 \\
-%0 &  0 &  1 &  0 \\
-%0 & \text{-}1 &  0 &  0 \\
-%1 &  0 &  0 &  0 \\
-%\end{bmatrix}.
-%\]
-
-All these permutation matrices are skew-symmetric (i.e. $\tr{\mt B_k} = -\mt B_k$) and also satisfy the relations\cite{Dichmann1999}
+\end{smallarray} \right]
+\]
+
+The product of $\mt B$ with an arbitrary vector $\vt x$ results in
+\begin{equation}
+\label{eq:product_B_vector}
+\mt B \vt x = \left[ \begin{smallarray}{ccc} \hat{\mt B}_1 \vt q & \hat{\mt B}_2 \vt q & \hat{\mt B}_3 \vt q \end{smallarray}\right] \left[ \begin{smallarray}{c} x_1 \\ x_2 \\ x_3 \end{smallarray} \right] = \sum_{j=1}^3 x_j \hat{\mt B}_j \vt q.
+\end{equation}
+
+Similarly, the product of $\tr{\mt B}$ with an arbitrary quaternion $\vt \pi$ results in
+\begin{equation}
+\label{eq:vector_entries}
+\tr{\mt B}\vt \pi =
+\left[\begin{smallarray}{c}
+\tr{\vt q}\tr{\hat{\mt B}}_1 \\
+\tr{\vt q}\tr{\hat{\mt B}}_2 \\
+\tr{\vt q}\tr{\hat{\mt B}}_3 \\
+\end{smallarray}\right]\vt \pi = 
+\left[\begin{smallarray}{c}
+\tr{\vt q}\tr{\hat{\mt B}}_1\vt \pi \\
+\tr{\vt q}\tr{\hat{\mt B}}_2\vt \pi \\
+\tr{\vt q}\tr{\hat{\mt B}}_3\vt \pi \\
+\end{smallarray}\right] = 
+\left[\begin{smallarray}{c}
+\tr{\vt \pi}{\hat{\mt B}}_1\vt q \\
+\tr{\vt \pi}{\hat{\mt B}}_2\vt q \\
+\tr{\vt \pi}{\hat{\mt B}}_3\vt q \\
+\end{smallarray}\right].
+\end{equation}
+
+We now turn the attention to the product $\tr{\mt B}\dot{\mt B}$, where $\dot{\mt B} = \left[\begin{smallarray}{ccc} \hat{\mt B}_1\dot{\vt q} & \hat{\mt B}_2\dot{\vt q} & \hat{\mt B}_3\dot{\vt q}\end{smallarray}\right]$ is the time-derivative of matrix $\mt B$.
+Each entry of this product is $\{\tr{\mt B}\dot{\mt B}\}_{ij} = \tr{\vt q}\tr{\hat{\mt B}_i} \hat{\mt B}_j\dot{\vt q}$.
+By noting that products involving the permutation matrices satisfy the relations\cite{Dichmann1999}
 \begin{equation}
@@
 \begin{aligned}
-\tr{\mt B_1}{\mt B_1} &= \tr{\mt B_2}{\mt B_2} = \tr{\mt B_3}{\mt B_3} = \mt 1_3 \\
-\tr{\mt B_1}{\mt B_2} &= -\tr{\mt B_2}{\mt B_1} = -\tr{\mt B_3} \\
-\tr{\mt B_2}{\mt B_3} &= -\tr{\mt B_3}{\mt B_2} = -\tr{\mt B_1} \\
-\tr{\mt B_3}{\mt B_1} &= -\tr{\mt B_1}{\mt B_3} = -\tr{\mt B_2}
+&\tr{\hat{\mt B}}_1 \hat{\mt B}_1 = \tr{\hat{\mt B}}_2 \hat{\mt B}_2 = \tr{\hat{\mt B}}_3 \hat{\mt B}_3 = \mt 1_4, \\
+&\tr{\hat{\mt B}}_1 \hat{\mt B}_2 = -\tr{\hat{\mt B}}_2 \hat{\mt B}_1 = -\tr{\hat{\mt B}_3}, \\
+&\tr{\hat{\mt B}}_2 \hat{\mt B}_3 = -\tr{\hat{\mt B}}_3 \hat{\mt B}_2 = -\tr{\hat{\mt B}_1}, \; \text{and} \\
+&\tr{\hat{\mt B}}_3 \hat{\mt B}_1 = -\tr{\hat{\mt B}}_1 \hat{\mt B}_3 = -\tr{\hat{\mt B}_2}, \\
 \end{aligned}
 \end{equation}
-
-The product of $\mt B$ with an arbitrary vector $\vt x$ results in
-\begin{equation}
-\label{eq:product_B_vector}
-\mt B \vt x = \left[ \begin{array}{ccc} \mt B_1 \vt q & \mt B_2 \vt q & \mt B_3 \vt q \end{array}\right] \left[ \begin{array}{c} x_1 \\ x_2 \\ x_3 \end{array} \right] = \sum_{k=1}^3 x_k {\mt B}_k \vt q.
-\end{equation}
-
-Similarly, the product of $\tr{\mt B}$ with an arbitrary quaternion $\vt y$ results in
-\begin{equation}
-\label{eq:vector_entries}
-\tr{\mt B}\vt y =
-\left[\begin{array}{c}
-\tr{\vt q}\tr{\mt B}_1 \\
-\tr{\vt q}\tr{\mt B}_2 \\
-\tr{\vt q}\tr{\mt B}_3 \\
-\end{array}\right]\vt y = 
-\left[\begin{array}{c}
-\tr{\vt q}\tr{\mt B}_1\vt y \\
-\tr{\vt q}\tr{\mt B}_2\vt y \\
-\tr{\vt q}\tr{\mt B}_3\vt y \\
-\end{array}\right] = 
-\left[\begin{array}{c}
-\tr{\vt y}{\mt B}_1\vt q \\
-\tr{\vt y}{\mt B}_2\vt q \\
-\tr{\vt y}{\mt B}_3\vt q \\
-\end{array}\right].
-\end{equation}
-
-Let us now consider the product $\tr{\mt B}\dot{\mt B}$, where $\dot{\mt B}$ is the time derivative of $\mt B$.
-Because the permutation matrices are constant, we can express $\dot{\mt B} = [\begin{array}{ccc}{\mt B}_1\dot{\vt q} & {\mt B}_2\dot{\vt q} & {\mt B}_3\dot{\vt q}\end{array}]$.
-By carrying out the multiplication and applying the relations of Eq.~\ref{eq:BB_products}, we observe that
-\[
+and recalling that $\vt q$ and $\dot{\vt q}$ are mutually orthogonal (Eq.~\ref{eq:diff_qTq}), we can obtain a simplified expression for $\tr{\mt B}\dot{\mt B}$, which is
+\begin{equation}
+\label{eq:BtBdot}
 \tr{\mt B}\dot{\mt B} = \left[
-\begin{array}{ccc}
-0 & -\tr{\vt q} \tr{\mt B_3}\dot{\vt q} & \tr{\vt q} \tr{\mt B_2}\dot{\vt q} \\
-\tr{\vt q} \tr{\mt B_3}\dot{\vt q} & 0 & -\tr{\vt q} \tr{\mt B_1}\dot{\vt q} \\
--\tr{\vt q} \tr{\mt B_2}\dot{\vt q} & \tr{\vt q} \tr{\mt B_1}\dot{\vt q}  & 0
-\end{array}
-\right],
-\]
-where the diagonal entries are zero as a consequence of the orthogonality between $\vt q$ and $\dot{\vt q}$. By contrasting this result with Eq.~\ref{eq:operator_S}, we draw the conclusion that
+\begin{smallarray}{ccc}
+0 & -\tr{\vt q} \tr{\hat{\mt B}_3}\dot{\vt q} &  \tr{\vt q} \tr{\hat{\mt B}_2}\dot{\vt q} \\
+\tr{\vt q} \tr{\hat{\mt B}_3}\dot{\vt q} & 0 & -\tr{\vt q} \tr{\hat{\mt B}_1}\dot{\vt q} \\
+-\tr{\vt q} \tr{\hat{\mt B}_2}\dot{\vt q} & \tr{\vt q} \tr{\hat{\mt B}_1}\dot{\vt q} & 0
+\end{smallarray}
+\right].
+\end{equation}
+
+Skew-symmetric matrices like the one above are associated to cross products.
+For arbitrary vectors $\vt x$ and $\vt y$ in $\mathbb R^3$, it turns out that $\vt x \times \vt y = \mt S(\vt x)\vt y$, where $\mt S(\cdot)$ is an operator defined as
+\begin{equation}
+\label{eq:operator_S}
+\mt S(\vt x) = \left[ \begin{smallarray}{ccc}
+0   & -x_3 &  x_2 \\
+x_3 &  0   & -x_1 \\
+-x_2 &  x_1 &  0
+\end{smallarray}\right].
+\end{equation}
+
+By contrasting the definition above with Eqs.~\ref{eq:vector_entries} and \ref{eq:BtBdot}, it becomes clear that
 \begin{equation}
@@
 
-All the conclusions drawn thus far for matrix $\mt B$ are equally valid for matrix $\mt C$.
-
-Matrix $\mt C$ and its time-derivative $\dot{\mt C}$ can also be expressed as $[\begin{array}{ccc}{\mt C}_1{\vt q} & {\mt C}_2{\vt q} & {\mt C}_3{\vt q}\end{array}]$ and $[\begin{array}{ccc}{\mt C}_1\dot{\vt q} & {\mt C}_2\dot{\vt q} & {\mt C}_3\dot{\vt q}\end{array}]$, respectively, where $\mt C_1$, $\mt C_2$, and $\mt C_3$ are
-\[
-{\mt C}_1 = \left[ \begin{smallmatrix}
+We can also express matrix $\mt C$ as $\left[\begin{smallarray}{ccc} \hat{\mt C}_1{\vt q} & \hat{\mt C}_2{\vt q} & \hat{\mt C}_3{\vt q}\end{smallarray}\right]$, where $\hat{\mt C}_1$, $\hat{\mt C}_2$, and $\hat{\mt C}_3$ are the following skew-symmetric permutation matrices:
+\[
+\hat{\mt C}_1 = \left[ \begin{smallarray}{cccc}
  0 & \text{-}1 &  0 &  0 \\
@@
  0 &  0 &  1 &  0 \\
-\end{smallmatrix} \right], \,
-{\mt C}_2 = \left[ \begin{smallmatrix}
+\end{smallarray} \right] \quad
+\hat{\mt C}_2 = \left[ \begin{smallarray}{cccc}
  0 &  0 & \text{-}1 &  0 \\
@@
  0 & \text{-}1 &  0 &  0 \\
-\end{smallmatrix} \right], \text{ and }
-{\mt C}_3 = \left[ \begin{smallmatrix}
+\end{smallarray} \right] \quad
+\hat{\mt C}_3 = \left[ \begin{smallarray}{cccc}
  0 &  0 &  0 & \text{-}1 \\
@@
  1 &  0 &  0 &  0 \\
-\end{smallmatrix} \right].
-\]
-
-A special connection between the permutation matrices for $\mt B$ and $\mt C$ is the fact that $\tr{\mt B_i}{\mt C_j}$ is symmetric for any combination of $i$ and $j$.
-Thus, each entry of $\tr{\mt B}\dot{\mt C}$ is
-\begin{align*}
-\{\tr{\mt B}\dot{\mt C}\}_{ij} &= \tr{\vt q}\tr{\mt B}_i{\mt C}_j\dot{\vt q} = \tr{\vt q}\tr{\mt C}_j{\mt B}_i\dot{\vt q} = \{\tr{\mt C}\dot{\mt B}\}_{ji} = \\
-&= \{\tr{\dot{\mt B}}\mt C\}_{ij}.
-\end{align*}
-
-This means that $\tr{\mt B}\dot{\mt C} = \tr{\dot{\mt B}}{\mt C}$. Proceeding further, we observe that
-\begin{align*}
-\mt A \tr{\dot{\mt A}} &= \tr{\mt B}\mt C \tr{(\tr{{\mt B}}\dot{\mt C} + \tr{\dot{\mt B}}{\mt C})} = \tr{\mt B}\mt C \tr{(2 \tr{\dot{\mt B}}{\mt C})} = \\
-&= 2 \tr{\mt B}\dot{\mt B},
-\end{align*}
-where we have applied the relations $\mt C \tr{\mt C} = \mt 1_4 - \vt q\tr{\vt q}$ and $\tr{\vt B}\vt q = \vt 0$.
-Finally, resorting to Eqs.~\ref{eq:relation_A_omega} and \ref{eq:relation_B_qdot}, the result above converts to
-\begin{equation}
-\vt \omega = 2\tr{\mt B}\dot{\vt q}.
-\end{equation}
-
-\begin{equation}
-\label{eq:operator_S}
-\mt S(\vt x) = \left[ \begin{array}{ccc}
-0   & -x_3 &  x_2 \\
-x_3 &  0   & -x_1 \\
--x_2 &  x_1 &  0
-\end{array}\right].
-\end{equation}
-
-The operator  $\mt S(\cdot)$ is able to represent the cross-product of two vectors $\vt x$ and $\vt y$ as a matrix-vector product, since $\vt x \times \vt y = \mt S(\vt x)\vt y$.
-This property is explored in the Appendix.
-
-Demonstration of Euler equation:
-
-\begin{align*}
-\diff{(\tr{\mt B}\vt \pi)}{t} &= \tr{\dot{\mt B}}\vt \pi + \tr{\mt B}\dot{\vt \pi} = \\
-&=\tr{\dot{\mt B}}{\mt B}\tr{\mt B}{\vt \pi}  + \dfrac{1}{2} \tr{\mt B}\mt \Omega \vt \omega + 2 \tr{\mt B}\mt C \vt \tau = \\
-&= -\frac{1}{2}{\mt S}(\vt \omega)\tr{\mt B}{\vt \pi} + \dfrac{1}{2} {\mt S}(\tr{\mt B}\mt \pi) \vt \omega + 2 \mt A \vt \tau = \\
-&= - \vt \omega \times (\tr{\mt B}{\vt \pi}) + 2 {\vt \tau}^\prime
-\end{align*}
-
-As $\tr{\mt B}\vt \pi = 2 \mt I \vt \omega$, the result above becomes
-\[
-\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = {\vt \tau}^\prime
-\]
+\end{smallarray} \right]
+\]
+
+However, products involving these matrices are slightly different from those in Eq.~\ref{eq:BB_products}. They are
+\begin{equation}
+\label{eq:CC_products}
+\begin{aligned}
+&\tr{\hat{\mt C}}_1 \hat{\mt C}_1 = \tr{\hat{\mt C}}_2 \hat{\mt C}_2 = \tr{\hat{\mt C}}_3 \hat{\mt C}_3 = \mt 1_4, \\
+&\tr{\hat{\mt C}}_1 \hat{\mt C}_2 = -\tr{\hat{\mt C}}_2 \hat{\mt C}_1 = \tr{\hat{\mt C}_3}, \\
+&\tr{\hat{\mt C}}_2 \hat{\mt C}_3 = -\tr{\hat{\mt C}}_3 \hat{\mt C}_2 = \tr{\hat{\mt C}_1}, \; \text{and} \\
+&\tr{\hat{\mt C}}_3 \hat{\mt C}_1 = -\tr{\hat{\mt C}}_1 \hat{\mt C}_3 = \tr{\hat{\mt C}_2}. \\
+\end{aligned}
+\end{equation}
+
+A special connection between the permutation matrices that make $\mt B$ and $\mt C$ is the fact that $\tr{\hat{\mt B}_i}{\hat{\mt C}_j} = \tr{\hat{\mt C}_j}{\hat{\mt B}_i}$ for any combination of $i$ and $j$.
+A direct consequence is that $\tr{\vt q}\tr{\hat{\mt B}}_i\hat{\mt C}_j\dot{\vt q} = \tr{\dot{\vt q}}\tr{\hat{\mt B}}_i\hat{\mt C}_j{\vt q}$, which means that $\tr{\mt B}\dot{\mt C} = \tr{\dot{\mt B}}{\mt C}$.
+Thus, differentiating Eq.~\ref{eq:factorization_of_A} with respect to time, we conclude that $\dot{\mt A} = \tr{\mt B}\dot{\mt C} + \tr{\dot{\mt B}}{\mt C} = 2 \tr{\dot{\mt B}}{\mt C}$.
+Note that Eq.~\ref{eq:basics_B} also applies for matrix $\mt C$, which means that, as far as $\tr{\vt q}\vt q = 1$, we can insert the product $\mt C\tr{\mt C}$ between $\tr{\mt B}$ and $\dot{\mt B}$ in Eq.~\ref{eq:relation_B_qdot} without altering its result.
+Then, using Eq.~\ref{eq:factorization_of_A} again and the conclusion above about $\dot{\mt A}$, we have that
+\begin{equation}
+\frac{1}{2} {\mt A}\tr{\dot{\mt A}} = {\mt S}\left( \tr{\mt B}\dot{\vt q} \right).
+\end{equation}
+
+As mentioned in Sec.~\ref{sec:kinematics}, ${\mt A}\tr{\dot{\mt A}} = \mt S(\vt \omega)$, which makes the equation above result in Eq.~\ref{eq:relation_qdot_omega}.
 
@@
 
-Care must be taken when differentiating the potential energy $U$ with respect to $\vt q$ because the Euler parameters are not all independent.
-Schay demonstrated in Ref.~\cite{Schay1995} that, when differentiating a scalar function $f$ with respect to a vector variable $\vt x$, subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ is given by
+Care must be taken when differentiating the potential energy $U$ with respect to $\vt q$ because the quaternion components are interdependent.
+Schay demonstrated in Ref.~\onlinecite{Schay1995} that, when differentiating a scalar function $f$ with respect to a vector variable $\vt x$, subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by
 \[
@@
 where $\vt n$ is the unit vector proportional to $(\partial g/\partial \vt x)^\ast$, with the asterisk denoting a differentiation performed without considering the interdependence of the entries of $\vt x$.
-We note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\cite{Schay1995}, where the Jacobian convention is used.
-In the  case of Euler parameters,
-\[
-g(\vt q) = \tr{\vt q}\vt q - 1 \; \rightarrow \; \left(\diff{g}{\vt q}\right)^\ast = 2 \vt q.
-\]
-
-Hence, the unit four-dimensional vector proportional to $(\partial g/\partial \vt x)^\ast$ is exactly $\vt q$, which means that the constrained derivative of $U$ is given by
+We note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\onlinecite{Schay1995}, where the Jacobian convention is used.
+In the case of a unit quaternion,
+\[
+g(\vt q) = \tr{\vt q}\vt q - 1 = 0 \; \rightarrow \; \left(\diff{g}{\vt q}\right)^\ast = 2 \vt q.
+\]
+
+Hence, the unit vector proportional to $(\partial g/\partial \vt x)^\ast$ is exactly $\vt q$, which means that the constrained derivative of $U$ is given by
 \begin{equation}
@@
 Therefore, the constrained derivative is the projection of the unconstrained one in the hyperplane orthogonal to $\vt q$.
-Recently, Nielsen and Krenk\cite{Nielsen2012} employed a Lagrange multiplier approach to demonstrate the need of carrying out such projection.
-Schay\cite{Schay1995, Schay1998} discussed on the equivalence of the two approaches.
+Recently, Nielsen and Krenk\cite{Nielsen2012} employed a Lagrange multiplier approach to demonstrate the need of carrying out such a projection.
+Schay\cite{Schay1995, Schay1998} discussed on the equivalence of these two approaches.
 
@@
 \[
-\left( \diff{U}{\vt q} \right)^\ast = \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast \diff{U}{\vt R_j} = - \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast {\vt F_j},
+\left( \diff{U}{\vt q} \right)^\ast = - \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast {\vt F_j},
 \]
@@
 
+\section{Derivation of the Euler equation}
+\label{eq:derivation_euler}
+
+Let us premultiply Eq.~\ref{eq:euler_eq_quat} by $\frac{1}{2} \tr{\mt B}$, which corresponds to a transformation from the quaternion-fixed reference frame to the body-fixed frame. Hence,
+\[
+\frac{1}{2} \tr{\mt B} \dot{\vt \pi} = \frac{1}{4} \tr{\mt B} \mt \Omega \vt \omega + \vt \tau^\prime,
+\]
+where $\vt \tau^\prime = \tr{\mt B}\mt C \vt \tau = \mt A \vt \tau$ is the body-fixed torque.
+Given that $\mt \Omega = \left[\begin{smallarray}{ccc} \hat{\mt B}_1{\vt \pi} & \hat{\mt B}_2{\vt \pi} & \hat{\mt B}_3{\vt \pi}\end{smallarray}\right]$, by analogy with Eq.~\ref{eq:BtBdot} we conclude that $\tr{\mt B} \mt \Omega = \mt S(\tr{\mt B}\vt \pi) = 2 \mt S(\mt I \vt \omega)$.
+In addition,
+\[
+\tr{\mt B} \dot{\vt \pi} = \frac{d(\tr{\mt B}\vt \pi)}{dt} - \tr{\dot{\mt B}}\vt \pi = 2 \mt I\dot{\vt \omega} - 2 \tr{\dot{\mt B}} \mt B \mt I \vt \omega,
+\]
+where we have inserted $\mt B \tr{\mt B}$ between $\tr{\dot{\mt B}}$ and $\vt \pi$ and substituted $\tr{\mt B} \vt \pi = 2 \mt I \vt \omega$.
+Finally, after transposing Eq.~\ref{eq:relation_B_qdot}, replacing $2 \tr{\mt B}\dot{\vt q} = \vt \omega$, and resorting to the relation between $\mt S(\cdot)$ and a cross product, we obtain
+\[
+\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = {\vt \tau}^\prime.
+\]
+
+This is the Euler equation of motion for rigid bodies.\cite{Goldstein2002}
+
 \bibliography{rigid_bodies}

### 2016-07-23 rigid_bodies.tex
Appendices finished

@@
 
-Hence, the unit vector proportional to $(\partial g/\partial \vt x)^\ast$ is exactly $\vt q$, which means that the constrained derivative of $U$ is given by
+Hence, the unit vector proportional to $(\partial g/\partial \vt x)^\ast$ is exactly $\vt q$.
+As the same constraint makes ${\mt C}\tr{\mt C} = \mt 1_4 - \vt q\tr{\vt q}$ (in analogy to Eq.~\ref{eq:basics_B}), the constrained derivative of $U$ can be represented by
 \begin{equation}
 \label{eq:diff_projection}
-\diff{U}{\vt q} = (\mt 1_4 - {\vt q}\tr{\vt q})\left( \diff{U}{\vt q} \right)^\ast.
+\diff{U}{\vt q} = {\mt C}\tr{\mt C} \left( \diff{U}{\vt q} \right)^\ast.
 \end{equation}
@@
 \[
-\left( \diff{U}{\vt q} \right)^\ast = - \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast {\vt F_j},
-\]
-where $\vt F_j = -\partial U/\partial \vt R_j$ is, by definition, the force applied on particle $j$. As $\vt R_j = \vt r + \tr{\mt A}\vt d_j$,
-\[
-\left( \diff{\vt R_j}{\vt q} \right)^\ast = \left( \diff{(\tr{\mt A}\vt d_j)}{\vt q} \right)^\ast.
-\]
-
-For simplicity, we will drop the index $j$ for the next few lines. Remember that $\tr{\mt A}\vt d = \tr{\mt C} \mt B\vt d$ and, according to Eq.~\ref{eq:product_B_vector},
-\[
-\mt B \vt d = \sum_{k=1}^3 d_k \mt B_k \vt q = \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q.
-\]
-
-An identity equivalent to Eq.~\ref{eq:vector_entries} holds for the product of $\tr{\mt C}$ with an arbitrary quaternion, so that we can write
-\[
-\tr{\mt A}{\vt d} =
-\left[\begin{array}{c}
-\tr{\vt q}\tr{\mt C}_1 \mt B \vt d \\
-\tr{\vt q}\tr{\mt C}_2 \mt B \vt d \\
-\tr{\vt q}\tr{\mt C}_3 \mt B \vt d \\
-\end{array}\right] = 
-\left[\begin{array}{c}
-\tr{\vt q} \tr{\mt C}_1 \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q \\
-\tr{\vt q} \tr{\mt C}_2 \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q \\
-\tr{\vt q} \tr{\mt C}_3 \left( \sum_{k=1}^3 d_k \mt B_k \right) \vt q \\
-\end{array}\right].
-\]
-
-As mentioned in Appendix \ref{sec:auxiliary_math}, $\tr{\mt C}_i \mt B_j = \tr{\mt B}_j \mt C_i$ for any pair $ij$, so that all the constant matrices above between $\tr{\vt q}$ and $\vt q$ are constant and symmetric, which facilitates differentiation.
-After carrying it out and replacing back $(\sum_{k=1}^3 d_k \mt B_k) \vt q$ by $\mt B \vt d$, we have
-\[
-\left( \diff{\vt R}{\vt q} \right)^\ast = 2 \left[\begin{array}{ccc}
-\tr{\mt C}_1 \mt B \vt d & \tr{\mt C}_2 \mt B \vt d & \tr{\mt C}_3 \mt B \vt d
-\end{array}\right]
-\]
-
-Considering that we deal exclusively with unit quaternions $\vt q$, so that $\mt C \tr{\mt C} = \mt 1_4 - \vt q \tr{\vt q}$ can be substituted in Eq.~\ref{eq:diff_projection}, it is useful to evaluate the product
-\[
-\tr{\mt C} \left( \diff{\vt R}{\vt q} \right)^\ast = 2 \left[\begin{array}{c}
-\tr{\vt q}\tr{\mt C_1} \\
-\tr{\vt q}\tr{\mt C_2} \\
-\tr{\vt q}\tr{\mt C_3}
-\end{array}\right] \left[\begin{array}{ccc}
-\tr{\mt C}_1 \mt B \vt d & \tr{\mt C}_2 \mt B \vt d & \tr{\mt C}_3 \mt B \vt d
-\end{array}\right]
-\]
-
-The permutation matrices $\mt C_1$, $\mt C_2$, and $\mt C_3$ have the following properties:
-\begin{equation}
-\begin{aligned}
-\tr{\mt C_1}\tr{\mt C_1} &= \tr{\mt C_2}\tr{\mt C_2} = \tr{\mt C_3}\tr{\mt C_3} = -\mt 1_4 \\
-\tr{\mt C_1}\tr{\mt C_2} &= -\tr{\mt C_2}\tr{\mt C_1} = -\tr{\mt C_3} \\
-\tr{\mt C_2}\tr{\mt C_3} &= -\tr{\mt C_3}\tr{\mt C_2} = -\tr{\mt C_1} \\
-\tr{\mt C_3}\tr{\mt C_1} &= -\tr{\mt C_1}\tr{\mt C_3} = -\tr{\mt C_2}
-\end{aligned}
-\end{equation}
-
-Along with the fact that $\mt B\vt d$ and $\vt q$ are necessarily orthogonal to each other, the properties above lead to
-\begin{align*}
-\tr{\mt C} \left( \diff{\vt R}{\vt q} \right)^\ast &= 2 \left[\begin{array}{ccc}
-0 & -\tr{\vt q}\tr{\mt C}_3 \mt B \vt d & \tr{\vt q}\tr{\mt C}_2 \mt B \vt d \\
-\tr{\vt q}\tr{\mt C}_3 \mt B \vt d & 0 & -\tr{\vt q}\tr{\mt C}_1 \mt B \vt d \\
--\tr{\vt q}\tr{\mt C}_2 \mt B \vt d & \tr{\vt q}\tr{\mt C}_1 \mt B \vt d & 0
-\end{array}\right] \\
-&= 2 \mt S(\tr{\mt C} \mt B \vt d) = 2 \mt S(\tr{\mt A}\vt d) = 2 \mt S(\vt \delta),
-\end{align*}
-where $\vt \delta$ is equivalent to $\vt d$, but expressed in the space-fixed frame of reference. Finally, we obtain
+\left( \diff{U}{\vt q} \right)^\ast = \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast \diff{U}{\vt R_j} = - \sum_{j=1}^{n_p} \left( \diff{\vt \delta_j}{\vt q} \right)^\ast {\vt F_j},
+\]
+where $\vt R_j = \vt r + \vt \delta_j$ and $\vt F_j = -\partial U/\partial \vt R_j$ is, by definition, the force applied on particle $j$.
+Temporarily dropping the particle index for simplicity, we recall from Sec.~\ref{sec:kinematics} that $\vt \delta = \tr{\mt A}\vt d$, where $\vt d$ is a constant vector.
+Because each entry of matrix $\mt A = \tr{\mt B}{\mt C}$ is $A_{ij}=\tr{\vt q} \tr{\hat{\mt B}}_i \hat{\mt C}_j {\vt q}$, an entry of $\vt \delta$ is given by
+\[
+\delta_j = \sum_{i=1}^{3} d_i \tr{\vt q} \tr{\hat{\mt C}}_j \hat{\mt B}_i {\vt q}.
+\]
+
+Because $\tr{\hat{\mt C}}_j \hat{\mt B}_i$ is symmetric, an unconstrained differentiation results in
+\[
+\left(\diff{\delta_j}{\vt q}\right)^\ast = 2 \sum_{i=1}^{3} d_i \tr{\hat{\mt C}}_j \hat{\mt B}_i {\vt q} = -2{\hat{\mt C}}_j {\mt B}{\vt d},
+\]
+where we have factored out $\tr{\hat{\mt C}}_j = -\hat{\mt C}_j$ from the summation and then employed Eq.~\ref{eq:product_B_vector}.
+The constrained differential $\partial \vt \delta/\partial \vt q$ is, therefore,
+\[
+\diff{\vt \delta}{\vt q} = -2{\mt C}\tr{\mt C}
+\left[\begin{smallarray}{ccc}
+{\hat{\mt C}}_1 {\mt B}{\vt d} & {\hat{\mt C}}_2 {\mt B}{\vt d} & {\hat{\mt C}}_3 {\mt B}{\vt d}
+\end{smallarray}\right] = 2{\mt C}\;{\mt S}\!\left(\tr{\mt C}{\mt B}\vt d\right),
+\]
+where we have carried out the second matrix product above, employed the properties presented in Eq.~\ref{eq:CC_products}, and then used the operator defined in Eq.~\ref{eq:operator_S}.
+Finally, after realizing that $\tr{\mt C}{\mt B}\vt d = \tr{\mt A}{\vt d} = \vt \delta$, we can used the result above to obtain
 \[

### 2016-07-27 rigid_bodies.tex
Some small changes

@@
 The formulation of the equations of motion in this frame of reference is notably simplified.
-Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\citep{Goldstein2002}
+Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
 \[
@@
 Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain.
-Appendix~\ref{sec:quat_from_A} describes how one can compute the unit quaternion related to a given rotation matrix.
+Methods are available for computing the unit quaternion related to a given rotation matrix (e.g.
+Ref.~\onlinecite{Shepperd1978}).
 Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
@@
 
-As explained in the mentioned appendix, premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$.
+As explained in the referred appendix, premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$.
 Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.
-Therefore, one can define an intermediary frame of reference, fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body, whose connection to the body-fixed and the space-fixed frames are summarized as follows:
+Therefore, one can define an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body.
+Its connection to the body-fixed and space-fixed frames are summarized as follows:
 \[
 \boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
-\mathrel{\mathop{\rightleftarrows}^{\frac{1}{2}\mt B}_{\frac{1}{2}\tr{\mt B}}}
+\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
 \boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
-\mathrel{\mathop{\rightleftarrows}^{2 \tr{\mt C}}_{2 \mt C}}
+\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
 \boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
@@
 
-Therefore, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply the angular velocity vector expressed in the quaternion-fixed frame.
-It also guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.
+Therefore, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply half the angular velocity vector expressed in the quaternion-fixed frame.
+Besides, it guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.
 
@@
 
-To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the quaternion momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By performing the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
+To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By carrying out the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
 \begin{equation}
@@
 
-Hence, the conjugated momentum $\vt \pi$ is also orthogonal to $\vt q$ (premultiplication of $\mt I \vt \omega$ by $\mt B$).
+Hence, $\vt \pi$ is twice the quaternion-fixed version of the angular momentum ($\mt I \vt \omega$) and, as such, it is also orthogonal to $\vt q$.
 Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
@@
 \label{eq:H_with_B}
-\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi + U(\vt r, \vt q).
+\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi}{8} + U(\vt r, \vt q).
 \end{equation}
@@
 \label{eq:H_with_Omega}
-\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + \frac{1}{8} \tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q + U(\vt r, \vt q).
+\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q}{8} + U(\vt r, \vt q).
 \end{equation}
@@
 where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame.
-Therefore, $-\partial U/\partial \vt q$ is simply the torque represented in the quaternion-fixed frame.
+Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in the quaternion-fixed frame.
 Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
@@
 \[
-\mathcal{H} = \frac{1}{2m} \tr{\vt p} \vt p + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
+\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
 \]
@@
 The exact solution for this case is
-\begin{align*}
-&{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0 \\
-&{\vt p}(t) = {\vt p}_0 \\
-&{\vt q}(t) = {\vt q}_0 \\
-&{\vt \pi}(t) = {\vt \pi}_0
-\end{align*}
-
+\[
+{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0,
+\]
+with ${\vt p}(t) = {\vt p}_0$, ${\vt q}(t) = {\vt q}_0$, and ${\vt \pi}(t) = {\vt \pi}_0$.
 This corresponds to a free translation (hence the subscript $t$).
@@
 \begin{align*}
-&{\vt r}(t) = {\vt r}_0 \\
-&{\vt p}(t) = {\vt p}_0 + t \vt F_0 \\
-&{\vt q}(t) = {\vt q}_0 \\
-&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C_0 \vt \tau_0
+&{\vt p}(t) = {\vt p}_0 + t \vt F \\
+&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C \vt \tau
 \end{align*}
-
-Note that $\vt r$ and $\vt q$ remain unaltered, thus keeping $\vt F$, $\mt C$, and $\vt \tau$ constant as well, which justifies the subscript $0$ assigned to them.
+with ${\vt r}(t) = {\vt r}_0$ and ${\vt q}(t) = {\vt q}_0$.
+As $\vt r$ and $\vt q$ remain unaltered, $\vt F$, $\mt C$, and $\vt \tau$ are constant as well.
 This solution clearly conserves $\mathcal{H}_b$ and marks the effect of a propagator $e^{t i\!L_b}$, obtained from the operator
@@
 As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form.
-The following analytical solution then takes place:\cite{Miller2002}
+The analytical solution that takes place is\cite{Miller2002}
 \begin{align*}
-&{\vt r}(t) = {\vt r}_0 \\
-&{\vt p}(t) = {\vt p}_0 \\
 &{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
@@
 \end{align*}
-
+with ${\vt r}(t) = {\vt r}_0$ and ${\vt p}(t) = {\vt p}_0$.
 These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{t i\!L_k}$ derived from the operator
@@
 
-\section{Obtaining quaternion components from a rotation matrix}
-\label{sec:quat_from_A}
-
-The method described here was introduced by Shepperd.\cite{Shepperd1978} As one can observe in Eq.~\ref{eq:A_from_q}, the three main diagonal components of $\mt A$ and the fact that $\lVert \vt q \lVert^2 = 1$ provide a linear system for the squares of the quaternion components, whose solution is $4 q_0^2 = 1 + \text{tr}\mt A$ and $4 q_k^2 = 1 + A_{k,k} - \text{tr}\mt A$ for $k = 1, 2, 3$, where $\text{tr}\mt A$ is the trace of matrix $\mt A$.
-As both $\vt q$ and $- \vt q$ correspond to the same rotation matrix, the sign of the largest component\cite{Shepperd1978} can be made positive and the others can be obtained from the off-diagonal entries of $\mt A$, which give $4 q_0 q_1 = A_{2,3} - A_{3,2}$, $4 q_0 q_2 = A_{3,1} - A_{1,3}$, $4 q_0 q_3 = A_{1,2} - A_{2,1}$, and $4 q_k q_l = A_{k,l} + A_{l,k}$, where $k,l = 1,2,3$ with $k \neq l$.
-
-\section{Differentiation rules with vectors}
+\section{Differentiation with vectors and matrices}
 \label{sec:Diff_Rules}
 
-Along the paper, the differentiation of scalar functions with respect to vectors are done directly using the rules
-\[
-\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a \quad \text{and} \quad \diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = (\mt A + \tr{\mt A})\vt x,
+Along the paper, we employ the gradient convention (rather than the Jacobian convention) for expressing derivatives with respect to vector variables.
+Hence, very often we differentiate a scalar function using one of the following rules:
+\[
+\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a \quad \text{or} \quad \diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = (\mt A + \tr{\mt A})\vt x,
 \]

### 2016-08-23 rigid_bodies.tex
Resumo em desenvolvimento

@@
-%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
-%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
 \documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
@@
 \usepackage[pdftex]{graphicx}
-%\usepackage{epstopdf}
-%\usepackage{caption}
-%\usepackage{subfig}
-%\usepackage[]{units}
-%\usepackage[]{natbib}
 \usepackage[]{threeparttable}
-%\usepackage{booktabs}
-%\usepackage{multirow}
 
@@
 
-\newenvironment{smallarray}[1]
+\newenvironment{smallarray}[1]                          % small arrays
 {\null\,\vcenter\bgroup\scriptsize
@@
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+Sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or molecular substructures in Molecular Dynamics.
+This strategy eliminates degrees of freedom and admits larger time steps without giving up the atomistic character of a model.
+In this paper, we rely on a factorization of the rotation matrix, expressed in terms of unit quaternion components, to simplify the microcanonical and canonical dynamics of systems containing rigid bodies.
 \end{abstract}
@@
 
-Improvements in molecular dynamics (MD) algorithms, along with advances in software and hardware, have allowed researchers to extend both the time and the length scales attainable in their atomistic simulations.
-This progress has fomented the ambitious attempt to elucidate complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
+Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations.
+This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
 Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.
@@
 However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
-It is expected to be computationally more efficient because the constraints are treated implicitly.
+This is expected to be computationally more efficient because the constraints are treated implicitly.
 In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
 Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
-This approach has been successfully used in pure MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so on.
+This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth.
 It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
 
-In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as corresponding numerical integrators.
-The contribution in the NVE case relies on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'{e}-Hoover chain (NHC) thermostat.
+In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases.
+Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
 This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom.
 We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
-For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and another one that corrects a pressure calculated in terms of interactions among free atoms.
+For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
 
@@
 In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions.
-In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT ensembles, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
+In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
 Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure.
@@
 Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector.
-An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal motion.
+An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
 
@@
 Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.
-Therefore, one can define an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body.
+Therefore, there exists an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body.
 Its connection to the body-fixed and space-fixed frames are summarized as follows:
@@
 
-Therefore, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply half the angular velocity vector expressed in the quaternion-fixed frame.
+Hence, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply half the angular velocity vector expressed in the quaternion-fixed frame.
 Besides, it guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.
@@
 
-To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\citep{Goldstein2002} By carrying out the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
+To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\cite{Goldstein2002} By carrying out the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
 \begin{equation}
@@
 \]
-where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed reference frame.
-Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in the quaternion-fixed frame.
+where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.
+Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in quaternion-fixed frame of reference.
 Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
@@
 Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
-However, no closed-form solution is known, even in the simplest case of free motion (i.e., when $U$ is constant), due to the dependence of $\omega_k$ with $\vt q$ and $\vt \pi$ (Eq.~\ref{eq:omega_entry}) and the consequent coupling of Eqs.~\ref{eq:edo_q} and \ref{eq:edo_pi}.
-Nevertheless, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010}
-
-For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
+Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
 \[
@@
 An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$).
-Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their conjugate momenta.
+Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta.
 Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.
@@
 
-Considering the system described above, with $\vt r$ and $\vt q$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
+Considering the system described above, with $\vt r$ and $\vt q$ now representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
 \begin{equation}
@@
 
-Therefore, a symplectic integrator can be devised using the same splitting scheme as described in Sec.~\ref{sec:symplectic}.
-The only difference is that the action of each propagator (boost, free translation, and free uniaxial rotation) must now occur simultaneously for all rigid bodies in the system.
+Therefore, a symplectic integrator can be devised using the same splitting scheme of Sec.~\ref{sec:symplectic}, with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system.
 If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$.
@@
 
-Kamberaj \textit{et al.}\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble...........
+Kamberaj \textit{et al}.\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble...........
 
@@
 Care must be taken when differentiating the potential energy $U$ with respect to $\vt q$ because the quaternion components are interdependent.
-Schay demonstrated in Ref.~\onlinecite{Schay1995} that, when differentiating a scalar function $f$ with respect to a vector variable $\vt x$, subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by
+Schay has demonstrated that, when a vector variable $\vt x$ is subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by\cite{Schay1995}
 \[

### 2016-08-25 rigid_bodies.tex
Last version without analytical rotation

@@
 \begin{abstract}
-Sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or molecular substructures in Molecular Dynamics.
-This strategy eliminates degrees of freedom and admits larger time steps without giving up the atomistic character of a model.
-In this paper, we rely on a factorization of the rotation matrix, expressed in terms of unit quaternion components, to simplify the microcanonical and canonical dynamics of systems containing rigid bodies.
+In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
+Such a coarse-graining strategy eliminates degrees of freedom and allows larger time steps without abandoning the atomistic character of a model.
+In this paper, we rely on a particular factorization of the rotation matrix in order to express the mechanics of systems containing rigid bodies.
+Then, we present simplified equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics.
+We also develop new expressions for computing the virial pressure of rigid- body systems under periodic boundary conditions.
 \end{abstract}
@@
 Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
-\begin{equation}
+\[
 \label{eq:A_from_q}
@@
 \end{smallarray} \right].
-\end{equation}
+\]
 
@@
 \]
-it follows that $\tr{\mt B}{\vt \pi} = \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
+it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
 \begin{equation}
@@
 Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in quaternion-fixed frame of reference.
-Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
+Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
 \begin{subequations}
@@
 
-Kamberaj \textit{et al}.\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble...........
-
 Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space.
@@
 will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric, that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$.
-We employ such technique to couple a unique Nos\'{e}-Hoover chain\cite{Martyna1992} with $M$ thermostats to a system with $N$ rigid bodies.
+We employ such technique to couple a unique Nos\'{e}-Hoover chain\cite{Martyna1992} (NHC) with $M$ thermostats to a system with $N$ rigid bodies.
+This is considerably simpler than the approach adopted by Kamberaj \textit{et al}.,\cite{Kamberaj2005} who empirically used two thermostat chains independently coupled to the translational and the rotational degrees of freedom.
 For this, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
@@
 \end{equation}
-where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE} and $L$ is a constant to be determined.
+where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $L$ is a constant to be determined, and $Q_j$ is an inertial parameter of thermostat $j$.
+As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_B T t_d^2$ and $Q_j = k_B T t_d^2$ for $j > 1$, where $t_d$ is a characteristic time scale of the thermostatted system.
 The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
@@
 
-Finally, as the last thermostat will just act on the penultimate one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
+Finally, as the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
 Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'{e}-Hoover chain become
@@
 &\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
-&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \\
+&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \label{eq:nhc_p} \\
 &\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
@@
 &\dot{\eta}_j = \alpha_j \\
-&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}
+&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}  \label{eq:nhc_p_eta}
 \end{align}
@@
 
-As recommended in Ref.~\onlinecite{Martyna1992}, the inertial parameters can be calculated from a characteristic time scale $t_d$ of the system by $Q_1 = L k_B T t_d^2$ and, for $j > 1$, $Q_j = k_B T t_d^2$.
-Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
-For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
-By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.
-
-\subsection{Integration of the Canonical Equations of Motion}
-
-As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion.
-First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components.
-In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator}.
-Retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full operator as\cite{Martyna1996}
+The system in Eq.~\ref{eq:nhc_system} has a non-vanishing phase-space compressibility $\kappa = \sum_k \partial \dot{x}_k/\partial x_k$ because some phase-space variables affect their own time-derivatives.
+This is the case for the $3N$ momentum components in Eq.~\ref{eq:nhc_p}, the $4N$ components in Eq.~\ref{eq:nhc_pi}, and the first $M-1$ thermostat momenta in Eq.~\ref{eq:nhc_p_eta} (recalling that $\alpha_{M+1} = 0$).
+Thus, we have
+\[
+\kappa = -7N \alpha_1 - \sum_{j=1}^{M-1} \alpha_{j+1} = -7N \dot{\eta}_1 - \sum_{j=2}^M \dot{\eta}_j.
+\]
+
+Therefore, besides the conserved quantity $H$ of Eq.~\ref{eq:H_nvt}, the non-Hamiltonian system in Eq.~\ref{eq:nhc_system} has an invariant volume form in phase space given by\cite{Tuckerman_1999, Tuckerman2001, Ezra2006}
+\begin{equation}
+\label{eq:volume_form}
+\bar{\omega} = e^{7N \eta_1 + \sum_{j=2}^M \eta_j} d{\vt r} \wedge d{\vt p} \wedge d{\vt q} \wedge d{\vt \pi} \wedge d{\vt \eta} \wedge d{\vt p_\eta}.
+\end{equation}
+
+Analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
+For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
+By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if we make $L = 6N$.
+
+\subsection{Time-Reversible and Measure-Preserving Integration of the Canonical Equations of Motion}
+
+As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} This implies that we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
+In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
+As a consequence, the whole propagator is symplectic and thus preserves the phase-space volume element.
+It is widely recognized that this property is important for numerical stability,\cite{Skeel1997} but it is restricted to Hamiltonian systems.
+For non-Hamiltonian ones, although the phase-space volume is no longer conserved, there exists an invariant measure in phase space, such as that present in the volume form shown in Eq.~\ref{eq:volume_form}.
+In this sense, Ezra\cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which, besides being time-reversible, preserve the measure associated with the original equations of motion.
+
+Here, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian ones.
+In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, from Eq.~\ref{eq:nhc_system}, it follows that
+\begin{equation}
+\label{eq:iL_NHC}
+\begin{split}
+i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
+&+ \sum_{j=1}^{M} \left[\alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}\right].
+\end{split}
+\end{equation}
+
+By retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator as
 \begin{equation}
@@
 
-It follows from Eq.~\ref{eq:nhc_system} that
-\begin{equation}
-\label{eq:iL_NHC}
-\begin{split}
-i\!L_\text{NHC} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
-&+ \sum_{j=1}^M \alpha_j \diff{}{\eta_j} + \sum_{j=1}^M \left[ (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
-\end{split}
-\end{equation}
-
-If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\eta}_j = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values.
+This factorization is similar to that proposed by Martyna \textit{et al}.\cite{Martyna1996} for general situations and differs from the one developed by Kamberaj \textit{et al}.\cite{Kamberaj2005} in the context of rigid bodies.........
+
+\begin{subequations}
+\begin{align}
+i\!L_\text{NHC}^{(0)} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \alpha_1 \diff{}{\eta_1} \\
+i\!L_\text{NHC}^{(j)} &= (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}} \\
+i\!L_\text{NHC}^{(M)} &= G_M \diff{}{p_{\eta_M}}.
+\end{align}
+\end{subequations}
+
+Kamberaj \textit{et al}.\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble.
+For this, they applied independent Nos\'e-Hoover chain\cite{Martyna1992} thermostats for the translational and rotational degrees of freedom.
+In addition,
+
+.............
+
+Here we present a different approach. ...........
+
+If the last summation was absent from Eq.~\ref{eq:iL_NHC}, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\eta}_j = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values.
 This system of equations can be solved analytically, which results in
@@
 \begin{align*}
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = &\prod_{j=M}^1 \exp \left[ \frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right] \\
-\times \exp \Bigg[ -\frac{\Delta t}{2} &\alpha_1 \sum_{i=1}^N \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j} \Bigg] \\
-\times &\prod_{j=1}^M \exp \left[\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].\\
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \exp \left( \frac{\Delta t}{4} G_M \diff{}{p_{\eta_M}} \right) \\
+\prod_{j=M-1}^1 \exp \left\{ \frac{\Delta t}{4} \left[ (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}} \right] \right\} \\
+\exp \Bigg[ -\frac{\Delta t}{2} \alpha_1 \sum_{i=1}^N \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j} \Bigg] \\
+\prod_{j=1}^M \exp \left[\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].\\
 \end{align*}
@@
 
-In the microcanonical case, each term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
-As a consequence, the whole propagator preserves the phase-space volume.
-Recall that the inverse statement is not necessarily true.
-It is now widely recognized that the symplectic property is important for numerical stability,\cite{Skeel1997} although it is restricted to Hamiltonian systems.
-For non-Hamiltonian equations of motion, the phase space volume is no longer conserved.
-Nevertheless, there exists an invariant measure on the phase space, and for the system in Eq.~\ref{eq:nhc_system}, it is given by
-\begin{align*}
-\bar{\omega} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j} \wedge {\vt r} \wedge {\vt p} \wedge {\vt q} \wedge {\vt \pi} \wedge {\vt \eta} \wedge {\vt p_\eta}
-\end{align*}
-
-In this sense, Ezra \cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which are both time-reversible and measure-preserving.
-In his work, he also pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to trajectories that are more ergodic.
-It can be shown that many of the individual operators in both Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:modified_splitting} do not preserve the invariant measure.
+Ezra\cite{Ezra2006} pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to trajectories that are more ergodic.
 Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.

### 2016-08-26 rigid_bodies.tex
Measure-preserving formulation for NHC

@@
 In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
-Such a coarse-graining strategy eliminates degrees of freedom and allows larger time steps without abandoning the atomistic character of a model.
-In this paper, we rely on a particular factorization of the rotation matrix in order to express the mechanics of systems containing rigid bodies.
-Then, we present simplified equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics.
-We also develop new expressions for computing the virial pressure of rigid- body systems under periodic boundary conditions.
+Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model.
+In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules.
+Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics.
+We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions.
+Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
 \end{abstract}
@@
 
-In Appendix \ref{eq:derivation_euler}, it is demonstrated that Eq.~\ref{eq:euler_eq_quat} is a quaternion-fixed frame representation of the well-known Euler equation.\cite{Goldstein2002} Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
+Anticipating a feature that will prove convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
 Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
@@
 where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
-\[
+\begin{equation}
+\label{eq:H_split_omega}
 \mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
-\]
+\end{equation}
 
@@
 
-An exact solution of Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
+An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
 Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
@@
 
-Now note that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
+Now note that the Hamiltonian in Eq.~\ref{eq:H_split_omega} is equal to the sum ${\mathcal H}_t + {\mathcal H}_b + \sum_{k=1}^3 {\mathcal H}_k$, so that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
 \begin{align}
@@
 
-The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}.
+The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion which preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}.
 For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
@@
 We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side.
-Splitting schemes with higher-order approximations have been used for rigid body dynamics,\cite{Omelyan2007, Omelyan2008, vanZon2008} but they result in methods which are more complex and often require expensive force-gradient calculations.
-Moreover, it is unclear whether their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than NVE.
-
-By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ commute, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
+By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ are commutable, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
 \begin{equation}
@@
 
+Finally, we remark that an analytical solution for the complete rotation exists\cite{vanZon2007} and that splitting schemes with higher-order approximations have been used for rigid body dynamics.\cite{Omelyan2007, Omelyan2008, vanZon2008} However, they result in methods that are more complex and often require expensive force-gradient calculations.
+Moreover, it is unclear how their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than the microcanonical one.
+
 \section{Molecular Dynamics}
@@
 
-Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space.
-Sergi and Ferrario\cite{Sergi2001} described a simple way of systematically generating conservative trajectories in an arbitrary phase space.
+Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
+Sergi and Ferrario\cite{Sergi2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.
 If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
@@
 will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric, that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$.
-We employ such technique to couple a unique Nos\'{e}-Hoover chain\cite{Martyna1992} (NHC) with $M$ thermostats to a system with $N$ rigid bodies.
-This is considerably simpler than the approach adopted by Kamberaj \textit{et al}.,\cite{Kamberaj2005} who empirically used two thermostat chains independently coupled to the translational and the rotational degrees of freedom.
-For this, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
+A generalized Liouville operator can then be defined as
+\[
+i\!L = \tr{\dot{\vt x}}\diff{}{\vt x} = \tr{\left(\diff{H}{\vt x}\right)} \tr{\boldsymbol{\mathcal B}} \diff{}{\vt x}.
+\]
+
+According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman2001} a non-Hamiltonian system might have a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative.
+In this case, phase-space volume is no longer conserved, but there is an invariant measure in phase space given by $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
+Using a generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
+\begin{equation}
+\label{eq:relation_kappa_w}
+\kappa = \dot w = i\!L w.
+\end{equation}
+
+We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a unique Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
+This is considerably simpler than the approach adopted by Kamberaj \textit{et al}.,\cite{Kamberaj2005} who empirically used two thermostat chains independently coupled to the translational and rotational degrees of freedom.
+To this end, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
 The conserved ``extended Hamiltonian'' in this case is
@@
 \label{eq:H_nvt}
-H = \mathcal{H} + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j,
-\end{equation}
-where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $L$ is a constant to be determined, and $Q_j$ is an inertial parameter of thermostat $j$.
-As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_B T t_d^2$ and $Q_j = k_B T t_d^2$ for $j > 1$, where $t_d$ is a characteristic time scale of the thermostatted system.
+H = \mathcal{H}(\vt r,\vt p,\vt q,\vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_b T\eta_1 + k_b T\sum_{j=2}^M \eta_j,
+\end{equation}
+where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
+As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostatted system.
 The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
@@
 
-In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the first thermostat.
-In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, plus a term regarding the action of a second thermostat.
+In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat.
+In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, along with a term regarding the action of a second thermostat.
 The result is
@@
 Finally, as the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
-Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'{e}-Hoover chain become
+Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain become
 \begin{subequations}
@@
 &\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
-&\dot{\eta}_j = \alpha_j \\
+&\dot{\eta}_j = \alpha_j \label{eq:nhc_eta} \\
 &{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}  \label{eq:nhc_p_eta}
@@
 \begin{align*}
-&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_B T \quad \text{and}\\
+&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_b T \quad \text{and}\\
 &G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
@@
 
-The system in Eq.~\ref{eq:nhc_system} has a non-vanishing phase-space compressibility $\kappa = \sum_k \partial \dot{x}_k/\partial x_k$ because some phase-space variables affect their own time-derivatives.
-This is the case for the $3N$ momentum components in Eq.~\ref{eq:nhc_p}, the $4N$ components in Eq.~\ref{eq:nhc_pi}, and the first $M-1$ thermostat momenta in Eq.~\ref{eq:nhc_p_eta} (recalling that $\alpha_{M+1} = 0$).
-Thus, we have
-\[
-\kappa = -7N \alpha_1 - \sum_{j=1}^{M-1} \alpha_{j+1} = -7N \dot{\eta}_1 - \sum_{j=2}^M \dot{\eta}_j.
-\]
-
-Therefore, besides the conserved quantity $H$ of Eq.~\ref{eq:H_nvt}, the non-Hamiltonian system in Eq.~\ref{eq:nhc_system} has an invariant volume form in phase space given by\cite{Tuckerman_1999, Tuckerman2001, Ezra2006}
-\begin{equation}
-\label{eq:volume_form}
-\bar{\omega} = e^{7N \eta_1 + \sum_{j=2}^M \eta_j} d{\vt r} \wedge d{\vt p} \wedge d{\vt q} \wedge d{\vt \pi} \wedge d{\vt \eta} \wedge d{\vt p_\eta}.
-\end{equation}
-
-Analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
+Note that the $3N$ momentum components in Eq.~\ref{eq:nhc_p}, the $4N$ components in Eq.~\ref{eq:nhc_pi}, and the thermostat momenta in Eq.~\ref{eq:nhc_p_eta} affect their own time-derivatives.
+Thus, the thermostatted system has a phase-space compressibility $\kappa = -7N \alpha_1 - \sum_j \alpha_{j+1}$.
+Resorting to Eq.~\ref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
+\begin{equation}
+\label{eq:nhc_measure}
+w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
+\end{equation}
+
+Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
 For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
@@
 
-\subsection{Time-Reversible and Measure-Preserving Integration of the Canonical Equations of Motion}
-
-As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} This implies that we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
+\subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}
+
+As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
 In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
-As a consequence, the whole propagator is symplectic and thus preserves the phase-space volume element.
-It is widely recognized that this property is important for numerical stability,\cite{Skeel1997} but it is restricted to Hamiltonian systems.
-For non-Hamiltonian ones, although the phase-space volume is no longer conserved, there exists an invariant measure in phase space, such as that present in the volume form shown in Eq.~\ref{eq:volume_form}.
-In this sense, Ezra\cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which, besides being time-reversible, preserve the measure associated with the original equations of motion.
-
-Here, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian ones.
-In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, from Eq.~\ref{eq:nhc_system}, it follows that
+As a result, the whole propagator is symplectic and thus preserves the phase-space volume element.
+It is widely recognized that this property is important for numerical stability.\cite{Skeel1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion.
+In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator does it as well.
+As the author demonstrated,\cite{Ezra2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
+Here, $\dot{\vt x}^{(k)}$ refers to the equations of motion obtained from the individual operator $i\!L^{(k)}$, which might display their own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$.
+By means of the product rule, it is straightforward to devise an equivalent but clearer criterion, which is
+\begin{equation}
+\label{eq:split_kappa_w}
+\kappa^{(k)} = \sum_j \diff{\dot{x}_j^{(k)}}{x_j} = i\!L^{(k)} w.
+\end{equation}
+
+Therefore, if the function $w(\vt x)$ that satisfies Eq.~\ref{eq:relation_kappa_w} also satisfies the equation above for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$, in accordance with Ref.~\onlinecite{Ezra2006}.
+We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj2005} does not fulfill this criterion.
+In the present paper, we propose and compare two distinct measure-preserving splittings for a rigid-body system coupled to a Nos\'e-Hoover chain thermostat.
+In the first one, we perform the usual separation of the non-Hamiltonian components of the equations of motion (those involving thermostat variables) from the Hamiltonian ones.
+In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, as it follows from Eq.~\ref{eq:nhc_system},
 \begin{equation}
@@
 
-By retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator as
+Then, by retaining the NVE factorization defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator as
 \begin{equation}
@@
 
-This factorization is similar to that proposed by Martyna \textit{et al}.\cite{Martyna1996} for general situations and differs from the one developed by Kamberaj \textit{et al}.\cite{Kamberaj2005} in the context of rigid bodies.........
-
+In order to satisfy the criterion in Eq.~\ref{eq:split_kappa_w} for all factors, given the function $w$ expressed in Eq.~\ref{eq:nhc_measure}, we further split the NHC contribution as $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_\text{NHC}^{(j)}$, where
 \begin{subequations}
 \begin{align}
-i\!L_\text{NHC}^{(0)} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \alpha_1 \diff{}{\eta_1} \\
-i\!L_\text{NHC}^{(j)} &= (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}} \\
-i\!L_\text{NHC}^{(M)} &= G_M \diff{}{p_{\eta_M}}.
+i\!L_\text{NHC}^{(0)} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \alpha_1 \diff{}{\eta_1}, \\
+i\!L_\text{NHC}^{(j)} &= (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}}, \label{eq:iL_NHC_j} \\
+i\!L_\text{NHC}^{(M)} &= G_M \diff{}{p_{\eta_M}},
 \end{align}
 \end{subequations}
-
-Kamberaj \textit{et al}.\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble.
-For this, they applied independent Nos\'e-Hoover chain\cite{Martyna1992} thermostats for the translational and rotational degrees of freedom.
-In addition,
-
-.............
-
-Here we present a different approach. ...........
-
-If the last summation was absent from Eq.~\ref{eq:iL_NHC}, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\eta}_j = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values.
-This system of equations can be solved analytically, which results in
-\begin{subequations}
-\begin{align}
-&\vt p_i(t) = \vt p_i^0 e^{-\alpha_1 t}, \\
-&\vt \pi_i(t) = \vt \pi_i^0 e^{-\alpha_1 t}, \quad \text{and}\\
-&\eta_j(t) = \eta_j^0 + \alpha_j t. \label{eq:solution_eta}
-\end{align}
-\end{subequations}
-
-Otherwise, if a single term $(G_j - \alpha_{j+1} p_{\eta_j}) \partial/\partial p_{\eta_j}$ was the only term present in Eq.~\ref{eq:iL_NHC}, then $e^{t i\!L_\text{NHC}}$ would behave according to the equation $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$, with all variables other than $p_{\eta_j}$ kept unchanged.
-The analytical solution of this single equation is
-\begin{equation}
-\label{eq:solution_p_eta}
-\begin{split}
-&p_{\eta_j}(t) = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1} p_{\eta_j}^0 \right) \phi\left(\alpha_{j+1} t\right) t, \\
-\end{split}
-\end{equation}
+with $j$ varying from $1$ to $M-1$ in Eq.~\ref{eq:iL_NHC_j}. Then, each propagator $e^{\frac{\Delta t}{2} i\!L_\text{NHC}}$ present in Eq.~\ref{eq:iL_NHC} is factorized as
+\[
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \Bigg[ \prod_{j=M}^1 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
+\]
+
+The action of the propagator $e^{t i\!L_\text{NHC}^{(0)}}$ is the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\eta}_1 = \alpha_1$, with all other variables remaining constant at their initial values.
+This system can be solved analytically, which results in
+\begin{align*}
+&\vt p_i(t) = \vt p_i^0 e^{-\alpha_1 t} \\
+&\vt \pi_i(t) = \vt \pi_i^0 e^{-\alpha_1 t} \\
+&\eta_1(t) = \eta_1^0 + \alpha_1 t
+\end{align*}
+
+In turn, the propagator $e^{t i\!L_\text{NHC}^{(j)}}$ defined in Eq.~\ref{eq:iL_NHC_j} behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged, whose analytical solution is
+\begin{align*}
+&p_{\eta_j}(t) = p_{\eta_j}^0 + \Big( G_j - \alpha_{j+1} p_{\eta_j}^0 \Big) \phi\left(\alpha_{j+1} t\right) t \\
+&\eta_{j+1}(t) = \eta_{j+1}^0 + \alpha_{j+1} t
+\end{align*}
 where $\phi(x) = (1-e^{-x})/x$.
@@
 
-Considering the analytical solutions described above, it is possible to split the NHC operator into $2M+1$ factors by applying the Trotter-Suzuki formula in a recursive way.
-The splitting we employ here, noting that the product signs run in reverse orders, is
+Finally, the propagator $e^{t i\!L_\text{NHC}^{(M)}}$ involves a unique differential equation $\dot{p}_{\eta_M} = G_M$, whose solution is
+\[
+p_{\eta_M}(t) = p_{\eta_M}^0 + G_M t.
+\]
+
+------------
+
+The second splitting scheme we propose here relies on the definition of the operator
 \begin{align*}
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \exp \left( \frac{\Delta t}{4} G_M \diff{}{p_{\eta_M}} \right) \\
-\prod_{j=M-1}^1 \exp \left\{ \frac{\Delta t}{4} \left[ (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}} \right] \right\} \\
-\exp \Bigg[ -\frac{\Delta t}{2} \alpha_1 \sum_{i=1}^N \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j} \Bigg] \\
-\prod_{j=1}^M \exp \left[\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].\\
+i\!L_b^\ast &= \sum_{i=1}^N \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} \\
+&+ \sum_{i=1}^N \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} + \alpha_1 \diff{}{\eta_1}.
 \end{align*}
@@
 These equations are equivalent in form to the one solved above, so that their solutions are
+\begin{align*}
+&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
+&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
+&\eta_1(t) = \eta_1^0 + \alpha_1 t
+\end{align*}
+
 \begin{subequations}
@@
 
+In the new propagator, $e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} }$ is removed:
+\[
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast} = \Bigg[ \prod_{j=M}^2 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(1)} } \Bigg[ \prod_{j=2}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
+\]
+
+---------
+
 Of course, Eqs.~\ref{eq:solution_eta} and \ref{eq:solution_p_eta} are the ones used to evaluate the propagators above. Finally, the complete scheme is
@@
 
-Ezra\cite{Ezra2006} pointed out that the use of an accurate measure-preserving integrator does not necessarily lead to trajectories that are more ergodic.
  Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.
@@
 
-\section{Derivation of the Euler equation}
-\label{eq:derivation_euler}
-
-Let us premultiply Eq.~\ref{eq:euler_eq_quat} by $\frac{1}{2} \tr{\mt B}$, which corresponds to a transformation from the quaternion-fixed reference frame to the body-fixed frame. Hence,
-\[
-\frac{1}{2} \tr{\mt B} \dot{\vt \pi} = \frac{1}{4} \tr{\mt B} \mt \Omega \vt \omega + \vt \tau^\prime,
-\]
-where $\vt \tau^\prime = \tr{\mt B}\mt C \vt \tau = \mt A \vt \tau$ is the body-fixed torque.
-Given that $\mt \Omega = \left[\begin{smallarray}{ccc} \hat{\mt B}_1{\vt \pi} & \hat{\mt B}_2{\vt \pi} & \hat{\mt B}_3{\vt \pi}\end{smallarray}\right]$, by analogy with Eq.~\ref{eq:BtBdot} we conclude that $\tr{\mt B} \mt \Omega = \mt S(\tr{\mt B}\vt \pi) = 2 \mt S(\mt I \vt \omega)$.
-In addition,
-\[
-\tr{\mt B} \dot{\vt \pi} = \frac{d(\tr{\mt B}\vt \pi)}{dt} - \tr{\dot{\mt B}}\vt \pi = 2 \mt I\dot{\vt \omega} - 2 \tr{\dot{\mt B}} \mt B \mt I \vt \omega,
-\]
-where we have inserted $\mt B \tr{\mt B}$ between $\tr{\dot{\mt B}}$ and $\vt \pi$ and substituted $\tr{\mt B} \vt \pi = 2 \mt I \vt \omega$.
-Finally, after transposing Eq.~\ref{eq:relation_B_qdot}, replacing $2 \tr{\mt B}\dot{\vt q} = \vt \omega$, and resorting to the relation between $\mt S(\cdot)$ and a cross product, we obtain
-\[
-\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = {\vt \tau}^\prime.
-\]
-
-This is the Euler equation of motion for rigid bodies.\cite{Goldstein2002}
-
 \bibliography{rigid_bodies}

### 2016-09-20 rigid_bodies.tex
Analytical solution - initial inclusion

@@
 
-For every non-collinear rigid body, there exists a particular frame of reference in which its inertia tensor is diagonal.
+For every non-collinear rigid body (for conciseness, the only case considered here), there exists a particular frame of reference in which its inertia tensor is diagonal.
 The formulation of the equations of motion in this frame of reference is notably simplified.
@@
 \]
-where $I_1$, $I_2$, and $I_3$ (the principal moments of inertia) are the eigenvalues of ${\mt I}^\ast$ and the columns of $\tr{\mt A}$ (the principal axes of rotation) are their corresponding eigenvectors.
-Algorithms for calculating $\mt A$ from $\mt I^\ast$ can be found in the literature.\cite{Kopp2008} Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$.
-Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents an identity matrix in $\mathbb{R}^{k \times k}$.
+where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation.
+With no loss of generality, we can consider that $I_1 > I_2 > I_3$.
+Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g.
+Ref.~\onlinecite{Kopp2008}).
+Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$.
+Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix.
 By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric.
@@
 
-Anticipating a feature that will prove convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
+Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
 Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
@@
 \end{equation}
-where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
+where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body can be rewritten as
 \begin{equation}
@@
 
-In order to rewrite Eq.~\ref{eq:EDO_system}, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to a new form for the equations of motion:
+Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eq.~\ref{eq:EDO_system}:
 \begin{subequations}
@@
 \end{subequations}
+
+\subsection{Analytical Solution for Free Rotation}
+\label{sec:free_rotation}
+
+If no forces act on a rigid body, then the rotational energy $K_r$ and the angular momentum norm $L$ are integrals of motion, so that we can write
+\[
+\begin{cases}
+I_1 \omega_1^2 + I_2 \omega_2^2 + I_3 \omega_3^2 = 2 K_r \\
+I_1^2 \omega_1^2 + I_2^2 \omega_2^2 + I_3^2 \omega_3^2 = L^2
+\end{cases}
+\]
+
+Evidently, one can promptly evaluate $K_r$ and $L$ if the initial angular velocity vector ${\vt \omega}_0 = [\omega_{1,0}]$ is known.
+By solving the equations above for $\omega_1^2$ and $\omega_3^2$, we obtain
+\begin{subequations}
+\begin{align}
+\omega_1^2 = \frac{L^2-2I_3K_r}{I_1(I_1-I_3)}\left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right] \label{eq:solution_omega_1} \\
+\omega_3^2 = \frac{2I_1K_r-L^2}{I_3(I_1-I_3)}\left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right] \label{eq:solution_omega_3}
+\end{align}
+\end{subequations}
+where
+\begin{equation}
+\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \quad \text{and} \quad \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
+\end{equation}
+
+By realizing that the prefactor in Eqs.~\ref{eq:solution_omega_1} and \ref{eq:solution_omega_3} are constant and that these equations are also valid at $t = 0$, we can rewrite then as
+\begin{subequations}
+	\begin{align}
+	\omega_1^2 = A_1^2 \left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right] \label{eq:solution_omega_1} \\
+	\omega_3^2 = A_2^2 \left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right] \label{eq:solution_omega_3}
+	\end{align}
+\end{subequations}
+where
+\begin{equation}
+A_1 = \frac{\lambda_1 \omega_{1,0}}{\sqrt{\lambda_1^2 - \omega_{2,0}^2}} \quad \text{and} \quad
+A_3 = \frac{\lambda_3 \omega_{3,0}}{\sqrt{\lambda_3^2 - \omega_{2,0}^2}}.
+\end{equation}
+
+Substituting in the Euler equation for $\omega_2$, we have
+\[
+I_2 \dot{\omega}_2 - (I_1-I_3) A_1 A_3 \sqrt{\left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right]\left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right]}
+\]
+
+By defining $x = \omega_2/\lambda_\text{min}$ and $\kappa = \lambda_\text{min}/\lambda_\text{max}$, where $\lambda_\text{min} = \min(\lambda_1,\lambda_3)$ and $\lambda_\text{max} = \max(\lambda_1,\lambda_3)$, we obtain the following initial value problem for $x$:
+\[
+\begin{cases}
+I_2 \lambda_\text{min} \dot{x} - (I_1-I_3) A_1 A_3 \sqrt{(1-x^2)(1-\kappa^2x^2)} = 0 \\
+x(0) = \omega_{2,0}/\lambda_\text{min}.
+\end{cases}
+\]
+
+The solution of the equation above can be expressed using the elliptic Jacobi function $\text{sn}$, so that we can calculate $\omega_2$ at any time as
+\[
+\omega_2(t) = \lambda_\text{min} \text{sn}(\alpha t + \epsilon,\kappa),
+\]
+where
+\[
+\alpha = -\frac{(I_1-I_3)A_1A_3}{\lambda_\text{min} I_2} \quad \text{and} \quad \epsilon = \text{arcsn}\left(\frac{\omega_{2,0}}{\lambda_\text{min}},\kappa\right).
+\]
 
@@
 
-Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
+Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
 Sergi and Ferrario\cite{Sergi2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.
@@
 \end{equation}
-will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric, that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$.
-A generalized Liouville operator can then be defined as
+will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric (that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$).
+A generalized Liouville operator can then be defined as\cite{Sergi2004}
 \[
@@
 
-According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman2001} a non-Hamiltonian system might have a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative.
-In this case, phase-space volume is no longer conserved, but there is an invariant measure in phase space given by $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
-Using a generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
+The system might exhibit a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative.
+According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman2001} although the phase-space volume element $d\vt x$ is no longer conserved, there exits an invariant measure $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
+Using the generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
 \begin{equation}
@@
 
-We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a unique Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
-This is considerably simpler than the approach adopted by Kamberaj \textit{et al}.,\cite{Kamberaj2005} who empirically used two thermostat chains independently coupled to the translational and rotational degrees of freedom.
-To this end, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
+We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a single Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
+This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom.
+To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
 The conserved ``extended Hamiltonian'' in this case is
@@
 \end{equation}
-where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
-As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostatted system.
+where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$.
+As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
 The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
@@
 Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
-For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
-By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if we make $L = 6N$.
+For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
+In the general case, by taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
+In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna1994}
 
@@
 It is widely recognized that this property is important for numerical stability.\cite{Skeel1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion.
-In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator does it as well.
-As the author demonstrated,\cite{Ezra2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
-Here, $\dot{\vt x}^{(k)}$ refers to the equations of motion obtained from the individual operator $i\!L^{(k)}$, which might display their own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$.
-By means of the product rule, it is straightforward to devise an equivalent but clearer criterion, which is
+In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator preserves the same measure.
+As the author has shown,\cite{Ezra2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
+Here, $\dot{\vt x}^{(k)}$ refers to the system of equations obtained from the individual operator $i\!L^{(k)}$, which might display its own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$.
+By means of the product rule, it is straightforward to devise an equivalent but more meaningful criterion, which is
 \begin{equation}
 \label{eq:split_kappa_w}
-\kappa^{(k)} = \sum_j \diff{\dot{x}_j^{(k)}}{x_j} = i\!L^{(k)} w.
-\end{equation}
-
-Therefore, if the function $w(\vt x)$ that satisfies Eq.~\ref{eq:relation_kappa_w} also satisfies the equation above for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$, in accordance with Ref.~\onlinecite{Ezra2006}.
-We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj2005} does not fulfill this criterion.
-In the present paper, we propose and compare two distinct measure-preserving splittings for a rigid-body system coupled to a Nos\'e-Hoover chain thermostat.
-In the first one, we perform the usual separation of the non-Hamiltonian components of the equations of motion (those involving thermostat variables) from the Hamiltonian ones.
-In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, as it follows from Eq.~\ref{eq:nhc_system},
+\kappa^{(k)} = i\!L^{(k)} w.
+\end{equation}
+
+Therefore, if the function $w(\vt x)$ that derives from Eq.~\ref{eq:relation_kappa_w} also satisfies Eq.~\ref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$.
+Here, we propose and compare two distinct measure-preserving integrators for a rigid-body system coupled to a single thermostat chain.
+In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, as it follows from Eq.~\ref{eq:nhc_system},
 \begin{equation}
@@
 
-Then, by retaining the NVE factorization defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator as
+Then, by retaining the NVE factorization presented in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator for each time step as
 \begin{equation}
@@
 \end{equation}
+
+The Nos\'e-Hoover chain propagator is further split by making $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_\text{NHC}^{(j)}$ and, afterwards,
+\[
+e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \Pi_{j=1}^n \Bigg[ \prod_{j=M}^1 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
+\]
+
+The boost, translation, and rotation propagators already satisfy the criterion in Eq.~\ref{eq:split_kappa_w} with the function $w$ expressed in Eq.~\ref{eq:nhc_measure} (they are all incompressible and keep $w$ unchanged).
+The NHC propagator is further split into factors that also satisfy Eq.~\ref{eq:split_kappa_w}.
 
@@
 ---------
+We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj2005} does not fulfill this criterion.
 
@@
 	\end{threeparttable}
+\end{table}
+
+\subsection{New Results}
+
+\begin{table*}
+\caption{Results.....}
+\label{table:NVE}
+\begin{ruledtabular}
+\begin{tabular}{cccccccc}
+$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+\hline
+1.0 & $305.95\pm0.03$ & $-8395.4\pm0.1$ & $1646.1\pm0.1$ & $823.3\pm0.1$ & $822.9\pm0.1$ & $151\pm2$ & $0.0471$ \\
+2.0 & $305.17\pm0.03$ & $-8389.0\pm0.1$ & $1641.9\pm0.1$ & $823.2\pm0.1$ & $818.8\pm0.1$ & $148\pm2$ & $0.853$ \\
+3.0 & $304.34\pm0.03$ & $-8374\pm2$ & $1637.5\pm0.2$ & $824.0\pm0.1$ & $813.4\pm0.1$ & $152\pm2$ & $3.96$ \\
+4.0 & $303.6\pm0.4$ & $-8349\pm5$ & $1633\pm2$ & $826.8\pm0.2$ & $806.5\pm0.4$ & $168\pm2$ & $8.25$ \\
+\end{tabular}
+\end{ruledtabular}
+\end{table*}
+
+\begin{table*}
+\caption{Results.....}
+\label{table:proposed_method}
+\begin{ruledtabular}
+\begin{tabular}{cccccccc}
+$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+
+\hline
+\multicolumn{8}{c}{Proposed Method} \\
+1.0 & $300.01\pm0.02$ & $-8465.1\pm0.5$ & $1614.2\pm0.1$ & $807.4\pm0.1$ & $806.8\pm0.1$ & $62\pm2$ & $0.0328$ \\
+2.0 & $300.01\pm0.03$ & $-8450.8\pm0.4$ & $1614.2\pm0.2$ & $809.4\pm0.1$ & $804.7\pm0.1$ & $73\pm2$ & $0.597$ \\
+3.0 & $299.97\pm0.04$ & $-8425.7\pm0.5$ & $1614.0\pm0.2$ & $812.8\pm0.1$ & $801.2\pm0.2$ & $90\pm2$ & $3.44$ \\
+4.0 & $299.93\pm0.04$ & $-8391.4\pm0.6$ & $1613.7\pm0.2$ & $817.1\pm0.2$ & $796.6\pm0.2$ & $119\pm2$ & $5.64$ \\
+
+\hline
+\multicolumn{8}{c}{Alternative Method} \\
+1.0 & $300.00\pm0.02$ & $-8465.5\pm0.5$ & $1614.1\pm0.1$ & $807.1\pm0.1$ & $807.0\pm0.1$ & $65\pm2$ & $0.108$ \\
+2.0 & $299.99\pm0.03$ & $-8450.5\pm0.5$ & $1614.1\pm0.2$ & $809.3\pm0.1$ & $804.7\pm0.1$ & $72\pm2$ & $1.31$ \\
+3.0 & $299.98\pm0.04$ & $-8425.9\pm0.5$ & $1614.0\pm0.2$ & $812.5\pm0.1$ & $801.5\pm0.2$ & $88\pm2$ & $2.53$ \\
+4.0 & $300.01\pm0.04$ & $-8390.4\pm0.6$ & $1614.2\pm0.2$ & $817.4\pm0.2$ & $796.8\pm0.2$ & $120\pm2$ & $7.31$ \\
+
+\hline
+\multicolumn{8}{c}{Method of Kamberaj \textit{et al}.\cite{Kamberaj2005}} \\
+1.0 & $299.53\pm0.02$ & $-8470.9\pm0.5$ & $1611.6\pm0.1$ & $806.4\pm0.1$ & $805.2\pm0.1$ & $52\pm2$ & $0.023$ \\
+2.0 & $298.24\pm0.03$ & $-8471.2\pm0.5$ & $1604.6\pm0.2$ & $805.4\pm0.1$ & $799.2\pm0.1$ & $47\pm2$ & $17.0$ \\
+3.0 & $295.94\pm0.04$ & $-8480.9\pm0.6$ & $1592.3\pm0.2$ & $804.2\pm0.1$ & $788.1\pm0.1$ & $28\pm2$ & $53.2$ \\
+4.0 & $292.72\pm0.04$ & $-8494.7\pm0.7$ & $1575.0\pm0.2$ & $802.4\pm0.1$ & $772.5\pm0.2$ & $-1\pm2$ & $60.6$ \\
+
+\hline
+\multicolumn{8}{c}{Double-Thermostat Analogue of the Proposed Method} \\
+1.0 & $300.00\pm0.02$ & $-8466.0\pm0.5$ & $1614.1\pm0.1$ & $806.6\pm0.1$ & $807.5\pm0.1$ & $60\pm2$ & $6.42$\\
+2.0 & $299.98\pm0.03$ & $-8452.0\pm0.5$ & $1614.0\pm0.2$ & $807.1\pm0.1$ & $806.9\pm0.1$ & $71\pm2$ & $80.4$ \\
+3.0 & $300.16\pm0.04$ & $-8426.1\pm0.5$ & $1615.0\pm0.2$ & $808.4\pm0.1$ & $806.6\pm0.1$ & $90\pm2$ & $1.38\times10^4$ \\
+4.0 & $300.75\pm0.04$ & $-8385.3\pm0.6$ & $1618.2\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $125\pm2$ & $4.29\times10^4$ \\
+
+\hline
+\multicolumn{8}{c}{Double-Thermostat Analogue of the Alternative Method} \\
+1.0 & $299.98\pm0.02$ & $-8465.6\pm0.5$ & $1614.0\pm0.1$ & $806.6\pm0.1$ & $807.4\pm0.1$ & $56\pm2$ & $4.87$ \\
+2.0 & $299.96\pm0.03$ & $-8452.0\pm0.5$ & $1613.9\pm0.2$ & $807.1\pm0.1$ & $806.8\pm0.1$ & $71\pm2$ & $85.1$ \\
+3.0 & $300.20\pm0.04$ & $-8426.0\pm0.5$ & $1615.2\pm0.2$ & $808.5\pm0.1$ & $806.7\pm0.1$ & $89\pm2$ & $1.50\times10^4$ \\
+4.0 & $300.75\pm0.04$ & $-8384.7\pm0.6$ & $1618.1\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $123\pm2$ & $4.33\times10^4$ \\
+
+\hline
+\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of Kamberaj \textit{et al}.\cite{Kamberaj2005}} \\
+1.0 & $299.52\pm0.02$ & $-8470.2\pm0.5$ & $1611.6\pm0.1$ & $805.9\pm0.1$ & $805.6\pm0.1$ & $58\pm2$ & $0.272$ \\
+2.0 & $298.18\pm0.03$ & $-8471.8\pm0.5$ & $1604.3\pm0.2$ & $804.4\pm0.1$ & $799.9\pm0.1$ & $44\pm2$ & $0.390$ \\
+3.0 & $295.97\pm0.04$ & $-8474.6\pm0.5$ & $1592.4\pm0.2$ & $801.6\pm0.1$ & $790.8\pm0.2$ & $29\pm2$ & $3.00$ \\
+4.0 & $292.65\pm0.04$ & $-8478.7\pm0.6$ & $1574.6\pm0.2$ & $797.6\pm0.2$ & $777.0\pm0.2$ & $10\pm2$ & $8.12$ \\
+
+\end{tabular}
+\end{ruledtabular}
+\end{table*}
+
+\begin{table}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
+	\label{table:new_results}
+	\begin{ruledtabular}
+		\begin{tabular}{cccc}
+			& \multicolumn{3}{c}{ Temperature (K) } \\
+			\cline{2-4}
+			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			\hline
+			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
+			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
+			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
+		\end{tabular}
+	\end{ruledtabular}
+\end{table}
+
+\begin{table}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
+	\label{table:new_results_2}
+	\begin{ruledtabular}
+		\begin{tabular}{cccc}
+			& \multicolumn{3}{c}{Drift rate [(kcal/mol)/ns]} \\
+			\cline{2-4}
+			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			\hline
+			1.0 & $0.0328$ & $0.108$ & $0.023$ \\
+			2.0 & $0.597$  & $1.31$  & $17.0$ \\
+			3.0 & $3.44$   & $2.53$  & $53.2$
+		\end{tabular}
+	\end{ruledtabular}
 \end{table}

### 2016-09-25 rigid_bodies.tex
Explicação de rotação combinada

@@
 An arbitrary rotation can be described using three independent parameters.
-One can use, for instance, the so-called Euler angles, but mathematical singularities arise when the equations of motion are described in such terms.
-Instead, we consider an alternative representation, in which the orientation of a rigid body is given by the Euler parameters $q_0$, $q_1$, $q_2$, and $q_3$, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_0 & q_1 & q_2 & q_3 \end{array}]}$.
-The mathematical relation between the Euler angles and the Euler parameters requires that\cite{Goldstein2002}
+One can use Euler angles, for instance, but mathematical singularities arise when the equations of motion are described in such terms.
+Instead, we consider an alternative representation, in which the orientation of a rigid body is given by four parameters $q_1$, $q_2$, $q_3$, and $q_4$, often referred to as Euler parameters, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ and satisfy the relation\cite{Goldstein2002}
 \begin{equation}
 \label{eq:norm_eq_1}
-\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_0^2 + q_1^2 + q_2^2 + q_3^2 = 1.
+\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1.
 \end{equation}
@@
 \label{eq:A_from_q}
-\mt A = \left[ \begin{smallarray}{ccc}
-q_0^2 + q_1^2 - q_2^2 - q_3^2 & 2(q_1 q_2 + q_0 q_3) & 2(q_1 q_3 - q_0 q_2) \\
-2(q_1 q_2 - q_0 q_3) & q_0^2 - q_1^2 + q_2^2 - q_3^2 & 2(q_2 q_3 + q_0 q_1) \\
-2(q_1 q_3 + q_0 q_2) & 2(q_2 q_3 - q_0 q_1) & q_0^2 - q_1^2 - q_2^2 + q_3^2  
+\mt A(\vt q) = \left[ \begin{smallarray}{ccc}
+q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
+2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
+2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
 \end{smallarray} \right].
@@
 
-Let us now introduce two new matrices that also depend on the quaternion components, which are
+Let us now introduce two new matrix-valued functions that depend on the quaternion components as
 \begin{equation}
 \label{eq:def_B_and_C}
-\mt B = \left[
+\mt B(\vt q) = \left[
 \begin{smallarray}{rrrr}
--q_1 & -q_2 & -q_3 \\
- q_0 & -q_3 &  q_2 \\
- q_3 &  q_0 & -q_1 \\
--q_2 &  q_1 &  q_0
+-q_2 & -q_3 & -q_4 \\
+ q_1 & -q_4 &  q_3 \\
+ q_4 &  q_1 & -q_2 \\
+-q_3 &  q_2 &  q_1
 \end{smallarray}
@@
 \quad \text{and} \quad
-\mt C = \left[
+\mt C(\vt q) = \left[
 \begin{smallarray}{rrrr}
--q_1 & -q_2 & -q_3 \\
- q_0 &  q_3 & -q_2 \\
--q_3 &  q_0 &  q_1 \\
- q_2 & -q_1 &  q_0
+-q_2 & -q_3 & -q_4 \\
+ q_1 &  q_4 & -q_3 \\
+-q_4 &  q_1 &  q_2 \\
+ q_3 & -q_2 &  q_1
 \end{smallarray}
@@
 
-As explained in the referred appendix, premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$.
-Both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.
+Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.
+It is explained in Appendix~\ref{sec:auxiliary_math} that premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$ and that both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.
 Therefore, there exists an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body.
@@
 
-Hence, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is simply half the angular velocity vector expressed in the quaternion-fixed frame.
+Hence, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is half the quaternion-fixed version of the angular velocity vector.
 Besides, it guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.
+
+To conclude this section, we discuss on the possibility of switching between alternative space-fixed frames when describing the motion of a rigid body.
+This will be useful when we deal with torque-free rotations in Sec.~\ref{sec:free_rotation}.
+Consider that a body-fixed vector $\vt d$ is, at the same instant $t$, observed as $\vt \delta = \tr{[\mt A(\vt q)]}{\vt d}$ in a space-fixed frame $\mathcal S_q$ and as $\vt \eta = \tr{[\mt A(\vt z)]}{\vt d}$ in another space-fixed frame $\mathcal S_z$.
+It thus follows that $\vt \eta = \tr{[\mt A(\vt z)]}{\mt A}(\vt q)\vt \delta$, which means that these two reference frames are rotated with respect to each other by a composite matrix $\mt A(\vt g) = \tr{[\mt A(\vt z)]}{\mt A}(\vt q)$.
+A simple relation between the unit quaternions $\vt g$, $\vt z$, and $\vt q$ is a key result in quaternion algebra.\cite{Dichmann1999} Adapted to our matrix/vector notation, such relation is
+\begin{subequations}
+\begin{equation}
+ \label{eq:comb_rot}
+ \vt g = \tr{\left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right]} \vt q.
+\end{equation}
+
+The block matrix above comprises an orthogonal transformation in $\mathbb{R}^4$, which means that the inverse transformation is simply given by
+\begin{equation}
+\label{eq:inv_comb_rot}
+\vt q = \left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right] \vt g.
+\end{equation}
+\end{subequations}
+
+As $\mathcal S_q$ and $\mathcal S_z$ are fixed in space, one can conclude that the quaternion $\vt g$ remains constant while both $\vt q$ and $\vt z$ vary in time as the body rotates.
+In addition, because Eq.~\ref{eq:relation_qdot_omega} was derived for an arbitrary space-fixed frame, it is clear that the equation $\dot{\vt z} = \tfrac{1}{2}{\vt B}(\vt z)\vt \omega$ is equally valid.
+Therefore, as an alternative to following the evolution of $\vt q$ along a period of time, one can select some convenient unit quaternion $\vt z$ at any instant, compute $\vt g$ using Eq.~\ref{eq:comb_rot}, and then follow the evolution of $\vt z$ from this instant on.
+Of course, Eq.~\ref{eq:inv_comb_rot} can be employed whenever the value of $\vt q$ is needed afterwards.
 
@@
 
-The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B$ depends on $\vt q$ only.
-However, if we define
-\[
-\mt \Omega = \left[
-\begin{smallarray}{rrrr}
--\pi_1 & -\pi_2 & -\pi_3 \\
- \pi_0 & -\pi_3 &  \pi_2 \\
- \pi_3 &  \pi_0 & -\pi_1 \\
--\pi_2 &  \pi_1 &  \pi_0
-\end{smallarray}
-\right],
-\]
-it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
+The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B(\vt q)$ does not depend on it.
+However, if we define $\mt \Omega = \mt B(\vt \pi)$, it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
 \begin{equation}
@@
 
-If no forces act on a rigid body, then the rotational energy $K_r$ and the angular momentum norm $L$ are integrals of motion, so that we can write
-\[
-\begin{cases}
-I_1 \omega_1^2 + I_2 \omega_2^2 + I_3 \omega_3^2 = 2 K_r \\
-I_1^2 \omega_1^2 + I_2^2 \omega_2^2 + I_3^2 \omega_3^2 = L^2
-\end{cases}
-\]
-
-Evidently, one can promptly evaluate $K_r$ and $L$ if the initial angular velocity vector ${\vt \omega}_0 = [\omega_{1,0}]$ is known.
-By solving the equations above for $\omega_1^2$ and $\omega_3^2$, we obtain
-\begin{subequations}
-\begin{align}
-\omega_1^2 = \frac{L^2-2I_3K_r}{I_1(I_1-I_3)}\left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right] \label{eq:solution_omega_1} \\
-\omega_3^2 = \frac{2I_1K_r-L^2}{I_3(I_1-I_3)}\left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right] \label{eq:solution_omega_3}
-\end{align}
-\end{subequations}
-where
-\begin{equation}
-\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \quad \text{and} \quad \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
-\end{equation}
-
-By realizing that the prefactor in Eqs.~\ref{eq:solution_omega_1} and \ref{eq:solution_omega_3} are constant and that these equations are also valid at $t = 0$, we can rewrite then as
-\begin{subequations}
-	\begin{align}
-	\omega_1^2 = A_1^2 \left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right] \label{eq:solution_omega_1} \\
-	\omega_3^2 = A_2^2 \left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right] \label{eq:solution_omega_3}
-	\end{align}
-\end{subequations}
-where
-\begin{equation}
-A_1 = \frac{\lambda_1 \omega_{1,0}}{\sqrt{\lambda_1^2 - \omega_{2,0}^2}} \quad \text{and} \quad
-A_3 = \frac{\lambda_3 \omega_{3,0}}{\sqrt{\lambda_3^2 - \omega_{2,0}^2}}.
-\end{equation}
-
-Substituting in the Euler equation for $\omega_2$, we have
-\[
-I_2 \dot{\omega}_2 - (I_1-I_3) A_1 A_3 \sqrt{\left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right]\left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right]}
-\]
-
-By defining $x = \omega_2/\lambda_\text{min}$ and $\kappa = \lambda_\text{min}/\lambda_\text{max}$, where $\lambda_\text{min} = \min(\lambda_1,\lambda_3)$ and $\lambda_\text{max} = \max(\lambda_1,\lambda_3)$, we obtain the following initial value problem for $x$:
-\[
-\begin{cases}
-I_2 \lambda_\text{min} \dot{x} - (I_1-I_3) A_1 A_3 \sqrt{(1-x^2)(1-\kappa^2x^2)} = 0 \\
-x(0) = \omega_{2,0}/\lambda_\text{min}.
-\end{cases}
-\]
-
-The solution of the equation above can be expressed using the elliptic Jacobi function $\text{sn}$, so that we can calculate $\omega_2$ at any time as
-\[
-\omega_2(t) = \lambda_\text{min} \text{sn}(\alpha t + \epsilon,\kappa),
-\]
-where
-\[
-\alpha = -\frac{(I_1-I_3)A_1A_3}{\lambda_\text{min} I_2} \quad \text{and} \quad \epsilon = \text{arcsn}\left(\frac{\omega_{2,0}}{\lambda_\text{min}},\kappa\right).
-\]
+Before describing the method, we consider the transformati
+
+Method based on work of Kosenko\cite{Kosenko_1998} and Celledoni \textit{et al}.\cite{Celledoni_2008} and also related to the work of van Zon \textit{et al}.\cite{vanZon2007}.
+
+Let $\vt L = \tr{[\mt A(\vt q)]}{\mt I}{\vt \omega}$ be the angular momentum vector
+
+During a torque-free rotational motion, the angular momentum vector $\vt L$ remains constant in any space-fixed frame of reference.
+
+In the body-fixed frame, although the
+
+$\vt L = \tr{[\mt A(\vt q)]}{\mt I}{\vt \omega}$ remains constant while both the quaternion $\vt q$ and the body-fixed angular velocity $\vt \omega$ vary.
+
+ Once $\vt L$ is computed from the initial condition and the evolution of $\vt \omega$ is known (after solving the Euler equation), this fact constrains the possible evolution of $\vt q$.
+By premultiplying $\vt L$ by $\vt C$, one has that $\vt C \vt L = \vt B \mt I \vt \omega$.
+In other words, $\vt q$ must vary in such a way that the mapping of both $\vt L$ and $\mt I\vt \omega$ in its orthogonal plane always coincide.
+Then, resorting to Eq.~\ref{eq:product_B_vector} we can write
+\begin{equation*}
+\sum_{k=1}^3\left(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k\right){\vt q} = \vt 0,
+\end{equation*}
+where $\hat{\vt C}_k$ and $\hat{\vt B}_k$ are the permutation matrices defined in Appendix~\ref{sec:auxiliary_math}.
+Therefore, $\vt q$ is constrained to lie in the null space of a matrix $\mt G = \sum_{k=1}^3(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k)$. The determinant of this matrix can be obtained as
+\begin{equation*}
+\det \mt G = (\tr{\vt \omega}{\mt I}^2{\vt \omega} - L^2)^2,
+\end{equation*}
+which is identically null in a free rotational motion.
+In fact, this is a requirement for a matrix to have a non-empty null space.
+Using a computer algebra package, it is straightforward to find out that $\vt G$ has a two-dimensional null space whose basis set is formed by the two vectors
+\begin{equation*}
+\vt v_1 = \left[\begin{smallarray}{c}
+L_2 + I_2 \omega_2 \\
+L_3 - I_3 \omega_3 \\
+0 \\
+L_1 - I_1 \omega_1
+\end{smallarray}\right]
+\quad \text{and} \quad
+\vt v_1 = \left[\begin{smallarray}{c}
+L_3 + I_3 \omega_3 \\
+L_2 - I_2 \omega_2 \\
+L_1 - I_1 \omega_1 \\
+0
+\end{smallarray}\right].
+\end{equation*}
 
@@
 Along the paper, we employ the gradient convention (rather than the Jacobian convention) for expressing derivatives with respect to vector variables.
-Hence, very often we differentiate a scalar function using one of the following rules:
+In this context, very often we differentiate a scalar function using one of the following rules:
 \[
@@
 orthogonality of $\vt y$ and $\vt q$) for the operation to be reversible by $\mt B$.
-In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999} As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = [\begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}]$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
+In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999}
+
+As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = [\begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}]$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
 \[
@@
 
-We now turn the attention to the product $\tr{\mt B}\dot{\mt B}$, where $\dot{\mt B} = \left[\begin{smallarray}{ccc} \hat{\mt B}_1\dot{\vt q} & \hat{\mt B}_2\dot{\vt q} & \hat{\mt B}_3\dot{\vt q}\end{smallarray}\right]$ is the time-derivative of matrix $\mt B$.
-Each entry of this product is $\{\tr{\mt B}\dot{\mt B}\}_{ij} = \tr{\vt q}\tr{\hat{\mt B}_i} \hat{\mt B}_j\dot{\vt q}$.
+With $\dot{\mt B} = \left[\begin{smallarray}{ccc} \hat{\mt B}_1\dot{\vt q} & \hat{\mt B}_2\dot{\vt q} & \hat{\mt B}_3\dot{\vt q}\end{smallarray}\right]$ being the time-derivative of matrix $\mt B$, we now focus on the product $\tr{\mt B}\dot{\mt B}$, whose each entry is $\{\tr{\mt B}\dot{\mt B}\}_{ij} = \tr{\vt q}\tr{\hat{\mt B}_i} \hat{\mt B}_j\dot{\vt q}$.
 By noting that products involving the permutation matrices satisfy the relations\cite{Dichmann1999}
@@
 
-We can also express matrix $\mt C$ as $\left[\begin{smallarray}{ccc} \hat{\mt C}_1{\vt q} & \hat{\mt C}_2{\vt q} & \hat{\mt C}_3{\vt q}\end{smallarray}\right]$, where $\hat{\mt C}_1$, $\hat{\mt C}_2$, and $\hat{\mt C}_3$ are the following skew-symmetric permutation matrices:
+We can also express $\mt C = \left[\begin{smallarray}{ccc} \hat{\mt C}_1{\vt q} & \hat{\mt C}_2{\vt q} & \hat{\mt C}_3{\vt q}\end{smallarray}\right]$, where $\hat{\mt C}_1$, $\hat{\mt C}_2$, and $\hat{\mt C}_3$ are the following skew-symmetric permutation matrices:
 \[
@@
 
-However, products involving these matrices are slightly different from those in Eq.~\ref{eq:BB_products}. They are
+Products involving these matrices are slightly different from those in Eq.~\ref{eq:BB_products}. They are
 \begin{equation}
@@
 
-A special connection between the permutation matrices that make $\mt B$ and $\mt C$ is the fact that $\tr{\hat{\mt B}_i}{\hat{\mt C}_j} = \tr{\hat{\mt C}_j}{\hat{\mt B}_i}$ for any combination of $i$ and $j$.
+A special connection between the permutation matrices that make $\mt B$ and $\mt C$ is the fact that $\tr{\hat{\mt B}_i}{\hat{\mt C}_j}$ is symmetric for any combination of $i$ and $j$.
 A direct consequence is that $\tr{\vt q}\tr{\hat{\mt B}}_i\hat{\mt C}_j\dot{\vt q} = \tr{\dot{\vt q}}\tr{\hat{\mt B}}_i\hat{\mt C}_j{\vt q}$, which means that $\tr{\mt B}\dot{\mt C} = \tr{\dot{\mt B}}{\mt C}$.
-Thus, differentiating Eq.~\ref{eq:factorization_of_A} with respect to time, we conclude that $\dot{\mt A} = \tr{\mt B}\dot{\mt C} + \tr{\dot{\mt B}}{\mt C} = 2 \tr{\dot{\mt B}}{\mt C}$.
+Thus, differentiating Eq.~\ref{eq:factorization_of_A} with respect to time yields $\dot{\mt A} = \tr{\mt B}\dot{\mt C} + \tr{\dot{\mt B}}{\mt C} = 2 \tr{\dot{\mt B}}{\mt C}$.
 Note that Eq.~\ref{eq:basics_B} also applies for matrix $\mt C$, which means that, as far as $\tr{\vt q}\vt q = 1$, we can insert the product $\mt C\tr{\mt C}$ between $\tr{\mt B}$ and $\dot{\mt B}$ in Eq.~\ref{eq:relation_B_qdot} without altering its result.
@@
 
-As mentioned in Sec.~\ref{sec:kinematics}, ${\mt A}\tr{\dot{\mt A}} = \mt S(\vt \omega)$, which makes the equation above result in Eq.~\ref{eq:relation_qdot_omega}.
+As mentioned in Sec.~\ref{sec:kinematics}, ${\mt A}\tr{\dot{\mt A}} = \mt S(\vt \omega)$, which makes the equation above equivalent to Eq.~\ref{eq:relation_qdot_omega}.
 
@@
 
-Care must be taken when differentiating the potential energy $U$ with respect to $\vt q$ because the quaternion components are interdependent.
-Schay has demonstrated that, when a vector variable $\vt x$ is subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by\cite{Schay1995}
+Care must be taken when differentiating the potential energy with respect to $\vt q$ because the quaternion components are interdependent.
+According to Schay,\cite{Schay1995} when a vector variable $\vt x$ is subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by
 \[
@@
 where $\vt n$ is the unit vector proportional to $(\partial g/\partial \vt x)^\ast$, with the asterisk denoting a differentiation performed without considering the interdependence of the entries of $\vt x$.
-We note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\onlinecite{Schay1995}, where the Jacobian convention is used.
+Note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\onlinecite{Schay1995}, where the Jacobian convention is used.
 In the case of a unit quaternion,
@@
 Temporarily dropping the particle index for simplicity, we recall from Sec.~\ref{sec:kinematics} that $\vt \delta = \tr{\mt A}\vt d$, where $\vt d$ is a constant vector.
-Because each entry of matrix $\mt A = \tr{\mt B}{\mt C}$ is $A_{ij}=\tr{\vt q} \tr{\hat{\mt B}}_i \hat{\mt C}_j {\vt q}$, an entry of $\vt \delta$ is given by
+As each entry of $\mt A = \tr{\mt B}{\mt C}$ is $A_{ij}=\tr{\vt q} \tr{\hat{\mt B}}_i \hat{\mt C}_j {\vt q}$, an entry of $\vt \delta$ is given by
 \[
@@
 
-Because $\tr{\hat{\mt C}}_j \hat{\mt B}_i$ is symmetric, an unconstrained differentiation results in
+Due to the symmetry of $\tr{\hat{\mt C}}_j \hat{\mt B}_i$, an unconstrained differentiation results in
 \[
@@
 where we have carried out the second matrix product above, employed the properties presented in Eq.~\ref{eq:CC_products}, and then used the operator defined in Eq.~\ref{eq:operator_S}.
-Finally, after realizing that $\tr{\mt C}{\mt B}\vt d = \tr{\mt A}{\vt d} = \vt \delta$, we can used the result above to obtain
+Finally, after realizing that $\tr{\mt C}{\mt B}\vt d = \tr{\mt A}{\vt d} = \vt \delta$, we can use the result above to obtain
 \[
@@
 
+Therefore, the constrained derivative of $U$ with respect to $\vt q$ is proportional to the quaternion-fixed representation of the resultant torque over the body.
+
+\section{Derivation of the Euler equation and its solution for a torque-free rotation}
+\label{eq:derivation_euler}
+
+Recall that $\vt \pi$ is twice the quaternion-fixed version of the angular momentum vector.
+Thus, it is expected that Eq.~\ref{eq:euler_eq_quat} will result in the well-known Euler equation, which describes the evolution of the angular momentum in the body-fixed frame, if $\frac{1}{2} \tr{\mt B}$ is premultiplied to both sides.
+In order to show that this actually occurs, we start by carrying out the premultiplication to obtain
+\[
+\frac{1}{2} \tr{\mt B} \dot{\vt \pi} = \frac{1}{4} \tr{\mt B} \mt \Omega \vt \omega + \mt A \vt \tau.
+\]
+
+Once $\mt \Omega = \left[\begin{smallarray}{ccc} \hat{\mt B}_1{\vt \pi} & \hat{\mt B}_2{\vt \pi} & \hat{\mt B}_3{\vt \pi}\end{smallarray}\right]$, in analogy with Eq.~\ref{eq:BtBdot} we conclude that $\tr{\mt B} \mt \Omega = \mt S(\tr{\mt B}\vt \pi) = 2 \mt S(\mt I \vt \omega)$.
+In addition,
+\[
+\tr{\mt B} \dot{\vt \pi} = \frac{d(\tr{\mt B}\vt \pi)}{dt} - \tr{\dot{\mt B}}\vt \pi = 2 \mt I\dot{\vt \omega} - 2 \tr{\dot{\mt B}} \mt B \mt I \vt \omega,
+\]
+where we have inserted $\mt B \tr{\mt B}$ between $\tr{\dot{\mt B}}$ and $\vt \pi$ and substituted $\tr{\mt B} \vt \pi = 2 \mt I \vt \omega$.
+Finally, after transposing Eq.~\ref{eq:relation_B_qdot} with $\tr{\mt B}\dot{\vt q}$ replaced by $\frac{1}{2} \vt \omega$ and resorting to the relation between $\mt S(\cdot)$ and a cross product, we obtain
+\[
+\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau,
+\]
+which is exactly the Euler equation of motion for rigid bodies.\cite{Goldstein2002} Its main advantage, in comparison with Eq.~\ref{eq:euler_eq_quat}, is that it completely decouples from the quaternion $\vt q$ when $\vt \tau = \vt 0$.
+In such a torque-free situation, the vector equation can be rewritten as the following system:
+\begin{align*}
+I_1 \dot{\omega}_1 - (I_2 - I_3) \omega_2 \omega_3 = 0\\
+I_2 \dot{\omega}_2 + (I_1 - I_3) \omega_1 \omega_3 = 0 \\
+I_3 \dot{\omega}_3 - (I_1 - I_2) \omega_1 \omega_2 = 0
+\end{align*}
+
+Analytical solutions for this system are a textbook subject.\cite{Landau1976} Here we present the solution for the case under consideration, when $I_1 > I_2 > I_3$ (an asymmetrical top), in a form that aims to facilitate computer implementation, inspired by the forms presented in Refs.~\onlinecite{Landau1976} and \onlinecite{Thorn2012}.
+In a free rotation, the kinetic energy $K_r = \tfrac{1}{2} \tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-frame angular momentum $\vt L = \tfrac{1}{2}\tr{\mt C}{\vt \pi} = \tr{\mt A}{\mt I}{\vt \omega}$ are integrals of motion.
+Note that the square norm of $\vt L$ does not depend explicitly on $\vt q$ in any frame of reference:
+\[
+L^2 = \tr{\vt L}{\vt L} = \tr{\vt \omega}{\mt I}^2{\vt \omega} = \frac{1}{4}\tr{\vt \pi}{\vt \pi}.
+\]
+
+Therefore, if we compute $K_r$ and $L^2$ from the initial angular velocity ${\vt \omega}^0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, we can apply the following constraints to the Euler equations:
+\begin{align*}
+I_1 \omega_1^2 + I_2 \omega_2^2 + I_3 \omega_3^2 = 2 K_r \\
+I_1^2 \omega_1^2 + I_2^2 \omega_2^2 + I_3^2 \omega_3^2 = L^2
+\end{align*}
+
+By solving these equations for $\omega_1^2$ and $\omega_3^2$, we obtain
+\begin{equation}
+\label{eq:omegas_1_and_3}
+\omega_1^2 = A_1^2 \left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right] \;\; \text{and} \;\; \omega_3^2 = A_3^2 \left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right],
+\end{equation}
+where
+\begin{equation}
+\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \;\; \text{and} \;\; \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
+\end{equation}
+
+In order to obtain the prefactors $A_1$ and $A_3$ with the correct signs, we realize that Eqs.~\ref{eq:omegas_1_and_3} must be valid at the initial instant ($t = 0$), which makes
+\begin{equation}
+A_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}} \;\; \text{and} \;\;
+A_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}.
+\end{equation}
+
+By defining $x = {\omega_2}/{\lambda_\text{min}}$ and $\kappa = {\lambda_\text{min}}/{\lambda_\text{max}}$, where $\lambda_\text{min} = \min(\lambda_1,\lambda_3)$ and $\lambda_\text{max} = \max(\lambda_1,\lambda_3)$, we can transform the Euler equation for $\omega_2$ into to the following initial value problem for $x$:
+\[
+\begin{cases}
+I_2 \lambda_\text{min} \dot{x} + (I_1-I_3) A_1 A_3 \sqrt{(1-x^2)(1-\kappa^2x^2)} = 0 \\
+x(0) = \omega_2^0/\lambda_\text{min}.
+\end{cases}
+\]
+
+The solution of the equation above and its application to
+
+can be expressed using the elliptic Jacobi function $\text{sn}$, so that we can calculate $\omega_2$ at any time as
+\[
+\omega_2(t) = \lambda_\text{min} \text{sn}(\alpha t + \epsilon,\kappa),
+\]
+where
+\[
+\alpha = -\frac{(I_1-I_3)A_1A_3}{\lambda_\text{min} I_2} \quad \text{and} \quad \epsilon = \text{arcsn}\left(\frac{\omega_{2,0}}{\lambda_\text{min}},\kappa\right).
+\]
+
 \bibliography{rigid_bodies}

### 2016-09-28 rigid_bodies.tex
Avanços na soluçao analítica

@@
 	\arraycolsep=.13885em
-	\hbox\bgroup$\array{@{}#1@{}}}
-{\endarray$\egroup\egroup\,\null}
+	\hbox\bgroup$\left[\array{@{}#1@{}}}
+{\endarray\right]$\egroup\egroup\,\null}
 
@@
 \[
-{\mt I}^\ast = \sum_{j=1}^N M_j \left[ \begin{smallarray}{ccc}
+{\mt I}^\ast = \sum_{j=1}^N M_j \begin{smallarray}{ccc}
 {\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
@@
 -{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
-\end{smallarray} \right].
+\end{smallarray}.
 \]
@@
 \[
-{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \left[ \begin{smallarray}{ccc}
+{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \begin{smallarray}{ccc}
 I_1 &   0 &   0 \\
@@
   0 &   0 & I_3 \\
-\end{smallarray} \right],
+\end{smallarray},
 \]
@@
 \label{eq:A_from_q}
-\mt A(\vt q) = \left[ \begin{smallarray}{ccc}
+\mt A(\vt q) = \begin{smallarray}{ccc}
 q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
@@
 2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
-\end{smallarray} \right].
+\end{smallarray}.
 \]
@@
 \label{eq:def_B_and_C}
-\mt B(\vt q) = \left[
-\begin{smallarray}{rrrr}
+\mt B(\vt q) = \begin{smallarray}{rrrr}
 -q_2 & -q_3 & -q_4 \\
@@
 \end{smallarray}
-\right]
-\quad \text{and} \quad
-\mt C(\vt q) = \left[
-\begin{smallarray}{rrrr}
+\; \text{and} \;
+\mt C(\vt q) = \begin{smallarray}{rrrr}
 -q_2 & -q_3 & -q_4 \\
@@
  q_3 & -q_2 &  q_1
-\end{smallarray}
-\right].
+\end{smallarray}.
 \end{equation}
@@
 Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.
-
 We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
@@
 
-Hence, $\vt \pi$ is twice the quaternion-fixed version of the angular momentum ($\mt I \vt \omega$) and, as such, it is also orthogonal to $\vt q$.
+Hence, $\vt \pi$ is twice the quaternion-fixed version of the angular momentum $\vt L = \mt I \vt \omega$ and, as such, it is also orthogonal to $\vt q$.
 Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
@@
 &\dot{\vt p} = \mt F \\
-&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \\
-&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:euler_eq_quat}
+&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
+&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
 \end{align}
@@
 
-Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eq.~\ref{eq:EDO_system}:
+Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
 \begin{subequations}
@@
 \begin{align}
-&\dot{\vt r} = \frac{1}{m} \vt p \\
-&\dot{\vt p} = \vt F \\
 &\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
@@
 
-Before describing the method, we consider the transformati
+It is not possible Eq.~\ref{eq:EDO_system}
 
@@
 
+Considering that $I_1 > I_2 > I_3$, the prefactors of the angular momenta $\omega_1$, $\omega_2$, and $\omega_3$ are, respectively,
+\begin{align}
+\mt A = \left[\begin{array}{ccc}
+\frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}} & 0 & 0 \\
+0 & \min(\lambda_1,\lambda_3) & 0 \\
+0 & 0 & \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
+\end{array}\right], \\
+\end{align}
+\begin{align}
+A_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}}, \\
+A_2 = \min(\lambda_1,\lambda_3), \; \text{and} \\
+A_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}},
+\end{align}
+where
+\begin{equation}
+\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \;\; \text{and} \;\; \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
+\end{equation}
+
+The solution is
+\begin{subequations}
+\begin{align}
+\omega_1(t) &= \begin{cases}
+A_1\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
+A_1\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
+\end{cases} \\
+\omega_2(t) &= A_2 \text{sn}(a t + b,k) \\
+\omega_3(t) &= \begin{cases}
+A_3\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
+A_3\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
+\end{cases}
+\end{align}
+\end{subequations}
+where
+\begin{align*}
+&a = -\frac{(I_1-I_3)A_1A_3}{I_2 A_2},\\
+&b = \text{sn}^{-1}\left(\frac{\omega_{2,0}}{A_2},k\right),\quad\text{and}\\
+&k = \frac{A_2}{\text{max}(\lambda_1,\lambda_3)}
+\end{align*}
+
+During a torque-free rotational motion, the angular momentum vector $\vt L$ remains constant in any space-fixed frame of reference.
+
+In the body-fixed frame, although the
+
 Let $\vt L = \tr{[\mt A(\vt q)]}{\mt I}{\vt \omega}$ be the angular momentum vector
-
-During a torque-free rotational motion, the angular momentum vector $\vt L$ remains constant in any space-fixed frame of reference.
-
-In the body-fixed frame, although the
 
@@
 \begin{equation*}
-\vt v_1 = \left[\begin{smallarray}{c}
+\vt v_1 = \begin{smallarray}{c}
 L_2 + I_2 \omega_2 \\
@@
 L_1 - I_1 \omega_1
-\end{smallarray}\right]
+\end{smallarray}
 \quad \text{and} \quad
-\vt v_1 = \left[\begin{smallarray}{c}
+\vt v_1 = \begin{smallarray}{c}
 L_3 + I_3 \omega_3 \\
@@
 0
-\end{smallarray}\right].
+\end{smallarray}.
 \end{equation*}
+
+\begin{equation*}
+\exp\left(-\begin{smallarray}{cc}
+ \alpha(t) & \phi(t) \\
+-\phi(t) & \alpha(t)
+\end{smallarray}\right) = 
+e^{-\alpha(t)} \begin{smallarray}{cc}
+\cos \phi(t) & -\sin \phi(t) \\
+\sin \phi(t) &  \cos \phi(t)
+\end{smallarray}
+\end{equation*}
+
+Here,
+\begin{equation}
+\alpha(t) = \frac{1}{2}\int_0^t \frac{I_1 \omega_1^\prime(\tau)}{L - I_1\omega_1(\tau)}d\tau =
+\frac{1}{2}\ln\left(\frac{L - I_1\omega_1}{L - I_1\omega_1^0}\right)
+\end{equation}
+
+\begin{equation}
+\varphi(t) = \frac{1}{2}\int_0^t \frac{2 K_r + L \omega_1(\tau)}{L + I_1\omega_1(\tau)}d\tau
+\end{equation}
+
+After some algebra\cite{Celledoni_2008}
+\begin{equation*}
+\varphi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2 I_1 L}\int_0^t \frac{1}{1+(I_1/L)\omega_1(\tau)}d\tau.
+\end{equation*}
+
+If $\lambda_1 < \lambda_3$, then $\omega_1(t) = A_1 \text{cn}(\alpha t + \epsilon)$
 
@@
 Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
-\[
+\begin{equation*}
 \dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
-\]
+\end{equation*}
 where $i\!L$ is the Liouville operator.
@@
 &{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
-&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
+&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
 &\eta_1(t) = \eta_1^0 + \alpha_1 t
@@
 &{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \quad \text{and} \label{eq:solution_p} \\
-&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t. \label{eq:solution_pi}
+&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t. \label{eq:solution_pi}
 \end{align}
@@
 
-As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = [\begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}]$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
-\[
-\hat{\mt B}_1 = \left[ \begin{smallarray}{cccc}
+As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = \begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
+\[
+\hat{\mt B}_1 = \begin{smallarray}{cccc}
  0 & \text{-}1 &  0 &  0 \\
@@
  0 &  0 & \text{-}1 &  0 \\
-\end{smallarray} \right] \quad
-\hat{\mt B}_2 = \left[ \begin{smallarray}{cccc}
+\end{smallarray} \quad
+\hat{\mt B}_2 = \begin{smallarray}{cccc}
  0 &  0 & \text{-}1 &  0 \\
@@
  0 &  1 &  0 &  0 \\
-\end{smallarray} \right] \quad
-\hat{\mt B}_3 = \left[ \begin{smallarray}{cccc}
+\end{smallarray} \quad
+\hat{\mt B}_3 = \begin{smallarray}{cccc}
  0 &  0 &  0 & \text{-}1 \\
@@
  1 &  0 &  0 &  0 \\
-\end{smallarray} \right]
+\end{smallarray}
 \]
@@
 \label{eq:product_B_vector}
-\mt B \vt x = \left[ \begin{smallarray}{ccc} \hat{\mt B}_1 \vt q & \hat{\mt B}_2 \vt q & \hat{\mt B}_3 \vt q \end{smallarray}\right] \left[ \begin{smallarray}{c} x_1 \\ x_2 \\ x_3 \end{smallarray} \right] = \sum_{j=1}^3 x_j \hat{\mt B}_j \vt q.
+\mt B \vt x = \begin{smallarray}{ccc} \hat{\mt B}_1 \vt q & \hat{\mt B}_2 \vt q & \hat{\mt B}_3 \vt q \end{smallarray} \begin{smallarray}{c} x_1 \\ x_2 \\ x_3 \end{smallarray} = \sum_{j=1}^3 x_j \hat{\mt B}_j \vt q.
 \end{equation}
@@
 \tr{\mt B}\vt \pi =
-\left[\begin{smallarray}{c}
+\begin{smallarray}{c}
 \tr{\vt q}\tr{\hat{\mt B}}_1 \\
@@
 \tr{\vt q}\tr{\hat{\mt B}}_3 \\
-\end{smallarray}\right]\vt \pi = 
-\left[\begin{smallarray}{c}
+\end{smallarray}\vt \pi = 
+\begin{smallarray}{c}
 \tr{\vt q}\tr{\hat{\mt B}}_1\vt \pi \\
@@
 \tr{\vt q}\tr{\hat{\mt B}}_3\vt \pi \\
-\end{smallarray}\right] = 
-\left[\begin{smallarray}{c}
+\end{smallarray} = 
+\begin{smallarray}{c}
 \tr{\vt \pi}{\hat{\mt B}}_1\vt q \\
@@
 \tr{\vt \pi}{\hat{\mt B}}_3\vt q \\
-\end{smallarray}\right].
-\end{equation}
-
-With $\dot{\mt B} = \left[\begin{smallarray}{ccc} \hat{\mt B}_1\dot{\vt q} & \hat{\mt B}_2\dot{\vt q} & \hat{\mt B}_3\dot{\vt q}\end{smallarray}\right]$ being the time-derivative of matrix $\mt B$, we now focus on the product $\tr{\mt B}\dot{\mt B}$, whose each entry is $\{\tr{\mt B}\dot{\mt B}\}_{ij} = \tr{\vt q}\tr{\hat{\mt B}_i} \hat{\mt B}_j\dot{\vt q}$.
+\end{smallarray}.
+\end{equation}
+
+With $\dot{\mt B} = \begin{smallarray}{ccc} \hat{\mt B}_1\dot{\vt q} & \hat{\mt B}_2\dot{\vt q} & \hat{\mt B}_3\dot{\vt q}\end{smallarray}$ being the time-derivative of matrix $\mt B$, we now focus on the product $\tr{\mt B}\dot{\mt B}$, whose each entry is $\{\tr{\mt B}\dot{\mt B}\}_{ij} = \tr{\vt q}\tr{\hat{\mt B}_i} \hat{\mt B}_j\dot{\vt q}$.
 By noting that products involving the permutation matrices satisfy the relations\cite{Dichmann1999}
@@
 \label{eq:BtBdot}
-\tr{\mt B}\dot{\mt B} = \left[
-\begin{smallarray}{ccc}
+\tr{\mt B}\dot{\mt B} = \begin{smallarray}{ccc}
 0 & -\tr{\vt q} \tr{\hat{\mt B}_3}\dot{\vt q} &  \tr{\vt q} \tr{\hat{\mt B}_2}\dot{\vt q} \\
@@
 -\tr{\vt q} \tr{\hat{\mt B}_2}\dot{\vt q} & \tr{\vt q} \tr{\hat{\mt B}_1}\dot{\vt q} & 0
-\end{smallarray}
-\right].
+\end{smallarray}.
 \end{equation}
@@
 \label{eq:operator_S}
-\mt S(\vt x) = \left[ \begin{smallarray}{ccc}
+\mt S(\vt x) = \begin{smallarray}{ccc}
 0   & -x_3 &  x_2 \\
@@
 -x_2 &  x_1 &  0
-\end{smallarray}\right].
+\end{smallarray}.
 \end{equation}
@@
 
-We can also express $\mt C = \left[\begin{smallarray}{ccc} \hat{\mt C}_1{\vt q} & \hat{\mt C}_2{\vt q} & \hat{\mt C}_3{\vt q}\end{smallarray}\right]$, where $\hat{\mt C}_1$, $\hat{\mt C}_2$, and $\hat{\mt C}_3$ are the following skew-symmetric permutation matrices:
-\[
-\hat{\mt C}_1 = \left[ \begin{smallarray}{cccc}
+We can also express $\mt C = \begin{smallarray}{ccc} \hat{\mt C}_1{\vt q} & \hat{\mt C}_2{\vt q} & \hat{\mt C}_3{\vt q}\end{smallarray}$, where $\hat{\mt C}_1$, $\hat{\mt C}_2$, and $\hat{\mt C}_3$ are the following skew-symmetric permutation matrices:
+\[
+\hat{\mt C}_1 = \begin{smallarray}{cccc}
  0 & \text{-}1 &  0 &  0 \\
@@
  0 &  0 &  1 &  0 \\
-\end{smallarray} \right] \quad
-\hat{\mt C}_2 = \left[ \begin{smallarray}{cccc}
+\end{smallarray} \quad
+\hat{\mt C}_2 = \begin{smallarray}{cccc}
  0 &  0 & \text{-}1 &  0 \\
@@
  0 & \text{-}1 &  0 &  0 \\
-\end{smallarray} \right] \quad
-\hat{\mt C}_3 = \left[ \begin{smallarray}{cccc}
+\end{smallarray} \quad
+\hat{\mt C}_3 = \begin{smallarray}{cccc}
  0 &  0 &  0 & \text{-}1 \\
@@
 \diff{\vt \delta}{\vt q} = -2{\mt C}\tr{\mt C}
-\left[\begin{smallarray}{ccc}
+\begin{smallarray}{ccc}
 {\hat{\mt C}}_1 {\mt B}{\vt d} & {\hat{\mt C}}_2 {\mt B}{\vt d} & {\hat{\mt C}}_3 {\mt B}{\vt d}
-\end{smallarray}\right] = 2{\mt C}\;{\mt S}\!\left(\tr{\mt C}{\mt B}\vt d\right),
+\end{smallarray} = 2{\mt C}\;{\mt S}\!\left(\tr{\mt C}{\mt B}\vt d\right),
 \]
@@
 Recall that $\vt \pi$ is twice the quaternion-fixed version of the angular momentum vector.
-Thus, it is expected that Eq.~\ref{eq:euler_eq_quat} will result in the well-known Euler equation, which describes the evolution of the angular momentum in the body-fixed frame, if $\frac{1}{2} \tr{\mt B}$ is premultiplied to both sides.
+Thus, it is expected that Eq.~\ref{eq:EDO_pi} will result in the well-known Euler equation, which describes the evolution of the angular momentum in the body-fixed frame, if $\frac{1}{2} \tr{\mt B}$ is premultiplied to both sides.
 In order to show that this actually occurs, we start by carrying out the premultiplication to obtain
@@
 
-Once $\mt \Omega = \left[\begin{smallarray}{ccc} \hat{\mt B}_1{\vt \pi} & \hat{\mt B}_2{\vt \pi} & \hat{\mt B}_3{\vt \pi}\end{smallarray}\right]$, in analogy with Eq.~\ref{eq:BtBdot} we conclude that $\tr{\mt B} \mt \Omega = \mt S(\tr{\mt B}\vt \pi) = 2 \mt S(\mt I \vt \omega)$.
+Once $\mt \Omega = \begin{smallarray}{ccc} \hat{\mt B}_1{\vt \pi} & \hat{\mt B}_2{\vt \pi} & \hat{\mt B}_3{\vt \pi}\end{smallarray}$, in analogy with Eq.~\ref{eq:BtBdot} we conclude that $\tr{\mt B} \mt \Omega = \mt S(\tr{\mt B}\vt \pi) = 2 \mt S(\mt I \vt \omega)$.
 In addition,
@@
 \]
-which is exactly the Euler equation of motion for rigid bodies.\cite{Goldstein2002} Its main advantage, in comparison with Eq.~\ref{eq:euler_eq_quat}, is that it completely decouples from the quaternion $\vt q$ when $\vt \tau = \vt 0$.
+which is exactly the Euler equation of motion for rigid bodies.\cite{Goldstein2002} Its main advantage, in comparison with Eq.~\ref{eq:EDO_pi}, is that it completely decouples from the quaternion $\vt q$ when $\vt \tau = \vt 0$.
 In such a torque-free situation, the vector equation can be rewritten as the following system:

### 2016-09-29 rigid_bodies.tex
Mais avanço na solução analítica

@@
 
-It is not possible Eq.~\ref{eq:EDO_system}
-
-Method based on work of Kosenko\cite{Kosenko_1998} and Celledoni \textit{et al}.\cite{Celledoni_2008} and also related to the work of van Zon \textit{et al}.\cite{vanZon2007}.
-
-Considering that $I_1 > I_2 > I_3$, the prefactors of the angular momenta $\omega_1$, $\omega_2$, and $\omega_3$ are, respectively,
-\begin{align}
-\mt A = \left[\begin{array}{ccc}
-\frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}} & 0 & 0 \\
-0 & \min(\lambda_1,\lambda_3) & 0 \\
-0 & 0 & \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
-\end{array}\right], \\
-\end{align}
-\begin{align}
-A_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}}, \\
-A_2 = \min(\lambda_1,\lambda_3), \; \text{and} \\
-A_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}},
-\end{align}
-where
-\begin{equation}
-\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \;\; \text{and} \;\; \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
-\end{equation}
-
-The solution is
+It is not possible to analytically integrate the equations of motion in Eq.~\ref{eq:EDO_system} for a general potential $U(\vt r,\vt q)$.
+Nonetheless, closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_q} are known for the particular case of torque-free rotation (i.e.
+when $\vt \tau = \vt 0$ at all times).
+Even though free rotations might seem of limited use in Molecular Dynamics at first sight, they are very useful as part of numerical integration methods based on splitting strategies.
+This is the topic addressed in Sec.~\ref{sec:symplectic}.
+As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} can be used to derive the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is completely decoupled from Eq.~\ref{eq:EDO_pi} when $\vt \tau = \vt 0$.
+It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum $\vt L = \tr{\mt A}{\mt I}{\vt \omega}$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation.
+This allows us to analytically solve the Euler equation for the angular momentum vector $\vt \omega$ in terms of elliptic functions.\cite{Byrd_1971, Landau1976, Kosenko_1998, vanZon2007, Celledoni_2008, Thorn2012} From an initial condition $\vt \omega(0) = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, evaluating such solution starts by computing three prefactors $a_1$, $a_2$, and $a_3$, defined as
+\begin{gather*}
+a_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}}, \quad a_2 = \min(\lambda_1,\lambda_3), \\
+a_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}},
+\end{gather*}
+where $\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}}$ and $\lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}$ with, as already mentioned, $I_1 > I_2 > I_3$.
+Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$.
+This also occurs for $\vt \omega$, whose functional form is
 \begin{subequations}
-\begin{align}
-\omega_1(t) &= \begin{cases}
-A_1\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
-A_1\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
-\end{cases} \\
-\omega_2(t) &= A_2 \text{sn}(a t + b,k) \\
-\omega_3(t) &= \begin{cases}
-A_3\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
-A_3\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
-\end{cases}
-\end{align}
+\begin{equation}
+\vt \omega(t) = \left[\begin{array}{c}
+a_1\text{cn}(b t + c, k^2) \\
+a_2\text{sn}(b t + c, k^2) \\
+a_3\text{dn}(b t + c, k^2)
+\end{array}\right] \qquad \text{if} \quad \lambda_1 < \lambda_3
+\end{equation}
+or
+\begin{equation}
+\vt \omega(t) = \left[\begin{array}{c}
+a_1\text{dn}(b t + c, k^2) \\
+a_2\text{sn}(b t + c, k^2) \\
+a_3\text{cn}(b t + c, k^2)
+\end{array}\right] \qquad \text{if} \quad \lambda_1 > \lambda_3
+\end{equation}
 \end{subequations}
-where
-\begin{align*}
-&a = -\frac{(I_1-I_3)A_1A_3}{I_2 A_2},\\
-&b = \text{sn}^{-1}\left(\frac{\omega_{2,0}}{A_2},k\right),\quad\text{and}\\
-&k = \frac{A_2}{\text{max}(\lambda_1,\lambda_3)}
-\end{align*}
-
-During a torque-free rotational motion, the angular momentum vector $\vt L$ remains constant in any space-fixed frame of reference.
-
-In the body-fixed frame, although the
+where $\text{sn}(u,m)$, $\text{cn}(u,m)$, and $\text{dn}(u,m)$ are the Jacobi elliptic functions, $k = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ is the elliptic modulus, and the constants $b$ and $c$ are given by
+\begin{equation*}
+b = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad c = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}, k^2\right).
+\end{equation*}
+
+In order to compute the inverse function $\text{sn}^{-1}$, one can use its relation with an elliptic integral in Carlson form, \cite{Carlson_1977} $\text{sn}^{-1}(x,m) = x R_F(1-x^2,1-mx^2,1)$.
+In this work, we employ the routines available in Ref.~\onlinecite{Galassi_2009} for the numerical evaluation of Jacobi functions and Carlson integrals.
+
+Once we know $\vt \omega$ as a function of time, we can also solve Eq.~\ref{eq:EDO_q} analytically for the quaternion $\vt q$.
+Based on an approach pioneered by Kosenko\cite{Kosenko_1998} and extended by Celledoni \textit{et al}.,\cite{Celledoni_2008} we employ the factorization in Eq.\ref{eq:factorization_of_A} to accomplish such task.
 
@@
 
+\begin{equation*}
+\Pi(x,n,m) = \text{sn}^{-1}(x,m) + \frac{1}{3}nx^3R_J(1-x^2,1-mx^2,1,1-nx^2).
+\end{equation*}
+
 \subsection{Symplectic Numerical Integration}
@@
 \section{Derivation of the Euler equation and its solution for a torque-free rotation}
-\label{eq:derivation_euler}
+\label{sec:derivation_euler}
 

### 2016-10-11 rigid_bodies.tex
Início da Correção da Solução Analítica

--- 2016-09-28 rigid_bodies.tex
@@
 
+\newcommand{\sn}{\text{sn}}                             % Jacobi function sn
+\newcommand{\cn}{\text{cn}}                             % Jacobi function cn
+\newcommand{\dn}{\text{dn}}                             % Jacobi function dn
+
 \newenvironment{smallarray}[1]                          % small arrays
 {\null\,\vcenter\bgroup\scriptsize
-	\renewcommand{\arraystretch}{0.7}
+	\renewcommand{\arraystretch}{1.5}
 	\arraycolsep=.13885em
@@
 The center of mass of this body is located at
-\[
+\begin{equation*}
 \vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
-\]
+\end{equation*}
 where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body.
@@
 In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
-\[
+\begin{equation*}
 {\mt I}^\ast = \sum_{j=1}^N M_j \begin{smallarray}{ccc}
@@
 \end{smallarray}.
-\]
+\end{equation*}
 
@@
 Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
-\[
+\begin{equation*}
 {\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \begin{smallarray}{ccc}
@@
 \end{smallarray},
-\]
+\end{equation*}
 where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation.
-With no loss of generality, we can consider that $I_1 > I_2 > I_3$.
+With no loss of generality, we can consider that $I_1 \geq I_2 \geq I_3$.
 Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g.
@@
 Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
-\[
+\begin{equation*}
 \label{eq:A_from_q}
@@
 To conclude this section, we discuss on the possibility of switching between alternative space-fixed frames when describing the motion of a rigid body.
-This will be useful when we deal with torque-free rotations in Sec.~\ref{sec:free_rotation}.
-Consider that a body-fixed vector $\vt d$ is, at the same instant $t$, observed as $\vt \delta = \tr{[\mt A(\vt q)]}{\vt d}$ in a space-fixed frame $\mathcal S_q$ and as $\vt \eta = \tr{[\mt A(\vt z)]}{\vt d}$ in another space-fixed frame $\mathcal S_z$.
-It thus follows that $\vt \eta = \tr{[\mt A(\vt z)]}{\mt A}(\vt q)\vt \delta$, which means that these two reference frames are rotated with respect to each other by a composite matrix $\mt A(\vt g) = \tr{[\mt A(\vt z)]}{\mt A}(\vt q)$.
-A simple relation between the unit quaternions $\vt g$, $\vt z$, and $\vt q$ is a key result in quaternion algebra.\cite{Dichmann1999} Adapted to our matrix/vector notation, such relation is
-\begin{subequations}
-\begin{equation}
- \label{eq:comb_rot}
- \vt g = \tr{\left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right]} \vt q.
-\end{equation}
-
-The block matrix above comprises an orthogonal transformation in $\mathbb{R}^4$, which means that the inverse transformation is simply given by
-\begin{equation}
-\label{eq:inv_comb_rot}
-\vt q = \left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right] \vt g.
-\end{equation}
-\end{subequations}
-
-As $\mathcal S_q$ and $\mathcal S_z$ are fixed in space, one can conclude that the quaternion $\vt g$ remains constant while both $\vt q$ and $\vt z$ vary in time as the body rotates.
-In addition, because Eq.~\ref{eq:relation_qdot_omega} was derived for an arbitrary space-fixed frame, it is clear that the equation $\dot{\vt z} = \tfrac{1}{2}{\vt B}(\vt z)\vt \omega$ is equally valid.
-Therefore, as an alternative to following the evolution of $\vt q$ along a period of time, one can select some convenient unit quaternion $\vt z$ at any instant, compute $\vt g$ using Eq.~\ref{eq:comb_rot}, and then follow the evolution of $\vt z$ from this instant on.
-Of course, Eq.~\ref{eq:inv_comb_rot} can be employed whenever the value of $\vt q$ is needed afterwards.
+This will be useful when we deal with torque-free rotations (Sec.~\ref{sec:free_rotation}).
+Consider that a body-fixed vector $\vt d$ is simultaneously observed as $\vt \delta = \tr{\mt A\!}(\vt q){\vt d}$ in a space-fixed frame $\mathcal S_q$ and as $\vt \xi = \tr{\mt A\!}(\vt z){\vt d}$ in another space-fixed frame $\mathcal S_z$.
+It follows that $\vt \xi = \mt A(\vt g)\vt \delta$, where the composite rotation matrix $\mt A(\vt g) = \tr{\mt A\!}(\vt z){\mt A}(\vt q)$ denotes the relative orientation between $\mathcal S_q$ and $\mathcal S_z$.
+One key result in quaternion algebra\cite{Dichmann1999} is a simple relation that exists between $\vt q$, $\vt z$, and $\vt g$, which can be adapted to our notation as $\vt g = \tr{[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}]} \vt q$.
+Note that both $\vt q$ and $\vt z$ vary while the body rotates, but the quaternion $\vt g$ remains constant because $\mathcal S_q$ and $\mathcal S_z$ are fixed in space.
+Moreover, as Eq.~\ref{eq:relation_qdot_omega} was derived for an arbitrary space-fixed frame, it is also true that $\dot{\vt z} = \tfrac{1}{2}{\vt B}(\vt z)\vt \omega$.
+In conclusion, as an alternative to tracking the evolution of $\vt q(t)$ after an instant $t_0$ in which $\vt q = \vt q_0$, one can select some other convenient unit quaternion $\vt z_0$ and then track the evolution of $\vt z(t)$, starting from $\vt z(t_0) = \vt z_0$.
+Then, whenever $\vt q$ is needed, one can simply make
+\begin{equation}
+\label{eq:relation_q_z}
+\vt q = [\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}] \tr{[\begin{array}{cc}\vt z_0 & \mt C(\vt z_0)\end{array}]} \vt q_0,
+\end{equation}
+which is correct because these block matrices represent orthogonal transformations in $\mathbb{R}^4$.
 
@@
 Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
-\[
+\begin{equation*}
 \mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
-\]
+\end{equation*}
 
@@
 
-It is not possible Eq.~\ref{eq:EDO_system}
-
-Method based on work of Kosenko\cite{Kosenko_1998} and Celledoni \textit{et al}.\cite{Celledoni_2008} and also related to the work of van Zon \textit{et al}.\cite{vanZon2007}.
-
-Considering that $I_1 > I_2 > I_3$, the prefactors of the angular momenta $\omega_1$, $\omega_2$, and $\omega_3$ are, respectively,
-\begin{align}
-\mt A = \left[\begin{array}{ccc}
-\frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}} & 0 & 0 \\
-0 & \min(\lambda_1,\lambda_3) & 0 \\
-0 & 0 & \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
-\end{array}\right], \\
-\end{align}
-\begin{align}
-A_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}}, \\
-A_2 = \min(\lambda_1,\lambda_3), \; \text{and} \\
-A_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}},
-\end{align}
-where
-\begin{equation}
-\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \;\; \text{and} \;\; \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
-\end{equation}
-
-The solution is
+It is not possible to analytically integrate the equations of motion for a general potential $U(\vt r,\vt q)$.
+Nonetheless, closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known for the particular case of torque-free rotation (i.e.
+when $\vt \tau = \vt 0$).
+Free rotations are very useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies, which is the topic addressed in Sec.~\ref{sec:symplectic}.
+As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is completely decoupled from Eq.~\ref{eq:EDO_pi} when $\vt \tau = \vt 0$.
+It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum vector $\vt L$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation.
+This allows us to analytically solve the Euler equation for the angular velocity vector $\vt \omega$.
+Here we present the solution for an asymmetric top ($I_1 > I_2 > I_3$) in a form that aims to facilitate computer implementation, inspired by the derivations presented in Refs.~\onlinecite{Landau1976} and \onlinecite{Thorn2012}.
+From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, such solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a vector of prefactors defined as
+\begin{equation*}
+\vt a = \begin{smallarray}{c}
+\frac{\omega_1^0}{\sqrt{1 - \left(\omega_2^0/\lambda_1\right)^2}} \\
+\min(\lambda_1,\lambda_3) \\
+\frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
+\end{smallarray},
+\; \text{where} \;
+\begin{cases}
+\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \\
+\lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}
+\end{cases}.
+\end{equation*}
+
+Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$. This also occurs for $\vt \omega$, whose functional form is
 \begin{subequations}
 \begin{align}
-\omega_1(t) &= \begin{cases}
-A_1\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
-A_1\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
-\end{cases} \\
-\omega_2(t) &= A_2 \text{sn}(a t + b,k) \\
-\omega_3(t) &= \begin{cases}
-A_3\text{dn}(a t + b,k) \quad \text{if} \quad \lambda_1 < \lambda_3 \\
-A_3\text{cn}(a t + b,k) \quad \text{if} \quad \lambda_1 > \lambda_3
-\end{cases}
+\vt \omega(t) &= \begin{smallarray}{c}
+a_1\text{cn}(\omega_p t + \epsilon, k^2) \\
+a_2\text{sn}(\omega_p t + \epsilon, k^2) \\
+a_3\text{dn}(\omega_p t + \epsilon, k^2)
+\end{smallarray} \quad \text{if} \quad \lambda_1 < \lambda_3 \quad \text{or} \\
+\vt \omega(t) &= \begin{smallarray}{c}
+a_1\text{dn}(\omega_p t + \epsilon, k^2) \\
+a_2\text{sn}(\omega_p t + \epsilon, k^2) \\
+a_3\text{cn}(\omega_p t + \epsilon, k^2)
+\end{smallarray} \quad \text{if} \quad \lambda_1 > \lambda_3,
 \end{align}
 \end{subequations}
-where
-\begin{align*}
-&a = -\frac{(I_1-I_3)A_1A_3}{I_2 A_2},\\
-&b = \text{sn}^{-1}\left(\frac{\omega_{2,0}}{A_2},k\right),\quad\text{and}\\
-&k = \frac{A_2}{\text{max}(\lambda_1,\lambda_3)}
-\end{align*}
-
-During a torque-free rotational motion, the angular momentum vector $\vt L$ remains constant in any space-fixed frame of reference.
-
-In the body-fixed frame, although the
+where $\text{sn}(u,m)$, $\text{cn}(u,m)$, and $\text{dn}(u,m)$ are the Jacobi elliptic functions,\cite{Byrd_1971} $k = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ is the elliptic modulus, and the constants $\omega_p$ and $\epsilon$ are given by
+\begin{equation*}
+\omega_p = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
+\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}, k^2\right).
+\end{equation*}
+
+In order to compute the inverse function $\text{sn}^{-1}$, one can use its relation with an elliptic integral in Carlson form, \cite{Carlson_1977} $\text{sn}^{-1}(x,m) = x R_F(1-x^2,1-mx^2,1)$.
+In this work, we employ the routines available in Ref.~\onlinecite{Galassi_2009} for numerically evaluating Jacobi functions and Carlson integrals.
+
+Once $\vt \omega$ is known as a function of time, we can solve Eq.~\ref{eq:EDO_q} with initial condition $\vt q(0) = \vt q_0$.
+The method we propose here, which is closely related to the one pioneered by Kosenko\cite{Kosenko_1998} and extended by Celledoni \textit{et al}.,\cite{Celledoni_2008} relies on the factorization in Eq.~\ref{eq:factorization_of_A} to facilitate the task.
+During a free rotation, any space-fixed representation of the angular momentum vector will remain constant.
+Thus, instead of solving Eq.~\ref{eq:EDO_q} directly for obtaining $\vt q$, we solve an analogous equation for obtaining another unit quaternion $\vt z$ specially chosen so that $\tr{[\mt A(\vt z)]}\mt I\vt \omega = \tr{[\begin{array}{ccc} L & 0 & 0 \end{array}]}$.
+In other words, we look for an alternative space-fixed frame whose first axis is aligned with the angular momentum vector.
+
+\begin{equation*}
+\mt B(\vt z)\mt I\vt \omega = \mt C(\vt z)\begin{smallarray}{c}
+L \\ 0 \\ 0
+\end{smallarray}
+\end{equation*}
+
+By making ${\mt A}(\vt z) = \mt A(\vt z)\mt C(\vt z)$
+
+factorizing $\mt A(\vt z)$ and applying Eq.~\ref{eq:product_B_vector}, we obtain a homogeneous linear system
+\begin{equation*}
+\Big(L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j\Big) \vt z = \vt 0,
+\end{equation*}
+where $\hat{\vt C}_1$ and $\hat{\vt B}_k$ are the permutation matrices introduced in Appendix~\ref{sec:auxiliary_math}.
+Hence, besides having unit norm, the quaternion $\vt z$ will always lie in the null space of a matrix $\mt G = L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j$, which is guaranteed to exist because $\det \mt G = (L^2 - \tr{\vt \omega}{\mt I}^2{\vt \omega})^2 = 0$.
+Such a null space turns out to be a moving hyperplane in $\mathbb{R}^4$ defined by the orthogonal basis vectors
+\begin{equation*}
+\vt v_1(t) = \begin{smallarray}{c}
+\frac{I_3\omega_3(t)}{L-I_1\omega_1(t)} \\
+\frac{I_2\omega_2(t)}{L-I_1\omega_1(t)} \\
+1 \\
+0
+\end{smallarray}
+\quad \text{and} \quad
+\vt v_2(t) = \begin{smallarray}{c}
+\frac{-I_2\omega_2(t)}{L-I_1\omega_1(t)} \\
+\frac{I_3\omega_3(t)}{L-I_1\omega_1(t)} \\
+0 \\
+1
+\end{smallarray}.
+\end{equation*}
+
+At all times, therefore, $\vt z$ will be a linear combination of $\vt v_1$ and $\vt v_2$.
+As these basis vectors are known functions of time, the problem of determining $\vt z(t)$ reduces to finding out how the two coefficients of the combination vary with time.
+Examination of $\vt v_1$ and $\vt v_2$ above makes it clear that these coefficients coincide with the entries $z_3$ and $z_4$ of $\vt z$, in such a way that
+\begin{equation}
+\label{eq:linear_comb_z}
+\vt z(t) = z_3(t) \vt v_1(t) + z_4(t) \vt v_2(t).
+\end{equation}
+
+Note that we are free to choose any unit quaternion in the hyperplane defined by $\vt v_1(0)$ and $\vt v_2(0)$ as the initial condition $\vt z(0) = \vt z_0$.
+In other words, we can select any space-fixed frame whose first axis aligns with the angular momentum vector.
+For convenience, we choose
+\begin{equation*}
+\vt z_0 = \frac{\vt v_1(0)}{\|\vt v_1(0)\|} = \sqrt{\frac{L - I_1\omega_1^0}{2L}} \vt v_1(0).
+\end{equation*}
+
+By substituting Eq.~\ref{eq:linear_comb_z} into $\dot{\vt z} = \frac{1}{2}{\mt B}(\vt z){\vt \omega}$, which is the equivalent of Eq.~\ref{eq:EDO_q} for $\vt z$, we obtain a vector of four differential equations containing $z_3$ and $z_4$, from which we pick the third and fourth ones, which are linearly independent.
+After replacing $(I_2 - I_3)\omega_2\omega_3 = I_1\dot{\omega}_1$ and $I_1\omega_1^2 + I_2\omega_2^2 + I_3\omega_3^2 = 2 K_r$, followed by defining
+\begin{equation*}
+\mt M(t) = \begin{smallarray}{rr} f(t) & -g(t) \\ g(t) & f(t) \end{smallarray} \; \text{with} \;
+\begin{cases}
+f(t) = \frac{-I_1 \dot{\omega}_1(t)}{L - I_1\omega_1(t)} \\
+g(t) = \frac{2K_r - L\omega_1(t)}{L - I_1\omega_1(t)}
+\end{cases},
+\end{equation*}
+we can represent these equations in matrix form as
+\begin{equation*}
+\begin{smallarray}{c} \dot{z}_3 \\ \dot{z}_4 \end{smallarray} = \frac{1}{2} {\mt M}(t) \begin{smallarray}{c} z_3 \\ z_4 \end{smallarray}.
+\end{equation*}
+
+It is fortunate that the product $\mt M(t_a)\mt M(t_b)$ commutes for any $t_a$ and $t_b$, since this is a sufficient condition for a solution to be obtained in terms of matrix exponential $\exp[\frac{1}{2}\int_0^t \mt M(\tau)d\tau]$.
+Such solution is
+\begin{equation*}
+\begin{smallarray}{c} z_3 \\ z_4 \end{smallarray} = \sqrt{\tfrac{L-I_1\omega_1(t)}{L-I_1\omega_1^0}} \begin{smallarray}{rr} \cos \phi(t) & -\sin \phi(t) \\ \sin \phi(t) & \cos \phi(t) \end{smallarray} \begin{smallarray}{c} z_3^0 \\ z_4^0 \end{smallarray},
+\end{equation*}
+where $\phi(t) = \frac{1}{2} \int_0^t g(\tau)d\tau$. Note that this consists in a counterclockwise rotation, followed by a scaling.
+
+\begin{align*}
+z_3(t) &= \sqrt{\tfrac{L-I_1\omega_1(t)}{2L}} \cos \phi(t) \\
+z_4(t) &= \sqrt{\tfrac{L-I_1\omega_1(t)}{2L}} \sin \phi(t)
+\end{align*}
+
+\begin{equation*}
+\phi(t) = \frac{L}{2 I_1\omega_p}\left\{(A_\phi + 1)\omega_p t + A_\phi[h(t) - h(0)]\right\},
+\end{equation*}
+where $A_\phi = \frac{2 I_1 K_r - L^2}{L^2 - I_1^2 a_1^2}$ and $B_\phi = \frac{\max(\lambda_1,\lambda_3)}{\sqrt{\lambda_1^2 + \eta \lambda_3^2}}$, $C_\phi = \frac{\lambda_1\lambda_3}{a_3 \sqrt{\lambda_1^2 + \eta \lambda_3^2}}$.
+The function $h(t)$ is defined as
+\begin{align*}
+h(t) &= \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) + \\
+&+ \frac{I_1}{L} \frac{a_1 a_3}{a_2} C_\phi \arctan\left(C_\phi\frac{\omega_3(t)}{\omega_2(t)}\right)
+\end{align*}
+
+But $\frac{a_1 a_3}{a_2} = - \frac{I_2}{I_1 - I_3} \omega_p$. Thus,
+\begin{align*}
+h(t) &= \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) \\
+&- \frac{I_1 \omega_p}{L}\frac{I_2 C_\phi}{I_1 - I_3} \arctan\left(C_\phi\frac{\omega_3(t)}{\omega_2(t)}\right)
+\end{align*}
+
+Rewriting $g(t)$:
+\begin{equation*}
+g(t) = \frac{L}{I_1} + \frac{2 I_1 K_r - L^2}{L I_1 \omega_p}\frac{\omega_p}{1 + \alpha[\omega_1(t)/a_1]},
+\end{equation*}
+where $I_1 a_1/L$. By defining $u = \omega_p \tau + \epsilon$, we have that $du = \omega_p d\tau$. Therefore, if $\lambda_1 < \lambda_3$:
+\begin{equation*}
+\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1 \omega_p}\int\limits_\epsilon^{\omega_p t + \epsilon} \frac{1}{1 + \alpha \cn u}du
+\end{equation*}
+
+Otherwise, if $\lambda_3 < \lambda_1$:
+\begin{equation*}
+\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1 \omega_p}\int\limits_\epsilon^{\omega_p t + \epsilon} \frac{1}{1 + \alpha \dn u}du
+\end{equation*}
+
+From Ref.~\onlinecite{Byrd_1971}, we deduce that
+\begin{align*}
+\int \frac{1}{1 + \alpha \cn u}du = (\eta + 1) \Bigg[\Pi(\sn u,-\eta,k^2) + \\
++ \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
+\end{align*}
+
+Given that $\cn(u,k^2) = \dn(ku,1/k^2)$, $\dn(u,k^2) = \cn(ku,1/k^2)$, and $\sn(ku,1/k^2) = k\sn(u,k^2)$
+\begin{align*}
+\int \frac{1}{1 + \alpha \dn u}du = \frac{\eta + 1}{k} \Bigg[\Pi(k\sn u,-\eta,1/k^2) + \\
++ \frac{\alpha}{\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
+\end{align*}
+where $\eta = \frac{\alpha^2}{1-\alpha^2}$. Moreover, as $\Pi(x,y,z) = k\Pi(x/k,yk^2,zk^4)$:
+\begin{align*}
+\int \frac{1}{1 + \alpha \dn u}du = (\eta + 1) \Bigg[\Pi(\sn u,-k^2\eta,k^2) + \\
++ \frac{\alpha}{k\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
+\end{align*}
+
+\begin{widetext}
+\begin{equation*}
+\phi(t) = \frac{L}{2 I_1\omega_p}\left\{(A_\phi + 1)\omega_p t + A_\phi [h(t) - h(0)]\right\},
+\end{equation*}
+\begin{equation}
+h(t) = \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) + \alpha B_\phi \arctan\left(B_\phi\frac{\omega_3(t)/a_3}{\omega_2(t)/a_2}\right)
+\end{equation}
+where $A_\phi = \frac{2 I_1 K_r - L^2}{L^2}(\eta + 1)$ and $B_\phi = \frac{\max(\lambda_1,\lambda_3)}{\sqrt{\lambda_1^2 + \eta \lambda_3^2}}$
+
+\begin{equation*}
+\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2 b I_1 L} (\eta + 1) \Bigg[bt + \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 \eta}{\lambda_1^2},k^2\right) + \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
+\end{equation*}
+
+\begin{equation*}
+\phi(t) = \frac{L}{2}\frac{2K_r - I_1 a_1^2}{L^2 - I_1^2 a_1^2}t + \frac{2 I_1 K_r - L^2}{2 b I_1 L} (\eta + 1) \Bigg[bt + \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 \eta}{\lambda_1^2},k^2\right) + \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
+\end{equation*}
+
+\end{widetext}
+where $\alpha = I_1 a_1/L$.
+
+\begin{align*}
+\int \frac{1}{1 + \alpha \cn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-\eta,k^2) + \\
++ \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
+\end{align*}
+
+\begin{align*}
+\int \frac{1}{1 + \alpha \dn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-k^2\eta,k^2) + \\
++ \frac{\alpha}{k\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
+\end{align*}
+
+-----
+
+\begin{equation*}
+\left[\begin{array}{c}
+\dot{z}_3 \\
+\dot{z}_4
+\end{array}\right] =
+\frac{1}{2}\left[\begin{array}{rr}
+f(t) & g(t) \\
+-g(t) & f(t)
+\end{array}\right]
+\left[\begin{array}{c}
+z_3 \\
+z_4
+\end{array}\right],
+\end{equation*}
+where $f(t) = \frac{-I_1 \dot{\omega}_1(t)}{L - I_1\omega_1(t)}$ and $g(t) = \frac{2K_r - L\omega_1(t)}{L - I_1\omega_1(t)}$.
+Because the product $\mt M(t_a)\mt M(t_b)$, where $\mt M(t)$ is the matrix above, commutes for any instants $t_a$ and $t_b$, since this is a sufficient condition for
+\begin{equation*}
+\left[\begin{array}{c}
+z_3 \\
+z_4
+\end{array}\right] =
+\exp \Bigg( \frac{1}{2}\left[\begin{array}{rr}
+f(t) & g(t) \\
+-g(t) & f(t)
+\end{array}\right] \Bigg)
+\left[\begin{array}{c}
+z_3^0 \\
+z_4^0
+\end{array}\right],
+\end{equation*}
+
+Note that
+
+The remaining terms of Eq.~\ref{eq:linear_comb_z} are
+
+\begin{align*}
+f(t) &= \frac{-I_1 \dot{\omega}_1(t)}{L - I_1\omega_1(t)} \\
+g(t) &= \frac{L}{I_1} + \frac{2 I_1 K_r - L^2}{b I_1 L}\frac{b}{1+(I_1 a_1/L)[\omega_1(t)/a_1]}
+\end{align*}
+
+ two linearly independent ones.
+
+Those corresponding to the third and forth elements of $\vt v_1$ and $\vt v_2$
+
+\begin{equation*}
+\dot{\vt z} = \frac{1}{2} {\mt B}(\vt z){\vt \omega} = \frac{1}{2}\left(\sum_{j=1}^3 \omega_j \hat{\mt B}_j\right){\vt z}.
+\end{equation*}
+
+\begin{equation*}
+\dot{z}_3 \vt v_1 + \dot{z}_4 \vt v_2 + z_3 \dot{\vt v}_1 + z_4 \dot{\vt v}_2 = \frac{1}{2}{\mt B}(z_3 \vt v_1 + z_4 \vt v_2){\vt \omega}.
+\end{equation*}
+
+------
+
+, which are, in fact, the elements $z_3(t)$ and $z_4(t)$.
+
+. Simple examination of these vectors makes evident that $\vt z = z_3 \vt v_1 + z_4 \vt v_2$, which means that
+
+the coefficients of such combination clearly coincide with the elements $z_3$ and $z_4$.
+
+\begin{equation*}
+\vt q = \left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right]\tr{\left[\begin{array}{cc}\vt z_0 & \mt C(\vt z_0)\end{array}\right]} \vt q_0.
+\end{equation*}
+
+Any unit quaternion in the hyperplane defined by $\vt v_1(0)$ and $\vt v_2(0)$ can be taken as $\vt z(0) = \vt z_0$.
+
+\begin{equation*}
+\vt z_0 = \frac{1}{\sqrt{2L(L - I_1\omega_1^0)}}
+\left[\begin{array}{c}
+I_3\omega_3^0 \\
+I_2\omega_2^0 \\
+L-I_1\omega_1^0 \\
+0
+\end{array}\right].
+\end{equation*}
+
+, that is, $\vt z = z_3 \vt v_1 + z_4 \vt v_2$.
+
+This reduces the problem to determining
+
+\begin{equation*}
+\vt z = z_3 \vt v_1 + z_4 \vt v_2 =
+\left[\begin{array}{c}
+\frac{I_3\omega_3z_3 - I_2\omega_2z_4}{L-I_1\omega_1} \\
+\frac{I_2\omega_2z_3 + I_3\omega_3z_4}{L-I_1\omega_1} \\
+z_3 \\
+z_4
+\end{array}\right]
+\end{equation*}
+Such null space certainly exists because .
+
+Given that $(I_2 - I_3)\omega_2\omega_3 = I_1 \dot{\omega}_1$ and $\tr{\vt \omega}\mt I\vt \omega = K_r$:
+\begin{align*}
+\dot{z}_3 = \frac{1}{2}\frac{I_1\dot{\omega}_1 z_3 + (2K_r + L\omega_1)z_4}{L + I_1\omega_1} \\
+\dot{z}_4 = \frac{1}{2}\frac{I_1\dot{\omega}_1 z_4 - (2K_r + L\omega_1)z_3}{L + I_1\omega_1}
+\end{align*}
+
+From Eq.~\ref{eq:factorization_of_A}, we observe that $\mt B(\vt z)\mt I\vt \omega = \mt C(\vt z)\tr{[L \; 0 \; 0]}$.
+
+A key aspect of the approach consists in defining is the choice of a temporary space-fixed frame of reference
 
@@
 If $\lambda_1 < \lambda_3$, then $\omega_1(t) = A_1 \text{cn}(\alpha t + \epsilon)$
+
+\begin{align*}
+\int \frac{1}{1 + \alpha \cn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-\eta,k^2) + \\
++ \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
+\end{align*}
+
+\begin{align*}
+\int \frac{1}{1 + \alpha \dn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-k^2\eta,k^2) + \\
++ \frac{\alpha}{k\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
+\end{align*}
+
+\begin{align*}
+\Theta(x,n,m) &= \Pi(x,n,m) - \text{sn}^{-1}(x,m) = \\
+&= \frac{1}{3}nx^3R_J(1-x^2,1-mx^2,1,1-nx^2).
+\end{align*}
 

### 2016-10-28 rigid_bodies.tex
Before removing drafts on exact solution

@@
 
-\subsection{Analytical Solution for Free Rotation}
+\subsection{Exact Solution for Free Rotation}
 \label{sec:free_rotation}
 
-It is not possible to analytically integrate the equations of motion for a general potential $U(\vt r,\vt q)$.
-Nonetheless, closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known for the particular case of torque-free rotation (i.e.
-when $\vt \tau = \vt 0$).
-Free rotations are very useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies, which is the topic addressed in Sec.~\ref{sec:symplectic}.
-As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is completely decoupled from Eq.~\ref{eq:EDO_pi} when $\vt \tau = \vt 0$.
+Eq.~\ref{eq:EDO_system} cannot be integrated analytically for a general potential $U(\vt r,\vt q)$, but closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known in the case when $\vt \tau = \vt 0$.
+Torque-free rotations are useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies.
+This is the topic addressed in Sec.~\ref{sec:symplectic}.
+As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is totally decoupled from Eq.~\ref{eq:EDO_q} when $\vt \tau = \vt 0$.
 It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum vector $\vt L$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation.
-This allows us to analytically solve the Euler equation for the angular velocity vector $\vt \omega$.
-Here we present the solution for an asymmetric top ($I_1 > I_2 > I_3$) in a form that aims to facilitate computer implementation, inspired by the derivations presented in Refs.~\onlinecite{Landau1976} and \onlinecite{Thorn2012}.
-From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, such solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a vector of prefactors defined as
-\begin{equation*}
-\vt a = \begin{smallarray}{c}
-\frac{\omega_1^0}{\sqrt{1 - \left(\omega_2^0/\lambda_1\right)^2}} \\
-\min(\lambda_1,\lambda_3) \\
-\frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}
-\end{smallarray},
-\; \text{where} \;
-\begin{cases}
-\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \\
-\lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}
-\end{cases}.
-\end{equation*}
-
-Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$. This also occurs for $\vt \omega$, whose functional form is
+This allows us to solve the Euler equation analytically for $\vt \omega$.
+In the case of a spherical top ($I_1 = I_2 = I_3$), the solution is simply $\vt \omega(t) = \vt \omega(0)$.
+Inspired by the derivations in Refs.~\onlinecite{Landau1976, Thorn2012, vanZon2007}, we present the general solution in a form that facilitates computer implementation and allows a unified treatment for both asymmetric ($I_1 > I_2 > I_3$) and symmetric tops (either $I_1 = I_2$ or $I_2 = I_3$) if infinities are dealt with properly (IEEE 754 standard).
+From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, the solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a set of prefactors defined as
+\begin{align*}
+a_1 &= \text{sgn}(\omega_1^0)\sqrt{\frac{L^2 - 2 I_3 K_r}{I_1(I_1 - I_3)}}, \\
+a_2 &= \sqrt{\min(\lambda_1,\lambda_3)}, \quad \text{and} \\
+a_3 &= \text{sgn}(\omega_3^0)\sqrt{\frac{2 I_1 K_r - L^2}{I_1(I_1 - I_3)}}, \\
+\end{align*}
+where $\lambda_1 = \frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}$ and $\lambda_3 = \frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}$.
+Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$.
+This also occurs for $\vt \omega$, whose functional form is
 \begin{subequations}
@@
 \vt \omega(t) &= \begin{smallarray}{c}
-a_1\text{cn}(\omega_p t + \epsilon, k^2) \\
-a_2\text{sn}(\omega_p t + \epsilon, k^2) \\
-a_3\text{dn}(\omega_p t + \epsilon, k^2)
+a_1\text{cn}(\omega_p t + \epsilon) \\
+a_2\text{sn}(\omega_p t + \epsilon) \\
+a_3\text{dn}(\omega_p t + \epsilon)
 \end{smallarray} \quad \text{if} \quad \lambda_1 < \lambda_3 \quad \text{or} \\
 \vt \omega(t) &= \begin{smallarray}{c}
-a_1\text{dn}(\omega_p t + \epsilon, k^2) \\
-a_2\text{sn}(\omega_p t + \epsilon, k^2) \\
-a_3\text{cn}(\omega_p t + \epsilon, k^2)
+a_1\text{dn}(\omega_p t + \epsilon) \\
+a_2\text{sn}(\omega_p t + \epsilon) \\
+a_3\text{cn}(\omega_p t + \epsilon)
 \end{smallarray} \quad \text{if} \quad \lambda_1 > \lambda_3,
@@
 \end{subequations}
-where $\text{sn}(u,m)$, $\text{cn}(u,m)$, and $\text{dn}(u,m)$ are the Jacobi elliptic functions,\cite{Byrd_1971} $k = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ is the elliptic modulus, and the constants $\omega_p$ and $\epsilon$ are given by
+where $\text{sn}$, $\text{cn}$, and $\text{dn}$ are the Jacobi elliptic functions\cite{Byrd_1971} with parameter $m = k^2 = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ and the constants $\omega_p$ and $\epsilon$ are given by
 \begin{equation*}
 \omega_p = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
-\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}, k^2\right).
-\end{equation*}
-
-In order to compute the inverse function $\text{sn}^{-1}$, one can use its relation with an elliptic integral in Carlson form, \cite{Carlson_1977} $\text{sn}^{-1}(x,m) = x R_F(1-x^2,1-mx^2,1)$.
-In this work, we employ the routines available in Ref.~\onlinecite{Galassi_2009} for numerically evaluating Jacobi functions and Carlson integrals.
-
-Once $\vt \omega$ is known as a function of time, we can solve Eq.~\ref{eq:EDO_q} with initial condition $\vt q(0) = \vt q_0$.
-The method we propose here, which is closely related to the one pioneered by Kosenko\cite{Kosenko_1998} and extended by Celledoni \textit{et al}.,\cite{Celledoni_2008} relies on the factorization in Eq.~\ref{eq:factorization_of_A} to facilitate the task.
+\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}\right).
+\end{equation*}
+
+In order to compute the inverse function $\text{sn}^{-1}$, one can resort to its relation with an elliptic integral in Carlson form,\cite{Carlson_1977, Gradshteyn2014} $\text{sn}^{-1}(x) = x R_F(1-x^2,1-mx^2,1)$.
+One can numerically evaluate Jacobi functions and Carlson integrals by using, for instance, the routines available in the GNU Scientific Library.\cite{Galassi_2009}
+
+Once $\vt \omega(t)$ is known, we can solve Eq.~\ref{eq:EDO_q} for the unit quaternion $\vt q$ with initial condition $\vt q(0) = \vt q_0$.
+Here we propose a new procedure that is closely linked to the methods by Kosenko\cite{Kosenko_1998} and by Celledoni \textit{et al}.,\cite{Celledoni_2008} but that relies on the rotation matrix factorization (Eq.~\ref{eq:factorization_of_A}) to facilitate the task.
+These methods are also related to the one by van Zon and Schofield,\cite{vanZon2007} as shown in Ref.~\onlinecite{Celledoni_2008}.
 During a free rotation, any space-fixed representation of the angular momentum vector will remain constant.
-Thus, instead of solving Eq.~\ref{eq:EDO_q} directly for obtaining $\vt q$, we solve an analogous equation for obtaining another unit quaternion $\vt z$ specially chosen so that $\tr{[\mt A(\vt z)]}\mt I\vt \omega = \tr{[\begin{array}{ccc} L & 0 & 0 \end{array}]}$.
+Thus, instead of solving Eq.~\ref{eq:EDO_q} directly for obtaining $\vt q(t)$, we solve an analogous equation $\dot{\vt z} = \frac{1}{2}{\mt B(\vt z)}{\vt \omega}$ for determining a unit quaternion $\vt z(t)$ specially chosen so that $\tr{\mt A\!}(\vt z)\mt I\vt \omega = \tr{[\begin{array}{ccc} L & 0 & 0 \end{array}]}$.
 In other words, we look for an alternative space-fixed frame whose first axis is aligned with the angular momentum vector.
-
-\begin{equation*}
-\mt B(\vt z)\mt I\vt \omega = \mt C(\vt z)\begin{smallarray}{c}
-L \\ 0 \\ 0
-\end{smallarray}
-\end{equation*}
-
-By making ${\mt A}(\vt z) = \mt A(\vt z)\mt C(\vt z)$
-
-factorizing $\mt A(\vt z)$ and applying Eq.~\ref{eq:product_B_vector}, we obtain a homogeneous linear system
-\begin{equation*}
-\Big(L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j\Big) \vt z = \vt 0,
-\end{equation*}
-where $\hat{\vt C}_1$ and $\hat{\vt B}_k$ are the permutation matrices introduced in Appendix~\ref{sec:auxiliary_math}.
-Hence, besides having unit norm, the quaternion $\vt z$ will always lie in the null space of a matrix $\mt G = L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j$, which is guaranteed to exist because $\det \mt G = (L^2 - \tr{\vt \omega}{\mt I}^2{\vt \omega})^2 = 0$.
-Such a null space turns out to be a moving hyperplane in $\mathbb{R}^4$ defined by the orthogonal basis vectors
-\begin{equation*}
-\vt v_1(t) = \begin{smallarray}{c}
-\frac{I_3\omega_3(t)}{L-I_1\omega_1(t)} \\
-\frac{I_2\omega_2(t)}{L-I_1\omega_1(t)} \\
-1 \\
-0
-\end{smallarray}
+After replacing $\tr{\mt A\!}(\vt z) = \tr{\mt C}(\vt z){\mt B}(\vt z)$, we premultiply both sides by $\mt C(\vt z)$ and then apply Eq.~\ref{eq:product_B_vector} to obtain the homogeneous linear system $\mt G \vt z = \vt 0$, where $\mt G = L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j$.
+This means that $\vt z$ will continuously lie in the kernel of matrix $\mt G$, which can be determined by a standard elimination method.
+Such kernel turns out to be a (moving) linear space defined by the orthogonal basis vectors
+\begin{equation*}
+\vt u(t) = \begin{smallarray}{c} I_3\omega_3(t) \\ I_2\omega_2(t) \\ L-I_1\omega_1(t) \\ 0 \end{smallarray}
 \quad \text{and} \quad
-\vt v_2(t) = \begin{smallarray}{c}
-\frac{-I_2\omega_2(t)}{L-I_1\omega_1(t)} \\
-\frac{I_3\omega_3(t)}{L-I_1\omega_1(t)} \\
-0 \\
-1
-\end{smallarray}.
-\end{equation*}
-
-At all times, therefore, $\vt z$ will be a linear combination of $\vt v_1$ and $\vt v_2$.
+\vt v(t) = \begin{smallarray}{c} -I_2\omega_2(t) \\ I_3\omega_3(t) \\ 0 \\ L-I_1\omega_1(t) \end{smallarray}.
+\end{equation*}
+
+At all times, therefore, $\vt z$ will be a linear combination of $\vt u$ and $\vt v$. We can then write the expression
+\begin{equation*}
+\vt z(t) = \frac{b(t) \vt u(t) + c(t) \vt v(t)}{\sqrt{2L[L-I_1\omega_1(t)]}},
+\end{equation*}
+whose denominator is the norm of both $\vt u$ and $\vt v$.
+The problem has thus been reduced to determining how the coefficients $b$ and $c$ vary with time.
+Further reduction is achieved by realizing that both the unit-norm constraint ($\tr{\vt z}{\vt z} = 1$) and the orthogonality condition ($\tr{\vt u}{\vt v} = 0$) hold at all times, which makes
+\begin{equation*}
+\tr{\vt z}{\vt z} = \frac{b^2 \tr{\vt u}{\vt u} + 2bc\tr{\vt u}{\vt v} + c^2 \tr{\vt v}{\vt v}}{2L[L-I_1\omega_1(t)]} = b^2 + c^2 = 1.
+\end{equation*}
+
+Therefore, the values of $b$ and $c$ are restricted to a unit circle, so that we can finally rewrite $\vt z(t)$ as
+\begin{equation*}
+\vt z(t) = \frac{\vt u_1(t) \cos \theta(t) + \vt u_2(t) \sin \theta(t)}{\sqrt{2L[L-I_1\omega_1(t)]}}.
+\end{equation*}
+
+$\dot{\theta} = \frac{2K_r - L\omega_1(t)}{2[L - I_1\omega_1(t)]}$
+
+\begin{equation*}
+\theta = \frac{L}{2I_1}(t - t_0) + \frac{2 I_1 K_r - L^2}{2 L I_1\omega_p} \int_{t_0}^t \frac{\omega_p}{1 - \alpha [\omega_1(s)/a_1]}ds,
+\end{equation*}
+where $\alpha = I_1a_1/L$.
+
+From Ref.~\onlinecite{Byrd_1971}, we deduce that
+\begin{gather*}
+\int \tfrac{1 - \alpha^2}{1 - \alpha \cn u}du = \Pi(\sn u,-\eta,k^2) + \frac{\alpha\arctan\left(\frac{\sn u\sqrt{k^2+\eta}}{\dn u}\right)}{\sqrt{k^2+\eta}}
+\end{gather*}
+
+\begin{gather*}
+\int \tfrac{1 - \alpha^2}{1 - \alpha \dn u}du = \Pi(\sn u,-k^2\eta,k^2)
++ \frac{\alpha\arctan\left(\frac{\sn u\sqrt{1 + k^2\eta}}{\cn u}\right)}{\sqrt{1 + k^2\eta}}
+\end{gather*}
+
+Finally, by defining
+\begin{equation*}
+\eta = \frac{(\alpha {a_2}/{\lambda_1})^2}{1-\alpha^2}
+\quad \text{and} \quad
+\beta = \sqrt{\eta + ({a_2}/{\lambda_3})^2}
+\end{equation*}
+
+We have
+\begin{equation*}
+\theta = \frac{L}{2I_1\omega_p}(\omega_p t + \epsilon) + \frac{2 I_1 K_r - L^2}{2LI_1\omega_p(1-\alpha^2)} \int_0^t \frac{(1-\alpha^2)\omega_p}{1 - \alpha [\omega_1(\tau)/a_1]}d\tau,
+\end{equation*}
+
+\begin{equation*}
+\theta = \frac{L}{2I_1} t + \frac{2 I_1 K_r - L^2}{2LI_1\omega_p(1-\alpha^2)} [F(\omega_p t + \epsilon) - F(\epsilon)],
+\end{equation*}
+
+\begin{equation*}
+\theta = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2LI_1\omega_p(1-\alpha^2)} \left[\omega_p t + h(t) - h(0)\right],
+\end{equation*}
+where
+\begin{equation*}
+h = \Theta\left({\omega_2}/{a_2},-\eta,k^2\right) + \frac{\alpha}{\beta} \arctan\left( \beta \frac{\omega_2/a_2}{\omega_3/a_3} \right)
+\end{equation*}
+
+\subsection{Integrals}
+
+\begin{gather*}
+C(u) = \int_0^u \frac{1 - \alpha^2}{1 - \alpha \cn x}dx = \\
+C(u) = u + \text{sgn}(\cn u) \Theta(\sn u,\eta) + 2\Theta(1,\eta) \left\lfloor \frac{u}{2 K} \right\rceil + \\
++ \frac{\alpha}{\sqrt{k^2+\eta}} \arctan\left(\frac{\sqrt{k^2+\eta} \; \sn u}{\dn u}\right)
+\end{gather*}
+
+\begin{gather*}
+D(u) = \int_0^u \frac{1 - \alpha^2}{1 - \alpha \dn x}dx = \\
+F_\dn(u) = u + \text{sgn}(\cn u) \Theta(\sn u,k^2\eta) + 2\Theta(1,k^2\eta) \left\lfloor\frac{u}{2 K}\right\rceil + \\
++ \frac{\alpha}{\sqrt{1 + k^2\eta}} \left[\arctan\left(\frac{\sqrt{1 + k^2\eta} \; \sn u}{\cn u}\right) + \pi \left\lfloor \frac{u}{2 K} \right\rceil\right]
+\end{gather*}
+
+\begin{align*}
+\Theta(x,n,m) &= \Pi(x,-n,m) - F(x,m) = \\
+&= -\frac{1}{3}nx^3R_J(1-x^2,1-mx^2,1,1+nx^2).
+\end{align*}
+
+\subsection{DRAFT}
+
 As these basis vectors are known functions of time, the problem of determining $\vt z(t)$ reduces to finding out how the two coefficients of the combination vary with time.
@@
 \subsection{New Results}
+
+\begin{table*}
+	\caption{Results with Exact Solution for Free Rotations}
+	\label{table:NVE}
+	\begin{ruledtabular}
+		\begin{tabular}{cccccccc}
+			$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+			\hline
+			1.0 & $	305.91	\pm	0.03	$ & $	-8395.1	\pm	0.1	$ & $	1645.9	\pm	0.1	$ & $	823.2	\pm	0.1	$ & $	822.7	\pm	0.1	$ & $	146	\pm	2	$ & $0.0456$ \\
+			2.0 & $	305.21	\pm	0.02	$ & $	-8389.5	\pm	0.1	$ & $	1642.1	\pm	0.1	$ & $	823.0	\pm	0.1	$ & $	819.1	\pm	0.1	$ & $	149	\pm	2	$ & $0.656$ \\
+			3.0 & $	304.30	\pm	0.03	$ & $	-8375.4	\pm	0.9	$ & $	1637.2	\pm	0.1	$ & $	824.0	\pm	0.1	$ & $	813.3	\pm	0.1	$ & $	156	\pm	2	$ & $2.87$ \\
+			4.0 & $	304.5	\pm	0.5	$ & $	-8339	\pm	6	$ & $	1638	\pm	3	$ & $	829	\pm	1	$ & $	808.8	\pm	0.6	$ & $	180	\pm	2	$ & $10.1$ \\
+		\end{tabular}
+	\end{ruledtabular}
+\end{table*}
 

### 2016-11-01 rigid_bodies.tex
Inclusion of definite integrals plot

@@
 In the case of a spherical top ($I_1 = I_2 = I_3$), the solution is simply $\vt \omega(t) = \vt \omega(0)$.
-Inspired by the derivations in Refs.~\onlinecite{Landau1976, Thorn2012, vanZon2007}, we present the general solution in a form that facilitates computer implementation and allows a unified treatment for both asymmetric ($I_1 > I_2 > I_3$) and symmetric tops (either $I_1 = I_2$ or $I_2 = I_3$) if infinities are dealt with properly (IEEE 754 standard).
+Inspired by the derivations in Refs.~\onlinecite{Landau1976, Thorn2012, vanZon2007}, we present the general solution in a form that facilitates computer implementation and allows a unified treatment for both asymmetric ($I_1 > I_2 > I_3$) and symmetric tops (either $I_1 = I_2$ or $I_2 = I_3$) if infinities are properly dealt with (IEEE 754 standard).
 From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, the solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a set of prefactors defined as
@@
 whose denominator is the norm of both $\vt u$ and $\vt v$.
-The problem has thus been reduced to determining how the coefficients $b$ and $c$ vary with time.
+The problem has now been reduced to determining how the coefficients $b$ and $c$ vary with time.
 Further reduction is achieved by realizing that both the unit-norm constraint ($\tr{\vt z}{\vt z} = 1$) and the orthogonality condition ($\tr{\vt u}{\vt v} = 0$) hold at all times, which makes
@@
 
-Therefore, the values of $b$ and $c$ are restricted to a unit circle, so that we can finally rewrite $\vt z(t)$ as
+Hence, the values of $b$ and $c$ are restricted to a unit circle, so that we can finally rewrite $\vt z(t)$ as
 \begin{equation*}
-\vt z(t) = \frac{\vt u_1(t) \cos \theta(t) + \vt u_2(t) \sin \theta(t)}{\sqrt{2L[L-I_1\omega_1(t)]}}.
+\vt z(t) = \frac{\vt u(t) \cos \theta(t) + \vt v(t) \sin \theta(t)}{\sqrt{2L[L-I_1\omega_1(t)]}}.
 \end{equation*}
 
-$\dot{\theta} = \frac{2K_r - L\omega_1(t)}{2[L - I_1\omega_1(t)]}$
-
+The next step consists in replacing this result into the equation $\dot{\vt z} = \frac{1}{2}{\mt B(\vt z)}{\vt \omega}$.
+This yields, after some algebra, a single differential equation whose solution can be expressed as
 \begin{equation*}
-\theta = \frac{L}{2I_1}(t - t_0) + \frac{2 I_1 K_r - L^2}{2 L I_1\omega_p} \int_{t_0}^t \frac{\omega_p}{1 - \alpha [\omega_1(s)/a_1]}ds,
+\theta(t) = \theta_0 + \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1}\int_0^t \frac{1}{1-(I_1/L)\omega_1(s)}ds.
 \end{equation*}
-where $\alpha = I_1a_1/L$.
-
-From Ref.~\onlinecite{Byrd_1971}, we deduce that
-\begin{gather*}
-\int \tfrac{1 - \alpha^2}{1 - \alpha \cn u}du = \Pi(\sn u,-\eta,k^2) + \frac{\alpha\arctan\left(\frac{\sn u\sqrt{k^2+\eta}}{\dn u}\right)}{\sqrt{k^2+\eta}}
-\end{gather*}
-
-\begin{gather*}
-\int \tfrac{1 - \alpha^2}{1 - \alpha \dn u}du = \Pi(\sn u,-k^2\eta,k^2)
-+ \frac{\alpha\arctan\left(\frac{\sn u\sqrt{1 + k^2\eta}}{\cn u}\right)}{\sqrt{1 + k^2\eta}}
-\end{gather*}
-
-Finally, by defining
-\begin{equation*}
-\eta = \frac{(\alpha {a_2}/{\lambda_1})^2}{1-\alpha^2}
-\quad \text{and} \quad
-\beta = \sqrt{\eta + ({a_2}/{\lambda_3})^2}
-\end{equation*}
-
-We have
-\begin{equation*}
-\theta = \frac{L}{2I_1\omega_p}(\omega_p t + \epsilon) + \frac{2 I_1 K_r - L^2}{2LI_1\omega_p(1-\alpha^2)} \int_0^t \frac{(1-\alpha^2)\omega_p}{1 - \alpha [\omega_1(\tau)/a_1]}d\tau,
-\end{equation*}
-
-\begin{equation*}
-\theta = \frac{L}{2I_1} t + \frac{2 I_1 K_r - L^2}{2LI_1\omega_p(1-\alpha^2)} [F(\omega_p t + \epsilon) - F(\epsilon)],
-\end{equation*}
-
-\begin{equation*}
-\theta = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2LI_1\omega_p(1-\alpha^2)} \left[\omega_p t + h(t) - h(0)\right],
-\end{equation*}
-where
-\begin{equation*}
-h = \Theta\left({\omega_2}/{a_2},-\eta,k^2\right) + \frac{\alpha}{\beta} \arctan\left( \beta \frac{\omega_2/a_2}{\omega_3/a_3} \right)
-\end{equation*}
-
-\subsection{Integrals}
-
-\begin{gather*}
-C(u) = \int_0^u \frac{1 - \alpha^2}{1 - \alpha \cn x}dx = \\
-C(u) = u + \text{sgn}(\cn u) \Theta(\sn u,\eta) + 2\Theta(1,\eta) \left\lfloor \frac{u}{2 K} \right\rceil + \\
-+ \frac{\alpha}{\sqrt{k^2+\eta}} \arctan\left(\frac{\sqrt{k^2+\eta} \; \sn u}{\dn u}\right)
-\end{gather*}
-
-\begin{gather*}
-D(u) = \int_0^u \frac{1 - \alpha^2}{1 - \alpha \dn x}dx = \\
-F_\dn(u) = u + \text{sgn}(\cn u) \Theta(\sn u,k^2\eta) + 2\Theta(1,k^2\eta) \left\lfloor\frac{u}{2 K}\right\rceil + \\
-+ \frac{\alpha}{\sqrt{1 + k^2\eta}} \left[\arctan\left(\frac{\sqrt{1 + k^2\eta} \; \sn u}{\cn u}\right) + \pi \left\lfloor \frac{u}{2 K} \right\rceil\right]
-\end{gather*}
-
-\begin{align*}
-\Theta(x,n,m) &= \Pi(x,-n,m) - F(x,m) = \\
-&= -\frac{1}{3}nx^3R_J(1-x^2,1-mx^2,1,1+nx^2).
-\end{align*}
-
-\subsection{DRAFT}
-
-As these basis vectors are known functions of time, the problem of determining $\vt z(t)$ reduces to finding out how the two coefficients of the combination vary with time.
-Examination of $\vt v_1$ and $\vt v_2$ above makes it clear that these coefficients coincide with the entries $z_3$ and $z_4$ of $\vt z$, in such a way that
-\begin{equation}
-\label{eq:linear_comb_z}
-\vt z(t) = z_3(t) \vt v_1(t) + z_4(t) \vt v_2(t).
-\end{equation}
-
-Note that we are free to choose any unit quaternion in the hyperplane defined by $\vt v_1(0)$ and $\vt v_2(0)$ as the initial condition $\vt z(0) = \vt z_0$.
-In other words, we can select any space-fixed frame whose first axis aligns with the angular momentum vector.
-For convenience, we choose
-\begin{equation*}
-\vt z_0 = \frac{\vt v_1(0)}{\|\vt v_1(0)\|} = \sqrt{\frac{L - I_1\omega_1^0}{2L}} \vt v_1(0).
-\end{equation*}
-
-By substituting Eq.~\ref{eq:linear_comb_z} into $\dot{\vt z} = \frac{1}{2}{\mt B}(\vt z){\vt \omega}$, which is the equivalent of Eq.~\ref{eq:EDO_q} for $\vt z$, we obtain a vector of four differential equations containing $z_3$ and $z_4$, from which we pick the third and fourth ones, which are linearly independent.
-After replacing $(I_2 - I_3)\omega_2\omega_3 = I_1\dot{\omega}_1$ and $I_1\omega_1^2 + I_2\omega_2^2 + I_3\omega_3^2 = 2 K_r$, followed by defining
-\begin{equation*}
-\mt M(t) = \begin{smallarray}{rr} f(t) & -g(t) \\ g(t) & f(t) \end{smallarray} \; \text{with} \;
-\begin{cases}
-f(t) = \frac{-I_1 \dot{\omega}_1(t)}{L - I_1\omega_1(t)} \\
-g(t) = \frac{2K_r - L\omega_1(t)}{L - I_1\omega_1(t)}
-\end{cases},
-\end{equation*}
-we can represent these equations in matrix form as
-\begin{equation*}
-\begin{smallarray}{c} \dot{z}_3 \\ \dot{z}_4 \end{smallarray} = \frac{1}{2} {\mt M}(t) \begin{smallarray}{c} z_3 \\ z_4 \end{smallarray}.
-\end{equation*}
-
-It is fortunate that the product $\mt M(t_a)\mt M(t_b)$ commutes for any $t_a$ and $t_b$, since this is a sufficient condition for a solution to be obtained in terms of matrix exponential $\exp[\frac{1}{2}\int_0^t \mt M(\tau)d\tau]$.
-Such solution is
-\begin{equation*}
-\begin{smallarray}{c} z_3 \\ z_4 \end{smallarray} = \sqrt{\tfrac{L-I_1\omega_1(t)}{L-I_1\omega_1^0}} \begin{smallarray}{rr} \cos \phi(t) & -\sin \phi(t) \\ \sin \phi(t) & \cos \phi(t) \end{smallarray} \begin{smallarray}{c} z_3^0 \\ z_4^0 \end{smallarray},
-\end{equation*}
-where $\phi(t) = \frac{1}{2} \int_0^t g(\tau)d\tau$. Note that this consists in a counterclockwise rotation, followed by a scaling.
-
-\begin{align*}
-z_3(t) &= \sqrt{\tfrac{L-I_1\omega_1(t)}{2L}} \cos \phi(t) \\
-z_4(t) &= \sqrt{\tfrac{L-I_1\omega_1(t)}{2L}} \sin \phi(t)
-\end{align*}
-
-\begin{equation*}
-\phi(t) = \frac{L}{2 I_1\omega_p}\left\{(A_\phi + 1)\omega_p t + A_\phi[h(t) - h(0)]\right\},
-\end{equation*}
-where $A_\phi = \frac{2 I_1 K_r - L^2}{L^2 - I_1^2 a_1^2}$ and $B_\phi = \frac{\max(\lambda_1,\lambda_3)}{\sqrt{\lambda_1^2 + \eta \lambda_3^2}}$, $C_\phi = \frac{\lambda_1\lambda_3}{a_3 \sqrt{\lambda_1^2 + \eta \lambda_3^2}}$.
-The function $h(t)$ is defined as
-\begin{align*}
-h(t) &= \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) + \\
-&+ \frac{I_1}{L} \frac{a_1 a_3}{a_2} C_\phi \arctan\left(C_\phi\frac{\omega_3(t)}{\omega_2(t)}\right)
-\end{align*}
-
-But $\frac{a_1 a_3}{a_2} = - \frac{I_2}{I_1 - I_3} \omega_p$. Thus,
-\begin{align*}
-h(t) &= \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) \\
-&- \frac{I_1 \omega_p}{L}\frac{I_2 C_\phi}{I_1 - I_3} \arctan\left(C_\phi\frac{\omega_3(t)}{\omega_2(t)}\right)
-\end{align*}
-
-Rewriting $g(t)$:
-\begin{equation*}
-g(t) = \frac{L}{I_1} + \frac{2 I_1 K_r - L^2}{L I_1 \omega_p}\frac{\omega_p}{1 + \alpha[\omega_1(t)/a_1]},
-\end{equation*}
-where $I_1 a_1/L$. By defining $u = \omega_p \tau + \epsilon$, we have that $du = \omega_p d\tau$. Therefore, if $\lambda_1 < \lambda_3$:
-\begin{equation*}
-\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1 \omega_p}\int\limits_\epsilon^{\omega_p t + \epsilon} \frac{1}{1 + \alpha \cn u}du
-\end{equation*}
-
-Otherwise, if $\lambda_3 < \lambda_1$:
-\begin{equation*}
-\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1 \omega_p}\int\limits_\epsilon^{\omega_p t + \epsilon} \frac{1}{1 + \alpha \dn u}du
-\end{equation*}
-
-From Ref.~\onlinecite{Byrd_1971}, we deduce that
-\begin{align*}
-\int \frac{1}{1 + \alpha \cn u}du = (\eta + 1) \Bigg[\Pi(\sn u,-\eta,k^2) + \\
-+ \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
-\end{align*}
-
-Given that $\cn(u,k^2) = \dn(ku,1/k^2)$, $\dn(u,k^2) = \cn(ku,1/k^2)$, and $\sn(ku,1/k^2) = k\sn(u,k^2)$
-\begin{align*}
-\int \frac{1}{1 + \alpha \dn u}du = \frac{\eta + 1}{k} \Bigg[\Pi(k\sn u,-\eta,1/k^2) + \\
-+ \frac{\alpha}{\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
-\end{align*}
-where $\eta = \frac{\alpha^2}{1-\alpha^2}$. Moreover, as $\Pi(x,y,z) = k\Pi(x/k,yk^2,zk^4)$:
-\begin{align*}
-\int \frac{1}{1 + \alpha \dn u}du = (\eta + 1) \Bigg[\Pi(\sn u,-k^2\eta,k^2) + \\
-+ \frac{\alpha}{k\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
-\end{align*}
+
+In order to solve this equation, we introduce the following functions derived from definite integrals: $C(x) = \int_0^x \frac{1}{1 - \alpha \cn y}dy$ and $D(x) = \int_0^x \frac{1}{1 - \alpha \dn y}dy$, where $\alpha = I_1a_1/L$.
 
 \begin{widetext}
-\begin{equation*}
-\phi(t) = \frac{L}{2 I_1\omega_p}\left\{(A_\phi + 1)\omega_p t + A_\phi [h(t) - h(0)]\right\},
-\end{equation*}
-\begin{equation}
-h(t) = \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 }{\lambda_1^2}\eta,k^2\right) + \alpha B_\phi \arctan\left(B_\phi\frac{\omega_3(t)/a_3}{\omega_2(t)/a_2}\right)
-\end{equation}
-where $A_\phi = \frac{2 I_1 K_r - L^2}{L^2}(\eta + 1)$ and $B_\phi = \frac{\max(\lambda_1,\lambda_3)}{\sqrt{\lambda_1^2 + \eta \lambda_3^2}}$
-
-\begin{equation*}
-\phi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2 b I_1 L} (\eta + 1) \Bigg[bt + \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 \eta}{\lambda_1^2},k^2\right) + \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
-\end{equation*}
-
-\begin{equation*}
-\phi(t) = \frac{L}{2}\frac{2K_r - I_1 a_1^2}{L^2 - I_1^2 a_1^2}t + \frac{2 I_1 K_r - L^2}{2 b I_1 L} (\eta + 1) \Bigg[bt + \Theta\left(\frac{\omega_2(t)}{a_2},-\frac{a_2^2 \eta}{\lambda_1^2},k^2\right) + \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
-\end{equation*}
-
+	\begin{subequations}
+\begin{gather}
+\label{eq:def_integral_C} \frac{C(x)}{\eta + 1} = x + \text{sgn}(\cn x) \Theta(\sn x,\eta) + 2\Theta(1,\eta) \left\lfloor \frac{x}{2 K} \right\rceil +
+\frac{\alpha}{\sqrt{m+\eta}} \arctan\left(\frac{\sqrt{m+\eta} \; \sn x}{\dn x}\right) \\
+\label{eq:def_integral_D} \frac{D(x)}{\eta + 1} = x + \text{sgn}(\cn x) \Theta(\sn x,m\eta) + 2\Theta(1,m\eta) \left\lfloor\frac{x}{2 K}\right\rceil +
+\frac{\alpha}{\sqrt{m\eta + 1}} \left\{\arctan\left(\frac{\sqrt{m\eta + 1} \; \sn x}{\cn x}\right) + \pi \left\lfloor \frac{x}{2 K} \right\rceil\right\}
+\end{gather}
+	\end{subequations}
 \end{widetext}
-where $\alpha = I_1 a_1/L$.
-
-\begin{align*}
-\int \frac{1}{1 + \alpha \cn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-\eta,k^2) + \\
-+ \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
-\end{align*}
-
-\begin{align*}
-\int \frac{1}{1 + \alpha \dn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-k^2\eta,k^2) + \\
-+ \frac{\alpha}{k\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
-\end{align*}
-
------
-
-\begin{equation*}
-\left[\begin{array}{c}
-\dot{z}_3 \\
-\dot{z}_4
-\end{array}\right] =
-\frac{1}{2}\left[\begin{array}{rr}
-f(t) & g(t) \\
--g(t) & f(t)
-\end{array}\right]
-\left[\begin{array}{c}
-z_3 \\
-z_4
-\end{array}\right],
-\end{equation*}
-where $f(t) = \frac{-I_1 \dot{\omega}_1(t)}{L - I_1\omega_1(t)}$ and $g(t) = \frac{2K_r - L\omega_1(t)}{L - I_1\omega_1(t)}$.
-Because the product $\mt M(t_a)\mt M(t_b)$, where $\mt M(t)$ is the matrix above, commutes for any instants $t_a$ and $t_b$, since this is a sufficient condition for
-\begin{equation*}
-\left[\begin{array}{c}
-z_3 \\
-z_4
-\end{array}\right] =
-\exp \Bigg( \frac{1}{2}\left[\begin{array}{rr}
-f(t) & g(t) \\
--g(t) & f(t)
-\end{array}\right] \Bigg)
-\left[\begin{array}{c}
-z_3^0 \\
-z_4^0
-\end{array}\right],
-\end{equation*}
-
-Note that
-
-The remaining terms of Eq.~\ref{eq:linear_comb_z} are
-
-\begin{align*}
-f(t) &= \frac{-I_1 \dot{\omega}_1(t)}{L - I_1\omega_1(t)} \\
-g(t) &= \frac{L}{I_1} + \frac{2 I_1 K_r - L^2}{b I_1 L}\frac{b}{1+(I_1 a_1/L)[\omega_1(t)/a_1]}
-\end{align*}
-
- two linearly independent ones.
-
-Those corresponding to the third and forth elements of $\vt v_1$ and $\vt v_2$
-
-\begin{equation*}
-\dot{\vt z} = \frac{1}{2} {\mt B}(\vt z){\vt \omega} = \frac{1}{2}\left(\sum_{j=1}^3 \omega_j \hat{\mt B}_j\right){\vt z}.
-\end{equation*}
-
-\begin{equation*}
-\dot{z}_3 \vt v_1 + \dot{z}_4 \vt v_2 + z_3 \dot{\vt v}_1 + z_4 \dot{\vt v}_2 = \frac{1}{2}{\mt B}(z_3 \vt v_1 + z_4 \vt v_2){\vt \omega}.
-\end{equation*}
-
-------
-
-, which are, in fact, the elements $z_3(t)$ and $z_4(t)$.
-
-. Simple examination of these vectors makes evident that $\vt z = z_3 \vt v_1 + z_4 \vt v_2$, which means that
-
-the coefficients of such combination clearly coincide with the elements $z_3$ and $z_4$.
-
-\begin{equation*}
-\vt q = \left[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}\right]\tr{\left[\begin{array}{cc}\vt z_0 & \mt C(\vt z_0)\end{array}\right]} \vt q_0.
-\end{equation*}
-
-Any unit quaternion in the hyperplane defined by $\vt v_1(0)$ and $\vt v_2(0)$ can be taken as $\vt z(0) = \vt z_0$.
-
-\begin{equation*}
-\vt z_0 = \frac{1}{\sqrt{2L(L - I_1\omega_1^0)}}
-\left[\begin{array}{c}
-I_3\omega_3^0 \\
-I_2\omega_2^0 \\
-L-I_1\omega_1^0 \\
-0
-\end{array}\right].
-\end{equation*}
-
-, that is, $\vt z = z_3 \vt v_1 + z_4 \vt v_2$.
-
-This reduces the problem to determining
-
-\begin{equation*}
-\vt z = z_3 \vt v_1 + z_4 \vt v_2 =
-\left[\begin{array}{c}
-\frac{I_3\omega_3z_3 - I_2\omega_2z_4}{L-I_1\omega_1} \\
-\frac{I_2\omega_2z_3 + I_3\omega_3z_4}{L-I_1\omega_1} \\
-z_3 \\
-z_4
-\end{array}\right]
-\end{equation*}
-Such null space certainly exists because .
-
-Given that $(I_2 - I_3)\omega_2\omega_3 = I_1 \dot{\omega}_1$ and $\tr{\vt \omega}\mt I\vt \omega = K_r$:
-\begin{align*}
-\dot{z}_3 = \frac{1}{2}\frac{I_1\dot{\omega}_1 z_3 + (2K_r + L\omega_1)z_4}{L + I_1\omega_1} \\
-\dot{z}_4 = \frac{1}{2}\frac{I_1\dot{\omega}_1 z_4 - (2K_r + L\omega_1)z_3}{L + I_1\omega_1}
-\end{align*}
-
-From Eq.~\ref{eq:factorization_of_A}, we observe that $\mt B(\vt z)\mt I\vt \omega = \mt C(\vt z)\tr{[L \; 0 \; 0]}$.
-
-A key aspect of the approach consists in defining is the choice of a temporary space-fixed frame of reference
-
-Let $\vt L = \tr{[\mt A(\vt q)]}{\mt I}{\vt \omega}$ be the angular momentum vector
-
-$\vt L = \tr{[\mt A(\vt q)]}{\mt I}{\vt \omega}$ remains constant while both the quaternion $\vt q$ and the body-fixed angular velocity $\vt \omega$ vary.
-
- Once $\vt L$ is computed from the initial condition and the evolution of $\vt \omega$ is known (after solving the Euler equation), this fact constrains the possible evolution of $\vt q$.
-By premultiplying $\vt L$ by $\vt C$, one has that $\vt C \vt L = \vt B \mt I \vt \omega$.
-In other words, $\vt q$ must vary in such a way that the mapping of both $\vt L$ and $\mt I\vt \omega$ in its orthogonal plane always coincide.
-Then, resorting to Eq.~\ref{eq:product_B_vector} we can write
-\begin{equation*}
-\sum_{k=1}^3\left(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k\right){\vt q} = \vt 0,
-\end{equation*}
-where $\hat{\vt C}_k$ and $\hat{\vt B}_k$ are the permutation matrices defined in Appendix~\ref{sec:auxiliary_math}.
-Therefore, $\vt q$ is constrained to lie in the null space of a matrix $\mt G = \sum_{k=1}^3(L_k \hat{\vt C}_k - I_k \omega_k \hat{\vt B}_k)$. The determinant of this matrix can be obtained as
-\begin{equation*}
-\det \mt G = (\tr{\vt \omega}{\mt I}^2{\vt \omega} - L^2)^2,
-\end{equation*}
-which is identically null in a free rotational motion.
-In fact, this is a requirement for a matrix to have a non-empty null space.
-Using a computer algebra package, it is straightforward to find out that $\vt G$ has a two-dimensional null space whose basis set is formed by the two vectors
-\begin{equation*}
-\vt v_1 = \begin{smallarray}{c}
-L_2 + I_2 \omega_2 \\
-L_3 - I_3 \omega_3 \\
-0 \\
-L_1 - I_1 \omega_1
-\end{smallarray}
-\quad \text{and} \quad
-\vt v_1 = \begin{smallarray}{c}
-L_3 + I_3 \omega_3 \\
-L_2 - I_2 \omega_2 \\
-L_1 - I_1 \omega_1 \\
-0
-\end{smallarray}.
-\end{equation*}
-
-\begin{equation*}
-\exp\left(-\begin{smallarray}{cc}
- \alpha(t) & \phi(t) \\
--\phi(t) & \alpha(t)
-\end{smallarray}\right) = 
-e^{-\alpha(t)} \begin{smallarray}{cc}
-\cos \phi(t) & -\sin \phi(t) \\
-\sin \phi(t) &  \cos \phi(t)
-\end{smallarray}
-\end{equation*}
-
-Here,
-\begin{equation}
-\alpha(t) = \frac{1}{2}\int_0^t \frac{I_1 \omega_1^\prime(\tau)}{L - I_1\omega_1(\tau)}d\tau =
-\frac{1}{2}\ln\left(\frac{L - I_1\omega_1}{L - I_1\omega_1^0}\right)
-\end{equation}
-
-\begin{equation}
-\varphi(t) = \frac{1}{2}\int_0^t \frac{2 K_r + L \omega_1(\tau)}{L + I_1\omega_1(\tau)}d\tau
-\end{equation}
-
-After some algebra\cite{Celledoni_2008}
-\begin{equation*}
-\varphi(t) = \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2 I_1 L}\int_0^t \frac{1}{1+(I_1/L)\omega_1(\tau)}d\tau.
-\end{equation*}
-
-If $\lambda_1 < \lambda_3$, then $\omega_1(t) = A_1 \text{cn}(\alpha t + \epsilon)$
-
-\begin{align*}
-\int \frac{1}{1 + \alpha \cn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-\eta,k^2) + \\
-+ \frac{\alpha}{\sqrt{k^2+\eta}}\arctan\left(\frac{\dn u}{\sn u \sqrt{k^2+\eta}}\right)\Bigg]
-\end{align*}
-
-\begin{align*}
-\int \frac{1}{1 + \alpha \dn u}du = (\eta + 1) \Bigg[u + \Theta(\sn u,-k^2\eta,k^2) + \\
-+ \frac{\alpha}{k\sqrt{1/k^2+\eta}}\arctan\left(\frac{\cn u}{k \sn u \sqrt{1/k^2+\eta}}\right)\Bigg]
-\end{align*}
-
-\begin{align*}
-\Theta(x,n,m) &= \Pi(x,n,m) - \text{sn}^{-1}(x,m) = \\
-&= \frac{1}{3}nx^3R_J(1-x^2,1-mx^2,1,1-nx^2).
-\end{align*}
+
+\begin{figure}
+	\includegraphics{definite_integrals.eps}
+	\caption{Illustrative plot of functions $c(x)=1/(1-\alpha\cn x)$ (dark cyan line) and $d(x)=1/(1-\alpha \dn x)$ (magenta line) with $\alpha = 9/10$ and elliptic parameter $m = 1/2$, together with their definite integrals $C(x)$ (blue line) and $D(x)$ (red line), defined in Eqs.~\ref{eq:def_integral_C} and \ref{eq:def_integral_D}, respectively.}
+	\label{fig:definite_integrals}
+\end{figure}
 

### 2016-11-01 rigid_bodies.tex
Order inversion of symplectic integration and exact rotation

@@
 \end{subequations}
+
+\subsection{Symplectic Numerical Integration}
+\label{sec:symplectic}
+
+An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
+Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
+Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
+\begin{equation*}
+\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
+\end{equation*}
+where $i\!L$ is the Liouville operator.
+This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$.
+As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$.
+Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.
+
+Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}.
+Let us analyze three of them, which will be useful later on to derive a complete approximate solution.
+In what follows, an index $0$ stands for ``initial value''.
+We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$.
+The exact solution for this case is
+\[
+{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0,
+\]
+with ${\vt p}(t) = {\vt p}_0$, ${\vt q}(t) = {\vt q}_0$, and ${\vt \pi}(t) = {\vt \pi}_0$.
+This corresponds to a free translation (hence the subscript $t$).
+As the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected.
+Of course, the solution above is the exact effect of the propagator $e^{t i\!L_t}$, where the Liouville operator $i\!L_t$ is, simply,
+\[
+i L_t = \frac{\tr{\vt p}}{m}\diff{}{\vt r}.
+\]
+
+In the second hypothetical case, referred to as the boost case, we consider a Hamiltonian $\mathcal{H}_b = U(\vt r, \vt q)$.
+The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
+\begin{align*}
+&{\vt p}(t) = {\vt p}_0 + t \vt F \\
+&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C \vt \tau
+\end{align*}
+with ${\vt r}(t) = {\vt r}_0$ and ${\vt q}(t) = {\vt q}_0$.
+As $\vt r$ and $\vt q$ remain unaltered, $\vt F$, $\mt C$, and $\vt \tau$ are constant as well.
+This solution clearly conserves $\mathcal{H}_b$ and marks the effect of a propagator $e^{t i\!L_b}$, obtained from the operator
+\[
+i\!L_b = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
+\]
+
+Finally, the third hypothetical case consists of a Hamiltonian $\mathcal{H}_k = (1/2) I_k \omega_k^2$.
+Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation.
+The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt q$, and $\dot{\vt \pi} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt \pi$.
+A special feature of this system facilitates its solution.
+It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well.
+As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form.
+The analytical solution that takes place is\cite{Miller2002}
+\begin{align*}
+&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
+&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
+\end{align*}
+with ${\vt r}(t) = {\vt r}_0$ and ${\vt p}(t) = {\vt p}_0$.
+These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{t i\!L_k}$ derived from the operator
+\[
+i\!L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{\hat{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{\hat{\mt B}_k} \diff{}{\vt \pi} \right).
+\]
+
+Now note that the Hamiltonian in Eq.~\ref{eq:H_split_omega} is equal to the sum ${\mathcal H}_t + {\mathcal H}_b + \sum_{k=1}^3 {\mathcal H}_k$, so that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
+\begin{align}
+\label{eq:full_operator}
+i\!L = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k.
+\end{align}
+
+The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion which preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}.
+For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
+\[
+e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^\mathcal{N} + \mathcal{O}(\Delta t^2),
+\]
+where $\Delta t = t/\mathcal N$ is the time step size and $\mathcal N$ is the number of steps.
+We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side.
+By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ are commutable, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
+\begin{equation}
+\label{eq:trotter_splitting_NVE}
+e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
+\end{equation}
+
+Update of forces and torques is required only after the free translation and free rotation have been carried out. Each complete rotation is split as\cite{Miller2002}
+\begin{equation}
+\label{eq:splitting_rot}
+e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3} \right),
+\end{equation}
+where $\sum_{j=1}^n {\delta t}_j = \Delta t$.
+This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
+Following Miller \textit{et al}.,\cite{Miller2002} one could simply make $\delta t_j = \Delta t/n$ for all $j$.
+Alternatively, one could use a higher-order factorization scheme, such as the one due to Yoshida\cite{Yoshida1990} and Suzuki,\cite{Suzuki1991a, Suzuki1991b} in which $\delta t_j = \omega_j \Delta t$ and the weights $\omega_j$ are carefully chosen to determine the order of approximation.
+It is our intention in this paper to investigate how the performance of the integrator as a whole is affected by using such a higher-order scheme.
+For that, we will consider $n = 1$ with $\omega_1 = 1$ (zero-order) and $n = 7$ with $\omega_1 = \omega_7 = 0.784513610477560$, $\omega_2 = \omega_6 = 0.235573213359357$, $\omega_3 = \omega_5 = -1.17767998417887$, and $\omega_4 = 1 - \sum_{j \neq 4} \omega_j$ (sixth-order).
+
+Finally, we remark that an analytical solution for the complete rotation exists\cite{vanZon2007} and that splitting schemes with higher-order approximations have been used for rigid body dynamics.\cite{Omelyan2007, Omelyan2008, vanZon2008} However, they result in methods that are more complex and often require expensive force-gradient calculations.
+Moreover, it is unclear how their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than the microcanonical one.
 
@@
 \begin{align}
-\vt \omega(t) &= \begin{smallarray}{c}
+\vt \omega(t) &= \left[\begin{array}{c}
 a_1\text{cn}(\omega_p t + \epsilon) \\
@@
 a_3\text{dn}(\omega_p t + \epsilon)
-\end{smallarray} \quad \text{if} \quad \lambda_1 < \lambda_3 \quad \text{or} \\
-\vt \omega(t) &= \begin{smallarray}{c}
+\end{array}\right] \quad \text{if} \quad \lambda_1 \le \lambda_3 \quad \text{or} \\
+\vt \omega(t) &= \left[\begin{array}{c}
 a_1\text{dn}(\omega_p t + \epsilon) \\
@@
 a_3\text{cn}(\omega_p t + \epsilon)
-\end{smallarray} \quad \text{if} \quad \lambda_1 > \lambda_3,
+\end{array}\right] \quad \text{if} \quad \lambda_1 > \lambda_3,
 \end{align}
@@
 \begin{equation*}
-\omega_p = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
+\omega_p = \frac{(I_3 - I_1)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
 \epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}\right).
@@
 
+In the equations above, $\eta = \frac{\alpha^2}{1-\alpha^2}$, the function $\text{sgn}(\cdot)$ returns $-1$ if its argument is negative or $+1$ otherwise, the operator $\lfloor \cdot \rceil$ rounds a real value to its nearest integer (with the unusual feature that halfway values are rounded towards zero), and the function $\Theta$ is defined in terms of the Carlson elliptic integral $R_J$ as
+\begin{equation*}
+\Theta(x,n) = -\frac{1}{3}nx^3 R_J(1-x^2,1-mx^2,1,1+nx^2).
+\end{equation*}
+
 \begin{figure}
@@
 \end{figure}
-
-\subsection{Symplectic Numerical Integration}
-\label{sec:symplectic}
-
-An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
-Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
-Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
-\begin{equation*}
-\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
-\end{equation*}
-where $i\!L$ is the Liouville operator.
-This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$.
-As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$.
-Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.
-
-Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}.
-Let us analyze three of them, which will be useful later on to derive a complete approximate solution.
-In what follows, an index $0$ stands for ``initial value''.
-We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$.
-The exact solution for this case is
-\[
-{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0,
-\]
-with ${\vt p}(t) = {\vt p}_0$, ${\vt q}(t) = {\vt q}_0$, and ${\vt \pi}(t) = {\vt \pi}_0$.
-This corresponds to a free translation (hence the subscript $t$).
-As the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected.
-Of course, the solution above is the exact effect of the propagator $e^{t i\!L_t}$, where the Liouville operator $i\!L_t$ is, simply,
-\[
-i L_t = \frac{\tr{\vt p}}{m}\diff{}{\vt r}.
-\]
-
-In the second hypothetical case, referred to as the boost case, we consider a Hamiltonian $\mathcal{H}_b = U(\vt r, \vt q)$.
-The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
-\begin{align*}
-&{\vt p}(t) = {\vt p}_0 + t \vt F \\
-&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C \vt \tau
-\end{align*}
-with ${\vt r}(t) = {\vt r}_0$ and ${\vt q}(t) = {\vt q}_0$.
-As $\vt r$ and $\vt q$ remain unaltered, $\vt F$, $\mt C$, and $\vt \tau$ are constant as well.
-This solution clearly conserves $\mathcal{H}_b$ and marks the effect of a propagator $e^{t i\!L_b}$, obtained from the operator
-\[
-i\!L_b = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
-\]
-
-Finally, the third hypothetical case consists of a Hamiltonian $\mathcal{H}_k = (1/2) I_k \omega_k^2$.
-Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation.
-The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt q$, and $\dot{\vt \pi} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt \pi$.
-A special feature of this system facilitates its solution.
-It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well.
-As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form.
-The analytical solution that takes place is\cite{Miller2002}
-\begin{align*}
-&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
-&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
-\end{align*}
-with ${\vt r}(t) = {\vt r}_0$ and ${\vt p}(t) = {\vt p}_0$.
-These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{t i\!L_k}$ derived from the operator
-\[
-i\!L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{\hat{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{\hat{\mt B}_k} \diff{}{\vt \pi} \right).
-\]
-
-Now note that the Hamiltonian in Eq.~\ref{eq:H_split_omega} is equal to the sum ${\mathcal H}_t + {\mathcal H}_b + \sum_{k=1}^3 {\mathcal H}_k$, so that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
-\begin{align}
-\label{eq:full_operator}
-i\!L = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k.
-\end{align}
-
-The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion which preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}.
-For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
-\[
-e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^\mathcal{N} + \mathcal{O}(\Delta t^2),
-\]
-where $\Delta t = t/\mathcal N$ is the time step size and $\mathcal N$ is the number of steps.
-We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side.
-By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ are commutable, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
-\begin{equation}
-\label{eq:trotter_splitting_NVE}
-e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
-\end{equation}
-
-Update of forces and torques is required only after the free translation and free rotation have been carried out. Each complete rotation is split as\cite{Miller2002}
-\begin{equation}
-\label{eq:splitting_rot}
-e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3} \right),
-\end{equation}
-where $\sum_{j=1}^n {\delta t}_j = \Delta t$.
-This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
-Following Miller \textit{et al}.,\cite{Miller2002} one could simply make $\delta t_j = \Delta t/n$ for all $j$.
-Alternatively, one could use a higher-order factorization scheme, such as the one due to Yoshida\cite{Yoshida1990} and Suzuki,\cite{Suzuki1991a, Suzuki1991b} in which $\delta t_j = \omega_j \Delta t$ and the weights $\omega_j$ are carefully chosen to determine the order of approximation.
-It is our intention in this paper to investigate how the performance of the integrator as a whole is affected by using such a higher-order scheme.
-For that, we will consider $n = 1$ with $\omega_1 = 1$ (zero-order) and $n = 7$ with $\omega_1 = \omega_7 = 0.784513610477560$, $\omega_2 = \omega_6 = 0.235573213359357$, $\omega_3 = \omega_5 = -1.17767998417887$, and $\omega_4 = 1 - \sum_{j \neq 4} \omega_j$ (sixth-order).
-
-Finally, we remark that an analytical solution for the complete rotation exists\cite{vanZon2007} and that splitting schemes with higher-order approximations have been used for rigid body dynamics.\cite{Omelyan2007, Omelyan2008, vanZon2008} However, they result in methods that are more complex and often require expensive force-gradient calculations.
-Moreover, it is unclear how their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than the microcanonical one.
 

### 2017-01-30 rigid_bodies.tex
first modification

--- 2016-11-01 rigid_bodies.tex
@@
 
-\section{Introduction}
-
-Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations.
-This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
-Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.
-
-One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
-Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion.
-This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles.
-However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
-This is expected to be computationally more efficient because the constraints are treated implicitly.
-In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
-Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
-This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth.
-It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
-
-In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases.
-Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
-This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom.
-We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
-For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
-
-The paper is organized as follows.
-In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions.
-In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
-Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure.
-Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
-
-Throughout this paper, we use the word ``quaternion'' as a synonym for a four-dimensional vector, avoiding other interpretations.
-Thus, only ordinary matrix/vector operations are present in the formulation.
-Some mathematical details are given in the Appendix.
-
-\section{Mechanics of Rigid Body Motion}
-\label{sec:mechanics}
-
-In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study.
-Unit quaternions are employed to represent the orientations of rigid bodies.
-Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector.
-An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
-
-\subsection{Kinematics}
-\label{sec:kinematics}
-
-Consider a rigid body composed of $n_p$ individual particles, with each particle $j$ having mass $M_j$ and being located at a Cartesian coordinate $\vt R_j$.
-The center of mass of this body is located at
-\begin{equation*}
-\vt r = \frac{1}{m}\sum_{j=1}^{n_p} M_j {\vt R}_j,
-\end{equation*}
-where $m = \sum_{j=1}^{n_p} M_j$ is the total mass of the body.
-The position of each particle $j$ in a space-fixed frame of reference, whose origin coincides with the body's center of mass, is ${\vt \delta}_j = \tr{ [\begin{array}{ccc} {\delta_j}_x & {\delta_j}_y & {\delta_j}_z \end{array}]} = {\vt R}_j - {\vt r}$.
-In such frame, the symmetric inertia tensor for rotations of the body about the origin is\cite{Goldstein2002}
-\begin{equation*}
-{\mt I}^\ast = \sum_{j=1}^N M_j \begin{smallarray}{ccc}
-{\delta_j}_y^2 + {\delta_j}_z^2 & -{\delta_j}_x {\delta_j}_y & -{\delta_j}_x {\delta_j}_z \\
--{\delta_j}_y {\delta_j}_x & {\delta_j}_x^2 + {\delta_j}_z^2 & -{\delta_j}_y {\delta_j}_z \\
--{\delta_j}_z {\delta_j}_x & -{\delta_j}_z {\delta_j}_y & {\delta_j}_x^2 + {\delta_j}_y^2 \\
-\end{smallarray}.
-\end{equation*}
-
-For every non-collinear rigid body (for conciseness, the only case considered here), there exists a particular frame of reference in which its inertia tensor is diagonal.
-The formulation of the equations of motion in this frame of reference is notably simplified.
-Conversions from the space-fixed frame to such a body-fixed frame are performed by a rotation matrix $\mt A$ that satisfies the similarity relation\cite{Goldstein2002}
-\begin{equation*}
-{\mt I} = {\mt A} {\mt I}^\ast \tr{\mt A} = \begin{smallarray}{ccc}
-I_1 &   0 &   0 \\
-0 & I_2 &   0 \\
-0 &   0 & I_3 \\
-\end{smallarray},
-\end{equation*}
-where $I_1$, $I_2$, and $I_3$ (the eigenvalues of ${\mt I}^\ast$) are the principal moments of inertia and the columns of $\tr{\mt A}$ (the corresponding eigenvectors) are the principal axes of rotation.
-With no loss of generality, we can consider that $I_1 \geq I_2 \geq I_3$.
-Algorithms for obtaining $\mt A$ from $\mt I^\ast$ can be found in the literature (e.g.
-Ref.~\onlinecite{Kopp2008}).
-Once this is done, the position of every particle $j$ in the body-fixed frame can be obtained by ${\vt d}_j = {\mt A} {\vt \delta}_j$.
-Since $\mt A$ represents an orthogonal transformation, a conversion in the opposite direction is done by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} This is so because $\mt A \tr{\mt A} = \mt 1_3$, where $\mt 1_k$ represents a $k \times k$ identity matrix.
-By differentiating both sides of this relation with respect to time, we conclude that ${\mt A} \tr{\dot{\mt A}} = -\dot{\mt A} \tr{\mt A}$, which means that the result of $\mt A \tr{\dot{\mt A}}$ is skew-symmetric.
-The angular velocity of the body expressed in the body-fixed frame, $\vt \omega$, is related to this product by $\mt S(\vt \omega) = \mt A \tr{\dot{\mt A}}$,\cite{Haug1989} where $\mt S(\cdot)$ is the skew-symmetric operator defined in Appendix \ref{sec:auxiliary_math}  (Eq.~\ref{eq:operator_S}).
-
-An arbitrary rotation can be described using three independent parameters.
-One can use Euler angles, for instance, but mathematical singularities arise when the equations of motion are described in such terms.
-Instead, we consider an alternative representation, in which the orientation of a rigid body is given by four parameters $q_1$, $q_2$, $q_3$, and $q_4$, often referred to as Euler parameters, which form a quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ and satisfy the relation\cite{Goldstein2002}
+\section{New Theory}
+Exact solution for torque-free rotation
+In a previous paper\cite{Abreu_2017} we presented a formulation for the motion of rigid bodies which was simplified by the introduction of a particular factorization of the rotational matrix. 
+quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ and satisfy the relation\cite{Goldstein2002}
 \begin{equation}
@@
 
-Therefore, $\vt q$ is restricted to the surface of a unit hypersphere in $\mathbb{R}^4$, which is itself a three-dimensional domain.
-Methods are available for computing the unit quaternion related to a given rotation matrix (e.g.
-Ref.~\onlinecite{Shepperd1978}).
 Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
@@
 
-Some important properties of these matrices (or transposes thereof) have been reported in the literature\cite{Haug1989, Shuster1993, Dichmann1999, Ravishankar2004, Nielsen2012} and are revisited with our particular notation in Appendix~\ref{sec:auxiliary_math}.
 The most remarkable feature is that
@@
 Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.
-It is explained in Appendix~\ref{sec:auxiliary_math} that premultiplication of a vector with either $\mt B$ or $\mt C$ transforms it into a quaternion that lies in the hyperplane orthogonal to $\vt q$ and that both transformations can be reversed by applying $\tr{\vt B}$ or $\tr{\vt C}$ accordingly.
-Therefore, there exists an intermediary frame of reference, which is fixed with respect to the unit quaternion $\vt q$ that represents the orientation of the body.
-Its connection to the body-fixed and space-fixed frames are summarized as follows:
-\[
-\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
-\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
-\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
-\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
-\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
-\]
-
-Hence, Eq.~\ref{eq:relation_qdot_omega} shows that $\dot{\vt q}$ is half the quaternion-fixed version of the angular velocity vector.
-Besides, it guarantees that the norm of $\vt q$ remains unchanged because Eq.~\ref{eq:diff_qTq} is identically satisfied.
-
-To conclude this section, we discuss on the possibility of switching between alternative space-fixed frames when describing the motion of a rigid body.
-This will be useful when we deal with torque-free rotations (Sec.~\ref{sec:free_rotation}).
-Consider that a body-fixed vector $\vt d$ is simultaneously observed as $\vt \delta = \tr{\mt A\!}(\vt q){\vt d}$ in a space-fixed frame $\mathcal S_q$ and as $\vt \xi = \tr{\mt A\!}(\vt z){\vt d}$ in another space-fixed frame $\mathcal S_z$.
-It follows that $\vt \xi = \mt A(\vt g)\vt \delta$, where the composite rotation matrix $\mt A(\vt g) = \tr{\mt A\!}(\vt z){\mt A}(\vt q)$ denotes the relative orientation between $\mathcal S_q$ and $\mathcal S_z$.
-One key result in quaternion algebra\cite{Dichmann1999} is a simple relation that exists between $\vt q$, $\vt z$, and $\vt g$, which can be adapted to our notation as $\vt g = \tr{[\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}]} \vt q$.
-Note that both $\vt q$ and $\vt z$ vary while the body rotates, but the quaternion $\vt g$ remains constant because $\mathcal S_q$ and $\mathcal S_z$ are fixed in space.
-Moreover, as Eq.~\ref{eq:relation_qdot_omega} was derived for an arbitrary space-fixed frame, it is also true that $\dot{\vt z} = \tfrac{1}{2}{\vt B}(\vt z)\vt \omega$.
-In conclusion, as an alternative to tracking the evolution of $\vt q(t)$ after an instant $t_0$ in which $\vt q = \vt q_0$, one can select some other convenient unit quaternion $\vt z_0$ and then track the evolution of $\vt z(t)$, starting from $\vt z(t_0) = \vt z_0$.
-Then, whenever $\vt q$ is needed, one can simply make
-\begin{equation}
-\label{eq:relation_q_z}
-\vt q = [\begin{array}{cc}\vt z & \mt C(\vt z)\end{array}] \tr{[\begin{array}{cc}\vt z_0 & \mt C(\vt z_0)\end{array}]} \vt q_0,
-\end{equation}
-which is correct because these block matrices represent orthogonal transformations in $\mathbb{R}^4$.
+
+\section{Introduction}
+\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
+\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
+Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations.
+This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
+Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.
+
+One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
+Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion.
+This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles.
+However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
+This is expected to be computationally more efficient because the constraints are treated implicitly.
+In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
+Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
+This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth.
+It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
+
+In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases.
+Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
+This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom.
+We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
+For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
+
+The paper is organized as follows.
+In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions.
+In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
+Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure.
+Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
+
+Throughout this paper, we use the word ``quaternion'' as a synonym for a four-dimensional vector, avoiding other interpretations.
+Thus, only ordinary matrix/vector operations are present in the formulation.
+Some mathematical details are given in the Appendix.
+
+\section{Mechanics of Rigid Body Motion}
+\label{sec:mechanics}
+
+In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study.
+Unit quaternions are employed to represent the orientations of rigid bodies.
+Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector.
+An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
+
+\subsection{Kinematics}
 
@@
 The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$.
-Several authors have obtained Hamiltonian equations for the rotation of rigid bodies using quaternions as generalized coordinates.\cite{Maciejewski1985, Dichmann1996, Miller2002, Ravishankar2004, Nielsen2012} Here we try to combine convenient features of their formulations.
-We start by defining the Lagrangian $\mathcal{L} = K - U$ as a function of $\vt r$, $\dot{\vt r}$, $\vt q$, and $\dot{\vt q}$, where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
-Since $K = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + \frac{1}{2} \tr{\vt \omega} \mt I \vt \omega$,\cite{Goldstein2002} we can resort to Eq.~\ref{eq:relation_qdot_omega} to write
-\begin{equation*}
-\mathcal{L} = \frac{1}{2} m \tr{\dot{\vt r}} \dot{\vt r} + 2 \tr{\dot{\vt q}} \mt B \mt I \tr{\mt B} \dot{\vt q} - U(\vt r, \vt q).
-\end{equation*}
-
-To employ the Hamiltonian approach, besides the linear momentum $\vt p = \partial \mathcal{L}/\partial \dot{\vt r} = m \dot{\vt r}$, conjugated to $\vt r$, we introduce the momentum $\vt \pi = \partial \mathcal{L}/\partial \dot{\vt q}$, conjugated to $\vt q$.\cite{Goldstein2002} By carrying out the differentiation (see Appendix \ref{sec:Diff_Rules}), we get
-\begin{equation}
-\label{eq:conj_momentum}
-\vt \pi = 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega.
-\end{equation}
-
-Hence, $\vt \pi$ is twice the quaternion-fixed version of the angular momentum $\vt L = \mt I \vt \omega$ and, as such, it is also orthogonal to $\vt q$.
-Using the relation above, we can write down the Hamiltonian $\mathcal{H} = K + U$ as a function of $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$, which is
-\begin{equation}
-\label{eq:H_with_B}
-\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt \pi} {\mt B} {\mt I}^{-1} \tr{\mt B} \vt \pi}{8} + U(\vt r, \vt q).
-\end{equation}
-
-The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002} Eq.~\ref{eq:H_with_B} is convenient for differentiation with respect to $\vt \pi$, since $\mt B(\vt q)$ does not depend on it.
-However, if we define $\mt \Omega = \mt B(\vt \pi)$, it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$ and, therefore, a convenient form of $\mathcal{H}$ for differentiation with respect to $\vt q$ is
-\begin{equation}
-\label{eq:H_with_Omega}
-\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + \frac{\tr{\vt q} {\mt \Omega} {\mt I}^{-1} \tr{\mt \Omega} \vt q}{8} + U(\vt r, \vt q).
-\end{equation}
-
-For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
-\[
-\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
-\]
-where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference (see Appendix \ref{sec:Diff_PotEng}), and thus $\vt F$ is the resultant force on the body.
-To obtain the derivative $\partial U/\partial \vt q$, one must take into account that the entries of $\vt q$ are not independent, due to the constraint imposed by Eq.~\ref{eq:norm_eq_1}.
-The correct way of performing differentiation with respect to a constrained vector is described in Ref.~\onlinecite{Schay1995}.
-As demonstrated in Appendix \ref{sec:Diff_PotEng}, we have
-\[
-\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
-\]
-where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.
-Therefore, $-\partial U/\partial \vt q$ is twice the torque represented in quaternion-fixed frame of reference.
-Finally, after carrying out the required differentiations of $\mathcal{H}$ using either Eq.~\ref{eq:H_with_B} or Eq.~\ref{eq:H_with_Omega} (see Appendix~\ref{sec:Diff_Rules}) and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
+we obtain the following Hamiltonian system of equations of motion for a rigid body:
 \begin{subequations}
@@
 
+\subsection{Symplectic Numerical Integration}
+\label{sec:symplectic}
+
+An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
+Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
 \subsection{Exact Solution for Free Rotation}
 \label{sec:free_rotation}
-
-Eq.~\ref{eq:EDO_system} cannot be integrated analytically for a general potential $U(\vt r,\vt q)$, but closed-form solutions for Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi} are known in the case when $\vt \tau = \vt 0$.
-Torque-free rotations are useful in Molecular Dynamics as part of numerical integration methods based on splitting strategies.
-This is the topic addressed in Sec.~\ref{sec:symplectic}.
-As shown in Appendix~\ref{sec:derivation_euler}, Eq.~\ref{eq:EDO_pi} results in the well-known Euler equation $\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau$, which is totally decoupled from Eq.~\ref{eq:EDO_q} when $\vt \tau = \vt 0$.
-It is a known fact that both the rotational energy $K_r = \frac{1}{2}\tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-fixed angular momentum vector $\vt L$ (whose squared norm is $L^2 = \tr{\vt \omega}{\mt I}^2{\vt \omega}$) are integrals of motion in a free rotation.
-This allows us to solve the Euler equation analytically for $\vt \omega$.
-In the case of a spherical top ($I_1 = I_2 = I_3$), the solution is simply $\vt \omega(t) = \vt \omega(0)$.
-Inspired by the derivations in Refs.~\onlinecite{Landau1976, Thorn2012, vanZon2007}, we present the general solution in a form that facilitates computer implementation and allows a unified treatment for both asymmetric ($I_1 > I_2 > I_3$) and symmetric tops (either $I_1 = I_2$ or $I_2 = I_3$) if infinities are properly dealt with (IEEE 754 standard).
-From an initial condition $\vt \omega(0) = \vt \omega_0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, the solution starts by computing $K_r = \frac{1}{2}\tr{\vt \omega_0}{\mt I}{\vt \omega}_0$ and $L = (\tr{\vt \omega_0}{\mt I}^2{\vt \omega_0})^{1/2}$, followed by evaluating a set of prefactors defined as
-\begin{align*}
-a_1 &= \text{sgn}(\omega_1^0)\sqrt{\frac{L^2 - 2 I_3 K_r}{I_1(I_1 - I_3)}}, \\
-a_2 &= \sqrt{\min(\lambda_1,\lambda_3)}, \quad \text{and} \\
-a_3 &= \text{sgn}(\omega_3^0)\sqrt{\frac{2 I_1 K_r - L^2}{I_1(I_1 - I_3)}}, \\
-\end{align*}
-where $\lambda_1 = \frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}$ and $\lambda_3 = \frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}$.
-Note that the value of $a_2$ depends on a comparison between $\lambda_1$ and $\lambda_3$.
-This also occurs for $\vt \omega$, whose functional form is
-\begin{subequations}
-\begin{align}
-\vt \omega(t) &= \begin{smallarray}{c}
-a_1\text{cn}(\omega_p t + \epsilon) \\
-a_2\text{sn}(\omega_p t + \epsilon) \\
-a_3\text{dn}(\omega_p t + \epsilon)
-\end{smallarray} \quad \text{if} \quad \lambda_1 < \lambda_3 \quad \text{or} \\
-\vt \omega(t) &= \begin{smallarray}{c}
-a_1\text{dn}(\omega_p t + \epsilon) \\
-a_2\text{sn}(\omega_p t + \epsilon) \\
-a_3\text{cn}(\omega_p t + \epsilon)
-\end{smallarray} \quad \text{if} \quad \lambda_1 > \lambda_3,
-\end{align}
-\end{subequations}
-where $\text{sn}$, $\text{cn}$, and $\text{dn}$ are the Jacobi elliptic functions\cite{Byrd_1971} with parameter $m = k^2 = {a_2}/{\text{max}(\lambda_1,\lambda_3)}$ and the constants $\omega_p$ and $\epsilon$ are given by
-\begin{equation*}
-\omega_p = -\frac{(I_1 - I_3)a_1 a_3}{I_2 a_2} \quad \text{and} \quad
-\epsilon = \text{sn}^{-1}\left(\frac{\omega_2^0}{a_2}\right).
-\end{equation*}
-
-In order to compute the inverse function $\text{sn}^{-1}$, one can resort to its relation with an elliptic integral in Carlson form,\cite{Carlson_1977, Gradshteyn2014} $\text{sn}^{-1}(x) = x R_F(1-x^2,1-mx^2,1)$.
-One can numerically evaluate Jacobi functions and Carlson integrals by using, for instance, the routines available in the GNU Scientific Library.\cite{Galassi_2009}
-
-Once $\vt \omega(t)$ is known, we can solve Eq.~\ref{eq:EDO_q} for the unit quaternion $\vt q$ with initial condition $\vt q(0) = \vt q_0$.
-Here we propose a new procedure that is closely linked to the methods by Kosenko\cite{Kosenko_1998} and by Celledoni \textit{et al}.,\cite{Celledoni_2008} but that relies on the rotation matrix factorization (Eq.~\ref{eq:factorization_of_A}) to facilitate the task.
-These methods are also related to the one by van Zon and Schofield,\cite{vanZon2007} as shown in Ref.~\onlinecite{Celledoni_2008}.
-During a free rotation, any space-fixed representation of the angular momentum vector will remain constant.
-Thus, instead of solving Eq.~\ref{eq:EDO_q} directly for obtaining $\vt q(t)$, we solve an analogous equation $\dot{\vt z} = \frac{1}{2}{\mt B(\vt z)}{\vt \omega}$ for determining a unit quaternion $\vt z(t)$ specially chosen so that $\tr{\mt A\!}(\vt z)\mt I\vt \omega = \tr{[\begin{array}{ccc} L & 0 & 0 \end{array}]}$.
-In other words, we look for an alternative space-fixed frame whose first axis is aligned with the angular momentum vector.
-After replacing $\tr{\mt A\!}(\vt z) = \tr{\mt C}(\vt z){\mt B}(\vt z)$, we premultiply both sides by $\mt C(\vt z)$ and then apply Eq.~\ref{eq:product_B_vector} to obtain the homogeneous linear system $\mt G \vt z = \vt 0$, where $\mt G = L \hat{\vt C}_1 - \sum_{j=1}^3 I_j \omega_j \hat{\vt B}_j$.
-This means that $\vt z$ will continuously lie in the kernel of matrix $\mt G$, which can be determined by a standard elimination method.
-Such kernel turns out to be a (moving) linear space defined by the orthogonal basis vectors
-\begin{equation*}
-\vt u(t) = \begin{smallarray}{c} I_3\omega_3(t) \\ I_2\omega_2(t) \\ L-I_1\omega_1(t) \\ 0 \end{smallarray}
-\quad \text{and} \quad
-\vt v(t) = \begin{smallarray}{c} -I_2\omega_2(t) \\ I_3\omega_3(t) \\ 0 \\ L-I_1\omega_1(t) \end{smallarray}.
-\end{equation*}
-
-At all times, therefore, $\vt z$ will be a linear combination of $\vt u$ and $\vt v$. We can then write the expression
-\begin{equation*}
-\vt z(t) = \frac{b(t) \vt u(t) + c(t) \vt v(t)}{\sqrt{2L[L-I_1\omega_1(t)]}},
-\end{equation*}
-whose denominator is the norm of both $\vt u$ and $\vt v$.
-The problem has now been reduced to determining how the coefficients $b$ and $c$ vary with time.
-Further reduction is achieved by realizing that both the unit-norm constraint ($\tr{\vt z}{\vt z} = 1$) and the orthogonality condition ($\tr{\vt u}{\vt v} = 0$) hold at all times, which makes
-\begin{equation*}
-\tr{\vt z}{\vt z} = \frac{b^2 \tr{\vt u}{\vt u} + 2bc\tr{\vt u}{\vt v} + c^2 \tr{\vt v}{\vt v}}{2L[L-I_1\omega_1(t)]} = b^2 + c^2 = 1.
-\end{equation*}
-
-Hence, the values of $b$ and $c$ are restricted to a unit circle, so that we can finally rewrite $\vt z(t)$ as
-\begin{equation*}
-\vt z(t) = \frac{\vt u(t) \cos \theta(t) + \vt v(t) \sin \theta(t)}{\sqrt{2L[L-I_1\omega_1(t)]}}.
-\end{equation*}
-
-The next step consists in replacing this result into the equation $\dot{\vt z} = \frac{1}{2}{\mt B(\vt z)}{\vt \omega}$.
-This yields, after some algebra, a single differential equation whose solution can be expressed as
-\begin{equation*}
-\theta(t) = \theta_0 + \frac{L}{2I_1}t + \frac{2 I_1 K_r - L^2}{2L I_1}\int_0^t \frac{1}{1-(I_1/L)\omega_1(s)}ds.
-\end{equation*}
-
-In order to solve this equation, we introduce the following functions derived from definite integrals: $C(x) = \int_0^x \frac{1}{1 - \alpha \cn y}dy$ and $D(x) = \int_0^x \frac{1}{1 - \alpha \dn y}dy$, where $\alpha = I_1a_1/L$.
-
-\begin{widetext}
-	\begin{subequations}
-\begin{gather}
-\label{eq:def_integral_C} \frac{C(x)}{\eta + 1} = x + \text{sgn}(\cn x) \Theta(\sn x,\eta) + 2\Theta(1,\eta) \left\lfloor \frac{x}{2 K} \right\rceil +
-\frac{\alpha}{\sqrt{m+\eta}} \arctan\left(\frac{\sqrt{m+\eta} \; \sn x}{\dn x}\right) \\
-\label{eq:def_integral_D} \frac{D(x)}{\eta + 1} = x + \text{sgn}(\cn x) \Theta(\sn x,m\eta) + 2\Theta(1,m\eta) \left\lfloor\frac{x}{2 K}\right\rceil +
-\frac{\alpha}{\sqrt{m\eta + 1}} \left\{\arctan\left(\frac{\sqrt{m\eta + 1} \; \sn x}{\cn x}\right) + \pi \left\lfloor \frac{x}{2 K} \right\rceil\right\}
-\end{gather}
-	\end{subequations}
-\end{widetext}
-
-\begin{figure}
-	\includegraphics{definite_integrals.eps}
-	\caption{Illustrative plot of functions $c(x)=1/(1-\alpha\cn x)$ (dark cyan line) and $d(x)=1/(1-\alpha \dn x)$ (magenta line) with $\alpha = 9/10$ and elliptic parameter $m = 1/2$, together with their definite integrals $C(x)$ (blue line) and $D(x)$ (red line), defined in Eqs.~\ref{eq:def_integral_C} and \ref{eq:def_integral_D}, respectively.}
-	\label{fig:definite_integrals}
-\end{figure}
-
-\subsection{Symplectic Numerical Integration}
-\label{sec:symplectic}
-
-An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
-Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
-Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
-\begin{equation*}
-\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
-\end{equation*}
-where $i\!L$ is the Liouville operator.
-This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$.
-As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$.
-Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.
-
-Analytical solutions exist for some simpler hypothetical cases related to Eq.~\ref{eq:EDO_system_alternative}.
-Let us analyze three of them, which will be useful later on to derive a complete approximate solution.
-In what follows, an index $0$ stands for ``initial value''.
-We start by considering a hypothetical Hamiltonian $\mathcal{H}_t = (1/2m) \tr{\vt p} \vt p$, whose resulting equations of motion are $\dot{\vt r} = {\vt p}/m$, $\dot{\vt p} = 0$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 0$.
-The exact solution for this case is
-\[
-{\vt r}(t) = {\vt r}_0 + \frac{t}{m} {\vt p}_0,
-\]
-with ${\vt p}(t) = {\vt p}_0$, ${\vt q}(t) = {\vt q}_0$, and ${\vt \pi}(t) = {\vt \pi}_0$.
-This corresponds to a free translation (hence the subscript $t$).
-As the momentum $\vt p$ remains constant, the Hamiltonian $\mathcal{H}_t$ is conserved, as expected.
-Of course, the solution above is the exact effect of the propagator $e^{t i\!L_t}$, where the Liouville operator $i\!L_t$ is, simply,
-\[
-i L_t = \frac{\tr{\vt p}}{m}\diff{}{\vt r}.
-\]
-
-In the second hypothetical case, referred to as the boost case, we consider a Hamiltonian $\mathcal{H}_b = U(\vt r, \vt q)$.
-The resulting equations of motion are $\dot{\vt r} = 0$, $\dot{\vt p} = \vt F$, $\dot{\vt q} = 0$, and $\dot{\vt \pi} = 2 \mt C \vt \tau$, whose exact solution is
-\begin{align*}
-&{\vt p}(t) = {\vt p}_0 + t \vt F \\
-&{\vt \pi}(t) = {\vt \pi}_0 +  2 t \mt C \vt \tau
-\end{align*}
-with ${\vt r}(t) = {\vt r}_0$ and ${\vt q}(t) = {\vt q}_0$.
-As $\vt r$ and $\vt q$ remain unaltered, $\vt F$, $\mt C$, and $\vt \tau$ are constant as well.
-This solution clearly conserves $\mathcal{H}_b$ and marks the effect of a propagator $e^{t i\!L_b}$, obtained from the operator
-\[
-i\!L_b = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C} \diff{}{\vt \pi}.
-\]
-
-Finally, the third hypothetical case consists of a Hamiltonian $\mathcal{H}_k = (1/2) I_k \omega_k^2$.
-Note that a single entry of $\vt \omega$ is accounted for in this case, so that it corresponds to a free uniaxial rotation.
-The equations of motion resulting from $\mathcal{H}_k$ are $\dot{\vt r} = 0$, $\dot{\vt p} = 0$, $\dot{\vt q} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt q$, and $\dot{\vt \pi} = \frac{1}{2} \omega_k \hat{\mt B}_k \vt \pi$.
-A special feature of this system facilitates its solution.
-It is known beforehand that the solution will keep $\mathcal{H}_k$ unaltered and, as a consequence, $\omega_k$ will be a constant as well.
-As noticed by Miller~\textit{el al.}~\cite{Miller2002}, this decouples the equations for $\dot{\vt q}$ and $\dot{\vt \pi}$, which are identical in form.
-The analytical solution that takes place is\cite{Miller2002}
-\begin{align*}
-&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
-&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
-\end{align*}
-with ${\vt r}(t) = {\vt r}_0$ and ${\vt p}(t) = {\vt p}_0$.
-These equations, with $\omega_k$ calculated from $\vt q_0$ and $\vt \pi_0$ using Eq.~\ref{eq:omega_entry}, represent the effect of a propagator $e^{t i\!L_k}$ derived from the operator
-\[
-i\!L_k = \frac{1}{2} \omega_k \left( \tr{\vt q}\tr{\hat{\mt B}_k} \diff{}{\vt q} + \tr{\vt \pi}\tr{\hat{\mt B}_k} \diff{}{\vt \pi} \right).
-\]
-
-Now note that the Hamiltonian in Eq.~\ref{eq:H_split_omega} is equal to the sum ${\mathcal H}_t + {\mathcal H}_b + \sum_{k=1}^3 {\mathcal H}_k$, so that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
-\begin{align}
-\label{eq:full_operator}
-i\!L = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k.
-\end{align}
-
-The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion which preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}.
-For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
-\[
-e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^\mathcal{N} + \mathcal{O}(\Delta t^2),
-\]
-where $\Delta t = t/\mathcal N$ is the time step size and $\mathcal N$ is the number of steps.
-We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side.
-By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ are commutable, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
-\begin{equation}
-\label{eq:trotter_splitting_NVE}
-e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
-\end{equation}
-
-Update of forces and torques is required only after the free translation and free rotation have been carried out. Each complete rotation is split as\cite{Miller2002}
-\begin{equation}
-\label{eq:splitting_rot}
-e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3} \right),
-\end{equation}
-where $\sum_{j=1}^n {\delta t}_j = \Delta t$.
-This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
-Following Miller \textit{et al}.,\cite{Miller2002} one could simply make $\delta t_j = \Delta t/n$ for all $j$.
-Alternatively, one could use a higher-order factorization scheme, such as the one due to Yoshida\cite{Yoshida1990} and Suzuki,\cite{Suzuki1991a, Suzuki1991b} in which $\delta t_j = \omega_j \Delta t$ and the weights $\omega_j$ are carefully chosen to determine the order of approximation.
-It is our intention in this paper to investigate how the performance of the integrator as a whole is affected by using such a higher-order scheme.
-For that, we will consider $n = 1$ with $\omega_1 = 1$ (zero-order) and $n = 7$ with $\omega_1 = \omega_7 = 0.784513610477560$, $\omega_2 = \omega_6 = 0.235573213359357$, $\omega_3 = \omega_5 = -1.17767998417887$, and $\omega_4 = 1 - \sum_{j \neq 4} \omega_j$ (sixth-order).
-
-Finally, we remark that an analytical solution for the complete rotation exists\cite{vanZon2007} and that splitting schemes with higher-order approximations have been used for rigid body dynamics.\cite{Omelyan2007, Omelyan2008, vanZon2008} However, they result in methods that are more complex and often require expensive force-gradient calculations.
-Moreover, it is unclear how their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than the microcanonical one.
 
@@
 
-\subsection{Pressure of a System of Rigid Bodies}
-\label{sec:pressure}
-
-The pressure of a system with constant NVT, for which $\sum_{i=1}^N{\vt F}_i = 0$, is calculated by\cite{Louwerse2006, Tuckerman2001}
-\[
-P = \frac{(N-1) k_B T + \left\langle W \right\rangle}{V},
-\]
-where $W$ is the internal virial of the system, defined as $W = - V \frac{dU}{dV}$.
-For simplicity, we consider a system of rigid bodies under periodic boundary conditions whose potential energy function is pairwise additive with respect to individual atoms, so that the approach of Louwerse and Baerends\cite{Louwerse2006} can be applied (otherwise, one can employ the approach of Thompson \textit{et al.}\cite{Thompson2009}).
-Due to the rigidity of the bodies, interactions amongst companion atoms (i.e., atoms belonging to the same body) are irrelevant.
-Assuming that the minimum image convention\cite{Allen1989} applies for entire bodies, meaning that the simulation box is large enough to preclude any atom from having their own images or images of its companion atoms within its cutoff radius, the potential energy per unit box is given by
-\begin{equation}
-\label{eq:U_pairwise_pbc}
-U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon(r_{ijkl}),
-\end{equation}
-where $\upsilon(r)$ is a pair interaction potential and $r_{ijkl}$ is the distance between the $k$-th atom of body $i$ to its nearest image of the $l$-th atom of body $j$.
-Such distance is the norm of a vector ${\vt r}_{ijkl} = \hat{\vt r}_{ij} + \tr{[{\mt A}({\vt q}_i)]}{{\vt d}_k}_i - \tr{[{\mt A}({\vt q}_j)]}{{\vt d}_l}_k$, where $\hat{\vt r}_{ij}$ is the minimum-image vector pointing from the center of mass of body $j$ to that of body $i$.
-We note that $U$ does not depend explicitly on $V$, which makes ${\partial U}/{\partial V} = 0$.
-If the atoms were free to move individually, an isotropic change in volume would displace all atoms and we would have ${\partial {\vt r}_{ijkl}}/{\partial V} = (3V)^{-1} \tr{\vt r}_{ijkl}$.\cite{Louwerse2006} In this event, the atomic internal virial would be computed by
-\begin{equation}
-\label{eq:virial_atoms}
-W_\text{a} = \frac{1}{3} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \tr{\vt r}_{ijkl} {\vt f}_{ijkl},
-\end{equation}
-where $\vt f_{ijkl} = -{\partial \upsilon}/{\partial \vt r_{ijkl}}$ is the minimum-image force exerted by the $l$-th atom of body $j$ on the $k$-th atom of body $i$.
-In the case of rigid bodies, considering that an isotropic change in volume would neither deform nor rotate them, but only promote center-of-mass displacements, we have that ${\partial \hat{\vt r}_{ij}}/{\partial V} = (3V)^{-1} \tr{\hat{\vt r}}_{ij}$ and ${\partial {\vt q}_i}/{\partial V} = {\partial {\vt q}_j}/{\partial V} = \tr{\vt 0}$.
-Therefore, by applying the chain rule to Eq.~\ref{eq:U_pairwise_pbc} and realizing that $\partial \vt r_{ijkl}/\partial \hat{\vt r}_{ij} = \vt 1_3$, we obtain
-\[
-\label{eq:dUdV_rigid_bodies}
-\frac{dU}{dV} = -\frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \tr{\hat{\vt r}}_{ij} \vt f_{ijkl}.
-\]
-
-There are two convenient ways of employing this result to compute the internal virial of a system of rigid bodies.
-On one hand, we can explore the fact that ${\hat{\vt r}}_{ij}$ does not depend on the indexes $k$ and $l$.
-On the other hand, we can replace $\hat{\vt r}_{ij} = {\vt r}_{ijkl} - {{\vt \delta}_k}_i + {{\vt \delta}_l}_k$ and make use of the fact that $\vt f_{ijkl} = -\vt f_{jilk}$.
-Finally, we have
 \begin{subequations}
@@
 from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $\mathcal N$.
-Both the instantaneous fluctuations, given by the term inside the summation, and the long-time conservation of energy, $D\!E$, are monitored.
-
-Recall that, when the splitting scheme of Eq.~\ref{eq:splitting_rot} is executed with $\delta t_j = \Delta t/n$ for all $j$, the method introduced here for the NVE dynamics is just a mathematical reformulation of the algorithm of Miller \textit{et al}.\cite{Miller2002} Therefore, short trajectories obtained with both methods should coincide up to round-off errors.
-In Fig.~\ref{fig:miller1fs}, we plot the instantaneous fluctuations of total energy obtained with $n = 1$ and a time step of 1~fs using both methods.
-The simulation with the algorithm of Ref.~\onlinecite{Miller2002} was performed using the LAMMPS software package.\cite{Plimpton1995} For the purpose of comparison, we did not carry out force shifting for the Lennard-Jones model, as this is not implemented in LAMMPS.
-It can be seen in Fig.~\ref{fig:miller1fs} that, in fact, the fluctuations of both methods are of similar magnitudes.
-
-\begin{figure}
-	\includegraphics{millerourmd}
-	\caption{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and the LAMMPS implementation of the algorithm of Ref.~\onlinecite{Miller2002} (dotted line) with $n = 1$, $\Delta t = 1$ fs, and $E_0 = -2799.87$ kcal/mol.}
-	\label{fig:miller1fs}
-\end{figure}
-
-In Table~\ref{table:nve}, we compare the energy drifts observed when the rotational operator in Eq.~\ref{eq:splitting_rot} is split using either $n = 1$ or the sixth-order Yoshida-Suzuki factorization scheme ($n = 7$).
-As observed, for a small time steps of 1~fs, increasing the order of factorization does not provide any improvement.
-As the time step size increases, trajectories with higher accuracy are achieved with $n = 7$.
-Nevertheless, the improvement is only marginal and all trajectories are stable, exhibiting a small long-time energy drift.
-Thus, we proceed by considering $n = 1$ in the analysis of the NVT dynamics.
-
-\begin{table}
-	\caption{Average total-energy deviations for liquid water in NVE simulations for Eq.~\ref{eq:splitting_rot} with either $n = 1$ or $n = 7$ with Yoshida-Suzuki factorization.}
-	\label{table:nve}
-	\begin{ruledtabular}
-		\begin{tabular}{ccc}
-			 & \multicolumn{2}{c}{ $D\!E$ } \\
-			 \cline{2-3}
-			 $\Delta t$/fs & $n = 1$ & $n = 7$ \\
-			 \hline
-			 1.0 & 0.000055 & 0.000074 \\
-			 2.0 & 0.00048  & 0.00031  \\
-			 3.0 & 0.0012   & 0.0011   \\
-		\end{tabular}
-	\end{ruledtabular}
-\end{table}
-
 It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
@@
 
-\begin{table}
-	\begin{threeparttable}
-		\caption{Average deviation of the conserved quantity ($DE$) and average temperatures from liquid water simulation with NVT integration \tnote{a}\tnote{b}}
-		\label{table:denvt}
-		\begin{ruledtabular}
-			\begin{tabular}{ccccc}
-				& \multicolumn{2}{c}{Eq.~\ref{eq:trotter_splitting_NHC}} & \multicolumn{2}{c}{Eq.~\ref{eq:modified_splitting}} \\
-				\cline{2-5}
-				$\Delta t$/fs & $D\!E$ & $T$/K & $D\!E$ & $T$/K \\
-				\hline
-				1.0 & 0.000043 & 298.16  & 0.000052  & 298.15 \\
-				2.0 & 0.00013  & 298.23  & 0.00017   & 298.06 \\
-				3.0 & 0.00089  & 298.17  & 0.0013    & 298.12
-			\end{tabular}
-		\end{ruledtabular}
-		\begin{tablenotes}
-			\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$.
-			\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs.
-		\end{tablenotes}
-	\end{threeparttable}
-\end{table}
-
 \subsection{New Results}
-
-\begin{table*}
-	\caption{Results with Exact Solution for Free Rotations}
-	\label{table:NVE}
-	\begin{ruledtabular}
-		\begin{tabular}{cccccccc}
-			$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
-			\hline
-			1.0 & $	305.91	\pm	0.03	$ & $	-8395.1	\pm	0.1	$ & $	1645.9	\pm	0.1	$ & $	823.2	\pm	0.1	$ & $	822.7	\pm	0.1	$ & $	146	\pm	2	$ & $0.0456$ \\
-			2.0 & $	305.21	\pm	0.02	$ & $	-8389.5	\pm	0.1	$ & $	1642.1	\pm	0.1	$ & $	823.0	\pm	0.1	$ & $	819.1	\pm	0.1	$ & $	149	\pm	2	$ & $0.656$ \\
-			3.0 & $	304.30	\pm	0.03	$ & $	-8375.4	\pm	0.9	$ & $	1637.2	\pm	0.1	$ & $	824.0	\pm	0.1	$ & $	813.3	\pm	0.1	$ & $	156	\pm	2	$ & $2.87$ \\
-			4.0 & $	304.5	\pm	0.5	$ & $	-8339	\pm	6	$ & $	1638	\pm	3	$ & $	829	\pm	1	$ & $	808.8	\pm	0.6	$ & $	180	\pm	2	$ & $10.1$ \\
-		\end{tabular}
-	\end{ruledtabular}
-\end{table*}
-
-\begin{table*}
-\caption{Results.....}
-\label{table:NVE}
-\begin{ruledtabular}
-\begin{tabular}{cccccccc}
-$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
-\hline
-1.0 & $305.95\pm0.03$ & $-8395.4\pm0.1$ & $1646.1\pm0.1$ & $823.3\pm0.1$ & $822.9\pm0.1$ & $151\pm2$ & $0.0471$ \\
-2.0 & $305.17\pm0.03$ & $-8389.0\pm0.1$ & $1641.9\pm0.1$ & $823.2\pm0.1$ & $818.8\pm0.1$ & $148\pm2$ & $0.853$ \\
-3.0 & $304.34\pm0.03$ & $-8374\pm2$ & $1637.5\pm0.2$ & $824.0\pm0.1$ & $813.4\pm0.1$ & $152\pm2$ & $3.96$ \\
-4.0 & $303.6\pm0.4$ & $-8349\pm5$ & $1633\pm2$ & $826.8\pm0.2$ & $806.5\pm0.4$ & $168\pm2$ & $8.25$ \\
-\end{tabular}
-\end{ruledtabular}
-\end{table*}
 
@@
 \label{sec:energypartition}
-
-Checking the consistency between the equipartition of energy and the simulated energies serves as another approach for investigating the precision of a simulation.
-In fact, as stated by Eastwood \textit{et al}.,\cite{Eastwood_2010} this kind of checking has been useful to improve methodologies related to the calculation of  electrostatic\cite{Levitt_1988, Guenot_1992, Arnold_1994} and dispersive\cite{Sagui_1999} interactions, as well as to detect errors in thermostats\cite{Harvey_1998, Mor_2008} and barostats.\cite{Feller_1995} The truncation error associated to the numerical schemes is a potential source of error, and is the focus of the work of Eastwood \textit{et al}.,\cite{Eastwood_2010} which investigates a solvated protein and a temperature difference of approximately 5.5~K is observed between the protein and the solvent using a time step of 2~fs.
-
-In this paper, we analyze the partition of energy between the translational and rotational degrees of freedom, which are related to the temperature by the equipartition theorem as follows
-\begin{subequations}
-\begin{align}
-\label{eq:tras_equipartition}
-2 \langle E_\text{trans} \rangle = (3N - 3) k_b T
-\end{align}
-\begin{align}
-\label{eq:rot_equipartition}
-2 \langle E_\text{rot} \rangle = 3N k_b T
-\end{align} 
-\end{subequations}
-
-In these equations, $N$ is the number of rigid bodies.
-In microcanonical MD simulations, conservation of total linear momentum $\boldsymbol{P}$ results in the loss of three translational degrees of freedom, which is the reason why the factor $3N-3$ appears in Eq.~\ref{eq:tras_equipartition}.
-For a cubic box with periodic boundary conditions, there is no rotational symmetry,\cite{Frenkel_2013, Kuzkin_2014} with which the total angular momentum is not conserved, giving a number of $3N$ rotational degrees of freedom.
-We calculate the average energy per DOF $\langle E_{cin} \rangle_{DOF} = \langle E_{cin} \rangle /N_{DOF}$ and then we calculate both the translational and rotational energies from equipartition as follows
-
-\begin{equation}
-\langle E_{kin}^{trans}\rangle =  \langle E_{kin} \rangle_{DOF} \, (3N-3)
-\end{equation}
-
-\begin{equation}
-\langle E_{kin}^{rot} \rangle =  \langle E_{kin} \rangle_{DOF} \, (3N)
-\end{equation}
-
-In the tables below we report the translational and rotational energies using Eqs.~\ref{eq:tras_equipartition} and~\ref{eq:rot_equipartition}, together with those obtained in the NVE simulations.
-In this case, we simulated 903 molecules of liquid water in order to examine the behavior for a cutoff distance of $13$ {\AA}, which is a more conservative criterion as it provides better numerical stability.
-In Table~\ref{table:water_partition_LJ-SF} we observe that for a small time step of 0.5 fs the energies are in good correspondence, which is lost when the time step is increased.
-This result is confirmed also in Table~\ref{table:water_partition_13A_LJ-SF} for the cutoff distance of 13 {\AA}.
-This allows us to prove the influence of the truncation error on in this effect, in accordance with what Eastwood~\textit{et al.}~\cite{Eastwood_2010} point out.
-For significant deviations, the main implication is that Eqs.~\ref{eq:tras_equipartition} and~\ref{eq:rot_equipartition} are no longer suitable for the temperature calculation, which becomes an undefined property.
-In the aforementioned paper, the authors introduce an alternative approach for calculating the temperature which is based on the generalized equipartition theorem applied to the Shadow Hamiltonian, which will be investigated in future work.
-
-\begin{table}
-	\begin{threeparttable}
-		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c$ = 9~{\r A} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
-		\label{table:water_partition_LJ-SF}
-		\begin{ruledtabular}
-			\begin{tabular}{ccccccc}
-				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
-%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
-				\cline{4-7}
-				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
-				\hline
-				0.5 & 305.20  & 1642.11  & 820.60  & 821.06  & 820.60 & 821.51  \\
-				1.0 & 304.92  & 1640.57  & 819.83  & 820.74  & 820.63 & 819.94  \\
-				2.0 & 304.20  & 1636.74  & 817.92  & 818.82  & 820.56 & 816.18  \\
-				3.0 & 302.97  & 1630.07  & 814.59  & 815.48  & 820.47 & 809.60
-			\end{tabular}
-		\end{ruledtabular}
-		\begin{tablenotes}
-			\item[a] Energies are in kcal/mol.
-		\end{tablenotes}
-	\end{threeparttable}
-\end{table}
-
-\begin{table}
-	\begin{threeparttable}
-		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c$ = 13~{\r A} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
-		\label{table:water_partition_13A_LJ-SF}
-		\begin{ruledtabular}
-			\begin{tabular}{ccccccc}
-				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
-%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
-				\cline{4-7}
-				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
-				\hline
-				0.5 & 307.17 & 1652.71  & 825.90  & 826.81  & 825.96 & 826.75  \\
-				1.0 & 306.98 & 1651.68  & 825.38  & 826.30  & 826.09 & 825.59  \\
-				2.0 & 306.04 & 1646.60  & 822.85  & 823.75  & 825.57 & 821.03  \\
-				3.0 & 304.79 & 1639.89  & 819.49  & 820.40  & 825.63 & 814.26
-			\end{tabular}
-		\end{ruledtabular}
-		\begin{tablenotes}
-			\item[a] Energies are in kcal/mol.
-		\end{tablenotes}
-	\end{threeparttable}
-\end{table}
 
@@
 
-\subsection{Pressure Calculations}
-
-In Fig.~\ref{fig:pressure} we compare the results using Eq.~\ref{eq:virial_rigid_bodies} with those from a simulation performed with LAMMPS, which uses a numerical estimation of the constraint forces in the pressure calculation.
-The simulations were run in the NVE ensemble, for which neither shifting to the LJ forces nor analytical tail corrections were considered.
-
-Our purpose is....
-
-As one can see in Fig.~\ref{fig:pressure}....
-
 %\begin{center}
@@
 
-\begin{figure}
-	\includegraphics[width=0.5\textwidth,keepaspectratio]{FiguraAna}
-	\caption{Histogram of pressure values obtained using Eq.~\ref{eq:virial_rigid_bodies} (filled bars), along with the results from a simulation using LAMMPS (unfilled bars).
-The average pressures are $\bar{P} = -0.60$~bar and $\bar{P} = 2.14$~bar, respectively.
-Simulation of liquid water with NVE integration, $\Delta t = 1$~fs, and $n = 1$.}
-	\label{fig:pressure}
-\end{figure}
-
 \section{Concluding Remarks}
@@
 
-\appendix
-
-\section{Differentiation with vectors and matrices}
-\label{sec:Diff_Rules}
-
-Along the paper, we employ the gradient convention (rather than the Jacobian convention) for expressing derivatives with respect to vector variables.
-In this context, very often we differentiate a scalar function using one of the following rules:
-\[
-\diff{(\tr{\vt x}\vt a)}{\vt x} = \diff{(\tr{\vt a}\vt x)}{\vt x} = \vt a \quad \text{or} \quad \diff{(\tr{\vt x}\mt A \vt x)}{\vt x} = (\mt A + \tr{\mt A})\vt x,
-\]
-where $\vt a$ is a constant vector and $\mt A$ is a constant matrix. If $\mt A$ is symmetric, then $\mt A + \tr{\mt A} = 2\mt A$.
-
-\section{Algebraic relations involving unit quaternions}
-\label{sec:auxiliary_math}
-
-We start by analyzing the matrix $\mt B$ as defined in Eq.~\ref{eq:def_B_and_C}.
-Note that $\tr{\mt B}\vt q = \vt 0$, where $\vt 0$ is a null vector.\cite{Haug1989, Shuster1993, Dichmann1999} This means that all columns of $\mt B$ are orthogonal to $\vt q$.
-As these columns have unit norm and are also orthogonal to each other, they form an orthonormal basis for the hyperplane orthogonal to $\mt q$, which is a subspace of $\mathbb{R}^4$.
-Therefore, the general operation $\vt y = \mt B \vt x$ transforms a vector $\vt x \in \mathbb{R}^3$ into a quaternion $\vt y \in \mathbb{R}^4$, but restricted to the mentioned hyperplane.
-Other two identities that hold for $\mt B$ are\citep{Haug1989}
-\begin{equation}
-\label{eq:basics_B}
-\begin{aligned}
-\tr{\mt B}\mt B &= (\tr{\vt q}{\vt q}) {\mt 1}_3 \text{ and} \\
-\mt B\tr{\mt B} &= (\tr{\vt q}{\vt q}) {\mt 1}_4 - {\vt q}\tr{\vt q}.
-\end{aligned}
-\end{equation}
-
-If $\tr{\vt q}{\vt q} = 1$, then the operation $\vt y = \mt B \vt x$ can always be reverted by $\tr{\mt B}$.
-This is not true for a general operation in the opposite direction, $\vt x = \tr{\mt B} \vt y$.
-Besides unit norm, it is necessary that $\tr{\vt q}\vt y = 0$ (i.e.
-orthogonality of $\vt y$ and $\vt q$) for the operation to be reversible by $\mt B$.
-In fact, the composite transformation $\mt B\tr{\mt B}$ corresponds to a projection into the hyperplane orthogonal to $\vt q$.\cite{Dichmann1999}
-
-As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = \begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
-\[
-\hat{\mt B}_1 = \begin{smallarray}{cccc}
- 0 & \text{-}1 &  0 &  0 \\
- 1 &  0 &  0 &  0 \\
- 0 &  0 &  0 &  1 \\
- 0 &  0 & \text{-}1 &  0 \\
-\end{smallarray} \quad
-\hat{\mt B}_2 = \begin{smallarray}{cccc}
- 0 &  0 & \text{-}1 &  0 \\
- 0 &  0 &  0 & \text{-}1 \\
- 1 &  0 &  0 &  0 \\
- 0 &  1 &  0 &  0 \\
-\end{smallarray} \quad
-\hat{\mt B}_3 = \begin{smallarray}{cccc}
- 0 &  0 &  0 & \text{-}1 \\
- 0 &  0 &  1 &  0 \\
- 0 & \text{-}1 &  0 &  0 \\
- 1 &  0 &  0 &  0 \\
-\end{smallarray}
-\]
-
-The product of $\mt B$ with an arbitrary vector $\vt x$ results in
-\begin{equation}
-\label{eq:product_B_vector}
-\mt B \vt x = \begin{smallarray}{ccc} \hat{\mt B}_1 \vt q & \hat{\mt B}_2 \vt q & \hat{\mt B}_3 \vt q \end{smallarray} \begin{smallarray}{c} x_1 \\ x_2 \\ x_3 \end{smallarray} = \sum_{j=1}^3 x_j \hat{\mt B}_j \vt q.
-\end{equation}
-
-Similarly, the product of $\tr{\mt B}$ with an arbitrary quaternion $\vt \pi$ results in
-\begin{equation}
-\label{eq:vector_entries}
-\tr{\mt B}\vt \pi =
-\begin{smallarray}{c}
-\tr{\vt q}\tr{\hat{\mt B}}_1 \\
-\tr{\vt q}\tr{\hat{\mt B}}_2 \\
-\tr{\vt q}\tr{\hat{\mt B}}_3 \\
-\end{smallarray}\vt \pi = 
-\begin{smallarray}{c}
-\tr{\vt q}\tr{\hat{\mt B}}_1\vt \pi \\
-\tr{\vt q}\tr{\hat{\mt B}}_2\vt \pi \\
-\tr{\vt q}\tr{\hat{\mt B}}_3\vt \pi \\
-\end{smallarray} = 
-\begin{smallarray}{c}
-\tr{\vt \pi}{\hat{\mt B}}_1\vt q \\
-\tr{\vt \pi}{\hat{\mt B}}_2\vt q \\
-\tr{\vt \pi}{\hat{\mt B}}_3\vt q \\
-\end{smallarray}.
-\end{equation}
-
-With $\dot{\mt B} = \begin{smallarray}{ccc} \hat{\mt B}_1\dot{\vt q} & \hat{\mt B}_2\dot{\vt q} & \hat{\mt B}_3\dot{\vt q}\end{smallarray}$ being the time-derivative of matrix $\mt B$, we now focus on the product $\tr{\mt B}\dot{\mt B}$, whose each entry is $\{\tr{\mt B}\dot{\mt B}\}_{ij} = \tr{\vt q}\tr{\hat{\mt B}_i} \hat{\mt B}_j\dot{\vt q}$.
-By noting that products involving the permutation matrices satisfy the relations\cite{Dichmann1999}
-\begin{equation}
-\label{eq:BB_products}
-\begin{aligned}
-&\tr{\hat{\mt B}}_1 \hat{\mt B}_1 = \tr{\hat{\mt B}}_2 \hat{\mt B}_2 = \tr{\hat{\mt B}}_3 \hat{\mt B}_3 = \mt 1_4, \\
-&\tr{\hat{\mt B}}_1 \hat{\mt B}_2 = -\tr{\hat{\mt B}}_2 \hat{\mt B}_1 = -\tr{\hat{\mt B}_3}, \\
-&\tr{\hat{\mt B}}_2 \hat{\mt B}_3 = -\tr{\hat{\mt B}}_3 \hat{\mt B}_2 = -\tr{\hat{\mt B}_1}, \; \text{and} \\
-&\tr{\hat{\mt B}}_3 \hat{\mt B}_1 = -\tr{\hat{\mt B}}_1 \hat{\mt B}_3 = -\tr{\hat{\mt B}_2}, \\
-\end{aligned}
-\end{equation}
-and recalling that $\vt q$ and $\dot{\vt q}$ are mutually orthogonal (Eq.~\ref{eq:diff_qTq}), we can obtain a simplified expression for $\tr{\mt B}\dot{\mt B}$, which is
-\begin{equation}
-\label{eq:BtBdot}
-\tr{\mt B}\dot{\mt B} = \begin{smallarray}{ccc}
-0 & -\tr{\vt q} \tr{\hat{\mt B}_3}\dot{\vt q} &  \tr{\vt q} \tr{\hat{\mt B}_2}\dot{\vt q} \\
-\tr{\vt q} \tr{\hat{\mt B}_3}\dot{\vt q} & 0 & -\tr{\vt q} \tr{\hat{\mt B}_1}\dot{\vt q} \\
--\tr{\vt q} \tr{\hat{\mt B}_2}\dot{\vt q} & \tr{\vt q} \tr{\hat{\mt B}_1}\dot{\vt q} & 0
-\end{smallarray}.
-\end{equation}
-
-Skew-symmetric matrices like the one above are associated to cross products.
-For arbitrary vectors $\vt x$ and $\vt y$ in $\mathbb R^3$, it turns out that $\vt x \times \vt y = \mt S(\vt x)\vt y$, where $\mt S(\cdot)$ is an operator defined as
-\begin{equation}
-\label{eq:operator_S}
-\mt S(\vt x) = \begin{smallarray}{ccc}
-0   & -x_3 &  x_2 \\
-x_3 &  0   & -x_1 \\
--x_2 &  x_1 &  0
-\end{smallarray}.
-\end{equation}
-
-By contrasting the definition above with Eqs.~\ref{eq:vector_entries} and \ref{eq:BtBdot}, it becomes clear that
-\begin{equation}
-\label{eq:relation_B_qdot}
-\tr{\mt B}\dot{\mt B} = {\mt S}\left( \tr{\mt B}\dot{\vt q} \right).
-\end{equation}
-
-We can also express $\mt C = \begin{smallarray}{ccc} \hat{\mt C}_1{\vt q} & \hat{\mt C}_2{\vt q} & \hat{\mt C}_3{\vt q}\end{smallarray}$, where $\hat{\mt C}_1$, $\hat{\mt C}_2$, and $\hat{\mt C}_3$ are the following skew-symmetric permutation matrices:
-\[
-\hat{\mt C}_1 = \begin{smallarray}{cccc}
- 0 & \text{-}1 &  0 &  0 \\
- 1 &  0 &  0 &  0 \\
- 0 &  0 &  0 & \text{-}1 \\
- 0 &  0 &  1 &  0 \\
-\end{smallarray} \quad
-\hat{\mt C}_2 = \begin{smallarray}{cccc}
- 0 &  0 & \text{-}1 &  0 \\
- 0 &  0 &  0 &  1 \\
- 1 &  0 &  0 &  0 \\
- 0 & \text{-}1 &  0 &  0 \\
-\end{smallarray} \quad
-\hat{\mt C}_3 = \begin{smallarray}{cccc}
- 0 &  0 &  0 & \text{-}1 \\
- 0 &  0 & \text{-}1 &  0 \\
- 0 &  1 &  0 &  0 \\
- 1 &  0 &  0 &  0 \\
-\end{smallarray}
-\]
-
-Products involving these matrices are slightly different from those in Eq.~\ref{eq:BB_products}. They are
-\begin{equation}
-\label{eq:CC_products}
-\begin{aligned}
-&\tr{\hat{\mt C}}_1 \hat{\mt C}_1 = \tr{\hat{\mt C}}_2 \hat{\mt C}_2 = \tr{\hat{\mt C}}_3 \hat{\mt C}_3 = \mt 1_4, \\
-&\tr{\hat{\mt C}}_1 \hat{\mt C}_2 = -\tr{\hat{\mt C}}_2 \hat{\mt C}_1 = \tr{\hat{\mt C}_3}, \\
-&\tr{\hat{\mt C}}_2 \hat{\mt C}_3 = -\tr{\hat{\mt C}}_3 \hat{\mt C}_2 = \tr{\hat{\mt C}_1}, \; \text{and} \\
-&\tr{\hat{\mt C}}_3 \hat{\mt C}_1 = -\tr{\hat{\mt C}}_1 \hat{\mt C}_3 = \tr{\hat{\mt C}_2}. \\
-\end{aligned}
-\end{equation}
-
-A special connection between the permutation matrices that make $\mt B$ and $\mt C$ is the fact that $\tr{\hat{\mt B}_i}{\hat{\mt C}_j}$ is symmetric for any combination of $i$ and $j$.
-A direct consequence is that $\tr{\vt q}\tr{\hat{\mt B}}_i\hat{\mt C}_j\dot{\vt q} = \tr{\dot{\vt q}}\tr{\hat{\mt B}}_i\hat{\mt C}_j{\vt q}$, which means that $\tr{\mt B}\dot{\mt C} = \tr{\dot{\mt B}}{\mt C}$.
-Thus, differentiating Eq.~\ref{eq:factorization_of_A} with respect to time yields $\dot{\mt A} = \tr{\mt B}\dot{\mt C} + \tr{\dot{\mt B}}{\mt C} = 2 \tr{\dot{\mt B}}{\mt C}$.
-Note that Eq.~\ref{eq:basics_B} also applies for matrix $\mt C$, which means that, as far as $\tr{\vt q}\vt q = 1$, we can insert the product $\mt C\tr{\mt C}$ between $\tr{\mt B}$ and $\dot{\mt B}$ in Eq.~\ref{eq:relation_B_qdot} without altering its result.
-Then, using Eq.~\ref{eq:factorization_of_A} again and the conclusion above about $\dot{\mt A}$, we have that
-\begin{equation}
-\frac{1}{2} {\mt A}\tr{\dot{\mt A}} = {\mt S}\left( \tr{\mt B}\dot{\vt q} \right).
-\end{equation}
-
-As mentioned in Sec.~\ref{sec:kinematics}, ${\mt A}\tr{\dot{\mt A}} = \mt S(\vt \omega)$, which makes the equation above equivalent to Eq.~\ref{eq:relation_qdot_omega}.
-
-\section{Potential energy differentiation}
-\label{sec:Diff_PotEng}
-
-Care must be taken when differentiating the potential energy with respect to $\vt q$ because the quaternion components are interdependent.
-According to Schay,\cite{Schay1995} when a vector variable $\vt x$ is subject to a constraint $g(\vt x) = 0$, the derivative of $f$ with respect to $\vt x$ must be computed by
-\[
-\diff{f}{\vt x} = (\vt 1 - \vt n \tr{\vt n})\left( \diff{f}{\vt x} \right)^\ast,
-\]
-where $\vt n$ is the unit vector proportional to $(\partial g/\partial \vt x)^\ast$, with the asterisk denoting a differentiation performed without considering the interdependence of the entries of $\vt x$.
-Note that transpositions are necessary when comparing the equation above, based on the gradient convention, with that in Ref.~\onlinecite{Schay1995}, where the Jacobian convention is used.
-In the case of a unit quaternion,
-\[
-g(\vt q) = \tr{\vt q}\vt q - 1 = 0 \; \rightarrow \; \left(\diff{g}{\vt q}\right)^\ast = 2 \vt q.
-\]
-
-Hence, the unit vector proportional to $(\partial g/\partial \vt x)^\ast$ is exactly $\vt q$.
-As the same constraint makes ${\mt C}\tr{\mt C} = \mt 1_4 - \vt q\tr{\vt q}$ (in analogy to Eq.~\ref{eq:basics_B}), the constrained derivative of $U$ can be represented by
-\begin{equation}
-\label{eq:diff_projection}
-\diff{U}{\vt q} = {\mt C}\tr{\mt C} \left( \diff{U}{\vt q} \right)^\ast.
-\end{equation}
-
-Therefore, the constrained derivative is the projection of the unconstrained one in the hyperplane orthogonal to $\vt q$.
-Recently, Nielsen and Krenk\cite{Nielsen2012} employed a Lagrange multiplier approach to demonstrate the need of carrying out such a projection.
-Schay\cite{Schay1995, Schay1998} discussed on the equivalence of these two approaches.
-
-By applying the chain rule, we have
-\[
-\left( \diff{U}{\vt q} \right)^\ast = \sum_{j=1}^{n_p} \left( \diff{\vt R_j}{\vt q} \right)^\ast \diff{U}{\vt R_j} = - \sum_{j=1}^{n_p} \left( \diff{\vt \delta_j}{\vt q} \right)^\ast {\vt F_j},
-\]
-where $\vt R_j = \vt r + \vt \delta_j$ and $\vt F_j = -\partial U/\partial \vt R_j$ is, by definition, the force applied on particle $j$.
-Temporarily dropping the particle index for simplicity, we recall from Sec.~\ref{sec:kinematics} that $\vt \delta = \tr{\mt A}\vt d$, where $\vt d$ is a constant vector.
-As each entry of $\mt A = \tr{\mt B}{\mt C}$ is $A_{ij}=\tr{\vt q} \tr{\hat{\mt B}}_i \hat{\mt C}_j {\vt q}$, an entry of $\vt \delta$ is given by
-\[
-\delta_j = \sum_{i=1}^{3} d_i \tr{\vt q} \tr{\hat{\mt C}}_j \hat{\mt B}_i {\vt q}.
-\]
-
-Due to the symmetry of $\tr{\hat{\mt C}}_j \hat{\mt B}_i$, an unconstrained differentiation results in
-\[
-\left(\diff{\delta_j}{\vt q}\right)^\ast = 2 \sum_{i=1}^{3} d_i \tr{\hat{\mt C}}_j \hat{\mt B}_i {\vt q} = -2{\hat{\mt C}}_j {\mt B}{\vt d},
-\]
-where we have factored out $\tr{\hat{\mt C}}_j = -\hat{\mt C}_j$ from the summation and then employed Eq.~\ref{eq:product_B_vector}.
-The constrained differential $\partial \vt \delta/\partial \vt q$ is, therefore,
-\[
-\diff{\vt \delta}{\vt q} = -2{\mt C}\tr{\mt C}
-\begin{smallarray}{ccc}
-{\hat{\mt C}}_1 {\mt B}{\vt d} & {\hat{\mt C}}_2 {\mt B}{\vt d} & {\hat{\mt C}}_3 {\mt B}{\vt d}
-\end{smallarray} = 2{\mt C}\;{\mt S}\!\left(\tr{\mt C}{\mt B}\vt d\right),
-\]
-where we have carried out the second matrix product above, employed the properties presented in Eq.~\ref{eq:CC_products}, and then used the operator defined in Eq.~\ref{eq:operator_S}.
-Finally, after realizing that $\tr{\mt C}{\mt B}\vt d = \tr{\mt A}{\vt d} = \vt \delta$, we can use the result above to obtain
-\[
-\diff{U}{\vt q} = - 2 \mt C \sum_{j=1}^{n_p} \mt S(\vt \delta_j) {\vt F_j}  = - 2 \mt C \sum_{j=1}^{n_p} \vt \delta_j \times {\vt F_j}.
-\]
-
-Therefore, the constrained derivative of $U$ with respect to $\vt q$ is proportional to the quaternion-fixed representation of the resultant torque over the body.
-
-\section{Derivation of the Euler equation and its solution for a torque-free rotation}
-\label{sec:derivation_euler}
-
-Recall that $\vt \pi$ is twice the quaternion-fixed version of the angular momentum vector.
-Thus, it is expected that Eq.~\ref{eq:EDO_pi} will result in the well-known Euler equation, which describes the evolution of the angular momentum in the body-fixed frame, if $\frac{1}{2} \tr{\mt B}$ is premultiplied to both sides.
-In order to show that this actually occurs, we start by carrying out the premultiplication to obtain
-\[
-\frac{1}{2} \tr{\mt B} \dot{\vt \pi} = \frac{1}{4} \tr{\mt B} \mt \Omega \vt \omega + \mt A \vt \tau.
-\]
-
-Once $\mt \Omega = \begin{smallarray}{ccc} \hat{\mt B}_1{\vt \pi} & \hat{\mt B}_2{\vt \pi} & \hat{\mt B}_3{\vt \pi}\end{smallarray}$, in analogy with Eq.~\ref{eq:BtBdot} we conclude that $\tr{\mt B} \mt \Omega = \mt S(\tr{\mt B}\vt \pi) = 2 \mt S(\mt I \vt \omega)$.
-In addition,
-\[
-\tr{\mt B} \dot{\vt \pi} = \frac{d(\tr{\mt B}\vt \pi)}{dt} - \tr{\dot{\mt B}}\vt \pi = 2 \mt I\dot{\vt \omega} - 2 \tr{\dot{\mt B}} \mt B \mt I \vt \omega,
-\]
-where we have inserted $\mt B \tr{\mt B}$ between $\tr{\dot{\mt B}}$ and $\vt \pi$ and substituted $\tr{\mt B} \vt \pi = 2 \mt I \vt \omega$.
-Finally, after transposing Eq.~\ref{eq:relation_B_qdot} with $\tr{\mt B}\dot{\vt q}$ replaced by $\frac{1}{2} \vt \omega$ and resorting to the relation between $\mt S(\cdot)$ and a cross product, we obtain
-\[
-\mt I \dot{\vt \omega} + \vt \omega \times (\mt I \vt \omega) = \mt A \vt \tau,
-\]
-which is exactly the Euler equation of motion for rigid bodies.\cite{Goldstein2002} Its main advantage, in comparison with Eq.~\ref{eq:EDO_pi}, is that it completely decouples from the quaternion $\vt q$ when $\vt \tau = \vt 0$.
-In such a torque-free situation, the vector equation can be rewritten as the following system:
-\begin{align*}
-I_1 \dot{\omega}_1 - (I_2 - I_3) \omega_2 \omega_3 = 0\\
-I_2 \dot{\omega}_2 + (I_1 - I_3) \omega_1 \omega_3 = 0 \\
-I_3 \dot{\omega}_3 - (I_1 - I_2) \omega_1 \omega_2 = 0
-\end{align*}
-
-Analytical solutions for this system are a textbook subject.\cite{Landau1976} Here we present the solution for the case under consideration, when $I_1 > I_2 > I_3$ (an asymmetrical top), in a form that aims to facilitate computer implementation, inspired by the forms presented in Refs.~\onlinecite{Landau1976} and \onlinecite{Thorn2012}.
-In a free rotation, the kinetic energy $K_r = \tfrac{1}{2} \tr{\vt \omega}{\mt I}{\vt \omega}$ and the space-frame angular momentum $\vt L = \tfrac{1}{2}\tr{\mt C}{\vt \pi} = \tr{\mt A}{\mt I}{\vt \omega}$ are integrals of motion.
-Note that the square norm of $\vt L$ does not depend explicitly on $\vt q$ in any frame of reference:
-\[
-L^2 = \tr{\vt L}{\vt L} = \tr{\vt \omega}{\mt I}^2{\vt \omega} = \frac{1}{4}\tr{\vt \pi}{\vt \pi}.
-\]
-
-Therefore, if we compute $K_r$ and $L^2$ from the initial angular velocity ${\vt \omega}^0 = \tr{[\omega_1^0 \; \omega_2^0 \; \omega_3^0]}$, we can apply the following constraints to the Euler equations:
-\begin{align*}
-I_1 \omega_1^2 + I_2 \omega_2^2 + I_3 \omega_3^2 = 2 K_r \\
-I_1^2 \omega_1^2 + I_2^2 \omega_2^2 + I_3^2 \omega_3^2 = L^2
-\end{align*}
-
-By solving these equations for $\omega_1^2$ and $\omega_3^2$, we obtain
-\begin{equation}
-\label{eq:omegas_1_and_3}
-\omega_1^2 = A_1^2 \left[1-\left(\frac{\omega_2}{\lambda_1}\right)^2\right] \;\; \text{and} \;\; \omega_3^2 = A_3^2 \left[1-\left(\frac{\omega_2}{\lambda_3}\right)^2\right],
-\end{equation}
-where
-\begin{equation}
-\lambda_1 = \sqrt{\frac{L^2 - 2 I_3 K_r}{I_2(I_2 - I_3)}} \;\; \text{and} \;\; \lambda_3 = \sqrt{\frac{2 I_1 K_r - L^2}{I_2(I_1 - I_2)}}.
-\end{equation}
-
-In order to obtain the prefactors $A_1$ and $A_3$ with the correct signs, we realize that Eqs.~\ref{eq:omegas_1_and_3} must be valid at the initial instant ($t = 0$), which makes
-\begin{equation}
-A_1 = \frac{\omega_1^0}{\sqrt{1 - (\omega_2^0/\lambda_1)^2}} \;\; \text{and} \;\;
-A_3 = \frac{\omega_3^0}{\sqrt{1 - (\omega_2^0/\lambda_3)^2}}.
-\end{equation}
-
-By defining $x = {\omega_2}/{\lambda_\text{min}}$ and $\kappa = {\lambda_\text{min}}/{\lambda_\text{max}}$, where $\lambda_\text{min} = \min(\lambda_1,\lambda_3)$ and $\lambda_\text{max} = \max(\lambda_1,\lambda_3)$, we can transform the Euler equation for $\omega_2$ into to the following initial value problem for $x$:
-\[
-\begin{cases}
-I_2 \lambda_\text{min} \dot{x} + (I_1-I_3) A_1 A_3 \sqrt{(1-x^2)(1-\kappa^2x^2)} = 0 \\
-x(0) = \omega_2^0/\lambda_\text{min}.
-\end{cases}
-\]
-
-The solution of the equation above and its application to
-
-can be expressed using the elliptic Jacobi function $\text{sn}$, so that we can calculate $\omega_2$ at any time as
-\[
-\omega_2(t) = \lambda_\text{min} \text{sn}(\alpha t + \epsilon,\kappa),
-\]
-where
-\[
-\alpha = -\frac{(I_1-I_3)A_1A_3}{\lambda_\text{min} I_2} \quad \text{and} \quad \epsilon = \text{arcsn}\left(\frac{\omega_{2,0}}{\lambda_\text{min}},\kappa\right).
-\]
-
 \bibliography{rigid_bodies}
-
+%\begin{table*}
+%	\caption{Results with Exact Solution for Free Rotations}
+%	\label{table:NVE}
+%	\begin{ruledtabular}
+%		\begin{tabular}{cccccccc}
+%			$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+%			\hline
+%			1.0 & $	305.91	\pm	0.03	$ & $	-8395.1	\pm	0.1	$ & $	1645.9	\pm	0.1	$ & $	823.2	\pm	0.1	$ & $	822.7	\pm	0.1	$ & $	146	\pm	2	$ & $0.0456$ \\
+%			2.0 & $	305.21	\pm	0.02	$ & $	-8389.5	\pm	0.1	$ & $	1642.1	\pm	0.1	$ & $	823.0	\pm	0.1	$ & $	819.1	\pm	0.1	$ & $	149	\pm	2	$ & $0.656$ \\
+%			3.0 & $	304.30	\pm	0.03	$ & $	-8375.4	\pm	0.9	$ & $	1637.2	\pm	0.1	$ & $	824.0	\pm	0.1	$ & $	813.3	\pm	0.1	$ & $	156	\pm	2	$ & $2.87$ \\
+%			4.0 & $	304.5	\pm	0.5	$ & $	-8339	\pm	6	$ & $	1638	\pm	3	$ & $	829	\pm	1	$ & $	808.8	\pm	0.6	$ & $	180	\pm	2	$ & $10.1$ \\
+%		\end{tabular}
+%	\end{ruledtabular}
+%\end{table*}
+
+%\begin{table*}
+%\caption{Results.....}
+%\label{table:NVE}
+%\begin{ruledtabular}
+%\begin{tabular}{cccccccc}
+%$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+%\hline
+%1.0 & $305.95\pm0.03$ & $-8395.4\pm0.1$ & $1646.1\pm0.1$ & $823.3\pm0.1$ & $822.9\pm0.1$ & $151\pm2$ & $0.0471$ \\
+%2.0 & $305.17\pm0.03$ & $-8389.0\pm0.1$ & $1641.9\pm0.1$ & $823.2\pm0.1$ & $818.8\pm0.1$ & $148\pm2$ & $0.853$ \\
+%3.0 & $304.34\pm0.03$ & $-8374\pm2$ & $1637.5\pm0.2$ & $824.0\pm0.1$ & $813.4\pm0.1$ & $152\pm2$ & $3.96$ \\
+%4.0 & $303.6\pm0.4$ & $-8349\pm5$ & $1633\pm2$ & $826.8\pm0.2$ & $806.5\pm0.4$ & $168\pm2$ & $8.25$ \\
+%\end{tabular}
+%\end{ruledtabular}
+%\end{table*}
+%\begin{table}
+%	\begin{threeparttable}
+%		\caption{Average deviation of the conserved quantity ($DE$) and average temperatures from liquid water simulation with NVT integration \tnote{a}\tnote{b}}
+%		\label{table:denvt}
+%		\begin{ruledtabular}
+%			\begin{tabular}{ccccc}
+%				& \multicolumn{2}{c}{Eq.~\ref{eq:trotter_splitting_NHC}} & \multicolumn{2}{c}{Eq.~%\ref{eq:modified_splitting}} \\
+%				\cline{2-5}
+%				$\Delta t$/fs & $D\!E$ & $T$/K & $D\!E$ & $T$/K \\
+%				\hline
+%				1.0 & 0.000043 & 298.16  & 0.000052  & 298.15 \\
+%				2.0 & 0.00013  & 298.23  & 0.00017   & 298.06 \\
+%				3.0 & 0.00089  & 298.17  & 0.0013    & 298.12
+%			\end{tabular}
+%		\end{ruledtabular}
+%		\begin{tablenotes}
+%			\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$.
+%			\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs.
+%		\end{tablenotes}
+%	\end{threeparttable}
+%\end{table}
 \end{document}

### 2017-01-31 rigid_bodies.tex
nve review

@@
 
-\section{New Theory}
-Exact solution for torque-free rotation
-In a previous paper\cite{Abreu_2017} we presented a formulation for the motion of rigid bodies which was simplified by the introduction of a particular factorization of the rotational matrix. 
-quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ and satisfy the relation\cite{Goldstein2002}
-\begin{equation}
-\label{eq:norm_eq_1}
-\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1.
-\end{equation}
+\section{Introduction}
+\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
+systematic errors
+\begin{itemize}
+\item[1] Kamberaj\cite{Kamberaj2005}\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
+assumption equipartition of the kinetic energy between the translational and rotational degrees of freedom. We demonstrate that this assumption is not valid and it dramatically affects the drift.
+to our knowledge, have not been reported in the
+order of the operators
+\item[2] our paper review
+\item[3] coarse grained: 
+Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations.
+This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
+Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.
+One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
+Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion.
+This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles.
+However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
+This is expected to be computationally more efficient because the constraints are treated implicitly.
+In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
+Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
+This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth.
+It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
 
-Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
+\end{itemize}
+
+In this paper, we report formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases.
+Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
+This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom.
+We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
+For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
+
+The paper is organized as follows.
+In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
+Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure.
+Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
+
+\subsection{Microcanonical Ensemble: A Review}
+algebraic manipulations
+Prompted by the introduction of a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
+Nos\'{e}-Hoover chain thermostat.
+the non-symplectic character, we used the approximate solution for rotation, as we high quality and non-Hamiltonian dynamics admitos no backward error analyisis.
+We now will briefly review the key expressions of the metioned paper.
+First, recall that the translational and rotational degrees of freedom are decoupled.
+Mathematicaly, the position of a particle $j$,  constituent of a rigid body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference.
+Note that unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ are employed to represent the orientations of rigid bodies, which satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\vt q} = 0$.
+The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
 \begin{equation*}
@@
 2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
-\end{smallarray}.
+\end{smallarray},
 \end{equation*}
-
-Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
+which admits the following factorization:
 \begin{equation}
-\label{eq:diff_qTq}
-\tr{\vt q}\dot{\vt q} = 0,
+\label{eq:factorization_of_A}
+{\mt A} = \tr{\mt B}{\mt C}.
 \end{equation} 
-which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, the rate of change $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.
-
-Let us now introduce two new matrix-valued functions that depend on the quaternion components as
+that depend on the quaternion components as
 \begin{equation}
@@
 \end{equation}
-
-The most remarkable feature is that
-\begin{equation}
-\label{eq:factorization_of_A}
-{\mt A} = \tr{\mt B}{\mt C}.
-\end{equation}
-
-The equation above is a simple and convenient factorization of matrix $\mt A$, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters.
-As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the quaternion components, which is
+Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} aconvenient factorization of matrix $\mt A$.
+This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rates of change of the quaternion components, which is
 \begin{equation}
@@
 \end{equation}
-
+It avoids the use of an ficticious angular velocity component, as Miller~\textit{et al.}\cite{Miller2002}.
+An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
 Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.
 
-\section{Introduction}
-\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
-\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
-Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations.
-This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
-Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.
+For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), and kinetic energy as follows:
 
-One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
-Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion.
-This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles.
-However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
-This is expected to be computationally more efficient because the constraints are treated implicitly.
-In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
-Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
-This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth.
-It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
+\begin{equation}
+\label{eq:H_NVE}
+\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right),
+\end{equation}
 
-In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases.
-Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
-This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom.
-We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
-For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
+\begin{equation}
+\label{eq:H_split_omega}
+\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
+\end{equation}
 
-The paper is organized as follows.
-In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions.
-In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
-Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure.
-Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
-
-Throughout this paper, we use the word ``quaternion'' as a synonym for a four-dimensional vector, avoiding other interpretations.
-Thus, only ordinary matrix/vector operations are present in the formulation.
-Some mathematical details are given in the Appendix.
-
-\section{Mechanics of Rigid Body Motion}
-\label{sec:mechanics}
-
-In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study.
-Unit quaternions are employed to represent the orientations of rigid bodies.
-Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector.
-An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
-
-\subsection{Kinematics}
-
-\subsection{Hamiltonian Dynamics}
-\label{sec:hamiltonian}
-
-The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body.
-As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively.
-The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$.
-we obtain the following Hamiltonian system of equations of motion for a rigid body:
+where an index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, with $m_i$ is the mass and  $\mt I_i$ the inertia tensor which in the principal frame is a diagonal matrix withe entries given by ($I_1$,$I_2$,$I_3$).
+The equations of motion
 \begin{subequations}
@@
 &\dot{\vt p} = \mt F \\
-&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
-&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
-\end{align}
-\end{subequations}
-
-Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually.
-Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
-\begin{equation}
-\label{eq:omega_entry}
-\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
-\end{equation}
-where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body can be rewritten as
-\begin{equation}
-\label{eq:H_split_omega}
-\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
-\end{equation}
-
-Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
-\begin{subequations}
-\label{eq:EDO_system_alternative}
-\begin{align}
 &\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
@@
 
-\subsection{Symplectic Numerical Integration}
-\label{sec:symplectic}
+consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
+\begin{equation}
+\label{eq:omega_entry}
+\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
+\end{equation}
+where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math} of the aforementione peper.
 
-An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$.
-Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element.
-\subsection{Exact Solution for Free Rotation}
-\label{sec:free_rotation}
+%Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
+%\begin{subequations}
+%\label{eq:EDO_system_alternative}
+%\begin{align}
+%&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
+%&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau %\label{eq:edo_pi}
+%\end{align}
+%\end{subequations}
 
-\section{Molecular Dynamics}
-\label{sec:molecular_dynamics}
-
-In this section, we present factorization schemes for generating molecular dynamics algorithms in the microcanonical (NVE) and canonical (NVT) ensembles.
-In both cases, a three-dimensional system is considered as consisting of $N$ interacting, non-collinear rigid bodies occupying a fixed volume $V$.
-An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$).
-Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta.
+The phase space of the whole system contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta.
 Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.
-
-\subsection{Microcanonical Ensemble}
-
-Considering the system described above, with $\vt r$ and $\vt q$ now representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
-\begin{equation}
-\label{eq:H_NVE}
-\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
-\end{equation}
-
-As usual, sampling of a microcanonical ensemble can be carried out by a Hamiltonian dynamics in phase space, which will have $\mathcal{H} = E$ as an integral of motion.
-The dynamics of this system is governed by a Liouville operator expressed as in Eq.~\ref{eq:full_operator}, but now with individual operators defined as
+A Hamiltonian dynamics in phase space will have $\mathcal{H} = E$ as an integral of motion.
+The dynamics of this system is governed by a Liouville operator with individual operators defined as
 \[
@@
 
-Therefore, a symplectic integrator can be devised using the same splitting scheme of Sec.~\ref{sec:symplectic}, with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system.
+Therefore, a symplectic integrator can be devised using a splitting scheme with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system.
 If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$.
 If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.
+
+By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete free rotation, the Liouville operator corresponding to the original case of Eq.~\eqref{eq:EDO_system} (as best noticed in its alternative form, Eq.~\eqref{eq:EDO_system_alternative}) can be written as
+\begin{equation}
+\label{eq:full_operator_uniaxial}
+i\!L = i\!L_t + i\!L_b + i\!L_r,
+\end{equation}
+where $i\!L_t$ corresponds to the first term in the right-hand of Eq.~\eqref{eq:tras_rot_op}. The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
+\begin{equation}
+\label{eq:trotter_splitting_NVE}
+e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
+\end{equation}
+
+Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is split as
+\begin{equation}
+\label{eq:splitting_rot}
+e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t}{2} i\!L_3} e^{\frac{\delta t}{2} i\!L_2} e^{\delta t i\!L_1} e^{\frac{\delta t}{2} i\!L_2} e^{\frac{\delta t}{2} i\!L_3} \right),
+\end{equation}
+where ${\delta t} = {\Delta t}/n$.
+This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
 

### 2017-01-31 rigid_bodies.tex
nve

@@
 
+The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002}.
+some algebra manipulations where we defined $\mt \Omega = \mt B(\vt \pi)$, it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$.
+For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
+\[
+\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
+\]
+where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference and thus $\vt F$ is the resultant force on the body. As demonstrated in Abreu, we have
+\[
+\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
+\]
+where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame.
+e.
+Finally, after carrying out the required differentiations of $\mathcal{H}$  and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
+\begin{subequations}
+\label{eq:EDO_system}
+\begin{align}
+&\dot{\vt r} = \frac{1}{m} \vt p \\
+&\dot{\vt p} = \mt F \\
+&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
+&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
+\end{align}
+\end{subequations}
+
+we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\eqref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
+\begin{equation}
+\label{eq:omega_entry}
+\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
+\end{equation}
+where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body can be rewritten as
 \begin{equation}
@@
 
-where an index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, with $m_i$ is the mass and  $\mt I_i$ the inertia tensor which in the principal frame is a diagonal matrix withe entries given by ($I_1$,$I_2$,$I_3$).
-The equations of motion
+Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\eqref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi}:
 \begin{subequations}
-\label{eq:EDO_system}
+\label{eq:EDO_system_alternative}
 \begin{align}
-&\dot{\vt r} = \frac{1}{m} \vt p \\
-&\dot{\vt p} = \mt F \\
-&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
-&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
+&\dot{\vt q} = \frac{1}{2} \left( \sum_{k=1}^3 \omega_k \hat{\mt B}_k \right) \vt q \label{eq:edo_q} \\
+&\dot{\vt \pi} = \frac{1}{2} \left( \sum_{k=1}^3 \omega_k \hat{\mt B}_k \right) \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
 \end{align}
@@
 
-consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
-\begin{equation}
-\label{eq:omega_entry}
-\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
-\end{equation}
-where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math} of the aforementione peper.
-
-%Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
-%\begin{subequations}
-%\label{eq:EDO_system_alternative}
-%\begin{align}
-%&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
-%&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau %\label{eq:edo_pi}
-%\end{align}
-%\end{subequations}
+where an index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, with $m_i$ is the mass and  $\mt I_i$ the inertia tensor which in the principal frame is a diagonal matrix withe entries given by ($I_1$,$I_2$,$I_3$).
 

### 2017-01-31 rigid_bodies.tex
nve review

@@
 
-\subsection{Microcanonical Ensemble: A Review}
-algebraic manipulations
-Prompted by the introduction of a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
-Nos\'{e}-Hoover chain thermostat.
-the non-symplectic character, we used the approximate solution for rotation, as we high quality and non-Hamiltonian dynamics admitos no backward error analyisis.
-We now will briefly review the key expressions of the metioned paper.
-First, recall that the translational and rotational degrees of freedom are decoupled.
-Mathematicaly, the position of a particle $j$,  constituent of a rigid body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference.
+\section{Theory}
+
+Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
+In the following, we skip over the mathematical details of the derivations, which can be found in Abreu~\textit{et al.}\cite{Abreu_2017}, and briefly reviewed some of the final expressions.
+In the aforementioned paper, we introduced an analytical solution for free-rotation, which would be usefull if any backward error analysis is to be carried out.
+However, the  Nos\'{e}-Hoover chain thermostat are non-Hamiltonian, so any backward error analysis it is not feasible anymore.
+Given that, in this paper we employ the approximate solution for the rotational motion , which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision, in terms of the integration erros, almost identical to that achieved using the analytical solution.
+
+\subsection{Microcanonical Ensemble}
+
+When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled.
+Mathematicaly, the position of a particle $j$,  constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference.
 Note that unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ are employed to represent the orientations of rigid bodies, which satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\vt q} = 0$.
@@
 
-For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), and kinetic energy as follows:
+For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows~\cite{Abreu_2017}
 
@@
 The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002}.
-some algebra manipulations where we defined $\mt \Omega = \mt B(\vt \pi)$, it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$.
+In order to Eq.~\eqref{eq:H_NVE} to be easily derivable by $\vt \pi$ we defined $\mt \Omega = \mt B(\vt \pi)$ and it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$.
 For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
@@
 
+As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = \begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
+\[
+\hat{\mt B}_1 = \begin{smallarray}{cccc}
+ 0 & \text{-}1 &  0 &  0 \\
+ 1 &  0 &  0 &  0 \\
+ 0 &  0 &  0 &  1 \\
+ 0 &  0 & \text{-}1 &  0 \\
+\end{smallarray} \quad
+\hat{\mt B}_2 = \begin{smallarray}{cccc}
+ 0 &  0 & \text{-}1 &  0 \\
+ 0 &  0 &  0 & \text{-}1 \\
+ 1 &  0 &  0 &  0 \\
+ 0 &  1 &  0 &  0 \\
+\end{smallarray} \quad
+\hat{\mt B}_3 = \begin{smallarray}{cccc}
+ 0 &  0 &  0 & \text{-}1 \\
+ 0 &  0 &  1 &  0 \\
+ 0 & \text{-}1 &  0 &  0 \\
+ 1 &  0 &  0 &  0 \\
+\end{smallarray}
+\]
+
 we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\eqref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
@@
 \end{equation}
-where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body can be rewritten as
+where $\hat{\mt B}_k$ is one of the permutation matrices. In this context, the Hamiltonian of the rigid body can be rewritten as
 \begin{equation}
@@
 \end{equation}
-where $i\!L_t$ corresponds to the first term in the right-hand of Eq.~\eqref{eq:tras_rot_op}. The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
+The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
 \begin{equation}
@@
 Here, we propose and compare two distinct measure-preserving integrators for a rigid-body system coupled to a single thermostat chain.
-In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, as it follows from Eq.~\ref{eq:nhc_system},
+In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:trotter_splitting_NVE} and, as it follows from Eq.~\ref{eq:nhc_system},
 \begin{equation}

### 2017-02-02 rigid_bodies.tex
canonical ensemble reviewed

--- 2017-01-31 rigid_bodies.tex
@@
 We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
-For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
+For this, we derive two alternative expressions: %one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
 
-The paper is organized as follows.
+%The paper is organized as follows.
 In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
@@
 
-\subsection{Microcanonical Ensemble: A Review}
-algebraic manipulations
-Prompted by the introduction of a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
-Nos\'{e}-Hoover chain thermostat.
-the non-symplectic character, we used the approximate solution for rotation, as we high quality and non-Hamiltonian dynamics admitos no backward error analyisis.
-We now will briefly review the key expressions of the metioned paper.
-First, recall that the translational and rotational degrees of freedom are decoupled.
-Mathematicaly, the position of a particle $j$,  constituent of a rigid body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference.
-Note that unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ are employed to represent the orientations of rigid bodies, which satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\vt q} = 0$.
+\section{Theory}
+
+Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
+In the aforementioned paper, we introduced an analytical solution for torque-free rotations, which would be particularly useful if any backward error analysis\cite{Hairer_2006} is to be carried out.
+However, the  Nos\'{e}-Hoover chain thermostat introduces non-Hamiltonian components, so any backward error analysis it is not feasible anymore.
+Given that, in this paper, the thermostatted version of the equations of motion will be generated considering the approximate solution for the rotational motion, which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision notably similar to that achieved using the analytical solution.
+In the next subsection, we will skip over the mathematical details and restrict ourselves to some fundamental expressions on which the dynamics in the canonical (NVT) ensemble will be developed.
+For full details in the NVE ensemble the reader is referred to our previous paper.\cite{Abreu_2017}
+
+\subsection{Microcanonical Ensemble}
+
+When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled.
+Mathematically, the position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$.
+In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\dot{\vt q}} = 0$.
+
 The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
@@
 \label{eq:factorization_of_A}
-{\mt A} = \tr{\mt B}{\mt C}.
+{\mt A} = \tr{\mt B}{\mt C},
 \end{equation} 
-that depend on the quaternion components as
+where
 \begin{equation}
@@
 \end{equation}
-Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} aconvenient factorization of matrix $\mt A$.
-This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rates of change of the quaternion components, which is
+Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} a convenient factorization of matrix $\mt A$.
+This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components, which is
 \begin{equation}
@@
 \end{equation}
-It avoids the use of an ficticious angular velocity component, as Miller~\textit{et al.}\cite{Miller2002}.
-An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
-Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.
+The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
+For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.
 
-For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), and kinetic energy as follows:
+In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}.
+In this equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a digonal matrix in the principal frame.
+For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}
 
@@
 \end{equation}
+where the index $i$ ranging from $1$ to $N$ indicates the quantities related to each body.
 
-\begin{equation}
-\label{eq:H_split_omega}
-\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
-\end{equation}
-
-where an index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, with $m_i$ is the mass and  $\mt I_i$ the inertia tensor which in the principal frame is a diagonal matrix withe entries given by ($I_1$,$I_2$,$I_3$).
-The equations of motion
+The equations of motion are obtained from the Hamiltonian by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}.
+For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
+\[
+\diff{U}{\vt r_i} = -\vt F_i = -\sum_{j=1}^{n_p} {\vt F_j},
+\]
+where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference and thus $\vt F_i$ is the resultant force on the body.
+As demonstrated in Abreu~\textit{et al.},\citep{Abreu_2017} $\partial{U}/\partial \vt q_i$ is given by
+\[
+\diff{U}{\vt q_i} = -2 \mt C_i \vt \tau_i = -2 \mt C_i \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
+\]
+where $\vt \tau_i$ is the resultant torque exerted on the body and ${\vt \delta}_j$ is the position of the particle $j$, both expressed in the space-fixed frame.
+The position ${\vt \delta}_j$ is related to ${\vt d}_j$ by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} After carrying out the required differentiations of $\mathcal{H}$  we obtain the following Hamiltonian system of equations of motion for a rigid body:
 \begin{subequations}
@@
 \begin{align}
-&\dot{\vt r} = \frac{1}{m} \vt p \\
-&\dot{\vt p} = \mt F \\
-&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
-&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
+&\dot{\vt r}_i = \frac{1}{m_i} \vt p_i \\
+&\dot{\vt p}_i = \mt F_i \\
+&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:EDO_q} \\
+&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i \label{eq:EDO_pi}
 \end{align}
 \end{subequations}
+with $\mt \Omega_i = \mt B_i(\vt \pi_i)$.
 
-consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
+The approximate solution for a complete rotation is obtained by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ employing each entry of $\vt \omega$: 
 \begin{equation}
@@
 \end{equation}
-where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math} of the aforementione peper.
+where $\hat{\mt B}_k$ is one of the following permutation matrices:
+\[
+\hat{\mt B}_1 = \begin{smallarray}{cccc}
+ 0 & \text{-}1 &  0 &  0 \\
+ 1 &  0 &  0 &  0 \\
+ 0 &  0 &  0 &  1 \\
+ 0 &  0 & \text{-}1 &  0 \\
+\end{smallarray} \quad
+\hat{\mt B}_2 = \begin{smallarray}{cccc}
+ 0 &  0 & \text{-}1 &  0 \\
+ 0 &  0 &  0 & \text{-}1 \\
+ 1 &  0 &  0 &  0 \\
+ 0 &  1 &  0 &  0 \\
+\end{smallarray} \quad
+\hat{\mt B}_3 = \begin{smallarray}{cccc}
+ 0 &  0 &  0 & \text{-}1 \\
+ 0 &  0 &  1 &  0 \\
+ 0 & \text{-}1 &  0 &  0 \\
+ 1 &  0 &  0 &  0 \\
+\end{smallarray}
+\]
+The new form of the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi} are then:
+\begin{subequations}
+\label{eq:EDO_system_alternative}
+\begin{align}
+&\dot{\vt q}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt q_i \label{eq:edo_q} \\
+&\dot{\vt \pi}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt \pi_i + 2 \mt C_i \vt \tau_i \label{eq:edo_pi}
+\end{align}
+\end{subequations}
 
-%Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
-%\begin{subequations}
-%\label{eq:EDO_system_alternative}
-%\begin{align}
-%&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
-%&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau %\label{eq:edo_pi}
-%\end{align}
-%\end{subequations}
-
-The phase space of the whole system contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta.
-Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.
-A Hamiltonian dynamics in phase space will have $\mathcal{H} = E$ as an integral of motion.
-The dynamics of this system is governed by a Liouville operator with individual operators defined as
+In this context, the Hamiltonian of the rigid body can be rewritten as
+\begin{equation}
+\label{eq:H_split_omega}
+\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
+\end{equation}
+from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions.
+In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$.
+Of coarse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$.
+The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
+\begin{equation}
+\label{eq:full_operator}
+i\!L_{NVE} = i\!L_t + i\!L_b + i\!L_r,
+\end{equation}
+where the individual operators are defined as follows:
 \[
 \begin{split}
+&boost:\\
 i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
+&translation: \\
  i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
-i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right).
+&uniaxial \; rotation:\\
+ i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
 \end{split}
 \]
-
-Therefore, a symplectic integrator can be devised using a splitting scheme with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system.
-If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$.
-If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.
-
-By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete free rotation, the Liouville operator corresponding to the original case of Eq.~\eqref{eq:EDO_system} (as best noticed in its alternative form, Eq.~\eqref{eq:EDO_system_alternative}) can be written as
-\begin{equation}
-\label{eq:full_operator_uniaxial}
-i\!L = i\!L_t + i\!L_b + i\!L_r,
-\end{equation}
-where $i\!L_t$ corresponds to the first term in the right-hand of Eq.~\eqref{eq:tras_rot_op}. The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
+using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation.
+The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
 \begin{equation}
 \label{eq:trotter_splitting_NVE}
-e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
+e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b},
 \end{equation}
-
-Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is split as
+Update of forces and torques is required only after the free translation and free rotation have been carried out.
+Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is splitted as
 \begin{equation}
 \label{eq:splitting_rot}
-e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t}{2} i\!L_3} e^{\frac{\delta t}{2} i\!L_2} e^{\delta t i\!L_1} e^{\frac{\delta t}{2} i\!L_2} e^{\frac{\delta t}{2} i\!L_3} \right),
+e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
 \end{equation}
-where ${\delta t} = {\Delta t}/n$.
-This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
+where ${\Delta t}$ is the time step size.
+
+As demonstrated in our previous paper,\cite{Abreu2017}, the action of the propagators 
+\begin{align*}
+&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0,
+\end{align*}
+
+\begin{align*}
+&{\vt p} = {\vt p}_0 + {\vt F}_0 t, \\
+&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t.
+\end{align*}
+
+\begin{align*}
+&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \ \text{and} \\
+&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0, 
+\end{align*}
 
@@
 \end{equation}
-where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$.
+where $\mathcal H$ is the same as in Eq.~\eqref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$.
 As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
@@
 
-Finally, as the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
-Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain become
+Given that the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
+Therefore, the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain become
 \begin{subequations}
@@
 
-In the equations above, ${\vt \omega}_i = \frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
+In the equations above, ${\vt \omega}_i$ $=$ $\frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
 \begin{align*}
@@
 
-Note that the $3N$ momentum components in Eq.~\ref{eq:nhc_p}, the $4N$ components in Eq.~\ref{eq:nhc_pi}, and the thermostat momenta in Eq.~\ref{eq:nhc_p_eta} affect their own time-derivatives.
+Note that the $3N$ momentum components in Eq.~\eqref{eq:nhc_p}, the $4N$ components in Eq.~\eqref{eq:nhc_pi}, and the thermostat momenta in Eq.~\eqref{eq:nhc_p_eta} affect their own time-derivatives.
 Thus, the thermostatted system has a phase-space compressibility $\kappa = -7N \alpha_1 - \sum_j \alpha_{j+1}$.
-Resorting to Eq.~\ref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
+Resorting to Eq.~\eqref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
 \begin{equation}
@@
 
-Therefore, if the function $w(\vt x)$ that derives from Eq.~\ref{eq:relation_kappa_w} also satisfies Eq.~\ref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$.
+Therefore, if the function $w(\vt x)$ that derives from Eq.~\eqref{eq:relation_kappa_w} also satisfies Eq.~\eqref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$.
 Here, we propose and compare two distinct measure-preserving integrators for a rigid-body system coupled to a single thermostat chain.
-In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, as it follows from Eq.~\ref{eq:nhc_system},
+In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\eqref{eq:full_operator} and, as it follows from Eq.~\eqref{eq:nhc_system},
 \begin{equation}
@@
 
-Then, by retaining the NVE factorization presented in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator for each time step as
+Then, by retaining the NVE factorization presented in Eq.~\eqref{eq:trotter_splitting_NVE}, we write the full exponential operator for each time step as
 \begin{equation}
@@
 
-The Nos\'e-Hoover chain propagator is further split by making $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_\text{NHC}^{(j)}$ and, afterwards,
-\[
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \Pi_{j=1}^n \Bigg[ \prod_{j=M}^1 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
-\]
+The boost, translation, and rotation propagators already satisfy the criterion in Eq.~\eqref{eq:split_kappa_w} with the function $w$ expressed in Eq.~\eqref{eq:nhc_measure} (they are all incompressible and keep $w$ unchanged).
+The NHC propagator is further split into factors that also satisfy Eq.~\eqref{eq:split_kappa_w}, whichnare given by
 
-The boost, translation, and rotation propagators already satisfy the criterion in Eq.~\ref{eq:split_kappa_w} with the function $w$ expressed in Eq.~\ref{eq:nhc_measure} (they are all incompressible and keep $w$ unchanged).
-The NHC propagator is further split into factors that also satisfy Eq.~\ref{eq:split_kappa_w}.
-
-In order to satisfy the criterion in Eq.~\ref{eq:split_kappa_w} for all factors, given the function $w$ expressed in Eq.~\ref{eq:nhc_measure}, we further split the NHC contribution as $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_\text{NHC}^{(j)}$, where
 \begin{subequations}
@@
 \end{subequations}
-with $j$ varying from $1$ to $M-1$ in Eq.~\ref{eq:iL_NHC_j}. Then, each propagator $e^{\frac{\Delta t}{2} i\!L_\text{NHC}}$ present in Eq.~\ref{eq:iL_NHC} is factorized as
+with $j$ varying from $1$ to $M-1$ in Eq.~\eqref{eq:iL_NHC_j}. Then, each propagator $e^{\frac{\Delta t}{2} i\!L_\text{NHC}}$ present in Eq.~\eqref{eq:iL_NHC} is factorized as
 \[
@@
 
-In turn, the propagator $e^{t i\!L_\text{NHC}^{(j)}}$ defined in Eq.~\ref{eq:iL_NHC_j} behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged, whose analytical solution is
+In turn, the propagator $e^{t i\!L_\text{NHC}^{(j)}}$ defined in Eq.~\eqref{eq:iL_NHC_j} behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged, whose analytical solution is
 \begin{align*}
@@
 
-------------
-
 The second splitting scheme we propose here relies on the definition of the operator
@@
 i\!L_b^\ast &= \sum_{i=1}^N \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} \\
-&+ \sum_{i=1}^N \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} + \alpha_1 \diff{}{\eta_1}.
+&+ \sum_{i=1}^N \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} + \alpha_1 \diff{}{\eta_1},
 \end{align*}
-
-It is worth noting that an analytical solution can also be derived for the system of equations $\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i$ for all $i \in [1,N]$ if all other variables are constant.
-These equations are equivalent in form to the one solved above, so that their solutions are
+with which analytical solutions can also be derived for the system of equations $\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i$ for all $i \in [1,N]$ if all other variables are constant.
+These equations are equivalent in form to the ones solved above, so we get
 \begin{align*}
@@
 
-\begin{subequations}
-\label{eq:solution_momenta}
-\begin{align}
-&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \quad \text{and} \label{eq:solution_p} \\
-&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t. \label{eq:solution_pi}
-\end{align}
-\end{subequations}
-
-By evaluating these functions simultaneously for all $i$, we accomplish the effect of a propagator $e^{t i\!L_b^\ast}$, where $i\!L_b^\ast$ is the Liouville operator given by
-\[
-i\!L_b^\ast = \sum_{i=1}^N \left[ \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} \right].
-\]
-
-This is a kind of boost that already incorporates the influence of the thermostat chain.
-One of the aims of this paper is to compare the performance of the integrator given by Eq.~\ref{eq:trotter_splitting_NHC} to that obtained when the solution in Eq.~\ref{eq:solution_momenta} is considered.
-For that, we also define a modified Nos\'e-Hoover chain propagator $e^{i\!L^\ast_\text{NHC}}$, as follows:
-\begin{equation}
-\begin{split}
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } = &\prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-\times &\exp\left(-\frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j}\right) \\
-\times &\prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
-\end{split}
-\end{equation}
-
 In the new propagator, $e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} }$ is removed:
@@
 
----------
 We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj2005} does not fulfill this criterion. 
-
-Of course, Eqs.~\ref{eq:solution_eta} and \ref{eq:solution_p_eta} are the ones used to evaluate the propagators above. Finally, the complete scheme is
+Finally, the complete scheme is
 \begin{equation}
@@
 \end{subequations}
-where $W_\text{a}$ is the atomic virial given by Eq.~\ref{eq:virial_atoms}, while $\vt F_{ij}$ is the total action of body $j$ (and its images) over body $i$ and $\vt f_{ik}$ is the resultant force exerted on the $k$-th atom of body $i$, respectively given by
+where $W_\text{a}$ is the atomic virial given by Eq.~\eqref{eq:virial_atoms}, while $\vt F_{ij}$ is the total action of body $j$ (and its images) over body $i$ and $\vt f_{ik}$ is the resultant force exerted on the $k$-th atom of body $i$, respectively given by
 \
@@
 \end{equation}
-from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $\mathcal N$.
+from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $\mathcal N$.
 It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
-\textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\ref{eq:splitting_rot}}.
-Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the average temperatures obtained in each case.
-It can be seen that both integrators, represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:modified_splitting}, systematically achieve the specified temperature, regardless of the time step size.
+\textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}.
+Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$, in Table~\eqref{table:denvt} we show the results of Eq.~\eqref{eq:performance} as well as the average temperatures obtained in each case.
+It can be seen that both integrators, represented by the Eq.~\eqref{eq:trotter_splitting_NHC} and Eq.~\eqref{eq:modified_splitting}, systematically achieve the specified temperature, regardless of the time step size.
 In addition, the energy drift is, at its worst, around $1 \%$.
@@
 \end{equation}
-Taking the logarithm of Eq.~\ref{eq:probability_ratio}, we obtain
+Taking the logarithm of Eq.~\eqref{eq:probability_ratio}, we obtain
 \begin{equation}
@@
 
-The Eq.~\ref{eq:ln_probability_ratio} is the basis for the statistical analysis in the ensemble validation, which determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
+The Eq.~\eqref{eq:ln_probability_ratio} is the basis for the statistical analysis in the ensemble validation, which determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
 Note that it is independent of the unknown free energies.
@@
 
-In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\ref{eq:trotter_splitting_NHC}.
+In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
 In figure legends we show the averages temperatures simulated.
 It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\ref{eq:modified_splitting}.
+Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.
 
@@
 	\caption{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
-Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
+Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
 	\label{fig:checkensemble}
@@
 	\includegraphics{maxwell-botlzamm-paper}
-	\caption{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\ref{eq:mb}.
-Simulation of liquid water using Eq.~\ref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}$ = 304.99~K.}
+	\caption{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\eqref{eq:mb}.
+Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}$ = 304.99~K.}
 \label{fig:mbdistribution}
@@
 				\hline % inserts single-line 
-				Eq.~\ref{eq:trotter_splitting_NHC} & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
-				Eq.~\ref{eq:modified_splitting}    & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14
+				Eq.~\eqref{eq:trotter_splitting_NHC} & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
+				Eq.~\eqref{eq:modified_splitting}    & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14
 		\end{tabular}
@@
 %\end{table}
+
+%The Nos\'e-Hoover chain propagator is further split by making $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_%\text{NHC}^{(j)}$ and, afterwards,
+%\[
+%e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \Pi_{j=1}^n \Bigg[ \prod_{j=M}^1 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
+%\]
+
+%In order to satisfy the criterion in Eq.~\eqref{eq:split_kappa_w} for all factors, given the function $w$ expressed in Eq.~\eqref{eq:nhc_measure}, we further split the NHC contribution as $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_\text{NHC}^{(j)}$, where
+
+%\begin{subequations}
+%\label{eq:solution_momenta}
+%\begin{align}
+%&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) %t \quad \text{and} \label{eq:solution_p} \\
+%&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi%\left(\alpha_1 t \right) t. \label{eq:solution_pi}
+%\end{align}
+%\end{subequations}
+
+%By evaluating these functions simultaneously for all $i$, we accomplish the effect of a propagator $e^{t i%\!L_b^\ast}$, where $i\!L_b^\ast$ is the Liouville operator given by
+%\[
+%i\!L_b^\ast = \sum_{i=1}^N \left[ \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} \right].
+%\]
+%We also define a modified Nos\'e-Hoover chain propagator $e^{i\!L^\ast_\text{NHC}}$, as follows:
+%\begin{equation}
+%\begin{split}
+%e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } = &\prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
+%\times &\exp\left(-\frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j}\right) \\
+%\times &\prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
+%\end{split}
+%\end{equation}
+
 \end{document}

### 2017-02-02 rigid_bodies.tex
nvt review
canonical review

@@
 &boost:\\
-i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
+&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
 &translation: \\
- i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
+& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
 &uniaxial \; rotation:\\
- i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
+& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
 \end{split}
 \]
-using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation.
-The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
+using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation. As explained in our previous paper,\cite{Abreu2017} those individual operators can be applied analytically according to
+\[
+\begin{split}
+&boost:\\
+&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
+&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
+&translation:\\
+&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
+&uniaxial \; rotation:\\
+&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
+&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
+\end{split}
+\]
+
+Applying the Trotter-susz The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
 \begin{equation}
@@
 Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is splitted as
-\begin{equation}
+\begin{align*}
 \label{eq:splitting_rot}
 e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
-\end{equation}
+\end{align*}
 where ${\Delta t}$ is the time step size.
 
-As demonstrated in our previous paper,\cite{Abreu2017}, the action of the propagators 
-\begin{align*}
-&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0,
-\end{align*}
-
-\begin{align*}
-&{\vt p} = {\vt p}_0 + {\vt F}_0 t, \\
-&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t.
-\end{align*}
-
-\begin{align*}
-&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \ \text{and} \\
-&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0, 
-\end{align*}
+equipartion of energy
 
@@
 \subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}
-
+\label{sec:numerical_solvers}
 As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
@@
 
+We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
+Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
+However...the energy drift...
+
+\section{Results}
+\label{sec:numerical_results}
  Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. 
-Note that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
-Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
-
-\begin{subequations}
-	\label{eq:virial_rigid_bodies}
-	\begin{align}
-	W &= \frac{1}{3} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \tr{\hat{\vt r}}_{ij} \vt F_{ij} \\
-	  &= W_\text{a} - \frac{1}{3} \sum_{i=1}^N \sum_{k=1}^{{n_p}_i} {\tr{\vt \delta}_k}_i {{\vt f}_k}_i,
-	\end{align}
-\end{subequations}
-where $W_\text{a}$ is the atomic virial given by Eq.~\eqref{eq:virial_atoms}, while $\vt F_{ij}$ is the total action of body $j$ (and its images) over body $i$ and $\vt f_{ik}$ is the resultant force exerted on the $k$-th atom of body $i$, respectively given by
-\
-\[
-\vt F_{ij} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \vt f_{ijkl} \quad \text{and} \quad \vt f_{ik} = \sum_{\substack{j=1\\j \neq i}}^N \sum_{l=1}^{{n_p}_j} \vt f_{ijkl}.
-\]
-
-\section{Numerical Results}
-\label{sec:numerical_results}
-
-In this section, we analyze some performance aspects of the integration schemes when applied to a molecular system in the liquid state.
-For this, we simulated a system containing 356 TIP3P water molecules\cite{Price2004} at a density of 999.56 $kg/m^3$.
+In this section, we analyze the integration schemes introduced in Subsection~\ref{sec:numerical_solvers} and compare the results with those obtained employing the method of Kamberaj~\cite{Kamberaj2005}.
+For that, we simulated a system containing 356 TIP3P water molecules\cite{Price2004} at a density of 999.56 $kg/m^3$.
 The intermolecular interactions were truncated at 9 \AA.
@@
 
-\subsection{Numerical Stability}
-\label{sec:performance}
-
-In order to evaluate the stability of the numerical integrators, we employ a measure given by\cite{Tuckerman2010}
-\begin{equation}
-\label{eq:performance}
-D\!E =  \frac{1}{\mathcal N} \sum_{t=1}^{\mathcal N} \left| \frac{E_t - E_0}{E_0} \right|,
-\end{equation}
-from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $\mathcal N$.
 It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
@@
 In addition, the energy drift is, at its worst, around $1 \%$.
-
-\subsection{New Results}
 
@@
 
-\subsection{Partition of Energy}
-\label{sec:energypartition}
-
-\subsection{Sampling Validation}
+\section{Sampling Validation}
 \label{sec:samplingvalidation}
@@
 
+%In order to evaluate the stability of the numerical integrators, we employ a measure given by%\cite{Tuckerman2010}
+%\begin{equation}
+%\label{eq:performance}
+%D\!E =  \frac{1}{\mathcal N} \sum_{t=1}^{\mathcal N} \left| \frac{E_t - E_0}{E_0} \right|,
+%\end{equation}
+%from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ %(either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ %after a certain number of time steps $\mathcal N$.
+
 \end{document}

### 2017-02-03 rigid_bodies.tex
check ensemble

@@
 \section{Introduction}
+Nos\'{e}-Hoover thermostat is routinely used in Molecular Dynamics to maintain the temperature
 \cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
@@
 
+\subsection{Sampling Validation}
+\label{sec:samplingvalidation}
+
+In this subsection we briefly review a simple method that serves for checking the consistency between the sampled and theoretical energy distributions.
+The procedures implemented by Shirts ~\cite{Shirts2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
+Fist recall that in the NVT ensemble the probability of observing an energy E is given by
+\[
+P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right),
+\]
+where $ \Omega(E) $ is the density of states, which does not depend on $ \beta $, and $ Q(\beta) $  is the canonical partition function, linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$.
+
+Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get
+\begin{equation}
+\frac{P(E/\beta_2)}{P(E/\beta_1)} = \exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
+\end{equation}
+Taking the logarithm of the equation above, we obtain
+\begin{equation}
+\label{eq:ln_probability_ratio}
+\ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E,
+\end{equation}
+which is the basis for the statistical analysis carried out in the ensemble validation that determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
+Note that it is independent of the unknown free energies.
+For a guidance on how to choose the temperature gap the reader is referred to the aforementioned paper.
+
+Furthermore,  it is possible to examine the consistency between the sampled kinetic energy $K$ and the corresponding Maxwell-Boltzmann distribution, which is given by the following expression:
+\begin{equation}
+\label{eq:mb}
+P(K) = \frac{\beta}{\sqrt{N_{dof}\pi}} \exp\left[-\frac{\left(\beta K - \frac{N_{dof}}{2}\right)^2}{N_{dof}} \right],
+\end{equation}
+where $N_{dof}$ is the number of degrees of freedom. Eq.~\eqref{eq:mb} is a normal distribution, with average  $ \langle K \rangle = N_{dof}/2\beta $ and variance $\sigma^2 = N_{dof}/2\beta^2$.
+
 \section{Results}
 \label{sec:numerical_results}
- Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. 
-In this section, we analyze the integration schemes introduced in Subsection~\ref{sec:numerical_solvers} and compare the results with those obtained employing the method of Kamberaj~\cite{Kamberaj2005}.
-For that, we simulated a system containing 356 TIP3P water molecules\cite{Price2004} at a density of 999.56 $kg/m^3$.
-The intermolecular interactions were truncated at 9 \AA.
-The shifted-force (SF) method\cite{Allen1989, Fennell2006, Toxvaerd_2011} was used to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
 
-It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
-\textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}.
-Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$, in Table~\eqref{table:denvt} we show the results of Eq.~\eqref{eq:performance} as well as the average temperatures obtained in each case.
-It can be seen that both integrators, represented by the Eq.~\eqref{eq:trotter_splitting_NHC} and Eq.~\eqref{eq:modified_splitting}, systematically achieve the specified temperature, regardless of the time step size.
-In addition, the energy drift is, at its worst, around $1 \%$.
+CÓDIGO: CHECAR FACTORIZACIÓN Y DELTA T DE LOS OPERADORES DE NHC
+
+In this section, we analyze the numerical integrators introduced in Subsection~\ref{sec:numerical_solvers}, and compare the results with those obtained employing the method of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS\cite{Plimpton1995} software package.
+For that, we considered a system of 900 TIP3P\cite{Jorgensen_1983} water molecules in a cubic box of length {\AA} subject to periodic boundary conditions.
+The intermolecular interactions were truncated at 9 \AA and to avoid a discontinuity in the force, we applied the so-called shifted-force (SF) method\cite{Allen1989, Fennell2006, Toxvaerd_2011} to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
+
+It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
+First we present some results as first mented.
+That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
+In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
+By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
+
+Resultados sobre o checkensemble: We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
+Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
+m-b distribution valid when equipartition of energy is satisfied.
+
+Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
+Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.
+
+\begin{table}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
+	\label{table:new_results}
+	\begin{ruledtabular}
+		\begin{tabular}{cccc}
+			& \multicolumn{3}{c}{ Temperature (K) } \\
+			\cline{2-4}
+			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			\hline
+			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
+			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
+			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
+		\end{tabular}
+	\end{ruledtabular}
+\end{table}
+
+\begin{table}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
+	\label{table:new_results_2}
+	\begin{ruledtabular}
+		\begin{tabular}{cccc}
+			& \multicolumn{3}{c}{Drift rate [(kcal/mol)/ns]} \\
+			\cline{2-4}
+			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			\hline
+			1.0 & $0.0328$ & $0.108$ & $0.023$ \\
+			2.0 & $0.597$  & $1.31$  & $17.0$ \\
+			3.0 & $3.44$   & $2.53$  & $53.2$
+		\end{tabular}
+	\end{ruledtabular}
+\end{table}
 
@@
 \end{table*}
-
-\begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
-	\label{table:new_results}
-	\begin{ruledtabular}
-		\begin{tabular}{cccc}
-			& \multicolumn{3}{c}{ Temperature (K) } \\
-			\cline{2-4}
-			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
-			\hline
-			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
-			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
-			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
-		\end{tabular}
-	\end{ruledtabular}
-\end{table}
-
-\begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
-	\label{table:new_results_2}
-	\begin{ruledtabular}
-		\begin{tabular}{cccc}
-			& \multicolumn{3}{c}{Drift rate [(kcal/mol)/ns]} \\
-			\cline{2-4}
-			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
-			\hline
-			1.0 & $0.0328$ & $0.108$ & $0.023$ \\
-			2.0 & $0.597$  & $1.31$  & $17.0$ \\
-			3.0 & $3.44$   & $2.53$  & $53.2$
-		\end{tabular}
-	\end{ruledtabular}
-\end{table}
-
-\section{Sampling Validation}
-\label{sec:samplingvalidation}
-
-In this subsection we present the results from a simple method that serves for checking the consistency between the sampled energy distributions and the theoretical ones.
-The procedure, introduced and implemented by Shirts ~\cite{Shirts2013}, relies on a quantitative statistical analysis of fitting and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
-This last approach does not involve any binning of the energies, hence no discretization error is present.
-We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
-
-Recall that in the NVT ensemble the probability of observing an energy E is given by
-\[
-P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right),
-\]
-where $ \beta $ is the Boltzmann's constant, $ \Omega(E) $ is the density of states which does not depend on $ \beta $ , and $ Q(\beta) $  is the canonical partition function, linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$.
-
-Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get
-\begin{equation}
-\label{eq:probability_ratio}
-\frac{P(E/\beta_2)}{P(E/\beta_1)} = \exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
-\end{equation}
-Taking the logarithm of Eq.~\eqref{eq:probability_ratio}, we obtain
-\begin{equation}
-\label{eq:ln_probability_ratio}
-\ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E.
-\end{equation}
-
-The Eq.~\eqref{eq:ln_probability_ratio} is the basis for the statistical analysis in the ensemble validation, which determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
-Note that it is independent of the unknown free energies.
-For a guidance on how to choose the temperature gap the reader is referred to the aforementioned paper.
-Given that the potential and kinetic energies are separable, the analysis can be made separately as follows
-
-\begin{equation}
-\label{eq:probability_ratio_pot}
-\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  \exp\left[(-\beta_2 - \beta_1)E_{pot}\right]
-\end{equation}
-
-\begin{equation}
-\label{eq:probability_ratio_kin}
-\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  \exp\left[(-\beta_2 - \beta_1)E_{kin}\right]
-\end{equation}
-Furthermore,  it is possible to examine the consistency between the sampled kinetic energy and the corresponding Maxwell-Boltzmann distribution, which is given by the following expression:
-\begin{equation}
-\label{eq:mb}
-P(E_{cin}) = \frac{\beta}{\sqrt{N_{DOF}\pi}} \exp\left[-\frac{\left(\beta E_{kin} - \frac{N_{DOF}}{2}\right)^2}{N_{DOF}} \right],
-\end{equation}
-where $N_{DOF}$ is the number of degrees of freedom.
-This last equations is a normal distribution characterized by an average of $ \langle E_{kin} \rangle = (N_{DOF}/2\beta) $ and variance $\sigma^2 = N_{DOF}/2\beta^2$, valid when equipartition of energy is satisfied.
 
@@
 \label{sec:conclusion}
-
+Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. 
+by contrast
 \begin{acknowledgments}
@@
 %from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ %(either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ %after a certain number of time steps $\mathcal N$.
+% \textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}.
 
+%\begin{equation}
+%\label{eq:probability_ratio_pot}
+%\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  \exp\left[(-\beta_2 - \beta_1)E_{pot}\right]
+%\end{equation}
+
+%\begin{equation}
+%\label{eq:probability_ratio_kin}
+%\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  \exp\left[(-%\beta_2 - \beta_1)E_{kin}\right]
+%\end{equation}
 \end{document}

### 2017-02-07 rigid_bodies.tex
intro

@@
 \begin{abstract}
-In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
+%In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
 Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model.
@@
 \section{Introduction}
-Nos\'{e}-Hoover thermostat is routinely used in Molecular Dynamics to maintain the temperature
-\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
-systematic errors
-\begin{itemize}
-\item[1] Kamberaj\cite{Kamberaj2005}\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
-assumption equipartition of the kinetic energy between the translational and rotational degrees of freedom. We demonstrate that this assumption is not valid and it dramatically affects the drift.
-to our knowledge, have not been reported in the
-order of the operators
-\item[2] our paper review
-\item[3] coarse grained: 
-Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations.
-This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few.
-Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.
-One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations.
-Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion.
-This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles.
-However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
+
+Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena.
+Under those approximations, certain Statistical Mechanics laws might not been necessarily translated into assumptions in Molecular Simulation.
+For instance, it has already been shown that the unavoidalbe truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and the consequence is that the temperature becomes an undefinite property.
+In a previous paper,\citep{Abreu_2017} we arrived same now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble.
+From that observation, we inspected the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, on which two thermostat chains act separately on the rotational and the translational degrees of freedom and it is not able to reproduced de specified temperature.
+
+The importance of rigid body MD stems from the fact that its employment is not limited to molecules modelled as rigid bodies, like the ubiquitous case of water,  but also it increasingly tool as a coarse-grained strategy for sistems designed as colletions of rigid bodies like molecular motors, nanoparticles, proteins, etc.
+The method of Kamberaj~\textit{et al.}\cite{Kamberaj2005} was actually used in the simulation of small molecules\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}, as well as in molecular machines,\cite{Akimov2008, Konyukhov2010} and nanoparticles.\cite{Knorowski2012, Patra2013}
+
+Nos\'{e}-Hoover thermostat is routinely used in Molecular Dynamics to maintain the temperature.
+It extended phase strategy on which ficticius variables are introduced in order to control the fluctuations of energy and, so the temperature as well.
+The dynamic of those variables are non-Hamiltonian in nature, which makes the evaluation more difficult, comparing with Hamiltonian dynamics where backward error analysis is done.
+In fact, it has been used....
+In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with SHAKE and Hybrid Monte Carlo (HMC).\cite{Duane1987}  HMC is an exact method in the sense that the truncation erros are not present.
+And SHAKE is a iterative method that employ lagrange multiplier to solve the holonomic constraints used in MD simulations.For molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
 This is expected to be computationally more efficient because the constraints are treated implicitly.
 In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
-Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions.
-This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth.
-It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.
 
-\end{itemize}
-
-In this paper, we report formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases.
-Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
-This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom.
-We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
+In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce the numerical integrators.
+both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
+This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which .
+%We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
 For this, we derive two alternative expressions: %one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
 
-%The paper is organized as follows.
+The paper is organized as follows.
 In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.

### 2017-02-07 rigid_bodies.tex
intr
introduction

@@
 \begin{abstract}
-%In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
+In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
 Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model.
@@
 
-Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena.
-Under those approximations, certain Statistical Mechanics laws might not been necessarily translated into assumptions in Molecular Simulation.
-For instance, it has already been shown that the unavoidalbe truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and the consequence is that the temperature becomes an undefinite property.
-In a previous paper,\citep{Abreu_2017} we arrived same now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble.
-From that observation, we inspected the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, on which two thermostat chains act separately on the rotational and the translational degrees of freedom and it is not able to reproduced de specified temperature.
+Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} and requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena.
+Under those approximations, certain Statistical Mechanics laws might not been necessarily traduced into assumptions in Molecular Simulation.
+For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}.
+In a previous paper,\citep{Abreu_2017} we observed the same problem now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble.
+That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, which couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.
 
-The importance of rigid body MD stems from the fact that its employment is not limited to molecules modelled as rigid bodies, like the ubiquitous case of water,  but also it increasingly tool as a coarse-grained strategy for sistems designed as colletions of rigid bodies like molecular motors, nanoparticles, proteins, etc.
-The method of Kamberaj~\textit{et al.}\cite{Kamberaj2005} was actually used in the simulation of small molecules\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}, as well as in molecular machines,\cite{Akimov2008, Konyukhov2010} and nanoparticles.\cite{Knorowski2012, Patra2013}
+The importance of rigid-body Molecular Dynamics stems from the fact that it is not limited to simulate molecules modelled as rigid bodies only, like the ubiquitous case of water,  but also because it is a coarse-grained strategy used for sistems designed as collections of rigid bodies like proteins and nanoparticiles\cite{Knorowski2012, Patra2013}.
+The approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has actually been used in the simulation of small molecules\cite{Aimoli_2014}, as well as in molecular motors\cite{Yan_2008,Akimov_2012,Geiger_2013}.
 
-Nos\'{e}-Hoover thermostat is routinely used in Molecular Dynamics to maintain the temperature.
-It extended phase strategy on which ficticius variables are introduced in order to control the fluctuations of energy and, so the temperature as well.
-The dynamic of those variables are non-Hamiltonian in nature, which makes the evaluation more difficult, comparing with Hamiltonian dynamics where backward error analysis is done.
-In fact, it has been used....
-In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with SHAKE and Hybrid Monte Carlo (HMC).\cite{Duane1987}  HMC is an exact method in the sense that the truncation erros are not present.
-And SHAKE is a iterative method that employ lagrange multiplier to solve the holonomic constraints used in MD simulations.For molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach.
+The Nos\'{e}-Hoover thermostat, routinely used in Molecular Dynamics, relies on the ``extended'' phase space on which ficticius variables are introduced in order to control the fluctuations of kinetic energy of the physical variables maintaining the temperature of the physical system in its specified value.
+The dynamics of those variables introduces non-Hamiltonian components to the equations of motion, which makes its evaluation more limited, comparing with Hamiltonian dynamics where backward error analysis is possible.
+
+In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce time reversible and measure-preserving numerical integrators as well.
+both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
+Our starting point to develop the thermostatized equations of motion begining from our previous work on the microcanonical ensemble.
+
+In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with results obtained with Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present, point of reference.
+We also using SHAKE\cite{Ryckaert1977}, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints used in MD simulations.
+For molecules modeled as rigid structures the dynamics of rigid bodies emerges as a suitable approach.
 This is expected to be computationally more efficient because the constraints are treated implicitly.
-In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.
-
-In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce the numerical integrators.
-both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
-This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which .
-%We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions.
-For this, we derive two alternative expressions: %one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.
+The property of being symplectic implies the conservation of volume in phase space.
+the inverse statement is not true.
+measure-preserving
 

### 2017-02-17 rigid_bodies.tex
aa

@@
 
-\title{A Simplified Formulation for Molecular Dynamics with Rigid Bodies}
+\title{Revisiting the Nos\'{e}-Hoover chain thermostat applied to the Rigid Bodies Molecular Dynamics}
 
@@
 
-Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} and requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena.
-Under those approximations, certain Statistical Mechanics laws might not been necessarily traduced into assumptions in Molecular Simulation.
-For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}.
-In a previous paper,\citep{Abreu_2017} we observed the same problem now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble.
-That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, which couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.
+Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that makes possible a microscopic insight into complex physical phenomena.
+To mimic ``real'' systems, numerical approximation are required under which certain Statistical Mechanics laws cannot be assumed to be valid also in Molecular Simulation.
+For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integration of the equations of motion in Molecular Dynamics disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}.
+In a previous paper,\citep{Abreu_2017} we observed the same problem, but regarding the partition of the kinetic energy between translational and rotational degrees of freedom in Molecular Dynamics of rigid bodies in the microcanonical ensemble.
+That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.
 
-The importance of rigid-body Molecular Dynamics stems from the fact that it is not limited to simulate molecules modelled as rigid bodies only, like the ubiquitous case of water,  but also because it is a coarse-grained strategy used for sistems designed as collections of rigid bodies like proteins and nanoparticiles\cite{Knorowski2012, Patra2013}.
-The approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has actually been used in the simulation of small molecules\cite{Aimoli_2014}, as well as in molecular motors\cite{Yan_2008,Akimov_2012,Geiger_2013}.
+The relevance of rigid-body Molecular Dynamics stems from the fact that its application is not limited to molecules actually modelled as a rigid body, like the ubiquitous case of water, but also it constitues a simple coarse-grain strategy for systems that might be designed as collection of rigid bodies interconected, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}.
+As a matter of fact, the approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has been applied in the simulation of small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as in membranes~\cite{Bucior_2012} molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}, DNA\cite{Fathizadeh_2012}
 
-The Nos\'{e}-Hoover thermostat, routinely used in Molecular Dynamics, relies on the ``extended'' phase space on which ficticius variables are introduced in order to control the fluctuations of kinetic energy of the physical variables maintaining the temperature of the physical system in its specified value.
-The dynamics of those variables introduces non-Hamiltonian components to the equations of motion, which makes its evaluation more limited, comparing with Hamiltonian dynamics where backward error analysis is possible.
+The Nos\'{e}-Hoover chain thermostat\cite{Martyna1992}, routinely used in Molecular Dynamics, employs the extended phase-space approach and introduces ficticius variables in order to ``control'' the kinetic energy fluctuations of the physical variables which allows to maintain the system at a specified temperature.
+The dynamics of those variables introduces non-Hamiltonian components which makes infeasible to carry out the backward error analysis.
+Altough it is not possible to devise symplectic integrator, one can rigurously obtain measure-preserving numerical solvers\cite{Sergi2001,Ezra2004,Ezra2006} and obtain numerical stability.
 
-In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce time reversible and measure-preserving numerical integrators as well.
-both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat.
-Our starting point to develop the thermostatized equations of motion begining from our previous work on the microcanonical ensemble.
-
-In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with results obtained with Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present, point of reference.
-We also using SHAKE\cite{Ryckaert1977}, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints used in MD simulations.
-For molecules modeled as rigid structures the dynamics of rigid bodies emerges as a suitable approach.
-This is expected to be computationally more efficient because the constraints are treated implicitly.
-The property of being symplectic implies the conservation of volume in phase space.
-the inverse statement is not true.
-measure-preserving
+In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble and introduce time reversible and measure-preserving numerical integrators as well.
+Our startint point is the symplectic integrator in the microcanonical ensemble we introduced in a previous paper\citep{Abreu_2017}, which employs the unit quaterion representation for the rotational degrees of freedom.
+Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at the specified value.
+The analysis of the thermostats will be rather heuristic reliying on comparisons with results obtained employing the Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present.
+We will also include results from the SHAKE\cite{Ryckaert1977} method, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints.
+unavoidable errors but we shoud be aware of the effect on the properties of interest.
 
 The paper is organized as follows.
-In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
-Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure.
+In Sec.~\ref{sec:theory} we briefly review the equations of motion in the microcanonical ensemble, which are the basis for developing the corresponding dynamics in the canonical ensemble.
+In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
 Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
@@
 \section{Theory}
-
+\label{sec:theory}
 Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.

### 2017-04-01 rigid_bodies.tex
march 24
april1

@@
 
-\begin{abstract}
-In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
+%\begin{abstract}
+%In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
 Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model.
@@
 Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
-\end{abstract}
+%\end{abstract}
 
@@
 Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that makes possible a microscopic insight into complex physical phenomena.
-To mimic ``real'' systems, numerical approximation are required under which certain Statistical Mechanics laws cannot be assumed to be valid also in Molecular Simulation.
+To mimic ``real'' systems, numerical approximation are required under which certain Statistical Mechanics laws cannot be assumed to be valid.
 For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integration of the equations of motion in Molecular Dynamics disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}.
@@
 
-CÓDIGO: CHECAR FACTORIZACIÓN Y DELTA T DE LOS OPERADORES DE NHC
+%CÓDIGO: CHECAR FACTORIZACIÓN Y DELTA T DE LOS OPERADORES DE NHC
 
@@
 By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
+%Resultados sobre o checkensemble:
 
-Resultados sobre o checkensemble: We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
+checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
 Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
@@
 \end{document}
+\grid

### 2017-04-02 rigid_bodies.tex
introduction

@@
 
-\title{Revisiting the Nos\'{e}-Hoover chain thermostat applied to the Rigid Bodies Molecular Dynamics}
+\title{Revisiting the Nos\'{e}-Hoover chain thermostat in the Rigid Bodies Molecular Dynamics}
 
@@
 
-Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that makes possible a microscopic insight into complex physical phenomena.
-To mimic ``real'' systems, numerical approximation are required under which certain Statistical Mechanics laws cannot be assumed to be valid.
-For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integration of the equations of motion in Molecular Dynamics disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}.
-In a previous paper,\citep{Abreu_2017} we observed the same problem, but regarding the partition of the kinetic energy between translational and rotational degrees of freedom in Molecular Dynamics of rigid bodies in the microcanonical ensemble.
-That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.
+Molecular Simulation, being the computational realization of Statistical Mechanics\cite{Tuckerman2010}, involves several types of approximations in order to mimic ``real'' physical systems.
+Naively, one would expect those approximations to only produce deviation from experimental values.
+However, they actually may give rise to systematic errors introducing non-physical behaviour.
+For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics, disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property\cite{Eastwood_2010}.
+In a previous paper,\citep{Abreu_2017} we reported the same artificiality to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics.
+That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
 
-The relevance of rigid-body Molecular Dynamics stems from the fact that its application is not limited to molecules actually modelled as a rigid body, like the ubiquitous case of water, but also it constitues a simple coarse-grain strategy for systems that might be designed as collection of rigid bodies interconected, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}.
-As a matter of fact, the approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has been applied in the simulation of small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as in membranes~\cite{Bucior_2012} molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}, DNA\cite{Fathizadeh_2012}
+The relevance of rigid-body Molecular Dynamics stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water, but also it is a simple coarse-grain strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}.
+As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented in the softwares packages \textit{blue}-HOOMD and LAMMPS\cite{Plimpton1995}, has been applied to simulate small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as membranes~\cite{Bucior_2012}, molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}.
 
-The Nos\'{e}-Hoover chain thermostat\cite{Martyna1992}, routinely used in Molecular Dynamics, employs the extended phase-space approach and introduces ficticius variables in order to ``control'' the kinetic energy fluctuations of the physical variables which allows to maintain the system at a specified temperature.
-The dynamics of those variables introduces non-Hamiltonian components which makes infeasible to carry out the backward error analysis.
-Altough it is not possible to devise symplectic integrator, one can rigurously obtain measure-preserving numerical solvers\cite{Sergi2001,Ezra2004,Ezra2006} and obtain numerical stability.
+The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in Molecular Dynamics, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
+The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with what backward error analysis is not feasible anymore.
+Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}
 
-In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble and introduce time reversible and measure-preserving numerical integrators as well.
-Our startint point is the symplectic integrator in the microcanonical ensemble we introduced in a previous paper\citep{Abreu_2017}, which employs the unit quaterion representation for the rotational degrees of freedom.
-Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at the specified value.
-The analysis of the thermostats will be rather heuristic reliying on comparisons with results obtained employing the Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present.
-We will also include results from the SHAKE\cite{Ryckaert1977} method, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints.
-unavoidable errors but we shoud be aware of the effect on the properties of interest.
+In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body Molecular Dynamics in the canonical ensemble.
+The starting point is our previous work on the microcanonical ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
+Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
+The analysis of the integrators will be rather heuristic, relying on comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy.
+We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
 

### 2017-04-04 rigid_bodies.tex
por se me assaltam

@@
 
-Molecular Simulation, being the computational realization of Statistical Mechanics\cite{Tuckerman2010}, involves several types of approximations in order to mimic ``real'' physical systems.
-Naively, one would expect those approximations to only produce deviation from experimental values.
-However, they actually may give rise to systematic errors introducing non-physical behaviour.
-For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics, disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property\cite{Eastwood_2010}.
-In a previous paper,\citep{Abreu_2017} we reported the same artificiality to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics.
+Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman2010} involves several types of approximations in order to mimic ``real'' physical systems.
+Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
+However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour.
+For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Abreu_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics.
 That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
 
-The relevance of rigid-body Molecular Dynamics stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water, but also it is a simple coarse-grain strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules\cite{Knorowski2012, Patra2013}.
-As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented in the softwares packages \textit{blue}-HOOMD and LAMMPS\cite{Plimpton1995}, has been applied to simulate small molecules\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2}, as well as membranes~\cite{Bucior_2012}, molecular motors\cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles\cite{Patra_2014}.
+The relevance of rigid-body Molecular Dynamics stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski2012, Patra2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton1995}, has been applied to simulate small molecules,\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
 
 The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in Molecular Dynamics, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
-The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with what backward error analysis is not feasible anymore.
+The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is not longer feasible.
 Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}
 
-In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body Molecular Dynamics in the canonical ensemble.
+In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body Molecular Dynamics in the canonical (NVT) ensemble.
 The starting point is our previous work on the microcanonical ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
 Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
-The analysis of the integrators will be rather heuristic, relying on comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy.
+The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy.
 We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
@@
 The paper is organized as follows.
-In Sec.~\ref{sec:theory} we briefly review the equations of motion in the microcanonical ensemble, which are the basis for developing the corresponding dynamics in the canonical ensemble.
+In Sec.~\ref{sec:theory} we briefly review the equations of motion in the microcanonical (NVT) ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble.
 In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
@@
 \label{sec:theory}
-Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
-In the aforementioned paper, we introduced an analytical solution for torque-free rotations, which would be particularly useful if any backward error analysis\cite{Hairer_2006} is to be carried out.
-However, the  Nos\'{e}-Hoover chain thermostat introduces non-Hamiltonian components, so any backward error analysis it is not feasible anymore.
-Given that, in this paper, the thermostatted version of the equations of motion will be generated considering the approximate solution for the rotational motion, which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision notably similar to that achieved using the analytical solution.
-In the next subsection, we will skip over the mathematical details and restrict ourselves to some fundamental expressions on which the dynamics in the canonical (NVT) ensemble will be developed.
-For full details in the NVE ensemble the reader is referred to our previous paper.\cite{Abreu_2017}
+
+Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we introduced symplectic numerical integrators for MD in the NVE ensemble, for which we derived an analytical, as well as an approximate solution for torque-free rotations.
+This last approach is a reformulation of the method introduced by Miller~\textit{et al.},\cite{Miller2002} and given that it demonstrated to be as accurate as the exact solution, the thermostatted version of the equations of motion will be derived considering the approximate solution.
+In the next subsection, we will skip over the mathematical derivations and restrict ourselves to some fundamental expressions.
 
@@
 When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled.
-Mathematically, the position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$.
+Mathematically, the position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$.
 In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\dot{\vt q}} = 0$.
@@
 \end{equation}
-Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} a convenient factorization of matrix $\mt A$.
-This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components, which is
+Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} a convenient factorization of matrix $\mt A$ and most important, it provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components, which is given by
 \begin{equation}
@@
 \end{equation}
-The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
  For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.
@@
 In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}.
-In this equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a digonal matrix in the principal frame.
+In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a digonal matrix in the principal frame.
 For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}
@@
 
-The equations of motion are obtained from the Hamiltonian by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}.
+The equations of motion are derived from the Hamiltonian Eq.\eqref{eq:H_NVE} by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}.
 For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
@@
 \]
-where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference and thus $\vt F_i$ is the resultant force on the body.
+where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference, and thus $\vt F_i$ is the resultant force on the body.
 As demonstrated in Abreu~\textit{et al.},\citep{Abreu_2017} $\partial{U}/\partial \vt q_i$ is given by
@@
 \]
-where $\vt \tau_i$ is the resultant torque exerted on the body and ${\vt \delta}_j$ is the position of the particle $j$, both expressed in the space-fixed frame.
-The position ${\vt \delta}_j$ is related to ${\vt d}_j$ by ${\vt \delta}_j = \tr{\mt A} {\vt d}_j$.\cite{Goldstein2002} After carrying out the required differentiations of $\mathcal{H}$  we obtain the following Hamiltonian system of equations of motion for a rigid body:
+where $\vt \tau_i$ is the resultant torque exerted on the body, and ${\vt \delta}_j$ $=$ $\tr{\mt A} {\vt d}_j$ is the position of the particle $j$, both expressed in the space-fixed frame.
+After carrying out the required differentiations of Eq.\eqref{eq:H_NVE}, we obtain the following system of equations for the rigid body motion:
 \begin{subequations}
@@
 
-The approximate solution for a complete rotation is obtained by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ employing each entry of $\vt \omega$: 
+The approximate solution for rotation is derived by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ from each entry of $\vt \omega$: 
 \begin{equation}
@@
 
-In this context, the Hamiltonian of the rigid body can be rewritten as
+In this context, the Hamiltonian of Eq.\eqref{eq:H_NVE} can be rewritten as
 \begin{equation}
@@
 In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$.
-Of coarse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$.
+Of corse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$.
 The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
@@
 \label{eq:full_operator}
-i\!L_{NVE} = i\!L_t + i\!L_b + i\!L_r,
+i\!L_\text{NVE} = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k,
 \end{equation}
@@
 &boost:\\
-&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
+&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
 &translation: \\
-& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
+& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
 &uniaxial \; rotation:\\
-& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right),
+& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
 \end{split}
 \]
-using $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete torque-free rotation. As explained in our previous paper,\cite{Abreu2017} those individual operators can be applied analytically according to
+The individual operators can be evaluated analytically according to
 \[
@@
 \]
-
-Applying the Trotter-susz The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
-\begin{equation}
-\label{eq:trotter_splitting_NVE}
-e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b},
-\end{equation}
-Update of forces and torques is required only after the free translation and free rotation have been carried out.
-Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is splitted as
+Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is factorized as
 \begin{align*}
@@
 \end{align*}
-where ${\Delta t}$ is the time step size.
-
-equipartion of energy
+where ${\Delta t}$ is the time step size. The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
+\begin{equation}
+\label{eq:trotter_splitting_NVE}
+e^{\Delta t i\!L_\text{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
+\end{equation}
 
@@
 %\end{equation}
+
+%The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
 \end{document}

### 2017-04-07 rigid_bodies.tex
NVE

@@
 However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour.
-For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Abreu_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics.
+For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Abreu_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
 That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
 
-The relevance of rigid-body Molecular Dynamics stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski2012, Patra2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton1995}, has been applied to simulate small molecules,\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
+The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski2012, Patra2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton1995} has been applied to simulate small molecules,\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
 
-The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in Molecular Dynamics, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
-The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is not longer feasible.
+The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
+The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible.
 Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}
 
-In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body Molecular Dynamics in the canonical (NVT) ensemble.
-The starting point is our previous work on the microcanonical ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
+In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble.
+The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
 Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
@@
 The paper is organized as follows.
-In Sec.~\ref{sec:theory} we briefly review the equations of motion in the microcanonical (NVT) ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble.
+In Sec.~\ref{sec:theory} we briefly review the equations of motion in the NVE ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble.
 In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
@@
 
-Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we introduced symplectic numerical integrators for MD in the NVE ensemble, for which we derived an analytical, as well as an approximate solution for torque-free rotations.
-This last approach is a reformulation of the method introduced by Miller~\textit{et al.},\cite{Miller2002} and given that it demonstrated to be as accurate as the exact solution, the thermostatted version of the equations of motion will be derived considering the approximate solution.
-In the next subsection, we will skip over the mathematical derivations and restrict ourselves to some fundamental expressions.
+Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we derived symplectic numerical integrators in the NVE ensemble: one that uses the exact solution for torque-free rotation and another approach, based on the method introduced by Miller~\textit{et al.},\cite{Miller2002}, that employs an approximate solution for rotation.
+For mathematical details the reader is referred to our paper.\cite{Abreu_2017} Next, we restrict ourselves to present the expressions required to develop the NVT case.
 
@@
 When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled.
-Mathematically, the position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$.
-In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\dot{\vt q}} = 0$.
+The position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$.
+In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$ and, as a consequence, $\tr{\vt q}{\dot{\vt q}} = 0$.
 
-The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
-\begin{equation*}
-\label{eq:A_from_q}
-\mt A(\vt q) = \begin{smallarray}{ccc}
-q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
-2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
-2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
-\end{smallarray},
-\end{equation*}
-which admits the following factorization:
+The matrix $\mt A$ which can be computed from a given unit quaternion $\vt q$ as explained in Refs.~\onlinecite{Allen1989,Miller2002} admits the following factorization:
 \begin{equation}
@@
 \end{equation}
-Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} a convenient factorization of matrix $\mt A$ and most important, it provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components, which is given by
+That factorization of matrix $\mt A$ provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components $\dot{\vt q}$, which is given by
 \begin{equation}
@@
 In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}.
-In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a digonal matrix in the principal frame.
+In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a diagonal matrix in the principal frame.
 For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}
@@
 & i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
-&uniaxial \; rotation:\\
+&rotation \; about \; a \; principal \; axis:\\
 & i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
@@
 &{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
-&uniaxial \; rotation:\\
+&rotation \; about \; a \; principal \; axis:\\
 &{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
@@
 Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
-However...the energy drift...
+However, as will be shown in the numerical results of Sec.~\ref{sec:numerical_results},
 
@@
 In this subsection we briefly review a simple method that serves for checking the consistency between the sampled and theoretical energy distributions.
-The procedures implemented by Shirts ~\cite{Shirts2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
-Fist recall that in the NVT ensemble the probability of observing an energy E is given by
+The procedures, implemented by Shirts,\cite{Shirts2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
+First, recall that in the NVT ensemble the probability of observing an energy E is given by
 \[

### 2018-02-04 rigid_bodies.tex
add figures
histograms
Change of format for the American Chemical Society

@@
-\documentclass[aip,jcp,reprint,amsmath,amssymb]{revtex4-1}
+%\documentclass[journal=jctcce,manuscript=article]{achemso}
+\documentclass[journal=jctcce,manuscript=article,layout=twocolumn]{achemso}
 
-\usepackage[pdftex]{graphicx}
-\usepackage[]{threeparttable}
+% NOTE: comment bellow before submitting
+\let\titlefont\undefined
+\makeatletter
+\let\l@addto@macro\relax
+\makeatother
+\usepackage[fontsize=11pt]{scrextend}
+% NOTE: comment above before submitting
+
+\usepackage{threeparttable}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage[T1]{fontenc}
+\usepackage[table]{xcolor}
+\definecolor{lightgray}{gray}{0.9}
 
@@
 \newcommand{\tr}[1]{#1^\text{t}}                % transposition
-\newcommand{\diff}[2]{\dfrac{\partial #1}{\partial #2}} % partial derivative
+\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
+%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
+\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
+\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
+\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
 
-\newcommand{\sn}{\text{sn}}                             % Jacobi function sn
-\newcommand{\cn}{\text{cn}}                             % Jacobi function cn
-\newcommand{\dn}{\text{dn}}                             % Jacobi function dn
+\usepackage{array}
+\newcolumntype{L}{>{$}l<{$}}
+\newcolumntype{C}{>{$}c<{$}}
+\newcolumntype{R}{>{$}r<{$}}
 
@@
 {\endarray\right]$\egroup\egroup\,\null}
-
-\begin{document}
-
-\title{Revisiting the Nos\'{e}-Hoover chain thermostat in the Rigid Bodies Molecular Dynamics}
+\listfiles
 
@@
 
-\date{\today}
+\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}
 
-%\begin{abstract}
-%In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof.
-Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model.
-In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules.
-Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics.
-We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions.
-Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
-%\end{abstract}
+%\abbreviations{AAA, BBB, CCC}
+\keywords{aaa, bbb, ccc}
 
-\maketitle
+\begin{document}
+
+\begin{tocentry}
+	Graphical Abstract
+\end{tocentry}
+
+\begin{abstract}
+	Abstract.
+\end{abstract}
 
@@
 However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour.
-For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Abreu_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
+For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
 That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
@@
 In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble.
-The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Abreu_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
+The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
 Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
@@
 
-Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we derived symplectic numerical integrators in the NVE ensemble: one that uses the exact solution for torque-free rotation and another approach, based on the method introduced by Miller~\textit{et al.},\cite{Miller2002}, that employs an approximate solution for rotation.
-For mathematical details the reader is referred to our paper.\cite{Abreu_2017} Next, we restrict ourselves to present the expressions required to develop the NVT case.
+Guided by a particular factorization of the rotation matrix expressed in terms of quaternion components, in a previous paper \cite{Silveira_2017} we derived symplectic numerical integrators for rigid bodies in the NVE ensemble.
+The first one uses an exact solution for torque-free rotations.
+The second approach is based on the method introduced by \citeauthor{Dullweber_1997} \cite{Dullweber_1997} and \citeauthor{Miller2002} \cite{Miller2002}, that employs an approximate solution for rotation.
+For mathematical details the reader is referred to our paper.\cite{Silveira_2017} Next, we restrict ourselves to present the expressions required to develop the NVT case.
 
@@
 
-The matrix $\mt A$ which can be computed from a given unit quaternion $\vt q$ as explained in Refs.~\onlinecite{Allen1989,Miller2002} admits the following factorization:
+The matrix $\mt A$ which can be computed from a given unit quaternion $\vt q$ as explained in Refs.~\citenum{Allen1989,Miller2002} admits the following factorization:
 \begin{equation}
@@
 In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a diagonal matrix in the principal frame.
-For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Abreu_2017}
+For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Silveira_2017}
 
@@
 where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference, and thus $\vt F_i$ is the resultant force on the body.
-As demonstrated in Abreu~\textit{et al.},\citep{Abreu_2017} $\partial{U}/\partial \vt q_i$ is given by
+As demonstrated in Abreu~\textit{et al.},\citep{Silveira_2017} $\partial{U}/\partial \vt q_i$ is given by
 \[
@@
 Of corse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$.
-The NVE dynamics of this system is governed by a Liouville operator given by\cite{Abreu_2017}
+The NVE dynamics of this system is governed by a Liouville operator given by\cite{Silveira_2017}
 \begin{equation}
@@
 where $\mathcal H$ is the same as in Eq.~\eqref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$.
-As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
+As recommended in Ref.~\citenum{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
 The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
@@
 For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
-In the general case, by taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
+In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
 In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna1994}
@@
 We might note that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^3 {(-1)^n x^n}/{(n+1)!}$ whenever $|x| \leq 10^{-4}$.
-The solution above is equivalent to that found in Ref.~\onlinecite{Martyna1996}.
+The solution above is equivalent to that found in Ref.~\citenum{Martyna1996}.
 
@@
 	\label{table:new_results}
-	\begin{ruledtabular}
+%	\begin{ruledtabular}
 		\begin{tabular}{cccc}
@@
 		\end{tabular}
-	\end{ruledtabular}
+%	\end{ruledtabular}
 \end{table}
@@
 	\label{table:new_results_2}
-	\begin{ruledtabular}
+%	\begin{ruledtabular}
 		\begin{tabular}{cccc}
@@
 \label{table:proposed_method}
-\begin{ruledtabular}
+%\begin{ruledtabular}
 \begin{tabular}{cccccccc}
@@
 \end{tabular}
-\end{ruledtabular}
+%\end{ruledtabular}
 \end{table*}
+
+\begin{figure}
+    \caption{SHAKE: $\Delta t$ = 1 fs ($\bar{T}$ = 299.98 K, $\bar{U}$ = -8428.86 kcal/mol) HMC: $\bar{U}$ = -8470.71 kcal/mol  }
+	\includegraphics{potenergy}
+\end{figure}
+
+\begin{figure}
+    \caption{$\Delta t$ = 1 fs ($\bar{T}$ = 300.04 K, $\bar{U}$ = -8465.63 kcal/mol)}
+	\includegraphics{potenergy2}
+\end{figure}
+
+\begin{figure}
+    \caption{$\Delta t$ = 0.5 fs ($\bar{T}$ = 299.99 K, $\bar{U}$ = -8468.19 kcal/mol)}
+	\includegraphics{potenergy_05}
+\end{figure}
+
+\begin{figure}
+    \caption{$\Delta t$ = 1 fs ($\bar{T}$ = 299.52 K, $\bar{U}$ = -8470.21 kcal/mol)}
+	\includegraphics{potenergy_kamb}
+\end{figure}
 
@@
 		\label{table:ensemblevalidation}
-		\begin{ruledtabular}
+%		\begin{ruledtabular}
 			\begin{tabular}{ccccc}
@@
 		\end{tabular}
-		\end{ruledtabular}
+%		\end{ruledtabular}
 		\begin{tablenotes}
@@
 by contrast
-\begin{acknowledgments}
+
+\begin{acknowledgement}
 The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
-\end{acknowledgments}
+\end{acknowledgement}
 
@@
 \end{document}
-\grid

### 2018-02-04 rigid_bodies.tex
Insertion of rudimentary text on shadow hamiltonian

@@
 where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference, and thus $\vt F_i$ is the resultant force on the body.
-As demonstrated in Abreu~\textit{et al.},\citep{Silveira_2017} $\partial{U}/\partial \vt q_i$ is given by
+As demonstrated in \citeauthor{Silveira_2017} \cite{Silveira_2017}, $\partial{U}/\partial \vt q_i$ is given by
 \[
@@
 \end{equation}
+
+\subsection{Shadow Hamiltonian}
+\label{sec:shadown hamiltonian}
+
+First-order derivative of a scalar function with respect to a vector variable (gradient vector):
+\begin{equation*}
+f_{\vt x} = \diff{f}{\vt x}
+\end{equation*}
+
+Second-order derivative of a scalar function with respect to two vector variables (Hessian matrix):
+\begin{equation*}
+f_{\vt x \vt y} = \diff{(\tr{f}_{\vt x})}{\vt y}.
+\end{equation*}
+
+Liouville operator:
+\begin{equation}
+\label{eq:general Lioville operator}
+\Liu{} = \tr{{\mathcal H}_{\vt p}} \diff{}{\vt r} - \tr{{\mathcal H}_{\vt r}} \diff{}{\vt p} = \{\circ,\Ham{}\}.
+\end{equation}
+
+For non-commutative operators $A$ and $B$ and a real constant $h$, the symmetric Baker-Campbell-Hausdorff (BCH) formula (Ref.~\citenum{Hairer_2006}, p.~86) can be expressed as
+\begin{equation}
+\label{eq:symmetric BCH exponential}
+e^{\frac{h}{2} B} e^{h A} e^{\frac{h}{2} B} = e^{h S},
+\end{equation}
+with
+\begin{equation}
+\label{eq:symmetric BCH operator}
+S = A + B + \frac{h^2}{24} \left[2 A + B,[A,B]\right] + \mathcal{O}(h^4).
+\end{equation}
+
+In this expression, $[A,B]$ is the commutator of operators $A$ and $B$, defined as $[A,B] = AB - BA$.
+The remaining terms of the infinite series involve nested commutators with increasing levels of recursion \cite{Hairer_2006}.
+If $A$ and $B$ are Liouville operators $\Liu A = \{\circ,\Ham A\}$ and $\Liu B = \{\circ,\Ham B\}$, respectively, then we have that
+\begin{align*}
+[\Liu A,\Liu B] &= \{\{\circ,\Ham B\},\Ham A\} - \{\{\circ,\Ham A\},\Ham B\} = \\
+&= -\{\Ham A,\{\circ,\Ham B\}\} - \{\Ham B,\{\Ham A,\circ\}\} = \\
+&= \{\circ,\{\Ham B,\Ham A\}\} = \\
+&= \{\circ,{\Liu A} {\Ham B}\}.
+\end{align*}
+
+In the expression above, the Jacobi identity \cite{Hairer_2006} has been employed after some convenient permutations.
+This result means that the commutator of two Liouville operators is, in fact, a new Liouville operator $\Liu C$ with Hamiltonian $\Ham C = {\Liu A}{\Ham B}$.
+By induction, we can conclude that the operator $S$ in Eq.~\eqref{eq:symmetric BCH operator} is also a Liouville operator $\Liu S$ with Hamiltonian
+\begin{equation}
+\label{eq:symmetric BCH hamiltonian}
+\Ham S = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
+\end{equation}
+
+Considering a separable Hamiltonian $\Ham{}(\vt r, \vt p) = K(\vt p) + U(\vt r)$, where $K$ and $U$ are the kinetic and potential energies, respectively, the Velocity Verlet integrator consists in splitting $\Ham{} = \Ham A + \Ham B$, where $\Ham A = K(\vt p)$ and $\Ham B = U(\vt r)$.
+Therefore, $\Liu A = \tr{K_{\vt p}} \diff{}{\vt r}$ and $\Liu B = -\tr{U_{\vt r}} \diff{}{\vt p}$, which makes ${\Liu A}{\Ham B} = \tr{K_{\vt p}} U_{\vt r} = \tr{U_{\vt r}} K_{\vt p}$.
+Hence, Eq.~\eqref{eq:symmetric BCH hamiltonian} becomes
+\begin{equation*}
+\Ham S = K + U + \frac{{\Delta t}^2}{24} \left(2\tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p} - \tr{U_{\vt r}} K_{\vt p \vt p} U_{\vt r} \right) + \mathcal{O}({\Delta t}^4).
+%\Ham S = K + U + \frac{h^2}{12} \tr{\left(\diff{K}{\vt p}\right)}\left(\sdiff{U}{\vt r}\right) \left(\diff{K}{\vt p}\right) + \frac{h^2}{24} \tr{\left(\diff{U}{\vt r}\right)}\left(\sdiff{K}{\vt p}\right) \left(\diff{U}{\vt r}\right) + \mathcal{O}(h^4).
+\end{equation*}
+
+The kinetic energy is given by $K = \frac{1}{2} \tr{\vt p}{\mt M}^{-1}{\vt p}$, where $\mt M$ is a diagonal matrix containing the mass associated to each degree of freedom.
+This makes $K_{\vt p} = {\mt M}^{-1}{\vt p}$ and $K_{\vt p \vt p} = {\mt M}^{-1}$.
+The vector containing the force acting on each degree of freedom is $\vt F = -U_{\vt r}$.
+Thus,
+\begin{align*}
+\Ham S &= K + U + \frac{{\Delta t}^2}{24} \left(2\tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p} - \tr{\vt F} {\mt M}^{-1} {\vt F} \right) \\ &+ \mathcal{O}({\Delta t}^4).
+\end{align*}
+
+Let us introduce $\tilde{\mt M}$ as a mass tensor\cite{Bennett_1975} whose definition is given by
+\begin{equation*}
+\tilde{\mt M}^{-1} = {\mt M}^{-1} + \frac{{\Delta t}^2}{6} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1}.
+\end{equation*}
+
+Therefore,
+\begin{equation}
+\Ham S = \frac{1}{2}\tr{\vt p} \tilde{\mt M}^{-1} {\vt p} + {\tilde U} + \mathcal{O}({\Delta t}^4),
+\end{equation}
+where
+\begin{equation*}
+{\tilde U} = U - \frac{{\Delta t}^2}{24}\tr{\vt F}{\mt M}^{-1} {\vt F}.
+\end{equation*}
+
+The important thing is that $\tilde{\mt M}$ depends exclusively on the coordinates, which makes $\diff{\tilde K}{\vt p} = \tilde{\mt M}^{-1} {\vt p}$.
+In this way, the equipartition theorem (valid at least in this direction?
+something to do with integrating through momenta at least internally to the integral in coordinates?) states that
+\begin{equation*}
+3N k_B T = \left\langle \tr{\vt p} \diff{\tilde K}{\vt p} \right\rangle = 2\langle \tilde K \rangle
+\end{equation*}
 
@@
 \hline
-\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of Kamberaj \textit{et al}.\cite{Kamberaj2005}} \\
+\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of \citeauthor{Kamberaj2005} \cite{Kamberaj2005}} \\
 1.0 & $299.52\pm0.02$ & $-8470.2\pm0.5$ & $1611.6\pm0.1$ & $805.9\pm0.1$ & $805.6\pm0.1$ & $58\pm2$ & $0.272$ \\
@@
 \end{acknowledgement}
+
+\appendix
+
+\section{Shadow Hamiltonian of Rigid Bodies}
+
+In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies. The Hamiltonian is:
+\begin{equation*}
+\Ham{}(\vt r, \vt p, \vt q, \vt \pi) = K(\vt p, \vt q, \vt \pi) + U(\vt r, \vt q),
+\end{equation*}
+where
+\begin{equation*}
+K = \frac{1}{2} \tr{\vt p}{\mt M}^{-1}{\vt p} + \frac{1}{8}\tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi,
+\end{equation*}
+with
+\begin{equation*}
+\mt B(\vt q) = \left[\begin{array}{rrrr}
+-q_2 & -q_3 & -q_4 \\
+q_1 & -q_4 &  q_3 \\
+q_4 &  q_1 & -q_2 \\
+-q_3 &  q_2 &  q_1
+\end{array}\right].
+\end{equation*}
+
+If a splitting is done so that $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$, then from Eq.~\eqref{eq:general Lioville operator} we have that $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$.
+Therefore, $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$.
+This makes
+\begin{align*}
+\Liu A \Liu A \Ham B &= \tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p}
++ 2 \tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p}
++ \tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} \\
+&+ \tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q}
+- \tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q},
+\end{align*}
+where the fact that $\tr{U_{\vt q \vt r}} = U_{\vt r \vt q}$ has been exploited.
+As reported in Ref.~\citenum{Silveira_2017}, the identity $\tr{\mt B}(\vt q) {\vt \pi} = -\tr{\mt B}(\vt \pi) {\vt q}$ is useful for performing differentiations.
+The first-order derivatives of $K$ with respect to $\vt p$, $\vt q$, and $\vt \pi$ are, respectively,
+\begin{align*}
+&K_{\vt p} = {\mt M}^{-1} {\vt p}, \\
+&K_{\vt q} = -\frac{1}{2} {\mt B}(\vt \pi) {\vt \omega}, \; \text{and} \\
+&K_{\vt \pi} = \frac{1}{2} {\mt B}(\vt q) {\vt \omega},
+\end{align*}
+where $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi = -\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt \pi) \vt q$ is the vector of angular velocities.\cite{Silveira_2017} Hence,
+\begin{gather*}
+\tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p} = \tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p}, \\
+\tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p} = \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}, \\
+\tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} = \frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
+\end{gather*}
+
+As $\diff{\tr{\vt \omega}}{\vt \pi} = \frac{1}{2} {\mt B}(\vt q) {\mt I}^{-1}$, the second-order derivative of $K$ with respect to $\vt \pi$ turns out to be
+\begin{equation*}
+K_{\vt \pi \vt \pi} = \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q).
+\end{equation*}
+
+Pre-multiplication by $\tr{K_{\vt q}}$ will introduce a product $\tr{\mt B}(\vt \pi) {\mt B}(\vt q)$, which has been shown in Ref.~\citenum{Silveira_2017} to be equal to ${\mt S}\left( \tr{\mt B}(\vt \pi) {\vt q} \right) = -2 {\mt S}({\mt I} {\vt \omega})$, where the operator ${\mt S}(\cdot)$ builds a skew-symmetric matrix from the entries of a vector.
+In addition, post-multiplication by $U_{\vt q}$ will introduce a product $\tr{\mt B}(\vt q) U_{\vt q}$, which is equal to $-2 {\vt \tau}^b$, where ${\vt \tau}^b$ is the body-fixed frame representation of the torque.
+Then, the fact that ${\mt S}(\vt x) {\vt y} = -{\mt S}(\vt y) {\vt x}$ leads to
+\begin{equation*}
+\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega}.
+\end{equation*}
+
+Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{k=1}^3 \omega_k \hat{\mt B}_k ) \vt q$, where each $\hat{\mt B}_k$ is a skew-symmetric permutation matrix (i.e.
+$\tr{\hat{\mt B}}_k = -\hat{\mt B}_k$).
+This allows us to obtain
+\begin{align*}
+K_{\vt \pi \vt q} &= \frac{1}{2} (\diff{\tr{\vt \omega}}{\vt q}) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{k=1}^3 \omega_k \tr{\hat{\mt B}}_k = \\
+&= -\frac{1}{4} {\mt B}(\vt \pi) {\mt I}^{-1} \tr{\mt B}(\vt q) - \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k.
+\end{align*}
+
+Pre-multiplication by $\tr{K_{\vt \pi}}$ and post-multiplying $U_{\vt q}$, followed by some algebraic transformations, ultimately lead to
+\begin{equation*}
+\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q} = -\frac{1}{2} \tr{\vt \omega} \tr{\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} + \frac{1}{2} \tr{\vt \omega} {\mt S}({\vt \tau}^b){\vt \omega}.
+\end{equation*}
+
+The second term in the right-hand side of the equation above is identically null whenever (a quadratic form whose matrix is skew-symmetric), which makes $\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = -\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q}$.
+Finally,
+\begin{align*}
+\Liu A \Liu A \Ham B = &\tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p} + \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}
++ \\
+&\frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega - \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}_b) {\mt I} {\vt \omega}.
+\end{align*}
+
+The last term above need to be double checked.
+
+For the second part:
+\begin{align*}
+\Liu B \Liu A \Ham B &= -\tr{U_{\vt r}} \tr{K_{\vt p \vt p}} U_{\vt r} - \tr{U_{\vt q}} \tr{K_{\vt \pi \vt \pi}} U_{\vt q} = \\
+&= -\tr{\vt F} {\mt M}^{-1} {\vt F} - \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b.
+\end{align*}
+
+Overall:
+\begin{equation*}
+\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\Delta t}^4),
+\end{equation*}
+where
+\begin{equation*}
+\tilde{U} = U - \frac{\Delta t^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right)
+\end{equation*}
+and
+\begin{equation*}
+\tilde{\mt \Omega} = {\mt \Omega} + \frac{\Delta t^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} + {\mt \Gamma} \right)
+\end{equation*}
+with
+\begin{equation*}
+\mt \Omega = \left[\begin{array}{cc}
+{\mt M}^{-1} & \mt 0 \\
+\mt 0 & \frac{1}{2} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
+\end{array}\right],
+\end{equation*}
+\begin{equation*}
+\mt \Gamma = \left[\begin{array}{cc}
+\mt 0 & \mt 0 \\
+\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} {\mt S}({\mt I}^{-1} {\vt \tau}_b) \tr{\mt B}(\vt q)
+\end{array}\right],
+\end{equation*}
+and
+\begin{equation*}
+{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
+U_{\vt r \vt r} & U_{\vt r \vt q} \\
+U_{\vt q \vt r} & U_{\vt q \vt q}
+\end{array}\right].
+\end{equation*}
 

### 2018-02-11 rigid_bodies.tex
New section with proposed method at the beginning

@@
 \newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
+\newcommand{\timestep}{h}
+\newcommand{\shadow}[1]{\widetilde{#1}}
 
@@
 
-\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}
+%\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}
+\title{Nos\'e-Hoover Chains Improved by Backward Error Analysis: Application to Rigid-Body Molecular Dynamics}
 
@@
 
-Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman2010} involves several types of approximations in order to mimic ``real'' physical systems.
+Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman_2010} involves several types of approximations in order to mimic ``real'' physical systems.
 Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
@@
 For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
-That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
-
-The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski2012, Patra2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton1995} has been applied to simulate small molecules,\cite{Geiger_2013,Aimoli_2014,Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
-
-The Nos\'{e}-Hoover chain thermostat,\cite{Martyna1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
+That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
+
+The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski_2012, Patra_2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj_2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton_1995} has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
+
+The Nos\'{e}-Hoover chain thermostat,\cite{Martyna_1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
 The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible.
-Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi2001,Ezra2004,Ezra2006}
+Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}
 
@@
 Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
-The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane1987} which is exact in the sense that truncation errors do not affect its accuracy.
-We also include results from simulations using the conventional SHAKE\cite{Ryckaert1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
+The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj_2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy.
+We also include results from simulations using the conventional SHAKE\cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
 
@@
 
-\section{Theory}
-\label{sec:theory}
-
-Guided by a particular factorization of the rotation matrix expressed in terms of quaternion components, in a previous paper \cite{Silveira_2017} we derived symplectic numerical integrators for rigid bodies in the NVE ensemble.
-The first one uses an exact solution for torque-free rotations.
-The second approach is based on the method introduced by \citeauthor{Dullweber_1997} \cite{Dullweber_1997} and \citeauthor{Miller2002} \cite{Miller2002}, that employs an approximate solution for rotation.
-For mathematical details the reader is referred to our paper.\cite{Silveira_2017} Next, we restrict ourselves to present the expressions required to develop the NVT case.
-
-\subsection{Microcanonical Ensemble}
-
-When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled.
-The position of a particle $j$ constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the center of mass position of the body, $\vt d_j$ is the particle position in the body-fixed frame of reference, and $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix $\mt A$, which in turn is a function of the unit quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$.
-In terms of $\vt q$, the dynamics must satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$ and, as a consequence, $\tr{\vt q}{\dot{\vt q}} = 0$.
-
-The matrix $\mt A$ which can be computed from a given unit quaternion $\vt q$ as explained in Refs.~\citenum{Allen1989,Miller2002} admits the following factorization:
-\begin{equation}
-\label{eq:factorization_of_A}
-{\mt A} = \tr{\mt B}{\mt C},
-\end{equation} 
-where
-\begin{equation}
+\section{Proposed Method}
+\label{sec:methodology}
+
+In a previous paper \cite{Silveira_2017}, we employed a particular factorization of the orientation matrix expressed in terms of quaternion components to 1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for rigid bodies in the NVE ensemble and 2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} as a sequence of uniform rotations around the principal axes of a body, thus unveiling the equivalence between this method and an older one by \citeauthor{Dullweber_1997} \cite{Dullweber_1997}
+
+The ordinary differential equation (ODE) system that describes a Hamiltonian rigid-body motion are \cite{Silveira_2017}
+\begin{subequations}
+	\label{eq:ODE system for NVE}
+	\begin{align}
+	&\dot{\vt r} =
+	%+\diff{\Ham{}}{\vt p} =
+	{\mt M}^{-1} {\vt p}, \\
+	&\dot{\vt p} =
+	%-\diff{\Ham{}}{\vt r} = 
+	{\vt F}, \\
+	&\dot{\vt q} =
+	%+\diff{\Ham{}}{\vt \pi} =
+	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
+	&\dot{\vt \pi} =
+	%-\diff{\Ham{}}{\vt q} =
+	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
+	\end{align}
+\end{subequations}
+with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$.
+In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines the body's orientation, and $\vt \pi$ is the momentum conjugated to such quaternion.
+Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity.
+Matrices $\mt M$ and $\mt I$ are diagonal ones, with the former containing the body's mass in all its diagonal entries and the latter containing the principal moments of inertia.
+Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
+\begin{equation*}
 \label{eq:def_B_and_C}
@@
 \end{smallarray}.
-\end{equation}
-That factorization of matrix $\mt A$ provides the mathematical relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rate of change of the quaternion components $\dot{\vt q}$, which is given by
-\begin{equation}
-\label{eq:relation_qdot_omega}
-\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
-\end{equation}
-For the sake of brevity, the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) might be omitted, when such argument is $\vt q$.
-
-In the Hamiltonian approach, we employ the linear momentum $\vt p = m \dot{\vt r}$, conjugated to $\vt r$, as well as the momentum $\vt \pi= 4 \mt B \mt I \tr{\mt B} \dot{\vt q} = 2 \mt B \mt I \vt \omega$, conjugated to $\vt q$\cite{Goldstein2002}.
-In these equations, $m$ is the mass of the body and $\mt I$ is the inertia tensor, which is a diagonal matrix in the principal frame.
-For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows:~\cite{Silveira_2017}
-
-\begin{equation}
-\label{eq:H_NVE}
-\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right),
-\end{equation}
-where the index $i$ ranging from $1$ to $N$ indicates the quantities related to each body.
-
-The equations of motion are derived from the Hamiltonian Eq.\eqref{eq:H_NVE} by $\dot{\vt r}_i = \partial \mathcal{H} / \partial \vt p_i$, $\dot{\vt p}_i = -\partial \mathcal{H} / \partial \vt r_i$, $\dot{\vt q}_i = \partial \mathcal{H} / \partial \vt \pi_i$, and $\dot{\vt \pi}_i = -\partial \mathcal{H} / \partial \vt q_i$.\cite{Goldstein2002}.
-For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
+\end{equation*}
+
+The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the principal-axis frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the principal-axis and space-fixed frames) expressed as a function of the quaternion components.
+The fact that such matrix admits the simple factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ played a central role in Ref.~\citenum{Silveira_2017}.
+
+The Hamiltonian conserved by Eq.~\eqref{eq:ODE system for NVE} is
+\begin{equation}
+\label{eq:Hamiltonian}
+\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
+\end{equation}
+where $K$ and $U$ are the body's kinetic and potential energies, respectively. While $U(\vt r, \vt q)$ can have any functional form, the kinetic energy is given by
+\begin{equation}
+\label{eq:kinetic energy}
+K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
+\end{equation}
+
+An advantage of this notation is that Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the respective properties of all bodies piled together.
+The same occurs for diagonal matrices $\mt M$ and $\mt I$, whose sizes increase to $3N \times 3N$ in this case.
+In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
+Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged to accommodate a number of free atoms.
+
+A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system's configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$.
+A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
+\begin{equation}
+\label{eq:splitting NVE}
+e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
+\end{equation}
+where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
+In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, where the asterisk denotes the state of the system immediately before a propagation takes place.
+The action of $e^{\timestep \Liu{A}}$ is a uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) with simultaneous torque-free rotation during a full time step.
+
+Description of analytical and NO\_SQUISH rotations...
+
+Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble.
+For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system.
+This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement.
+To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$.
+The equations of motion, in this case, are
+\begin{subequations}
+	\label{eq:ODE system for NVT}
+	\begin{align}
+	&\dot{\vt r} =
+	{\mt M}^{-1} {\vt p}, \\
+	&\dot{\vt p} =
+	{\vt F} - \alpha_1 {\vt p}, \\
+	&\dot{\vt q} =
+	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \\
+	&\dot{\vt \pi} =
+	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
+	&\dot{\eta}_j = \alpha_j, \text{ and} \\
+	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M,
+	\end{align}
+\end{subequations}
+
+In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where each $Q_j$ is an inertial parameter of the corresponding thermostat.
+Besides, each $G_j$ is a generalized force acting on thermostat $j$, defined as
+\begin{subequations}
+\begin{align}
+\label{eq:generalized force NVT}
+&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
+&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1,
+\end{align}
+\end{subequations}
+where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set as null, as usual (otherwise, $L = 6N$).\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
+%Finally, $\shadow K$ is the total kinetic energy of the system, whose computation (not using Eq.~\ref{eq:kinetic energy}, hence the distinct symbol) is the central feature of the method we propose.
+
+The ``extended energy'' which is conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
+\begin{equation}
+\label{eq:nvt extended energy}
+H_\text{\tiny NVT} = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
+\end{equation}
+
+As in the previous case, it is possible to devise a splitting method to solve Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
+By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
+non-Hamiltonian) Liouville operator, the splitting goes as
+\begin{equation*}
+e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
+\end{equation*}
+where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
+Of course, the NVE part can be integrated exactly as described before.
+The NHC part, in turn, can be split even further as
+\begin{equation*}
+e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
+\end{equation*}
+
+In our scheme, the first propagator to act, $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
+Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied sequentially with $j$ ranging from $M-1$ down to $1$.
+With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as
+\begin{align*}
+&p_{\eta_j} = p_{\eta_j}^\ast + \Big( G_j^\ast - \alpha_{j+1}^\ast p_{\eta_j}^\ast \Big) \phi\left(\frac{\alpha_{j+1}^\ast \timestep}{4n}\right) \frac{\timestep}{4n} \; \text{and} \\
+&\eta_{j+1} = \eta_{j+1}^\ast + \alpha_{j+1}^\ast \frac{\timestep}{4n}.
+\end{align*}
+
+After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ enacts the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of the thermostat coordinate $\eta_1$.
+Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
+%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\shadow K$ to be scaled by the square of it, hence ${\shadow K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\shadow K}^\ast$.
+In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\shadow K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
+Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_1}$ to $e^{\frac{\timestep}{4n} \mathcal{L}_M}$ are applied again, but now in the increasing order of their indices.
+It is worth noting that the $\eta$'s are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H_\text{\tiny NVT}$.
+
+The whole procedure described above is repeated $n$ times.
+Because of its low computational cost, the compensation required for increasing the time step $\timestep$ without degrading the accuracy of the NHC part (that is, employing a larger $n$) causes a small impact in the overall integration effort.
+In contrast, evaluating the NVE part is expensive for involving force/torque computations and its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
+A high-order integration scheme\cite{Omelyan_2007, vanZon_2008} could possibly admit larger time steps, but it would raise both cost and complexity because for entailing the computation (or numerical estimation) of force/torque gradients.
+The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
+
+Here we take a simpler path.
+Instead of attempting to improve the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we try to quantify the obtained deviations and then use them to tune the thermostatting procedure and properly weight the sampled configurations when canonical averages are computed.
+
+A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are also Hamiltonian.
+Thus, a trajectory generated to be roughly consistent with the specified Hamiltonian $\Ham{}$ will be exactly consistent with a nearby, albeit unknown, function $\shadow{\mathcal H}$.
+It is also known that $\shadow{\mathcal H}$, referred to as a shadow Hamiltonian, depends on the time step size $\timestep$ and that, for a time-reversible splitting like Eq.~\eqref{eq:splitting NVE}, $\shadow{\mathcal H} = \Ham{} + \mathcal{O}(\timestep^2)$.
+Estimating the trailing term of this equality, either analytically or numerically, is a task of backward error analysis.
+As demonstrated in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be better approximated as $\shadow{\Ham{}} = \shadow K + \shadow U + \mathcal{O}(h^4)$, where $\shadow K$ and $\shadow U$ will be called shadow kinetic and shadow potential energies, respectively.
+The former is given by
+\begin{equation}
+\label{eq:shadow kinetic energy}
+\shadow K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\end{equation}
+where $\shadow{\mathbf \Omega}$ is a complicated matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, but independent of both $\vt p$ and $\vt \pi$.
+The shadow potential is also independent of these momenta and is given by
+\begin{equation}
+\label{eq:shadow potential energy}
+\shadow U = U - \frac{\timestep^2}{24} \left[ \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A}(\vt q) {\mt I}^{-1} {\mt A}(\vt q) {\vt \tau} \right].
+\end{equation}
+
+Therefore, the outcome of solving Eq.~\eqref{eq:ODE system for NVE} using a Trotter-Suzuki splitting will be close to the exact solution of Eq.~\eqref{eq:ODE system for NVE}, but it will be even closer to the solution of a system composed of equations
+\begin{equation}
+\label{eq:shadow ODE system}
+\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] =
+\left[\begin{array}{c} \diff{\shadow K}{\vt p} \\ \diff{\shadow K}{\vt \pi} \end{array}\right] =
+\shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\end{equation}
+as well as $\dot{\vt p} = -\diff{\shadow{\mathcal H}}{\vt r}$ and $\dot{\vt \pi} = -\diff{\shadow{\mathcal H}}{\vt q}$.
+In our proposed method, we acknowledge this fact and thus use the Nos\'e-Hoover chain to provide a canonical distribution consistent with $\shadow{\mathcal H}$ (rather than $\Ham{}$) at temperature $T$.
+Note that this entails solving Eq.~\eqref{eq:ODE system for NVT} exactly as described before, with the only requirement of replacing $K$ by $\shadow K$ in the definition of the generalized force $G_1$.
+By contrasting Eqs.~\eqref{eq:generalized force NVT}, \eqref{eq:shadow kinetic energy}, and \eqref{eq:shadow ODE system}, we can see that
+\begin{equation}
+G_1 = 2 \shadow K - L k_\text{\tiny B} T.
+\end{equation}
+
+Direct evaluation of Eq.~\eqref{eq:shadow kinetic energy} for computing $\shadow K$ is quite complex and expensive.
+However, one can do it very easily using the approach devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the derivatives in the left-hand side of Eq.~\eqref{eq:shadow ODE system}.
+Denoting the obtained estimators as $\dot{\vt r}_\text{\tiny NVE}$ and $\dot{\vt q}_\text{\tiny NVE}$ and substituting Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:shadow kinetic energy}, we find out that
+\begin{equation}
+\shadow K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}_\text{\tiny NVE} + \tr{\vt \pi} \dot{\vt q}_\text{\tiny NVE} \right).
+\end{equation}
+
+- Pratical estimation of numerical derivatives: virtual steps and interpolation
+
+- Scaling of $\shadow K$ during the NHC integration.
+
+Reweighting:
+\begin{equation}
+\label{eq:reweighting}
+\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
+\end{equation}
+
+\section{Theoretical Background}
+\label{sec:theory}
+
+\subsection{Microcanonical Ensemble}
+
+When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of a center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t) = \tr{[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}.
+In this way, the position of any constituent particle $j$ will vary with time as $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in a body-fixed frame of reference centered in $\vt r$ and ${\mt A}(\vt q)$ is the body's orientation matrix (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components.
+The superscript $t$ stands for transposition, which is equivalent to inversion in the case of a canonical transformation like $\mt A$ \cite{Goldstein_2002}.
+The central feature explored in Ref.~\citenum{Silveira_2017} is the fact that $\mt A$ admits the simple factorization
+\begin{equation}
+\label{eq:factorization_of_A}
+{\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q),
+\end{equation}
+where
+\begin{equation*}
+\label{eq:def_B_and_C}
+\mt B(\vt q) = \begin{smallarray}{rrrr}
+-q_2 & -q_3 & -q_4 \\
+ q_1 & -q_4 &  q_3 \\
+ q_4 &  q_1 & -q_2 \\
+-q_3 &  q_2 &  q_1
+\end{smallarray}
+\; \text{and} \;
+\mt C(\vt q) = \begin{smallarray}{rrrr}
+-q_2 & -q_3 & -q_4 \\
+ q_1 &  q_4 & -q_3 \\
+-q_4 &  q_1 &  q_2 \\
+ q_3 & -q_2 &  q_1
+\end{smallarray}.
+\end{equation*}
+
+This factorization creates an intermediate frame of reference for representing three-dimensional vectors in $\mathbb{R}^3$ as quaternions that lie in the hyperplane orthogonal to $\vt q$, since the columns of matrix $\mt B$ and those of matrix $\mt C$ form two different bases for such a subspace of $\mathbb{R}^4$.
+Transformations amongst three distinct reference frames are enacted by multiplications with these matrices or their transposes, such as represented in the following scheme:
 \[
-\diff{U}{\vt r_i} = -\vt F_i = -\sum_{j=1}^{n_p} {\vt F_j},
+\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
+\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
+\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
+\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
+\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
 \]
-where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference, and thus $\vt F_i$ is the resultant force on the body.
-As demonstrated in \citeauthor{Silveira_2017} \cite{Silveira_2017}, $\partial{U}/\partial \vt q_i$ is given by
-\[
-\diff{U}{\vt q_i} = -2 \mt C_i \vt \tau_i = -2 \mt C_i \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
-\]
-where $\vt \tau_i$ is the resultant torque exerted on the body, and ${\vt \delta}_j$ $=$ $\tr{\mt A} {\vt d}_j$ is the position of the particle $j$, both expressed in the space-fixed frame.
-After carrying out the required differentiations of Eq.\eqref{eq:H_NVE}, we obtain the following system of equations for the rigid body motion:
+
+The intermediate frame is exactly the subspace of $\mathbb{R}^4$ that must contain the rate-of-change vector $\dot{\vt q}$ if the norm of $\vt q$ is to remain constant, since $\tr{\vt q} {\vt q} = 1$ implies that $\tr{\vt q} \dot{\vt q} = 0$.
+In the Hamiltonian approach for the motion of a rigid body, the momentum $\vt \pi$ conjugated to $\vt q$ is given by $\vt \pi = 2 \mt B(\vt q) \mt I \vt \omega$, where $\vt \omega$ is the three-dimensional angular velocity and $\mt I$ is a diagonal matrix containing the body's principal moments of inertia.\cite{Silveira_2017} This means that $\vt \pi$ is the quaternion-fixed representation of the angular momentum $\vt L = \mt I \vt \omega$, thus being orthogonal to $\vt q$ as well.
+After this notation has been introduced, the Hamiltonian of a rigid body can be written as
+\begin{equation}
+\label{eq:Hamiltonian}
+\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
+\end{equation}
+where $K$ and $U$ are the kinetic and potential energies of the body, respectively. While the potential energy can have any functional form, the kinetic energy is given by
+\begin{equation}
+\label{eq:kinetic energy}
+K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi},
+\end{equation}
+where $\vt p$ is the linear momentum of the body and $\mt M$ is a diagonal matrix containing the body's mass in all three diagonal entries.
+Note that the kinetic energy is composed of two independent contributions, a translational and a rotational one.
+An advantage of this notation is that Eqs.~\eqref{eq:Hamiltonian} and \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies once we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the respective properties of all bodies piled together.
+The same occurs for matrices $\mt M$ and $\mt I$, which turn into diagonal matrices of size $3N \times 3N$.
+In addition, $\mt B(\vt q)$ becomes a block-diagonal matrix with $4N$ rows and $3N$ columns.
+
+The Liouville operator \cite{Tuckerman_2010} that stems from the Hamiltonian in Eq.~\eqref{eq:Hamiltonian} is\cite{Silveira_2017}
+\begin{align*}
+\label{eq:Lioville operator}
+\Liu{\tiny NVE} &= \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} \\
+&+ \tr{\vt F} \diff{}{\vt p} + \left[ \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \right] \diff{}{\vt \pi},
+\end{align*}
+where $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body (or on $N$ bodies), both represented in the space-fixed frame of reference, whereas $\vt \omega$ can be obtained by
+\begin{equation*}
+\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
+\end{equation*}
+
+The time evolution of any phase-space vector $\vt x$ is governed by $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ \cite{Tuckerman_2010}.
+Therefore, the equations of motion that describe the rigid-body motion are \cite{Silveira_2017}
 \begin{subequations}
@@
 	\begin{align}
-&\dot{\vt r}_i = \frac{1}{m_i} \vt p_i \\
-&\dot{\vt p}_i = \mt F_i \\
-&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:EDO_q} \\
-&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i \label{eq:EDO_pi}
+	&\dot{\vt r} =
+	%+\diff{\Ham{}}{\vt p} =
+	{\mt M}^{-1} {\vt p}, \\
+	&\dot{\vt p} =
+	%-\diff{\Ham{}}{\vt r} = 
+	{\vt F}, \\
+	&\dot{\vt q} =
+	%+\diff{\Ham{}}{\vt \pi} =
+	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
+	&\dot{\vt \pi} =
+	%-\diff{\Ham{}}{\vt q} =
+	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
 	\end{align}
 \end{subequations}
-with $\mt \Omega_i = \mt B_i(\vt \pi_i)$.
-
-The approximate solution for rotation is derived by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ from each entry of $\vt \omega$: 
-\begin{equation}
-\label{eq:omega_entry}
-\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
-\end{equation}
-where $\hat{\mt B}_k$ is one of the following permutation matrices:
-\[
-\hat{\mt B}_1 = \begin{smallarray}{cccc}
- 0 & \text{-}1 &  0 &  0 \\
- 1 &  0 &  0 &  0 \\
- 0 &  0 &  0 &  1 \\
- 0 &  0 & \text{-}1 &  0 \\
-\end{smallarray} \quad
-\hat{\mt B}_2 = \begin{smallarray}{cccc}
- 0 &  0 & \text{-}1 &  0 \\
- 0 &  0 &  0 & \text{-}1 \\
- 1 &  0 &  0 &  0 \\
- 0 &  1 &  0 &  0 \\
-\end{smallarray} \quad
-\hat{\mt B}_3 = \begin{smallarray}{cccc}
- 0 &  0 &  0 & \text{-}1 \\
- 0 &  0 &  1 &  0 \\
- 0 & \text{-}1 &  0 &  0 \\
- 1 &  0 &  0 &  0 \\
-\end{smallarray}
-\]
-The new form of the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi} are then:
+
+The solution for a system like $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ with initial condition $\vt x(0) = \vt x_0$ can be formally expressed as $\tr{\vt x}(t) = e^{t \Liu{\tiny NVE}} \tr{\vt x}_0$.
+Although the exact effect of the exponential operator %$e^{t \Liu{\tiny NVE}}$
+cannot be determined for an arbitrary potential $U(\vt r, \vt q)$, a symplectic numerical solution can be achieved by a Verlet-type integration.
+For two non-commutative Liouville operators $\Liu A$ and $\Liu B$ and a real constant $\timestep$, the symmetric Baker-Campbell-Hausdorff (BCH) formula can be expressed as \cite{Hairer_2006}
+\begin{align}
+\label{eq:symmetric BCH}
+&e^{\frac{\timestep}{2} \Liu B} e^{\timestep \Liu A} e^{\frac{\timestep}{2} \Liu B} = \nonumber \\
+&= e^{\timestep (\Liu A + \Liu B) + \frac{\timestep^3}{24} \left[2 \Liu A + \Liu B,[\Liu A,\Liu B]\right] + \mathcal{O}(\timestep^5)},
+\end{align}
+where $[X,Y]$ is the commutator of operators $X$ and $Y$, defined as $[X,Y] = XY - YX$.
+The remaining terms of the infinite series involve ever-deepening, nested commutators \cite{Hairer_2006}.
+With $\Liu{\tiny NVE} = \Liu{A} + \Liu{B}$ and a small $\timestep$ value, we can approximate the BCH equation as
+\begin{equation*}
+e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
+\end{equation*}
+which is known as the Trotter-Suzuki splitting formula \cite{Trotter_1959, Suzuki_1976}. In the rigid-body case, by making
 \begin{subequations}
-\label{eq:EDO_system_alternative}
-\begin{align}
-&\dot{\vt q}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt q_i \label{eq:edo_q} \\
-&\dot{\vt \pi}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt \pi_i + 2 \mt C_i \vt \tau_i \label{eq:edo_pi}
-\end{align}
+\label{eq:Liouvile operators K and U}
+\begin{equation}
+\label{eq:Lioville operator K}
+\Liu{A} = \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) \diff{}{\vt \pi}
+\end{equation}
+and
+\begin{equation}
+\label{eq:Lioville operator U}
+\Liu{B} = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \diff{}{\vt \pi},
+\end{equation}
 \end{subequations}
-
-In this context, the Hamiltonian of Eq.\eqref{eq:H_NVE} can be rewritten as
-\begin{equation}
-\label{eq:H_split_omega}
-\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
-\end{equation}
-from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions.
+it becomes possible to determine exactly the individual actions of operators $e^{\timestep \Liu{A}}$ and $e^{\frac{\timestep}{2} \Liu{B}}$, with $\timestep$ being a small time step $\Delta t$.
+The latter operator promotes a kick in the linear and quaternion momenta given by ${\vt p} \leftarrow {\vt p} + \frac{\timestep}{2} {\vt F}$ and ${\vt \pi} \leftarrow {\vt \pi} + \timestep {\mt C}({\vt q}) {\vt \tau}$.
+The former, in turn, performs a simultaneous force-free translation and torque-free rotation during the length of a time step.
+%
+%
+%first one is the solution of
+%\begin{subequations}
+%	\label{eq:EDO translation and rotation}
+%	\begin{align}
+%	&\dot{\vt r} =
+%	%+\diff{\Ham{}}{\vt p} =
+%	{\mt M}^{-1} {\vt p}, \\
+%	&\dot{\vt q} =
+%	%+\diff{\Ham{}}{\vt \pi} =
+%	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
+%	&\dot{\vt \pi} =
+%	%-\diff{\Ham{}}{\vt q} =
+%	\frac{1}{2} \mt B(\vt \pi) \vt \omega, \label{eq:EDO_pi}
+%	\end{align}
+%\end{subequations}
+%
+% and $N_s$ is the number of steps. We emphasize that each exponential operator above will act on (that is, have as initial condition) the result of the operator at its right-hand side.
+%
+%
+%
+%for this system of equations e rigid body motion which preserves the symplectic structure of Eq.~\eqref{eq:EDO_system}.
+For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula isThe can be used to devise
+%The full integrator is second-order in $\timestep$ because the leading-order term of the error at every step is proportional to $\timestep^3$ and, in addition, $N_s = t/\timestep$.\cite{Tuckerman_2010} The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
+%\begin{equation}
+%\label{eq:trotter_exact_rotation}
+%e^{\timestep i\!L} = e^{\frac{\timestep}{2} i\!L_b} e^{\timestep i\!L_{t+r}} e^{\frac{\timestep}{2} i\!L_b} + \mathcal{O}(\timestep^3).
+%\end{equation}
+%
+%This is the Velocity Verlet algorithm\cite{Tuckerman_2010} with center-of-mass translations and rigid-body rotations taking place simultaneously.
+Update of forces and torques is required only after the step of translation \& rotation has been completed.
+Because the solution of torque-free rotation involves a single step in the numerical solver, the scheme represented by Eq.~\ref{eq:trotter_exact_rotation} will be referred to as the unsplit rotation integrator.
+We remark that van Zon and co-workers\cite{vanZon2007PRE, vanZon2008} have already employed this algorithm using the formulation proposed in Ref.~\citenum{vanZon2007}.
+Other splitting schemes with higher-order approximations have also been used for rigid bodies.\cite{Omelyan2007, Omelyan2008, vanZon2008} However, they result in methods that are more complex and often require expensive force-gradient calculations.
+Moreover, it is unclear how their benefits would extend to non-Hamiltonian numerical integrators, such as those often employed for molecular dynamics in ensembles other than the microcanonical one.
+%
+%
+%
+%
+%
+%
+%
+%The approximate solution for rotation is derived by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ from each entry of $\vt \omega$: 
+%\begin{equation}
+%\label{eq:omega_entry}
+%\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
+%\end{equation}
+%where $\hat{\mt B}_k$ is one of the following permutation matrices:
+%\[
+%\hat{\mt B}_1 = \begin{smallarray}{cccc}
+% 0 & \text{-}1 &  0 &  0 \\
+% 1 &  0 &  0 &  0 \\
+% 0 &  0 &  0 &  1 \\
+% 0 &  0 & \text{-}1 &  0 \\
+%\end{smallarray} \quad
+%\hat{\mt B}_2 = \begin{smallarray}{cccc}
+% 0 &  0 & \text{-}1 &  0 \\
+% 0 &  0 &  0 & \text{-}1 \\
+% 1 &  0 &  0 &  0 \\
+% 0 &  1 &  0 &  0 \\
+%\end{smallarray} \quad
+%\hat{\mt B}_3 = \begin{smallarray}{cccc}
+% 0 &  0 &  0 & \text{-}1 \\
+% 0 &  0 &  1 &  0 \\
+% 0 & \text{-}1 &  0 &  0 \\
+% 1 &  0 &  0 &  0 \\
+%\end{smallarray}
+%\]
+%The new form of the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi} are then:
+%\begin{subequations}
+%\label{eq:EDO_system_alternative}
+%\begin{align}
+%&\dot{\vt q}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt q_i \label{eq:edo_q} \\
+%&\dot{\vt \pi}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt \pi_i + 2 \mt C_i \vt \tau_i \label{eq:edo_pi}
+%\end{align}
+%\end{subequations}
+%
+%In this context, the Hamiltonian of Eq.\eqref{eq:H_NVE} can be rewritten as
+%\begin{equation}
+%\label{eq:H_split_omega}
+%\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
+%\end{equation}
+%from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions.
 In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$.
@@
 The NVE dynamics of this system is governed by a Liouville operator given by\cite{Silveira_2017}
-\begin{equation}
-\label{eq:full_operator}
-i\!L_\text{NVE} = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k,
-\end{equation}
-where the individual operators are defined as follows:
-\[
-\begin{split}
-&boost:\\
-&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
-&translation: \\
-& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
-&rotation \; about \; a \; principal \; axis:\\
-& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
-\end{split}
-\]
-The individual operators can be evaluated analytically according to
-\[
-\begin{split}
-&boost:\\
-&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
-&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
-&translation:\\
-&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
-&rotation \; about \; a \; principal \; axis:\\
-&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
-&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
-\end{split}
-\]
-Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is factorized as
-\begin{align*}
-\label{eq:splitting_rot}
-e^{\Delta t i\!L_r} =  e^{\frac{\Delta t}{2} i\!L_3} e^{\frac{\Delta t}{2} i\!L_2} e^{\Delta t i\!L_1} e^{\frac{\Delta t}{2} i\!L_2} e^{\frac{\Delta t}{2} i\!L_3},
-\end{align*}
-where ${\Delta t}$ is the time step size. The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
-\begin{equation}
-\label{eq:trotter_splitting_NVE}
-e^{\Delta t i\!L_\text{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
-\end{equation}
-
-\subsection{Shadow Hamiltonian}
-\label{sec:shadown hamiltonian}
-
-First-order derivative of a scalar function with respect to a vector variable (gradient vector):
-\begin{equation*}
-f_{\vt x} = \diff{f}{\vt x}
-\end{equation*}
-
-Second-order derivative of a scalar function with respect to two vector variables (Hessian matrix):
-\begin{equation*}
-f_{\vt x \vt y} = \diff{(\tr{f}_{\vt x})}{\vt y}.
-\end{equation*}
-
-Liouville operator:
-\begin{equation}
-\label{eq:general Lioville operator}
-\Liu{} = \tr{{\mathcal H}_{\vt p}} \diff{}{\vt r} - \tr{{\mathcal H}_{\vt r}} \diff{}{\vt p} = \{\circ,\Ham{}\}.
-\end{equation}
-
-For non-commutative operators $A$ and $B$ and a real constant $h$, the symmetric Baker-Campbell-Hausdorff (BCH) formula (Ref.~\citenum{Hairer_2006}, p.~86) can be expressed as
-\begin{equation}
-\label{eq:symmetric BCH exponential}
-e^{\frac{h}{2} B} e^{h A} e^{\frac{h}{2} B} = e^{h S},
-\end{equation}
-with
-\begin{equation}
-\label{eq:symmetric BCH operator}
-S = A + B + \frac{h^2}{24} \left[2 A + B,[A,B]\right] + \mathcal{O}(h^4).
-\end{equation}
-
-In this expression, $[A,B]$ is the commutator of operators $A$ and $B$, defined as $[A,B] = AB - BA$.
-The remaining terms of the infinite series involve nested commutators with increasing levels of recursion \cite{Hairer_2006}.
-If $A$ and $B$ are Liouville operators $\Liu A = \{\circ,\Ham A\}$ and $\Liu B = \{\circ,\Ham B\}$, respectively, then we have that
+%\begin{equation}
+%\label{eq:full_operator}
+%i\!L_\text{NVE} = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k,
+%\end{equation}
+%where the individual operators are defined as follows:
+%\[
+%\begin{split}
+%&boost:\\
+%&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
+%&translation: \\
+%& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
+%&rotation \; about \; a \; principal \; axis:\\
+%& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
+%\end{split}
+%\]
+%The individual operators can be evaluated analytically according to
+%\[
+%\begin{split}
+%&boost:\\
+%&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
+%&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
+%&translation:\\
+%&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
+%&rotation \; about \; a \; principal \; axis:\\
+%&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
+%&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
+%\end{split}
+%\]
+%Following Miller \textit{et al}.,\cite{Miller_2002} each complete rotation is factorized as
+%\begin{align*}
+%\label{eq:splitting_rot}
+%e^{\timestep i\!L_r} =  e^{\frac{\timestep}{2} i\!L_3} e^{\frac{\timestep}{2} i\!L_2} e^{\timestep i\!L_1} e^{\frac{\timestep}{2} i\!L_2} e^{\frac{\timestep}{2} i\!L_3},
+%\end{align*}
+%where ${\timestep}$ is the time step size. The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
+%\begin{equation}
+%\label{eq:trotter_splitting_NVE}
+%e^{\timestep i\!L_\text{NVE}} = e^{\frac{\timestep}{2} i\!L_b} e^{\timestep i\!L_r} e^{\timestep i\!L_t} e^{\frac{\timestep}{2} i\!L_b}.
+%\end{equation}
+
+\subsection{Backward Error Analysis}
+\label{sec:backward error analysis}
+
+Let us now represent the operators $\Liu A$ and $\Liu B$ using Poisson brackets involving Hamiltonians $\Ham A$ and $\Ham B$, that is, $\Liu A = \{\circ,\Ham A\}$ and $\Liu B = \{\circ,\Ham B\}$.
+By employing the properties of anti-commutativity and Jacobi identity \cite{Hairer_2006}, we can deduce that
 \begin{align*}
@@
 
-In the expression above, the Jacobi identity \cite{Hairer_2006} has been employed after some convenient permutations.
-This result means that the commutator of two Liouville operators is, in fact, a new Liouville operator $\Liu C$ with Hamiltonian $\Ham C = {\Liu A}{\Ham B}$.
-By induction, we can conclude that the operator $S$ in Eq.~\eqref{eq:symmetric BCH operator} is also a Liouville operator $\Liu S$ with Hamiltonian
-\begin{equation}
-\label{eq:symmetric BCH hamiltonian}
-\Ham S = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
-\end{equation}
-
-Considering a separable Hamiltonian $\Ham{}(\vt r, \vt p) = K(\vt p) + U(\vt r)$, where $K$ and $U$ are the kinetic and potential energies, respectively, the Velocity Verlet integrator consists in splitting $\Ham{} = \Ham A + \Ham B$, where $\Ham A = K(\vt p)$ and $\Ham B = U(\vt r)$.
-Therefore, $\Liu A = \tr{K_{\vt p}} \diff{}{\vt r}$ and $\Liu B = -\tr{U_{\vt r}} \diff{}{\vt p}$, which makes ${\Liu A}{\Ham B} = \tr{K_{\vt p}} U_{\vt r} = \tr{U_{\vt r}} K_{\vt p}$.
-Hence, Eq.~\eqref{eq:symmetric BCH hamiltonian} becomes
-\begin{equation*}
-\Ham S = K + U + \frac{{\Delta t}^2}{24} \left(2\tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p} - \tr{U_{\vt r}} K_{\vt p \vt p} U_{\vt r} \right) + \mathcal{O}({\Delta t}^4).
-%\Ham S = K + U + \frac{h^2}{12} \tr{\left(\diff{K}{\vt p}\right)}\left(\sdiff{U}{\vt r}\right) \left(\diff{K}{\vt p}\right) + \frac{h^2}{24} \tr{\left(\diff{U}{\vt r}\right)}\left(\sdiff{K}{\vt p}\right) \left(\diff{U}{\vt r}\right) + \mathcal{O}(h^4).
-\end{equation*}
-
-The kinetic energy is given by $K = \frac{1}{2} \tr{\vt p}{\mt M}^{-1}{\vt p}$, where $\mt M$ is a diagonal matrix containing the mass associated to each degree of freedom.
-This makes $K_{\vt p} = {\mt M}^{-1}{\vt p}$ and $K_{\vt p \vt p} = {\mt M}^{-1}$.
-The vector containing the force acting on each degree of freedom is $\vt F = -U_{\vt r}$.
-Thus,
-\begin{align*}
-\Ham S &= K + U + \frac{{\Delta t}^2}{24} \left(2\tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p} - \tr{\vt F} {\mt M}^{-1} {\vt F} \right) \\ &+ \mathcal{O}({\Delta t}^4).
-\end{align*}
-
-Let us introduce $\tilde{\mt M}$ as a mass tensor\cite{Bennett_1975} whose definition is given by
-\begin{equation*}
-\tilde{\mt M}^{-1} = {\mt M}^{-1} + \frac{{\Delta t}^2}{6} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1}.
-\end{equation*}
-
-Therefore,
-\begin{equation}
-\Ham S = \frac{1}{2}\tr{\vt p} \tilde{\mt M}^{-1} {\vt p} + {\tilde U} + \mathcal{O}({\Delta t}^4),
-\end{equation}
+This means that the commutator $[\Liu A,\Liu B]$ is, in fact, a new Liouville operator $\Liu C$ whose associated Hamiltonian is $\Ham C = {\Liu A}{\Ham B}$.
+By applying this procedure recursively, we can rewrite the right-hand side of Eq.~\eqref{eq:symmetric BCH} as $e^{\timestep \Liu S}$, where $\Liu S$ is the Liouville operator whose associated Hamiltonian is
+\begin{equation*}
+\label{eq:general shadow hamiltonian}
+\shadow{\Ham{}} = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
+\end{equation*}
+
+Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\shadow{\Ham{}}$ as above.
+In the literature, $\shadow{\Ham{}}$ is usually referred to as a shadow Hamiltonian.
+
+The first new result we present here is an analytical form for the shadow Hamiltonian concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~\citenum{Silveira_2017}.
+A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and we summarize the final results as follows.
+First of all, we rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
+\begin{equation*}
+K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\end{equation*}
+where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
+\begin{equation*}
+{\mt \Omega} = \left[\begin{array}{cc}
+{\mt M}^{-1} & \mt 0 \\
+\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
+\end{array}\right].
+\end{equation*}
+
+With $\nabla^2 U$ being the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, we define a shadow kinetic energy $\shadow K$ as
+\begin{equation*}
+\shadow K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \shadow{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\end{equation*}
 where
 \begin{equation*}
-{\tilde U} = U - \frac{{\Delta t}^2}{24}\tr{\vt F}{\mt M}^{-1} {\vt F}.
-\end{equation*}
-
-The important thing is that $\tilde{\mt M}$ depends exclusively on the coordinates, which makes $\diff{\tilde K}{\vt p} = \tilde{\mt M}^{-1} {\vt p}$.
+\shadow{\mathbf \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} {\mt \Omega} (\nabla^2 U) \tr{\mt \Omega}.
+\end{equation*}
+
+Although $\shadow{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
+As a result, the shadow kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
+
+In addition to $\shadow K$, we also define a shadow potential energy $\shadow U$ as
+\begin{equation*}
+\shadow U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
+\end{equation*}
+
+Finally, the shadow Hamiltonian for a systems of rigid body becomes
+\begin{equation}
+\shadow{\Ham{}} = \shadow K + \shadow U + \mathcal{O}(h^4).
+\end{equation}
+
+\subsection{Generalized Equipartition}
+
+The important thing is that $\shadow{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\shadow K}{\vt p} \\ \diff{\shadow K}{\vt \pi} \end{array}\right] = \shadow{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
 In this way, the equipartition theorem (valid at least in this direction?
@@
 
-Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
-Sergi and Ferrario\cite{Sergi2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.
+Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
+Sergi and Ferrario\cite{Sergi_2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.
 If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
@@
 will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric (that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$).
-A generalized Liouville operator can then be defined as\cite{Sergi2004}
+A generalized Liouville operator can then be defined as\cite{Sergi_2004}
 \[
@@
 The system might exhibit a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative.
-According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman2001} although the phase-space volume element $d\vt x$ is no longer conserved, there exits an invariant measure $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
+According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman_2001} although the phase-space volume element $d\vt x$ is no longer conserved, there exits an invariant measure $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
 Using the generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
@@
 
-We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a single Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
-This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom.
+We employ the technique of Sergi and Ferrario\cite{Sergi_2001} to couple a single Nos\'e-Hoover chain\cite{Martyna_1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
+This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom.
 To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
@@
 where $\mathcal H$ is the same as in Eq.~\eqref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$.
-As recommended in Ref.~\citenum{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
+As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
 The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
@@
 
-Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated.
-For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
-In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
-In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna1994}
+Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables must be eliminated.
+For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman_2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
+In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
+In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna_1994}
 
@@
 \label{sec:numerical_solvers}
-As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
+As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman_2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
 In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
 As a result, the whole propagator is symplectic and thus preserves the phase-space volume element.
-It is widely recognized that this property is important for numerical stability.\cite{Skeel1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion.
+It is widely recognized that this property is important for numerical stability.\cite{Skeel_1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion.
 In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator preserves the same measure.
-As the author has shown,\cite{Ezra2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
+As the author has shown,\cite{Ezra_2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
 Here, $\dot{\vt x}^{(k)}$ refers to the system of equations obtained from the individual operator $i\!L^{(k)}$, which might display its own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$.
@@
 \label{eq:trotter_splitting_NHC}
-e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
+e^{\timestep i\!L} = e^{\frac{\timestep}{2} i\!L_\text{NHC}} e^{\frac{\timestep}{2} i\!L_b} e^{\timestep i\!L_r} e^{\timestep i\!L_t}  e^{\frac{\timestep}{2} i\!L_b} e^{\frac{\timestep}{2} i\!L_\text{NHC}}.
 \end{equation}
@@
 \end{subequations}
-with $j$ varying from $1$ to $M-1$ in Eq.~\eqref{eq:iL_NHC_j}. Then, each propagator $e^{\frac{\Delta t}{2} i\!L_\text{NHC}}$ present in Eq.~\eqref{eq:iL_NHC} is factorized as
+with $j$ varying from $1$ to $M-1$ in Eq.~\eqref{eq:iL_NHC_j}. Then, each propagator $e^{\frac{\timestep}{2} i\!L_\text{NHC}}$ present in Eq.~\eqref{eq:iL_NHC} is factorized as
 \[
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \Bigg[ \prod_{j=M}^1 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
+e^{\frac{\timestep}{2} i\!L_\text{NHC}} = \Bigg[ \prod_{j=M}^1 e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
 \]
@@
 We might note that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^3 {(-1)^n x^n}/{(n+1)!}$ whenever $|x| \leq 10^{-4}$.
-The solution above is equivalent to that found in Ref.~\citenum{Martyna1996}.
+The solution above is equivalent to that found in Ref.~\citenum{Martyna_1996}.
 
@@
 
-In the new propagator, $e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} }$ is removed:
+In the new propagator, $e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(0)} }$ is removed:
 \[
-e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast} = \Bigg[ \prod_{j=M}^2 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(1)} } \Bigg[ \prod_{j=2}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
+e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast} = \Bigg[ \prod_{j=M}^2 e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(1)} } \Bigg[ \prod_{j=2}^M e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
 \]
 
-We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj2005} does not fulfill this criterion. 
+We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj_2005} does not fulfill this criterion. 
 Finally, the complete scheme is
@@
 \begin{split}
-e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast} e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b^\ast} e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast}.
+e^{\timestep i\!L} = e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast} e^{\frac{\timestep}{2} i\!L_b^\ast} e^{\timestep i\!L_r} e^{\timestep i\!L_t}  e^{\frac{\timestep}{2} i\!L_b^\ast} e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast}.
 \end{split}
@@
 In this subsection we briefly review a simple method that serves for checking the consistency between the sampled and theoretical energy distributions.
-The procedures, implemented by Shirts,\cite{Shirts2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
+The procedures, implemented by Shirts,\cite{Shirts_2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
 First, recall that in the NVT ensemble the probability of observing an energy E is given by
@@
 
-In this section, we analyze the numerical integrators introduced in Subsection~\ref{sec:numerical_solvers}, and compare the results with those obtained employing the method of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS\cite{Plimpton1995} software package.
+In this section, we analyze the numerical integrators introduced in Subsection~\ref{sec:numerical_solvers}, and compare the results with those obtained employing the method of Kamberaj~\textit{et al.}\cite{Kamberaj_2005}, which is implemented in the LAMMPS\cite{Plimpton_1995} software package.
 For that, we considered a system of 900 TIP3P\cite{Jorgensen_1983} water molecules in a cubic box of length {\AA} subject to periodic boundary conditions.
-The intermolecular interactions were truncated at 9 \AA and to avoid a discontinuity in the force, we applied the so-called shifted-force (SF) method\cite{Allen1989, Fennell2006, Toxvaerd_2011} to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
+The intermolecular interactions were truncated at 9 \AA and to avoid a discontinuity in the force, we applied the so-called shifted-force (SF) method\cite{Allen_1989, Fennell_2006, Toxvaerd_2011} to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
 
@@
 In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
-By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
+By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
 %Resultados sobre o checkensemble:
@@
 \begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
 	\label{table:new_results}
@@
 			\cline{2-4}
-			$\Delta t$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
 			\hline
@@
 \begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj2005}}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
 	\label{table:new_results_2}
@@
 \begin{tabular}{cccccccc}
-$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
 
@@
 \hline
-\multicolumn{8}{c}{Method of Kamberaj \textit{et al}.\cite{Kamberaj2005}} \\
+\multicolumn{8}{c}{Method of Kamberaj \textit{et al}.\cite{Kamberaj_2005}} \\
 1.0 & $299.53\pm0.02$ & $-8470.9\pm0.5$ & $1611.6\pm0.1$ & $806.4\pm0.1$ & $805.2\pm0.1$ & $52\pm2$ & $0.023$ \\
@@
 \hline
-\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of \citeauthor{Kamberaj2005} \cite{Kamberaj2005}} \\
+\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}} \\
 1.0 & $299.52\pm0.02$ & $-8470.2\pm0.5$ & $1611.6\pm0.1$ & $805.9\pm0.1$ & $805.6\pm0.1$ & $58\pm2$ & $0.272$ \\
@@
 \begin{figure}
-    \caption{SHAKE: $\Delta t$ = 1 fs ($\bar{T}$ = 299.98 K, $\bar{U}$ = -8428.86 kcal/mol) HMC: $\bar{U}$ = -8470.71 kcal/mol  }
+    \caption{SHAKE: $\timestep$ = 1 fs ($\bar{T}$ = 299.98 K, $\bar{U}$ = -8428.86 kcal/mol) HMC: $\bar{U}$ = -8470.71 kcal/mol  }
 	\includegraphics{potenergy}
@@
 \begin{figure}
-    \caption{$\Delta t$ = 1 fs ($\bar{T}$ = 300.04 K, $\bar{U}$ = -8465.63 kcal/mol)}
+    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 300.04 K, $\bar{U}$ = -8465.63 kcal/mol)}
 	\includegraphics{potenergy2}
@@
 \begin{figure}
-    \caption{$\Delta t$ = 0.5 fs ($\bar{T}$ = 299.99 K, $\bar{U}$ = -8468.19 kcal/mol)}
+    \caption{$\timestep$ = 0.5 fs ($\bar{T}$ = 299.99 K, $\bar{U}$ = -8468.19 kcal/mol)}
 	\includegraphics{potenergy_05}
@@
 \begin{figure}
-    \caption{$\Delta t$ = 1 fs ($\bar{T}$ = 299.52 K, $\bar{U}$ = -8470.21 kcal/mol)}
+    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 299.52 K, $\bar{U}$ = -8470.21 kcal/mol)}
 	\includegraphics{potenergy_kamb}
@@
 	\caption{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
-Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
+Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
 	\label{fig:checkensemble}
@@
 	\caption{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\eqref{eq:mb}.
-Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\Delta t$ = 1~fs, $\bar{T}$ = 304.99~K.}
+Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}$ = 304.99~K.}
 \label{fig:mbdistribution}
@@
 Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
-It can be seen that the deviations from the true slope, $\Delta T$ = 8.970~K, are less than $1\sigma$.
+It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
 This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
@@
 			\begin{tabular}{ccccc}
-				& \multicolumn{4}{c}{true $\Delta T$ = 8.970 K} \\
+				& \multicolumn{4}{c}{true $\timestep$ = 8.970 K} \\
 				\cline{2-5}
@@
 				\hline
-				Method  &$\Delta T/K$ & $\sigma$ dev & $\Delta T/K$ & $\sigma$ dev \\
+				Method  &$\timestep/K$ & $\sigma$ dev & $\timestep/K$ & $\sigma$ dev \\
 				\hline % inserts single-line 
@@
 %  \includegraphics{pressure2}
-%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars). Simulation of liquid water with NVE integration. $\Delta t : 1 fs$  }
+%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars). Simulation of liquid water with NVE integration. $\timestep : 1 fs$  }
 %\end{center}
@@
 \section{Shadow Hamiltonian of Rigid Bodies}
+\label{sec:rigid body shadow hamiltonian}
 
@@
 \begin{equation*}
-\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\Delta t}^4),
+\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
 \end{equation*}
@@
 \begin{equation*}
-\tilde{U} = U - \frac{\Delta t^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right)
+\tilde{U} = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right)
 \end{equation*}
@@
 \begin{equation*}
-\tilde{\mt \Omega} = {\mt \Omega} + \frac{\Delta t^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} + {\mt \Gamma} \right)
+\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} + {\mt \Gamma} \right)
 \end{equation*}
@@
 %		\begin{tabular}{cccccccc}
-%			$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+%			$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
 %			\hline
@@
 %\begin{tabular}{cccccccc}
-%$\Delta t$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+%$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
 %\hline
@@
 %				\cline{2-5}
-%				$\Delta t$/fs & $D\!E$ & $T$/K & $D\!E$ & $T$/K \\
+%				$\timestep$/fs & $D\!E$ & $T$/K & $D\!E$ & $T$/K \\
 %				\hline
@@
 %\[
-%e^{\frac{\Delta t}{2} i\!L_\text{NHC}} = \Pi_{j=1}^n \Bigg[ \prod_{j=M}^1 e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\Delta t}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\Delta t}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
+%e^{\frac{\timestep}{2} i\!L_\text{NHC}} = \Pi_{j=1}^n \Bigg[ \prod_{j=M}^1 e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
 %\]
@@
 %\begin{split}
-%e^{\frac{\Delta t}{2} i\!L_\text{NHC}^\ast } = &\prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-%\times &\exp\left(-\frac{\Delta t}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j}\right) \\
-%\times &\prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
+%e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast } = &\prod_{j=M}^1 \exp\left[\frac{\timestep}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
+%\times &\exp\left(-\frac{\timestep}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j}\right) \\
+%\times &\prod_{j=1}^M \exp\left[\frac{\timestep}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
 %\end{split}
@@
 
-%In order to evaluate the stability of the numerical integrators, we employ a measure given by%\cite{Tuckerman2010}
+%In order to evaluate the stability of the numerical integrators, we employ a measure given by%\cite{Tuckerman_2010}
 %\begin{equation}
@@
 %from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ %(either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ %after a certain number of time steps $\mathcal N$.
-% \textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}.
+% \textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\timestep/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}.
 
@@
 
-%The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
+%The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller_2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
 \end{document}

### 2018-02-12 rigid_bodies.tex
Organization of theory section

@@
 \newcommand{\timestep}{h}
-\newcommand{\shadow}[1]{\widetilde{#1}}
+\newcommand{\modified}[1]{\widetilde{#1}}
 
@@
 	\begin{align}
+%
 	&\dot{\vt r} =
@@
 	{\mt M}^{-1} {\vt p}, \\
+%
 	&\dot{\vt p} =
@@
 	{\vt F}, \\
+%
 	&\dot{\vt q} =
@@
 	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
+%
 	&\dot{\vt \pi} =
@@
 	\begin{align}
+%
+\label{eq:nhc_r}
 	&\dot{\vt r} =
 	{\mt M}^{-1} {\vt p}, \\
+%
+\label{eq:nhc_p} 
 	&\dot{\vt p} =
-	{\vt F} - \alpha_1 {\vt p}, \\
+	{\vt F} - \alpha_1 {\vt p},\\
+%
+\label{eq:nhc_q}
 	&\dot{\vt q} =
 	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \\
+%
+\label{eq:nhc_pi}
 	&\dot{\vt \pi} =
 	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
+%
+\label{eq:nhc_eta}
 	&\dot{\eta}_j = \alpha_j, \text{ and} \\
+%
+\label{eq:nhc_p_eta}
 	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M,
@@
 where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set as null, as usual (otherwise, $L = 6N$).\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
-%Finally, $\shadow K$ is the total kinetic energy of the system, whose computation (not using Eq.~\ref{eq:kinetic energy}, hence the distinct symbol) is the central feature of the method we propose.
-
+%Finally, $\modified K$ is the total kinetic energy of the system, whose computation (not using Eq.~\eqref{eq:kinetic energy}, hence the distinct symbol) is the central feature of the method we propose.
 The ``extended energy'' which is conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
@@
 \label{eq:nvt extended energy}
-H_\text{\tiny NVT} = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
+H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
 \end{equation}
@@
 non-Hamiltonian) Liouville operator, the splitting goes as
-\begin{equation*}
+\begin{equation}
+\label{eq:splitting NVT}
 e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
-\end{equation*}
+\end{equation}
 where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
@@
 Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
-%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\shadow K$ to be scaled by the square of it, hence ${\shadow K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\shadow K}^\ast$.
-In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\shadow K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
+%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\modified K$ to be scaled by the square of it, hence ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$.
+In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\modified K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
 Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_1}$ to $e^{\frac{\timestep}{4n} \mathcal{L}_M}$ are applied again, but now in the increasing order of their indices.
-It is worth noting that the $\eta$'s are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H_\text{\tiny NVT}$.
+It is worth noting that the $\eta$'s are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H$.
 
 The whole procedure described above is repeated $n$ times.
-Because of its low computational cost, the compensation required for increasing the time step $\timestep$ without degrading the accuracy of the NHC part (that is, employing a larger $n$) causes a small impact in the overall integration effort.
+Because of its low computational cost, the compensation required for increasing the time step $\timestep$ without degrading the accuracy of the NHC part (i.e.
+employing a larger $n$) causes a small impact in the overall integration effort.
 In contrast, evaluating the NVE part is expensive for involving force/torque computations and its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
-A high-order integration scheme\cite{Omelyan_2007, vanZon_2008} could possibly admit larger time steps, but it would raise both cost and complexity because for entailing the computation (or numerical estimation) of force/torque gradients.
+A high-order integration scheme\cite{Omelyan_2007, vanZon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
 The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
-
-Here we take a simpler path.
+Nevertheless, we choose to take a simpler path.
 Instead of attempting to improve the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we try to quantify the obtained deviations and then use them to tune the thermostatting procedure and properly weight the sampled configurations when canonical averages are computed.
 
-A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are also Hamiltonian.
-Thus, a trajectory generated to be roughly consistent with the specified Hamiltonian $\Ham{}$ will be exactly consistent with a nearby, albeit unknown, function $\shadow{\mathcal H}$.
-It is also known that $\shadow{\mathcal H}$, referred to as a shadow Hamiltonian, depends on the time step size $\timestep$ and that, for a time-reversible splitting like Eq.~\eqref{eq:splitting NVE}, $\shadow{\mathcal H} = \Ham{} + \mathcal{O}(\timestep^2)$.
-Estimating the trailing term of this equality, either analytically or numerically, is a task of backward error analysis.
-As demonstrated in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be better approximated as $\shadow{\Ham{}} = \shadow K + \shadow U + \mathcal{O}(h^4)$, where $\shadow K$ and $\shadow U$ will be called shadow kinetic and shadow potential energies, respectively.
-The former is given by
+A known property of splitting methods applied to Hamiltonian ODE systems is that they provide Hamiltonian approximate solutions.
+In other words, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian.
+Thus, whereas $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors).
+It is also known that $\Ham{S}$ depends on the time step size $\timestep$ and that, for a time-reversible splitting such as Eq.~\eqref{eq:splitting NVE}, one can write $\Ham{S} = \Ham{} + \mathcal{O}(\timestep^2)$.
+Estimating the trailing term is a task of backward error analysis.
+
+As shown in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be better approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are named modified kinetic and modified potential energies, respectively.
+The former can be expressed as
 \begin{equation}
 \label{eq:shadow kinetic energy}
-\shadow K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation}
-where $\shadow{\mathbf \Omega}$ is a complicated matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, but independent of both $\vt p$ and $\vt \pi$.
-The shadow potential is also independent of these momenta and is given by
+\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\end{equation}
+where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$. The modified potential, which is also independent of $\vt p$ and $\vt \pi$, is given by
 \begin{equation}
 \label{eq:shadow potential energy}
-\shadow U = U - \frac{\timestep^2}{24} \left[ \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A}(\vt q) {\mt I}^{-1} {\mt A}(\vt q) {\vt \tau} \right].
-\end{equation}
-
-Therefore, the outcome of solving Eq.~\eqref{eq:ODE system for NVE} using a Trotter-Suzuki splitting will be close to the exact solution of Eq.~\eqref{eq:ODE system for NVE}, but it will be even closer to the solution of a system composed of equations
+\modified U = U - \frac{\timestep^2}{24} \left[ \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right],
+\end{equation}
+where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the torques.
+Therefore, the outcome of solving Eq.~\eqref{eq:ODE system for NVE} using a Trotter-Suzuki splitting will be close to the exact solution of Eq.~\eqref{eq:ODE system for NVE}, but it will be even closer to the solution of
 \begin{equation}
@@
 \left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] =
-\left[\begin{array}{c} \diff{\shadow K}{\vt p} \\ \diff{\shadow K}{\vt \pi} \end{array}\right] =
-\shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation}
-as well as $\dot{\vt p} = -\diff{\shadow{\mathcal H}}{\vt r}$ and $\dot{\vt \pi} = -\diff{\shadow{\mathcal H}}{\vt q}$.
-In our proposed method, we acknowledge this fact and thus use the Nos\'e-Hoover chain to provide a canonical distribution consistent with $\shadow{\mathcal H}$ (rather than $\Ham{}$) at temperature $T$.
-Note that this entails solving Eq.~\eqref{eq:ODE system for NVT} exactly as described before, with the only requirement of replacing $K$ by $\shadow K$ in the definition of the generalized force $G_1$.
-By contrasting Eqs.~\eqref{eq:generalized force NVT}, \eqref{eq:shadow kinetic energy}, and \eqref{eq:shadow ODE system}, we can see that
-\begin{equation}
-G_1 = 2 \shadow K - L k_\text{\tiny B} T.
-\end{equation}
-
-Direct evaluation of Eq.~\eqref{eq:shadow kinetic energy} for computing $\shadow K$ is quite complex and expensive.
-However, one can do it very easily using the approach devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the derivatives in the left-hand side of Eq.~\eqref{eq:shadow ODE system}.
-Denoting the obtained estimators as $\dot{\vt r}_\text{\tiny NVE}$ and $\dot{\vt q}_\text{\tiny NVE}$ and substituting Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:shadow kinetic energy}, we find out that
-\begin{equation}
-\shadow K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}_\text{\tiny NVE} + \tr{\vt \pi} \dot{\vt q}_\text{\tiny NVE} \right).
+\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] =
+\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\end{equation}
+together with $\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}$ and $\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}$, where $\modified{\Ham{}} = \modified K + \modified U$.
+Our proposal consists in acknowledging this fact and then tuning the Nos\'e-Hoover chain to provide a canonical distribution consistent with $\modified{\Ham{}}$ (instead of $\Ham{}$) at temperature $T$.
+Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly as described before.
+The only requirement is to replace $K$ by $\modified K$ in the definition of the generalized force in thermostat $1$, given by Eq.~\eqref{eq:generalized force NVT}.
+By contrasting this one with Eqs.~\eqref{eq:shadow kinetic energy} and \eqref{eq:shadow ODE system}, we observe that such force becomes
+\begin{equation}
+G_1 = 2 \modified K - L k_\text{\tiny B} T.
+\end{equation}
+
+It is possible to evaluate Eq.~\eqref{eq:shadow kinetic energy} directly for computing $\modified K$, but this is quite complex and expensive.
+Instead, one can do it very easily by means of a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the derivatives in the left-hand side of Eq.~\eqref{eq:shadow ODE system}.
+By denoting the obtained estimators as $\dot{\vt r}_\text{\tiny NVE}$ and $\dot{\vt q}_\text{\tiny NVE}$, and substituting Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:shadow kinetic energy}, we find out that
+\begin{equation}
+\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}_\text{\tiny NVE} + \tr{\vt \pi} \dot{\vt q}_\text{\tiny NVE} \right).
 \end{equation}
@@
 
-- Scaling of $\shadow K$ during the NHC integration.
+- Scaling of $\modified K$ during the NHC integration.
 
@@
 \label{eq:reweighting}
-\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
+\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
 \end{equation}
@@
 
-\subsection{Microcanonical Ensemble}
-
-When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of a center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t) = \tr{[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}.
-In this way, the position of any constituent particle $j$ will vary with time as $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in a body-fixed frame of reference centered in $\vt r$ and ${\mt A}(\vt q)$ is the body's orientation matrix (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components.
-The superscript $t$ stands for transposition, which is equivalent to inversion in the case of a canonical transformation like $\mt A$ \cite{Goldstein_2002}.
-The central feature explored in Ref.~\citenum{Silveira_2017} is the fact that $\mt A$ admits the simple factorization
-\begin{equation}
-\label{eq:factorization_of_A}
-{\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q),
-\end{equation}
-where
-\begin{equation*}
-\label{eq:def_B_and_C}
-\mt B(\vt q) = \begin{smallarray}{rrrr}
--q_2 & -q_3 & -q_4 \\
- q_1 & -q_4 &  q_3 \\
- q_4 &  q_1 & -q_2 \\
--q_3 &  q_2 &  q_1
-\end{smallarray}
-\; \text{and} \;
-\mt C(\vt q) = \begin{smallarray}{rrrr}
--q_2 & -q_3 & -q_4 \\
- q_1 &  q_4 & -q_3 \\
--q_4 &  q_1 &  q_2 \\
- q_3 & -q_2 &  q_1
-\end{smallarray}.
-\end{equation*}
-
+\subsection{Hamiltonian Dynamics}
+\label{eq:hamiltonian dynamics}
+
+When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of the center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t)$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}.
+In this way, the connection between the body-fixed and space-fixed frames of reference can be recovered at any time by the rotation matrix $\mt A(\vt q)$, which admits the factorization mentioned in Sec.~\ref{sec:methodology}.
 This factorization creates an intermediate frame of reference for representing three-dimensional vectors in $\mathbb{R}^3$ as quaternions that lie in the hyperplane orthogonal to $\vt q$, since the columns of matrix $\mt B$ and those of matrix $\mt C$ form two different bases for such a subspace of $\mathbb{R}^4$.
 Transformations amongst three distinct reference frames are enacted by multiplications with these matrices or their transposes, such as represented in the following scheme:
-\[
+\begin{equation*}
 \boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
@@
 \boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
-\]
+\end{equation*}
 
 The intermediate frame is exactly the subspace of $\mathbb{R}^4$ that must contain the rate-of-change vector $\dot{\vt q}$ if the norm of $\vt q$ is to remain constant, since $\tr{\vt q} {\vt q} = 1$ implies that $\tr{\vt q} \dot{\vt q} = 0$.
-In the Hamiltonian approach for the motion of a rigid body, the momentum $\vt \pi$ conjugated to $\vt q$ is given by $\vt \pi = 2 \mt B(\vt q) \mt I \vt \omega$, where $\vt \omega$ is the three-dimensional angular velocity and $\mt I$ is a diagonal matrix containing the body's principal moments of inertia.\cite{Silveira_2017} This means that $\vt \pi$ is the quaternion-fixed representation of the angular momentum $\vt L = \mt I \vt \omega$, thus being orthogonal to $\vt q$ as well.
-After this notation has been introduced, the Hamiltonian of a rigid body can be written as
-\begin{equation}
-\label{eq:Hamiltonian}
-\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
-\end{equation}
-where $K$ and $U$ are the kinetic and potential energies of the body, respectively. While the potential energy can have any functional form, the kinetic energy is given by
-\begin{equation}
-\label{eq:kinetic energy}
-K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi},
-\end{equation}
-where $\vt p$ is the linear momentum of the body and $\mt M$ is a diagonal matrix containing the body's mass in all three diagonal entries.
-Note that the kinetic energy is composed of two independent contributions, a translational and a rotational one.
-An advantage of this notation is that Eqs.~\eqref{eq:Hamiltonian} and \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies once we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the respective properties of all bodies piled together.
-The same occurs for matrices $\mt M$ and $\mt I$, which turn into diagonal matrices of size $3N \times 3N$.
-In addition, $\mt B(\vt q)$ becomes a block-diagonal matrix with $4N$ rows and $3N$ columns.
-
-The Liouville operator \cite{Tuckerman_2010} that stems from the Hamiltonian in Eq.~\eqref{eq:Hamiltonian} is\cite{Silveira_2017}
-\begin{align*}
+In the Hamiltonian approach for the motion of a rigid body, the momentum $\vt \pi$ conjugated to $\vt q$ is given by $\vt \pi = 2 \mt B(\vt q) \mt I \vt \omega$.\cite{Silveira_2017} This means that $\vt \pi$ is the quaternion-fixed representation of the angular momentum $\vt L = \mt I \vt \omega$, thus being orthogonal to $\vt q$ as well.
+
+The Liouville operator that stems from the Hamiltonian in Eq.~\eqref{eq:Hamiltonian} is\cite{Silveira_2017}
+\begin{equation}
 \label{eq:Lioville operator}
-\Liu{\tiny NVE} &= \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} \\
-&+ \tr{\vt F} \diff{}{\vt p} + \left[ \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \right] \diff{}{\vt \pi},
-\end{align*}
-where $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body (or on $N$ bodies), both represented in the space-fixed frame of reference, whereas $\vt \omega$ can be obtained by
-\begin{equation*}
-\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
-\end{equation*}
-
-The time evolution of any phase-space vector $\vt x$ is governed by $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ \cite{Tuckerman_2010}.
-Therefore, the equations of motion that describe the rigid-body motion are \cite{Silveira_2017}
-\begin{subequations}
-	\label{eq:EDO_system}
-	\begin{align}
-	&\dot{\vt r} =
-	%+\diff{\Ham{}}{\vt p} =
-	{\mt M}^{-1} {\vt p}, \\
-	&\dot{\vt p} =
-	%-\diff{\Ham{}}{\vt r} = 
-	{\vt F}, \\
-	&\dot{\vt q} =
-	%+\diff{\Ham{}}{\vt \pi} =
-	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
-	&\dot{\vt \pi} =
-	%-\diff{\Ham{}}{\vt q} =
-	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
-	\end{align}
-\end{subequations}
-
+\begin{split}
+\Liu{\tiny NVE} &= \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \tr{\vt F} \diff{}{\vt p} \\
+&+ \left[ \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \right] \diff{}{\vt \pi}.
+\end{split}
+\end{equation}
+
+Because the time evolution of any phase-space vector $\vt x$ is governed by $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ \cite{Tuckerman_2010}, the equations of motion that describe the rigid-body motion are those in Eq.~\eqref{eq:ODE system for NVE}.
 The solution for a system like $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ with initial condition $\vt x(0) = \vt x_0$ can be formally expressed as $\tr{\vt x}(t) = e^{t \Liu{\tiny NVE}} \tr{\vt x}_0$.
@@
 For two non-commutative Liouville operators $\Liu A$ and $\Liu B$ and a real constant $\timestep$, the symmetric Baker-Campbell-Hausdorff (BCH) formula can be expressed as \cite{Hairer_2006}
-\begin{align}
+\begin{equation}
 \label{eq:symmetric BCH}
-&e^{\frac{\timestep}{2} \Liu B} e^{\timestep \Liu A} e^{\frac{\timestep}{2} \Liu B} = \nonumber \\
+\begin{split}
+&e^{\frac{\timestep}{2} \Liu B} e^{\timestep \Liu A} e^{\frac{\timestep}{2} \Liu B} = \\
 &= e^{\timestep (\Liu A + \Liu B) + \frac{\timestep^3}{24} \left[2 \Liu A + \Liu B,[\Liu A,\Liu B]\right] + \mathcal{O}(\timestep^5)},
-\end{align}
+\end{split}
+\end{equation}
 where $[X,Y]$ is the commutator of operators $X$ and $Y$, defined as $[X,Y] = XY - YX$.
 The remaining terms of the infinite series involve ever-deepening, nested commutators \cite{Hairer_2006}.
-With $\Liu{\tiny NVE} = \Liu{A} + \Liu{B}$ and a small $\timestep$ value, we can approximate the BCH equation as
-\begin{equation*}
-e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
-\end{equation*}
-which is known as the Trotter-Suzuki splitting formula \cite{Trotter_1959, Suzuki_1976}. In the rigid-body case, by making
+With $\Liu{\tiny NVE} = \Liu{A} + \Liu{B}$ and a small value for $\timestep$, we can approximate the BCH equation by the Trotter-Suzuki splitting formula, Eq.~\eqref{eq:splitting NVE}.
+In the rigid-body case, by making
 \begin{subequations}
 	\label{eq:Liouvile operators K and U}
-\begin{equation}
+	\begin{equation*}
 	\label{eq:Lioville operator K}
 	\Liu{A} = \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) \diff{}{\vt \pi}
-\end{equation}
+	\end{equation*}
 	and
-\begin{equation}
+	\begin{equation*}
 	\label{eq:Lioville operator U}
 	\Liu{B} = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \diff{}{\vt \pi},
-\end{equation}
+	\end{equation*}
 \end{subequations}
-it becomes possible to determine exactly the individual actions of operators $e^{\timestep \Liu{A}}$ and $e^{\frac{\timestep}{2} \Liu{B}}$, with $\timestep$ being a small time step $\Delta t$.
-The latter operator promotes a kick in the linear and quaternion momenta given by ${\vt p} \leftarrow {\vt p} + \frac{\timestep}{2} {\vt F}$ and ${\vt \pi} \leftarrow {\vt \pi} + \timestep {\mt C}({\vt q}) {\vt \tau}$.
-The former, in turn, performs a simultaneous force-free translation and torque-free rotation during the length of a time step.
-%
-%
-%first one is the solution of
-%\begin{subequations}
-%	\label{eq:EDO translation and rotation}
-%	\begin{align}
-%	&\dot{\vt r} =
-%	%+\diff{\Ham{}}{\vt p} =
-%	{\mt M}^{-1} {\vt p}, \\
-%	&\dot{\vt q} =
-%	%+\diff{\Ham{}}{\vt \pi} =
-%	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
-%	&\dot{\vt \pi} =
-%	%-\diff{\Ham{}}{\vt q} =
-%	\frac{1}{2} \mt B(\vt \pi) \vt \omega, \label{eq:EDO_pi}
-%	\end{align}
-%\end{subequations}
-%
-% and $N_s$ is the number of steps. We emphasize that each exponential operator above will act on (that is, have as initial condition) the result of the operator at its right-hand side.
-%
-%
-%
-%for this system of equations e rigid body motion which preserves the symplectic structure of Eq.~\eqref{eq:EDO_system}.
-For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula isThe can be used to devise
-%The full integrator is second-order in $\timestep$ because the leading-order term of the error at every step is proportional to $\timestep^3$ and, in addition, $N_s = t/\timestep$.\cite{Tuckerman_2010} The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
-%\begin{equation}
-%\label{eq:trotter_exact_rotation}
-%e^{\timestep i\!L} = e^{\frac{\timestep}{2} i\!L_b} e^{\timestep i\!L_{t+r}} e^{\frac{\timestep}{2} i\!L_b} + \mathcal{O}(\timestep^3).
-%\end{equation}
-%
-%This is the Velocity Verlet algorithm\cite{Tuckerman_2010} with center-of-mass translations and rigid-body rotations taking place simultaneously.
-Update of forces and torques is required only after the step of translation \& rotation has been completed.
-Because the solution of torque-free rotation involves a single step in the numerical solver, the scheme represented by Eq.~\ref{eq:trotter_exact_rotation} will be referred to as the unsplit rotation integrator.
-We remark that van Zon and co-workers\cite{vanZon2007PRE, vanZon2008} have already employed this algorithm using the formulation proposed in Ref.~\citenum{vanZon2007}.
-Other splitting schemes with higher-order approximations have also been used for rigid bodies.\cite{Omelyan2007, Omelyan2008, vanZon2008} However, they result in methods that are more complex and often require expensive force-gradient calculations.
-Moreover, it is unclear how their benefits would extend to non-Hamiltonian numerical integrators, such as those often employed for molecular dynamics in ensembles other than the microcanonical one.
-%
-%
-%
-%
-%
-%
-%
-%The approximate solution for rotation is derived by opening the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ from each entry of $\vt \omega$: 
-%\begin{equation}
-%\label{eq:omega_entry}
-%\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
-%\end{equation}
-%where $\hat{\mt B}_k$ is one of the following permutation matrices:
-%\[
-%\hat{\mt B}_1 = \begin{smallarray}{cccc}
-% 0 & \text{-}1 &  0 &  0 \\
-% 1 &  0 &  0 &  0 \\
-% 0 &  0 &  0 &  1 \\
-% 0 &  0 & \text{-}1 &  0 \\
-%\end{smallarray} \quad
-%\hat{\mt B}_2 = \begin{smallarray}{cccc}
-% 0 &  0 & \text{-}1 &  0 \\
-% 0 &  0 &  0 & \text{-}1 \\
-% 1 &  0 &  0 &  0 \\
-% 0 &  1 &  0 &  0 \\
-%\end{smallarray} \quad
-%\hat{\mt B}_3 = \begin{smallarray}{cccc}
-% 0 &  0 &  0 & \text{-}1 \\
-% 0 &  0 &  1 &  0 \\
-% 0 & \text{-}1 &  0 &  0 \\
-% 1 &  0 &  0 &  0 \\
-%\end{smallarray}
-%\]
-%The new form of the equations of motion in Eqs.~\eqref{eq:EDO_q} and \eqref{eq:EDO_pi} are then:
-%\begin{subequations}
-%\label{eq:EDO_system_alternative}
-%\begin{align}
-%&\dot{\vt q}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt q_i \label{eq:edo_q} \\
-%&\dot{\vt \pi}_i = \frac{1}{2} \left( \sum_{k=1}^3 \omega_{ik} \hat{\mt B}_k \right) \vt \pi_i + 2 \mt C_i \vt \tau_i \label{eq:edo_pi}
-%\end{align}
-%\end{subequations}
-%
-%In this context, the Hamiltonian of Eq.\eqref{eq:H_NVE} can be rewritten as
-%\begin{equation}
-%\label{eq:H_split_omega}
-%\mathcal{H} =  U(\vt r, \vt q) +  \sum_{i=1}^N \left( \frac{\tr{\vt p_i} \vt p_i}{2m_i} + \frac{1}{2} \sum_{k=1}^3 I_{ik} \omega_{ik}^2\right),
-%\end{equation}
-%from which we recognize partial Hamiltonians, each of one generates equations of motion with known analytical solutions.
-In particular, we have $\mathcal{H}_b = U(\vt r, \vt q)$, $\mathcal{H}_{t} = \frac{1}{2m}\tr{\vt p} \vt p$, and $\mathcal{H}_k = (1/2) I_k \omega_k^2$.
-Of corse, $\mathcal{H} = \mathcal{H}_b + \mathcal{H}_t + \sum_{k=1}^3 \mathcal{H}_k$.
-The NVE dynamics of this system is governed by a Liouville operator given by\cite{Silveira_2017}
-%\begin{equation}
-%\label{eq:full_operator}
-%i\!L_\text{NVE} = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k,
-%\end{equation}
-%where the individual operators are defined as follows:
-%\[
-%\begin{split}
-%&boost:\\
-%&i\!L_b = \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right) \\
-%&translation: \\
-%& i\!L_t = \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i} \\
-%&rotation \; about \; a \; principal \; axis:\\
-%& i\!L_k = \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right)
-%\end{split}
-%\]
-%The individual operators can be evaluated analytically according to
-%\[
-%\begin{split}
-%&boost:\\
-%&{\vt p} = {\vt p}_0 + {\vt F}_0 t \\
-%&{\vt \pi} = {\vt \pi}_0 +  2 {\mt C}_0 {\vt \tau}_0 t\\
-%&translation:\\
-%&{\vt r} = {\vt r}_0 + (t/m) {\vt p}_0\\
-%&rotation \; about \; a \; principal \; axis:\\
-%&{\vt q}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt q_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt q_0 \\
-%&{\vt \pi}(t) = \cos\left(\frac{\omega_k t}{2}\right) \vt \pi_0 + \sin\left(\frac{\omega_k t}{2}\right) \hat{\mt B}_k \vt \pi_0
-%\end{split}
-%\]
-%Following Miller \textit{et al}.,\cite{Miller_2002} each complete rotation is factorized as
-%\begin{align*}
-%\label{eq:splitting_rot}
-%e^{\timestep i\!L_r} =  e^{\frac{\timestep}{2} i\!L_3} e^{\frac{\timestep}{2} i\!L_2} e^{\timestep i\!L_1} e^{\frac{\timestep}{2} i\!L_2} e^{\frac{\timestep}{2} i\!L_3},
-%\end{align*}
-%where ${\timestep}$ is the time step size. The splitting of the Liouville operator of Eq.~\eqref{eq:full_operator} we employ in this paper is
-%\begin{equation}
-%\label{eq:trotter_splitting_NVE}
-%e^{\timestep i\!L_\text{NVE}} = e^{\frac{\timestep}{2} i\!L_b} e^{\timestep i\!L_r} e^{\timestep i\!L_t} e^{\frac{\timestep}{2} i\!L_b}.
-%\end{equation}
+it becomes possible to determine exactly the individual actions of operators $e^{\timestep \Liu{A}}$ and $e^{\frac{\timestep}{2} \Liu{B}}$ because they yield two separate, analytically solvable systems of equations.
+The one that originates from $e^{\timestep \Liu{A}}$ is made up of equations $\dot{\vt r} = {\mt M}^{-1} {\vt p}$, $\dot{\vt p} = {\vt 0}$, $\dot{\vt q} = \frac{1}{2} \mt B(\vt q) \vt \omega$, and $\dot{\vt \pi} = \frac{1}{2} \mt B(\vt \pi) \vt \omega$, while the one that stems from $e^{\frac{\timestep}{2} \Liu{B}}$ is composed of $\dot{\vt r} = {\vt 0}$, $\dot{\vt p} = {\vt F}$, $\dot{\vt q} = {\vt 0}$, and $\dot{\vt \pi} = 2 \mt C(\vt q) \vt \tau$.
+Their solutions provides us with the propagator actions described in Sec.~\ref{sec:methodology}.
+
+\subsection{Nos\'e-Hoover Chain Applied to Rigid Bodies}
+\label{sec:canonical}
+
+Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
+Sergi and Ferrario\cite{Sergi_2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.
+If vector $\vt x$ represents a point in phase space, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
+\begin{equation}
+\label{eq:eq_of_motion}
+\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
+\end{equation}
+will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric (that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$).
+A generalized Liouville operator can then be defined as\cite{Sergi_2004}
+\begin{equation*}
+{\mathcal L} = \tr{\dot{\vt x}}\diff{}{\vt x} = \tr{\left(\diff{H}{\vt x}\right)} \tr{\boldsymbol{\mathcal B}} \diff{}{\vt x}.
+\end{equation*}
+
+The system might exhibit a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative.
+According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman_2001} although the phase-space volume element $d\vt x$ is no longer conserved, there exits an invariant measure $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
+Using the generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
+\begin{equation}
+\label{eq:relation_kappa_w}
+\kappa = \dot w = {\mathcal L} w.
+\end{equation}
+
+We employ the technique of Sergi and Ferrario\cite{Sergi_2001} to couple a single Nos\'e-Hoover chain\cite{Martyna_1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
+This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom.
+To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
+The conserved ``extended Hamiltonian'' in this case is given by Eq.~\eqref{eq:nvt extended energy}.
+The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
+$\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$.
+However, the equations of motion for generalized momenta contain non-Hamiltonian terms.
+For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
+\begin{equation*}
+\begin{split}
+\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
+\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
+\end{split}
+\end{equation*}
+
+In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat.
+In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, along with a term regarding the action of a second thermostat.
+The result is
+\begin{equation*}
+{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
+\end{equation*}
+
+Each additional thermostat will then act on the preceding one, and be acted upon by the succeeding one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
+\begin{equation*}
+{\dot p}_{\eta_j} = -\diff{H}{\eta_j} + p_{\eta_{j-1}} \diff{H}{p_{\eta_{j-1}}} - p_{\eta_j} \diff{H}{p_{\eta_{j+1}}}.
+\end{equation*}
+
+Given that the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
+As a result, the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain are those in Eq.~\eqref{eq:ODE system for NVT}.
+Note that the $3N$ momentum components in Eq.~\eqref{eq:nhc_p}, the $4N$ components in Eq.~\eqref{eq:nhc_pi}, and the thermostat momenta in Eq.~\eqref{eq:nhc_p_eta} affect their own time-derivatives.
+Thus, the thermostatted system has a phase-space compressibility $\kappa = -7N \alpha_1 - \sum_j \alpha_{j+1}$.
+Resorting to Eq.~\eqref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
+\begin{equation}
+\label{eq:nhc_measure}
+w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
+\end{equation}
+
+Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables must be eliminated.
+For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman_2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
+In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
+In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna_1994}
+
+\subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}
+\label{sec:numerical_solvers}
+
+As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t {\mathcal L}}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman_2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
+In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
+As a result, the whole propagator is symplectic and thus preserves the phase-space volume element.
+It is widely recognized that this property is important for numerical stability.\cite{Skeel_1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion.
+In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator preserves the same measure.
+As the author has shown,\cite{Ezra_2006} for a particular splitting ${\mathcal L} = \sum_k {\mathcal L}^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
+Here, $\dot{\vt x}^{(k)}$ refers to the system of equations obtained from the individual operator $i\!L^{(k)}$, which might display its own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$.
+By means of the product rule, it is straightforward to devise an equivalent but more meaningful criterion, which is
+\begin{equation}
+\label{eq:split_kappa_w}
+\kappa^{(k)} = {\mathcal L}^{(k)} w.
+\end{equation}
+
+Therefore, if the function $w(\vt x)$ that derives from Eq.~\eqref{eq:relation_kappa_w} also satisfies Eq.~\eqref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$.
+We are now in a position to prove that the integration procedure described in Sec.~\ref{sec:methodology} is measure-preserving.
+As mentioned in that section, the generalized Liouville operator can be written as ${\mathcal L} = \Liu{\tiny NVE} + {\mathcal L}_\text{\tiny NHC}$, where $\Liu{\tiny NVE}$ is the same operator given by Eq.~\eqref{eq:Lioville operator} and
+\begin{equation}
+\label{eq:iL_NHC}
+\begin{split}
+{\mathcal L}_\text{\tiny NHC} = &-\alpha_1 \left( \tr{\vt p} \diff{}{\vt p} + \tr{\vt \pi} \diff{}{\vt \pi}\right) + \sum_{j=1}^{M} \alpha_j \diff{}{\eta_j} \\
+&+ \sum_{j=1}^{M} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}.
+\end{split}
+\end{equation}
+
+Then, by retaining the NVE factorization presented in Eq.~\eqref{eq:splitting NVE}, we write the full exponential operator for each time step as in Eq.~\eqref{eq:splitting NVT}.
+The Hamiltonian propagators introduced in Sec.~\ref{eq:hamiltonian dynamics} already satisfy the criterion in Eq.~\eqref{eq:split_kappa_w} with the function $w$ expressed in Eq.~\eqref{eq:nhc_measure} (they are all incompressible and keep $w$ unchanged).
+The NHC propagator is further split into factors that also satisfy Eq.~\eqref{eq:split_kappa_w}, which are given by
+\begin{equation*}
+\begin{split}
+&{\mathcal L}_0 = -\alpha_1 \left( \tr{\vt p} \diff{}{\vt p} + \tr{\vt \pi} \diff{}{\vt \pi}\right) + \alpha_1 \diff{}{\eta_1}, \\
+&{\mathcal L}_j = (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}}, \label{eq:iL_NHC_j} \\
+&{\mathcal L}_M = G_M \diff{}{p_{\eta_M}},
+\end{split}
+\end{equation*}
+with $j$ varying from $1$ to $M-1$.
+The action of the propagator $e^{t {\mathcal L}_0}$ is the solution of the equations $\dot{\vt p} = -\alpha_1 \vt p$ and $\dot{\vt \pi} = -\alpha_1 \vt \pi$ and $\dot{\eta}_1 = \alpha_1$, with all other variables remaining constant at their initial values.
+This results in the scaling operations described in Sec.~\ref{sec:methodology}.
+In turn, the propagator $e^{t {\mathcal L}_j}$ behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged.
+Finally, the propagator $e^{t {\mathcal L}_M}$ involves a unique meaningful differential equation $\dot{p}_{\eta_M} = G_M$.
+
+We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj_2005} does not fulfill this criterion.
+
+We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
+Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
+However, as will be shown in the numerical results of Sec.~\ref{sec:numerical_results},
 
@@
 \label{eq:general shadow hamiltonian}
-\shadow{\Ham{}} = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
-\end{equation*}
-
-Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\shadow{\Ham{}}$ as above.
-In the literature, $\shadow{\Ham{}}$ is usually referred to as a shadow Hamiltonian.
+\modified{\Ham{}} = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
+\end{equation*}
+
+Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\modified{\Ham{}}$ as above.
+In the literature, $\modified{\Ham{}}$ is usually referred to as a shadow Hamiltonian.
 
@@
 
-With $\nabla^2 U$ being the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, we define a shadow kinetic energy $\shadow K$ as
-\begin{equation*}
-\shadow K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \shadow{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+With $\nabla^2 U$ being the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, we define a shadow kinetic energy $\modified K$ as
+\begin{equation*}
+\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
 \end{equation*}
@@
 \begin{equation*}
-\shadow{\mathbf \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} {\mt \Omega} (\nabla^2 U) \tr{\mt \Omega}.
-\end{equation*}
-
-Although $\shadow{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
+\modified{\mathbf \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} {\mt \Omega} (\nabla^2 U) \tr{\mt \Omega}.
+\end{equation*}
+
+Although $\modified{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
 As a result, the shadow kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
 
-In addition to $\shadow K$, we also define a shadow potential energy $\shadow U$ as
-\begin{equation*}
-\shadow U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
+In addition to $\modified K$, we also define a shadow potential energy $\modified U$ as
+\begin{equation*}
+\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
 \end{equation*}
@@
 \begin{equation}
-\shadow{\Ham{}} = \shadow K + \shadow U + \mathcal{O}(h^4).
-\end{equation}
-
-\subsection{Generalized Equipartition}
-
-The important thing is that $\shadow{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\shadow K}{\vt p} \\ \diff{\shadow K}{\vt \pi} \end{array}\right] = \shadow{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
-In this way, the equipartition theorem (valid at least in this direction?
-something to do with integrating through momenta at least internally to the integral in coordinates?) states that
-\begin{equation*}
-3N k_B T = \left\langle \tr{\vt p} \diff{\tilde K}{\vt p} \right\rangle = 2\langle \tilde K \rangle
-\end{equation*}
-
-\subsection{Canonical Ensemble}
-\label{sec:canonical}
-
-Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
-Sergi and Ferrario\cite{Sergi_2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.
-If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
-\begin{equation} \label{eq:eq_of_motion}
-\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
-\end{equation}
-will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric (that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$).
-A generalized Liouville operator can then be defined as\cite{Sergi_2004}
-\[
-i\!L = \tr{\dot{\vt x}}\diff{}{\vt x} = \tr{\left(\diff{H}{\vt x}\right)} \tr{\boldsymbol{\mathcal B}} \diff{}{\vt x}.
-\]
-
-The system might exhibit a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative.
-According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman_2001} although the phase-space volume element $d\vt x$ is no longer conserved, there exits an invariant measure $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
-Using the generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
-\begin{equation}
-\label{eq:relation_kappa_w}
-\kappa = \dot w = i\!L w.
-\end{equation}
-
-We employ the technique of Sergi and Ferrario\cite{Sergi_2001} to couple a single Nos\'e-Hoover chain\cite{Martyna_1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
-This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom.
-To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
-The conserved ``extended Hamiltonian'' in this case is
-\begin{equation}
-\label{eq:H_nvt}
-H = \mathcal{H}(\vt r,\vt p,\vt q,\vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_b T\eta_1 + k_b T\sum_{j=2}^M \eta_j,
-\end{equation}
-where $\mathcal H$ is the same as in Eq.~\eqref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter of thermostat $j$.
-As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
-The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
-$\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$.
-However, the equations of motion for generalized momenta contain non-Hamiltonian terms.
-For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
-\[
-\begin{split}
-\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
-\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
-\end{split}
-\]
-
-In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat.
-In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, along with a term regarding the action of a second thermostat.
-The result is
-\[
-{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
-\]
-
-Each additional thermostat will then act on the preceding one, and be acted upon by the succeeding one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
-\[
-{\dot p}_{\eta_j} = -\diff{H}{\eta_j} + p_{\eta_{j-1}} \diff{H}{p_{\eta_{j-1}}} - p_{\eta_j} \diff{H}{p_{\eta_{j+1}}}.
-\]
-
-Given that the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
-Therefore, the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain become
-\begin{subequations}
-\label{eq:nhc_system}
-\begin{align}
-&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
-&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \label{eq:nhc_p} \\
-&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
-&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
-&\dot{\eta}_j = \alpha_j \label{eq:nhc_eta} \\
-&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}  \label{eq:nhc_p_eta}
-\end{align}
-\end{subequations}
-
-In the equations above, ${\vt \omega}_i$ $=$ $\frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
-\begin{align*}
-&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_b T \quad \text{and}\\
-&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
-\end{align*}
-
-Note that the $3N$ momentum components in Eq.~\eqref{eq:nhc_p}, the $4N$ components in Eq.~\eqref{eq:nhc_pi}, and the thermostat momenta in Eq.~\eqref{eq:nhc_p_eta} affect their own time-derivatives.
-Thus, the thermostatted system has a phase-space compressibility $\kappa = -7N \alpha_1 - \sum_j \alpha_{j+1}$.
-Resorting to Eq.~\eqref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
-\begin{equation}
-\label{eq:nhc_measure}
-w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
-\end{equation}
-
-Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables must be eliminated.
-For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman_2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
-In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
-In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna_1994}
-
-\subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}
-\label{sec:numerical_solvers}
-As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman_2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
-In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
-As a result, the whole propagator is symplectic and thus preserves the phase-space volume element.
-It is widely recognized that this property is important for numerical stability.\cite{Skeel_1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion.
-In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator preserves the same measure.
-As the author has shown,\cite{Ezra_2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
-Here, $\dot{\vt x}^{(k)}$ refers to the system of equations obtained from the individual operator $i\!L^{(k)}$, which might display its own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$.
-By means of the product rule, it is straightforward to devise an equivalent but more meaningful criterion, which is
-\begin{equation}
-\label{eq:split_kappa_w}
-\kappa^{(k)} = i\!L^{(k)} w.
-\end{equation}
-
-Therefore, if the function $w(\vt x)$ that derives from Eq.~\eqref{eq:relation_kappa_w} also satisfies Eq.~\eqref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$.
-Here, we propose and compare two distinct measure-preserving integrators for a rigid-body system coupled to a single thermostat chain.
-In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\eqref{eq:full_operator} and, as it follows from Eq.~\eqref{eq:nhc_system},
-\begin{equation}
-\label{eq:iL_NHC}
-\begin{split}
-i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
-&+ \sum_{j=1}^{M} \left[\alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}\right].
-\end{split}
-\end{equation}
-
-Then, by retaining the NVE factorization presented in Eq.~\eqref{eq:trotter_splitting_NVE}, we write the full exponential operator for each time step as
-\begin{equation}
-\label{eq:trotter_splitting_NHC}
-e^{\timestep i\!L} = e^{\frac{\timestep}{2} i\!L_\text{NHC}} e^{\frac{\timestep}{2} i\!L_b} e^{\timestep i\!L_r} e^{\timestep i\!L_t}  e^{\frac{\timestep}{2} i\!L_b} e^{\frac{\timestep}{2} i\!L_\text{NHC}}.
-\end{equation}
-
-The boost, translation, and rotation propagators already satisfy the criterion in Eq.~\eqref{eq:split_kappa_w} with the function $w$ expressed in Eq.~\eqref{eq:nhc_measure} (they are all incompressible and keep $w$ unchanged).
-The NHC propagator is further split into factors that also satisfy Eq.~\eqref{eq:split_kappa_w}, whichnare given by
-
-\begin{subequations}
-\begin{align}
-i\!L_\text{NHC}^{(0)} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \alpha_1 \diff{}{\eta_1}, \\
-i\!L_\text{NHC}^{(j)} &= (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}}, \label{eq:iL_NHC_j} \\
-i\!L_\text{NHC}^{(M)} &= G_M \diff{}{p_{\eta_M}},
-\end{align}
-\end{subequations}
-with $j$ varying from $1$ to $M-1$ in Eq.~\eqref{eq:iL_NHC_j}. Then, each propagator $e^{\frac{\timestep}{2} i\!L_\text{NHC}}$ present in Eq.~\eqref{eq:iL_NHC} is factorized as
-\[
-e^{\frac{\timestep}{2} i\!L_\text{NHC}} = \Bigg[ \prod_{j=M}^1 e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
-\]
-
-The action of the propagator $e^{t i\!L_\text{NHC}^{(0)}}$ is the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\eta}_1 = \alpha_1$, with all other variables remaining constant at their initial values.
-This system can be solved analytically, which results in
-\begin{align*}
-&\vt p_i(t) = \vt p_i^0 e^{-\alpha_1 t} \\
-&\vt \pi_i(t) = \vt \pi_i^0 e^{-\alpha_1 t} \\
-&\eta_1(t) = \eta_1^0 + \alpha_1 t
-\end{align*}
-
-In turn, the propagator $e^{t i\!L_\text{NHC}^{(j)}}$ defined in Eq.~\eqref{eq:iL_NHC_j} behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged, whose analytical solution is
-\begin{align*}
-&p_{\eta_j}(t) = p_{\eta_j}^0 + \Big( G_j - \alpha_{j+1} p_{\eta_j}^0 \Big) \phi\left(\alpha_{j+1} t\right) t \\
-&\eta_{j+1}(t) = \eta_{j+1}^0 + \alpha_{j+1} t
-\end{align*}
-where $\phi(x) = (1-e^{-x})/x$.
-We might note that, in order to avoid numerical issues with $x \rightarrow 0$, we evaluate $\phi(x)$ as $\sum_{n=0}^3 {(-1)^n x^n}/{(n+1)!}$ whenever $|x| \leq 10^{-4}$.
-The solution above is equivalent to that found in Ref.~\citenum{Martyna_1996}.
-
-Finally, the propagator $e^{t i\!L_\text{NHC}^{(M)}}$ involves a unique differential equation $\dot{p}_{\eta_M} = G_M$, whose solution is
-\[
-p_{\eta_M}(t) = p_{\eta_M}^0 + G_M t.
-\]
-
-The second splitting scheme we propose here relies on the definition of the operator
-\begin{align*}
-i\!L_b^\ast &= \sum_{i=1}^N \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} \\
-&+ \sum_{i=1}^N \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} + \alpha_1 \diff{}{\eta_1},
-\end{align*}
-with which analytical solutions can also be derived for the system of equations $\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i$ for all $i \in [1,N]$ if all other variables are constant.
-These equations are equivalent in form to the ones solved above, so we get
-\begin{align*}
-&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
-&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi\left(\alpha_1 t \right) t \\
-&\eta_1(t) = \eta_1^0 + \alpha_1 t
-\end{align*}
-
-In the new propagator, $e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(0)} }$ is removed:
-\[
-e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast} = \Bigg[ \prod_{j=M}^2 e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(1)} } \Bigg[ \prod_{j=2}^M e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
-\]
-
-We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj_2005} does not fulfill this criterion. 
-Finally, the complete scheme is
-\begin{equation}
-\label{eq:modified_splitting}
-\begin{split}
-e^{\timestep i\!L} = e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast} e^{\frac{\timestep}{2} i\!L_b^\ast} e^{\timestep i\!L_r} e^{\timestep i\!L_t}  e^{\frac{\timestep}{2} i\!L_b^\ast} e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast}.
-\end{split}
-\end{equation}
-
-We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
-Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
-However, as will be shown in the numerical results of Sec.~\ref{sec:numerical_results},
-
-\subsection{Sampling Validation}
-\label{sec:samplingvalidation}
-
-In this subsection we briefly review a simple method that serves for checking the consistency between the sampled and theoretical energy distributions.
-The procedures, implemented by Shirts,\cite{Shirts_2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
-First, recall that in the NVT ensemble the probability of observing an energy E is given by
-\[
-P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right),
-\]
-where $ \Omega(E) $ is the density of states, which does not depend on $ \beta $, and $ Q(\beta) $  is the canonical partition function, linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$.
-
-Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get
-\begin{equation}
-\frac{P(E/\beta_2)}{P(E/\beta_1)} = \exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
-\end{equation}
-Taking the logarithm of the equation above, we obtain
-\begin{equation}
-\label{eq:ln_probability_ratio}
-\ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E,
-\end{equation}
-which is the basis for the statistical analysis carried out in the ensemble validation that determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
-Note that it is independent of the unknown free energies.
-For a guidance on how to choose the temperature gap the reader is referred to the aforementioned paper.
-
-Furthermore,  it is possible to examine the consistency between the sampled kinetic energy $K$ and the corresponding Maxwell-Boltzmann distribution, which is given by the following expression:
-\begin{equation}
-\label{eq:mb}
-P(K) = \frac{\beta}{\sqrt{N_{dof}\pi}} \exp\left[-\frac{\left(\beta K - \frac{N_{dof}}{2}\right)^2}{N_{dof}} \right],
-\end{equation}
-where $N_{dof}$ is the number of degrees of freedom. Eq.~\eqref{eq:mb} is a normal distribution, with average  $ \langle K \rangle = N_{dof}/2\beta $ and variance $\sigma^2 = N_{dof}/2\beta^2$.
-
-\section{Results}
-\label{sec:numerical_results}
-
-%CÓDIGO: CHECAR FACTORIZACIÓN Y DELTA T DE LOS OPERADORES DE NHC
-
-In this section, we analyze the numerical integrators introduced in Subsection~\ref{sec:numerical_solvers}, and compare the results with those obtained employing the method of Kamberaj~\textit{et al.}\cite{Kamberaj_2005}, which is implemented in the LAMMPS\cite{Plimpton_1995} software package.
-For that, we considered a system of 900 TIP3P\cite{Jorgensen_1983} water molecules in a cubic box of length {\AA} subject to periodic boundary conditions.
-The intermolecular interactions were truncated at 9 \AA and to avoid a discontinuity in the force, we applied the so-called shifted-force (SF) method\cite{Allen_1989, Fennell_2006, Toxvaerd_2011} to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
-
-It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
-First we present some results as first mented.
-That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
-In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
-By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
-%Resultados sobre o checkensemble:
-
-checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
-Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
-m-b distribution valid when equipartition of energy is satisfied.
-
-Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
-Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.
-
-\begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
-	\label{table:new_results}
-%	\begin{ruledtabular}
-		\begin{tabular}{cccc}
-			& \multicolumn{3}{c}{ Temperature (K) } \\
-			\cline{2-4}
-			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
-			\hline
-			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
-			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
-			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
-		\end{tabular}
-%	\end{ruledtabular}
-\end{table}
-
-\begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
-	\label{table:new_results_2}
-%	\begin{ruledtabular}
-		\begin{tabular}{cccc}
-			& \multicolumn{3}{c}{Drift rate [(kcal/mol)/ns]} \\
-			\cline{2-4}
-			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
-			\hline
-			1.0 & $0.0328$ & $0.108$ & $0.023$ \\
-			2.0 & $0.597$  & $1.31$  & $17.0$ \\
-			3.0 & $3.44$   & $2.53$  & $53.2$
-		\end{tabular}
-%	\end{ruledtabular}
-\end{table}
-
-\begin{table*}
-\caption{Results.....}
-\label{table:proposed_method}
-%\begin{ruledtabular}
-\begin{tabular}{cccccccc}
-$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
-
-\hline
-\multicolumn{8}{c}{Proposed Method} \\
-1.0 & $300.01\pm0.02$ & $-8465.1\pm0.5$ & $1614.2\pm0.1$ & $807.4\pm0.1$ & $806.8\pm0.1$ & $62\pm2$ & $0.0328$ \\
-2.0 & $300.01\pm0.03$ & $-8450.8\pm0.4$ & $1614.2\pm0.2$ & $809.4\pm0.1$ & $804.7\pm0.1$ & $73\pm2$ & $0.597$ \\
-3.0 & $299.97\pm0.04$ & $-8425.7\pm0.5$ & $1614.0\pm0.2$ & $812.8\pm0.1$ & $801.2\pm0.2$ & $90\pm2$ & $3.44$ \\
-4.0 & $299.93\pm0.04$ & $-8391.4\pm0.6$ & $1613.7\pm0.2$ & $817.1\pm0.2$ & $796.6\pm0.2$ & $119\pm2$ & $5.64$ \\
-
-\hline
-\multicolumn{8}{c}{Alternative Method} \\
-1.0 & $300.00\pm0.02$ & $-8465.5\pm0.5$ & $1614.1\pm0.1$ & $807.1\pm0.1$ & $807.0\pm0.1$ & $65\pm2$ & $0.108$ \\
-2.0 & $299.99\pm0.03$ & $-8450.5\pm0.5$ & $1614.1\pm0.2$ & $809.3\pm0.1$ & $804.7\pm0.1$ & $72\pm2$ & $1.31$ \\
-3.0 & $299.98\pm0.04$ & $-8425.9\pm0.5$ & $1614.0\pm0.2$ & $812.5\pm0.1$ & $801.5\pm0.2$ & $88\pm2$ & $2.53$ \\
-4.0 & $300.01\pm0.04$ & $-8390.4\pm0.6$ & $1614.2\pm0.2$ & $817.4\pm0.2$ & $796.8\pm0.2$ & $120\pm2$ & $7.31$ \\
-
-\hline
-\multicolumn{8}{c}{Method of Kamberaj \textit{et al}.\cite{Kamberaj_2005}} \\
-1.0 & $299.53\pm0.02$ & $-8470.9\pm0.5$ & $1611.6\pm0.1$ & $806.4\pm0.1$ & $805.2\pm0.1$ & $52\pm2$ & $0.023$ \\
-2.0 & $298.24\pm0.03$ & $-8471.2\pm0.5$ & $1604.6\pm0.2$ & $805.4\pm0.1$ & $799.2\pm0.1$ & $47\pm2$ & $17.0$ \\
-3.0 & $295.94\pm0.04$ & $-8480.9\pm0.6$ & $1592.3\pm0.2$ & $804.2\pm0.1$ & $788.1\pm0.1$ & $28\pm2$ & $53.2$ \\
-4.0 & $292.72\pm0.04$ & $-8494.7\pm0.7$ & $1575.0\pm0.2$ & $802.4\pm0.1$ & $772.5\pm0.2$ & $-1\pm2$ & $60.6$ \\
-
-\hline
-\multicolumn{8}{c}{Double-Thermostat Analogue of the Proposed Method} \\
-1.0 & $300.00\pm0.02$ & $-8466.0\pm0.5$ & $1614.1\pm0.1$ & $806.6\pm0.1$ & $807.5\pm0.1$ & $60\pm2$ & $6.42$\\
-2.0 & $299.98\pm0.03$ & $-8452.0\pm0.5$ & $1614.0\pm0.2$ & $807.1\pm0.1$ & $806.9\pm0.1$ & $71\pm2$ & $80.4$ \\
-3.0 & $300.16\pm0.04$ & $-8426.1\pm0.5$ & $1615.0\pm0.2$ & $808.4\pm0.1$ & $806.6\pm0.1$ & $90\pm2$ & $1.38\times10^4$ \\
-4.0 & $300.75\pm0.04$ & $-8385.3\pm0.6$ & $1618.2\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $125\pm2$ & $4.29\times10^4$ \\
-
-\hline
-\multicolumn{8}{c}{Double-Thermostat Analogue of the Alternative Method} \\
-1.0 & $299.98\pm0.02$ & $-8465.6\pm0.5$ & $1614.0\pm0.1$ & $806.6\pm0.1$ & $807.4\pm0.1$ & $56\pm2$ & $4.87$ \\
-2.0 & $299.96\pm0.03$ & $-8452.0\pm0.5$ & $1613.9\pm0.2$ & $807.1\pm0.1$ & $806.8\pm0.1$ & $71\pm2$ & $85.1$ \\
-3.0 & $300.20\pm0.04$ & $-8426.0\pm0.5$ & $1615.2\pm0.2$ & $808.5\pm0.1$ & $806.7\pm0.1$ & $89\pm2$ & $1.50\times10^4$ \\
-4.0 & $300.75\pm0.04$ & $-8384.7\pm0.6$ & $1618.1\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $123\pm2$ & $4.33\times10^4$ \\
-
-\hline
-\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}} \\
-1.0 & $299.52\pm0.02$ & $-8470.2\pm0.5$ & $1611.6\pm0.1$ & $805.9\pm0.1$ & $805.6\pm0.1$ & $58\pm2$ & $0.272$ \\
-2.0 & $298.18\pm0.03$ & $-8471.8\pm0.5$ & $1604.3\pm0.2$ & $804.4\pm0.1$ & $799.9\pm0.1$ & $44\pm2$ & $0.390$ \\
-3.0 & $295.97\pm0.04$ & $-8474.6\pm0.5$ & $1592.4\pm0.2$ & $801.6\pm0.1$ & $790.8\pm0.2$ & $29\pm2$ & $3.00$ \\
-4.0 & $292.65\pm0.04$ & $-8478.7\pm0.6$ & $1574.6\pm0.2$ & $797.6\pm0.2$ & $777.0\pm0.2$ & $10\pm2$ & $8.12$ \\
-
-\end{tabular}
-%\end{ruledtabular}
-\end{table*}
-
-\begin{figure}
-    \caption{SHAKE: $\timestep$ = 1 fs ($\bar{T}$ = 299.98 K, $\bar{U}$ = -8428.86 kcal/mol) HMC: $\bar{U}$ = -8470.71 kcal/mol  }
-	\includegraphics{potenergy}
-\end{figure}
-
-\begin{figure}
-    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 300.04 K, $\bar{U}$ = -8465.63 kcal/mol)}
-	\includegraphics{potenergy2}
-\end{figure}
-
-\begin{figure}
-    \caption{$\timestep$ = 0.5 fs ($\bar{T}$ = 299.99 K, $\bar{U}$ = -8468.19 kcal/mol)}
-	\includegraphics{potenergy_05}
-\end{figure}
-
-\begin{figure}
-    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 299.52 K, $\bar{U}$ = -8470.21 kcal/mol)}
-	\includegraphics{potenergy_kamb}
-\end{figure}
-
-In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
-In figure legends we show the averages temperatures simulated.
-It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.
-
-\begin{figure}
-	\includegraphics{checkensemble}
-	\caption{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
-Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
-	\label{fig:checkensemble}
-\end{figure}
-
-\begin{figure}
-	\includegraphics{maxwell-botlzamm-paper}
-	\caption{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\eqref{eq:mb}.
-Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}$ = 304.99~K.}
-\label{fig:mbdistribution}
-\end{figure}
-
-Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
-It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
-This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
-
-\begin{table}
-	\begin{threeparttable}
-		\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach \tnote{a} \tnote{b}}
-		\label{table:ensemblevalidation}
-%		\begin{ruledtabular}
-			\begin{tabular}{ccccc}
-				& \multicolumn{4}{c}{true $\timestep$ = 8.970 K} \\
-				\cline{2-5}
-				& \multicolumn{2}{c}{potential} & \multicolumn{2}{c}{kinetic}\\
-				\hline
-				Method  &$\timestep/K$ & $\sigma$ dev & $\timestep/K$ & $\sigma$ dev \\
-				\hline % inserts single-line 
-				Eq.~\eqref{eq:trotter_splitting_NHC} & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
-				Eq.~\eqref{eq:modified_splitting}    & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14
-		\end{tabular}
-%		\end{ruledtabular}
-		\begin{tablenotes}
-			\item[a] The standard deviation of the mean temperature for all cases is $\sigma$ = 0.03.
-			\item[b] $\bar{T}_1$ = 296.03~K and $\bar{T}_2$ = 304.99~K.
-		\end{tablenotes}
-	\end{threeparttable}
-\end{table}
-
-%\begin{center}
-%  \label{fig:pressure}
-%  \includegraphics{pressure2}
-%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars). Simulation of liquid water with NVE integration. $\timestep : 1 fs$  }
-%\end{center}
-
-\section{Concluding Remarks}
-\label{sec:conclusion}
-Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. 
-by contrast
-
-\begin{acknowledgement}
-	The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
-\end{acknowledgement}
-
-\appendix
-
-\section{Shadow Hamiltonian of Rigid Bodies}
-\label{sec:rigid body shadow hamiltonian}
+\modified{\Ham{}} = \modified K + \modified U + \mathcal{O}(h^4).
+\end{equation}
 
@@
 
+\subsection{Generalized Equipartition}
+
+The important thing is that $\modified{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] = \modified{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
+In this way, the equipartition theorem (valid at least in this direction?
+something to do with integrating through momenta at least internally to the integral in coordinates?) states that
+\begin{equation*}
+3N k_B T = \left\langle \tr{\vt p} \diff{\tilde K}{\vt p} \right\rangle = 2\langle \tilde K \rangle
+\end{equation*}
+
+\subsection{Sampling Validation}
+\label{sec:samplingvalidation}
+
+In this subsection we briefly review a simple method that serves for checking the consistency between the sampled and theoretical energy distributions.
+The procedures, implemented by Shirts,\cite{Shirts_2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
+First, recall that in the NVT ensemble the probability of observing an energy E is given by
+\begin{equation*}
+P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right),
+\end{equation*}
+where $ \Omega(E) $ is the density of states, which does not depend on $ \beta $, and $ Q(\beta) $  is the canonical partition function, linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$.
+
+Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get
+\begin{equation}
+\frac{P(E/\beta_2)}{P(E/\beta_1)} = \exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
+\end{equation}
+Taking the logarithm of the equation above, we obtain
+\begin{equation}
+\label{eq:ln_probability_ratio}
+\ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E,
+\end{equation}
+which is the basis for the statistical analysis carried out in the ensemble validation that determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
+Note that it is independent of the unknown free energies.
+For a guidance on how to choose the temperature gap the reader is referred to the aforementioned paper.
+
+Furthermore,  it is possible to examine the consistency between the sampled kinetic energy $K$ and the corresponding Maxwell-Boltzmann distribution, which is given by the following expression:
+\begin{equation}
+\label{eq:mb}
+P(K) = \frac{\beta}{\sqrt{N_{dof}\pi}} \exp\left[-\frac{\left(\beta K - \frac{N_{dof}}{2}\right)^2}{N_{dof}} \right],
+\end{equation}
+where $N_{dof}$ is the number of degrees of freedom. Eq.~\eqref{eq:mb} is a normal distribution, with average  $ \langle K \rangle = N_{dof}/2\beta $ and variance $\sigma^2 = N_{dof}/2\beta^2$.
+
+\section{Results}
+\label{sec:numerical_results}
+
+%CÓDIGO: CHECAR FACTORIZACIÓN Y DELTA T DE LOS OPERADORES DE NHC
+
+In this section, we analyze the numerical integrators introduced in Subsection~\ref{sec:numerical_solvers}, and compare the results with those obtained employing the method of Kamberaj~\textit{et al.}\cite{Kamberaj_2005}, which is implemented in the LAMMPS\cite{Plimpton_1995} software package.
+For that, we considered a system of 900 TIP3P\cite{Jorgensen_1983} water molecules in a cubic box of length {\AA} subject to periodic boundary conditions.
+The intermolecular interactions were truncated at 9 \AA and to avoid a discontinuity in the force, we applied the so-called shifted-force (SF) method\cite{Allen_1989, Fennell_2006, Toxvaerd_2011} to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
+
+It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
+First we present some results as first mented.
+That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
+In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
+By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
+%Resultados sobre o checkensemble:
+
+checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
+Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
+m-b distribution valid when equipartition of energy is satisfied.
+
+Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
+Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.
+
+\begin{table}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
+	\label{table:new_results}
+%	\begin{ruledtabular}
+		\begin{tabular}{cccc}
+			& \multicolumn{3}{c}{ Temperature (K) } \\
+			\cline{2-4}
+			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			\hline
+			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
+			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
+			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
+		\end{tabular}
+%	\end{ruledtabular}
+\end{table}
+
+\begin{table}
+	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
+	\label{table:new_results_2}
+%	\begin{ruledtabular}
+		\begin{tabular}{cccc}
+			& \multicolumn{3}{c}{Drift rate [(kcal/mol)/ns]} \\
+			\cline{2-4}
+			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			\hline
+			1.0 & $0.0328$ & $0.108$ & $0.023$ \\
+			2.0 & $0.597$  & $1.31$  & $17.0$ \\
+			3.0 & $3.44$   & $2.53$  & $53.2$
+		\end{tabular}
+%	\end{ruledtabular}
+\end{table}
+
+\begin{table*}
+\caption{Results.....}
+\label{table:proposed_method}
+%\begin{ruledtabular}
+\begin{tabular}{cccccccc}
+$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
+
+\hline
+\multicolumn{8}{c}{Proposed Method} \\
+1.0 & $300.01\pm0.02$ & $-8465.1\pm0.5$ & $1614.2\pm0.1$ & $807.4\pm0.1$ & $806.8\pm0.1$ & $62\pm2$ & $0.0328$ \\
+2.0 & $300.01\pm0.03$ & $-8450.8\pm0.4$ & $1614.2\pm0.2$ & $809.4\pm0.1$ & $804.7\pm0.1$ & $73\pm2$ & $0.597$ \\
+3.0 & $299.97\pm0.04$ & $-8425.7\pm0.5$ & $1614.0\pm0.2$ & $812.8\pm0.1$ & $801.2\pm0.2$ & $90\pm2$ & $3.44$ \\
+4.0 & $299.93\pm0.04$ & $-8391.4\pm0.6$ & $1613.7\pm0.2$ & $817.1\pm0.2$ & $796.6\pm0.2$ & $119\pm2$ & $5.64$ \\
+
+\hline
+\multicolumn{8}{c}{Alternative Method} \\
+1.0 & $300.00\pm0.02$ & $-8465.5\pm0.5$ & $1614.1\pm0.1$ & $807.1\pm0.1$ & $807.0\pm0.1$ & $65\pm2$ & $0.108$ \\
+2.0 & $299.99\pm0.03$ & $-8450.5\pm0.5$ & $1614.1\pm0.2$ & $809.3\pm0.1$ & $804.7\pm0.1$ & $72\pm2$ & $1.31$ \\
+3.0 & $299.98\pm0.04$ & $-8425.9\pm0.5$ & $1614.0\pm0.2$ & $812.5\pm0.1$ & $801.5\pm0.2$ & $88\pm2$ & $2.53$ \\
+4.0 & $300.01\pm0.04$ & $-8390.4\pm0.6$ & $1614.2\pm0.2$ & $817.4\pm0.2$ & $796.8\pm0.2$ & $120\pm2$ & $7.31$ \\
+
+\hline
+\multicolumn{8}{c}{Method of Kamberaj \textit{et al}.\cite{Kamberaj_2005}} \\
+1.0 & $299.53\pm0.02$ & $-8470.9\pm0.5$ & $1611.6\pm0.1$ & $806.4\pm0.1$ & $805.2\pm0.1$ & $52\pm2$ & $0.023$ \\
+2.0 & $298.24\pm0.03$ & $-8471.2\pm0.5$ & $1604.6\pm0.2$ & $805.4\pm0.1$ & $799.2\pm0.1$ & $47\pm2$ & $17.0$ \\
+3.0 & $295.94\pm0.04$ & $-8480.9\pm0.6$ & $1592.3\pm0.2$ & $804.2\pm0.1$ & $788.1\pm0.1$ & $28\pm2$ & $53.2$ \\
+4.0 & $292.72\pm0.04$ & $-8494.7\pm0.7$ & $1575.0\pm0.2$ & $802.4\pm0.1$ & $772.5\pm0.2$ & $-1\pm2$ & $60.6$ \\
+
+\hline
+\multicolumn{8}{c}{Double-Thermostat Analogue of the Proposed Method} \\
+1.0 & $300.00\pm0.02$ & $-8466.0\pm0.5$ & $1614.1\pm0.1$ & $806.6\pm0.1$ & $807.5\pm0.1$ & $60\pm2$ & $6.42$\\
+2.0 & $299.98\pm0.03$ & $-8452.0\pm0.5$ & $1614.0\pm0.2$ & $807.1\pm0.1$ & $806.9\pm0.1$ & $71\pm2$ & $80.4$ \\
+3.0 & $300.16\pm0.04$ & $-8426.1\pm0.5$ & $1615.0\pm0.2$ & $808.4\pm0.1$ & $806.6\pm0.1$ & $90\pm2$ & $1.38\times10^4$ \\
+4.0 & $300.75\pm0.04$ & $-8385.3\pm0.6$ & $1618.2\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $125\pm2$ & $4.29\times10^4$ \\
+
+\hline
+\multicolumn{8}{c}{Double-Thermostat Analogue of the Alternative Method} \\
+1.0 & $299.98\pm0.02$ & $-8465.6\pm0.5$ & $1614.0\pm0.1$ & $806.6\pm0.1$ & $807.4\pm0.1$ & $56\pm2$ & $4.87$ \\
+2.0 & $299.96\pm0.03$ & $-8452.0\pm0.5$ & $1613.9\pm0.2$ & $807.1\pm0.1$ & $806.8\pm0.1$ & $71\pm2$ & $85.1$ \\
+3.0 & $300.20\pm0.04$ & $-8426.0\pm0.5$ & $1615.2\pm0.2$ & $808.5\pm0.1$ & $806.7\pm0.1$ & $89\pm2$ & $1.50\times10^4$ \\
+4.0 & $300.75\pm0.04$ & $-8384.7\pm0.6$ & $1618.1\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $123\pm2$ & $4.33\times10^4$ \\
+
+\hline
+\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}} \\
+1.0 & $299.52\pm0.02$ & $-8470.2\pm0.5$ & $1611.6\pm0.1$ & $805.9\pm0.1$ & $805.6\pm0.1$ & $58\pm2$ & $0.272$ \\
+2.0 & $298.18\pm0.03$ & $-8471.8\pm0.5$ & $1604.3\pm0.2$ & $804.4\pm0.1$ & $799.9\pm0.1$ & $44\pm2$ & $0.390$ \\
+3.0 & $295.97\pm0.04$ & $-8474.6\pm0.5$ & $1592.4\pm0.2$ & $801.6\pm0.1$ & $790.8\pm0.2$ & $29\pm2$ & $3.00$ \\
+4.0 & $292.65\pm0.04$ & $-8478.7\pm0.6$ & $1574.6\pm0.2$ & $797.6\pm0.2$ & $777.0\pm0.2$ & $10\pm2$ & $8.12$ \\
+
+\end{tabular}
+%\end{ruledtabular}
+\end{table*}
+
+\begin{figure}
+    \caption{SHAKE: $\timestep$ = 1 fs ($\bar{T}$ = 299.98 K, $\bar{U}$ = -8428.86 kcal/mol) HMC: $\bar{U}$ = -8470.71 kcal/mol  }
+	\includegraphics{potenergy}
+\end{figure}
+
+\begin{figure}
+    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 300.04 K, $\bar{U}$ = -8465.63 kcal/mol)}
+	\includegraphics{potenergy2}
+\end{figure}
+
+\begin{figure}
+    \caption{$\timestep$ = 0.5 fs ($\bar{T}$ = 299.99 K, $\bar{U}$ = -8468.19 kcal/mol)}
+	\includegraphics{potenergy_05}
+\end{figure}
+
+\begin{figure}
+    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 299.52 K, $\bar{U}$ = -8470.21 kcal/mol)}
+	\includegraphics{potenergy_kamb}
+\end{figure}
+
+In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
+In figure legends we show the averages temperatures simulated.
+It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
+Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.
+
+\begin{figure}
+	\includegraphics{checkensemble}
+	\caption{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
+Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
+	\label{fig:checkensemble}
+\end{figure}
+
+\begin{figure}
+	\includegraphics{maxwell-botlzamm-paper}
+	\caption{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\eqref{eq:mb}.
+Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}$ = 304.99~K.}
+\label{fig:mbdistribution}
+\end{figure}
+
+Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
+It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
+This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
+
+\begin{table}
+	\begin{threeparttable}
+		\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach \tnote{a} \tnote{b}}
+		\label{table:ensemblevalidation}
+%		\begin{ruledtabular}
+			\begin{tabular}{ccccc}
+				& \multicolumn{4}{c}{true $\timestep$ = 8.970 K} \\
+				\cline{2-5}
+				& \multicolumn{2}{c}{potential} & \multicolumn{2}{c}{kinetic}\\
+				\hline
+				Method  &$\timestep/K$ & $\sigma$ dev & $\timestep/K$ & $\sigma$ dev \\
+				\hline % inserts single-line 
+				Eq.~\eqref{eq:trotter_splitting_NHC} & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
+				Eq.~\eqref{eq:modified_splitting}    & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14
+		\end{tabular}
+%		\end{ruledtabular}
+		\begin{tablenotes}
+			\item[a] The standard deviation of the mean temperature for all cases is $\sigma$ = 0.03.
+			\item[b] $\bar{T}_1$ = 296.03~K and $\bar{T}_2$ = 304.99~K.
+		\end{tablenotes}
+	\end{threeparttable}
+\end{table}
+
+%\begin{center}
+%  \label{fig:pressure}
+%  \includegraphics{pressure2}
+%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars). Simulation of liquid water with NVE integration. $\timestep : 1 fs$  }
+%\end{center}
+
+\section{Concluding Remarks}
+\label{sec:conclusion}
+Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$. 
+by contrast
+
+\begin{acknowledgement}
+	The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
+\end{acknowledgement}
+
+\appendix
+
 \bibliography{rigid_bodies}
@@
 %			\begin{tabular}{ccccc}
-%				& \multicolumn{2}{c}{Eq.~\ref{eq:trotter_splitting_NHC}} & \multicolumn{2}{c}{Eq.~%\ref{eq:modified_splitting}} \\
+%				& \multicolumn{2}{c}{Eq.~\eqref{eq:trotter_splitting_NHC}} & \multicolumn{2}{c}{Eq.~%\eqref{eq:modified_splitting}} \\
 %				\cline{2-5}
@@
 %The Nos\'e-Hoover chain propagator is further split by making $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_%\text{NHC}^{(j)}$ and, afterwards,
-%\[
+%\begin{equation*}
 %e^{\frac{\timestep}{2} i\!L_\text{NHC}} = \Pi_{j=1}^n \Bigg[ \prod_{j=M}^1 e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
-%\]
+%\end{equation*}
 
@@
 %By evaluating these functions simultaneously for all $i$, we accomplish the effect of a propagator $e^{t i%\!L_b^\ast}$, where $i\!L_b^\ast$ is the Liouville operator given by
-%\[
+%\begin{equation*}
 %i\!L_b^\ast = \sum_{i=1}^N \left[ \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} \right].
-%\]
+%\end{equation*}
 %We also define a modified Nos\'e-Hoover chain propagator $e^{i\!L^\ast_\text{NHC}}$, as follows:

### 2018-03-18 rigid_bodies.tex
Rearrangement of further development of proposed method section

@@
-%\documentclass[journal=jctcce,manuscript=article]{achemso}
-\documentclass[journal=jctcce,manuscript=article,layout=twocolumn]{achemso}
-
-% NOTE: comment bellow before submitting
+\documentclass[
+journal=jctcce,
+layout=twocolumn
+]{achemso}
+
+\setkeys{acs}{
+	abbreviations = false,
+	articletitle  = false,
+	keywords      = false,
+	maxauthors    = 10,
+	super         = true
+}
+
+% Comment below before submitting
 \let\titlefont\undefined
-\makeatletter
-\let\l@addto@macro\relax
-\makeatother
 \usepackage[fontsize=11pt]{scrextend}
-% NOTE: comment above before submitting
+\usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
+%\flushbottom
+% Comment above before submitting
 
@@
 %\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}
-\title{Nos\'e-Hoover Chains Improved by Backward Error Analysis: Application to Rigid-Body Molecular Dynamics}
+\title{Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis}
 
@@
 
-In a previous paper \cite{Silveira_2017}, we employed a particular factorization of the orientation matrix expressed in terms of quaternion components to 1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for rigid bodies in the NVE ensemble and 2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} as a sequence of uniform rotations around the principal axes of a body, thus unveiling the equivalence between this method and an older one by \citeauthor{Dullweber_1997} \cite{Dullweber_1997}
-
-The ordinary differential equation (ODE) system that describes a Hamiltonian rigid-body motion are \cite{Silveira_2017}
+In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to 1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and 2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} as a sequence of uniform rotations around the principal axes of a body, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.
+
+The Hamiltonian system of ordinary differential equations (ODE) that describes a rigid-body motion are \cite{Silveira_2017}
 \begin{subequations}
@@
 with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$.
-In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines the body's orientation, and $\vt \pi$ is the momentum conjugated to such quaternion.
+In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugated momentum of such quaternion.
 Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity.
-Matrices $\mt M$ and $\mt I$ are diagonal ones, with the former containing the body's mass in all its diagonal entries and the latter containing the principal moments of inertia.
+Matrices $\mt M$ and $\mt I$ are diagonal ones.
+The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
 Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
@@
 
-The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the principal-axis frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the principal-axis and space-fixed frames) expressed as a function of the quaternion components.
-The fact that such matrix admits the simple factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ played a central role in Ref.~\citenum{Silveira_2017}.
+The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the body-fixed frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components.
+The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$.
 
@@
 \end{equation}
-where $K$ and $U$ are the body's kinetic and potential energies, respectively. While $U(\vt r, \vt q)$ can have any functional form, the kinetic energy is given by
+where $K$ and $U$ are the kinetic and potential energies of the body, respectively. While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
 \begin{equation}
@@
 
-An advantage of this notation is that Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the respective properties of all bodies piled together.
+An advantage of this notation is that Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
 The same occurs for diagonal matrices $\mt M$ and $\mt I$, whose sizes increase to $3N \times 3N$ in this case.
 In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
-Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged to accommodate a number of free atoms.
-
-A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system's configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$.
+Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further to accommodate a number of freely moving atoms.
+
+A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$.
 A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
@@
 where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
-In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, where the asterisk denotes the state of the system immediately before a propagation takes place.
-The action of $e^{\timestep \Liu{A}}$ is a uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) with simultaneous torque-free rotation during a full time step.
+In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state immediately before the propagation occurs.
+The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step.
 
@@
 
-In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where each $Q_j$ is an inertial parameter of the corresponding thermostat.
-Besides, each $G_j$ is a generalized force acting on thermostat $j$, defined as
+In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where $Q_j$ is an inertial parameter for the corresponding thermostat.
+Besides, $G_j$ is a generalized force acting on thermostat $j$, defined as
 \begin{subequations}
@@
 \end{subequations}
-where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set as null, as usual (otherwise, $L = 6N$).\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
-%Finally, $\modified K$ is the total kinetic energy of the system, whose computation (not using Eq.~\eqref{eq:kinetic energy}, hence the distinct symbol) is the central feature of the method we propose.
-The ``extended energy'' which is conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
+where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set to zero at the onset or $L = 6N$ otherwise.\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, we make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.
+The ``extended energy'' conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
 \begin{equation}
@@
 
-In our scheme, the first propagator to act, $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
-Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied sequentially with $j$ ranging from $M-1$ down to $1$.
+In the scheme above, the first propagator to act is $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, which promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
+Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ act sequentially, with $j$ varying from $M-1$ down to $1$.
 With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as
@@
 
-After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ enacts the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of the thermostat coordinate $\eta_1$.
+After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$.
 Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
@@
 In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\modified K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
-Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_1}$ to $e^{\frac{\timestep}{4n} \mathcal{L}_M}$ are applied again, but now in the increasing order of their indices.
-It is worth noting that the $\eta$'s are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H$.
-
-The whole procedure described above is repeated $n$ times.
-Because of its low computational cost, the compensation required for increasing the time step $\timestep$ without degrading the accuracy of the NHC part (i.e.
-employing a larger $n$) causes a small impact in the overall integration effort.
-In contrast, evaluating the NVE part is expensive for involving force/torque computations and its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
-A high-order integration scheme\cite{Omelyan_2007, vanZon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
+Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied again, but now with $j$ varying from $1$ up to $M$.
+It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H$.
+
+For having a low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort, making it feasible to increase the time step $\timestep$ without degrading the accuracy of the NHC part.
+In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
+A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
 The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
-Nevertheless, we choose to take a simpler path.
-Instead of attempting to improve the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we try to quantify the obtained deviations and then use them to tune the thermostatting procedure and properly weight the sampled configurations when canonical averages are computed.
-
-A known property of splitting methods applied to Hamiltonian ODE systems is that they provide Hamiltonian approximate solutions.
-In other words, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian.
-Thus, whereas $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors).
-It is also known that $\Ham{S}$ depends on the time step size $\timestep$ and that, for a time-reversible splitting such as Eq.~\eqref{eq:splitting NVE}, one can write $\Ham{S} = \Ham{} + \mathcal{O}(\timestep^2)$.
-Estimating the trailing term is a task of backward error analysis.
-
-As shown in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be better approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are named modified kinetic and modified potential energies, respectively.
-The former can be expressed as
-\begin{equation}
-\label{eq:shadow kinetic energy}
+Here we choose to take a simpler path.
+Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied deviations, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when computing ensemble averages.
+
+In order to explain our proposal, we turn the attention again to the NVE case.
+A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
+Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian.
+In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors).
+It is also known that $\Ham{S}$ depends on the time step size $\timestep$.
+As shown in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are modified kinetic and potential energies.
+The latter is given by
+\begin{equation}
+\label{eq:modified potential energy}
+\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right),
+\end{equation}
+where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the resultant torque vector.
+Note that $\modified U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
+In turn, the modified kinetic energy can be expressed as
+\begin{equation}
+\label{eq:modified kinetic energy}
 \modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
 \end{equation}
-where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$. The modified potential, which is also independent of $\vt p$ and $\vt \pi$, is given by
-\begin{equation}
-\label{eq:shadow potential energy}
-\modified U = U - \frac{\timestep^2}{24} \left[ \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right],
-\end{equation}
-where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the torques.
-Therefore, the outcome of solving Eq.~\eqref{eq:ODE system for NVE} using a Trotter-Suzuki splitting will be close to the exact solution of Eq.~\eqref{eq:ODE system for NVE}, but it will be even closer to the solution of
+where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$.
+Therefore, a Trotter-Suzuki trajectory, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
+\begin{equation*}
+\dot{\vt r} = \diff{\modified K}{\vt p}, \;
+\dot{\vt q} = \diff{\modified K}{\vt \pi}, \;
+\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}, \; \text{and} \;
+\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q},
+\end{equation*}
+where $\modified{\Ham{}} = \modified K + \modified U$ is a modified Hamiltonian.
+The first two of these equations can be combined together so that we can use Eq.~\eqref{eq:modified kinetic energy} to obtain
 \begin{equation}
 \label{eq:shadow ODE system}
-\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] =
-\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] =
-\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation}
-together with $\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}$ and $\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}$, where $\modified{\Ham{}} = \modified K + \modified U$.
-Our proposal consists in acknowledging this fact and then tuning the Nos\'e-Hoover chain to provide a canonical distribution consistent with $\modified{\Ham{}}$ (instead of $\Ham{}$) at temperature $T$.
+\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
+\end{equation}
+
+Analytical calculation of $\modified U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
+In contrast, although it is possible to compute $\modified K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
+Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained Trotter-Suzuki trajectory.
+Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
+\begin{equation}
+\label{eq:modified kinetic energy estimator}
+\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
+\end{equation}
+
+For reasons that will become clear shortly, we employ an asymmetric, four-point stencil formula for estimating $\dot{\vt r}$ at a given instant $t$, which is
+\begin{equation*}
+\dot{\vt r}^{[3]} = \frac{{\vt r}_{t-2\timestep} - 6 {\vt r}_{t-\timestep} + 3 {\vt r}_t + 2 {\vt r}_{t+\timestep}}{6\timestep}.
+\end{equation*}
+
+In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each rigid body $i$, such as required for satisfying the unit-norm condition \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of\cite{Schay_1995}
+\begin{equation*}
+\dot{\vt q}^{[3]} = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
+\end{equation*}
+where ${\mt \Pi}$ is a matrix-valued function given by either product ${\mt B}\tr{\mt B}$ or ${\mt C}\tr{\mt C}$ \cite{Silveira_2017}.
+In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.
+
+Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
+Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \modified{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
 Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly as described before.
-The only requirement is to replace $K$ by $\modified K$ in the definition of the generalized force in thermostat $1$, given by Eq.~\eqref{eq:generalized force NVT}.
-By contrasting this one with Eqs.~\eqref{eq:shadow kinetic energy} and \eqref{eq:shadow ODE system}, we observe that such force becomes
+The only required modification is that $\modified K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
+By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that such force becomes
 \begin{equation}
@@
 
-It is possible to evaluate Eq.~\eqref{eq:shadow kinetic energy} directly for computing $\modified K$, but this is quite complex and expensive.
-Instead, one can do it very easily by means of a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the derivatives in the left-hand side of Eq.~\eqref{eq:shadow ODE system}.
-By denoting the obtained estimators as $\dot{\vt r}_\text{\tiny NVE}$ and $\dot{\vt q}_\text{\tiny NVE}$, and substituting Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:shadow kinetic energy}, we find out that
-\begin{equation}
-\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}_\text{\tiny NVE} + \tr{\vt \pi} \dot{\vt q}_\text{\tiny NVE} \right).
-\end{equation}
-
-- Pratical estimation of numerical derivatives: virtual steps and interpolation
-
-- Scaling of $\modified K$ during the NHC integration.
-
-Reweighting:
+One instant in which $\modified K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
+Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
+Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
+This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
+Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
+Fortunately, only positions and orientations are required for the numerical differentiation.
+This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time.
+By incomplete we mean that the step is aborted once the positions and orientations have been updated.
+A drawback of this approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
+As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
+In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.
+
+In addition, $\modified K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
+Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\modified K$ goes as simply as ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$.
+
+Finally, reweighting:
 \begin{equation}
@@
 
-With $\nabla^2 U$ being the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, we define a shadow kinetic energy $\modified K$ as
+With $\nabla^2 U$ being the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, we define a modified kinetic energy $\modified K$ as
 \begin{equation*}
@@
 Although $\modified{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
-As a result, the shadow kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
-
-In addition to $\modified K$, we also define a shadow potential energy $\modified U$ as
+As a result, the modified kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
+
+In addition to $\modified K$, we also define a modified potential energy $\modified U$ as
 \begin{equation*}

### 2018-03-18 rigid_bodies.tex
Preparation for a short version

@@
 
-\section{Proposed Method}
+\section{Proposed Methodology}
 \label{sec:methodology}
+
+\subsection{NVE Dynamics: Notation and Symplectic Integration}
 
@@
 \end{smallarray}
+%
 \; \text{and} \;
+%
 \mt C(\vt q) = \begin{smallarray}{rrrr}
@@
 
-Description of analytical and NO\_SQUISH rotations...
+Brief description of analytical and NO-SQUISH rotations...
+
+\subsection{NVT Dynamics with a Single Thermostat Chain}
 
@@
 To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$.
-The equations of motion, in this case, are
+The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
+\begin{equation}
+\label{eq:nvt extended energy}
+H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
+\end{equation}
+where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
+As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.
+
+Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010_2} Invaluable tools have been devised for the development and analysis of extended phase space methods.
+By employing the method of Sergi and Ferrario \cite{Sergi_2001}, we obtain the equations of motion for the NVT dynamics with a single thermostat chain, which are
 \begin{subequations}
@@
 	&\dot{\vt q} =
-	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \\
+	\frac{1}{2} \mt B(\vt q) \vt \omega, \\
 %
@@
 \label{eq:nhc_p_eta}
-	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M,
+	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M.
 	\end{align}
@@
 
-In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where $Q_j$ is an inertial parameter for the corresponding thermostat.
-Besides, $G_j$ is a generalized force acting on thermostat $j$, defined as
+In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$. Besides, $G_j$ is a generalized force acting on thermostat $j$, defined as
 \begin{subequations}
@@
 &G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
-&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1,
+&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
 \end{align}
 \end{subequations}
-where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set to zero at the onset or $L = 6N$ otherwise.\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, we make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.
-The ``extended energy'' conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
-\begin{equation}
-\label{eq:nvt extended energy}
-H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
-\end{equation}
-
-As in the previous case, it is possible to devise a splitting method to solve Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
+
+With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
+It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but are often computed for checking how well a numerical integrator conserves $H$.
+Analysis of the probability distribution generated by Eq.~\eqref{eq:ODE system for NVT} can be done through the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables eliminated.
+For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates.
+Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
+In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if one makes $L = 6N$.
+In the particular situation in which one sets $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of a simulation, one must make $L = 6N - 3$ \cite{Martyna_1994}.
+
+\subsection{Measure-Preserving Integration for the NVT Dynamics}
+
+As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
+Ezra \cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion.
+In short, a splitting scheme will preserve a given measure if each individual propagator does it as well.
 By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
@@
 Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
-%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\modified K$ to be scaled by the square of it, hence ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$.
-In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\modified K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
 Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied again, but now with $j$ varying from $1$ up to $M$.
-It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H$.
 
@@
 Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied deviations, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when computing ensemble averages.
+
+\subsection{Backward Error Analysis}
 
@@
 \begin{equation*}
-\dot{\vt r}^{[3]} = \frac{{\vt r}_{t-2\timestep} - 6 {\vt r}_{t-\timestep} + 3 {\vt r}_t + 2 {\vt r}_{t+\timestep}}{6\timestep}.
+\dot{\vt r}^{[3]}_t = \frac{{\vt r}_{t-2\timestep} - 6 {\vt r}_{t-\timestep} + 3 {\vt r}_t + 2 {\vt r}_{t+\timestep}}{6\timestep}.
 \end{equation*}
@@
 \begin{equation*}
-\dot{\vt q}^{[3]} = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
+\dot{\vt q}^{[3]}_t = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
 \end{equation*}
 where ${\mt \Pi}$ is a matrix-valued function given by either product ${\mt B}\tr{\mt B}$ or ${\mt C}\tr{\mt C}$ \cite{Silveira_2017}.
-In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.
+In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.
+
+\subsection{Adjustment of the NVT Dynamics}
 
@@
 
+Speak about the Generalized Equipartition Theorem....
+
 Finally, reweighting:
@@
 
-Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
+Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010_2} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
 Sergi and Ferrario\cite{Sergi_2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.

### 2018-03-25 rigid_bodies.tex
Small changes

@@
 \label{eq:nvt extended energy}
-H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
+H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
 \end{equation}
@@
 
-In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$. Besides, $G_j$ is a generalized force acting on thermostat $j$, defined as
+In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
 \begin{subequations}
@@
 Analysis of the probability distribution generated by Eq.~\eqref{eq:ODE system for NVT} can be done through the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables eliminated.
-For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates.
+If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are integrals of motion, while the system's center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates.
 Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
@@
 Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
-Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied again, but now with $j$ varying from $1$ up to $M$.
+Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
 
@@
 
-Speak about the Generalized Equipartition Theorem....
+Tolman's generalized equipartition theorem \cite{Uline_2008}
+\begin{equation*}
+\left\langle \tr{\vt p} \diff{\modified{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\modified{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T
+\end{equation*}
 

### 2018-05-07 rigid_bodies.tex
corrected factor in omega
Separation of sentences for better diff tracking
Changes in citations in the introduction section

@@
 
-\author{Ana J. Silveira}
+\author{Ana J.
+Silveira}
 \email{asilveira@plapiqui.edu.ar}
@@
 
-\author{Charlles R. A. Abreu}
+\author{Charlles R.
+A.
+Abreu}
 \email{abreu@eq.ufrj.br}
@@
 
-Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman_2010} involves several types of approximations in order to mimic ``real'' physical systems.
+Molecular Simulation, being the computational realization of Statistical Mechanics \cite{Tuckerman_2010}, involves several types of approximations in order to mimic ``real'' physical systems.
 Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
-However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour.
-For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
-That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
-
-The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski_2012, Patra_2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj_2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton_1995} has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
-
-The Nos\'{e}-Hoover chain thermostat,\cite{Martyna_1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
+However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
+For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property \cite{Eastwood_2010}.
+In a previous paper \cite{Silveira_2017}, we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
+That finding brings into question the method introduced by Kamberaj and co-workers \cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
+
+The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
+As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the softwares packages HOOMD-\textit{blue} \cite{Anderson_2008} and LAMMPS \cite{Plimpton_1995}, has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008} and nanoparticles \cite{Patra_2014}.
+
+The Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
 The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible.
-Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}
+Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
 
 In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble.
-The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
+The starting point is our previous work on the microcanonical (NVE) ensemble \cite{Silveira_2017}, which employs the unit quaternion representation for the rotational degrees of freedom.
 Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
-The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj_2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy.
-We also include results from simulations using the conventional SHAKE\cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
+The analysis of the integrators includes comparisons with results from the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy.
+We also include results from simulations using the conventional SHAKE \cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
 
@@
 
-In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to 1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and 2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} as a sequence of uniform rotations around the principal axes of a body, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.
+In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
+1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and
+2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} as a sequence of uniform rotations around the principal axes of a body, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.
 
@@
 \end{equation}
-where $K$ and $U$ are the kinetic and potential energies of the body, respectively. While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
+where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
+While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
 \begin{equation}
@@
 
-Each additional thermostat will then act on the preceding one, and be acted upon by the succeeding one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
+Each additional thermostat will then act on the preceding one, and be acted upon by the succeeding one.
+Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
 \begin{equation*}
@@
 
-In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies. The Hamiltonian is:
+In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies.
+The Hamiltonian is:
 \begin{equation*}
@@
 {\mt M}^{-1} & \mt 0 \\
-\mt 0 & \frac{1}{2} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
+\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
 \end{array}\right],
@@
 The important thing is that $\modified{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] = \modified{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
-In this way, the equipartition theorem (valid at least in this direction?
-something to do with integrating through momenta at least internally to the integral in coordinates?) states that
+In this way, the equipartition theorem (valid at least in this direction? something to do with integrating through momenta at least internally to the integral in coordinates?) states that
 \begin{equation*}
@@
 \end{equation}
-where $N_{dof}$ is the number of degrees of freedom. Eq.~\eqref{eq:mb} is a normal distribution, with average  $ \langle K \rangle = N_{dof}/2\beta $ and variance $\sigma^2 = N_{dof}/2\beta^2$.
+where $N_{dof}$ is the number of degrees of freedom.
+Eq.~\eqref{eq:mb} is a normal distribution, with average  $ \langle K \rangle = N_{dof}/2\beta $ and variance $\sigma^2 = N_{dof}/2\beta^2$.
 
@@
 			\cline{2-4}
-			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}. \\
+			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}.
+\\
 			\hline
@@
 %  \includegraphics{pressure2}
-%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars). Simulation of liquid water with NVE integration. $\timestep : 1 fs$  }
+%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars).
+%Simulation of liquid water with NVE integration.
+%$\timestep : 1 fs$  }
 %\end{center}
@@
 Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.
+
 by contrast
@@
 %&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) %t \quad \text{and} \label{eq:solution_p} \\
-%&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi%\left(\alpha_1 t \right) t. \label{eq:solution_pi}
+%&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi%\left(\alpha_1 t \right) t.
+\label{eq:solution_pi}
 %\end{align}

### 2018-05-12 rigid_bodies.tex
some small changes

--- 2018-03-25 rigid_bodies.tex
@@
 
+backward error analysis
+
 The paper is organized as follows.
@@
 
-\section{Proposed Methodology}
+\section{Methodology}
 \label{sec:methodology}
@@
 
-In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to 1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and 2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} as a sequence of uniform rotations around the principal axes of a body, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.
-
-The Hamiltonian system of ordinary differential equations (ODE) that describes a rigid-body motion are \cite{Silveira_2017}
+In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
+1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and
+2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002}, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.
+
+The Hamiltonian system of ordinary differential equations (ODE) that describes a rigid-body motion is given by \cite{Silveira_2017}
 \begin{subequations}
@@
 In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugated momentum of such quaternion.
-Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity.
+Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity in the body-fixed frame.
 Matrices $\mt M$ and $\mt I$ are diagonal ones.
@@
 
-The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the body-fixed frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components.
-The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$.
+%The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the body-fixed frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components.
+The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, with ${\mt A}(\vt q)$ being the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames).
 
@@
 \end{equation}
-
-An advantage of this notation is that Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
+%An advantage of this notation is that
+Of course, Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
 The same occurs for diagonal matrices $\mt M$ and $\mt I$, whose sizes increase to $3N \times 3N$ in this case.
@@
 
-A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$.
+A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$, which is
+\begin{equation}
+\label{eq:Lioville operator}
+\begin{split}
+\Liu{\tiny NVE} &= \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \tr{\vt F} \diff{}{\vt p} \\
+&+ \left[ \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \right] \diff{}{\vt \pi}.
+\end{split}
+\end{equation}
+
 A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
@@
 In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state immediately before the propagation occurs.
-The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step.
-
-Brief description of analytical and NO-SQUISH rotations...
+The action of $e^{\timestep \Liu{A}}$ is the simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step.
+For this rotational part of the integration, in Ref.~\cite{Silveira_2017} we derived a general solution in a form that facilitates computer implementation and allows, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
+While the approximate solutions, such as the NO-SQUISH \cite{Miller_2002} method, only involve sine and cosine terms, the exact solution comprise ``less common'' functions, such as the Jacobi elliptic functions and Carlson integrals.
+This contributes to the fact that the approximate solution is more commonly used \cite{vanZon2007}. 
+For mathematical details, the reader is referred to our previos work \cite{Silveira_2017}, where we have verified that in liquid-phase simulations there is no significan gain in accuracy by using the exact solution in substitution to the NO-SQUISH as part of a symplectic integration scheme.
+However, in the NO-SQUISH method an overall rotation is given by a sequence of uniform rotations around the principal axes of a body, which results in a factorization of the rotational operator given by $ e^{\frac{h}{2} i\!L_3} e^{\frac{h}{2} i\!L_2} e^{h i\!L_1} e^{\frac{h}{2} i\!L_2} e^{\frac{h}{2} i\!L_3}$, with each factor representing a rotation around a given axe.
+The analytical solution, on the other hand, entails a single operator, thus making the application of backward error analysis considerably simpler.
 
@@
 Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ act sequentially, with $j$ varying from $M-1$ down to $1$.
-With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as
+With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as \cite{Martyna_1996}
 \begin{align*}
@@
 Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
+In the following, the NVT integrator introduced in the present section will be denoted by MKKT as its development was inspired by Ref.~\cite{Martyna_1996}.
 
@@
 
-\subsection{Sampling Validation}
-\label{sec:samplingvalidation}
-
-In this subsection we briefly review a simple method that serves for checking the consistency between the sampled and theoretical energy distributions.
-The procedures, implemented by Shirts,\cite{Shirts_2013} rely on quantitative statistical analysis of fitting, and it is possible to perform linear and nonlinear fitting or a maximum likelihood approach too.
-First, recall that in the NVT ensemble the probability of observing an energy E is given by
-\begin{equation*}
-P(E/\beta) = Q(\beta)^{-1}\Omega(E)\exp\left(-\beta E\right),
-\end{equation*}
-where $ \Omega(E) $ is the density of states, which does not depend on $ \beta $, and $ Q(\beta) $  is the canonical partition function, linked to the Helmholtz Free Energy $ (A) $ by $A = -\beta^{-1}\ln(Q)$.
-
-Taking the ratio of the probability distributions at different temperatures, the unknown density of states cancels out and we get
-\begin{equation}
-\frac{P(E/\beta_2)}{P(E/\beta_1)} = \exp\left[(\beta_2 A_2 - \beta_1 A_1) - (\beta_2 - \beta_1)E\right].
-\end{equation}
-Taking the logarithm of the equation above, we obtain
-\begin{equation}
-\label{eq:ln_probability_ratio}
-\ln\frac{P(E/\beta_2)}{P(E/\beta_1)} = \left(\beta_2 A_2 - \beta_1 A_1\right) - \left(\beta_2 - \beta_1\right)E,
-\end{equation}
-which is the basis for the statistical analysis carried out in the ensemble validation that determines how many standard deviations the calculated slope is from the specified one,  $ \beta_2 - \beta_1 $.
-Note that it is independent of the unknown free energies.
-For a guidance on how to choose the temperature gap the reader is referred to the aforementioned paper.
-
-Furthermore,  it is possible to examine the consistency between the sampled kinetic energy $K$ and the corresponding Maxwell-Boltzmann distribution, which is given by the following expression:
-\begin{equation}
-\label{eq:mb}
-P(K) = \frac{\beta}{\sqrt{N_{dof}\pi}} \exp\left[-\frac{\left(\beta K - \frac{N_{dof}}{2}\right)^2}{N_{dof}} \right],
-\end{equation}
-where $N_{dof}$ is the number of degrees of freedom. Eq.~\eqref{eq:mb} is a normal distribution, with average  $ \langle K \rangle = N_{dof}/2\beta $ and variance $\sigma^2 = N_{dof}/2\beta^2$.
+Sampling validation: Shirts \cite{Shirts_2013}
 

### 2018-05-28 rigid_bodies.tex
added supplementary material

--- 2018-05-12 rigid_bodies.tex
@@
 % Comment above before submitting
-
+\usepackage{booktabs}
 \usepackage{threeparttable}
@@
 \definecolor{lightgray}{gray}{0.9}
-
+\usepackage{tikz}
+\usetikzlibrary{decorations.text}
 \newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
@@
 \newcolumntype{R}{>{$}r<{$}}
+\usepackage{booktabs}    
+\usepackage[exponent-product=\times]{siunitx}
 
@@
 \subsection{NVE Dynamics: Notation and Symplectic Integration}
-
+\label{sec:nve}
 In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
@@
 
-%The position of any constituent particle $j$ of a rigid body will vary with time according to $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in the body-fixed frame of reference and ${\mt A}(\vt q)$ is the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components.
 The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, with ${\mt A}(\vt q)$ being the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames).
@@
 \subsection{Measure-Preserving Integration for the NVT Dynamics}
+\label{sec:mttk}
+As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
+For that, we employed the systematic method introduced by Ezra \cite{Ezra_2006} for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion, which is
+In short, a splitting scheme will preserve a given measure if each individual propagator does it as well. 
+It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure is given by
+\begin{equation}
+\label{eq:nhc_measure}
+w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
+\end{equation}
 
-As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
-Ezra \cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion.
-In short, a splitting scheme will preserve a given measure if each individual propagator does it as well.
+will preserve a given measure $e^{-w(\vt x)}d\vt x$
+
 By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
@@
 Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
-In the following, the NVT integrator introduced in the present section will be denoted by MKKT as its development was inspired by Ref.~\cite{Martyna_1996}.
+In the following, the NVT integrator introduced in the present section will be denoted as MTTK as its development was inspired by Ref.~\cite{Martyna_1996}.
 
@@
 \subsection{Backward Error Analysis}
+\label{sec:modified_h}
 
@@
 A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
-Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian.
+Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian \cite{Hairer_2006}.
 In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors). 
-It is also known that $\Ham{S}$ depends on the time step size $\timestep$.
-As shown in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are modified kinetic and potential energies.
+It is also known that $\Ham{S}$ explicity depends on the time step size $\timestep$.
+
+The first new result we present in this paper is an analytical form for the shadow Hamiltonian concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~ \citenum{Silveira_2017}.
+A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and here we summarize the final expressions.
+
+As shown in Appendix~\ref{sec:rigid body shadow hamiltonian}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are modified kinetic and potential energies.
 The latter is given by
@@
 \end{equation}
-where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$.
-Therefore, a Trotter-Suzuki trajectory, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
+where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
+\begin{equation*}
+{\mt \Omega} = \left[\begin{array}{cc}
+{\mt M}^{-1} & \mt 0 \\
+\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
+\end{array}\right]
+\end{equation*}
+and ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, given by
+\begin{equation*}
+{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
+U_{\vt r \vt r} & U_{\vt r \vt q} \\
+U_{\vt q \vt r} & U_{\vt q \vt q}
+\end{array}\right].
+\end{equation*}
+
+Hence, the trajectory generated by the splitting scheme of Eq.~\ref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
 \begin{equation*}
@@
 \dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}, \; \text{and} \;
-\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q},
+\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}.
 \end{equation*}
-where $\modified{\Ham{}} = \modified K + \modified U$ is a modified Hamiltonian.
+
 The first two of these equations can be combined together so that we can use Eq.~\eqref{eq:modified kinetic energy} to obtain
@@
 In contrast, although it is possible to compute $\modified K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
-Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained Trotter-Suzuki trajectory.
+Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
 Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
@@
 \subsection{Adjustment of the NVT Dynamics}
-
+\label{sec:smttk}
 Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
 Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \modified{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
-Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly as described before.
+Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
 The only required modification is that $\modified K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
-By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that such force becomes
+By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
 \begin{equation}
+\label{eq:thermostat_force}
 G_1 = 2 \modified K - L k_\text{\tiny B} T.
@@
 By incomplete we mean that the step is aborted once the positions and orientations have been updated.
+\begin{figure}
+\begin{tikzpicture}
+%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](3,.2) -- (3,1.5);
+%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
+\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
+\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
+\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
+\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
+\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
+\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
+\node at (0.0,-0.2)[below]{$r_{t-h}$};
+\node at (2.2,-0.2)[below]{$r_{t}$};
+\node at (4.4,-0.2)[below]{$r_{t+h}$};
+\node at (6.6,-0.2)[below]{$r_{t+2h}$};
+\path[->,thick] (1) edge [bend left=45] (2);
+\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
+\path[dashed,->,thick] (1) edge [bend right=45] (0);
+\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
+\path[dashed,->,thick] (2) edge [bend left=45] (3);
+\end{tikzpicture}
+\caption{Schematic representation}
+\label{fig:virtual}
+\end{figure}
+%\begin{figure}
+%\begin{tikzpicture}
+%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](2,.2) -- (2,1.5);
+%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
+%\node(0) at (0,0)[shape=circle,draw, minimum width=1cm, minimum height=1cm]        {$r_{t-h}$};
+%\node(1) at (2.5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_t$};
+%\node(2) at (5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_{t+h}$};
+%\node(3) at (7.5,0)[shape=circle,draw,minimum width=1cm, minimum height=1cm] {$r_{t+ 2h}$};
+%\node(4) at (3.75,1)[] {$e^{\timestep \Liu{\tiny NVE}}$};
+%\path [dashed,->]
+%  (1)      edge                 node [above]  {virtual}     (0)
+%  (2)      edge                 node [above] {virtual}     (3);
+%\path [->,thick] 
+%  (1)      edge                 node [above]  {}            (2);
+%\end{tikzpicture}
+%\caption{Schematic representation}
+%\label{fig:virtual}
+%\end{figure}
+
 A drawback of this approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
@@
 Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\modified K$ goes as simply as ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$.
+
+$\widetilde{H}$, but an ``extended energy'' given by \cite{Martyna_1992}
+\begin{equation}
+\label{eq:this extended energy}
+\widetilde{H} = \modified K + \modified U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
+\end{equation}
 
@@
 \section{Theoretical Background}
-\label{sec:theory}
 
-\subsection{Hamiltonian Dynamics}
-\label{eq:hamiltonian dynamics}
+We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj_2005} does not fulfill this criterion.
 
-When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of the center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t)$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}.
-In this way, the connection between the body-fixed and space-fixed frames of reference can be recovered at any time by the rotation matrix $\mt A(\vt q)$, which admits the factorization mentioned in Sec.~\ref{sec:methodology}.
-This factorization creates an intermediate frame of reference for representing three-dimensional vectors in $\mathbb{R}^3$ as quaternions that lie in the hyperplane orthogonal to $\vt q$, since the columns of matrix $\mt B$ and those of matrix $\mt C$ form two different bases for such a subspace of $\mathbb{R}^4$.
-Transformations amongst three distinct reference frames are enacted by multiplications with these matrices or their transposes, such as represented in the following scheme:
-\begin{equation*}
-\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
-\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
-\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
-\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
-\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
-\end{equation*}
+We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
+Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
+However, as will be shown in the numerical results of Sec.~\ref{sec:numerical_results},
 
-The intermediate frame is exactly the subspace of $\mathbb{R}^4$ that must contain the rate-of-change vector $\dot{\vt q}$ if the norm of $\vt q$ is to remain constant, since $\tr{\vt q} {\vt q} = 1$ implies that $\tr{\vt q} \dot{\vt q} = 0$.
-In the Hamiltonian approach for the motion of a rigid body, the momentum $\vt \pi$ conjugated to $\vt q$ is given by $\vt \pi = 2 \mt B(\vt q) \mt I \vt \omega$.\cite{Silveira_2017} This means that $\vt \pi$ is the quaternion-fixed representation of the angular momentum $\vt L = \mt I \vt \omega$, thus being orthogonal to $\vt q$ as well.
+\subsection{Generalized Equipartition}
 
-The Liouville operator that stems from the Hamiltonian in Eq.~\eqref{eq:Hamiltonian} is\cite{Silveira_2017}
+The important thing is that $\modified{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] = \modified{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
+In this way, the equipartition theorem (valid at least in this direction? something to do with integrating through momenta at least internally to the integral in coordinates?) states that
 \begin{equation}
-\label{eq:Lioville operator}
-\begin{split}
-\Liu{\tiny NVE} &= \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \tr{\vt F} \diff{}{\vt p} \\
-&+ \left[ \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \right] \diff{}{\vt \pi}.
-\end{split}
+\label{eq:equipartition}
+3N k_B T = \left\langle \tr{\vt p} \diff{\tilde K}{\vt p} \right\rangle = 2\langle \tilde K \rangle
 \end{equation}
 
-Because the time evolution of any phase-space vector $\vt x$ is governed by $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ \cite{Tuckerman_2010}, the equations of motion that describe the rigid-body motion are those in Eq.~\eqref{eq:ODE system for NVE}.
-The solution for a system like $\tr{\dot{\vt x}} = \Liu{\tiny NVE} \tr{\vt x}$ with initial condition $\vt x(0) = \vt x_0$ can be formally expressed as $\tr{\vt x}(t) = e^{t \Liu{\tiny NVE}} \tr{\vt x}_0$.
-Although the exact effect of the exponential operator %$e^{t \Liu{\tiny NVE}}$
-cannot be determined for an arbitrary potential $U(\vt r, \vt q)$, a symplectic numerical solution can be achieved by a Verlet-type integration.
+Sampling validation: Shirts \cite{Shirts_2013}
+
+\section{Results}
+\label{sec:numerical_results}
+
+In the following, we analyze the performance of the proposed NVT integrators introduced in Secs.
+\ref{sec:mttk}-\ref{sec:smttk}, and  compare the results with those obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
+The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA}.\cite{Silveira_2017} 
+The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.
+
+Before showing our main results, it may be instructive first to give a picture of the problem in controlling the temperature when discretization errors are non-negligible. 
+The most direct way (the most senstive indicator of the presence of discretization errors) to determine the importance of these errors is by examining the departure of the ``modified Hamiltonian'' $\modified{\Ham{}}$ from the physical Hamiltonian $\Ham{}$\cite{Engle_2005}.
+Fortunately, with the analytical form of $\modified{\Ham{}}$ derived in this paper we are able to perform such an analysis for the NVE integrator which is essential, as will be shown below, to understand the results of the NVT methods.
+In Fig.~\ref{fig:nve} we depict the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different step sizes ($\timestep$).
+The numerical integrator correspons to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation. 
+As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
+It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
+As a result of employing such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ fs and $\timestep$ fs (add in conclusions that we should use high-order approximations).
+
+\begin{figure}
+	\includegraphics{Figures/NVE.eps}
+	\caption{Total energy $\Ham{}$ (solid lines) and modified Hamiltonian $\modified{\Ham{}}$ (dotted lines) of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVE MD simulations for different step sizes $\timestep$.
+The numerical integrator correspons to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation.}
+	\label{fig:nve}
+\end{figure}
+
+For our implementation for rigid bodies of the numerical scheme due to Martyna \textit{et al.} \cite{Martyna_1996} (Eq.~\ref{eq:splitting NVT}), in Fig.~\ref{fig:props_k_m} we depict the computed averages of temperature ($\langle T \rangle$) and potential energy per molecule ($\langle U \rangle$) as a function of $\timestep$.
+The specified temperature corresponds to T = 298 K.
+Fig.~\ref{fig:props_k_m} clearly shows that a suitable temperature control leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
+As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
+Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
+These observations motivated the development of the scheme of Sec.~\ref{sec:smttk}, whose promising results we show below.
+
+\begin{figure}
+	\includegraphics{Figures/thermodynamic_properties_intro.eps}
+    \caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
+	\label{fig:props_k_m}
+\end{figure}
+
+\subsection*{Numerical Stability}
+
+First, we report related the numerical stability of the NVT integrators.
+
+examine the numerical stability of the NVT integrators by simply monitoring the quantity expected to be conserved along the simulation in each case: $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for the proposed method, and $H$ (Eq.~\ref{eq:nvt extended energy}) for the integrators due to Martyna \textit{et al.} \cite{Martyna_1996} and Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+In Table~\ref{table:stability} we report the energy drift, denoted by $R$, as the the rate of change of $\widetilde{H}$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting between those variables.
+As it is already known that \cite{Davidchack_2009,Bond_2007}
+
+we plot these extended energies for different time step sizes.
+From Figs.~\ref{fig:num_stab}(a) and (b) which, respectively, correspond to the integrator proposed in this work and our implementation of the scheme due to Martyna \textit{et al.} \cite{Martyna_1996}, one notes that the energy appears to remain bounded for time step sizes up to 5 fs, and a continous drift takes place for $\timestep$ = 6 fs.
+The numerical stability is only modestly improved by our method.
+The result that stands out, because of its anomalous feature, corresponds to Fig.~\ref{fig:num_stab} (c), which was obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+Note that for each time step size, the extended energy continously increases until reaching a point at which the integrator becomes stable.
+It should be remarked that these results were double-checked with the LAMMPS software package \cite{Plimpton_1995}. 
+As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
+Davidchack \cite{Davidchack_2009} also implemented such a separate thermostatization and discovered the existance of energy flows from one thermostat to the other.
+To the best of our knowledge, the work of Davidchak \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+As can be seen in Fig.~\ref{fig:ther_energies}, where we depict the energy of each thermostat for different time step sizes $h$, that flux actually exists.
+Also, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, while the positive values are associated to the thermostat coupled to the rotational degrees of freedom, and thus the energy is flowing from translational to rotational degrees of freedom.
+Interestingly, Fig.~\ref{fig:ther_energies} suggests that a unique steady state is established, and the time to reach the steady state decreases with $h$.
+It should be noted that all the simulations were carried out considering $t_d = 100$ fs.
+As far as we know, this behavior has not been reported before.
+reformulate(The drift is localized within the thermostat variables and thus is not indicative of the non-stationary evolution within the system itself).
+
+\begin{table}
+	\caption{Effect of the time-step size on ....
+of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.}
+    \label{table:stability}
+    \begin{tabular}{SS[table-number-alignment = center,table-figures-exponent = 1]S[table-number-alignment = center,table-figures-exponent = 1]}
+     & \multicolumn{2}{c}{Drift rate [(kcal/mol)/ns]}  \\
+     \cmidrule[0.5mm](lr){2-3} 
+     $\timestep$ $\text{(fs)}$  & $\modified{H}$  & $\modified{\Ham{}}$   \\
+	1 &  4.6e-4  & 8.5e-2 \\
+	2 & -8.2e-3 & -4.3e-2 \\
+	3 & -3.2e-2 & 7.3e-2\\
+	4 & 8.1e-3 & -1.1e-1 \\
+	5 & 1.3e0 & -2.2e-2\\
+    6 & 4.4e1 & -3.4e-3 \\
+	7 & 8.8e2 & 2.2e-2 \\
+	8 & 9.8e3 & 7.6e-2 \\
+	9 & 7.1e4 & -2.5e-2 \\
+	10 & 3.2e5 & -3.2e-2  \\
+	\end{tabular}
+\end{table}
+
+\begin{figure}
+	\includegraphics{Figures/numerical_stability.eps}
+    \caption{Total extended energy of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
+step sizes, employing (a) the proposed method represented by Eq.
+; (b) our implementation for rigid bodies of the integrator due to Martyna \textit{et al.} \cite{Martyna_1996} and (c) the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.Energy of thermostats for 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
+step sizes, employing the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.}
+	\label{fig:num_stab}
+\end{figure}
+
+\subsection*{Effect of Discretization Errors on Equilibrium Thermodynamic Properties}
+
+We now turn our focus to discretization errors.
+In Fig.~\ref{fig:properties} we show the computed average of some thermodynamic properties as a function of $\timestep$.
+The simulations were carried out at a specified temperature of T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_v$).
+The temperature was calculated from the kinetic energy assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
+The lines are a guide for the eye. 
+Also, the horizontal dotted lines (black) help to visualize the departure of the computed averages with increasing $\timestep$.
+They correspond to the results obtained employing $h$ = 1 fs in the numerical scheme proposed in this paper, which from now on will be referred as ``This work''.
+In addition, we have included the results of applying our proposal but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
+Despite the non-equilibrium process ocurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}, because that is not the only problematic issue related to the method in question.
+Particularly, from Fig.~\ref{fig:properties} (a) we note that the simulated temperature decreases systematically with $\timestep$.
+This kind of behavior is observed in NVE simulations\cite{Davidchack_2009,Silveira_2017}, reflecting that the factorization employed in Kamberaj \textit{et al.} \cite{Kamberaj_2005} in not suitable. 
+As a result of this inefficient temperature control, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Fig.~\ref{fig:properties} (b) and (c)).
+From Fig.~\ref{fig:properties} it becomes clear that our method yields significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
+This is true also for the results corresponding to the use of the NO-SQUISH method\cite{Miller_2002} in the NVE part of the integrator.
+Mathematically, the derivation of the refined Hamiltonian was facilitated by the fewer number of operators associated with the exact solution of rotation.
+However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement.
+
+With the aim of highlighting the importance of reweighting, in Table~\ref{table:reweight} we present the calculated temperature and potential energy for each $\timestep$ before and after applying Eq.~\ref{eq:reweighting}.
+The ``refined'' temperature $\modified{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
+Of course, as $\modified{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
+Indeed, for time step sizes up to, the agreement is excelent.
+Note that in the abscence of reweighting, the calculated temperatures are notably influenced by $h$.
+The free energy difference between the sampled state $\modified{\Ham{}}$ and the state we would like to sample, represented by the physical Hamiltonian $\Ham{}$.
+As expected, the $\Delta F$ is small, as it should be
+Note that a small, still clearly influencing systematic error is present in our scheme.
+
+\begin{table*}
+	\caption{Effect of the time-step size on ....
+of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.
+uncertainties in free energy never exced 1e-4}
+    \label{table:reweight}
+    \sisetup{
+table-number-alignment = center
+}
+    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]S[separate-uncertainty,table-figures-uncertainty=2]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S[separate-uncertainty,table-figures-uncertainty= 1]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S}
+     & &  \multicolumn{2}{c}{no reweighting} & \multicolumn{2}{c}{after reweighting} & \\
+     \cmidrule[0.5mm](lr){3-4} \cmidrule[0.5mm](lr){5-6}
+     $\timestep$ & $\modified{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F$} \\
+        $\text{(fs)}$  & $\text{(K)}$ & $\text{(K)}$ & $\text{(kcal/mol)}$  & $\text{(K)}$ & $\text{(kcal/mol)}$ & $\text{(kcal/mol\,N)}$ \\
+	1 & 298.00(1) & 297.71(1) & -9.1015(3) & 298.00(1) & -9.1028(3) & -0.001 \\
+	2 & 297.99(1) & 296.82(1) & -9.0970(3)  & 297.98(1)  & -9.1024(3) & -0.006 \\
+	3 & 297.98(1) & 295.38(1) & -9.0885(3) & 297.92(1) & -9.1005(3) & -0.013 \\
+	4 & 298.01(1) & 293.49(1)& -9.0764(3) & 297.85(3) & -9.0969(3) & -0.022 \\
+	5 & 297.99(1)& 291.13(1)& -9.0577(3) & 297.6(2) & -9.091(1) & -0.033 \\
+    6 & 298.00(1) & 288.46(1) & -9.0314(3) & 298.8(3) & -9.075(3) &  -0.045 \\
+	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
+
+	\end{tabular}
+\end{table*}
+
+To end this section,  Kinetic Energy Partition
+
+\begin{figure*}
+	\includegraphics{Figures/thermodynamic_properties.eps}
+	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
+	\label{fig:properties}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics{Figures/energy_partition.eps}
+\end{figure*}
+
+\begin{table*}
+	\caption{Effect of the time-step size on ....
+of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes....}
+	\label{table:check_ens}
+	\sisetup{table-number-alignment = center}
+    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]SS[separate-uncertainty,table-figures-uncertainty= 1]SS[separate-uncertainty,table-figures-uncertainty= 1]S}
+	{$\timestep$} & {estimated $\Delta T$} & {$\sigma$ deviations}  & {estimated $\Delta T$}  & {$\sigma$ deviations}  &    {estimated $\Delta T$}  & {$\sigma$ deviations} \\
+	{(fs)} & {(K)} &   & {(K)}  &  & {(K)}  & \\
+			\hline
+			\multicolumn{7}{c}{``This work (Unsplit)''}\\
+1.0	& 3.02(2) & 1.07 & & & &  \\
+2.0 & 3.02(2) & 0.81 & & & &  \\
+3.0 & 2.98(2) & 0.73 & & & &  \\
+4.0 & 3.01(2) & 0.41 & & & &  \\
+5.0 & 3.02(2) & 0.79 & & & &  \\
+6.0 & 3.03(3) & 1.23 & & & &  \\
+7.0 & 2.98(2) & 0.74 & & & &  \\
+            \hline
+			\multicolumn{7}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}\\
+1.0	& 2.98(2) & 0.66 & 3.01(2) & 0.49 & 2.96(3) & 0.90 \\
+2.0 & 2.98(2) & 0.73 & 2.99(2) & 0.03 & 3.02(3) & 0.66 \\
+3.0 & 2.95(2) & 2.33 & 3.01(2) & 0.65 & 2.91(3) & 2.47 \\
+4.0 & 3.00(2) & 0.01 & 3.01(2) & 0.72 & 2.92(3) & 2.19  \\
+5.0 & 2.93(2) & 3.05 & 3.01(2) & 0.70 & 2.94(3) & 1.58  \\
+6.0 & 2.92(2) & 3.87 & 3.00(2) & 0.22 & 2.91(3) & 2.76  \\
+7.0 & 2.89(2) & 5.35 & 3.02(1) & 1.33 & 2.87(2) & 4.33  \\
+            \hline
+			\multicolumn{7}{c}{Kamberaj \textit{et al.} \cite{Kamberaj_2005}}\\
+1.0	& 2.97(2) & 1.16 & 2.99(2) & 0.16 & 2.89(3) & 2.84  \\
+2.0 & 3.00(2) & 0.02 & 3.02(2) & 0.99 & 2.00(4) & 0.25  \\
+3.0 & 2.99(4) & 0.18 & 3.08(4) & 4.09 & 3.1(1) & 1.17  \\
+4.0 & 2.92(8) & 0.99 & 3.12(2) & 6.01 & 2.8(2) & 0.95  \\
+5.0 & 3.02(5) & 0.40 & 3.2(2)  & 9.50 &  2.9(1) & 0.28  \\
+6.0 & 3.14(5) & 2.69 & 3.29(2) & 14.68 & 3.21(7) & 2.90  \\
+7.0 & 3.03(4) & 0.73 & 3.35(2) & 17.37 & 2.99(6) & 0.16  \\
+    \end{tabular}
+\end{table*}
+%It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
+%First we present some results as first mented.
+%That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
+%In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
+%By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
+%Resultados sobre o checkensemble:
+
+%checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
+%Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
+%m-b distribution valid when equipartition of energy is satisfied.
+
+%Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
+%Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.
+
+%In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
+%In figure legends we show the averages temperatures simulated.
+%It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
+%Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.
+
+%Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
+%It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
+%This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
+
+\section{Concluding Remarks}
+\label{sec:conclusion}
+Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.
+
+by contrast
+
+\begin{acknowledgement}
+	The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
+\end{acknowledgement}
+
+\appendix
+
+\section{Backward Error Analysis}
+\label{sec:rigid body shadow hamiltonian}
+
 For two non-commutative Liouville operators $\Liu A$ and $\Liu B$ and a real constant $\timestep$, the symmetric Baker-Campbell-Hausdorff (BCH) formula can be expressed as \cite{Hairer_2006}
@@
 The remaining terms of the infinite series involve ever-deepening, nested commutators \cite{Hairer_2006}.
-With $\Liu{\tiny NVE} = \Liu{A} + \Liu{B}$ and a small value for $\timestep$, we can approximate the BCH equation by the Trotter-Suzuki splitting formula, Eq.~\eqref{eq:splitting NVE}.
-In the rigid-body case, by making
-\begin{subequations}
-	\label{eq:Liouvile operators K and U}
-	\begin{equation*}
-	\label{eq:Lioville operator K}
-	\Liu{A} = \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) \diff{}{\vt \pi}
-	\end{equation*}
-	and
-	\begin{equation*}
-	\label{eq:Lioville operator U}
-	\Liu{B} = \tr{\vt F} \diff{}{\vt p} + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \diff{}{\vt \pi},
-	\end{equation*}
-\end{subequations}
-it becomes possible to determine exactly the individual actions of operators $e^{\timestep \Liu{A}}$ and $e^{\frac{\timestep}{2} \Liu{B}}$ because they yield two separate, analytically solvable systems of equations.
-The one that originates from $e^{\timestep \Liu{A}}$ is made up of equations $\dot{\vt r} = {\mt M}^{-1} {\vt p}$, $\dot{\vt p} = {\vt 0}$, $\dot{\vt q} = \frac{1}{2} \mt B(\vt q) \vt \omega$, and $\dot{\vt \pi} = \frac{1}{2} \mt B(\vt \pi) \vt \omega$, while the one that stems from $e^{\frac{\timestep}{2} \Liu{B}}$ is composed of $\dot{\vt r} = {\vt 0}$, $\dot{\vt p} = {\vt F}$, $\dot{\vt q} = {\vt 0}$, and $\dot{\vt \pi} = 2 \mt C(\vt q) \vt \tau$.
-Their solutions provides us with the propagator actions described in Sec.~\ref{sec:methodology}.
-
-\subsection{Nos\'e-Hoover Chain Applied to Rigid Bodies}
-\label{sec:canonical}
-
-Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010_2} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space.
-Sergi and Ferrario\cite{Sergi_2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space.
-If vector $\vt x$ represents a point in phase space, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
-\begin{equation}
-\label{eq:eq_of_motion}
-\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
-\end{equation}
-will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric (that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$).
-A generalized Liouville operator can then be defined as\cite{Sergi_2004}
-\begin{equation*}
-{\mathcal L} = \tr{\dot{\vt x}}\diff{}{\vt x} = \tr{\left(\diff{H}{\vt x}\right)} \tr{\boldsymbol{\mathcal B}} \diff{}{\vt x}.
-\end{equation*}
-
-The system might exhibit a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative.
-According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman_2001} although the phase-space volume element $d\vt x$ is no longer conserved, there exits an invariant measure $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$.
-Using the generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
-\begin{equation}
-\label{eq:relation_kappa_w}
-\kappa = \dot w = {\mathcal L} w.
-\end{equation}
-
-We employ the technique of Sergi and Ferrario\cite{Sergi_2001} to couple a single Nos\'e-Hoover chain\cite{Martyna_1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies.
-This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed two thermostat chains independently coupled to the translational and rotational degrees of freedom.
-To this end, we consider extra coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$.
-The conserved ``extended Hamiltonian'' in this case is given by Eq.~\eqref{eq:nvt extended energy}.
-The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e.
-$\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$.
-However, the equations of motion for generalized momenta contain non-Hamiltonian terms.
-For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
-\begin{equation*}
-\begin{split}
-\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
-\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
-\end{split}
-\end{equation*}
-
-In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat.
-In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, along with a term regarding the action of a second thermostat.
-The result is
-\begin{equation*}
-{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
-\end{equation*}
-
-Each additional thermostat will then act on the preceding one, and be acted upon by the succeeding one.
-Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
-\begin{equation*}
-{\dot p}_{\eta_j} = -\diff{H}{\eta_j} + p_{\eta_{j-1}} \diff{H}{p_{\eta_{j-1}}} - p_{\eta_j} \diff{H}{p_{\eta_{j+1}}}.
-\end{equation*}
-
-Given that the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above.
-As a result, the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain are those in Eq.~\eqref{eq:ODE system for NVT}.
-Note that the $3N$ momentum components in Eq.~\eqref{eq:nhc_p}, the $4N$ components in Eq.~\eqref{eq:nhc_pi}, and the thermostat momenta in Eq.~\eqref{eq:nhc_p_eta} affect their own time-derivatives.
-Thus, the thermostatted system has a phase-space compressibility $\kappa = -7N \alpha_1 - \sum_j \alpha_{j+1}$.
-Resorting to Eq.~\eqref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
-\begin{equation}
-\label{eq:nhc_measure}
-w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
-\end{equation}
-
-Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables must be eliminated.
-For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i = \vt K$ are integrals of motion, while the center-of-mass position can be considered as a driven variable.\cite{Tuckerman_2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$.
-In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if one makes $L = 6N$.
-In the particular (and rather common) situation in which one sets $\vt P = \vt 0$ at the beginning of a simulation, one must make $L = 6N - 3$ because all three elements of $\vt P$ will be conserved.\cite{Martyna_1994}
-
-\subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}
-\label{sec:numerical_solvers}
-
-As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t {\mathcal L}}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman_2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion.
-In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator.
-As a result, the whole propagator is symplectic and thus preserves the phase-space volume element.
-It is widely recognized that this property is important for numerical stability.\cite{Skeel_1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra_2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion.
-In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator preserves the same measure.
-As the author has shown,\cite{Ezra_2006} for a particular splitting ${\mathcal L} = \sum_k {\mathcal L}^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$.
-Here, $\dot{\vt x}^{(k)}$ refers to the system of equations obtained from the individual operator $i\!L^{(k)}$, which might display its own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$.
-By means of the product rule, it is straightforward to devise an equivalent but more meaningful criterion, which is
-\begin{equation}
-\label{eq:split_kappa_w}
-\kappa^{(k)} = {\mathcal L}^{(k)} w.
-\end{equation}
-
-Therefore, if the function $w(\vt x)$ that derives from Eq.~\eqref{eq:relation_kappa_w} also satisfies Eq.~\eqref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$.
-We are now in a position to prove that the integration procedure described in Sec.~\ref{sec:methodology} is measure-preserving.
-As mentioned in that section, the generalized Liouville operator can be written as ${\mathcal L} = \Liu{\tiny NVE} + {\mathcal L}_\text{\tiny NHC}$, where $\Liu{\tiny NVE}$ is the same operator given by Eq.~\eqref{eq:Lioville operator} and
-\begin{equation}
-\label{eq:iL_NHC}
-\begin{split}
-{\mathcal L}_\text{\tiny NHC} = &-\alpha_1 \left( \tr{\vt p} \diff{}{\vt p} + \tr{\vt \pi} \diff{}{\vt \pi}\right) + \sum_{j=1}^{M} \alpha_j \diff{}{\eta_j} \\
-&+ \sum_{j=1}^{M} (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}.
-\end{split}
-\end{equation}
-
-Then, by retaining the NVE factorization presented in Eq.~\eqref{eq:splitting NVE}, we write the full exponential operator for each time step as in Eq.~\eqref{eq:splitting NVT}.
-The Hamiltonian propagators introduced in Sec.~\ref{eq:hamiltonian dynamics} already satisfy the criterion in Eq.~\eqref{eq:split_kappa_w} with the function $w$ expressed in Eq.~\eqref{eq:nhc_measure} (they are all incompressible and keep $w$ unchanged).
-The NHC propagator is further split into factors that also satisfy Eq.~\eqref{eq:split_kappa_w}, which are given by
-\begin{equation*}
-\begin{split}
-&{\mathcal L}_0 = -\alpha_1 \left( \tr{\vt p} \diff{}{\vt p} + \tr{\vt \pi} \diff{}{\vt \pi}\right) + \alpha_1 \diff{}{\eta_1}, \\
-&{\mathcal L}_j = (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}}, \label{eq:iL_NHC_j} \\
-&{\mathcal L}_M = G_M \diff{}{p_{\eta_M}},
-\end{split}
-\end{equation*}
-with $j$ varying from $1$ to $M-1$.
-The action of the propagator $e^{t {\mathcal L}_0}$ is the solution of the equations $\dot{\vt p} = -\alpha_1 \vt p$ and $\dot{\vt \pi} = -\alpha_1 \vt \pi$ and $\dot{\eta}_1 = \alpha_1$, with all other variables remaining constant at their initial values.
-This results in the scaling operations described in Sec.~\ref{sec:methodology}.
-In turn, the propagator $e^{t {\mathcal L}_j}$ behaves according to the equations $\dot{p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}$ and $\dot{\eta}_{j+1} = \alpha_{j+1}$, with all variables other than $p_{\eta_j}$ and $\eta_{j+1}$ kept unchanged.
-Finally, the propagator $e^{t {\mathcal L}_M}$ involves a unique meaningful differential equation $\dot{p}_{\eta_M} = G_M$.
-
-We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj_2005} does not fulfill this criterion.
-
-We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
-Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
-However, as will be shown in the numerical results of Sec.~\ref{sec:numerical_results},
-
-\subsection{Backward Error Analysis}
-\label{sec:backward error analysis}
 
@@
 
-The first new result we present here is an analytical form for the shadow Hamiltonian concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~\citenum{Silveira_2017}.
-A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and we summarize the final results as follows.
-First of all, we rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
-\begin{equation*}
-K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation*}
-where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
-\begin{equation*}
-{\mt \Omega} = \left[\begin{array}{cc}
-{\mt M}^{-1} & \mt 0 \\
-\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
-\end{array}\right].
-\end{equation*}
+In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies.
 
-With $\nabla^2 U$ being the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, we define a modified kinetic energy $\modified K$ as
-\begin{equation*}
-\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation*}
-where
-\begin{equation*}
-\modified{\mathbf \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} {\mt \Omega} (\nabla^2 U) \tr{\mt \Omega}.
-\end{equation*}
-
-Although $\modified{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
-As a result, the modified kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
-
-In addition to $\modified K$, we also define a modified potential energy $\modified U$ as
-\begin{equation*}
-\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
-\end{equation*}
-
-Finally, the shadow Hamiltonian for a systems of rigid body becomes
-\begin{equation}
-\modified{\Ham{}} = \modified K + \modified U + \mathcal{O}(h^4).
-\end{equation}
-
-In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies.
-The Hamiltonian is:
-\begin{equation*}
-\Ham{}(\vt r, \vt p, \vt q, \vt \pi) = K(\vt p, \vt q, \vt \pi) + U(\vt r, \vt q),
-\end{equation*}
-where
-\begin{equation*}
-K = \frac{1}{2} \tr{\vt p}{\mt M}^{-1}{\vt p} + \frac{1}{8}\tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi,
-\end{equation*}
-with
-\begin{equation*}
-\mt B(\vt q) = \left[\begin{array}{rrrr}
--q_2 & -q_3 & -q_4 \\
-q_1 & -q_4 &  q_3 \\
-q_4 &  q_1 & -q_2 \\
--q_3 &  q_2 &  q_1
-\end{array}\right].
-\end{equation*}
-
-If a splitting is done so that $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$, then from Eq.~\eqref{eq:general Lioville operator} we have that $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$.
-Therefore, $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$.
-This makes
+For the splitting introduced in Section~\ref{sec:nve}, $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$ with the corresponding Liouviller operators $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$, with $K_{x}$ and $U_{x}$ being the gradient of the kinetic and potential energies, respectively.
+We first obtain $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$, and we find that
 \begin{align*}
@@
 \end{align*}
-where $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi = -\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt \pi) \vt q$ is the vector of angular velocities.\cite{Silveira_2017} Hence,
+where $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi = -\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt \pi) \vt q$.\cite{Silveira_2017} Hence,
 \begin{gather*}
@@
 \begin{equation*}
-\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega}.
+\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} = 0,
 \end{equation*}
-
+which is a null term as it corresponds to the quadratic form of a skew-symmetric matrix.
 Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{k=1}^3 \omega_k \hat{\mt B}_k ) \vt q$, where each $\hat{\mt B}_k$ is a skew-symmetric permutation matrix (i.e.
@@
 \end{equation*}
-
-The second term in the right-hand side of the equation above is identically null whenever (a quadratic form whose matrix is skew-symmetric), which makes $\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = -\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q}$.
-Finally,
+Both terms in the right-hand side of the equation above are identically null. 
+Now, we are able to evaluate $\Liu A \Liu A \Ham B $, which is
 \begin{align*}
@@
 + \\
-&\frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega - \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}_b) {\mt I} {\vt \omega}.
+&\frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
 \end{align*}
-
-The last term above need to be double checked.
-
-For the second part:
+The remaining term to obtain $H_s$ is simply given by 
 \begin{align*}
@@
 
-Overall:
+As will become clear below, it is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
 \begin{equation*}
-\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
+K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\end{equation*}
+where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
+\begin{equation*}
+{\mt \Omega} = \left[\begin{array}{cc}
+{\mt M}^{-1} & \mt 0 \\
+\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
+\end{array}\right].
+\end{equation*}
+
+We now proceed to show the final expression of $H_s$, which results from grouping the terms corresponding to forces and torques in the \textit{modified} potential energy $\tilde{U}$, given by
+\begin{equation*}
+\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
+\end{equation*}
+while the remaining terms can be recast in a compact matrix, as follows
+\begin{equation*}
+\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
 \end{equation*}
@@
 \begin{equation*}
-\tilde{U} = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right)
-\end{equation*}
-and
-\begin{equation*}
-\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} + {\mt \Gamma} \right)
-\end{equation*}
-with
-\begin{equation*}
-\mt \Omega = \left[\begin{array}{cc}
-{\mt M}^{-1} & \mt 0 \\
-\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
-\end{array}\right],
-\end{equation*}
-\begin{equation*}
-\mt \Gamma = \left[\begin{array}{cc}
-\mt 0 & \mt 0 \\
-\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} {\mt S}({\mt I}^{-1} {\vt \tau}_b) \tr{\mt B}(\vt q)
-\end{array}\right],
+\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)
 \end{equation*}
@@
 
-\subsection{Generalized Equipartition}
-
-The important thing is that $\modified{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] = \modified{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
-In this way, the equipartition theorem (valid at least in this direction? something to do with integrating through momenta at least internally to the integral in coordinates?) states that
+Finally, we have that
 \begin{equation*}
-3N k_B T = \left\langle \tr{\vt p} \diff{\tilde K}{\vt p} \right\rangle = 2\langle \tilde K \rangle
+\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
 \end{equation*}
 
-Sampling validation: Shirts \cite{Shirts_2013}
+Although $\modified{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
+As a result, the modified kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
 
-\section{Results}
-\label{sec:numerical_results}
-
-%CÓDIGO: CHECAR FACTORIZACIÓN Y DELTA T DE LOS OPERADORES DE NHC
-
-In this section, we analyze the numerical integrators introduced in Subsection~\ref{sec:numerical_solvers}, and compare the results with those obtained employing the method of Kamberaj~\textit{et al.}\cite{Kamberaj_2005}, which is implemented in the LAMMPS\cite{Plimpton_1995} software package.
-For that, we considered a system of 900 TIP3P\cite{Jorgensen_1983} water molecules in a cubic box of length {\AA} subject to periodic boundary conditions.
-The intermolecular interactions were truncated at 9 \AA and to avoid a discontinuity in the force, we applied the so-called shifted-force (SF) method\cite{Allen_1989, Fennell_2006, Toxvaerd_2011} to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.
-
-It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
-First we present some results as first mented.
-That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
-In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
-By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
-%Resultados sobre o checkensemble:
-
-checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
-Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
-m-b distribution valid when equipartition of energy is satisfied.
-
-Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
-Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.
+\suppinfo
 
 \begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
-	\label{table:new_results}
-%	\begin{ruledtabular}
-		\begin{tabular}{cccc}
-			& \multicolumn{3}{c}{ Temperature (K) } \\
-			\cline{2-4}
-			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}.
-\\
-			\hline
-			1.0 & $300.01 \pm 0.02$ & $300.00 \pm 0.02$ & $299.52 \pm 0.02$ \\
-			2.0 & $300.01 \pm 0.03$ & $299.99 \pm 0.03$ & $298.24 \pm 0.03$ \\
-			3.0 & $299.97 \pm 0.04$ & $299.98 \pm 0.04$ & $295.94 \pm 0.04$
+	\caption{Effect of the time-step size on ....
+of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.}
+    \label{table:stability_martyna}
+    \begin{tabular}{SS[table-number-alignment = center,table-figures-exponent = 1]S[table-number-alignment = center,table-figures-exponent = 1]}
+     & \multicolumn{2}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}  \\
+     \cmidrule[0.5mm](lr){2-3} 
+     $\timestep$ $\text{(fs)}$  & $\modified{H}$  & $\modified{\Ham{}}$     \\
+	1 &  -1.6e-2  & 4.5e-3\\
+	2 & 9.5e-2 & 1.4e-1\\
+	3 & 2.8e-1 & 5.6e-2\\
+	4 & 1.1e0 & 5.6e-2\\
+	5 & 3.3e0 & -3.7e-2\\
+    6 & 6.1e1 & 6.9e-2 \\
+	7 & 9.8e2 & -3.6e-3 \\
+	8 & 1.1e4 & 1.8e-2 \\
+	9 & 7.5e4 & 3.8e-3 \\
+	10 & 3.3e5 & 1.9e-2  \\
 	\end{tabular}
-%	\end{ruledtabular}
 \end{table}
 
-\begin{table}
-	\caption{New results involving Kamberaj \text{et al}.\cite{Kamberaj_2005}}
-	\label{table:new_results_2}
-%	\begin{ruledtabular}
-		\begin{tabular}{cccc}
-			& \multicolumn{3}{c}{Drift rate [(kcal/mol)/ns]} \\
-			\cline{2-4}
-			$\timestep$/fs & Proposed & Alternative & Kamberaj \text{et al}.
-\\
-			\hline
-			1.0 & $0.0328$ & $0.108$ & $0.023$ \\
-			2.0 & $0.597$  & $1.31$  & $17.0$ \\
-			3.0 & $3.44$   & $2.53$  & $53.2$
-		\end{tabular}
-%	\end{ruledtabular}
-\end{table}
-
-\begin{table*}
-\caption{Results.....}
-\label{table:proposed_method}
-%\begin{ruledtabular}
-\begin{tabular}{cccccccc}
-$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
-
-\hline
-\multicolumn{8}{c}{Proposed Method} \\
-1.0 & $300.01\pm0.02$ & $-8465.1\pm0.5$ & $1614.2\pm0.1$ & $807.4\pm0.1$ & $806.8\pm0.1$ & $62\pm2$ & $0.0328$ \\
-2.0 & $300.01\pm0.03$ & $-8450.8\pm0.4$ & $1614.2\pm0.2$ & $809.4\pm0.1$ & $804.7\pm0.1$ & $73\pm2$ & $0.597$ \\
-3.0 & $299.97\pm0.04$ & $-8425.7\pm0.5$ & $1614.0\pm0.2$ & $812.8\pm0.1$ & $801.2\pm0.2$ & $90\pm2$ & $3.44$ \\
-4.0 & $299.93\pm0.04$ & $-8391.4\pm0.6$ & $1613.7\pm0.2$ & $817.1\pm0.2$ & $796.6\pm0.2$ & $119\pm2$ & $5.64$ \\
-
-\hline
-\multicolumn{8}{c}{Alternative Method} \\
-1.0 & $300.00\pm0.02$ & $-8465.5\pm0.5$ & $1614.1\pm0.1$ & $807.1\pm0.1$ & $807.0\pm0.1$ & $65\pm2$ & $0.108$ \\
-2.0 & $299.99\pm0.03$ & $-8450.5\pm0.5$ & $1614.1\pm0.2$ & $809.3\pm0.1$ & $804.7\pm0.1$ & $72\pm2$ & $1.31$ \\
-3.0 & $299.98\pm0.04$ & $-8425.9\pm0.5$ & $1614.0\pm0.2$ & $812.5\pm0.1$ & $801.5\pm0.2$ & $88\pm2$ & $2.53$ \\
-4.0 & $300.01\pm0.04$ & $-8390.4\pm0.6$ & $1614.2\pm0.2$ & $817.4\pm0.2$ & $796.8\pm0.2$ & $120\pm2$ & $7.31$ \\
-
-\hline
-\multicolumn{8}{c}{Method of Kamberaj \textit{et al}.\cite{Kamberaj_2005}} \\
-1.0 & $299.53\pm0.02$ & $-8470.9\pm0.5$ & $1611.6\pm0.1$ & $806.4\pm0.1$ & $805.2\pm0.1$ & $52\pm2$ & $0.023$ \\
-2.0 & $298.24\pm0.03$ & $-8471.2\pm0.5$ & $1604.6\pm0.2$ & $805.4\pm0.1$ & $799.2\pm0.1$ & $47\pm2$ & $17.0$ \\
-3.0 & $295.94\pm0.04$ & $-8480.9\pm0.6$ & $1592.3\pm0.2$ & $804.2\pm0.1$ & $788.1\pm0.1$ & $28\pm2$ & $53.2$ \\
-4.0 & $292.72\pm0.04$ & $-8494.7\pm0.7$ & $1575.0\pm0.2$ & $802.4\pm0.1$ & $772.5\pm0.2$ & $-1\pm2$ & $60.6$ \\
-
-\hline
-\multicolumn{8}{c}{Double-Thermostat Analogue of the Proposed Method} \\
-1.0 & $300.00\pm0.02$ & $-8466.0\pm0.5$ & $1614.1\pm0.1$ & $806.6\pm0.1$ & $807.5\pm0.1$ & $60\pm2$ & $6.42$\\
-2.0 & $299.98\pm0.03$ & $-8452.0\pm0.5$ & $1614.0\pm0.2$ & $807.1\pm0.1$ & $806.9\pm0.1$ & $71\pm2$ & $80.4$ \\
-3.0 & $300.16\pm0.04$ & $-8426.1\pm0.5$ & $1615.0\pm0.2$ & $808.4\pm0.1$ & $806.6\pm0.1$ & $90\pm2$ & $1.38\times10^4$ \\
-4.0 & $300.75\pm0.04$ & $-8385.3\pm0.6$ & $1618.2\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $125\pm2$ & $4.29\times10^4$ \\
-
-\hline
-\multicolumn{8}{c}{Double-Thermostat Analogue of the Alternative Method} \\
-1.0 & $299.98\pm0.02$ & $-8465.6\pm0.5$ & $1614.0\pm0.1$ & $806.6\pm0.1$ & $807.4\pm0.1$ & $56\pm2$ & $4.87$ \\
-2.0 & $299.96\pm0.03$ & $-8452.0\pm0.5$ & $1613.9\pm0.2$ & $807.1\pm0.1$ & $806.8\pm0.1$ & $71\pm2$ & $85.1$ \\
-3.0 & $300.20\pm0.04$ & $-8426.0\pm0.5$ & $1615.2\pm0.2$ & $808.5\pm0.1$ & $806.7\pm0.1$ & $89\pm2$ & $1.50\times10^4$ \\
-4.0 & $300.75\pm0.04$ & $-8384.7\pm0.6$ & $1618.1\pm0.2$ & $811.6\pm0.1$ & $806.6\pm0.2$ & $123\pm2$ & $4.33\times10^4$ \\
-
-\hline
-\multicolumn{8}{c}{Single-Thermostat Analogue of the Method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}} \\
-1.0 & $299.52\pm0.02$ & $-8470.2\pm0.5$ & $1611.6\pm0.1$ & $805.9\pm0.1$ & $805.6\pm0.1$ & $58\pm2$ & $0.272$ \\
-2.0 & $298.18\pm0.03$ & $-8471.8\pm0.5$ & $1604.3\pm0.2$ & $804.4\pm0.1$ & $799.9\pm0.1$ & $44\pm2$ & $0.390$ \\
-3.0 & $295.97\pm0.04$ & $-8474.6\pm0.5$ & $1592.4\pm0.2$ & $801.6\pm0.1$ & $790.8\pm0.2$ & $29\pm2$ & $3.00$ \\
-4.0 & $292.65\pm0.04$ & $-8478.7\pm0.6$ & $1574.6\pm0.2$ & $797.6\pm0.2$ & $777.0\pm0.2$ & $10\pm2$ & $8.12$ \\
-
-\end{tabular}
-%\end{ruledtabular}
-\end{table*}
-
-\begin{figure}
-    \caption{SHAKE: $\timestep$ = 1 fs ($\bar{T}$ = 299.98 K, $\bar{U}$ = -8428.86 kcal/mol) HMC: $\bar{U}$ = -8470.71 kcal/mol  }
-	\includegraphics{potenergy}
-\end{figure}
-
-\begin{figure}
-    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 300.04 K, $\bar{U}$ = -8465.63 kcal/mol)}
-	\includegraphics{potenergy2}
-\end{figure}
-
-\begin{figure}
-    \caption{$\timestep$ = 0.5 fs ($\bar{T}$ = 299.99 K, $\bar{U}$ = -8468.19 kcal/mol)}
-	\includegraphics{potenergy_05}
-\end{figure}
-
-\begin{figure}
-    \caption{$\timestep$ = 1 fs ($\bar{T}$ = 299.52 K, $\bar{U}$ = -8470.21 kcal/mol)}
-	\includegraphics{potenergy_kamb}
-\end{figure}
-
-In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
-In figure legends we show the averages temperatures simulated.
-It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.
-
-\begin{figure}
-	\includegraphics{checkensemble}
-	\caption{Linear plot of the log ratio probabilities with the true slope (solid line) and the measured slope from the linear regression analysis (symbols).
-Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}_1$ = 296.03~K, $\bar{T}_2$ = 304.99~K.}
-	\label{fig:checkensemble}
-\end{figure}
-
-\begin{figure}
-	\includegraphics{maxwell-botlzamm-paper}
-	\caption{Probability distribution of the kinetic energy with sampled values and those calculated with Eq.~\eqref{eq:mb}.
-Simulation of liquid water using Eq.~\eqref{eq:trotter_splitting_NHC}, $\timestep$ = 1~fs, $\bar{T}$ = 304.99~K.}
-\label{fig:mbdistribution}
-\end{figure}
-
-Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
-It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
-This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
-
-\begin{table}
-	\begin{threeparttable}
-		\caption{Ensemble Validation of NVT algorithms for liquid water using the Maximum Likelihood Approach \tnote{a} \tnote{b}}
-		\label{table:ensemblevalidation}
-%		\begin{ruledtabular}
-			\begin{tabular}{ccccc}
-				& \multicolumn{4}{c}{true $\timestep$ = 8.970 K} \\
-				\cline{2-5}
-				& \multicolumn{2}{c}{potential} & \multicolumn{2}{c}{kinetic}\\
-				\hline
-				Method  &$\timestep/K$ & $\sigma$ dev & $\timestep/K$ & $\sigma$ dev \\
-				\hline % inserts single-line 
-				Eq.~\eqref{eq:trotter_splitting_NHC} & 9.022 $\pm$ 0.117 & 0.57 & 8.928 $\pm$ 0.051 & 0.63 \\
-				Eq.~\eqref{eq:modified_splitting}    & 8.965 $\pm$ 0.101 & 0.05 & 8.967 $\pm$ 0.052 & 0.14
-		\end{tabular}
-%		\end{ruledtabular}
-		\begin{tablenotes}
-			\item[a] The standard deviation of the mean temperature for all cases is $\sigma$ = 0.03.
-			\item[b] $\bar{T}_1$ = 296.03~K and $\bar{T}_2$ = 304.99~K.
-		\end{tablenotes}
-	\end{threeparttable}
-\end{table}
-
-%\begin{center}
-%  \label{fig:pressure}
-%  \includegraphics{pressure2}
-%  \captionof{figure}{Histograms of pressure obtained in this work (filled bars) and using LAMMPS %(unfilled bars).
-%Simulation of liquid water with NVE integration.
-%$\timestep : 1 fs$  }
-%\end{center}
-
-\section{Concluding Remarks}
-\label{sec:conclusion}
-Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.
-
-by contrast
-
-\begin{acknowledgement}
-	The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
-\end{acknowledgement}
-
-\appendix
-
 \bibliography{rigid_bodies}
-%\begin{table*}
-%	\caption{Results with Exact Solution for Free Rotations}
-%	\label{table:NVE}
-%	\begin{ruledtabular}
-%		\begin{tabular}{cccccccc}
-%			$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
-%			\hline
-%			1.0 & $	305.91	\pm	0.03	$ & $	-8395.1	\pm	0.1	$ & $	1645.9	\pm	0.1	$ & $	823.2	\pm	0.1	$ & $	822.7	\pm	0.1	$ & $	146	\pm	2	$ & $0.0456$ \\
-%			2.0 & $	305.21	\pm	0.02	$ & $	-8389.5	\pm	0.1	$ & $	1642.1	\pm	0.1	$ & $	823.0	\pm	0.1	$ & $	819.1	\pm	0.1	$ & $	149	\pm	2	$ & $0.656$ \\
-%			3.0 & $	304.30	\pm	0.03	$ & $	-8375.4	\pm	0.9	$ & $	1637.2	\pm	0.1	$ & $	824.0	\pm	0.1	$ & $	813.3	\pm	0.1	$ & $	156	\pm	2	$ & $2.87$ \\
-%			4.0 & $	304.5	\pm	0.5	$ & $	-8339	\pm	6	$ & $	1638	\pm	3	$ & $	829	\pm	1	$ & $	808.8	\pm	0.6	$ & $	180	\pm	2	$ & $10.1$ \\
-%		\end{tabular}
-%	\end{ruledtabular}
-%\end{table*}
-
-%\begin{table*}
-%\caption{Results.....}
-%\label{table:NVE}
-%\begin{ruledtabular}
-%\begin{tabular}{cccccccc}
-%$\timestep$ (fs) & $T$ (K) & $\langle U\rangle$ (kcal/mol) & $\langle K\rangle$ (kcal/mol) & $\langle K_t\rangle$ (kcal/mol) & $\langle K_r\rangle$ (kcal/mol) & P (atm) & $R$ (kcal/mol.ns) \\
-%\hline
-%1.0 & $305.95\pm0.03$ & $-8395.4\pm0.1$ & $1646.1\pm0.1$ & $823.3\pm0.1$ & $822.9\pm0.1$ & $151\pm2$ & $0.0471$ \\
-%2.0 & $305.17\pm0.03$ & $-8389.0\pm0.1$ & $1641.9\pm0.1$ & $823.2\pm0.1$ & $818.8\pm0.1$ & $148\pm2$ & $0.853$ \\
-%3.0 & $304.34\pm0.03$ & $-8374\pm2$ & $1637.5\pm0.2$ & $824.0\pm0.1$ & $813.4\pm0.1$ & $152\pm2$ & $3.96$ \\
-%4.0 & $303.6\pm0.4$ & $-8349\pm5$ & $1633\pm2$ & $826.8\pm0.2$ & $806.5\pm0.4$ & $168\pm2$ & $8.25$ \\
-%\end{tabular}
-%\end{ruledtabular}
-%\end{table*}
-%\begin{table}
-%	\begin{threeparttable}
-%		\caption{Average deviation of the conserved quantity ($DE$) and average temperatures from liquid water simulation with NVT integration \tnote{a}\tnote{b}}
-%		\label{table:denvt}
-%		\begin{ruledtabular}
-%			\begin{tabular}{ccccc}
-%				& \multicolumn{2}{c}{Eq.~\eqref{eq:trotter_splitting_NHC}} & \multicolumn{2}{c}{Eq.~%\eqref{eq:modified_splitting}} \\
-%				\cline{2-5}
-%				$\timestep$/fs & $D\!E$ & $T$/K & $D\!E$ & $T$/K \\
-%				\hline
-%				1.0 & 0.000043 & 298.16  & 0.000052  & 298.15 \\
-%				2.0 & 0.00013  & 298.23  & 0.00017   & 298.06 \\
-%				3.0 & 0.00089  & 298.17  & 0.0013    & 298.12
-%			\end{tabular}
-%		\end{ruledtabular}
-%		\begin{tablenotes}
-%			\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$.
-%			\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs.
-%		\end{tablenotes}
-%	\end{threeparttable}
-%\end{table}
-
-%The Nos\'e-Hoover chain propagator is further split by making $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_%\text{NHC}^{(j)}$ and, afterwards,
-%\begin{equation*}
-%e^{\frac{\timestep}{2} i\!L_\text{NHC}} = \Pi_{j=1}^n \Bigg[ \prod_{j=M}^1 e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg] e^{\frac{\timestep}{2} i\!L_\text{NHC}^{(0)} } \Bigg[ \prod_{j=1}^M e^{\frac{\timestep}{4} i\!L_\text{NHC}^{(j)} } \Bigg].
-%\end{equation*}
-
-%In order to satisfy the criterion in Eq.~\eqref{eq:split_kappa_w} for all factors, given the function $w$ expressed in Eq.~\eqref{eq:nhc_measure}, we further split the NHC contribution as $i\!L_\text{NHC} = \sum_{j=0}^M i\!L_\text{NHC}^{(j)}$, where
-
-%\begin{subequations}
-%\label{eq:solution_momenta}
-%\begin{align}
-%&{\vt p}_i(t) = {\vt p}_i^0 + \left({\vt F}_i - \alpha_1 {\vt p}_i^0 \right) \phi\left(\alpha_1 t \right) %t \quad \text{and} \label{eq:solution_p} \\
-%&{\vt \pi}_i(t) = {\vt \pi}_i^0 + \left(2 {\mt C}_i{\vt \tau}_i - \alpha_1 {\vt \pi}_i^0 \right) \phi%\left(\alpha_1 t \right) t.
-\label{eq:solution_pi}
-%\end{align}
-%\end{subequations}
-
-%By evaluating these functions simultaneously for all $i$, we accomplish the effect of a propagator $e^{t i%\!L_b^\ast}$, where $i\!L_b^\ast$ is the Liouville operator given by
-%\begin{equation*}
-%i\!L_b^\ast = \sum_{i=1}^N \left[ \left( \tr{\vt F}_i - \alpha_1 \tr{\vt p}_i \right) \diff{}{\vt p_i} + \left(2 \tr{\vt \tau}_i \tr{\mt C}_i - \alpha_1 \tr{\vt \pi}_i \right) \diff{}{\vt \pi_i} \right].
-%\end{equation*}
-%We also define a modified Nos\'e-Hoover chain propagator $e^{i\!L^\ast_\text{NHC}}$, as follows:
-%\begin{equation}
-%\begin{split}
-%e^{\frac{\timestep}{2} i\!L_\text{NHC}^\ast } = &\prod_{j=M}^1 \exp\left[\frac{\timestep}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
-%\times &\exp\left(-\frac{\timestep}{2} \sum_{j=1}^M \alpha_j \diff{}{\eta_j}\right) \\
-%\times &\prod_{j=1}^M \exp\left[\frac{\timestep}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right]
-%\end{split}
-%\end{equation}
-
-%In order to evaluate the stability of the numerical integrators, we employ a measure given by%\cite{Tuckerman_2010}
-%\begin{equation}
-%\label{eq:performance}
-%D\!E =  \frac{1}{\mathcal N} \sum_{t=1}^{\mathcal N} \left| \frac{E_t - E_0}{E_0} \right|,
-%\end{equation}
-%from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ %(either $\mathcal{H}$ of Eq.~\eqref{eq:H_NVE} or $H$ of Eq.~\eqref{eq:H_nvt}) from its initial value $E_0$ %after a certain number of time steps $\mathcal N$.
-% \textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\timestep/4)$, we factorized those operators analogously to the scheme given by Eq.~\eqref{eq:splitting_rot}}.
-
-%\begin{equation}
-%\label{eq:probability_ratio_pot}
-%\frac{P_{pot}(E/\beta_2)}{P_{pot}(E/\beta_1)} = \frac{Q_{pot}(\beta_1)}{Q_{pot}(\beta_2)}  \exp\left[(-\beta_2 - \beta_1)E_{pot}\right]
-%\end{equation}
-
-%\begin{equation}
-%\label{eq:probability_ratio_kin}
-%\frac{P_{kin}(E/\beta_2)}{P_{kin}(E/\beta_1)} = \frac{Q_{kin}(\beta_1)}{Q_{kin}(\beta_2)}  \exp\left[(-%\beta_2 - \beta_1)E_{kin}\right]
-%\end{equation}
-
-%The equation above avoids the introduction of a ficticious angular velocity component, as required in the formulation of Miller~\textit{et al.}\cite{Miller_2002} An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.
 \end{document}

### 2018-05-28 rigid_bodies_supp_info.tex
added supplementary material

--- 2018-05-28 rigid_bodies.tex
@@
 journal=jctcce,
-layout=twocolumn
-]{achemso}
+manuscript=suppinfo]{achemso}
 
@@
 \begin{document}
-
-\begin{tocentry}
-	Graphical Abstract
-\end{tocentry}
-
-\begin{abstract}
-	Abstract.
-\end{abstract}
-
-\section{Introduction}
-
-Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman_2010} involves several types of approximations in order to mimic ``real'' physical systems.
-Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
-However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour.
-For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
-That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
-
-The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski_2012, Patra_2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj_2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton_1995} has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
-
-The Nos\'{e}-Hoover chain thermostat,\cite{Martyna_1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
-The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible.
-Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}
-
-In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble.
-The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
-Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
-The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj_2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy.
-We also include results from simulations using the conventional SHAKE\cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
-
-backward error analysis
-
-The paper is organized as follows.
-In Sec.~\ref{sec:theory} we briefly review the equations of motion in the NVE ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble.
-In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
-Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
-
-\section{Methodology}
-\label{sec:methodology}
-
-\subsection{NVE Dynamics: Notation and Symplectic Integration}
-\label{sec:nve}
-In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
-1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and
-2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002}, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.
-
-The Hamiltonian system of ordinary differential equations (ODE) that describes a rigid-body motion is given by \cite{Silveira_2017}
-\begin{subequations}
-	\label{eq:ODE system for NVE}
-	\begin{align}
-%
-	&\dot{\vt r} =
-	%+\diff{\Ham{}}{\vt p} =
-	{\mt M}^{-1} {\vt p}, \\
-%
-	&\dot{\vt p} =
-	%-\diff{\Ham{}}{\vt r} = 
-	{\vt F}, \\
-%
-	&\dot{\vt q} =
-	%+\diff{\Ham{}}{\vt \pi} =
-	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
-%
-	&\dot{\vt \pi} =
-	%-\diff{\Ham{}}{\vt q} =
-	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
-	\end{align}
-\end{subequations}
-with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$.
-In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugated momentum of such quaternion.
-Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity in the body-fixed frame.
-Matrices $\mt M$ and $\mt I$ are diagonal ones.
-The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
-Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
-\begin{equation*}
-\label{eq:def_B_and_C}
-\mt B(\vt q) = \begin{smallarray}{rrrr}
--q_2 & -q_3 & -q_4 \\
- q_1 & -q_4 &  q_3 \\
- q_4 &  q_1 & -q_2 \\
--q_3 &  q_2 &  q_1
-\end{smallarray}
-%
-\; \text{and} \;
-%
-\mt C(\vt q) = \begin{smallarray}{rrrr}
--q_2 & -q_3 & -q_4 \\
- q_1 &  q_4 & -q_3 \\
--q_4 &  q_1 &  q_2 \\
- q_3 & -q_2 &  q_1
-\end{smallarray}.
-\end{equation*}
-
-The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, with ${\mt A}(\vt q)$ being the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames).
-
-The Hamiltonian conserved by Eq.~\eqref{eq:ODE system for NVE} is
-\begin{equation}
-\label{eq:Hamiltonian}
-\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
-\end{equation}
-where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
-While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
-\begin{equation}
-\label{eq:kinetic energy}
-K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
-\end{equation}
-%An advantage of this notation is that
-Of course, Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
-The same occurs for diagonal matrices $\mt M$ and $\mt I$, whose sizes increase to $3N \times 3N$ in this case.
-In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
-Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further to accommodate a number of freely moving atoms.
-
-A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$, which is
-\begin{equation}
-\label{eq:Lioville operator}
-\begin{split}
-\Liu{\tiny NVE} &= \tr{\vt p} {\mt M}^{-1} \diff{}{\vt r} + \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) \diff{}{\vt q} + \tr{\vt F} \diff{}{\vt p} \\
-&+ \left[ \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt \pi) + 2 \tr{\vt \tau} \tr{\mt C}(\vt q) \right] \diff{}{\vt \pi}.
-\end{split}
-\end{equation}
-
-A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
-\begin{equation}
-\label{eq:splitting NVE}
-e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
-\end{equation}
-where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
-In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state immediately before the propagation occurs.
-The action of $e^{\timestep \Liu{A}}$ is the simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step.
-For this rotational part of the integration, in Ref.~\cite{Silveira_2017} we derived a general solution in a form that facilitates computer implementation and allows, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
-While the approximate solutions, such as the NO-SQUISH \cite{Miller_2002} method, only involve sine and cosine terms, the exact solution comprise ``less common'' functions, such as the Jacobi elliptic functions and Carlson integrals.
-This contributes to the fact that the approximate solution is more commonly used \cite{vanZon2007}. 
-For mathematical details, the reader is referred to our previos work \cite{Silveira_2017}, where we have verified that in liquid-phase simulations there is no significan gain in accuracy by using the exact solution in substitution to the NO-SQUISH as part of a symplectic integration scheme.
-However, in the NO-SQUISH method an overall rotation is given by a sequence of uniform rotations around the principal axes of a body, which results in a factorization of the rotational operator given by $ e^{\frac{h}{2} i\!L_3} e^{\frac{h}{2} i\!L_2} e^{h i\!L_1} e^{\frac{h}{2} i\!L_2} e^{\frac{h}{2} i\!L_3}$, with each factor representing a rotation around a given axe.
-The analytical solution, on the other hand, entails a single operator, thus making the application of backward error analysis considerably simpler.
-
-\subsection{NVT Dynamics with a Single Thermostat Chain}
-
-Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble.
-For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system.
-This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement.
-To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$.
-The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
-\begin{equation}
-\label{eq:nvt extended energy}
-H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
-\end{equation}
-where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
-As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.
-
-Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010_2} Invaluable tools have been devised for the development and analysis of extended phase space methods.
-By employing the method of Sergi and Ferrario \cite{Sergi_2001}, we obtain the equations of motion for the NVT dynamics with a single thermostat chain, which are
-\begin{subequations}
-	\label{eq:ODE system for NVT}
-	\begin{align}
-%
-\label{eq:nhc_r}
-	&\dot{\vt r} =
-	{\mt M}^{-1} {\vt p}, \\
-%
-\label{eq:nhc_p} 
-	&\dot{\vt p} =
-	{\vt F} - \alpha_1 {\vt p},\\
-%
-\label{eq:nhc_q}
-	&\dot{\vt q} =
-	\frac{1}{2} \mt B(\vt q) \vt \omega, \\
-%
-\label{eq:nhc_pi}
-	&\dot{\vt \pi} =
-	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
-%
-\label{eq:nhc_eta}
-	&\dot{\eta}_j = \alpha_j, \text{ and} \\
-%
-\label{eq:nhc_p_eta}
-	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M.
-	\end{align}
-\end{subequations}
-
-In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
-\begin{subequations}
-\begin{align}
-\label{eq:generalized force NVT}
-&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
-&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
-\end{align}
-\end{subequations}
-
-With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
-It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but are often computed for checking how well a numerical integrator conserves $H$.
-Analysis of the probability distribution generated by Eq.~\eqref{eq:ODE system for NVT} can be done through the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables eliminated.
-If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are integrals of motion, while the system's center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates.
-Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
-In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if one makes $L = 6N$.
-In the particular situation in which one sets $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of a simulation, one must make $L = 6N - 3$ \cite{Martyna_1994}.
-
-\subsection{Measure-Preserving Integration for the NVT Dynamics}
-\label{sec:mttk}
-As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
-For that, we employed the systematic method introduced by Ezra \cite{Ezra_2006} for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion, which is
-In short, a splitting scheme will preserve a given measure if each individual propagator does it as well. 
-It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure is given by
-\begin{equation}
-\label{eq:nhc_measure}
-w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
-\end{equation}
-
-will preserve a given measure $e^{-w(\vt x)}d\vt x$
-
-By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
-non-Hamiltonian) Liouville operator, the splitting goes as
-\begin{equation}
-\label{eq:splitting NVT}
-e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
-\end{equation}
-where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
-Of course, the NVE part can be integrated exactly as described before.
-The NHC part, in turn, can be split even further as
-\begin{equation*}
-e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
-\end{equation*}
-
-In the scheme above, the first propagator to act is $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, which promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
-Then, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ act sequentially, with $j$ varying from $M-1$ down to $1$.
-With $\phi(x) = \frac{1-e^{-x}}{x}$, the effect of each one is translated as \cite{Martyna_1996}
-\begin{align*}
-&p_{\eta_j} = p_{\eta_j}^\ast + \Big( G_j^\ast - \alpha_{j+1}^\ast p_{\eta_j}^\ast \Big) \phi\left(\frac{\alpha_{j+1}^\ast \timestep}{4n}\right) \frac{\timestep}{4n} \; \text{and} \\
-&\eta_{j+1} = \eta_{j+1}^\ast + \alpha_{j+1}^\ast \frac{\timestep}{4n}.
-\end{align*}
-
-After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ establishes the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of coordinate $\eta_1$.
-Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
-Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
-In the following, the NVT integrator introduced in the present section will be denoted as MTTK as its development was inspired by Ref.~\cite{Martyna_1996}.
-
-For having a low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort, making it feasible to increase the time step $\timestep$ without degrading the accuracy of the NHC part.
-In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
-A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
-The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
-Here we choose to take a simpler path.
-Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied deviations, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when computing ensemble averages.
-
-\subsection{Backward Error Analysis}
-\label{sec:modified_h}
-
-In order to explain our proposal, we turn the attention again to the NVE case.
-A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
-Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian \cite{Hairer_2006}.
-In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors). 
-It is also known that $\Ham{S}$ explicity depends on the time step size $\timestep$.
-
-The first new result we present in this paper is an analytical form for the shadow Hamiltonian concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~ \citenum{Silveira_2017}.
-A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and here we summarize the final expressions.
-
-As shown in Appendix~\ref{sec:rigid body shadow hamiltonian}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are modified kinetic and potential energies.
-The latter is given by
-\begin{equation}
-\label{eq:modified potential energy}
-\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right),
-\end{equation}
-where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the resultant torque vector.
-Note that $\modified U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
-In turn, the modified kinetic energy can be expressed as
-\begin{equation}
-\label{eq:modified kinetic energy}
-\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation}
-where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
-\begin{equation*}
-{\mt \Omega} = \left[\begin{array}{cc}
-{\mt M}^{-1} & \mt 0 \\
-\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
-\end{array}\right]
-\end{equation*}
-and ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, given by
-\begin{equation*}
-{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
-U_{\vt r \vt r} & U_{\vt r \vt q} \\
-U_{\vt q \vt r} & U_{\vt q \vt q}
-\end{array}\right].
-\end{equation*}
-
-Hence, the trajectory generated by the splitting scheme of Eq.~\ref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
-\begin{equation*}
-\dot{\vt r} = \diff{\modified K}{\vt p}, \;
-\dot{\vt q} = \diff{\modified K}{\vt \pi}, \;
-\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}, \; \text{and} \;
-\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}.
-\end{equation*}
-
-The first two of these equations can be combined together so that we can use Eq.~\eqref{eq:modified kinetic energy} to obtain
-\begin{equation}
-\label{eq:shadow ODE system}
-\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
-\end{equation}
-
-Analytical calculation of $\modified U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
-In contrast, although it is possible to compute $\modified K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
-Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
-Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
-\begin{equation}
-\label{eq:modified kinetic energy estimator}
-\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
-\end{equation}
-
-For reasons that will become clear shortly, we employ an asymmetric, four-point stencil formula for estimating $\dot{\vt r}$ at a given instant $t$, which is
-\begin{equation*}
-\dot{\vt r}^{[3]}_t = \frac{{\vt r}_{t-2\timestep} - 6 {\vt r}_{t-\timestep} + 3 {\vt r}_t + 2 {\vt r}_{t+\timestep}}{6\timestep}.
-\end{equation*}
-
-In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each rigid body $i$, such as required for satisfying the unit-norm condition \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of\cite{Schay_1995}
-\begin{equation*}
-\dot{\vt q}^{[3]}_t = {\mt \Pi}({\vt q}_t) \frac{{\vt q}_{t-2\timestep} - 6 {\vt q}_{t-\timestep} + 3 {\vt q}_t + 2 {\vt q}_{t+\timestep}}{6\timestep},
-\end{equation*}
-where ${\mt \Pi}$ is a matrix-valued function given by either product ${\mt B}\tr{\mt B}$ or ${\mt C}\tr{\mt C}$ \cite{Silveira_2017}.
-In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.
-
-\subsection{Adjustment of the NVT Dynamics}
-\label{sec:smttk}
-Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
-Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \modified{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
-Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
-The only required modification is that $\modified K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
-By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
-\begin{equation}
-\label{eq:thermostat_force}
-G_1 = 2 \modified K - L k_\text{\tiny B} T.
-\end{equation}
-
-One instant in which $\modified K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
-Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
-Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
-This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
-Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
-Fortunately, only positions and orientations are required for the numerical differentiation.
-This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time.
-By incomplete we mean that the step is aborted once the positions and orientations have been updated.
-\begin{figure}
-\begin{tikzpicture}
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](3,.2) -- (3,1.5);
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
-\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
-\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
-\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
-\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
-\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
-\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
-\node at (0.0,-0.2)[below]{$r_{t-h}$};
-\node at (2.2,-0.2)[below]{$r_{t}$};
-\node at (4.4,-0.2)[below]{$r_{t+h}$};
-\node at (6.6,-0.2)[below]{$r_{t+2h}$};
-\path[->,thick] (1) edge [bend left=45] (2);
-\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
-\path[dashed,->,thick] (1) edge [bend right=45] (0);
-\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
-\path[dashed,->,thick] (2) edge [bend left=45] (3);
-\end{tikzpicture}
-\caption{Schematic representation}
-\label{fig:virtual}
-\end{figure}
-%\begin{figure}
-%\begin{tikzpicture}
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](2,.2) -- (2,1.5);
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
-%\node(0) at (0,0)[shape=circle,draw, minimum width=1cm, minimum height=1cm]        {$r_{t-h}$};
-%\node(1) at (2.5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_t$};
-%\node(2) at (5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_{t+h}$};
-%\node(3) at (7.5,0)[shape=circle,draw,minimum width=1cm, minimum height=1cm] {$r_{t+ 2h}$};
-%\node(4) at (3.75,1)[] {$e^{\timestep \Liu{\tiny NVE}}$};
-%\path [dashed,->]
-%  (1)      edge                 node [above]  {virtual}     (0)
-%  (2)      edge                 node [above] {virtual}     (3);
-%\path [->,thick] 
-%  (1)      edge                 node [above]  {}            (2);
-%\end{tikzpicture}
-%\caption{Schematic representation}
-%\label{fig:virtual}
-%\end{figure}
-
-A drawback of this approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
-As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
-In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.
-
-In addition, $\modified K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
-Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\modified K$ goes as simply as ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$.
-
-$\widetilde{H}$, but an ``extended energy'' given by \cite{Martyna_1992}
-\begin{equation}
-\label{eq:this extended energy}
-\widetilde{H} = \modified K + \modified U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
-\end{equation}
-
-Tolman's generalized equipartition theorem \cite{Uline_2008}
-\begin{equation*}
-\left\langle \tr{\vt p} \diff{\modified{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\modified{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T
-\end{equation*}
-
-Finally, reweighting:
-\begin{equation}
-\label{eq:reweighting}
-\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
-\end{equation}
-
-\section{Theoretical Background}
-
-We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj_2005} does not fulfill this criterion.
-
-We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
-Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
-However, as will be shown in the numerical results of Sec.~\ref{sec:numerical_results},
-
-\subsection{Generalized Equipartition}
-
-The important thing is that $\modified{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] = \modified{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
-In this way, the equipartition theorem (valid at least in this direction? something to do with integrating through momenta at least internally to the integral in coordinates?) states that
-\begin{equation}
-\label{eq:equipartition}
-3N k_B T = \left\langle \tr{\vt p} \diff{\tilde K}{\vt p} \right\rangle = 2\langle \tilde K \rangle
-\end{equation}
-
-Sampling validation: Shirts \cite{Shirts_2013}
-
-\section{Results}
-\label{sec:numerical_results}
-
-In the following, we analyze the performance of the proposed NVT integrators introduced in Secs.
-\ref{sec:mttk}-\ref{sec:smttk}, and  compare the results with those obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
-The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA}.\cite{Silveira_2017} 
-The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.
-
-Before showing our main results, it may be instructive first to give a picture of the problem in controlling the temperature when discretization errors are non-negligible. 
-The most direct way (the most senstive indicator of the presence of discretization errors) to determine the importance of these errors is by examining the departure of the ``modified Hamiltonian'' $\modified{\Ham{}}$ from the physical Hamiltonian $\Ham{}$\cite{Engle_2005}.
-Fortunately, with the analytical form of $\modified{\Ham{}}$ derived in this paper we are able to perform such an analysis for the NVE integrator which is essential, as will be shown below, to understand the results of the NVT methods.
-In Fig.~\ref{fig:nve} we depict the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different step sizes ($\timestep$).
-The numerical integrator correspons to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation. 
-As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
-It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
-As a result of employing such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ fs and $\timestep$ fs (add in conclusions that we should use high-order approximations).
-
-\begin{figure}
-	\includegraphics{Figures/NVE.eps}
-	\caption{Total energy $\Ham{}$ (solid lines) and modified Hamiltonian $\modified{\Ham{}}$ (dotted lines) of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVE MD simulations for different step sizes $\timestep$.
-The numerical integrator correspons to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation.}
-	\label{fig:nve}
-\end{figure}
-
-For our implementation for rigid bodies of the numerical scheme due to Martyna \textit{et al.} \cite{Martyna_1996} (Eq.~\ref{eq:splitting NVT}), in Fig.~\ref{fig:props_k_m} we depict the computed averages of temperature ($\langle T \rangle$) and potential energy per molecule ($\langle U \rangle$) as a function of $\timestep$.
-The specified temperature corresponds to T = 298 K.
-Fig.~\ref{fig:props_k_m} clearly shows that a suitable temperature control leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
-As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
-Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
-These observations motivated the development of the scheme of Sec.~\ref{sec:smttk}, whose promising results we show below.
-
-\begin{figure}
-	\includegraphics{Figures/thermodynamic_properties_intro.eps}
-    \caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
-	\label{fig:props_k_m}
-\end{figure}
-
-\subsection*{Numerical Stability}
-
-First, we report related the numerical stability of the NVT integrators.
-
-examine the numerical stability of the NVT integrators by simply monitoring the quantity expected to be conserved along the simulation in each case: $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for the proposed method, and $H$ (Eq.~\ref{eq:nvt extended energy}) for the integrators due to Martyna \textit{et al.} \cite{Martyna_1996} and Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-In Table~\ref{table:stability} we report the energy drift, denoted by $R$, as the the rate of change of $\widetilde{H}$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting between those variables.
-As it is already known that \cite{Davidchack_2009,Bond_2007}
-
-we plot these extended energies for different time step sizes.
-From Figs.~\ref{fig:num_stab}(a) and (b) which, respectively, correspond to the integrator proposed in this work and our implementation of the scheme due to Martyna \textit{et al.} \cite{Martyna_1996}, one notes that the energy appears to remain bounded for time step sizes up to 5 fs, and a continous drift takes place for $\timestep$ = 6 fs.
-The numerical stability is only modestly improved by our method.
-The result that stands out, because of its anomalous feature, corresponds to Fig.~\ref{fig:num_stab} (c), which was obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-Note that for each time step size, the extended energy continously increases until reaching a point at which the integrator becomes stable.
-It should be remarked that these results were double-checked with the LAMMPS software package \cite{Plimpton_1995}. 
-As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Davidchack \cite{Davidchack_2009} also implemented such a separate thermostatization and discovered the existance of energy flows from one thermostat to the other.
-To the best of our knowledge, the work of Davidchak \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-As can be seen in Fig.~\ref{fig:ther_energies}, where we depict the energy of each thermostat for different time step sizes $h$, that flux actually exists.
-Also, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, while the positive values are associated to the thermostat coupled to the rotational degrees of freedom, and thus the energy is flowing from translational to rotational degrees of freedom.
-Interestingly, Fig.~\ref{fig:ther_energies} suggests that a unique steady state is established, and the time to reach the steady state decreases with $h$.
-It should be noted that all the simulations were carried out considering $t_d = 100$ fs.
-As far as we know, this behavior has not been reported before.
-reformulate(The drift is localized within the thermostat variables and thus is not indicative of the non-stationary evolution within the system itself).
-
-\begin{table}
-	\caption{Effect of the time-step size on ....
-of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.}
-    \label{table:stability}
-    \begin{tabular}{SS[table-number-alignment = center,table-figures-exponent = 1]S[table-number-alignment = center,table-figures-exponent = 1]}
-     & \multicolumn{2}{c}{Drift rate [(kcal/mol)/ns]}  \\
-     \cmidrule[0.5mm](lr){2-3} 
-     $\timestep$ $\text{(fs)}$  & $\modified{H}$  & $\modified{\Ham{}}$   \\
-	1 &  4.6e-4  & 8.5e-2 \\
-	2 & -8.2e-3 & -4.3e-2 \\
-	3 & -3.2e-2 & 7.3e-2\\
-	4 & 8.1e-3 & -1.1e-1 \\
-	5 & 1.3e0 & -2.2e-2\\
-    6 & 4.4e1 & -3.4e-3 \\
-	7 & 8.8e2 & 2.2e-2 \\
-	8 & 9.8e3 & 7.6e-2 \\
-	9 & 7.1e4 & -2.5e-2 \\
-	10 & 3.2e5 & -3.2e-2  \\
-	\end{tabular}
-\end{table}
-
-\begin{figure}
-	\includegraphics{Figures/numerical_stability.eps}
-    \caption{Total extended energy of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
-step sizes, employing (a) the proposed method represented by Eq.
-; (b) our implementation for rigid bodies of the integrator due to Martyna \textit{et al.} \cite{Martyna_1996} and (c) the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.Energy of thermostats for 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
-step sizes, employing the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.}
-	\label{fig:num_stab}
-\end{figure}
-
-\subsection*{Effect of Discretization Errors on Equilibrium Thermodynamic Properties}
-
-We now turn our focus to discretization errors.
-In Fig.~\ref{fig:properties} we show the computed average of some thermodynamic properties as a function of $\timestep$.
-The simulations were carried out at a specified temperature of T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_v$).
-The temperature was calculated from the kinetic energy assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
-The lines are a guide for the eye. 
-Also, the horizontal dotted lines (black) help to visualize the departure of the computed averages with increasing $\timestep$.
-They correspond to the results obtained employing $h$ = 1 fs in the numerical scheme proposed in this paper, which from now on will be referred as ``This work''.
-In addition, we have included the results of applying our proposal but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
-Despite the non-equilibrium process ocurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}, because that is not the only problematic issue related to the method in question.
-Particularly, from Fig.~\ref{fig:properties} (a) we note that the simulated temperature decreases systematically with $\timestep$.
-This kind of behavior is observed in NVE simulations\cite{Davidchack_2009,Silveira_2017}, reflecting that the factorization employed in Kamberaj \textit{et al.} \cite{Kamberaj_2005} in not suitable. 
-As a result of this inefficient temperature control, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Fig.~\ref{fig:properties} (b) and (c)).
-From Fig.~\ref{fig:properties} it becomes clear that our method yields significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
-This is true also for the results corresponding to the use of the NO-SQUISH method\cite{Miller_2002} in the NVE part of the integrator.
-Mathematically, the derivation of the refined Hamiltonian was facilitated by the fewer number of operators associated with the exact solution of rotation.
-However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement.
-
-With the aim of highlighting the importance of reweighting, in Table~\ref{table:reweight} we present the calculated temperature and potential energy for each $\timestep$ before and after applying Eq.~\ref{eq:reweighting}.
-The ``refined'' temperature $\modified{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
-Of course, as $\modified{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
-Indeed, for time step sizes up to, the agreement is excelent.
-Note that in the abscence of reweighting, the calculated temperatures are notably influenced by $h$.
-The free energy difference between the sampled state $\modified{\Ham{}}$ and the state we would like to sample, represented by the physical Hamiltonian $\Ham{}$.
-As expected, the $\Delta F$ is small, as it should be
-Note that a small, still clearly influencing systematic error is present in our scheme.
-
-\begin{table*}
-	\caption{Effect of the time-step size on ....
-of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.
-uncertainties in free energy never exced 1e-4}
-    \label{table:reweight}
-    \sisetup{
-table-number-alignment = center
-}
-    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]S[separate-uncertainty,table-figures-uncertainty=2]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S[separate-uncertainty,table-figures-uncertainty= 1]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S}
-     & &  \multicolumn{2}{c}{no reweighting} & \multicolumn{2}{c}{after reweighting} & \\
-     \cmidrule[0.5mm](lr){3-4} \cmidrule[0.5mm](lr){5-6}
-     $\timestep$ & $\modified{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F$} \\
-        $\text{(fs)}$  & $\text{(K)}$ & $\text{(K)}$ & $\text{(kcal/mol)}$  & $\text{(K)}$ & $\text{(kcal/mol)}$ & $\text{(kcal/mol\,N)}$ \\
-	1 & 298.00(1) & 297.71(1) & -9.1015(3) & 298.00(1) & -9.1028(3) & -0.001 \\
-	2 & 297.99(1) & 296.82(1) & -9.0970(3)  & 297.98(1)  & -9.1024(3) & -0.006 \\
-	3 & 297.98(1) & 295.38(1) & -9.0885(3) & 297.92(1) & -9.1005(3) & -0.013 \\
-	4 & 298.01(1) & 293.49(1)& -9.0764(3) & 297.85(3) & -9.0969(3) & -0.022 \\
-	5 & 297.99(1)& 291.13(1)& -9.0577(3) & 297.6(2) & -9.091(1) & -0.033 \\
-    6 & 298.00(1) & 288.46(1) & -9.0314(3) & 298.8(3) & -9.075(3) &  -0.045 \\
-	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
-
-	\end{tabular}
-\end{table*}
-
-To end this section,  Kinetic Energy Partition
-
-\begin{figure*}
-	\includegraphics{Figures/thermodynamic_properties.eps}
-	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
-	\label{fig:properties}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics{Figures/energy_partition.eps}
-\end{figure*}
-
-\begin{table*}
-	\caption{Effect of the time-step size on ....
-of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes....}
-	\label{table:check_ens}
-	\sisetup{table-number-alignment = center}
-    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]SS[separate-uncertainty,table-figures-uncertainty= 1]SS[separate-uncertainty,table-figures-uncertainty= 1]S}
-	{$\timestep$} & {estimated $\Delta T$} & {$\sigma$ deviations}  & {estimated $\Delta T$}  & {$\sigma$ deviations}  &    {estimated $\Delta T$}  & {$\sigma$ deviations} \\
-	{(fs)} & {(K)} &   & {(K)}  &  & {(K)}  & \\
-			\hline
-			\multicolumn{7}{c}{``This work (Unsplit)''}\\
-1.0	& 3.02(2) & 1.07 & & & &  \\
-2.0 & 3.02(2) & 0.81 & & & &  \\
-3.0 & 2.98(2) & 0.73 & & & &  \\
-4.0 & 3.01(2) & 0.41 & & & &  \\
-5.0 & 3.02(2) & 0.79 & & & &  \\
-6.0 & 3.03(3) & 1.23 & & & &  \\
-7.0 & 2.98(2) & 0.74 & & & &  \\
-            \hline
-			\multicolumn{7}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}\\
-1.0	& 2.98(2) & 0.66 & 3.01(2) & 0.49 & 2.96(3) & 0.90 \\
-2.0 & 2.98(2) & 0.73 & 2.99(2) & 0.03 & 3.02(3) & 0.66 \\
-3.0 & 2.95(2) & 2.33 & 3.01(2) & 0.65 & 2.91(3) & 2.47 \\
-4.0 & 3.00(2) & 0.01 & 3.01(2) & 0.72 & 2.92(3) & 2.19  \\
-5.0 & 2.93(2) & 3.05 & 3.01(2) & 0.70 & 2.94(3) & 1.58  \\
-6.0 & 2.92(2) & 3.87 & 3.00(2) & 0.22 & 2.91(3) & 2.76  \\
-7.0 & 2.89(2) & 5.35 & 3.02(1) & 1.33 & 2.87(2) & 4.33  \\
-            \hline
-			\multicolumn{7}{c}{Kamberaj \textit{et al.} \cite{Kamberaj_2005}}\\
-1.0	& 2.97(2) & 1.16 & 2.99(2) & 0.16 & 2.89(3) & 2.84  \\
-2.0 & 3.00(2) & 0.02 & 3.02(2) & 0.99 & 2.00(4) & 0.25  \\
-3.0 & 2.99(4) & 0.18 & 3.08(4) & 4.09 & 3.1(1) & 1.17  \\
-4.0 & 2.92(8) & 0.99 & 3.12(2) & 6.01 & 2.8(2) & 0.95  \\
-5.0 & 3.02(5) & 0.40 & 3.2(2)  & 9.50 &  2.9(1) & 0.28  \\
-6.0 & 3.14(5) & 2.69 & 3.29(2) & 14.68 & 3.21(7) & 2.90  \\
-7.0 & 3.03(4) & 0.73 & 3.35(2) & 17.37 & 2.99(6) & 0.16  \\
-    \end{tabular}
-\end{table*}
-%It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
-%First we present some results as first mented.
-%That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
-%In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
-%By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
-%Resultados sobre o checkensemble:
-
-%checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
-%Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
-%m-b distribution valid when equipartition of energy is satisfied.
-
-%Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
-%Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.
-
-%In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
-%In figure legends we show the averages temperatures simulated.
-%It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-%Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.
-
-%Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
-%It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
-%This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
-
-\section{Concluding Remarks}
-\label{sec:conclusion}
-Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.
-
-by contrast
-
-\begin{acknowledgement}
-	The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
-\end{acknowledgement}
-
-\appendix
-
-\section{Backward Error Analysis}
-\label{sec:rigid body shadow hamiltonian}
-
-For two non-commutative Liouville operators $\Liu A$ and $\Liu B$ and a real constant $\timestep$, the symmetric Baker-Campbell-Hausdorff (BCH) formula can be expressed as \cite{Hairer_2006}
-\begin{equation}
-\label{eq:symmetric BCH}
-\begin{split}
-&e^{\frac{\timestep}{2} \Liu B} e^{\timestep \Liu A} e^{\frac{\timestep}{2} \Liu B} = \\
-&= e^{\timestep (\Liu A + \Liu B) + \frac{\timestep^3}{24} \left[2 \Liu A + \Liu B,[\Liu A,\Liu B]\right] + \mathcal{O}(\timestep^5)},
-\end{split}
-\end{equation}
-where $[X,Y]$ is the commutator of operators $X$ and $Y$, defined as $[X,Y] = XY - YX$.
-The remaining terms of the infinite series involve ever-deepening, nested commutators \cite{Hairer_2006}.
-
-Let us now represent the operators $\Liu A$ and $\Liu B$ using Poisson brackets involving Hamiltonians $\Ham A$ and $\Ham B$, that is, $\Liu A = \{\circ,\Ham A\}$ and $\Liu B = \{\circ,\Ham B\}$.
-By employing the properties of anti-commutativity and Jacobi identity \cite{Hairer_2006}, we can deduce that
-\begin{align*}
-[\Liu A,\Liu B] &= \{\{\circ,\Ham B\},\Ham A\} - \{\{\circ,\Ham A\},\Ham B\} = \\
-&= -\{\Ham A,\{\circ,\Ham B\}\} - \{\Ham B,\{\Ham A,\circ\}\} = \\
-&= \{\circ,\{\Ham B,\Ham A\}\} = \\
-&= \{\circ,{\Liu A} {\Ham B}\}.
-\end{align*}
-
-This means that the commutator $[\Liu A,\Liu B]$ is, in fact, a new Liouville operator $\Liu C$ whose associated Hamiltonian is $\Ham C = {\Liu A}{\Ham B}$.
-By applying this procedure recursively, we can rewrite the right-hand side of Eq.~\eqref{eq:symmetric BCH} as $e^{\timestep \Liu S}$, where $\Liu S$ is the Liouville operator whose associated Hamiltonian is
-\begin{equation*}
-\label{eq:general shadow hamiltonian}
-\modified{\Ham{}} = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
-\end{equation*}
-
-Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\modified{\Ham{}}$ as above.
-In the literature, $\modified{\Ham{}}$ is usually referred to as a shadow Hamiltonian.
-
-In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies.
-
-For the splitting introduced in Section~\ref{sec:nve}, $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$ with the corresponding Liouviller operators $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$, with $K_{x}$ and $U_{x}$ being the gradient of the kinetic and potential energies, respectively.
-We first obtain $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$, and we find that
-\begin{align*}
-\Liu A \Liu A \Ham B &= \tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p}
-+ 2 \tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p}
-+ \tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} \\
-&+ \tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q}
-- \tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q},
-\end{align*}
-where the fact that $\tr{U_{\vt q \vt r}} = U_{\vt r \vt q}$ has been exploited.
-As reported in Ref.~\citenum{Silveira_2017}, the identity $\tr{\mt B}(\vt q) {\vt \pi} = -\tr{\mt B}(\vt \pi) {\vt q}$ is useful for performing differentiations.
-The first-order derivatives of $K$ with respect to $\vt p$, $\vt q$, and $\vt \pi$ are, respectively,
-\begin{align*}
-&K_{\vt p} = {\mt M}^{-1} {\vt p}, \\
-&K_{\vt q} = -\frac{1}{2} {\mt B}(\vt \pi) {\vt \omega}, \; \text{and} \\
-&K_{\vt \pi} = \frac{1}{2} {\mt B}(\vt q) {\vt \omega},
-\end{align*}
-where $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) \vt \pi = -\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt \pi) \vt q$.\cite{Silveira_2017} Hence,
-\begin{gather*}
-\tr{K_{\vt p}} U_{\vt r \vt r} K_{\vt p} = \tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p}, \\
-\tr{K_{\vt \pi}} U_{\vt r \vt q} K_{\vt p} = \frac{1}{2} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}, \\
-\tr{K_{\vt \pi}} U_{\vt q \vt q} K_{\vt \pi} = \frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
-\end{gather*}
-
-As $\diff{\tr{\vt \omega}}{\vt \pi} = \frac{1}{2} {\mt B}(\vt q) {\mt I}^{-1}$, the second-order derivative of $K$ with respect to $\vt \pi$ turns out to be
-\begin{equation*}
-K_{\vt \pi \vt \pi} = \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q).
-\end{equation*}
-
-Pre-multiplication by $\tr{K_{\vt q}}$ will introduce a product $\tr{\mt B}(\vt \pi) {\mt B}(\vt q)$, which has been shown in Ref.~\citenum{Silveira_2017} to be equal to ${\mt S}\left( \tr{\mt B}(\vt \pi) {\vt q} \right) = -2 {\mt S}({\mt I} {\vt \omega})$, where the operator ${\mt S}(\cdot)$ builds a skew-symmetric matrix from the entries of a vector.
-In addition, post-multiplication by $U_{\vt q}$ will introduce a product $\tr{\mt B}(\vt q) U_{\vt q}$, which is equal to $-2 {\vt \tau}^b$, where ${\vt \tau}^b$ is the body-fixed frame representation of the torque.
-Then, the fact that ${\mt S}(\vt x) {\vt y} = -{\mt S}(\vt y) {\vt x}$ leads to
-\begin{equation*}
-\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} = 0,
-\end{equation*}
-which is a null term as it corresponds to the quadratic form of a skew-symmetric matrix.
-Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{k=1}^3 \omega_k \hat{\mt B}_k ) \vt q$, where each $\hat{\mt B}_k$ is a skew-symmetric permutation matrix (i.e.
-$\tr{\hat{\mt B}}_k = -\hat{\mt B}_k$).
-This allows us to obtain
-\begin{align*}
-K_{\vt \pi \vt q} &= \frac{1}{2} (\diff{\tr{\vt \omega}}{\vt q}) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{k=1}^3 \omega_k \tr{\hat{\mt B}}_k = \\
-&= -\frac{1}{4} {\mt B}(\vt \pi) {\mt I}^{-1} \tr{\mt B}(\vt q) - \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k.
-\end{align*}
-
-Pre-multiplication by $\tr{K_{\vt \pi}}$ and post-multiplying $U_{\vt q}$, followed by some algebraic transformations, ultimately lead to
-\begin{equation*}
-\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q} = -\frac{1}{2} \tr{\vt \omega} \tr{\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} + \frac{1}{2} \tr{\vt \omega} {\mt S}({\vt \tau}^b){\vt \omega}.
-\end{equation*}
-Both terms in the right-hand side of the equation above are identically null. 
-Now, we are able to evaluate $\Liu A \Liu A \Ham B $, which is
-\begin{align*}
-\Liu A \Liu A \Ham B = &\tr{\vt p} {\mt M}^{-1} U_{\vt r \vt r} {\mt M}^{-1} {\vt p} + \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt r \vt q} {\mt M}^{-1} {\vt p}
-+ \\
-&\frac{1}{4} \tr{\vt \omega} \tr{\mt B}(\vt q) U_{\vt q \vt q} {\mt B}(\vt q) \vt \omega.
-\end{align*}
-The remaining term to obtain $H_s$ is simply given by 
-\begin{align*}
-\Liu B \Liu A \Ham B &= -\tr{U_{\vt r}} \tr{K_{\vt p \vt p}} U_{\vt r} - \tr{U_{\vt q}} \tr{K_{\vt \pi \vt \pi}} U_{\vt q} = \\
-&= -\tr{\vt F} {\mt M}^{-1} {\vt F} - \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b.
-\end{align*}
-
-As will become clear below, it is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
-\begin{equation*}
-K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation*}
-where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
-\begin{equation*}
-{\mt \Omega} = \left[\begin{array}{cc}
-{\mt M}^{-1} & \mt 0 \\
-\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
-\end{array}\right].
-\end{equation*}
-
-We now proceed to show the final expression of $H_s$, which results from grouping the terms corresponding to forces and torques in the \textit{modified} potential energy $\tilde{U}$, given by
-\begin{equation*}
-\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
-\end{equation*}
-while the remaining terms can be recast in a compact matrix, as follows
-\begin{equation*}
-\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation*}
-where
-\begin{equation*}
-\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)
-\end{equation*}
-and
-\begin{equation*}
-{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
-U_{\vt r \vt r} & U_{\vt r \vt q} \\
-U_{\vt q \vt r} & U_{\vt q \vt q}
-\end{array}\right].
-\end{equation*}
-
-Finally, we have that
-\begin{equation*}
-\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
-\end{equation*}
-
-Although $\modified{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
-As a result, the modified kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
-
-\suppinfo
+\begin{suppinfo}
 
@@
 
+\end{suppinfo}
+
 \bibliography{rigid_bodies}

### 2018-06-01 rigid_bodies.tex
results

@@
 Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
-However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behaviour.
+However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
 For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
-That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time step sizes.
+That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time-step sizes.
 
@@
 The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
-Simultaneosuly coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to mantain the temperature of the system at its specified value.
+Simultaneously coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to maintain the temperature of the system at its specified value.
 The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj_2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy.
@@
 
-A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system configuration, where $\timestep$ is the time step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$, which is
+A numerical solution of Eq.~\eqref{eq:ODE system for NVE} is usually represented by a stepwise application of the classical time-evolution propagator \cite{Tuckerman_2010} $e^{\timestep \Liu{\tiny NVE}}$ to the system configuration, where $\timestep$ is the time-step size and $\Liu{\tiny NVE}$ is the Liouville operator associated to $\Ham{}$, which is
 \begin{equation}
@@
 In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors). 
-It is also known that $\Ham{S}$ explicity depends on the time step size $\timestep$.
+It is also known that $\Ham{S}$ explicity depends on the time-step size $\timestep$.
 
@@
 
-\section{Results}
+\section{Application to Water and Discussion}
 \label{sec:numerical_results}
@@
 \ref{sec:mttk}-\ref{sec:smttk}, and  compare the results with those obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+It should be noted that all the simulations were carried out considering $t_d = 100$ fs.
 The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
@@
 Fortunately, with the analytical form of $\modified{\Ham{}}$ derived in this paper we are able to perform such an analysis for the NVE integrator which is essential, as will be shown below, to understand the results of the NVT methods.
-In Fig.~\ref{fig:nve} we depict the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different step sizes ($\timestep$).
-The numerical integrator correspons to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation. 
+In Fig.~\ref{fig:nve} we depict the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
+The numerical integrator correspond to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation. 
 As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
 It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
-As a result of employing such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ fs and $\timestep$ fs (add in conclusions that we should use high-order approximations).
+As a result of employing such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).
 
@@
 	\includegraphics{Figures/NVE.eps}
-	\caption{Total energy $\Ham{}$ (solid lines) and modified Hamiltonian $\modified{\Ham{}}$ (dotted lines) of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVE MD simulations for different step sizes $\timestep$.
-The numerical integrator correspons to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation.}
+	\caption{Total energy $\Ham{}$ (solid lines) and modified Hamiltonian $\modified{\Ham{}}$ (dotted lines) of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVE MD simulations for different time-step sizes $\timestep$.
+The numerical integrator correspond to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation.}
 	\label{fig:nve}
@@
 
-For our implementation for rigid bodies of the numerical scheme due to Martyna \textit{et al.} \cite{Martyna_1996} (Eq.~\ref{eq:splitting NVT}), in Fig.~\ref{fig:props_k_m} we depict the computed averages of temperature ($\langle T \rangle$) and potential energy per molecule ($\langle U \rangle$) as a function of $\timestep$.
+For our implementation for rigid bodies of the numerical scheme due to Martyna \textit{et al.} \cite{Martyna_1996} (Eq.~\ref{eq:splitting NVT}), in Fig.~\ref{fig:props_k_m} we depict the computed averages of temperature ($\langle T \rangle$) and potential energy per molecule ($\langle U \rangle$) as a function of the time-step size.
 The specified temperature corresponds to T = 298 K.
@@
 
-First, we report related the numerical stability of the NVT integrators.
+The explicit inclusion of the ``refined'' Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
+This poses the question whether the symmetry loss will influence the long-term stability.
+In this paper, we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
+Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
+On the other hand, the symmetric scheme due to Martyna \textit{et al.} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
+In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
+Also, the long-term stability is only modestly improved by our method so, evidently, the lack of symmetry has a negligible impact on $R$.
+Fig.~\ref{fig:energy_drift} also contains the drift in $\Ham{}$ computed for the NVE method\cite{Silveira_2017}, which results to be almost identical to our method.
+It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
+The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
+We have computed $R$ also for $\Ham{}$, which is the contribution to the extended energy from the physical degrees of freedom, namely the NVE part of the integrator.
+In this case, $\Ham{}$ remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration, which is also true for the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Table of Supp.
+Material).
+This has already been noticed by Davidchack\cite{Davidchack_2009}, and previously by Bond \textit{et al.} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.
 
-examine the numerical stability of the NVT integrators by simply monitoring the quantity expected to be conserved along the simulation in each case: $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for the proposed method, and $H$ (Eq.~\ref{eq:nvt extended energy}) for the integrators due to Martyna \textit{et al.} \cite{Martyna_1996} and Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-In Table~\ref{table:stability} we report the energy drift, denoted by $R$, as the the rate of change of $\widetilde{H}$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting between those variables.
-As it is already known that \cite{Davidchack_2009,Bond_2007}
+\begin{figure}
+	\includegraphics{Figures/energy_drift.eps}
+    \caption{Effect of the time-step size of 903 TIP3P \cite{Jorgensen_1983} water molecules in NVT MD simulations.
+$R$ for $\widetilde{H}$ and $\Ham{}$, which is the contribution to the extended energy from the physical degrees of freedom.
+We included the results for the integrator due to Martyna \textit{et al.} \cite{Martyna_1996}
+}
+	\label{fig:energy_drift}
+\end{figure}
 
-we plot these extended energies for different time step sizes.
-From Figs.~\ref{fig:num_stab}(a) and (b) which, respectively, correspond to the integrator proposed in this work and our implementation of the scheme due to Martyna \textit{et al.} \cite{Martyna_1996}, one notes that the energy appears to remain bounded for time step sizes up to 5 fs, and a continous drift takes place for $\timestep$ = 6 fs.
-The numerical stability is only modestly improved by our method.
-The result that stands out, because of its anomalous feature, corresponds to Fig.~\ref{fig:num_stab} (c), which was obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-Note that for each time step size, the extended energy continously increases until reaching a point at which the integrator becomes stable.
-It should be remarked that these results were double-checked with the LAMMPS software package \cite{Plimpton_1995}. 
+To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+Particularly, one can see from Fig.~\ref{fig:num_stab} (a) that for each time-step size the extended energy drifts upwards until reaching a point at which the integrator becomes stable.
+It should be remarked that these results were double-checked by using the LAMMPS software package \cite{Plimpton_1995}. 
 As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Davidchack \cite{Davidchack_2009} also implemented such a separate thermostatization and discovered the existance of energy flows from one thermostat to the other.
-To the best of our knowledge, the work of Davidchak \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-As can be seen in Fig.~\ref{fig:ther_energies}, where we depict the energy of each thermostat for different time step sizes $h$, that flux actually exists.
-Also, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, while the positive values are associated to the thermostat coupled to the rotational degrees of freedom, and thus the energy is flowing from translational to rotational degrees of freedom.
-Interestingly, Fig.~\ref{fig:ther_energies} suggests that a unique steady state is established, and the time to reach the steady state decreases with $h$.
-It should be noted that all the simulations were carried out considering $t_d = 100$ fs.
+Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} also implemented such a separate thermostatization and discovered the existence of energy flows from one thermostat to the other.
+To the best of our knowledge, the work of Davidchack \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat for different time-step sizes, that flux is also observed for the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.
+Interestingly, Fig.~\ref{fig:num_stab} (b) suggests that a unique steady state is established, and the time to reach the steady state decreases with $h$.
 As far as we know, this behavior has not been reported before.
-reformulate(The drift is localized within the thermostat variables and thus is not indicative of the non-stationary evolution within the system itself).
-
-\begin{table}
-	\caption{Effect of the time-step size on ....
-of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.}
-    \label{table:stability}
-    \begin{tabular}{SS[table-number-alignment = center,table-figures-exponent = 1]S[table-number-alignment = center,table-figures-exponent = 1]}
-     & \multicolumn{2}{c}{Drift rate [(kcal/mol)/ns]}  \\
-     \cmidrule[0.5mm](lr){2-3} 
-     $\timestep$ $\text{(fs)}$  & $\modified{H}$  & $\modified{\Ham{}}$   \\
-	1 &  4.6e-4  & 8.5e-2 \\
-	2 & -8.2e-3 & -4.3e-2 \\
-	3 & -3.2e-2 & 7.3e-2\\
-	4 & 8.1e-3 & -1.1e-1 \\
-	5 & 1.3e0 & -2.2e-2\\
-    6 & 4.4e1 & -3.4e-3 \\
-	7 & 8.8e2 & 2.2e-2 \\
-	8 & 9.8e3 & 7.6e-2 \\
-	9 & 7.1e4 & -2.5e-2 \\
-	10 & 3.2e5 & -3.2e-2  \\
-	\end{tabular}
-\end{table}
 
@@
 
-We now turn our focus to discretization errors.
-In Fig.~\ref{fig:properties} we show the computed average of some thermodynamic properties as a function of $\timestep$.
-The simulations were carried out at a specified temperature of T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_v$).
-The temperature was calculated from the kinetic energy assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
+In the previous section we showed that the energy drift is mainly localized within the thermostat variables.
+It is then reasonable to ask if this leads to a more ``accurate'' sampling of the physical system.
+To answer this question, we turn our attention to the discretization errors.
+
+In Fig.~\ref{fig:properties} we show the computed average of some thermodynamic properties as a function of $h$.
+The simulations were carried out specifying T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_v$).
+The temperature values of Fig.~\ref{fig:properties} (a) were computed from the simulated kinetic energies assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
 The lines are a guide for the eye. 
 Also, the horizontal dotted lines (black) help to visualize the departure of the computed averages with increasing $\timestep$.
-They correspond to the results obtained employing $h$ = 1 fs in the numerical scheme proposed in this paper, which from now on will be referred as ``This work''.
-In addition, we have included the results of applying our proposal but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
+They correspond to the results obtained employing $h$ = 1 fs for the numerical scheme proposed in this paper.
+In addition, we have included the results of applying our method but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
 Despite the non-equilibrium process ocurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}, because that is not the only problematic issue related to the method in question.
 Particularly, from Fig.~\ref{fig:properties} (a) we note that the simulated temperature decreases systematically with $\timestep$.
-This kind of behavior is observed in NVE simulations\cite{Davidchack_2009,Silveira_2017}, reflecting that the factorization employed in Kamberaj \textit{et al.} \cite{Kamberaj_2005} in not suitable. 
+This kind of behavior is observed in NVE simulations\cite{Davidchack_2009,Silveira_2017}, reflecting that the factorization employed in Kamberaj \textit{et al.} \cite{Kamberaj_2005} is not suitable. 
 As a result of this inefficient temperature control, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Fig.~\ref{fig:properties} (b) and (c)).
@@
 Of course, as $\modified{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
-Indeed, for time step sizes up to, the agreement is excelent.
-Note that in the abscence of reweighting, the calculated temperatures are notably influenced by $h$.
-The free energy difference between the sampled state $\modified{\Ham{}}$ and the state we would like to sample, represented by the physical Hamiltonian $\Ham{}$.
-As expected, the $\Delta F$ is small, as it should be
-Note that a small, still clearly influencing systematic error is present in our scheme.
+Indeed, for all the time-step sizes the agreement is excellent.
+Note that in the absence of reweighting, the computed temperatures are notably influenced by $h$, in comparison with the potential energy.
+Also, a small, still clearly influencing systematic error is present in our scheme.
+
+One direct way to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\modified{\Ham{}}$ is by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\modified{\Ham{}}$, and the state of interest.
+From Table~\ref{table:reweight}, one verifies that $\Delta F$ is only negligible for $h$ = 1 fs and $h$ = 2 fs.
 
@@
      \cmidrule[0.5mm](lr){3-4} \cmidrule[0.5mm](lr){5-6}
-     $\timestep$ & $\modified{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F$} \\
-        $\text{(fs)}$  & $\text{(K)}$ & $\text{(K)}$ & $\text{(kcal/mol)}$  & $\text{(K)}$ & $\text{(kcal/mol)}$ & $\text{(kcal/mol\,N)}$ \\
-	1 & 298.00(1) & 297.71(1) & -9.1015(3) & 298.00(1) & -9.1028(3) & -0.001 \\
-	2 & 297.99(1) & 296.82(1) & -9.0970(3)  & 297.98(1)  & -9.1024(3) & -0.006 \\
-	3 & 297.98(1) & 295.38(1) & -9.0885(3) & 297.92(1) & -9.1005(3) & -0.013 \\
-	4 & 298.01(1) & 293.49(1)& -9.0764(3) & 297.85(3) & -9.0969(3) & -0.022 \\
+     $\timestep$ & $\modified{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F/N$} \\
+        $\text{(fs)}$  & $\text{(K)}$ & $\text{(K)}$ & $\text{(kcal/mol)}$  & $\text{(K)}$ & $\text{(kcal/mol)}$ & $\text{(kcal/mol)}$ \\
+	1 & 298.00(1) & 297.71(1) & -9.1015(3) & 298.00(1) & -9.1028(3) & -0.0006 \\
+	2 & 297.99(1) & 296.82(1) & -9.0970(3)  & 297.98(1)  & -9.1024(3) & -0.004 \\
+	3 & 297.98(1) & 295.38(1) & -9.0885(3) & 297.92(1) & -9.1005(3) & -0.008 \\
+	4 & 298.01(1) & 293.49(1)& -9.0764(3) & 297.85(3) & -9.0969(3) & -0.013 \\
 	5 & 297.99(1)& 291.13(1)& -9.0577(3) & 297.6(2) & -9.091(1) & -0.033 \\
-    6 & 298.00(1) & 288.46(1) & -9.0314(3) & 298.8(3) & -9.075(3) &  -0.045 \\
-	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
-
+    6 & 298.00(1) & 288.46(1) & -9.0314(3) & 296.8(3) & -9.075(3) &  -0.026 \\
+%	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
 	\end{tabular}
@@
 
+reformulate(The method controls the temperature of the physical variables well, even as the auxiliary thermostat variables show large and growing deviations from the initial state)
+
 \begin{table*}
@@
 			\hline
-			\multicolumn{7}{c}{``This work (Unsplit)''}\\
+			\multicolumn{7}{c}{This work (Unsplit)}\\
 1.0	& 3.02(2) & 1.07 & & & &  \\
@@
 
-\begin{table}
-	\caption{Effect of the time-step size on ....
-of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.}
-    \label{table:stability_martyna}
-    \begin{tabular}{SS[table-number-alignment = center,table-figures-exponent = 1]S[table-number-alignment = center,table-figures-exponent = 1]}
-     & \multicolumn{2}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}  \\
-     \cmidrule[0.5mm](lr){2-3} 
-     $\timestep$ $\text{(fs)}$  & $\modified{H}$  & $\modified{\Ham{}}$     \\
-	1 &  -1.6e-2  & 4.5e-3\\
-	2 & 9.5e-2 & 1.4e-1\\
-	3 & 2.8e-1 & 5.6e-2\\
-	4 & 1.1e0 & 5.6e-2\\
-	5 & 3.3e0 & -3.7e-2\\
-    6 & 6.1e1 & 6.9e-2 \\
-	7 & 9.8e2 & -3.6e-3 \\
-	8 & 1.1e4 & 1.8e-2 \\
-	9 & 7.5e4 & 3.8e-3 \\
-	10 & 3.3e5 & 1.9e-2  \\
-	\end{tabular}
-\end{table}
-
 \bibliography{rigid_bodies}

### 2018-06-04 rigid_bodies.tex
intro

@@
 }
-
+\usepackage{xr}
+\externaldocument{rigid_bodies_supp_info}
 % Comment below before submitting
@@
 
-Molecular Simulation, being the computational realization of Statistical Mechanics,\cite{Tuckerman_2010} involves several types of approximations in order to mimic ``real'' physical systems.
+Molecular Simulation, being the computational realization of Statistical Mechanics \cite{Tuckerman_2010}, involves several types of approximations in order to mimic ``real'' physical systems.
 Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
 However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
-For instance, it has already been shown that the unavoidable truncation errors, associated with the numerical integration in Molecular Dynamics (MD), disrupt the equipartition of kinetic energy and, as a consequence, the temperature becomes an undefined property.\cite{Eastwood_2010} In a previous paper,\citep{Silveira_2017} we reported the same artifact to occur on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body MD.
-That finding brings into question the method introduced by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom, but fails to reproduce the specified temperature for certain time-step sizes.
+For instance, it has already been shown that the unavoidable errors in the discretization of the equations of motion disrupt the equipartition of kinetic energy \cite{Eastwood_2010}.
+The same artifact occurs on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics (MD) \cite{Davidchack_2009,Silveira_2017}.
+This finding brings into question the NVT integrator developed by Kamberaj and co-workers\cite{Kamberaj_2005} that, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
+In addition, this method fails to reproduce the specified temperature for time-step sizes usually employed in MD.
 
-The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water,\cite{Jorgensen_1983} but also it is a simple coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, like proteins and nanoparticules.\cite{Knorowski_2012, Patra_2013} As a matter of fact, the approach of Kamberaj~\textit{et al.},\cite{Kamberaj_2005} which is implemented, for instance, in the softwares packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS,\cite{Plimpton_1995} has been applied to simulate small molecules,\cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2} as well as membranes,\cite{Bucior_2012} molecular motors,\cite{Akimov_2012} micelles\cite{Yan_2008} and nanoparticles.\cite{Patra_2014}
+The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water \cite{Jorgensen_1983}, but also it can be employed as a coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, such as proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
+As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the software packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes\cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles \cite{Patra_2014}.
 
-The Nos\'{e}-Hoover chain thermostat,\cite{Martyna_1992} routinely used in MD, is based on the extended phase-space approach that includes both the system of interest and extra degrees of freedom, which in turn are intended to control the kinetic energy fluctuations of the physical variables.
-The dynamics of those extra degrees of freedom introduces non-Hamiltonian components to the equations of motion, with which backward error analysis is no longer feasible.
-Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}
+The Nos\'{e}-Hoover chain (NHC) thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach, in which auxiliary degrees of freedom act regulating the kinetic energy of the physical variables so as to be consistent with the Maxwell-Boltzmann distribution at a given target temperature.
+To do so the NHC thermostat requires an expression relating the temperature, a macroscopic parameter, with the kinetic energy, which is a property of each microstate.
+Usually, in MD the equipartition theorem plays this role. 
+However, as already mentioned, this theorem is disrupted in numerical schemes with non-negligible discretization errors.
+This is because numerical trajectories actually sample from the so-called \textit{shadow} Hamiltonian, which cannot be merely separated into momentum and position dependent parts.
+Instead, what is usually identified as the \textit{shadow} kinetic energy contains quadratic terms in particle momenta, but also depends on particle positions\cite{Eastwood_2010}.
+It seems reasonable to argue that the thermostat should control this kinetic energy, instead of the physical one.
+However, the theory underlying the concept of shadow Hamiltonians correspond to backward error analysis, which is only applicable to symplectic integrators.
+This is not the case of NHC numerical schemes, given that the auxiliary variables introduce non-Hamiltonian components to the equations of motion.
+In this paper we introduce a procedure that partially overcomes this limitation by performing the backward error analysis on the NVE part of the numerical integrator with minimal computational overheads.
+The resulting NHC scheme, which simultaneously couples the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, is formulated as a function on an approximation to the \textit{shadow} kinetic energy.
+As will be shown later, this is only possible for splitting schemes in which the NVE Liouville operator is placed in the middle between the thermostat propagators. 
+This new method is particularly appealing for rigid-body MD, primarily because in terms of numerical stability one is able to employ considerable time-step sizes for Verlet-type numerical schemes.
+However, the discretization errors begin to manifest much earlier.
+The case of rigid water exemplifies this behavior and serves for an excellent case test for our method. 
+In fact, we have been able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by Martyna \citeauthor{Martyna_1996} \cite{Martyna_1996}.
+It should be emphasize that this novel technique is not limited to rigid bodies, neither to the NHC method.
+The only requirement is that
 
-In this paper, we derive time reversible and measure-preserving numerical integrators for rigid-body MD in the canonical (NVT) ensemble.
-The starting point is our previous work on the microcanonical (NVE) ensemble,\citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom.
-Simultaneously coupling the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, we are able to maintain the temperature of the system at its specified value.
-The analysis of the integrators includes comparisons with results from the method introduced by Kamberaj~\textit{et al.},\cite{Kamberaj_2005} as well as with the Hybrid Monte Carlo (HMC) method,\cite{Duane_1987} which is exact in the sense that truncation errors do not affect its accuracy.
-We also include results from simulations using the conventional SHAKE\cite{Ryckaert_1977} procedure, also coupled to a unique Nos\'{e}-Hoover chain.
-
-backward error analysis
-
-The paper is organized as follows.
-In Sec.~\ref{sec:theory} we briefly review the equations of motion in the NVE ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble.
-In Sec.~\ref{sec:numerical_solvers} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
+The paper proceeds as follows.
+In Sec.~\ref{sec:nve} we briefly review the equations of motion in the NVE ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble.
+In Sec.~\ref{sec:refined_method} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
 Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
-
+%The starting point is  microcanonical (NVE) ensemble \citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom. 
 \section{Methodology}
@@
 
-The Hamiltonian system of ordinary differential equations (ODE) that describes a rigid-body motion is given by \cite{Silveira_2017}
+The Hamiltonian system of ordinary differential equations (ODE) that describes the rigid-body motion is given by \cite{Silveira_2017}
 \begin{subequations}
@@
 with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$.
-In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugated momentum of such quaternion.
+In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugate momenta of the quaternion.
 Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity in the body-fixed frame.
-Matrices $\mt M$ and $\mt I$ are diagonal ones.
+Matrices $\mt M$ and $\mt I$ are diagonal.
 The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
@@
 The action of $e^{\timestep \Liu{A}}$ is the simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step.
-For this rotational part of the integration, in Ref.~\cite{Silveira_2017} we derived a general solution in a form that facilitates computer implementation and allows, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
-While the approximate solutions, such as the NO-SQUISH \cite{Miller_2002} method, only involve sine and cosine terms, the exact solution comprise ``less common'' functions, such as the Jacobi elliptic functions and Carlson integrals.
-This contributes to the fact that the approximate solution is more commonly used \cite{vanZon2007}. 
-For mathematical details, the reader is referred to our previos work \cite{Silveira_2017}, where we have verified that in liquid-phase simulations there is no significan gain in accuracy by using the exact solution in substitution to the NO-SQUISH as part of a symplectic integration scheme.
-However, in the NO-SQUISH method an overall rotation is given by a sequence of uniform rotations around the principal axes of a body, which results in a factorization of the rotational operator given by $ e^{\frac{h}{2} i\!L_3} e^{\frac{h}{2} i\!L_2} e^{h i\!L_1} e^{\frac{h}{2} i\!L_2} e^{\frac{h}{2} i\!L_3}$, with each factor representing a rotation around a given axe.
+In Ref.~\citenum{Silveira_2017} we derived a general solution of torque-free rotation in a form that facilitates computer implementation allowing, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
+For mathematical details, the reader is referred to our previos work \cite{Silveira_2017}.
+The exact solution comprise ``less familiar'' mathematical functions, such as the Jacobi elliptic functions and Carlson integrals.
+On the other hand, the approximate solutions, such as the NO-SQUISH method  \cite{Miller_2002}, have proven to be very accurate in liquid-phase simulations \cite{Silveira_2017} and its formulation only involve sine and cosine terms.
+This may explain why that the approximate solution is more commonly used \cite{vanZon2007}.
+This pictures completely changes in the context of backward error analysis: in the approximate methods, an overall rotation is given by a sequence of uniform rotations around the principal axes of a body, which results in a factorization of the rotational operator given by $ e^{\frac{h}{2} i\!L_3} e^{\frac{h}{2} i\!L_2} e^{h i\!L_1} e^{\frac{h}{2} i\!L_2} e^{\frac{h}{2} i\!L_3}$, with each factor representing a rotation around a given axes.
 The analytical solution, on the other hand, entails a single operator, thus making the application of backward error analysis considerably simpler.
@@
 \label{sec:mttk}
+
+Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}
+
 As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
@@
 In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors). 
-It is also known that $\Ham{S}$ explicity depends on the time-step size $\timestep$.
+It is also known that $\Ham{S}$ explicitly depends on the time-step size $\timestep$.
 
@@
 \subsection{Adjustment of the NVT Dynamics}
-\label{sec:smttk}
+\label{sec:refined_method}
 Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
@@
 In the following, we analyze the performance of the proposed NVT integrators introduced in Secs.
-\ref{sec:mttk}-\ref{sec:smttk}, and  compare the results with those obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+\ref{sec:mttk}-\ref{sec:refined_method}, and  compare the results with those obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
 It should be noted that all the simulations were carried out considering $t_d = 100$ fs.
@@
 Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
-These observations motivated the development of the scheme of Sec.~\ref{sec:smttk}, whose promising results we show below.
+These observations motivated the development of the scheme of Sec.~\ref{sec:refined_method}, whose promising results we show below.
 
@@
 We have computed $R$ also for $\Ham{}$, which is the contribution to the extended energy from the physical degrees of freedom, namely the NVE part of the integrator.
-In this case, $\Ham{}$ remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration, which is also true for the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Table of Supp.
+In this case, $\Ham{}$ remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration, which is also true for the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Table~\ref{table:stability} of Supp.
 Material).
@@
 
-With the aim of highlighting the importance of reweighting, in Table~\ref{table:reweight} we present the calculated temperature and potential energy for each $\timestep$ before and after applying Eq.~\ref{eq:reweighting}.
+We have also (check ensemble results)\ref{table:check_ens}
+
+With the aim of highlighting the importance of reweighting to recover averages of the physical system, in Table~\ref{table:reweight} we present the calculated temperature and potential energy for each $\timestep$ before and after applying Eq.~\ref{eq:reweighting} for time-step sizes smaller 7 fs, as we saw in Fig.
+that for ths case, the temperature does not reproduce the specified temperature.
 The ``refined'' temperature $\modified{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
 Of course, as $\modified{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
-Indeed, for all the time-step sizes the agreement is excellent.
-Note that in the absence of reweighting, the computed temperatures are notably influenced by $h$, in comparison with the potential energy.
-Also, a small, still clearly influencing systematic error is present in our scheme.
+Indeed, for all the time-step sizes considered the agreement is excellent.
+We see that in the absence of reweighting, the computed temperatures are notably influenced by $h$.
+Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $h$.
+This could, in principle, be remediated by using a higher-order ``refined'' Hamiltonian.
 
-One direct way to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\modified{\Ham{}}$ is by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\modified{\Ham{}}$, and the state of interest.
-From Table~\ref{table:reweight}, one verifies that $\Delta F$ is only negligible for $h$ = 1 fs and $h$ = 2 fs.
+In addition, we are able to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\modified{\Ham{}}$ by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\modified{\Ham{}}$, and that state of interest.
+$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
+As expected, from Table~\ref{table:reweight} one verifies that $\Delta F$ is only negligible for $h$ = 1 fs and systematically increases with $h$.
 
@@
     6 & 298.00(1) & 288.46(1) & -9.0314(3) & 296.8(3) & -9.075(3) &  -0.026 \\
-%	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
+	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
 	\end{tabular}
@@
 
-To end this section,  Kinetic Energy Partition
+We end this section with a discussion of the effect of discretization errors on the kinetic energy partition.
+In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition, which are calculated at each time-step size from the total kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times (3N - 3)$ and $\langle K_r \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times 3N$.
+
+In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because is the rotational degrees of freedom are the fastest modes in the rigid model of water.
+We see that the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
+From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption and a small deviation is only observed for 7 fs.
+On the other hand, the results depicted in Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of Martyna \textit{et al.} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c) in the sense that the simulated translational and rotational kinetic energies deviate in opposite directions from the expected values and they do so contrary to what is predicted by the equipartition theorem.
+Since we have a smaller number of translational degrees of freedom, the corresponding kinetic energy should be smaller than the rotational part.
 
@@
 	\includegraphics{Figures/energy_partition.eps}
+		\caption{Effect of the time-step size on ... 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
+	\label{fig:energy_partition}
 \end{figure*}
 
-reformulate(The method controls the temperature of the physical variables well, even as the auxiliary thermostat variables show large and growing deviations from the initial state)
-
-\begin{table*}
-	\caption{Effect of the time-step size on ....
-of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes....}
-	\label{table:check_ens}
-	\sisetup{table-number-alignment = center}
-    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]SS[separate-uncertainty,table-figures-uncertainty= 1]SS[separate-uncertainty,table-figures-uncertainty= 1]S}
-	{$\timestep$} & {estimated $\Delta T$} & {$\sigma$ deviations}  & {estimated $\Delta T$}  & {$\sigma$ deviations}  &    {estimated $\Delta T$}  & {$\sigma$ deviations} \\
-	{(fs)} & {(K)} &   & {(K)}  &  & {(K)}  & \\
-			\hline
-			\multicolumn{7}{c}{This work (Unsplit)}\\
-1.0	& 3.02(2) & 1.07 & & & &  \\
-2.0 & 3.02(2) & 0.81 & & & &  \\
-3.0 & 2.98(2) & 0.73 & & & &  \\
-4.0 & 3.01(2) & 0.41 & & & &  \\
-5.0 & 3.02(2) & 0.79 & & & &  \\
-6.0 & 3.03(3) & 1.23 & & & &  \\
-7.0 & 2.98(2) & 0.74 & & & &  \\
-            \hline
-			\multicolumn{7}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}\\
-1.0	& 2.98(2) & 0.66 & 3.01(2) & 0.49 & 2.96(3) & 0.90 \\
-2.0 & 2.98(2) & 0.73 & 2.99(2) & 0.03 & 3.02(3) & 0.66 \\
-3.0 & 2.95(2) & 2.33 & 3.01(2) & 0.65 & 2.91(3) & 2.47 \\
-4.0 & 3.00(2) & 0.01 & 3.01(2) & 0.72 & 2.92(3) & 2.19  \\
-5.0 & 2.93(2) & 3.05 & 3.01(2) & 0.70 & 2.94(3) & 1.58  \\
-6.0 & 2.92(2) & 3.87 & 3.00(2) & 0.22 & 2.91(3) & 2.76  \\
-7.0 & 2.89(2) & 5.35 & 3.02(1) & 1.33 & 2.87(2) & 4.33  \\
-            \hline
-			\multicolumn{7}{c}{Kamberaj \textit{et al.} \cite{Kamberaj_2005}}\\
-1.0	& 2.97(2) & 1.16 & 2.99(2) & 0.16 & 2.89(3) & 2.84  \\
-2.0 & 3.00(2) & 0.02 & 3.02(2) & 0.99 & 2.00(4) & 0.25  \\
-3.0 & 2.99(4) & 0.18 & 3.08(4) & 4.09 & 3.1(1) & 1.17  \\
-4.0 & 2.92(8) & 0.99 & 3.12(2) & 6.01 & 2.8(2) & 0.95  \\
-5.0 & 3.02(5) & 0.40 & 3.2(2)  & 9.50 &  2.9(1) & 0.28  \\
-6.0 & 3.14(5) & 2.69 & 3.29(2) & 14.68 & 3.21(7) & 2.90  \\
-7.0 & 3.03(4) & 0.73 & 3.35(2) & 17.37 & 2.99(6) & 0.16  \\
-    \end{tabular}
-\end{table*}
 %It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
@@
 by contrast
-
+reformulate(The method controls the temperature of the physical variables well, even as the auxiliary thermostat variables show large and growing deviations from the initial state)
 \begin{acknowledgement}
@@
 \end{document}
+%A geometric integrator is a numerical method which preserves certain geometric structures of the exact flow of a differential equation.
+%the issues arising in the design and analysis of numerical methods for molecular dynamics
+%(symplecticness) of the phase flow.
+We will later show (in Chap.
+3) that such symplectic methods preserve a perturbed energy-like invariant (to very high accuracy); the preservation of this modified energy then ensures that the original energy is nearly conserved
+%A benefit of respecting the symplectic structure lies in the prospect of using backward error analysis to analyze the behavior of the method
+%It seems intuitively correct to demand that a numerical method should mimic as much as possible the available structure of the dynamical system, when this is possible to do without incurring large computational overheads in the process.
+So much the better that discrete dynamical systems designed in this way appear to be related to modified continuous dynamics with appropriate properties.
+This observation is purely aesthetic.
+%These methods obtain their energy preservation properties indirectly: we design the methods to exactly conserve a certain geometric property
+%The purpose of this section is to demonstrate that, by using the principles of geometric integration and the backward error analysis, it is possible to compute effective estimates of the error in numerically computed averages, and even to correct those averages by a straightforward reweighting

### 2018-06-05 rigid_bodies.tex
general revision

@@
 In fact, we have been able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by Martyna \citeauthor{Martyna_1996} \cite{Martyna_1996}.
-It should be emphasize that this novel technique is not limited to rigid bodies, neither to the NHC method.
+It should be emphasize that this novel technique is not limited to rigid bodies, neither to the NHC method. \citeauthor{Bussi_2007} \cite{Bussi_2007}
 The only requirement is that
@@
 
-\subsection{NVT Dynamics with a Single Thermostat Chain}
+\subsection{NVT Dynamics with a Single Thermostat Chain: Notation and Measure-Preserving Integration}
 
-Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble.
-For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system.
-This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement.
-To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$.
+In this work, we couple a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} to both the translational and rotational degrees of freedom of the rigid-body system.
+To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momenta $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$.
 The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
@@
 \end{equation}
-where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
+where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the target temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
 As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.
 
-Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has advanced considerably in recent years.\cite{Tuckerman_1999, Tuckerman_2001, Sergi_2001, Sergi_2003, Ezra_2004, Sergi_2004, Ezra_2006, Sergi_2010_2} Invaluable tools have been devised for the development and analysis of extended phase space methods.
 By employing the method of Sergi and Ferrario \cite{Sergi_2001}, we obtain the equations of motion for the NVT dynamics with a single thermostat chain, which are
@@
 	\begin{align}
-%
 \label{eq:nhc_r}
@@
 \end{subequations}
+With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
 
-With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
-It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics, but are often computed for checking how well a numerical integrator conserves $H$.
-Analysis of the probability distribution generated by Eq.~\eqref{eq:ODE system for NVT} can be done through the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman_2001} For this, conservation laws must be identified and driven variables eliminated.
 If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are integrals of motion, while the system's center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates.
@@
 
-\subsection{Measure-Preserving Integration for the NVT Dynamics}
-\label{sec:mttk}
-
-Altough it is not possible to devise symplectic integrators, one can rigurously obtain measure-preserving numerical solvers.\cite{Sergi_2001, Ezra_2004, Ezra_2006}
-
-As in the Hamiltonian case, it is possible to devise a splitting scheme for solving Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}.
-For that, we employed the systematic method introduced by Ezra \cite{Ezra_2006} for deriving accurate integrators which preserve the phase-space measure associated with the original equations of motion, which is
-In short, a splitting scheme will preserve a given measure if each individual propagator does it as well. 
-It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure is given by
-\begin{equation}
-\label{eq:nhc_measure}
-w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
-\end{equation}
-
-will preserve a given measure $e^{-w(\vt x)}d\vt x$
+Altough it is not possible to devise symplectic integrators for Eq.~\eqref{eq:ODE system for NVT}, one can rigurously obtain measure-preserving numerical solvers \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
+It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure $e^{-w(\vt x)}d\vt x$ of Eq.~\ref{eq:ODE system for NVT} is given by $w = -7N \eta_1 - \sum_{j=2}^M \eta_j$.
+It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics so, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
+The $\eta$ coordinates are often computed for checking how well a numerical integrator conserves $H$.
 
@@
 where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
-Of course, the NVE part can be integrated exactly as described before.
-The NHC part, in turn, can be split even further as
+The NHC part can be split even further as
 \begin{equation*}
@@
 Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
-In the following, the NVT integrator introduced in the present section will be denoted as MTTK as its development was inspired by Ref.~\cite{Martyna_1996}.
+This numerical scheme is simply a measure-preserving version of the NVT integrator introduced by \citeauthor{Martyna_1996} \cite{Martyna_1996}.
 
@@
 A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
-Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian \cite{Hairer_2006}.
-In this way, while $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors). 
-It is also known that $\Ham{S}$ explicitly depends on the time-step size $\timestep$.
+Hence, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian \cite{Hairer_2006}, which explicitly depends on the time-step size $\timestep$.
 
-The first new result we present in this paper is an analytical form for the shadow Hamiltonian concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~ \citenum{Silveira_2017}.
+The first new result we present in this paper is an analytical form of an approximation to the shadow Hamiltonian, referred here as \textit{refine}d Hamiltonian, concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~ \citenum{Silveira_2017}.
 A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and here we summarize the final expressions.
 
-As shown in Appendix~\ref{sec:rigid body shadow hamiltonian}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are modified kinetic and potential energies.
+As demonstrated in Appendix~\ref{sec:rigid body shadow hamiltonian}, the \textit{refined} Hamiltonian for a Verlet-type method applied to a system of rigid bodies is given by $\modified{\Ham{}} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are \textit{refined} kinetic and potential energies.
 The latter is given by
@@
 \label{eq:modified potential energy}
-\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right),
+\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
 \end{equation}
-where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the resultant torque vector.
 Note that $\modified U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
-In turn, the modified kinetic energy can be expressed as
+In turn, the \textit{refined} kinetic energy can be expressed as
 \begin{equation}
@@
 \end{equation}
+Also, the ``extended energy'' preserved by this approach is now given by
+\begin{equation}
+\label{eq:this extended energy}
+\widetilde{H} = \modified K + \modified U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
+\end{equation}
 
@@
 Fortunately, only positions and orientations are required for the numerical differentiation.
-This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time.
+This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time, as schematically illustrated in Fig.~\ref{fig:virtual}.
 By incomplete we mean that the step is aborted once the positions and orientations have been updated.
@@
 \begin{tikzpicture}
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](3,.2) -- (3,1.5);
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
 \draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
@@
 \end{figure}
-%\begin{figure}
-%\begin{tikzpicture}
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](2,.2) -- (2,1.5);
-%\draw[dash pattern=on 2pt off 3pt on 4pt off 4pt](5.5,.2) -- (5.5,1.5);
-%\node(0) at (0,0)[shape=circle,draw, minimum width=1cm, minimum height=1cm]        {$r_{t-h}$};
-%\node(1) at (2.5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_t$};
-%\node(2) at (5,0)[shape=circle,draw,double,minimum width=1cm, minimum height=1cm]        {$r_{t+h}$};
-%\node(3) at (7.5,0)[shape=circle,draw,minimum width=1cm, minimum height=1cm] {$r_{t+ 2h}$};
-%\node(4) at (3.75,1)[] {$e^{\timestep \Liu{\tiny NVE}}$};
-%\path [dashed,->]
-%  (1)      edge                 node [above]  {virtual}     (0)
-%  (2)      edge                 node [above] {virtual}     (3);
-%\path [->,thick] 
-%  (1)      edge                 node [above]  {}            (2);
-%\end{tikzpicture}
-%\caption{Schematic representation}
-%\label{fig:virtual}
-%\end{figure}
-
-A drawback of this approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
-As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
-In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.
 
@@
 
-$\widetilde{H}$, but an ``extended energy'' given by \cite{Martyna_1992}
-\begin{equation}
-\label{eq:this extended energy}
-\widetilde{H} = \modified K + \modified U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
-\end{equation}
-
-Tolman's generalized equipartition theorem \cite{Uline_2008}
+Please note that in Eq.~\ref{eq:generalized force NVT} we have employed Tolman's generalized equipartition theorem \cite{Uline_2008}, which states that
 \begin{equation*}
-\left\langle \tr{\vt p} \diff{\modified{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\modified{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T
+\left\langle \tr{\vt p} \diff{\modified{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\modified{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T.
 \end{equation*}
+This definition, with the \textit{refined} kinetic energy between brackets, provides a generalization for $h > 0$ of the simulated temperature \cite{Eastwood_2010}.
 
@@
 
-\section{Theoretical Background}
-
-We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj_2005} does not fulfill this criterion.
-
-We end this section by noting that the term that premultiplies the extended phase-space volume in the invariant-measure above is a function of $\vt \eta$.
-Given that the time evolution of the physical variables and the thermostat momenta does not depend on $\vt \eta$, we might conclude that, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
-However, as will be shown in the numerical results of Sec.~\ref{sec:numerical_results},
-
-\subsection{Generalized Equipartition}
-
-The important thing is that $\modified{\mathbf \Omega}$ depends exclusively on the positions and orientations, which makes $\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] = \modified{\mathbf \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]$.
-In this way, the equipartition theorem (valid at least in this direction? something to do with integrating through momenta at least internally to the integral in coordinates?) states that
-\begin{equation}
-\label{eq:equipartition}
-3N k_B T = \left\langle \tr{\vt p} \diff{\tilde K}{\vt p} \right\rangle = 2\langle \tilde K \rangle
-\end{equation}
-
-Sampling validation: Shirts \cite{Shirts_2013}
+A drawback of our approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
+As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
+In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.
 
@@
 
-We have also (check ensemble results)\ref{table:check_ens}
+We have also (check ensemble results)\ref{table:check_ens} Sampling validation: Shirts \cite{Shirts_2013}
 

### 2018-06-06 rigid_bodies.tex
theory revision I

@@
 \newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
-%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
 \newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
@@
 
-%\title{Revisiting the Nos\'{e}-Hoover chain thermostat for Rigid-Body Molecular Dynamics}
 \title{Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis}
 
-%\abbreviations{AAA, BBB, CCC}
 \keywords{aaa, bbb, ccc}
@@
 The same artifact occurs on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics (MD) \cite{Davidchack_2009,Silveira_2017}.
-This finding brings into question the NVT integrator developed by Kamberaj and co-workers\cite{Kamberaj_2005} that, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-In addition, this method fails to reproduce the specified temperature for time-step sizes usually employed in MD.
+This finding brings into question the NVT integrator developed by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
+In addition, as will be shown later, this method fails to reproduce the specified temperature for time-step sizes usually employed in MD.
 
-The relevance of rigid-body MD stems from the fact that not only some molecules are modelled as rigid bodies, like the ubiquitous case of water \cite{Jorgensen_1983}, but also it can be employed as a coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, such as proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
+The relevance of rigid-body MD stems from the fact that not only some molecules are usually modelled as rigid bodies, like the ubiquitous case of water \cite{Jorgensen_1983}, but also it can be employed as a coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, such as proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
 As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the software packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes\cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles \cite{Patra_2014}.
@@
 The Nos\'{e}-Hoover chain (NHC) thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach, in which auxiliary degrees of freedom act regulating the kinetic energy of the physical variables so as to be consistent with the Maxwell-Boltzmann distribution at a given target temperature.
-To do so the NHC thermostat requires an expression relating the temperature, a macroscopic parameter, with the kinetic energy, which is a property of each microstate.
+This requires an expression relating the temperature, a macroscopic parameter, with the kinetic energy, which is a property of each microstate.
 Usually, in MD the equipartition theorem plays this role. 
@@
 This is because numerical trajectories actually sample from the so-called \textit{shadow} Hamiltonian, which cannot be merely separated into momentum and position dependent parts.
-Instead, what is usually identified as the \textit{shadow} kinetic energy contains quadratic terms in particle momenta, but also depends on particle positions\cite{Eastwood_2010}.
-It seems reasonable to argue that the thermostat should control this kinetic energy, instead of the physical one.
-However, the theory underlying the concept of shadow Hamiltonians correspond to backward error analysis, which is only applicable to symplectic integrators.
+Instead, what is usually identified as the \textit{shadow} kinetic energy contains quadratic terms in particle momenta accompanied by particle positions dependency \cite{Eastwood_2010}.
+It seems then reasonable to argue that the thermostat should control this kinetic energy, instead of the physical one.
+However, the theory underlying the concept of shadow Hamiltonian correspond to backward error analysis, which is only applicable to symplectic integrators.
 This is not the case of NHC numerical schemes, given that the auxiliary variables introduce non-Hamiltonian components to the equations of motion.
 In this paper we introduce a procedure that partially overcomes this limitation by performing the backward error analysis on the NVE part of the numerical integrator with minimal computational overheads.
-The resulting NHC scheme, which simultaneously couples the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, is formulated as a function on an approximation to the \textit{shadow} kinetic energy.
-As will be shown later, this is only possible for splitting schemes in which the NVE Liouville operator is placed in the middle between the thermostat propagators. 
-This new method is particularly appealing for rigid-body MD, primarily because in terms of numerical stability one is able to employ considerable time-step sizes for Verlet-type numerical schemes.
-However, the discretization errors begin to manifest much earlier.
-The case of rigid water exemplifies this behavior and serves for an excellent case test for our method. 
-In fact, we have been able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by Martyna \citeauthor{Martyna_1996} \cite{Martyna_1996}.
-It should be emphasize that this novel technique is not limited to rigid bodies, neither to the NHC method. \citeauthor{Bussi_2007} \cite{Bussi_2007}
-The only requirement is that
+The resulting NHC scheme, which simultaneously couples the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, controls the fluctuations of an approximation to the \textit{shadow} kinetic energy.
 
-The paper proceeds as follows.
-In Sec.~\ref{sec:nve} we briefly review the equations of motion in the NVE ensemble, which are the basis for developing the corresponding dynamics in the NVT ensemble.
-In Sec.~\ref{sec:refined_method} we devise the corresponding numerical integrators, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}.
+Our new method is particularly appealing for rigid-body MD, primarily because discretization errors begin to manifest much earlier than any numerical stability caused by the time-step size. 
+The case of rigid water exemplifies this behavior and serves for an excellent test case for our proposal. 
+In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by \citeauthor{Martyna_1996} \cite{Martyna_1996}.
+
+It should be emphasize that our novel technique is not limited to rigid bodies, neither to the NHC method. 
+The only requirement is imposed on the splitting scheme, which must have the NVE Liouville operator in the middle between the thermostat propagators.
+
+This paper proceeds as follows.
+In Secs.~\ref{sec:nve}-\ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
+This is followed by our results on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive our new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
 Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
-%The starting point is  microcanonical (NVE) ensemble \citep{Silveira_2017} which employs the unit quaternion representation for the rotational degrees of freedom. 
+Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the approximation to the shadow Hamiltonian.
+
 \section{Methodology}
@@
 \end{subequations}
-with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$.
 In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugate momenta of the quaternion.
-Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity in the body-fixed frame.
+Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference; and $\vt \omega$ $=$ $\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
 Matrices $\mt M$ and $\mt I$ are diagonal.
@@
 \end{equation}
-%An advantage of this notation is that
+
 Of course, Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
@@
 The action of $e^{\timestep \Liu{A}}$ is the simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step.
-In Ref.~\citenum{Silveira_2017} we derived a general solution of torque-free rotation in a form that facilitates computer implementation allowing, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
+In Ref.~\citenum{Silveira_2017} we derived an exact solution of torque-free rotation in a form that facilitates the computer implementation allowing, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
 For mathematical details, the reader is referred to our previos work \cite{Silveira_2017}.
 The exact solution comprise ``less familiar'' mathematical functions, such as the Jacobi elliptic functions and Carlson integrals.
-On the other hand, the approximate solutions, such as the NO-SQUISH method  \cite{Miller_2002}, have proven to be very accurate in liquid-phase simulations \cite{Silveira_2017} and its formulation only involve sine and cosine terms.
+On the other hand, the approximate solutions, such as the NO-SQUISH method  \cite{Miller_2002}, have proven to be very accurate in liquid-phase simulations \cite{Silveira_2017} and its formulation only involves sine and cosine terms.
 This may explain why that the approximate solution is more commonly used \cite{vanZon2007}.
@@
 \subsection{NVT Dynamics with a Single Thermostat Chain: Notation and Measure-Preserving Integration}
-
+\label{sec:nvt}
 In this work, we couple a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} to both the translational and rotational degrees of freedom of the rigid-body system.
@@
 Altough it is not possible to devise symplectic integrators for Eq.~\eqref{eq:ODE system for NVT}, one can rigurously obtain measure-preserving numerical solvers \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
-It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure $e^{-w(\vt x)}d\vt x$ of Eq.~\ref{eq:ODE system for NVT} is given by $w = -7N \eta_1 - \sum_{j=2}^M \eta_j$.
-It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics so, in this case, the measure-preserving condition for every individual operator can actually be relaxed.
-The $\eta$ coordinates are often computed for checking how well a numerical integrator conserves $H$.
+It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure $e^{-w(\vt x)}d\vt x$ for Eq.~\ref{eq:ODE system for NVT} is given by $w = -7N \eta_1 - \sum_{j=2}^M \eta_j$.
+It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics so, in this case, the measure-preserving condition for every individual operator could be relaxed.
 
@@
 where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
-The NHC part can be split even further as
+The operator can be split even further as
 \begin{equation*}
@@
 \end{equation*}
-
 In the scheme above, the first propagator to act is $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, which promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
@@
 
-Please note that in Eq.~\ref{eq:generalized force NVT} we have employed Tolman's generalized equipartition theorem \cite{Uline_2008}, which states that
+Please note that Eq.~\ref{eq:generalized force NVT} is closely related to Tolman's generalized equipartition theorem \cite{Uline_2008}, which states that
 \begin{equation*}
@@
 \end{equation*}
-This definition, with the \textit{refined} kinetic energy between brackets, provides a generalization for $h > 0$ of the simulated temperature \cite{Eastwood_2010}.
+By substituting $\modified{\Ham{}}$ in this equation, we obtain that
+\begin{equation}
+2\left\langle \modified{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
+\end{equation}
+This result, with the \textit{refined} kinetic energy between brackets, provides a generalization for $h > 0$ of the simulated temperature \cite{Eastwood_2010}.
 
-Finally, reweighting:
+Finally, given that configurations $x_i$ are sampled from a distribution proportional to $\exp(-\beta \modified{\Ham{}})$, we need to reweight the computed observables $A(x_i)$ to recover the original distribution $\exp (-\beta \Ham{})$.
+This is done as follows:
 \begin{equation}

### 2018-06-06 rigid_bodies_supp_info.tex
results

@@
 of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.}
-    \label{table:stability_martyna}
-    \begin{tabular}{SS[table-number-alignment = center,table-figures-exponent = 1]S[table-number-alignment = center,table-figures-exponent = 1]}
-     & \multicolumn{2}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}  \\
-     \cmidrule[0.5mm](lr){2-3} 
-     $\timestep$ $\text{(fs)}$  & $\modified{H}$  & $\modified{\Ham{}}$     \\
-	1 &  -1.6e-2  & 4.5e-3\\
-	2 & 9.5e-2 & 1.4e-1\\
-	3 & 2.8e-1 & 5.6e-2\\
-	4 & 1.1e0 & 5.6e-2\\
-	5 & 3.3e0 & -3.7e-2\\
-    6 & 6.1e1 & 6.9e-2 \\
-	7 & 9.8e2 & -3.6e-3 \\
-	8 & 1.1e4 & 1.8e-2 \\
-	9 & 7.5e4 & 3.8e-3 \\
-	10 & 3.3e5 & 1.9e-2  \\
+    \label{table:stability}
+        \sisetup{table-number-alignment = center}
+    \begin{tabular}{SS[table-figures-decimal=1,table-figures-exponent = 1]S[table-figures-exponent = 1,table-figures-decimal=1]S[table-figures-decimal=1,table-figures-exponent = 1]S[table-figures-exponent = 1,table-figures-decimal=1]S[table-figures-exponent = 1,table-figures-decimal=1]}
+     & \multicolumn{5}{c}{Drift rate [(kcal/mol)/ns]}  \\
+     & \multicolumn{2}{c}{This work} & \multicolumn{2}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}   & \multicolumn{1}{c}{NVE \cite{Silveira_2017}} \\
+     \cmidrule[0.5mm](lr){2-3} \cmidrule[0.5mm](lr){4-5} \cmidrule[0.5mm](lr){6-6}  
+     $\timestep$ $\; \text{(fs)}$  & $\modified{H}$  & $\modified{\Ham{}}$  & $H$  & $\modified{\Ham{}}$     & $\modified{\Ham{}}$   \\
+	1 &  4.6e-4  & 8.5e-2 & -1.6e-2  & 4.5e-3 & -1.e-3 \\
+	2 & -8.2e-3 & -4.3e-2 &  9.5e-2 & 1.4e-1 &  9.3e-3 \\
+	3 & -3.2e-2 & 7.3e-2 & 2.8e-1 & 5.6e-2 & -5.0e-3 \\
+	4 & 8.1e-3 & -1.1e-1 & 1.1e0 & 5.6e-2 & 8.5e-2 \\
+	5 & 1.3e0 & -2.2e-2 & 3.3e0 & -3.7e-2 & 1.1e0 \\
+    6 & 4.4e1 & -3.4e-3 & 6.1e1 & 6.9e-2 & 6.0e1 \\
+	7 & 8.8e2 & 2.2e-2  & 9.8e2 & -3.6e-3 & \text{-} \\
+	8 & 9.8e3 & 7.6e-2  & 1.1e4 & 1.8e-2 & \text{-} \\
+	9 & 7.1e4 & -2.5e-2  & 7.5e4 & 3.8e-3 & \text{-} \\
+	10 & 3.2e5 & -3.2e-2  & 3.3e5 & 1.9e-2  & \text{-} \\
 	\end{tabular}
 \end{table}
+
+ADD: COMPARE MILLER WITH AND WITHOUT RESPA
+ADD: CORRELATION FUNCTION AS A FUNCTION OF MASS PARAMETER OF THERMOSTAT
+
+\begin{table*}
+	\caption{Effect of the time-step size on ....
+of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes....}
+	\label{table:check_ens}
+	\sisetup{table-number-alignment = center}
+    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]SS[separate-uncertainty,table-figures-uncertainty= 1]SS[separate-uncertainty,table-figures-uncertainty= 1]S}
+	{$\timestep$} & {estimated $\Delta T$} & {$\sigma$ deviations}  & {estimated $\Delta T$}  & {$\sigma$ deviations}  &    {estimated $\Delta T$}  & {$\sigma$ deviations} \\
+	{(fs)} & {(K)} &   & {(K)}  &  & {(K)}  & \\
+			\hline
+			\multicolumn{7}{c}{This work (Unsplit)}\\
+1.0	& 3.02(2) & 1.07 & & & &  \\
+2.0 & 3.02(2) & 0.81 & & & &  \\
+3.0 & 2.98(2) & 0.73 & & & &  \\
+4.0 & 3.01(2) & 0.41 & & & &  \\
+5.0 & 3.02(2) & 0.79 & & & &  \\
+6.0 & 3.03(3) & 1.23 & & & &  \\
+7.0 & 2.98(2) & 0.74 & & & &  \\
+            \hline
+			\multicolumn{7}{c}{Martyna \textit{et al.} \cite{Martyna_1996}}\\
+1.0	& 2.98(2) & 0.66 & 3.01(2) & 0.49 & 2.96(3) & 0.90 \\
+2.0 & 2.98(2) & 0.73 & 2.99(2) & 0.03 & 3.02(3) & 0.66 \\
+3.0 & 2.95(2) & 2.33 & 3.01(2) & 0.65 & 2.91(3) & 2.47 \\
+4.0 & 3.00(2) & 0.01 & 3.01(2) & 0.72 & 2.92(3) & 2.19  \\
+5.0 & 2.93(2) & 3.05 & 3.01(2) & 0.70 & 2.94(3) & 1.58  \\
+6.0 & 2.92(2) & 3.87 & 3.00(2) & 0.22 & 2.91(3) & 2.76  \\
+7.0 & 2.89(2) & 5.35 & 3.02(1) & 1.33 & 2.87(2) & 4.33  \\
+            \hline
+			\multicolumn{7}{c}{Kamberaj \textit{et al.} \cite{Kamberaj_2005}}\\
+1.0	& 2.97(2) & 1.16 & 2.99(2) & 0.16 & 2.89(3) & 2.84  \\
+2.0 & 3.00(2) & 0.02 & 3.02(2) & 0.99 & 2.00(4) & 0.25  \\
+3.0 & 2.99(4) & 0.18 & 3.08(4) & 4.09 & 3.1(1) & 1.17  \\
+4.0 & 2.92(8) & 0.99 & 3.12(2) & 6.01 & 2.8(2) & 0.95  \\
+5.0 & 3.02(5) & 0.40 & 3.2(2)  & 9.50 &  2.9(1) & 0.28  \\
+6.0 & 3.14(5) & 2.69 & 3.29(2) & 14.68 & 3.21(7) & 2.90  \\
+7.0 & 3.03(4) & 0.73 & 3.35(2) & 17.37 & 2.99(6) & 0.16  \\
+    \end{tabular}
+\end{table*}
 

### 2018-06-07 rigid_bodies.tex
Initial revision of the introduction

@@
 }
-\usepackage{xr}
-\externaldocument{rigid_bodies_supp_info}
 % Comment below before submitting
@@
 % Comment above before submitting
+
+\usepackage{xr}
+\externaldocument{rigid_bodies_supp_info}
 \usepackage{booktabs}
@@
 
-\author{Ana J.
-Silveira}
+\author{Ana J. Silveira}
 \email{asilveira@plapiqui.edu.ar}
@@
 
-\author{Charlles R.
-A.
-Abreu}
+\author{Charlles R. A. Abreu}
 \email{abreu@eq.ufrj.br}
@@
 
-\title{Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis}
+\title{Improvement of Thermostated Molecular Dynamics via Backward Error Analysis}
 
@@
 
-Molecular Simulation, being the computational realization of Statistical Mechanics \cite{Tuckerman_2010}, involves several types of approximations in order to mimic ``real'' physical systems.
-Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
-However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
-For instance, it has already been shown that the unavoidable errors in the discretization of the equations of motion disrupt the equipartition of kinetic energy \cite{Eastwood_2010}.
-The same artifact occurs on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics (MD) \cite{Davidchack_2009,Silveira_2017}.
-This finding brings into question the NVT integrator developed by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-In addition, as will be shown later, this method fails to reproduce the specified temperature for time-step sizes usually employed in MD.
+Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
+Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the force field.
+However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
+For instance, it has been shown \cite{Eastwood_2010} that errors in the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
+A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
+This brings into question the NVT-MD integrator developed by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
+In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
+In fact, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
+Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
+This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods routinely used in MD are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.
 
-The relevance of rigid-body MD stems from the fact that not only some molecules are usually modelled as rigid bodies, like the ubiquitous case of water \cite{Jorgensen_1983}, but also it can be employed as a coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, such as proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
-As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the software packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes\cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles \cite{Patra_2014}.
+For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
+The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
+Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
+As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
+The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
+Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.
 
-The Nos\'{e}-Hoover chain (NHC) thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach, in which auxiliary degrees of freedom act regulating the kinetic energy of the physical variables so as to be consistent with the Maxwell-Boltzmann distribution at a given target temperature.
-This requires an expression relating the temperature, a macroscopic parameter, with the kinetic energy, which is a property of each microstate.
-Usually, in MD the equipartition theorem plays this role. 
-However, as already mentioned, this theorem is disrupted in numerical schemes with non-negligible discretization errors.
-This is because numerical trajectories actually sample from the so-called \textit{shadow} Hamiltonian, which cannot be merely separated into momentum and position dependent parts.
-Instead, what is usually identified as the \textit{shadow} kinetic energy contains quadratic terms in particle momenta accompanied by particle positions dependency \cite{Eastwood_2010}.
-It seems then reasonable to argue that the thermostat should control this kinetic energy, instead of the physical one.
-However, the theory underlying the concept of shadow Hamiltonian correspond to backward error analysis, which is only applicable to symplectic integrators.
-This is not the case of NHC numerical schemes, given that the auxiliary variables introduce non-Hamiltonian components to the equations of motion.
-In this paper we introduce a procedure that partially overcomes this limitation by performing the backward error analysis on the NVE part of the numerical integrator with minimal computational overheads.
-The resulting NHC scheme, which simultaneously couples the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, controls the fluctuations of an approximation to the \textit{shadow} kinetic energy.
+It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified, ``physical'' one.
+However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
+In this paper, we devise a new procedure that partially overcomes this limitation.
+It consists in using a splitting-based numerical integrator that entails a Hamiltonian component, and then carrying out backward error analysis on this NVE part only.
+This allows us to evaluate, with minimal computational overhead, a \textit{refined} kinetic energy for the due replacement in the thermostat equations.
+It is worth mentioning that all steps of this procedure are theoretically justified.
 
 Our new method is particularly appealing for rigid-body MD, primarily because discretization errors begin to manifest much earlier than any numerical stability caused by the time-step size.
+
+Rigid-body MD is relevant not only because some molecules are usually modeled as rigid bodies, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can be employed as a coarse-graining strategy for molecules that can be designed as collection of interconnected rigid bodies, such as proteins and nanoparticles \cite{Knorowski_2012, Patra_2013}.
+In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, for instance, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
+
 The case of rigid water exemplifies this behavior and serves for an excellent test case for our proposal. 

### 2018-06-07 rigid_bodies.tex
Further revision of the introduction

@@
 However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
-In this paper, we devise a new procedure that partially overcomes this limitation.
-It consists in using a splitting-based numerical integrator that entails a Hamiltonian component, and then carrying out backward error analysis on this NVE part only.
-This allows us to evaluate, with minimal computational overhead, a \textit{refined} kinetic energy for the due replacement in the thermostat equations.
-It is worth mentioning that all steps of this procedure are theoretically justified.
+In this paper, we devise a novel procedure that partially overcomes this limitation.
+It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis on this NVE part only.
+This allows us to evaluate, with minimal computational overhead, a \textit{refined kinetic energy} for replacement in the thermostat equations.
+As we will demonstrate, this procedure succeeds in both predicting equipartition and reproducing a canonical ensemble consistent with the \textit{refined Hamiltonian}.
+Finally, reweighting is done so that ensemble averages are consistent with the specified one.
 

### 2018-06-07 rigid_bodies.tex
New proposal for the introduction

@@
 However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
-For instance, it has been shown \cite{Eastwood_2010} that errors in the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
+For instance, it has been shown \cite{Eastwood_2010} that errors due to the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
 A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
@@
 
-It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified, ``physical'' one.
+It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified one.
 However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
 In this paper, we devise a novel procedure that partially overcomes this limitation.
-It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis on this NVE part only.
-This allows us to evaluate, with minimal computational overhead, a \textit{refined kinetic energy} for replacement in the thermostat equations.
-As we will demonstrate, this procedure succeeds in both predicting equipartition and reproducing a canonical ensemble consistent with the \textit{refined Hamiltonian}.
-Finally, reweighting is done so that ensemble averages are consistent with the specified one.
+It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis out of this NVE part only.
+This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
+By recasting the thermostat equations with this refined kinetic energy, we are able to predict its equipartition and to reproduce a canonical ensemble consistent with the refined Hamiltonian.
+Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one.
+It should be emphasized that this approach is not limited to deterministic thermostats.
+It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of \citeauthor{Bussi_2007} \cite{Bussi_2007} and \citeauthor{Leimkuhler_2009} \cite{Leimkuhler_2009}, for instance.
 
-Our new method is particularly appealing for rigid-body MD, primarily because discretization errors begin to manifest much earlier than any numerical stability caused by the time-step size.
+Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
+Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
+In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
 
-Rigid-body MD is relevant not only because some molecules are usually modeled as rigid bodies, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can be employed as a coarse-graining strategy for molecules that can be designed as collection of interconnected rigid bodies, such as proteins and nanoparticles \cite{Knorowski_2012, Patra_2013}.
-In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, for instance, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
-
-The case of rigid water exemplifies this behavior and serves for an excellent test case for our proposal. 
-In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by \citeauthor{Martyna_1996} \cite{Martyna_1996}.
-
-It should be emphasize that our novel technique is not limited to rigid bodies, neither to the NHC method. 
-The only requirement is imposed on the splitting scheme, which must have the NVE Liouville operator in the middle between the thermostat propagators.
+The case of rigid water exemplifies this behavior and serves as an excellent testbed for our proposal.
+In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
 

### 2018-06-07 rigid_bodies.tex
results rev I

--- 2018-06-07 rigid_bodies.tex
@@
 }
+\usepackage{xr}
+\externaldocument{rigid_bodies_supp_info}
 % Comment below before submitting
@@
 % Comment above before submitting
-
-\usepackage{xr}
-\externaldocument{rigid_bodies_supp_info}
 \usepackage{booktabs}
@@
 
-\title{Improvement of Thermostated Molecular Dynamics via Backward Error Analysis}
+\title{Improvement of Deterministic NVT Molecular Dynamics via Backward Error Analysis}
 
@@
 
-Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
-Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the force field.
-However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
-For instance, it has been shown \cite{Eastwood_2010} that errors in the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
-A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
-This brings into question the NVT-MD integrator developed by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
-In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
-In fact, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
-Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
-This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods routinely used in MD are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.
+Molecular Simulation, being the computational realization of Statistical Mechanics \cite{Tuckerman_2010}, involves several types of approximations in order to mimic ``real'' physical systems.
+Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
+However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
+For instance, it has already been shown that the unavoidable errors in the discretization of the equations of motion disrupt the equipartition of kinetic energy \cite{Eastwood_2010}.
+The same artifact occurs on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics (MD) \cite{Davidchack_2009,Silveira_2017}.
+This finding brings into question the NVT integrator developed by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
+In addition, as will be shown later, this method fails to reproduce the specified temperature for time-step sizes usually employed in MD.
 
-For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
-The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
-Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
-As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
-The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
-Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.
+The relevance of rigid-body MD stems from the fact that not only some molecules are usually modelled as rigid bodies, like the ubiquitous case of water \cite{Jorgensen_1983}, but also it can be employed as a coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, such as proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
+As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the software packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes\cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles \cite{Patra_2014}.
 
-It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified, ``physical'' one.
-However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
-In this paper, we devise a new procedure that partially overcomes this limitation.
-It consists in using a splitting-based numerical integrator that entails a Hamiltonian component, and then carrying out backward error analysis on this NVE part only.
-This allows us to evaluate, with minimal computational overhead, a \textit{refined} kinetic energy for the due replacement in the thermostat equations.
-It is worth mentioning that all steps of this procedure are theoretically justified.
+The Nos\'{e}-Hoover chain (NHC) thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach, in which auxiliary degrees of freedom act regulating the kinetic energy of the physical variables so as to be consistent with the Maxwell-Boltzmann distribution at a given target temperature.
+This requires an expression relating the temperature, a macroscopic parameter, with the kinetic energy, which is a property of each microstate.
+Usually, in MD the equipartition theorem plays this role. 
+However, as already mentioned, this theorem is disrupted in numerical schemes with non-negligible discretization errors.
+This is because numerical trajectories actually sample from the so-called \textit{shadow} Hamiltonian, which cannot be merely separated into momentum and position dependent parts.
+Instead, what is usually identified as the \textit{shadow} kinetic energy contains quadratic terms in particle momenta accompanied by particle positions dependency \cite{Eastwood_2010}.
+It seems then reasonable to argue that the thermostat should control this kinetic energy, instead of the physical one.
+However, the theory underlying the concept of shadow Hamiltonian correspond to backward error analysis, which is only applicable to symplectic integrators.
+This is not the case of NHC numerical schemes, given that the auxiliary variables introduce non-Hamiltonian components to the equations of motion.
+In this paper we introduce a procedure that partially overcomes this limitation by performing the backward error analysis on the NVE part of the numerical integrator with minimal computational overheads.
+The resulting NHC scheme, which simultaneously couples the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, controls the fluctuations of an approximation to the \textit{shadow} kinetic energy.
 
 Our new method is particularly appealing for rigid-body MD, primarily because discretization errors begin to manifest much earlier than any numerical stability caused by the time-step size. 
-
-Rigid-body MD is relevant not only because some molecules are usually modeled as rigid bodies, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can be employed as a coarse-graining strategy for molecules that can be designed as collection of interconnected rigid bodies, such as proteins and nanoparticles \cite{Knorowski_2012, Patra_2013}.
-In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, for instance, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
-
 The case of rigid water exemplifies this behavior and serves for an excellent test case for our proposal. 
@@
 
-In the following, we analyze the performance of the proposed NVT integrators introduced in Secs.
-\ref{sec:mttk}-\ref{sec:refined_method}, and  compare the results with those obtained employing the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-It should be noted that all the simulations were carried out considering $t_d = 100$ fs.
+In the following, we analyze the performance of the NVT scheme proposed in  this paper and compare the results with those obtained employing the measure-preserving version of the integrator due to \citeauthor{Martyna_1996} \cite{Martyna_1996} as well as the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}
+
 The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
-The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA}.\cite{Silveira_2017} 
+The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA} as done in Ref.~\citenum{Silveira_2017}.
 The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.
+In addition, all the simulations were carried out considering $t_d = 100$ fs.
 
-Before showing our main results, it may be instructive first to give a picture of the problem in controlling the temperature when discretization errors are non-negligible. 
-The most direct way (the most senstive indicator of the presence of discretization errors) to determine the importance of these errors is by examining the departure of the ``modified Hamiltonian'' $\modified{\Ham{}}$ from the physical Hamiltonian $\Ham{}$\cite{Engle_2005}.
-Fortunately, with the analytical form of $\modified{\Ham{}}$ derived in this paper we are able to perform such an analysis for the NVE integrator which is essential, as will be shown below, to understand the results of the NVT methods.
-In Fig.~\ref{fig:nve} we depict the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
-The numerical integrator correspond to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation. 
-As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
+Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}.
+This is essential for the discussion of the NVT case.
+
+In Fig.~\ref{fig:nve} we show the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
+For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation.
+As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$.
 It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
-As a result of employing such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).
+As a result of such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).
 
@@
 	\includegraphics{Figures/NVE.eps}
-	\caption{Total energy $\Ham{}$ (solid lines) and modified Hamiltonian $\modified{\Ham{}}$ (dotted lines) of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVE MD simulations for different time-step sizes $\timestep$.
-The numerical integrator correspond to Eq.~\ref{eq:splitting NVE} with the exact solution of free rotation.}
+	\caption{Total energy $\Ham{}$ (solid lines) and \textit{refined} Hamiltonian $\modified{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit rotation method.}
 	\label{fig:nve}
-\end{figure}
-
-For our implementation for rigid bodies of the numerical scheme due to Martyna \textit{et al.} \cite{Martyna_1996} (Eq.~\ref{eq:splitting NVT}), in Fig.~\ref{fig:props_k_m} we depict the computed averages of temperature ($\langle T \rangle$) and potential energy per molecule ($\langle U \rangle$) as a function of the time-step size.
-The specified temperature corresponds to T = 298 K.
-Fig.~\ref{fig:props_k_m} clearly shows that a suitable temperature control leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
-As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
-Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
-These observations motivated the development of the scheme of Sec.~\ref{sec:refined_method}, whose promising results we show below.
-
-\begin{figure}
-	\includegraphics{Figures/thermodynamic_properties_intro.eps}
-    \caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
-	\label{fig:props_k_m}
 \end{figure}
@@
 
-The explicit inclusion of the ``refined'' Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
+The explicit inclusion of the \textit{refined} Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
 This poses the question whether the symmetry loss will influence the long-term stability.
-In this paper, we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
+
+Here we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
 Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
@@
 In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
-Also, the long-term stability is only modestly improved by our method so, evidently, the lack of symmetry has a negligible impact on $R$.
-Fig.~\ref{fig:energy_drift} also contains the drift in $\Ham{}$ computed for the NVE method\cite{Silveira_2017}, which results to be almost identical to our method.
+Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
+Fig.~\ref{fig:energy_drift} also contains the drift in $\modified{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method.
 It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
 The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
-We have computed $R$ also for $\Ham{}$, which is the contribution to the extended energy from the physical degrees of freedom, namely the NVE part of the integrator.
-In this case, $\Ham{}$ remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration, which is also true for the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Table~\ref{table:stability} of Supp.
-Material).
+As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the \textit{refined} Hamiltonian part for our NVT method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
 This has already been noticed by Davidchack\cite{Davidchack_2009}, and previously by Bond \textit{et al.} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.
@@
 	\includegraphics{Figures/energy_drift.eps}
-    \caption{Effect of the time-step size of 903 TIP3P \cite{Jorgensen_1983} water molecules in NVT MD simulations.
-$R$ for $\widetilde{H}$ and $\Ham{}$, which is the contribution to the extended energy from the physical degrees of freedom.
-We included the results for the integrator due to Martyna \textit{et al.} \cite{Martyna_1996}
-}
+    \caption{Influence of the time-step size ($h$) on the long-time drift ($R$) in: the extended energy ($\modified{H}$, down triangles) and its \textit{refined} Hamiltonian ($\modified{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\modified{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}
@@
 As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} also implemented such a separate thermostatization and discovered the existence of energy flows from one thermostat to the other.
+Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows from one thermostat to the other.
 To the best of our knowledge, the work of Davidchack \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
@@
 	\includegraphics{Figures/numerical_stability.eps}
-    \caption{Total extended energy of 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
-step sizes, employing (a) the proposed method represented by Eq.
-; (b) our implementation for rigid bodies of the integrator due to Martyna \textit{et al.} \cite{Martyna_1996} and (c) the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.Energy of thermostats for 903 TIP3P\cite{Jorgensen_1983} water molecules obtained by NVT MD simulations with different time
-step sizes, employing the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.}
+    \caption{Long-time drift in (a) the extended energy and (b) in each thermostat energy for the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing to the thermostat coupled to the rotational degrees of freedom.}
 	\label{fig:num_stab}
@@
 However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement.
+
+The specified temperature corresponds to T = 298 K.
+Fig. clearly shows that a temperature control as described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
+As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
+Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
+These observations motivated the development of the scheme of Sec.~\ref{sec:refined_method}, whose promising results we show below.
 
@@
 
-%It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to prove that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble.
-%First we present some results as first mented.
-%That is, the proposed methods coupling to a single thermostat, while in Kamberaj the translational and rotational degrees of freedom are separatedly coupled.
-%In Table~\ref{table:new_results}, it can be seen that the integrators proposed here, which are represented by the Eqs.~\eqref{eq:trotter_splitting_NHC} and ~\eqref{eq:modified_splitting}, systemcatically achive the specifed temperature, regardless of the time step size.
-%By contrast, the numerical solver introduced by Kamberaj and co-workers\cite{Kamberaj_2005} fails to yield the specified temperature for even the smaller time step size of 1 fs.
-%Resultados sobre o checkensemble:
-
-%checkensemble:We report the results corresponding to that method, and since it is instructive to have a graphical insight, we also show some figures of the linear fitting.
 %Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
@@
 
-%Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$ of Eq.~\eqref{eq:H_nvt}, in Table~\ref{table:new_results_2} we report the energy drift, denoted by $R$, as the the rate of change of $H$ (kcal/mol) with time (ns), which is the angular coefficient obtained from linear regression fitting  between those variables.
-%Except for the case of $\Delta$ = 1 fs, there is also a clear correspondece between an incresing $R$ with the failure in keeping the temperature at its specified value.
-
-%In Fig.~\ref{fig:checkensemble} we show the result from the linear regression analysis for the potential energies sampled using Eq.~\eqref{eq:trotter_splitting_NHC}.
-%In figure legends we show the averages temperatures simulated.
 %It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-%Likewise, in Fig.~\ref{fig:mbdistribution} we observe a good correspondence between the kinetic energy distribution at $T= 304.99 K$ and the corresponding Maxwell-Boltzmann distribution We arrive at the same conclusions for the approach given by Eq.~\eqref{eq:modified_splitting}.
-
 %Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
@@
 \end{document}
-%A geometric integrator is a numerical method which preserves certain geometric structures of the exact flow of a differential equation.
-%the issues arising in the design and analysis of numerical methods for molecular dynamics
-%(symplecticness) of the phase flow.
-We will later show (in Chap.
-3) that such symplectic methods preserve a perturbed energy-like invariant (to very high accuracy); the preservation of this modified energy then ensures that the original energy is nearly conserved
-%A benefit of respecting the symplectic structure lies in the prospect of using backward error analysis to analyze the behavior of the method
-%It seems intuitively correct to demand that a numerical method should mimic as much as possible the available structure of the dynamical system, when this is possible to do without incurring large computational overheads in the process.
-So much the better that discrete dynamical systems designed in this way appear to be related to modified continuous dynamics with appropriate properties.
-This observation is purely aesthetic.
-%These methods obtain their energy preservation properties indirectly: we design the methods to exactly conserve a certain geometric property
-%The purpose of this section is to demonstrate that, by using the principles of geometric integration and the backward error analysis, it is possible to compute effective estimates of the error in numerically computed averages, and even to correct those averages by a straightforward reweighting

### 2018-06-07 rigid_bodies.tex

@@
 
-Molecular Simulation, being the computational realization of Statistical Mechanics \cite{Tuckerman_2010}, involves several types of approximations in order to mimic ``real'' physical systems.
-Naively, one would expect those approximations to only produce quantitative deviation from experimental values, and usually we tend to attribute any disagreement to a lack of accuracy of the force field.
-However, what can be hidden in the deviations are actually systematic errors, which may introduce non-physical behavior.
-For instance, it has already been shown that the unavoidable errors in the discretization of the equations of motion disrupt the equipartition of kinetic energy \cite{Eastwood_2010}.
-The same artifact occurs on the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body Molecular Dynamics (MD) \cite{Davidchack_2009,Silveira_2017}.
-This finding brings into question the NVT integrator developed by Kamberaj and co-workers\cite{Kamberaj_2005} which, assuming equipartition, couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-In addition, as will be shown later, this method fails to reproduce the specified temperature for time-step sizes usually employed in MD.
+Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
+Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the force field.
+However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
+For instance, it has been shown \cite{Eastwood_2010} that errors due to the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
+A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
+This brings into question the NVT-MD integrator developed by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
+In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
+In fact, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
+Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
+This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods routinely used in MD are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.
 
-The relevance of rigid-body MD stems from the fact that not only some molecules are usually modelled as rigid bodies, like the ubiquitous case of water \cite{Jorgensen_1983}, but also it can be employed as a coarse-graining strategy for systems that might be designed as collection of interconected rigid bodies, such as proteins and nanoparticules \cite{Knorowski_2012, Patra_2013}.
-As a matter of fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented, for instance, in the software packages HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes\cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles\cite{Yan_2008} and nanoparticles \cite{Patra_2014}.
+For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
+The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
+Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
+As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
+The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
+Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.
 
-The Nos\'{e}-Hoover chain (NHC) thermostat \cite{Martyna_1992}, routinely used in MD, is based on the extended phase-space approach, in which auxiliary degrees of freedom act regulating the kinetic energy of the physical variables so as to be consistent with the Maxwell-Boltzmann distribution at a given target temperature.
-This requires an expression relating the temperature, a macroscopic parameter, with the kinetic energy, which is a property of each microstate.
-Usually, in MD the equipartition theorem plays this role. 
-However, as already mentioned, this theorem is disrupted in numerical schemes with non-negligible discretization errors.
-This is because numerical trajectories actually sample from the so-called \textit{shadow} Hamiltonian, which cannot be merely separated into momentum and position dependent parts.
-Instead, what is usually identified as the \textit{shadow} kinetic energy contains quadratic terms in particle momenta accompanied by particle positions dependency \cite{Eastwood_2010}.
-It seems then reasonable to argue that the thermostat should control this kinetic energy, instead of the physical one.
-However, the theory underlying the concept of shadow Hamiltonian correspond to backward error analysis, which is only applicable to symplectic integrators.
-This is not the case of NHC numerical schemes, given that the auxiliary variables introduce non-Hamiltonian components to the equations of motion.
-In this paper we introduce a procedure that partially overcomes this limitation by performing the backward error analysis on the NVE part of the numerical integrator with minimal computational overheads.
-The resulting NHC scheme, which simultaneously couples the translational and rotational degrees of freedom to a unique Nos\'{e}-Hoover chain thermostat, controls the fluctuations of an approximation to the \textit{shadow} kinetic energy.
+It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified one.
+However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
+In this paper, we devise a novel procedure that partially overcomes this limitation.
+It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis out of this NVE part only.
+This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
+By recasting the thermostat equations with this refined kinetic energy, we are able to predict its equipartition and to reproduce a canonical ensemble consistent with the refined Hamiltonian.
+Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one.
+It should be emphasized that this approach is not limited to deterministic thermostats.
+It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of \citeauthor{Bussi_2007} \cite{Bussi_2007} and \citeauthor{Leimkuhler_2009} \cite{Leimkuhler_2009}, for instance.
 
-Our new method is particularly appealing for rigid-body MD, primarily because discretization errors begin to manifest much earlier than any numerical stability caused by the time-step size. 
-The case of rigid water exemplifies this behavior and serves for an excellent test case for our proposal. 
-In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming the method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, as well as our implementation for rigid bodies of the numerical scheme developed by \citeauthor{Martyna_1996} \cite{Martyna_1996}.
+Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
+Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
+In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
 
-It should be emphasize that our novel technique is not limited to rigid bodies, neither to the NHC method. 
-The only requirement is imposed on the splitting scheme, which must have the NVE Liouville operator in the middle between the thermostat propagators.
+The case of rigid water exemplifies this behavior and serves as an excellent testbed for our proposal.
+In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
 

### 2018-06-08 rigid_bodies.tex

@@
 
-\begin{tocentry}
-	Graphical Abstract
-\end{tocentry}
+%\begin{tocentry}
+%\end{tocentry}
 
-\begin{abstract}
-	Abstract.
-\end{abstract}
+%\begin{abstract}
+%\end{abstract}
 
@@
 In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
-
 The case of rigid water exemplifies this behavior and serves as an excellent testbed for our proposal.
-In fact, we are able to eliminate the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
+In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
 
@@
 In Secs.~\ref{sec:nve}-\ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
-This is followed by our results on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive our new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
-Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
-Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the approximation to the shadow Hamiltonian.
+This is followed by our developments on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive our new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
+Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}. Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the shadow Hamiltonian approximation.
 
@@
 
-\subsection{NVE Dynamics: Notation and Symplectic Integration}
+\subsection{Hamiltonian Dynamics with Rigid Bodies: Notation and Symplectic Integration}
 \label{sec:nve}
+
 In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
-1) derive a new exact solution for torque-free rotations and use it as part of a symplectic numerical integrator for the motion of rigid bodies and
-2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002}, thus unveiling its equivalence to a method previously developed by Dullweber and co-workers \cite{Dullweber_1997}.
+1) derive a new exact solution for torque-free rotations and use it as part of a symplectic integrator for rigid bodies and
+2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} and unveil its equivalence to the earlier method by Dullweber and co-workers \cite{Dullweber_1997}.
 
@@
 	%-\diff{\Ham{}}{\vt q} =
-	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
+	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau. \label{eq:EDO_pi}
 	\end{align}
 \end{subequations}
-In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the conjugate momenta of the quaternion.
-Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference; and $\vt \omega$ $=$ $\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
-Matrices $\mt M$ and $\mt I$ are diagonal.
+
+In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
+Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega$ $=$ $\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
+Matrices $\mt M$ and $\mt I$ are diagonal ones.
 The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
@@
 
-The factorization mentioned above, which played a central role in Ref.~\citenum{Silveira_2017}, is as simple as ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, with ${\mt A}(\vt q)$ being the orientation matrix of the body (that is, the rotation matrix connecting the body-fixed and space-fixed frames).
+The factorization which played a central role in Ref.~\citenum{Silveira_2017} is ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, where ${\mt A}(\vt q)$ is the rotation matrix that connects the body-fixed and space-fixed frames.
 
@@
 
-Of course, Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
-The same occurs for diagonal matrices $\mt M$ and $\mt I$, whose sizes increase to $3N \times 3N$ in this case.
+Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
+The size of diagonal matrices $\mt M$ and $\mt I$ also increases to $3N \times 3N$ in this case.
 In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
-Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further to accommodate a number of freely moving atoms.
+Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further so as to accommodate a number of freely moving atoms.
 
@@
 where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
-In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state immediately before the propagation occurs.
-The action of $e^{\timestep \Liu{A}}$ is the simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation over a full time step.
-In Ref.~\citenum{Silveira_2017} we derived an exact solution of torque-free rotation in a form that facilitates the computer implementation allowing, differently to existing schemes \cite{Kosenko_1998,vanZon2007,Celledoni_2008}, a unified treatment for both asymmetric ($I_1$ $>$ $I_2$ $>$ $I_3$) and symmetric tops (either $I_1$ = $I_2$ or $I_2$ = $I_3$).
-For mathematical details, the reader is referred to our previos work \cite{Silveira_2017}.
-The exact solution comprise ``less familiar'' mathematical functions, such as the Jacobi elliptic functions and Carlson integrals.
-On the other hand, the approximate solutions, such as the NO-SQUISH method  \cite{Miller_2002}, have proven to be very accurate in liquid-phase simulations \cite{Silveira_2017} and its formulation only involves sine and cosine terms.
-This may explain why that the approximate solution is more commonly used \cite{vanZon2007}.
-This pictures completely changes in the context of backward error analysis: in the approximate methods, an overall rotation is given by a sequence of uniform rotations around the principal axes of a body, which results in a factorization of the rotational operator given by $ e^{\frac{h}{2} i\!L_3} e^{\frac{h}{2} i\!L_2} e^{h i\!L_1} e^{\frac{h}{2} i\!L_2} e^{\frac{h}{2} i\!L_3}$, with each factor representing a rotation around a given axes.
-The analytical solution, on the other hand, entails a single operator, thus making the application of backward error analysis considerably simpler.
+In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state of the system immediately before the propagation occurs.
+The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
+We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
+This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
+Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
+This justifies its more widespread use.
+However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:modified_h} and Appendix \ref{sec:rigid body shadow hamiltonian}.
 

### 2018-06-08 rigid_bodies.tex
revision num stability

@@
 
-In the following, we analyze the performance of the NVT scheme proposed in  this paper and compare the results with those obtained employing the measure-preserving version of the integrator due to \citeauthor{Martyna_1996} \cite{Martyna_1996} as well as the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}
+In the following, we analyze the performance of the NVT scheme proposed in  this paper and compare the results with those obtained employing the integrator due to \citeauthor{Martyna_1996} \cite{Martyna_1996} as well as the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}
 
@@
 
-Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}.
-This is essential for the discussion of the NVT case.
-
+Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}, which is essential for discussing the NVT case.
 In Fig.~\ref{fig:nve} we show the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
@@
 As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows from one thermostat to the other.
-To the best of our knowledge, the work of Davidchack \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat for different time-step sizes, that flux is also observed for the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows between the two thermostats.
+To the best of our knowledge, this finding did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat, that flux is also observed for the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
 In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.

### 2018-06-08 rigid_bodies.tex
Revision of the NVT integration section

--- 2018-06-08 rigid_bodies.tex
@@
 In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
-Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega$ $=$ $\frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
+Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
 Matrices $\mt M$ and $\mt I$ are diagonal ones.
@@
 The factorization which played a central role in Ref.~\citenum{Silveira_2017} is ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, where ${\mt A}(\vt q)$ is the rotation matrix that connects the body-fixed and space-fixed frames.
-
-The Hamiltonian conserved by Eq.~\eqref{eq:ODE system for NVE} is
+Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
 \begin{equation}
@@
 \label{sec:nvt}
+
 In this work, we couple a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} to both the translational and rotational degrees of freedom of the rigid-body system.
-To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momenta $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$.
-The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an ``extended energy'' given by \cite{Martyna_1992}
+To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j = 1, \cdots, M$.
+The flow in the extended phase-space no longer conserves the Hamiltonian $\Ham{}$, but an \textit{extended energy} given by \cite{Martyna_1992}
 \begin{equation}
@@
 
-In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
+In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $j \le M$ and $\alpha_{M+1} = 0$, while $G_j$ is a generalized force acting on thermostat $j$, defined as
 \begin{subequations}
@@
 \end{subequations}
+
 With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
+If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as a driven variables \cite{Tuckerman_2001}.
+Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
+In the general case, by taking all these facts into account and applying the analysis of \citeauthor{Tuckerman_2001} \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
+In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.
 
-If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are integrals of motion, while the system's center-of-mass position can be considered as a driven variable,\cite{Tuckerman_2001} such as the $\eta$ coordinates.
-Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
-In the general case, by taking all these facts into account and applying the analysis of Ref.~\citenum{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if one makes $L = 6N$.
-In the particular situation in which one sets $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of a simulation, one must make $L = 6N - 3$ \cite{Martyna_1994}.
+Eq.~\eqref{eq:ODE system for NVT} defines an invariant measure in the extended phase-space \cite{Tuckerman_1999}, and it is possible to devise numerical solvers that preserve such measure \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
+We have done it by adapting the integrator introduced by \citeauthor{Martyna_1996} \cite{Martyna_1996} in light of a criterion explained by \citeauthor{Ezra_2006} \cite{Ezra_2006}.
+It is worth noting, however, that such adaptation required us to alter the integration of the $\eta$ coordinates only.
+Since these are driven variables, with no influence on the dynamics of the system, we can conclude that the practical importance of such alteration is small.
 
-Altough it is not possible to devise symplectic integrators for Eq.~\eqref{eq:ODE system for NVT}, one can rigurously obtain measure-preserving numerical solvers \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
-It can be shown that the function $w(\vt x)$ that determines the invariant phase-space measure $e^{-w(\vt x)}d\vt x$ for Eq.~\ref{eq:ODE system for NVT} is given by $w = -7N \eta_1 - \sum_{j=2}^M \eta_j$.
-It is worth noting that the $\eta$ coordinates are driven variables with no influence on the dynamics so, in this case, the measure-preserving condition for every individual operator could be relaxed.
-
-By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e.
-non-Hamiltonian) Liouville operator, the splitting goes as
+By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
 \begin{equation}
@@
 \end{equation}
-where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
-The operator can be split even further as
+where $\Liu{\tiny NVE}$ is the same operator of Sec.~\ref{sec:nve} and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
+This one can be split even further as
 \begin{equation*}
@@
 \end{equation*}
+
 In the scheme above, the first propagator to act is $e^{\frac{\timestep}{4n} \mathcal{L}_M}$, which promotes a kick in the momentum of thermostat $M$ expressed as $p_{\eta_M} = p_{\eta_M}^\ast + G_M^\ast \frac{\timestep}{4n}$.
@@
 Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_j}$ are applied once again, but now with index $j$ ascending from $1$ to $M$.
-This numerical scheme is simply a measure-preserving version of the NVT integrator introduced by \citeauthor{Martyna_1996} \cite{Martyna_1996}.
 
-For having a low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort, making it feasible to increase the time step $\timestep$ without degrading the accuracy of the NHC part.
+Due to its low computational cost, the whole procedure described above can be repeated $n$ times with little impact on the overall integration effort.
+This makes it feasible to increase the time step $\timestep$ without ruining the accuracy of the NHC part.
 In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
 A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
-The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
-Here we choose to take a simpler path.
-Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied deviations, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when computing ensemble averages.
+\citeauthor{Omelyan_2008}'s processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
+Here we choose to take a simpler path, though.
+Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble average are computed.
 

### 2018-06-08 rigid_bodies.tex
Change command from modified to refined

@@
 \newcommand{\timestep}{h}
-\newcommand{\modified}[1]{\widetilde{#1}}
+\newcommand{\refined}[1]{\widetilde{#1}}
 
@@
 
-As demonstrated in Appendix~\ref{sec:rigid body shadow hamiltonian}, the \textit{refined} Hamiltonian for a Verlet-type method applied to a system of rigid bodies is given by $\modified{\Ham{}} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are \textit{refined} kinetic and potential energies.
+As demonstrated in Appendix~\ref{sec:rigid body shadow hamiltonian}, the \textit{refined} Hamiltonian for a Verlet-type method applied to a system of rigid bodies is given by $\refined{\Ham{}} = \refined K + \refined U + \mathcal{O}(h^4)$, where $\refined K$ and $\refined U$ are \textit{refined} kinetic and potential energies.
 The latter is given by
@@
 \label{eq:modified potential energy}
-\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
+\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
 \end{equation}
-Note that $\modified U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
+Note that $\refined U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
 In turn, the \textit{refined} kinetic energy can be expressed as
@@
 \label{eq:modified kinetic energy}
-\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
 \end{equation}
-where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
+where $\refined{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
 \begin{equation*}
@@
 \begin{equation*}
-\dot{\vt r} = \diff{\modified K}{\vt p}, \;
-\dot{\vt q} = \diff{\modified K}{\vt \pi}, \;
-\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}, \; \text{and} \;
-\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}.
+\dot{\vt r} = \diff{\refined K}{\vt p}, \;
+\dot{\vt q} = \diff{\refined K}{\vt \pi}, \;
+\dot{\vt p} = -\diff{\refined{\Ham{}}}{\vt r}, \; \text{and} \;
+\dot{\vt \pi} = -\diff{\refined{\Ham{}}}{\vt q}.
 \end{equation*}
@@
 \label{eq:shadow ODE system}
-\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
+\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] = \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right].
 \end{equation}
 
-Analytical calculation of $\modified U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
-In contrast, although it is possible to compute $\modified K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
+Analytical calculation of $\refined U$ via Eq.~\eqref{eq:modified potential energy} is straightforward.
+In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
 Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
@@
 \label{eq:modified kinetic energy estimator}
-\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
+\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
 \end{equation}
@@
 Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
-Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \modified{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
+Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
 Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
-The only required modification is that $\modified K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
+The only required modification is that $\refined K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
 By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
@@
 \label{eq:thermostat_force}
-G_1 = 2 \modified K - L k_\text{\tiny B} T.
+G_1 = 2 \refined K - L k_\text{\tiny B} T.
 \end{equation}
@@
 \label{eq:this extended energy}
-\widetilde{H} = \modified K + \modified U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
+\widetilde{H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
 \end{equation}
 
-One instant in which $\modified K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
+One instant in which $\refined K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
 Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
@@
 
-In addition, $\modified K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
-Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\modified K$ goes as simply as ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$.
+In addition, $\refined K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
+Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ goes as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.
 
@@
 \begin{equation*}
-\left\langle \tr{\vt p} \diff{\modified{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\modified{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T.
+\left\langle \tr{\vt p} \diff{\refined{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\refined{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T.
 \end{equation*}
-By substituting $\modified{\Ham{}}$ in this equation, we obtain that
+By substituting $\refined{\Ham{}}$ in this equation, we obtain that
 \begin{equation}
-2\left\langle \modified{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
+2\left\langle \refined{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
 \end{equation}
@@
 
-Finally, given that configurations $x_i$ are sampled from a distribution proportional to $\exp(-\beta \modified{\Ham{}})$, we need to reweight the computed observables $A(x_i)$ to recover the original distribution $\exp (-\beta \Ham{})$.
+Finally, given that configurations $x_i$ are sampled from a distribution proportional to $\exp(-\beta \refined{\Ham{}})$, we need to reweight the computed observables $A(x_i)$ to recover the original distribution $\exp (-\beta \Ham{})$.
 This is done as follows:
@@
 \label{eq:reweighting}
-\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
+\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
 \end{equation}
@@
 
-In Fig.~\ref{fig:nve} we show the time evolution of both $\modified{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
+In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
 For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation.
-As expected, $\modified{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$.
-It should be remarked that $\modified{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
-As a result of such approximation, we have larger fluctuations in $\modified{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).
+As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$.
+It should be remarked that $\refined{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
+As a result of such approximation, we have larger fluctuations in $\refined{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).
 
@@
 	\includegraphics{Figures/NVE.eps}
-	\caption{Total energy $\Ham{}$ (solid lines) and \textit{refined} Hamiltonian $\modified{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit rotation method.}
+	\caption{Total energy $\Ham{}$ (solid lines) and \textit{refined} Hamiltonian $\refined{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit rotation method.}
 	\label{fig:nve}
@@
 Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
-Fig.~\ref{fig:energy_drift} also contains the drift in $\modified{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method.
+Fig.~\ref{fig:energy_drift} also contains the drift in $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method.
 It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
@@
 	\includegraphics{Figures/energy_drift.eps}
-    \caption{Influence of the time-step size ($h$) on the long-time drift ($R$) in: the extended energy ($\modified{H}$, down triangles) and its \textit{refined} Hamiltonian ($\modified{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\modified{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
+    \caption{Influence of the time-step size ($h$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its \textit{refined} Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}
@@
 that for ths case, the temperature does not reproduce the specified temperature.
-The ``refined'' temperature $\modified{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
-Of course, as $\modified{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
+The ``refined'' temperature $\refined{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
+Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
 Indeed, for all the time-step sizes considered the agreement is excellent.
@@
 
-In addition, we are able to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\modified{\Ham{}}$ by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\modified{\Ham{}}$, and that state of interest.
+In addition, we are able to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\refined{\Ham{}}$ by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\refined{\Ham{}}$, and that state of interest.
 $\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
@@
      \cmidrule[0.5mm](lr){3-4} \cmidrule[0.5mm](lr){5-6}
-     $\timestep$ & $\modified{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F/N$} \\
+     $\timestep$ & $\refined{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F/N$} \\
         $\text{(fs)}$  & $\text{(K)}$ & $\text{(K)}$ & $\text{(kcal/mol)}$  & $\text{(K)}$ & $\text{(kcal/mol)}$ & $\text{(kcal/mol)}$ \\
@@
 \label{eq:general shadow hamiltonian}
-\modified{\Ham{}} = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
+\refined{\Ham{}} = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
 \end{equation*}
 
-Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\modified{\Ham{}}$ as above.
-In the literature, $\modified{\Ham{}}$ is usually referred to as a shadow Hamiltonian.
+Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\refined{\Ham{}}$ as above.
+In the literature, $\refined{\Ham{}}$ is usually referred to as a shadow Hamiltonian.
 
@@
 \begin{equation*}
-\modified U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
+\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
 \end{equation*}
@@
 \begin{equation*}
-\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
 \end{equation*}
@@
 
-Although $\modified{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
+Although $\refined{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
 As a result, the modified kinetic energy cannot, in principle, be split into independent translational and rotational contributions.

### 2018-06-08 rigid_bodies.tex
revision I of disc. errors

--- 2018-06-08 rigid_bodies.tex
@@
 Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
-On the other hand, the symmetric scheme due to Martyna \textit{et al.} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
+On the other hand, the symmetric scheme due to \citeauthor{Martyna_1996} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
 In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
@@
 
-To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
 Particularly, one can see from Fig.~\ref{fig:num_stab} (a) that for each time-step size the extended energy drifts upwards until reaching a point at which the integrator becomes stable.
 It should be remarked that these results were double-checked by using the LAMMPS software package \cite{Plimpton_1995}. 
-As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
+As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
 Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows between the two thermostats.
-To the best of our knowledge, this finding did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat, that flux is also observed for the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+To the best of our knowledge, this finding did not result in a revision in the literature of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
+As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat, that flux is also observed for the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
 In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.
@@
 	\includegraphics{Figures/numerical_stability.eps}
-    \caption{Long-time drift in (a) the extended energy and (b) in each thermostat energy for the NVT method introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+    \caption{Long-time drift in (a) the extended energy and (b) in each thermostat energy for the NVT method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
 In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing to the thermostat coupled to the rotational degrees of freedom.}
@@
 The simulations were carried out specifying T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_v$).
-The temperature values of Fig.~\ref{fig:properties} (a) were computed from the simulated kinetic energies assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
 The lines are a guide for the eye. 
@@
 They correspond to the results obtained employing $h$ = 1 fs for the numerical scheme proposed in this paper.
-In addition, we have included the results of applying our method but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
-Despite the non-equilibrium process ocurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}, because that is not the only problematic issue related to the method in question.
+We have included the results of applying our method but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
+The temperature values of Fig.~\ref{fig:properties} (a) were computed from the simulated kinetic energies assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
+
+Despite the non-equilibrium process ocurring within the thermostats, it is worth showing the results obtained with the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, because that is not the only problematic issue.
 Particularly, from Fig.~\ref{fig:properties} (a) we note that the simulated temperature decreases systematically with $\timestep$.
-This kind of behavior is observed in NVE simulations\cite{Davidchack_2009,Silveira_2017}, reflecting that the factorization employed in Kamberaj \textit{et al.} \cite{Kamberaj_2005} is not suitable. 
-As a result of this inefficient temperature control, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al.} \cite{Martyna_1996} (see Fig.~\ref{fig:properties} (b) and (c)).
+This behavior is observed in NVE simulations\cite{Davidchack_2009,Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
+As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}.
+Figs.~\ref{fig:properties} (a)-(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
+As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
+Thus, the results for the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
+
 From Fig.~\ref{fig:properties} it becomes clear that our method yields significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
 This is true also for the results corresponding to the use of the NO-SQUISH method\cite{Miller_2002} in the NVE part of the integrator.
-Mathematically, the derivation of the refined Hamiltonian was facilitated by the fewer number of operators associated with the exact solution of rotation.
+Mathematically, the derivation of the refined Hamiltonian was facilitated by the fewer number of operators associated with the unsplit solution of rotation.
 However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement.
-
-The specified temperature corresponds to T = 298 K.
-Fig. clearly shows that a temperature control as described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
-As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
-Thus, the results of Fig.~\ref{fig:props_k_m} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the specified temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
-These observations motivated the development of the scheme of Sec.~\ref{sec:refined_method}, whose promising results we show below.
 
@@
 In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because is the rotational degrees of freedom are the fastest modes in the rigid model of water.
-We see that the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
+We see that the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
 From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption and a small deviation is only observed for 7 fs.
-On the other hand, the results depicted in Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of Martyna \textit{et al.} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c) in the sense that the simulated translational and rotational kinetic energies deviate in opposite directions from the expected values and they do so contrary to what is predicted by the equipartition theorem.
+On the other hand, the results depicted in Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c) in the sense that the simulated translational and rotational kinetic energies deviate in opposite directions from the expected values and they do so contrary to what is predicted by the equipartition theorem.
 Since we have a smaller number of translational degrees of freedom, the corresponding kinetic energy should be smaller than the rotational part.
@@
 \label{sec:conclusion}
-Nevertheless, as will be shown in the forthcoming section, those numerical schemes successfully achieve the specified temperature and an acceptable long-term conservation of the extended energy $H$.
 
-by contrast
-reformulate(The method controls the temperature of the physical variables well, even as the auxiliary thermostat variables show large and growing deviations from the initial state)
 \begin{acknowledgement}

### 2018-06-08 rigid_bodies.tex

--- 2018-06-08 rigid_bodies.tex
@@
 
-Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}.
-This is essential for the discussion of the NVT case.
-
+Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}, which is essential for discussing the NVT case.
 In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
@@
 
-To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
 Particularly, one can see from Fig.~\ref{fig:num_stab} (a) that for each time-step size the extended energy drifts upwards until reaching a point at which the integrator becomes stable.
 It should be remarked that these results were double-checked by using the LAMMPS software package \cite{Plimpton_1995}. 
-As already mentioned, the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows from one thermostat to the other.
-To the best of our knowledge, the work of Davidchack \cite{Davidchack_2009} did not result in a revision in the literature of the method of Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
-As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat for different time-step sizes, that flux is also observed for the scheme due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}.
+As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
+Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows between the two thermostats.
+To the best of our knowledge, this finding did not result in a revision in the literature of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
+As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat, that flux is also observed for the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
 In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.

### 2018-06-08 rigid_bodies.tex
Revision of backward error analysis and NVT Adjustment

@@
 Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
-As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
+As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
 The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
@@
 
-The first new result we present in this paper is an analytical form of an approximation to the shadow Hamiltonian, referred here as \textit{refine}d Hamiltonian, concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~ \citenum{Silveira_2017}.
-A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian} and here we summarize the final expressions.
-
-As demonstrated in Appendix~\ref{sec:rigid body shadow hamiltonian}, the \textit{refined} Hamiltonian for a Verlet-type method applied to a system of rigid bodies is given by $\refined{\Ham{}} = \refined K + \refined U + \mathcal{O}(h^4)$, where $\refined K$ and $\refined U$ are \textit{refined} kinetic and potential energies.
+A new result we present here is an analytically derived approximation for $\Ham{S}$, referred to here as a \textit{refined Hamiltonian} $\refined{\Ham{}}$, concerning a system of rigid bodies whose dynamics is integrated via the unsplit rotation method of Ref.~\citenum{Silveira_2017}.
+A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian}.
+As demonstrated therein, such approximation is given by $\Ham{S} = \refined K + \refined U + \mathcal{O}(h^4)$, where $\refined K$ and $\refined U$ are the refined versions of the kinetic and potential energies, respectively.
 The latter is given by
@@
 \end{equation}
-Note that $\refined U$, such as $U$, is independent of $\vt p$ and $\vt \pi$.
-In turn, the \textit{refined} kinetic energy can be expressed as
+
+Note that $\refined U$ becomes dependent on $\timestep^2$, but keeps being independent of $\vt p$ and $\vt \pi$.
+In turn, the \textit{refined kinetic energy} can be expressed as
 \begin{equation}
 \label{eq:modified kinetic energy}
-\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\refined K = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
 \end{equation}
-where $\refined{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, given by $\tilde{\mt \Omega} = {\mt \Omega} + {\timestep^2}/{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)$, where
-\begin{equation*}
-{\mt \Omega} = \left[\begin{array}{cc}
-{\mt M}^{-1} & \mt 0 \\
-\mt 0 & \frac{1}{4} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q)
-\end{array}\right]
-\end{equation*}
-and ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to the entries of $\vt r$ and $\vt q$, given by
-\begin{equation*}
-{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
-U_{\vt r \vt r} & U_{\vt r \vt q} \\
-U_{\vt q \vt r} & U_{\vt q \vt q}
-\end{array}\right].
-\end{equation*}
+where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega}$.
+In this definition, ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to $\vt r$ and $\vt q$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
+Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
+As a result, one cannot generally split the refined kinetic energy into independent translational and rotational contributions.
 
-Hence, the trajectory generated by the splitting scheme of Eq.~\ref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will approach even more closely the solution of a modified ODE system given by
+Now notice that a trajectory generated by the splitting scheme of Eq.~\eqref{eq:splitting NVE}, which is intended to approach the exact solution of Eq.~\eqref{eq:ODE system for NVE}, will even more closely approach the solution of a modified ODE system given by
 \begin{equation*}
@@
 In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
-Fortunately, we can estimate it rather easily though a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
+Fortunately, we can estimate it rather easily by following \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
 Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
@@
 
-In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each rigid body $i$, such as required for satisfying the unit-norm condition \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of\cite{Schay_1995}
+In the case of quaternions, as a simple polynomial interpolation is insufficient to ensure that $\tr{\vt q}_i \dot{\vt q}_i = 0$ at all times for each body $i$, such as required for preserving the unit-norm constraint \cite{Silveira_2017}, we estimate the time-derivative $\dot{\vt q}$ by means of \cite{Schay_1995}
 \begin{equation*}
@@
 \end{equation*}
-where ${\mt \Pi}$ is a matrix-valued function given by either product ${\mt B}\tr{\mt B}$ or ${\mt C}\tr{\mt C}$ \cite{Silveira_2017}.
-In practice, we evaluate the equation above individually for each body $i$ with ${\mt \Pi}({\vt q}_i) = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix, thus projecting the interpolation-based derivative quaternion into the hyperplane orthogonal to ${\vt q}_i$.
+where ${\mt \Pi}(\vt q)$ is a block-diagonal projection matrix, with each block given by ${\mt \Pi}_i = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix.
 
@@
 \label{sec:refined_method}
-Bringing the attention back to the NVT case, our proposal consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, produces a phase-space move that is more consistent with the modified Hamiltonian ODE system than with the original one.
-Thus, we should adjust the Nos\'e-Hoover chain so as to provide a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
-Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting scheme, as described before.
-The only required modification is that $\refined K$ replaces $K$ in Eq.~\eqref{eq:generalized force NVT}, which defines the force $G_1$.
-By contrasting such definition with Eqs.~\eqref{eq:modified kinetic energy} and \eqref{eq:shadow ODE system}, we observe that $G_1$ becomes
+
+Bringing the attention back to the NVT case, the observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
+It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a canonical ensemble for a system with this Hamiltonian.
+Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $6N-3$ degrees of freedom is that
+\begin{equation*}
+\left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T,
+\end{equation*}
+which becomes $2\langle \refined{K} \rangle = (6N - 3) k_\text{\tiny B} T$ if the refined kinetic energy is given by Eq.~\eqref{eq:modified kinetic energy}.
+
+Our proposal for dealing with discretization effects on NVT simulations goes as follows.
+It consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
+Thus, we should adjust the Nos\'e-Hoover chain with the aim to obtain a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
+Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
+The only required modification is that $\refined K$ should replace $K$ in Eq.~\eqref{eq:generalized force NVT}, which then becomes
 \begin{equation}
@@
 \end{equation}
-Also, the ``extended energy'' preserved by this approach is now given by
+
+Also, the \textit{refined extended energy} preserved by this approach is now given by
 \begin{equation}
@@
 \node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
-\node at (0.0,-0.2)[below]{$r_{t-h}$};
-\node at (2.2,-0.2)[below]{$r_{t}$};
-\node at (4.4,-0.2)[below]{$r_{t+h}$};
-\node at (6.6,-0.2)[below]{$r_{t+2h}$};
+\node at (0.0,-0.2)[below]{$r_{t-2h}$};
+\node at (2.2,-0.2)[below]{$r_{t-h}$};
+\node at (4.4,-0.2)[below]{$r_{t}$};
+\node at (6.6,-0.2)[below]{$r_{t+h}$};
 \path[->,thick] (1) edge [bend left=45] (2);
@@
 \appendix
+\numberwithin{equation}{section}
 
-\section{Backward Error Analysis}
+\section{Shadow Hamiltonian Approximation for Rigid Bodies}
 \label{sec:rigid body shadow hamiltonian}
@@
 where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
-\begin{equation*}
+\begin{equation}
+\label{eq:block-diagonal inverse mass tensor}
 {\mt \Omega} = \left[\begin{array}{cc}
@@
 \end{array}\right].
-\end{equation*}
+\end{equation}
 

### 2018-06-08 rigid_bodies.tex
Revision of NVT adjustment

@@
 \end{equation}
-where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega}$.
-In this definition, ${\mt U}_{\vt x \vt x}$ is the Hessian matrix of the potential energy with respect to $\vt r$ and $\vt q$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
+where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt \Xi} \tr{\mt \Omega}$.
+In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to $\vt r$ and $\vt q$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
 Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
@@
 \end{equation*}
-where ${\mt \Pi}(\vt q)$ is a block-diagonal projection matrix, with each block given by ${\mt \Pi}_i = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix.
+where ${\mt \Pi}(\vt q)$ is a block-diagonal projection matrix, with each diagonal block given by ${\mt \Pi}_i = {\mt 1} - {\vt q}_i \tr{\vt q}_i$, where $\mt 1$ is the identity matrix.
 
@@
 It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a canonical ensemble for a system with this Hamiltonian.
-Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $6N-3$ degrees of freedom is that
+Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
 \begin{equation*}
-\left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T,
+\left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle = L k_\text{\tiny B} T,
 \end{equation*}
-which becomes $2\langle \refined{K} \rangle = (6N - 3) k_\text{\tiny B} T$ if the refined kinetic energy is given by Eq.~\eqref{eq:modified kinetic energy}.
+which becomes $2\langle \refined{K} \rangle = L k_\text{\tiny B} T$ if the refined kinetic energy is given by Eq.~\eqref{eq:modified kinetic energy}.
 
@@
 
-One instant in which $\refined K$ must be evaluated is right after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
-Notice that such action takes place only once per time step, intercalating with non-Hamiltonian propagations.
-Therefore, there is no record beforehand of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$ with high accuracy.
-This could be solved by executing a number of virtual Verlet steps, both forward and backward in time, so as to generate a virtual Hamiltonian trajectory that coincides with the actual trajectory in the two phase-space points which are input and output of the central propagator in Eq.~\eqref{eq:splitting NVT}.
-Nonetheless, this might require additional force computations per actual time step, thus increasing the cost of the simulation.
+One instant in which $\refined K$ must be evaluated is immediately after the action of propagator $e^{\timestep \Liu{\tiny NVE}}$.
+This takes place once per time step, intercalating with non-Hamiltonian propagations.
+Therefore, at such instant there is no previous record of a Hamiltonian trajectory from which we can estimate $\dot{\vt r}$ and $\dot{\vt q}$ with high accuracy.
 Fortunately, only positions and orientations are required for the numerical differentiation.
-This allows us to obtain a four-point stencil without any additional force evaluation, just by executing one incomplete virtual step for each direction in time, as schematically illustrated in Fig.~\ref{fig:virtual}.
-By incomplete we mean that the step is aborted once the positions and orientations have been updated.
+This allows us to generate a four-point stencil by executing two virtual, incomplete Verlet steps, backward and forward in time, such as illustrated in Fig.~\ref{fig:virtual steps}.
+By virtual and incomplete we mean, respectively, that these steps are never incorporated to the trajectory and that they are aborted once the positions and orientations have been updated.
+By this means, we can obtain estimates $\dot{\vt r}^{[3]}$ and $\dot{\vt q}^{[3]}$ without any additional force evaluation, which results in a small computational overhead.
+
 \begin{figure}
 \begin{tikzpicture}
-\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{$r(t)$};
+\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{${\vt r}(t)$};
 \node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
@@
 \node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
-\node at (0.0,-0.2)[below]{$r_{t-2h}$};
-\node at (2.2,-0.2)[below]{$r_{t-h}$};
-\node at (4.4,-0.2)[below]{$r_{t}$};
-\node at (6.6,-0.2)[below]{$r_{t+h}$};
+\node at (0.0,-0.2)[below]{${\vt r}_{t-2h}$};
+\node at (2.2,-0.2)[below]{${\vt r}_{t-h}$};
+\node at (4.4,-0.2)[below]{${\vt r}_{t}$};
+\node at (6.6,-0.2)[below]{${\vt r}_{t+h}$};
 \path[->,thick] (1) edge [bend left=45] (2);
@@
 \caption{Schematic representation}
-\label{fig:virtual}
+\label{fig:virtual steps}
 \end{figure}
 
-In addition, $\refined K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales the momentum vectors $\vt p$ and $\vt \pi$ by a common factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
-Because such propagator leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ goes as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.
+In addition, $\refined K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales both $\vt p$ and $\vt \pi$ by a factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
+Because such action leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ is done as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.
 
-Please note that Eq.~\ref{eq:generalized force NVT} is closely related to Tolman's generalized equipartition theorem \cite{Uline_2008}, which states that
-\begin{equation*}
-\left\langle \tr{\vt p} \diff{\refined{\Ham{}}}{\vt p} + \tr{\vt \pi} \diff{\refined{\Ham{}}}{\vt \pi} \right\rangle = (6N - 3) k_\text{\tiny B} T.
-\end{equation*}
-By substituting $\refined{\Ham{}}$ in this equation, we obtain that
-\begin{equation}
-2\left\langle \refined{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
-\end{equation}
-This result, with the \textit{refined} kinetic energy between brackets, provides a generalization for $h > 0$ of the simulated temperature \cite{Eastwood_2010}.
-
-Finally, given that configurations $x_i$ are sampled from a distribution proportional to $\exp(-\beta \refined{\Ham{}})$, we need to reweight the computed observables $A(x_i)$ to recover the original distribution $\exp (-\beta \Ham{})$.
-This is done as follows:
+Finally, given that configurations $\vt x$ are sampled from a distribution proportional to $e^{-\beta \refined{\Ham{}}(\vt x)}$, we need to reweight any computed observable $A(\vt x)$ so that its average is consistent with the original distribution.
+Such reweighting is done by \cite{Torrie_1977}
 \begin{equation}
@@
 
-A drawback of our approach is that the resulting NVT integrator lacks time-reversal symmetry due to the employed differentiation formulas.
-As a matter of fact, long-term reversibility is not generally achievable with MD codes that rely on floating-point arithmetics.
-In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per actual time step.
+A drawback of our approach is that the resulting integrator lacks time-reversal symmetry due to the asymmetric differentiation formulas.
+As a matter of fact, long-term reversibility is not generally achievable with software that relies on floating-point arithmetics.
+In any event, if short-term reversibility is a requirement, one can simply employ a symmetric five-point stencil formula, at the price of one additional force computation round per time step.
+Higher-order derivative estimates can also be obtained in this fashion if necessary.
 

### 2018-06-08 rigid_bodies.tex
add figures and files

--- 2018-06-08 rigid_bodies.tex
@@
 \begin{equation}
-2\left\langle \refined{K} \right\rangle = (6N - 3) k_\text{\tiny B} T.
+\label{eq:equipartition}
+2\left\langle \refined{K} \right\rangle = (6N - 3) k_\text{\tiny B} \refined{T}.
 \end{equation}
@@
 However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement.
+Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $h$.
+This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
 
-We have also (check ensemble results)\ref{table:check_ens} Sampling validation: Shirts \cite{Shirts_2013}
+With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for each time-step size the free energy difference per water molecule ($\Delta F/N$, circles) between the sampled and target states.
+$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
+As expected, the free energy is only negligible for $h$ = 1 fs.
+In Fig.~\ref{fig:free_energy} we also depict the \textit{refined} temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
+Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the target temperature.
+Indeed, for all the time-step sizes considered the agreement is excellent.
 
-With the aim of highlighting the importance of reweighting to recover averages of the physical system, in Table~\ref{table:reweight} we present the calculated temperature and potential energy for each $\timestep$ before and after applying Eq.~\ref{eq:reweighting} for time-step sizes smaller 7 fs, as we saw in Fig.
-that for ths case, the temperature does not reproduce the specified temperature.
-The ``refined'' temperature $\refined{T}$ was calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
-Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the specified temperature.
-Indeed, for all the time-step sizes considered the agreement is excellent.
-We see that in the absence of reweighting, the computed temperatures are notably influenced by $h$.
-Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $h$.
-This could, in principle, be remediated by using a higher-order ``refined'' Hamiltonian.
-
-In addition, we are able to ``measure'' how far our sample is from the state of interest with physical Hamiltonian $\refined{\Ham{}}$ by calculating the free energy differences ($\Delta F$) between the sampled state, whose Hamiltonian is given by $\refined{\Ham{}}$, and that state of interest.
-$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
-As expected, from Table~\ref{table:reweight} one verifies that $\Delta F$ is only negligible for $h$ = 1 fs and systematically increases with $h$.
-
-\begin{table*}
-	\caption{Effect of the time-step size on ....
-of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing the unsplit solution for free rotations, given by the numerical schemes.
-uncertainties in free energy never exced 1e-4}
-    \label{table:reweight}
-    \sisetup{
-table-number-alignment = center
-}
-    \begin{tabular}{SS[separate-uncertainty,table-figures-uncertainty = 2]S[separate-uncertainty,table-figures-uncertainty=2]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S[separate-uncertainty,table-figures-uncertainty= 1]S[separate-uncertainty,table-figures-uncertainty= 1,table-figures-decimal=4]S}
-     & &  \multicolumn{2}{c}{no reweighting} & \multicolumn{2}{c}{after reweighting} & \\
-     \cmidrule[0.5mm](lr){3-4} \cmidrule[0.5mm](lr){5-6}
-     $\timestep$ & $\refined{T}$  & $T$  & {$\langle U/N \rangle$}  & $T$ & {$\langle U/N \rangle$}   & {$\Delta F/N$} \\
-        $\text{(fs)}$  & $\text{(K)}$ & $\text{(K)}$ & $\text{(kcal/mol)}$  & $\text{(K)}$ & $\text{(kcal/mol)}$ & $\text{(kcal/mol)}$ \\
-	1 & 298.00(1) & 297.71(1) & -9.1015(3) & 298.00(1) & -9.1028(3) & -0.0006 \\
-	2 & 297.99(1) & 296.82(1) & -9.0970(3)  & 297.98(1)  & -9.1024(3) & -0.004 \\
-	3 & 297.98(1) & 295.38(1) & -9.0885(3) & 297.92(1) & -9.1005(3) & -0.008 \\
-	4 & 298.01(1) & 293.49(1)& -9.0764(3) & 297.85(3) & -9.0969(3) & -0.013 \\
-	5 & 297.99(1)& 291.13(1)& -9.0577(3) & 297.6(2) & -9.091(1) & -0.033 \\
-    6 & 298.00(1) & 288.46(1) & -9.0314(3) & 296.8(3) & -9.075(3) &  -0.026 \\
-	7 & 297.99(1) & 285.57(1) & -8.9944(3)  & 295.0(5)  & -9.040(4) & -0.056 \\
-	\end{tabular}
-\end{table*}
+\begin{figure}
+	\includegraphics{Figures/Free_Energy.eps}
+	\caption{Free Energy}
+	\label{fig:free_energy}
+\end{figure}
 
 We end this section with a discussion of the effect of discretization errors on the kinetic energy partition.
-In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition, which are calculated at each time-step size from the total kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times (3N - 3)$ and $\langle K_r \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times 3N$.
+In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition, which are calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times (3N - 3)$ and $\langle K_r \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times 3N$.
 
-In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because is the rotational degrees of freedom are the fastest modes in the rigid model of water.
+In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water.
 We see that the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
+The results of Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
+Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
 From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption and a small deviation is only observed for 7 fs.
-On the other hand, the results depicted in Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c) in the sense that the simulated translational and rotational kinetic energies deviate in opposite directions from the expected values and they do so contrary to what is predicted by the equipartition theorem.
-Since we have a smaller number of translational degrees of freedom, the corresponding kinetic energy should be smaller than the rotational part.
 
 \begin{figure*}
-	\includegraphics{Figures/thermodynamic_properties.eps}
+	\includegraphics{Figures/props.eps}
 	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
@@
 \end{figure*}
-
-%Given that the potential and kinetic energies are separable, the analysis can be made separately as follows.This last approach does not involve any binning of the energies, hence no discretization error is present.
-%m-b distribution valid when equipartition of energy is satisfied.
-
-%It can be seen that the mentioned algorithm generates a distribution consistent with the canonical ensemble, and that is also true for the kinetic energies.
-%Finally, in Table~\ref{table:ensemblevalidation} we present the results of the maximum likelihood approach applied for both NVT integration schemes.
-%It can be seen that the deviations from the true slope, $\timestep$ = 8.970~K, are less than $1\sigma$.
-%This means that effectively, the sampled values do not deviate from the correct distribution to a statistically noticeable level.
 

### 2018-06-08 Figures/Free_Energy.py

@@ -1,0 +1,40 @@
+import numpy as np
+import utility_belt as ub
+import matplotlib.pyplot as plt
+from matplotlib.font_manager import FontProperties
+import mics
+import pandas as pd
+plt.rcParams.update({'font.size':8})
+
+fig = plt.figure('axes',figsize=(2.5,2))
+ax = fig.add_subplot(111)
+ax.set_xlabel('$h$ (fs)')
+ax.set_ylabel('$\Delta F/N$ (kcal/mol)')
+timesteps = [1,2,3,4,5,6,7]
+color = ['blue', 'green', 'red', 'black' ]
+ax.set_xlim(0.5,7.5)
+xticks = [1,2,3,4,5,6,7]
+ax.set_xticks(xticks)
+
+Ts = [298.00,297.99,297.98,298.01,297.99,298.00,297.99]
+Ts_error = [0.01,0.01,0.01,0.01,0.01,0.01,0.01]
+lw = 0.5
+
+
+NB = 903
+kB = 0.0019872041 #Boltzmann's constant (kcal/mol/K).
+beta = 1./(kB*298.0)
+data = np.genfromtxt('S1.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','temperatura',\
+                     'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
+                     'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
+ax.errorbar(timesteps, data['f'][0:7]/(NB*beta),yerr = data['df'][0:7]/(NB*beta),linestyle ='-', color = 'black', marker = 'o',markersize = 4, linewidth = lw)
+ax2 = ax.twinx()
+ax2.set_ylim(297.0,299.0)
+yticks = [297.0,298.0,299.0]
+ax2.set_yticks(yticks)
+ax2.set_ylabel('$T$ (K)')
+ax2.errorbar(timesteps, Ts, yerr=Ts_error,linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw)
+
+
+#ax.legend(loc='upper center', bbox_to_anchor=(0.5,0.7), ncol=2, fancybox=True, frameon=False, numpoints = 1)
+fig.savefig('Free_Energy.eps', format='eps', dpi=600, bbox_inches='tight')

### 2018-06-08 Figures/thermodynamic_properties.py

@@ -1,0 +1,52 @@
+import numpy as np
+import utility_belt as ub
+import pandas as pd
+import matplotlib.pyplot as plt
+from matplotlib import rc
+plt.rcParams.update({'font.size':8})
+rc('text', usetex=True)
+
+fig,((ax1,ax2),(ax3,ax4)) = plt.subplots(2,2,sharex=True,sharey=False)  
+ax = (ax1,ax2,ax3,ax4)
+fig.subplots_adjust(hspace=0.1,wspace=0.22)
+fig.set_size_inches(w=7.5,h=4)
+
+ax[2].set_xlabel('$h$ (fs)')
+ax[3].set_xlabel('$h$ (fs)')
+
+ax[0].set_ylabel(r'T (K)')
+ax[1].set_ylabel(r'$<U/N>$ (kcal/mol)')
+ax[2].set_ylabel(r'$<W>/N$ (kcal/mol)')
+ax[3].set_ylabel(r'$C_v$ (kcal/mol K)')
+
+
+ax[1].set_xlim(0.5,7.5)
+ax[2].set_xlim(0.5,7.5)
+xticks = [1,2,3,4,5,6,7]
+ax[1].set_xticks(xticks)
+ax[2].set_xticks(xticks)
+timesteps = [1,2,3,4,5,6,7]
+lw = 0.5
+NB = 903
+kB = 0.0019872041 #Boltzmann's constant (kcal/mol/K).
+beta = 1./(kB*298.0)
+integrators = ['S1','S1m','P1','K2']
+colors = ['blue','magenta','green','red']
+labels = ['This work (Unsplit)','This work (NO-SQUISH)', r'Martyna \textit{et al.}', r'Kamberaj \textit{et al.}']
+k = 0
+for i in integrators:
+    data = np.genfromtxt(i + '.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','temperatura',\
+                     'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
+                     'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
+    ax[0].errorbar(timesteps, data['temperatura'][0:7],yerr = data['dtemperatura'][0:7],linestyle ='-', color = colors[k], marker = 's',markersize = 4, linewidth = lw, label = labels[k])
[…]

### 2018-06-08 rigid_bodies.tex
Changes in NVE results

--- 2018-06-08 rigid_bodies.tex
@@
 
-\subsection{Adjustment of the NVT Dynamics}
+\subsection{Refinement of the NVT Dynamics}
 \label{sec:refined_method}
@@
 
-\section{Application to Water and Discussion}
+\section{Application to the Simulation of Liquid Water}
 \label{sec:numerical_results}
 
-In the following, we analyze the performance of the NVT scheme proposed in  this paper and compare the results with those obtained employing the integrator due to \citeauthor{Martyna_1996} \cite{Martyna_1996} as well as the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}
+In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} with our described adaptations.
 
-The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
-The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA} as done in Ref.~\citenum{Silveira_2017}.
-The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.
-In addition, all the simulations were carried out considering $t_d = 100$ fs.
+The system under study has periodic boundary conditions and contains 903 TIP3P \cite{Jorgensen_1983} water molecules at a density of 970~kg/m\textsuperscript{3}.
+The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at 10 {\AA} and smoothed by a switching function over the range from 9.5 \AA to 10 \AA, such as done in Ref.~\citenum{Silveira_2017}.
+The damping of Coulombic interactions was done by a factor $\text{erfc}(-\alpha r)$, with $\alpha = 0.29~\text{\AA}^{-1}$.
+In addition, all thermostated simulations were carried out considering a time-scale constant $t_d = 100~\text{fs}$.
 
-Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}, which is essential for discussing the NVT case.
-In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
-For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation.
-As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
-It should be remarked that $\refined{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
-As a result of such approximation, we have larger fluctuations in $\refined{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).
+We start by examining the departure of the \textit{refined} Hamiltonian from the specified one in NVE simulations.
+This is the most direct way of determining the importance of the discretization errors \cite{Engle_2005}, which is essential for discussing the NVT case.
+In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system using different time-step sizes ($\timestep$).
+For this, we used the numerical scheme of Eq.~\eqref{eq:splitting NVE} with the unsplit solution of Ref.~\citenum{Silveira_2017} applied for the rotations.
+The four-point differentiation formulas presented in Sec.~\ref{sec:refined_method} were employed for computing the refined kinetic energy.
+As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep = 1~\text{fs}$, but rapidly departs from $\Ham{}$ when $\timestep$ increases.
+It is worth remarking that $\refined{\Ham{}}$ corresponds to a simple and, most importantly, computationally inexpensive approximation of the truly conserved shadow Hamiltonian.
+As a result, we observe increasing fluctuations in $\refined{\Ham{}}$ for $\timestep = 3~\text{fs}$ and $\timestep = 4~\text{fs}$.
+This is an indication that
+1) the employed four-point formulas do not provide derivatives with the required accuracy or
+2) the high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would have a non-negligible impact on the shadow Hamiltonian.
 

### 2018-06-09 rigid_bodies.tex
Several changes in multiple parts

@@
 
-\title{Improvement of Thermostated Molecular Dynamics via Backward Error Analysis}
+\title{Refinement of Thermostated Molecular Dynamics via Backward Error Analysis}
 
@@
 A detailed derivation is provided in Appendix~\ref{sec:rigid body shadow hamiltonian}.
-As demonstrated therein, such approximation is given by $\Ham{S} = \refined K + \refined U + \mathcal{O}(h^4)$, where $\refined K$ and $\refined U$ are the refined versions of the kinetic and potential energies, respectively.
+As demonstrated therein, such approximation is given by $\Ham{S} = \refined K + \refined U + \mathcal{O}(\timestep^4)$, where $\refined K$ and $\refined U$ are the refined versions of the kinetic and potential energies, respectively.
 The latter is given by
@@
 where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt \Xi} \tr{\mt \Omega}$.
-In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to $\vt r$ and $\vt q$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
+In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to the stacked vector $[\substack{\vt p \\ \vt \pi}]$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
 Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
@@
 \label{eq:modified kinetic energy estimator}
-\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(h^n).
+\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(\timestep^n).
 \end{equation}
@@
 
+The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
+It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
+Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
+\begin{equation*}
+\label{eq:equipartition}
+L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
+\end{equation*}
+where subscript $\timestep$ points out the dependency of this average on the employed time step size.
+The expression above extends the main result of \citeauthor{Eastwood_2010} \cite{Eastwood_2010} to systems containing rigid bodies.
+It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
+We emphasize that such temperature concerns the system that has actually been simulated, rather than the one initially specified.
+
 \subsection{Refinement of the NVT Dynamics}
@@
 
-Bringing the attention back to the NVT case, the observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
-It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a canonical ensemble for a system with this Hamiltonian.
-Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
-\begin{equation*}
-\left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle = L k_\text{\tiny B} T,
-\end{equation*}
-which becomes $2\langle \refined{K} \rangle = L k_\text{\tiny B} T$ if the refined kinetic energy is given by Eq.~\eqref{eq:modified kinetic energy}.
-
-Our proposal for dealing with discretization effects on NVT simulations goes as follows.
+Bringing the attention back to the NVT case, our proposal for dealing with discretization effects goes as follows.
 It consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
-Thus, we should adjust the Nos\'e-Hoover chain with the aim to obtain a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
+Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
 Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
@@
 \label{eq:this extended energy}
-\widetilde{H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
+{\refined H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
 \end{equation}
@@
 \node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
-\node at (0.0,-0.2)[below]{${\vt r}_{t-2h}$};
-\node at (2.2,-0.2)[below]{${\vt r}_{t-h}$};
+\node at (0.0,-0.2)[below]{${\vt r}_{t-2\timestep}$};
+\node at (2.2,-0.2)[below]{${\vt r}_{t-\timestep}$};
 \node at (4.4,-0.2)[below]{${\vt r}_{t}$};
-\node at (6.6,-0.2)[below]{${\vt r}_{t+h}$};
+\node at (6.6,-0.2)[below]{${\vt r}_{t+\timestep}$};
 \path[->,thick] (1) edge [bend left=45] (2);
@@
 
-Finally, given that configurations $\vt x$ are sampled from a distribution proportional to $e^{-\beta \refined{\Ham{}}(\vt x)}$, we need to reweight any computed observable $A(\vt x)$ so that its average is consistent with the original distribution.
-Such reweighting is done by \cite{Torrie_1977}
+Finally, we recall that configurations are sampled from a distribution proportional to $e^{-\beta \refined{\Ham{}}}$.
+Notwithstanding, it is possible to compute ensemble averages consistent with a distribution proportional to $e^{-\beta \Ham{}}$, which would have been obtained if $\timestep \to 0$.
+For this, we simply need to reweight any computed observable $A$ in accordance with \cite{Torrie_1977}
 \begin{equation}
 \label{eq:reweighting}
-\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
+%\langle A \rangle_0 = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] },
+\langle A \rangle_0 = \frac{\left\langle A \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}{\left\langle \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}.
 \end{equation}
@@
 The system under study has periodic boundary conditions and contains 903 TIP3P \cite{Jorgensen_1983} water molecules at a density of 970~kg/m\textsuperscript{3}.
-The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at 10 {\AA} and smoothed by a switching function over the range from 9.5 \AA to 10 \AA, such as done in Ref.~\citenum{Silveira_2017}.
+The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at 10 \AA and smoothed by a switching function over the range from 9.5 \AA to 10 \AA, such as done in Ref.~\citenum{Silveira_2017}.
 The damping of Coulombic interactions was done by a factor $\text{erfc}(-\alpha r)$, with $\alpha = 0.29~\text{\AA}^{-1}$.
@@
 1) the employed four-point formulas do not provide derivatives with the required accuracy or
-2) the high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would have a non-negligible impact on the shadow Hamiltonian.
+2) the high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would have a non-negligible contribution to the shadow Hamiltonian.
 
@@
 	\includegraphics{Figures/NVE.eps}
-	\caption{Total energy $\Ham{}$ (solid lines) and \textit{refined} Hamiltonian $\refined{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit rotation method.}
+	\caption{Specified Hamiltonian $\Ham{}$ (solid lines) and refined Hamiltonian $\refined{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\eqref{eq:splitting NVE} with the unsplit rotation method \cite{Silveira_2017}.}
 	\label{fig:nve}
@@
 On the other hand, the symmetric scheme due to \citeauthor{Martyna_1996} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
-In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $h$ = 5 fs. 
+In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $\timestep$ = 5 fs.
 Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
 Fig.~\ref{fig:energy_drift} also contains the drift in $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method.
-It is worth mentioning that the stability threshold for the NVE method corresponds to $h$ = 6 fs.
+It is worth mentioning that the stability threshold for the NVE method corresponds to $\timestep$ = 6 fs.
 The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
@@
 	\includegraphics{Figures/energy_drift.eps}
-    \caption{Influence of the time-step size ($h$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its \textit{refined} Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
+    \caption{Influence of the time-step size ($\timestep$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its \textit{refined} Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}
@@
 In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.
-Interestingly, Fig.~\ref{fig:num_stab} (b) suggests that a unique steady state is established, and the time to reach the steady state decreases with $h$.
+Interestingly, Fig.~\ref{fig:num_stab} (b) suggests that a unique steady state is established, and the time to reach the steady state decreases with $\timestep$.
 As far as we know, this behavior has not been reported before.
@@
 
-In Fig.~\ref{fig:properties} we show the computed average of some thermodynamic properties as a function of $h$.
+In Fig.~\ref{fig:properties} we show the computed average of some thermodynamic properties as a function of $\timestep$.
 The simulations were carried out specifying T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_v$).
@@
 Also, the horizontal dotted lines (black) help to visualize the departure of the computed averages with increasing $\timestep$.
-They correspond to the results obtained employing $h$ = 1 fs for the numerical scheme proposed in this paper.
+They correspond to the results obtained employing $\timestep$ = 1 fs for the numerical scheme proposed in this paper.
 We have included the results of applying our method but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
@@
 However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement.
-Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $h$.
+Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $\timestep$.
 This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
@@
 $\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
-As expected, the free energy is only negligible for $h$ = 1 fs.
+As expected, the free energy is only negligible for $\timestep$ = 1 fs.
 In Fig.~\ref{fig:free_energy} we also depict the \textit{refined} temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
@@
 \begin{figure*}
-	\includegraphics{Figures/props.eps}
+	\includegraphics{Figures/thermodynamic_properties.eps}
 	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
@@
 \label{eq:general shadow hamiltonian}
-\refined{\Ham{}} = \Ham A + \Ham B + \frac{h^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(h^4).
+\refined{\Ham{}} = \Ham A + \Ham B + \frac{\timestep^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(\timestep^4).
 \end{equation*}

### 2018-06-09 rigid_bodies.tex
files added

--- 2018-06-08 rigid_bodies.tex
@@
 
-\subsection{Refinement of the NVT Dynamics}
+\subsection{Adjustment of the NVT Dynamics}
 \label{sec:refined_method}
@@
 
-\section{Application to the Simulation of Liquid Water}
+\section{Application to Water and Discussion}
 \label{sec:numerical_results}
 
-In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} with our described adaptations.
+In the following, we analyze the performance of the NVT scheme proposed in  this paper and compare the results with those obtained employing the integrator due to \citeauthor{Martyna_1996} \cite{Martyna_1996} as well as the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}
 
-The system under study has periodic boundary conditions and contains 903 TIP3P \cite{Jorgensen_1983} water molecules at a density of 970~kg/m\textsuperscript{3}.
-The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at 10 {\AA} and smoothed by a switching function over the range from 9.5 \AA to 10 \AA, such as done in Ref.~\citenum{Silveira_2017}.
-The damping of Coulombic interactions was done by a factor $\text{erfc}(-\alpha r)$, with $\alpha = 0.29~\text{\AA}^{-1}$.
-In addition, all thermostated simulations were carried out considering a time-scale constant $t_d = 100~\text{fs}$.
+The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
+The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA} as done in Ref.~\citenum{Silveira_2017}.
+The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.
+In addition, all the simulations were carried out considering $t_d = 100$ fs.
 
-We start by examining the departure of the \textit{refined} Hamiltonian from the specified one in NVE simulations.
-This is the most direct way of determining the importance of the discretization errors \cite{Engle_2005}, which is essential for discussing the NVT case.
-In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system using different time-step sizes ($\timestep$).
-For this, we used the numerical scheme of Eq.~\eqref{eq:splitting NVE} with the unsplit solution of Ref.~\citenum{Silveira_2017} applied for the rotations.
-The four-point differentiation formulas presented in Sec.~\ref{sec:refined_method} were employed for computing the refined kinetic energy.
-As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep = 1~\text{fs}$, but rapidly departs from $\Ham{}$ when $\timestep$ increases.
-It is worth remarking that $\refined{\Ham{}}$ corresponds to a simple and, most importantly, computationally inexpensive approximation of the truly conserved shadow Hamiltonian.
-As a result, we observe increasing fluctuations in $\refined{\Ham{}}$ for $\timestep = 3~\text{fs}$ and $\timestep = 4~\text{fs}$.
-This is an indication that
-1) the employed four-point formulas do not provide derivatives with the required accuracy or
-2) the high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would have a non-negligible impact on the shadow Hamiltonian.
+Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}, which is essential for discussing the NVT case.
+In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
+For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation.
+As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
+It should be remarked that $\refined{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
+As a result of such approximation, we have larger fluctuations in $\refined{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).
 
@@
 With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for each time-step size the free energy difference per water molecule ($\Delta F/N$, circles) between the sampled and target states.
-$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
+$\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\ref{eq:reweighting}.
 As expected, the free energy is only negligible for $h$ = 1 fs.
@@
 Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
-From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption and a small deviation is only observed for 7 fs.
+From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption for time-step sizes up to 5 fs.
 
 \begin{figure*}
-	\includegraphics{Figures/props.eps}
+	\includegraphics{Figures/thermodynamic_properties.eps}
 	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}

### 2018-06-09 Figures/Free_Energy.py

@@ -6,7 +6,7 @@ import mics
 import pandas as pd
 plt.rcParams.update({'font.size':8})
 
-fig = plt.figure('axes',figsize=(2.5,2))
+fig = plt.figure('axes',figsize=(3,2))
 ax = fig.add_subplot(111)
 ax.set_xlabel('$h$ (fs)')
 ax.set_ylabel('$\Delta F/N$ (kcal/mol)')
@@ -28,12 +28,12 @@ data = np.genfromtxt('S1.csv', delimiter=',', skip_header=1, names=[' ','T','f',
                      'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
                      'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
 ax.errorbar(timesteps, data['f'][0:7]/(NB*beta),yerr = data['df'][0:7]/(NB*beta),linestyle ='-', color = 'black', marker = 'o',markersize = 4, linewidth = lw)
-ax2 = ax.twinx()
-ax2.set_ylim(297.0,299.0)
-yticks = [297.0,298.0,299.0]
-ax2.set_yticks(yticks)
-ax2.set_ylabel('$T$ (K)')
-ax2.errorbar(timesteps, Ts, yerr=Ts_error,linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw)
+#ax2 = ax.twinx()
+#ax2.set_ylim(297.0,299.0)
+#yticks = [297.0,298.0,299.0]
+#ax2.set_yticks(yticks)
+#ax2.set_ylabel('$T$ (K)')
+#ax2.errorbar(timesteps, Ts, yerr=Ts_error,linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw)
 
 
 #ax.legend(loc='upper center', bbox_to_anchor=(0.5,0.7), ncol=2, fancybox=True, frameon=False, numpoints = 1)

### 2018-06-09 Figures/thermodynamic_properties.py

@@ -15,9 +15,9 @@ ax[2].set_xlabel('$h$ (fs)')
 ax[3].set_xlabel('$h$ (fs)')
 
 ax[0].set_ylabel(r'T (K)')
-ax[1].set_ylabel(r'$<U/N>$ (kcal/mol)')
-ax[2].set_ylabel(r'$<W>/N$ (kcal/mol)')
-ax[3].set_ylabel(r'$C_v$ (kcal/mol K)')
+ax[1].set_ylabel(r'{$\langle U/N \rangle$} (kcal/mol)')
+ax[2].set_ylabel(r'{$\langle W/N \rangle$} (kcal/mol)')
+ax[3].set_ylabel(r'$C_V$ (kcal/mol K)')
 
 
 ax[1].set_xlim(0.5,7.5)
@@ -30,23 +30,33 @@ lw = 0.5
 NB = 903
 kB = 0.0019872041 #Boltzmann's constant (kcal/mol/K).
 beta = 1./(kB*298.0)
-integrators = ['S1','S1m','P1','K2']
-colors = ['blue','magenta','green','red']
-labels = ['This work (Unsplit)','This work (NO-SQUISH)', r'Martyna \textit{et al.}', r'Kamberaj \textit{et al.}']
+integrators = ['S1','S1mr','P1','K2']
+colors = ['black','magenta','green','blue']
+markers = ['s','^','o','v']
+labels = ['Refined/Unsplit','Refined/NO-SQUISH', r'Martyna', r'Kamberaj']
 k = 0
 for i in integrators:
     data = np.genfromtxt(i + '.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','temperatura',\
                      'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
                      'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
-    ax[0].errorbar(timesteps, data['temperatura'][0:7],yerr = data['dtemperatura'][0:7],linestyle ='-', color = colors[k], marker = 's',markersize = 4, linewidth = lw, label = labels[k])
-    ax[1].errorbar(timesteps, data['P'][0:7],yerr = data['dP'][0:7],linestyle ='-', color = colors[k], marker = 's',markersize = 4, linewidth = lw, label = labels[k])
[…]

### 2018-06-09 rigid_bodies.tex

--- 2018-06-09 rigid_bodies.tex
@@
 With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for each time-step size the free energy difference per water molecule ($\Delta F/N$, circles) between the sampled and target states.
-$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
-As expected, the free energy is only negligible for $\timestep$ = 1 fs.
+$\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\ref{eq:reweighting}.
+As expected, the free energy is only negligible for $h$ = 1 fs.
+
 In Fig.~\ref{fig:free_energy} we also depict the \textit{refined} temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
@@
 Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
-From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption and a small deviation is only observed for 7 fs.
+From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption for time-step sizes up to 5 fs.
 

### 2018-06-09 Figures/KE_partition.py
modified partition figure

@@ -6,24 +6,24 @@ from matplotlib import rc
 plt.rcParams.update({'font.size':8})
 rc('text', usetex=True)
 
-fig,(ax1,ax2,ax3) = plt.subplots(1,3,sharex=True,sharey=True)  
-ax = (ax1,ax2,ax3)
+fig,((ax1,ax2),(ax3,ax4)) = plt.subplots(2,2,sharex=True,sharey=True)  
+ax = (ax1,ax2,ax3,ax4)
 fig.subplots_adjust(hspace=0.1,wspace=0.1)
-fig.set_size_inches(w=7.5,h=2)
+fig.set_size_inches(w=7.5,h=4)
 
-ax[0].set_xlabel('$h$ (fs)')
-ax[1].set_xlabel('$h$ (fs)')
+
 ax[2].set_xlabel('$h$ (fs)')
+ax[3].set_xlabel('$h$ (fs)')
 
 ax[0].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
+ax[2].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
 ax[0].set_xlim(0.5,7.5)
-ax[1].set_xlim(0.5,7.5)
-ax[2].set_xlim(0.5,7.5)
+ax[0].set_ylim(740,825)
 
 xticks = [1,2,3,4,5,6,7]
-ax[0].set_xticks(xticks)
-ax[1].set_xticks(xticks)
+
 ax[2].set_xticks(xticks)
+ax[3].set_xticks(xticks)
 timesteps = [1,2,3,4,5,6,7]
 lw = 0.5
 kB = 0.0019872041 #Boltzmann's constant (kcal/mol/K).
@@ -32,18 +32,34 @@ beta = 1./(kB*298.0)
 integrators = ['S1','P1','K2']
 colors = ['black','green','blue']
 markers = ['s','^','o','v']
-labels = ['Refined/Unsplit', 'Martyna', 'Kamberaj']
-k = 0
+
+data = np.genfromtxt('S1_KEpartition.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','KEt', \
+                     'dKEt','KEr','dKEr','Kt_eq','dKt_eq','Kr_eq','dKr_eq'])
[…]

### 2018-06-09 rigid_bodies.tex
Revision of the Appendix

@@
 \end{equation}
-where $[X,Y]$ is the commutator of operators $X$ and $Y$, defined as $[X,Y] = XY - YX$.
-The remaining terms of the infinite series involve ever-deepening, nested commutators \cite{Hairer_2006}.
+where $[X,Y] = XY - YX$ is the commutator of operators $X$ and $Y$.
+The remaining terms of the infinite series involve ever-deepening nested commutators \cite{Hairer_2006}.
 
 Let us now represent the operators $\Liu A$ and $\Liu B$ using Poisson brackets involving Hamiltonians $\Ham A$ and $\Ham B$, that is, $\Liu A = \{\circ,\Ham A\}$ and $\Liu B = \{\circ,\Ham B\}$.
-By employing the properties of anti-commutativity and Jacobi identity \cite{Hairer_2006}, we can deduce that
+By employing the property of anti-commutativity and the Jacobi identity \cite{Hairer_2006}, we can deduce that
 \begin{align*}
@@
 \label{eq:general shadow hamiltonian}
-\refined{\Ham{}} = \Ham A + \Ham B + \frac{\timestep^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(\timestep^4).
+\Ham{S} = \Ham A + \Ham B + \frac{\timestep^2}{24} (2 \Liu A + \Liu B){\Liu A}{\Ham B} + \mathcal{O}(\timestep^4).
 \end{equation*}
 
-Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (putting round-off issues aside) the dynamics of another system with Hamiltonian $\refined{\Ham{}}$ as above.
-In the literature, $\refined{\Ham{}}$ is usually referred to as a shadow Hamiltonian.
+Therefore, a splitting method meant to approximately reproduce the dynamics of a system with Hamiltonian $\Ham{} = \Ham A + \Ham B$ will, in fact, reproduce exactly (round-off issues aside) the dynamics of another system with Hamiltonian $\Ham{S}$ as above.
+In the literature, $\Ham{S}$ is usually referred to as a shadow Hamiltonian.
 
-In the development that follows, we consider a single rigid body and, in the end, we generalize the result for a system of rigid bodies.
+We remark that a single rigid body is considered in the development that follows, but the final result can be readily generalized for a system with multiple bodies.
 
-For the splitting introduced in Section~\ref{sec:nve}, $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$ with the corresponding Liouviller operators $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$, with $K_{x}$ and $U_{x}$ being the gradient of the kinetic and potential energies, respectively.
-We first obtain $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$, and we find that
+For the splitting introduced in Sec.~\ref{sec:nve}, $\Ham A = K(\vt p, \vt q, \vt \pi)$ and $\Ham B = U(\vt r, \vt q)$ with the corresponding Liouville operators $\Liu{A} = \tr{K_{\vt p}}\diff{}{\vt r} + \tr{K_{\vt \pi}}\diff{}{\vt q} - \tr{K_{\vt q}}\diff{}{\vt \pi}$ and $\Liu{B} = -\tr{U_{\vt r}}\diff{}{\vt p} - \tr{U_{\vt q}}\diff{}{\vt \pi}$.
+Note that any gradient like $\diff{f}{\vt x}$ is now represented as $f_{\vt x}$ for the sake of simplicity.
+We start by obtaining $\Liu A \Ham B = \tr{K_{\vt p}} U_{\vt r} + \tr{K_{\vt \pi}} U_{\vt q}$. Next, we deduce that
 \begin{align*}
@@
 \end{align*}
-where the fact that $\tr{U_{\vt q \vt r}} = U_{\vt r \vt q}$ has been exploited.
+where the fact that $\tr{U_{\vt q \vt r}} = U_{\vt r \vt q}$ has been used.
 As reported in Ref.~\citenum{Silveira_2017}, the identity $\tr{\mt B}(\vt q) {\vt \pi} = -\tr{\mt B}(\vt \pi) {\vt q}$ is useful for performing differentiations.
@@
 
-Pre-multiplication by $\tr{K_{\vt q}}$ will introduce a product $\tr{\mt B}(\vt \pi) {\mt B}(\vt q)$, which has been shown in Ref.~\citenum{Silveira_2017} to be equal to ${\mt S}\left( \tr{\mt B}(\vt \pi) {\vt q} \right) = -2 {\mt S}({\mt I} {\vt \omega})$, where the operator ${\mt S}(\cdot)$ builds a skew-symmetric matrix from the entries of a vector.
-In addition, post-multiplication by $U_{\vt q}$ will introduce a product $\tr{\mt B}(\vt q) U_{\vt q}$, which is equal to $-2 {\vt \tau}^b$, where ${\vt \tau}^b$ is the body-fixed frame representation of the torque.
+Pre-multiplication by $\tr{K_{\vt q}}$ will introduce a product $\tr{\mt B}(\vt \pi) {\mt B}(\vt q)$.
+This has been shown in Ref.~\citenum{Silveira_2017} to be equal to ${\mt S}\left( \tr{\mt B}(\vt \pi) {\vt q} \right) = -2 {\mt S}({\mt I} {\vt \omega})$, where the operator ${\mt S}(\cdot)$ builds a skew-symmetric matrix from the entries of a vector.
+In addition, post-multiplication by $U_{\vt q}$ will introduce a product $\tr{\mt B}(\vt q) U_{\vt q}$, which is equal to $-2 {\vt \tau}_b$ \cite{Silveira_2017}, where ${\vt \tau}_b$ is the body-fixed frame representation of the torque.
 Then, the fact that ${\mt S}(\vt x) {\vt y} = -{\mt S}(\vt y) {\vt x}$ leads to
 \begin{equation*}
-\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} = 0,
+\tr{K_{\vt q}} K_{\vt \pi \vt \pi} U_{\vt q} = \frac{1}{2} \tr{\vt \omega} {\mt S}({\mt I}^{-1} {\vt \tau}_b) {\mt I} {\vt \omega}.
 \end{equation*}
-which is a null term as it corresponds to the quadratic form of a skew-symmetric matrix.
+
+The term above is identically null, as it corresponds to the quadratic form built with a skew-symmetric matrix.
 Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{k=1}^3 \omega_k \hat{\mt B}_k ) \vt q$, where each $\hat{\mt B}_k$ is a skew-symmetric permutation matrix (i.e.
@@
 \begin{align*}
-K_{\vt \pi \vt q} &= \frac{1}{2} (\diff{\tr{\vt \omega}}{\vt q}) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{k=1}^3 \omega_k \tr{\hat{\mt B}}_k = \\
+K_{\vt \pi \vt q} &= \frac{1}{2} \left(\diff{\tr{\vt \omega}}{\vt q}\right) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{k=1}^3 \omega_k \tr{\hat{\mt B}}_k = \\
 &= -\frac{1}{4} {\mt B}(\vt \pi) {\mt I}^{-1} \tr{\mt B}(\vt q) - \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k.
@@
 
-Pre-multiplication by $\tr{K_{\vt \pi}}$ and post-multiplying $U_{\vt q}$, followed by some algebraic transformations, ultimately lead to
+Pre-multiplication by $\tr{K_{\vt \pi}}$ and post-multiplication by $U_{\vt q}$, followed by some algebraic transformations, ultimately lead to
 \begin{equation*}
-\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q} = -\frac{1}{2} \tr{\vt \omega} \tr{\mt S}({\mt I}^{-1} {\vt \tau}^b) {\mt I} {\vt \omega} + \frac{1}{2} \tr{\vt \omega} {\mt S}({\vt \tau}^b){\vt \omega}.
+\tr{K_{\vt \pi}} K_{\vt \pi \vt q} U_{\vt q} = -\frac{1}{2} \tr{\vt \omega} \tr{\mt S}({\mt I}^{-1} {\vt \tau}_b) {\mt I} {\vt \omega} + \frac{1}{2} \tr{\vt \omega} {\mt S}({\vt \tau}_b){\vt \omega}.
 \end{equation*}
-Both terms in the right-hand side of the equation above are identically null. 
-Now, we are able to evaluate $\Liu A \Liu A \Ham B $, which is
+
+For the same reason explained above, both terms in the right-hand side of the equation above are identically null.
+We are now able to evaluate $\Liu A \Liu A \Ham B $, which is
 \begin{align*}
@@
 \end{align*}
-The remaining term to obtain $H_s$ is simply given by 
+
+Finally, the term that remains for obtaining $\Ham{S}$ is given by
 \begin{align*}
@@
 
-As will become clear below, it is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form
-\begin{equation*}
-K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} {\mt \Omega}(\vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
-\end{equation*}
-where $\mt \Omega$ is a symmetric, block-diagonal matrix of size $7N \times 7N$ defined as
+It is convenient to rewrite the total kinetic energy of the original system, Eq.~\eqref{eq:kinetic energy}, as a quadratic form $K = \frac{1}{2}\tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega}(\vt q) [\substack{\vt p \\ \vt \pi}]$, where $\mt \Omega$ is a symmetric, block-diagonal matrix defined as
 \begin{equation}
@@
 
-We now proceed to show the final expression of $H_s$, which results from grouping the terms corresponding to forces and torques in the \textit{modified} potential energy $\tilde{U}$, given by
+We now proceed to present the final expression of $\Ham{S}$, which results from grouping the terms corresponding to forces and torques in the refined potential energy $\refined U(\vt r, \vt q, h^2)$, given by
 \begin{equation*}
-\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right).
+\refined U = U - \frac{\timestep^2}{24} \left( \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A} {\mt I}^{-1} {\mt A} {\vt \tau} \right),
 \end{equation*}
-while the remaining terms can be recast in a compact matrix, as follows
+while the terms that depend on the momentum vectors $\vt p$ and $\vt q$ are grouped together in the refined kinetic energy $\refined K$, which can be written in a compact matrix form as
 \begin{equation*}
-\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
+\refined K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \refined{\mathbf \Omega}(\vt r, \vt q, h^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
 \end{equation*}
@@
 \begin{equation*}
-\tilde{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} \left( {\mt \Omega} {\mt U}_{\vt x \vt x} \tr{\mt \Omega} \right)
-\end{equation*}
-and
-\begin{equation*}
-{\mt U}_{\vt x \vt x} = \left[\begin{array}{cc}
+\refined{\mt \Omega} = {\mt \Omega} + \frac{\timestep^2}{6} {\mt \Omega} \left[\begin{array}{cc}
 U_{\vt r \vt r} & U_{\vt r \vt q} \\
 U_{\vt q \vt r} & U_{\vt q \vt q}
-\end{array}\right].
+\end{array}\right] \tr{\mt \Omega}.
 \end{equation*}
 
-Finally, we have that
-\begin{equation*}
-\Ham S = \frac{1}{2} \tr{\left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \tilde{\mt \Omega} \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right] + \tilde{U} + \mathcal{O}({\timestep}^4),
-\end{equation*}
-
-Although $\refined{\mathbf \Omega}$ is still symmetric, it is no longer a block-diagonal matrix like $\mt \Omega$, due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both positions and orientations.
-As a result, the modified kinetic energy cannot, in principle, be split into independent translational and rotational contributions.
+Finally, the shadow Hamiltonian is given by $\Ham{S} = {\refined U} + {\refined K} + {\mathcal O}(\timestep^4)$.
 

### 2018-06-10 rigid_bodies.tex
Some small changes

@@
 
- \subsection*{Numerical Stability}
+\subsection{Numerical Stability}
 
-The explicit inclusion of the \textit{refined} Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
+The explicit inclusion of the refined Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
 This poses the question whether the symmetry loss will influence the long-term stability.
@@
 Here we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
-Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\ref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
-On the other hand, the symmetric scheme due to \citeauthor{Martyna_1996} \cite{Martyna_1996} should preserve $H$ (Eq.~\ref{eq:nvt extended energy}).
+Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\eqref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
+On the other hand, the symmetric scheme due to \citeauthor{Martyna_1996} \cite{Martyna_1996} should preserve $H$ (Eq.~\eqref{eq:nvt extended energy}).
 In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $\timestep$ = 5 fs.
@@
 The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
-As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the \textit{refined} Hamiltonian part for our NVT method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
-This has already been noticed by Davidchack\cite{Davidchack_2009}, and previously by Bond \textit{et al.} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.
+As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the refined Hamiltonian part for our NVT method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
+This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010}, and previously by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.
 
@@
 	\includegraphics{Figures/energy_drift.eps}
-    \caption{Influence of the time-step size ($\timestep$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its \textit{refined} Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the \textit{refined} Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\ref{eq:splitting NVE} with unsplit solution of rotation.}
+    \caption{Influence of the time-step size ($\timestep$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}
@@
 As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2009} discovered the existence of energy flows between the two thermostats.
+Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
 To the best of our knowledge, this finding did not result in a revision in the literature of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
@@
 
-\subsection*{Effect of Discretization Errors on Equilibrium Thermodynamic Properties}
+\subsection{Effect of Discretization Errors on Equilibrium Thermodynamic Properties}
 
@@
 They correspond to the results obtained employing $\timestep$ = 1 fs for the numerical scheme proposed in this paper.
-We have included the results of applying our method but using the approximate solution of rotation due to Miller \textit{et al.} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
+We have included the results of applying our method but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
 The temperature values of Fig.~\ref{fig:properties} (a) were computed from the simulated kinetic energies assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
@@
 Particularly, from Fig.~\ref{fig:properties} (a) we note that the simulated temperature decreases systematically with $\timestep$.
-This behavior is observed in NVE simulations\cite{Davidchack_2009,Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
+This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
 As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}.
@@
 With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for each time-step size the free energy difference per water molecule ($\Delta F/N$, circles) between the sampled and target states.
-$\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\ref{eq:reweighting}.
+$\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\eqref{eq:reweighting}.
 As expected, the free energy is only negligible for $h$ = 1 fs.
 
-In Fig.~\ref{fig:free_energy} we also depict the \textit{refined} temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
-Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the target temperature.
+In Fig.~\ref{fig:free_energy} we also depict the refined temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\eqref{eq:equipartition}.
+Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\eqref{eq:thermostat_force}, it is expected to reproduce the target temperature.
 Indeed, for all the time-step sizes considered the agreement is excellent.
@@
 We end this section with a discussion of the effect of discretization errors on the kinetic energy partition.
-In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition, which are calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times (3N - 3)$ and $\langle K_r \rangle_{eq}$ = $(\langle K \rangle /(6N -3)) \times 3N$.
+In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition, which are calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
 

### 2018-06-10 rigid_bodies.tex
energy drift figure corrected

@@
 Here we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
-Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\eqref{eq:this extended energy}) for our method, which from now on will be referred as ``This work''.
+Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\eqref{eq:this extended energy}) for our method, which from now on will be referred as ``Refined/Unsplit''.
 On the other hand, the symmetric scheme due to \citeauthor{Martyna_1996} \cite{Martyna_1996} should preserve $H$ (Eq.~\eqref{eq:nvt extended energy}).
@@
 Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
-Fig.~\ref{fig:energy_drift} also contains the drift in $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of our method.
+Fig.~\ref{fig:energy_drift} also contains the drift in $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of the Refined/Unsplit scheme.
 It is worth mentioning that the stability threshold for the NVE method corresponds to $\timestep$ = 6 fs.
 The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
-As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the refined Hamiltonian part for our NVT method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
+As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the refined Hamiltonian part for the Refined/Unsplit method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
 This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010}, and previously by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.
@@
 	\includegraphics{Figures/energy_drift.eps}
-    \caption{Influence of the time-step size ($\timestep$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
+    \caption{Influence of the time-step size ($\timestep$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for the Refined/Unsplit method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}

### 2018-06-10 Figures/Drift.py

@@ -10,10 +10,10 @@ from matplotlib import rc
 # activate latex text rendering
 rc('text', usetex=True)
 
-fig = plt.figure('axes',figsize=(3,2))
+fig = plt.figure('axes',figsize=(3.37,2))
 ax = fig.add_subplot(111)
 ax.set_xlabel('$h$ (fs)')
-ax.set_ylabel('$R$ [(kcal/mol) ns]')
+ax.set_ylabel('$R$ (kcal/mol.ns)')
 
 
 ax.set_xlim(0.5,10.5)
@@ -23,9 +23,9 @@ ax.set_xticks(xticks)
 lw = 0.5
 
 data = np.genfromtxt('Drift.csv', delimiter=',', skip_header=1, names=['timestep','thermo','Hs_nhc_thiswork','Hs_thiswork','H_nhc_martyna','Hs_nve'])
-ax.plot(data['timestep'], data['Hs_nhc_thiswork']/data['thermo'],linestyle ='-', color = 'black', marker = 'v',markersize = 4, linewidth = lw, label = r'$\widetilde{H}$ (This work) ')
-ax.plot(data['timestep'], data['H_nhc_martyna']/data['thermo'],linestyle ='-', color = 'green', marker = '^',markersize = 4, linewidth = lw, label = r'$H$ (Martyna \textit{et al.}) ')
-ax.plot(data['timestep'], data['Hs_thiswork']/data['thermo'],linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw,label = r'$\widetilde{\mathcal{H}}$ (This work)')
+ax.plot(data['timestep'], data['Hs_nhc_thiswork']/data['thermo'],linestyle ='-', color = 'black', marker = 'v',markersize = 4, linewidth = lw, label = r'$\widetilde{H}$ (Refined/Unsplit) ')
+ax.plot(data['timestep'], data['H_nhc_martyna']/data['thermo'],linestyle ='-', color = 'green', marker = '^',markersize = 4, linewidth = lw, label = r'$H$ (Martyna) ')
+ax.plot(data['timestep'], data['Hs_thiswork']/data['thermo'],linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw,label = r'$\widetilde{\mathcal{H}}$ (Refined/Unsplit)')
[…]

### 2018-06-10 rigid_bodies.tex
some corrections on disc. error

@@
 
-In the previous section we showed that the energy drift is mainly localized within the thermostat variables.
-It is then reasonable to ask if this leads to a more ``accurate'' sampling of the physical system.
+In the previous section we showed that the energy drift rate is mainly localized within the thermostat variables.
+It is then reasonable to ask if this leads to a more accurate sampling of the physical system.
 To answer this question, we turn our attention to the discretization errors.
 
-In Fig.~\ref{fig:properties} we show the computed average of some thermodynamic properties as a function of $\timestep$.
-The simulations were carried out specifying T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_v$).
+In Fig.~\ref{fig:properties} we show the computed averages of some thermodynamic properties as a function of $\timestep$.
+The simulations were carried out specifying T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_V$).
 The lines are a guide for the eye. 
 Also, the horizontal dotted lines (black) help to visualize the departure of the computed averages with increasing $\timestep$.
-They correspond to the results obtained employing $\timestep$ = 1 fs for the numerical scheme proposed in this paper.
-We have included the results of applying our method but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``This work (NO-SQUISH)''.
-The temperature values of Fig.~\ref{fig:properties} (a) were computed from the simulated kinetic energies assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
+They correspond to the results obtained employing $\timestep$ = 1 fs in the Refined/Unsplit numerical scheme.
+We have included the results of applying our method but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
+Recall that the computed averages for both Refined schemes are obtained via reweighting employing Eq.~\eqref{eq:reweighting}. 
+In addition, it should be pointed out that the temperature values in Fig.~\ref{fig:properties} (a) were computed from the simulated kinetic energies assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
 
@@
 
-From Fig.~\ref{fig:properties} it becomes clear that our method yields significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
-This is true also for the results corresponding to the use of the NO-SQUISH method\cite{Miller_2002} in the NVE part of the integrator.
+From Fig.~\ref{fig:properties} it becomes clear that the Refined schemes yields significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
+Note that a small, still clearly influencing systematic error is present, as the computed averages are slightly modified for $\timestep$ > 5.
+This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
 Mathematically, the derivation of the refined Hamiltonian was facilitated by the fewer number of operators associated with the unsplit solution of rotation.
-However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement.
-Also note that a small, still clearly influencing systematic error is present in our scheme, as both temperature and potential energy are slightly modified by $\timestep$.
-This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
+However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement between schemes Refined/Unsplit and Refined/NO-SQUISH.
+
+In Fig.~\ref{fig:properties} (a) we also depict the refined temperature, calculated by means of the generalized equipartition theorem, given by Eq.~\eqref{eq:equipartition}.
+Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\eqref{eq:thermostat_force} we expect the Refined schemes to reproduce the target temperature.
+Indeed, for all the time-step sizes considered the agreement is excellent.
 
@@
 As expected, the free energy is only negligible for $h$ = 1 fs.
-
-In Fig.~\ref{fig:free_energy} we also depict the refined temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\eqref{eq:equipartition}.
-Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\eqref{eq:thermostat_force}, it is expected to reproduce the target temperature.
-Indeed, for all the time-step sizes considered the agreement is excellent.
 
@@
 Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
-From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption for time-step sizes up to 5 fs.
+Fig.~\ref{fig:energy_partition} (a) was obtained directly from the contribution of translational and rotational degrees of freedom to the refined kinetic energy.
+It confirms that our approximation bla bla bla
+After reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects, for time-step sizes up to 5 fs, the serious disruption observed in the schemes of \citeauthor{Martyna_1996} \cite{Martyna_1996} and  \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}
 

### 2018-06-10 rigid_bodies.tex
correction on energy partition discussion

@@
 It is then reasonable to ask if this leads to a more accurate sampling of the physical system.
-To answer this question, we turn our attention to the discretization errors.
+To answer this question, we turn our attention to discretization effects.
 
@@
 
-From Fig.~\ref{fig:properties} it becomes clear that the Refined schemes yields significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
+From Fig.~\ref{fig:properties} it becomes clear that the refined schemes yield significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
 Note that a small, still clearly influencing systematic error is present, as the computed averages are slightly modified for $\timestep$ > 5.
@@
 In Fig.~\ref{fig:properties} (a) we also depict the refined temperature, calculated by means of the generalized equipartition theorem, given by Eq.~\eqref{eq:equipartition}.
-Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\eqref{eq:thermostat_force} we expect the Refined schemes to reproduce the target temperature.
+Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\eqref{eq:thermostat_force} we expect the refined schemes to reproduce the target temperature.
 Indeed, for all the time-step sizes considered the agreement is excellent.
@@
 
-We end this section with a discussion of the effect of discretization errors on the kinetic energy partition.
-In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition, which are calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
+We end this section with a discussion of discretization effects on the kinetic energy partition.
+In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition.
+In Figs.~\ref{fig:energy_partition} (a)-(c) these were calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
+For Fig.~\ref{fig:energy_partition} (d) we substituted $\langle K \rangle$ for $\langle \refined{K} \rangle$ in the right-hand side of the equations above.
 
@@
 We see that the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
-The results of Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
+The results of Fig.~\ref{fig:energy_partition} (a), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
 Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
-Fig.~\ref{fig:energy_partition} (a) was obtained directly from the contribution of translational and rotational degrees of freedom to the refined kinetic energy.
-It confirms that our approximation bla bla bla
-After reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects, for time-step sizes up to 5 fs, the serious disruption observed in the schemes of \citeauthor{Martyna_1996} \cite{Martyna_1996} and  \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}
+From Fig.~\ref{fig:energy_partition} (d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to 5 fs.
+For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects the serious disruption observed in the schemes of \citeauthor{Martyna_1996} \cite{Martyna_1996} and  \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
 

### 2018-06-10 Figures/KE_partition.py

@@ -17,6 +17,7 @@ ax[3].set_xlabel('$h$ (fs)')
 
 ax[0].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
 ax[2].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
+ax[3].set_ylabel(r'{$\langle \widetilde{K} \rangle$} (kcal/mol)')
 ax[0].set_xlim(0.5,7.5)
 ax[0].set_ylim(740,825)
 
@@ -29,18 +30,18 @@ lw = 0.5
 kB = 0.0019872041 #Boltzmann's constant (kcal/mol/K).
 N = 903
 beta = 1./(kB*298.0)
-integrators = ['S1','P1','K2']
+integrators = ['P1','K2','S1']
 colors = ['black','green','blue']
 markers = ['s','^','o','v']
 
 data = np.genfromtxt('S1_KEpartition.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','KEt', \
                      'dKEt','KEr','dKEr','Kt_eq','dKt_eq','Kr_eq','dKr_eq'])
-ax[0].errorbar(timesteps, data['KEt'][0:7],yerr = data['dKEt'][0:7],linestyle ='-', color = 'black', marker = 'v', markersize = 4, linewidth = lw, label = r'{$\langle K_t \rangle$}',capsize=2, markeredgewidth=1)
-ax[0].errorbar(timesteps, data['KEr'][0:7],yerr = data['dKEr'][0:7],linestyle ='-', color = 'black', marker = '^', markersize = 4, linewidth = lw, label = r'{$\langle K_r \rangle$}',capsize=2, markeredgewidth=1)
-ax[0].errorbar(timesteps, data['Kt_eq'][0:7],yerr =data['dKt_eq'][0:7],linestyle ='-', color = 'green', marker = 's', markersize = 4, linewidth = lw, label = r'{$\langle K_t \rangle_{eq}$}', capsize=2,  markeredgewidth=1)
-ax[0].errorbar(timesteps, data['Kr_eq'][0:7],yerr = data['dKr_eq'][0:7],linestyle ='-', color = 'red', marker = 'x', markersize = 4, linewidth = lw, label = r'{$\langle K_r \rangle_{eq}$}', capsize=2, markeredgewidth=1)
[…]

### 2018-06-11 Figures/energy_drift.py
Inclusion of Makefile

@@ -1,0 +1,41 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.font_manager import FontProperties
+plt.rcParams.update({'font.size':8})
+from matplotlib import rc
+
+# activate latex text rendering
+rc('text', usetex=True)
+
+fig = plt.figure('axes',figsize=(3.37,2))
+ax = fig.add_subplot(111)
+ax.set_xlabel('$h$ (fs)')
+ax.set_ylabel('$R$ (kcal/mol.ns)')
+
+
+ax.set_xlim(0.5,10.5)
+xticks = [1,2,3,4,5,6,7,8,9,10]
+ax.set_xticks(xticks)
+
+lw = 0.5
+
+data = np.genfromtxt('Drift.csv', delimiter=',', skip_header=1, names=['timestep','thermo','Hs_nhc_thiswork','Hs_thiswork','H_nhc_martyna','Hs_nve'])
+ax.plot(data['timestep'], data['Hs_nhc_thiswork']/data['thermo'],linestyle ='-', color = 'black', marker = 's',markersize = 4, linewidth = lw, label = r'$\widetilde{H}$ (Refined) ')
+ax.plot(data['timestep'], data['H_nhc_martyna']/data['thermo'],linestyle ='-', color = 'red', marker = 'x',markersize = 4, linewidth = lw, label = r'$H$ (Martyna) ')
+ax.plot(data['timestep'], data['Hs_nve']/data['thermo'],linestyle ='-', color = 'blue', marker = 'o',markersize = 4, linewidth = lw,label = r'$\widetilde{\mathcal{H}}$ (NVE)')
+ax.plot(data['timestep'], data['Hs_thiswork']/data['thermo'],linestyle ='-', color = 'green', marker = '^',markersize = 4, linewidth = lw,label = r'$\widetilde{\mathcal{H}}$ (Refined)')
+
+
+ax.set_ylim(-0.5,5)
+ax.legend(loc='upper left', ncol=1, fancybox=True, frameon=False, numpoints = 1)
+
+inset = fig.add_axes([.58, .5, .3, .3])
+inset.semilogy(data['timestep'], data['Hs_nhc_thiswork']/data['thermo'],linestyle ='-', color = 'black', marker = 's',markersize = 4, linewidth = lw, label = r'$\widetilde{H}$ (Refined) ')
[…]

### 2018-06-11 rigid_bodies.tex
Changes in Numerical Stability section

@@
 
-The explicit inclusion of the refined Hamiltonian in a computationally cheap NVT integrator is achieved at the cost of losing the time-reversibility.
-This poses the question whether the symmetry loss will influence the long-term stability.
+As described in Sec.~\ref{sec:refined_method}, the refined kinetic energy $\refined K$ can be explicitly included in an NVT integrator without a substantial increase in computational effort.
+However, this can only be achieved by limiting the order of approximation of numerical derivatives and sacrificing short-term reversibility.
+This poses the question whether these conditions will influence the long-term stability of the method, especially if one tries to employ long time steps.
 
-Here we define the energy drift, denoted by $R$, as the the rate of change of energy with time, which corresponds to the angular coefficient obtained from linear regression fitting.
-Recall that the quantity expected to be conserved along the simulation is $\widetilde{H}$ (Eq.~\eqref{eq:this extended energy}) for our method, which from now on will be referred as ``Refined/Unsplit''.
-On the other hand, the symmetric scheme due to \citeauthor{Martyna_1996} \cite{Martyna_1996} should preserve $H$ (Eq.~\eqref{eq:nvt extended energy}).
-In Fig.~\ref{fig:energy_drift}, we present the computed $R$ as a function of the time-step size, and one can see that for these numerical schemes the energy starts to drift upwards from $\timestep$ = 5 fs.
-Also, the long-term stability is only modestly improved by our method. Evidently, the impact of the lack of symmetry on $R$, if any, is negligible.
-Fig.~\ref{fig:energy_drift} also contains the drift in $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of the Refined/Unsplit scheme.
-It is worth mentioning that the stability threshold for the NVE method corresponds to $\timestep$ = 6 fs.
+Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in an equilibrated MD simulation.
+This corresponds to the angular coefficient of such quantity \textit{vs}. time as determined by linear regression.
+Recall that the symmetric scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
+In the case of our proposed method, which we refer to here as Refined NHC, the conserved extended energy is $\refined H$, such as defined by Eq.~\eqref{eq:this extended energy}.
+In Fig.~\ref{fig:energy_drift}, we present drift rates of these quantities computed in liquid water simulations carried out with different time-step sizes.
+As one can observe in the inset, very large drifts occur for both methods when $\timestep \ge 6~\text{fs}$.
+Up to $\timestep = 5~\text{fs}$, the long-term stability is slightly improved by our method.
+Evidently, there is little or no impact of the lack of time-reversal symmetry on $R$.
+Fig.~\ref{fig:energy_drift} also contains the drift in the refined Hamiltonian $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of the Refined NHC scheme.
+It is worth mentioning that $\timestep = 6~\text{fs}$ marks a stability threshold for the NVE method.
+
 The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
@@
 	\includegraphics{Figures/energy_drift.eps}
-    \caption{Influence of the time-step size ($\timestep$) on the long-time drift ($R$) in: the extended energy ($\refined{H}$, down triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for the Refined/Unsplit method, as well as in the extended energy ($H$, up triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
+	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, down-pointing triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up-pointing triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}
@@
 As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Considering a factorization scheme different to the one employed by those authors, Davidchack \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
+Considering a factorization scheme different to the one employed by those authors, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
 To the best of our knowledge, this finding did not result in a revision in the literature of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
@@
 	\includegraphics{Figures/numerical_stability.eps}
-    \caption{Long-time drift in (a) the extended energy and (b) in each thermostat energy for the NVT method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
+	\caption{Long-time drift rate in (a) the extended energy and (b) in each thermostat energy for the NVT method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
 In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing to the thermostat coupled to the rotational degrees of freedom.}

### 2018-06-11 Figures/Free_Energy.py
Inclusion of numerical stability and NVE figures

@@ -1,5 +1,4 @@
 import numpy as np
-import utility_belt as ub
 import matplotlib.pyplot as plt
 from matplotlib.font_manager import FontProperties
 import mics

### 2018-06-11 Figures/NVE.py

@@ -1,5 +1,4 @@
 import numpy as np
-import utility_belt as ub
 import matplotlib.pyplot as plt
 from matplotlib.font_manager import FontProperties
 import mics
@@ -23,9 +22,7 @@ captions = [r'$h=1$ fs',r'$h=2$ fs',r'$h=3$ fs',r'$h=4$ fs']
 color_c = ['black', 'red', 'green', 'blue' ]
 k = 0
 for i in reversed(timesteps):
-    a = ub.lammps_parser('NVE_' + str(i) + 'fs.log', run=2)
-    p = a.data_frame()
-    df = p.iloc[::150, :]
+    df = pd.read_csv('NVE_' + str(i) + 'fs.csv')
     ax.plot(df['Step']*i*1e-6-0.9, df['TotEng'], color = color[k],label=r'$\mathcal{H}$')
     ax.plot(df['Step']*i*1e-6-0.9, df['Hs'],linestyle = 'dotted', color = color[k],label=r'$\widetilde{\mathcal{H}}$')
     ax.annotate(captions[k], xy=(0, 0) , xycoords= 'axes fraction', xytext=xy[0][k],  \

### 2018-06-11 rigid_bodies.tex
abstract

--- 2018-06-10 rigid_bodies.tex
@@
 
-%\begin{abstract}
-%\end{abstract}
+\begin{abstract}
+
+Deterministic and stochastic Molecular Dynamics methods aimed at reproducing a target temperature share the equipartition theorem as a premise.
+While this intends to mimic the real world physics, discretization errors, which are intrinsic to numerical integrators, may severely distort such assumption.
+Fortunately, Backward Error Analysis is a powerful tool that allows to identify the quantity that is actually subject to equipartition, which is a shadow Hamiltonian that only coincides with the specified Hamiltonian
+when the time-step size approaches zero.
+This paper deals with discretization effects in a straightforward way: with a minimal computational overhead, we obtain a refined kinetic energy and a refined Hamiltonian as suitable estimators of their
+shadow counterparts and then we tune the thermostatting procedure with this refined kinetic energy.
+The case of rigid water has served for an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
+By applying our new method to the Nos\'{e}-Hoover chain thermostat, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 6 fs.
+\end{abstract}
 

### 2018-06-11 rigid_bodies.tex
Revision of the Numerical Stability section

--- 2018-06-11 rigid_bodies.tex
@@
 In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
-The case of rigid water exemplifies this behavior and serves as an excellent testbed for our proposal.
+The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
 In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
@@
 
+\subsection{Refined Hamiltonian in NVE Simulations}
+
 We start by examining the departure of the \textit{refined} Hamiltonian from the specified one in NVE simulations.
@@
 Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in an equilibrated MD simulation.
-This corresponds to the angular coefficient of such quantity \textit{vs}. time as determined by linear regression.
+This corresponds to the angular coefficient of such quantity \textit{vs}. time and is determined by linear regression.
 Recall that the symmetric scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
-In the case of our proposed method, which we refer to here as Refined NHC, the conserved extended energy is $\refined H$, such as defined by Eq.~\eqref{eq:this extended energy}.
-In Fig.~\ref{fig:energy_drift}, we present drift rates of these quantities computed in liquid water simulations carried out with different time-step sizes.
-As one can observe in the inset, very large drifts occur for both methods when $\timestep \ge 6~\text{fs}$.
-Up to $\timestep = 5~\text{fs}$, the long-term stability is slightly improved by our method.
-Evidently, there is little or no impact of the lack of time-reversal symmetry on $R$.
-Fig.~\ref{fig:energy_drift} also contains the drift in the refined Hamiltonian $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017} (circles), which almost coincides with the results of the Refined NHC scheme.
-It is worth mentioning that $\timestep = 6~\text{fs}$ marks a stability threshold for the NVE method.
+In the case of our proposed method, which we refer to as Refined NHC, the conserved extended energy is $\refined H$, such as defined by Eq.~\eqref{eq:this extended energy}.
+In Fig.~\ref{fig:energy_drift}, we present drift rates of these quantities computed in liquid water simulations, with $18~\text{ns}$ of production time, carried out with different time-step sizes.
+As one can observe, long-term stability is slightly improved by our method for $\timestep \le 5~\text{fs}$, indicating that there is little or no impact of the lack of time-reversal symmetry on $R$.
 
-The NVT methods discussed so far allow to extend that limit to 10 fs (see Table~\ref{table:stability} in the Supplementary Material).
-As one can notice in Fig.~\ref{fig:energy_drift}, the drift in the refined Hamiltonian part for the Refined/Unsplit method (squares) remains bounded for all the time-step sizes considered, so the thermostat is stabilizing the numerical integration.
-This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010}, and previously by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical error''.
+Fig.~\ref{fig:energy_drift} also contains drift rates in the refined Hamiltonian $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017}, which almost coincide with the results for the Refined NHC scheme up to $h = 6~\text{fs}$.
+However, the NVE integration becomes fully unstable for time step sizes above this value.
+Each NVT integrator, on the other hand, manages to remain numerically stable at least up to $h = 10~\text{fs}$, but displays an exceedingly large drift in its corresponding extended energy, as one can see the inset of Fig.~\ref{fig:energy_drift}.
+Trying to look for the cause of this behavior, we also present drift rates computed for the Hamiltonian $\refined{\Ham{}}$ in the simulations carried out with the Refined NHC method.
+Recall that this part of $\refined H$ represents the total energy of the physical system.
+Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time-step size, which shows that the thermostat stabilizes the numerical integration of the physical variables.
+This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010} and also by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
 
@@
 
-To end this section, we present in Fig.~\ref{fig:num_stab} the anomalous behavior encountered when using the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
-Particularly, one can see from Fig.~\ref{fig:num_stab} (a) that for each time-step size the extended energy drifts upwards until reaching a point at which the integrator becomes stable.
-It should be remarked that these results were double-checked by using the LAMMPS software package \cite{Plimpton_1995}. 
-As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom.
-Considering a factorization scheme different to the one employed by those authors, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
-To the best of our knowledge, this finding did not result in a revision in the literature of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
-As can be seen in Fig.~\ref{fig:num_stab} (b), where we depict the energy of each thermostat, that flux is also observed for the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
-In addition, the negative values of energy correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing from this thermostat to the one coupled to the rotational degrees of freedom.
-Interestingly, Fig.~\ref{fig:num_stab} (b) suggests that a unique steady state is established, and the time to reach the steady state decreases with $\timestep$.
-As far as we know, this behavior has not been reported before.
+To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior encountered when the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} is employed.
+Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each considered time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
+These results were obtained with our own simulation code and double-checked using LAMMPS \cite{Plimpton_1995} with a custom pair style.
+As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate NHC thermostats to the rotational and translational degrees of freedom.
+Considering a different factorization scheme, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
+%Nevertheless, this finding did not result in a revision of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in the literature.
+This is confirmed for the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in Fig.~\ref{fig:num_stab}(b), where we depict separately the contribution of each thermostat to the extended energy $H$.
+In this figure, the negative energy values always correspond to the thermostat coupled to the translational degrees of freedom.
+Therefore, energy flows from this thermostat to the one coupled to the rotational degrees.
+Interestingly, Fig.~\ref{fig:num_stab} shows that the rate at which the thermostats approach the steady state decreases with $\timestep$. As far as we know, this behavior has not been reported before.
 

### 2018-06-12 rigid_bodies.tex
Further revision of two result sections

--- 2018-06-11 rigid_bodies.tex
@@
 
-%\begin{abstract}
-%\end{abstract}
+\begin{abstract}
+%Many deterministic and stochastic molecular dynamics methods that aim at reproducing a target temperature have equipartition of kinetic energy as a premise.
+%While this intends to mimic the real world, discretization errors, which are intrinsic to numerical integrators, may distort such assumption.
+%Fortunately, backward error analysis is a powerful tool that allows us to identify the quantity actually subject to equipartition, which is a shadow Hamiltonian that only coincides with the specified Hamiltonian when the time-step size approaches zero.
+%In this paper,  deals with discretization effects in a straightforward way: with a minimal computational overhead, we obtain a refined kinetic energy and a refined Hamiltonian as suitable estimators of their
+%shadow counterparts and then we tune the thermostatting procedure with this refined kinetic energy.
+%The case of rigid water has served for an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
+%By applying our new method to the Nos\'{e}-Hoover chain thermostat, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 5 fs.
+\end{abstract}
 
@@
 The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
-In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 6 fs, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
+In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
 
@@
 The system under study has periodic boundary conditions and contains 903 TIP3P \cite{Jorgensen_1983} water molecules at a density of 970~kg/m\textsuperscript{3}.
-The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at 10 \AA and smoothed by a switching function over the range from 9.5 \AA to 10 \AA, such as done in Ref.~\citenum{Silveira_2017}.
+The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at 10 {\AA} and smoothed by a switching function over the range from 9.5 {\AA} to 10 {\AA}, such as done in Ref.~\citenum{Silveira_2017}.
 The damping of Coulombic interactions was done by a factor $\text{erfc}(-\alpha r)$, with $\alpha = 0.29~\text{\AA}^{-1}$.
@@
 
-\subsection{Refined Hamiltonian in NVE Simulations}
+\subsection{Refined Hamiltonian Calculation in NVE Simulations}
 
-We start by examining the departure of the \textit{refined} Hamiltonian from the specified one in NVE simulations.
-This is the most direct way of determining the importance of the discretization errors \cite{Engle_2005}, which is essential for discussing the NVT case.
-In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system using different time-step sizes ($\timestep$).
+We start by examining the departure of the refined Hamiltonian from the specified one in NVE simulations.
+This is the most direct way of determining the magnitude of the discretization errors \cite{Engle_2005}, which is essential for discussing the NVT case.
+In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system using different time-step sizes.
 For this, we used the numerical scheme of Eq.~\eqref{eq:splitting NVE} with the unsplit solution of Ref.~\citenum{Silveira_2017} applied for the rotations.
 The four-point differentiation formulas presented in Sec.~\ref{sec:refined_method} were employed for computing the refined kinetic energy.
-As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep = 1~\text{fs}$, but rapidly departs from $\Ham{}$ when $\timestep$ increases.
+As expected, $\refined{\Ham{}}$ is very well conserved and is close to $\Ham{}$ when $\timestep = 1~\text{fs}$, but rapidly departs from $\Ham{}$ when $\timestep$ increases.
 It is worth remarking that $\refined{\Ham{}}$ corresponds to a simple and, most importantly, computationally inexpensive approximation of the truly conserved shadow Hamiltonian.
@@
 1) the employed four-point formulas do not provide derivatives with the required accuracy or
-2) the high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would have a non-negligible contribution to the shadow Hamiltonian.
+2) the neglected high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would contribute notably to the shadow Hamiltonian.
 
@@
 
-As described in Sec.~\ref{sec:refined_method}, the refined kinetic energy $\refined K$ can be explicitly included in an NVT integrator without a substantial increase in computational effort.
+As described in Sec.~\ref{sec:refined_method}, it is possible to introduce the refined kinetic energy $\refined K$ in an NVT integrator without a substantial increase in computational effort.
 However, this can only be achieved by limiting the order of approximation of numerical derivatives and sacrificing short-term reversibility.
-This poses the question whether these conditions will influence the long-term stability of the method, especially if one tries to employ long time steps.
+This poses the question whether these conditions will influence the long-term stability of the method, especially if one tries to make use of large time steps.
 
 Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in an equilibrated MD simulation.
-This corresponds to the angular coefficient of such quantity \textit{vs}. time and is determined by linear regression.
+This corresponds to the angular coefficient of such quantity expressed as a function of time and is determined by least-square regression.
 Recall that the symmetric scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
@@
 In Fig.~\ref{fig:energy_drift}, we present drift rates of these quantities computed in liquid water simulations, with $18~\text{ns}$ of production time, carried out with different time-step sizes.
-As one can observe, long-term stability is slightly improved by our method for $\timestep \le 5~\text{fs}$, indicating that there is little or no impact of the lack of time-reversal symmetry on $R$.
-
-Fig.~\ref{fig:energy_drift} also contains drift rates in the refined Hamiltonian $\refined{\Ham{}}$ for the NVE method \cite{Silveira_2017}, which almost coincide with the results for the Refined NHC scheme up to $h = 6~\text{fs}$.
-However, the NVE integration becomes fully unstable for time step sizes above this value.
-Each NVT integrator, on the other hand, manages to remain numerically stable at least up to $h = 10~\text{fs}$, but displays an exceedingly large drift in its corresponding extended energy, as one can see the inset of Fig.~\ref{fig:energy_drift}.
-Trying to look for the cause of this behavior, we also present drift rates computed for the Hamiltonian $\refined{\Ham{}}$ in the simulations carried out with the Refined NHC method.
-Recall that this part of $\refined H$ represents the total energy of the physical system.
-Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time-step size, which shows that the thermostat stabilizes the numerical integration of the physical variables.
-This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010} and also by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
+As one can observe, long-term stability is slightly improved by our method if $\timestep \le 5~\text{fs}$, indicating that there is little or no impact of the lack of time-reversal symmetry on $R$.
 
@@
 
-To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior encountered when the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} is employed.
+Fig.~\ref{fig:energy_drift} also contains drift rates computed for the refined Hamiltonian $\refined{\Ham{}}$ in an NVE simulation \cite{Silveira_2017}.
+Up to $h = 6~\text{fs}$, these rates practically coincide with those obtained for $\refined H$ with the Refined NHC scheme.
+This indicates that the major causes for the observed drift took place during evaluations of the NVE propagator.
+As the interaction potential we use is very smooth at cutoff, a possible cause is the amplification of round-off errors which can occur in events of close interatomic approaches \cite{Engle_2005}.
+For time steps larger than $6~\text{fs}$, the NVE integration becomes fully unstable.
+In contrast, each NVT integrator manages to remain numerically stable at least up to $h = 10~\text{fs}$, but at the cost of displaying an exceedingly large drift in its corresponding extended energy, as one can see the inset of Fig.~\ref{fig:energy_drift}.
+Trying to look for the cause of this behavior, we also present drift rates computed for the Hamiltonian $\refined{\Ham{}}$ in the simulations carried out with the Refined NHC method.
+Recall that this part of $\refined H$ represents the total energy of the physical system.
+Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time steps, which shows that the thermostat stabilizes the numerical integration of the physical variables.
+This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010} and also by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
+
+To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} is employed.
 Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each considered time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
@@
 	\includegraphics{Figures/numerical_stability.eps}
-	\caption{Long-time drift rate in (a) the extended energy and (b) in each thermostat energy for the NVT method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
-In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom, so the energy is flowing to the thermostat coupled to the rotational degrees of freedom.}
+	\caption{Long-term drift rate in (a) the extended energy and (b) each thermostat energy for the NVT method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
+In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom.}
 	\label{fig:num_stab}
@@
 
-\subsection{Effect of Discretization Errors on Equilibrium Thermodynamic Properties}
+\subsection{Discretization Effects on Equilibrium Thermodynamic Properties (Work in Progress)}
 
-In the previous section we showed that the energy drift rate is mainly localized within the thermostat variables.
-It is then reasonable to ask if this leads to a more accurate sampling of the physical system.
-To answer this question, we turn our attention to discretization effects.
+%In the previous section, we have shown that the extended energy drifts in NVT simulations are localized within the thermostat variables.
+%It is then reasonable to ask if this leads to a more accurate sampling of the physical system.
+%To answer this question, we turn our attention to discretization effects.
 
-In Fig.~\ref{fig:properties} we show the computed averages of some thermodynamic properties as a function of $\timestep$.
-The simulations were carried out specifying T = 298 K and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume heat capacity ($C_V$).
-The lines are a guide for the eye. 
+In Fig.~\ref{fig:properties}, we show the computed averages of some thermodynamic properties as a function of $\timestep$.
+The simulations were carried out by specifying $T = 298~\text{K}$ and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume specific heat capacity ($C_V$).
+The lines are a guide for the eyes. 
 Also, the horizontal dotted lines (black) help to visualize the departure of the computed averages with increasing $\timestep$.
-They correspond to the results obtained employing $\timestep$ = 1 fs in the Refined/Unsplit numerical scheme.
-We have included the results of applying our method but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
-Recall that the computed averages for both Refined schemes are obtained via reweighting employing Eq.~\eqref{eq:reweighting}. 
-In addition, it should be pointed out that the temperature values in Fig.~\ref{fig:properties} (a) were computed from the simulated kinetic energies assuming equipartition, ie $2 \langle K \rangle = (6N -3) k_b T$.
+They correspond to the results obtained employing $\timestep = 1~\text{fs}$ in the Refined NHC numerical scheme.
+We have included the results of applying our method, but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
+Recall that the computed averages for both refined schemes are obtained via reweighting by employing Eq.~\eqref{eq:reweighting}. 
+In addition, it should be pointed out that the temperature values in Fig.~\ref{fig:properties}(a) were computed from the simulated kinetic energies assuming equipartition, i.e.
+$2 \langle K \rangle = (6N -3) k_b T$.
 
-Despite the non-equilibrium process ocurring within the thermostats, it is worth showing the results obtained with the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, because that is not the only problematic issue.
-Particularly, from Fig.~\ref{fig:properties} (a) we note that the simulated temperature decreases systematically with $\timestep$.
+Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, because that is not the only problematic issue.
+Particularly, from Fig.~\ref{fig:properties}(a) we note that the simulated temperature decreases systematically with $\timestep$.
 This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
 As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}.
-Figs.~\ref{fig:properties} (a)-(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
+Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
 As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
@@
 
-In Fig.~\ref{fig:properties} (a) we also depict the refined temperature, calculated by means of the generalized equipartition theorem, given by Eq.~\eqref{eq:equipartition}.
+In Fig.~\ref{fig:properties}(a) we also depict the refined temperature, calculated by means of the generalized equipartition theorem, given by Eq.~\eqref{eq:equipartition}.
 Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\eqref{eq:thermostat_force} we expect the refined schemes to reproduce the target temperature.
@@
 \label{sec:conclusion}
+
+To be included.
 
@@
 
-\suppinfo
+%\suppinfo
 

### 2018-06-12 Figures/thermodynamic_properties.py

@@ -16,7 +16,7 @@ ax[3].set_xlabel('$h$ (fs)')
 ax[0].set_ylabel(r'T (K)')
 ax[1].set_ylabel(r'{$\langle U/N \rangle$} (kcal/mol)')
 ax[2].set_ylabel(r'{$\langle W/N \rangle$} (kcal/mol)')
-ax[3].set_ylabel(r'$C_V$ (kcal/mol K)')
+ax[3].set_ylabel(r'$C_V$ (cal/mol K)')
 
 
 ax[1].set_xlim(0.5,7.5)
@@ -41,7 +41,7 @@ for i in integrators:
     ax[0].errorbar(timesteps, data['temperatura'][0:7],yerr = data['dtemperatura'][0:7],linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2,   markeredgewidth=1)
     ax[1].errorbar(timesteps, data['P'][0:7],yerr = data['dP'][0:7],linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2,  markeredgewidth=1)
     ax[2].errorbar(timesteps, data['virial'][0:7],yerr = data['dvirial'][0:7],linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2,  markeredgewidth=1)
-    ax[3].errorbar(timesteps, data['Cv1'][0:7],yerr = data['dCv1'][0:7],linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2, markeredgewidth=1)
+    ax[3].errorbar(timesteps, data['Cv1'][0:7]/0.903,yerr = data['dCv1'][0:7]/0.903,linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2, markeredgewidth=1)
 
     k = k + 1
 
@@ -50,12 +50,12 @@ data = np.genfromtxt('S1wr.csv', delimiter=',', skip_header=1, names=[' ','T','f
                      'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
[…]

### 2018-06-14 rigid_bodies.tex
Changes in Discretization Errors section

@@
 }
+
 % Comment below before submitting
@@
 \usetikzlibrary{decorations.text}
+
 \newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
@@
 \newcommand{\refined}[1]{\widetilde{#1}}
+\newcommand{\mini}[1]{\text{\tiny{#1}}}
 
@@
 \keywords{aaa, bbb, ccc}
+
+\date{\today}
 
@@
 
-\subsection{Discretization Effects on Equilibrium Thermodynamic Properties (Work in Progress)}
+\subsection{Correction of Discretization Effects on Thermodynamic Properties}
 
-%In the previous section, we have shown that the extended energy drifts in NVT simulations are localized within the thermostat variables.
-%It is then reasonable to ask if this leads to a more accurate sampling of the physical system.
-%To answer this question, we turn our attention to discretization effects.
+In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing full-fledged instability from happening.
+This allows simulations to be executed with large time steps, which could in principle save computer time.
+However, as we will show now, discretization errors can become manifest much earlier than any risk of instability.
+Fortunately, our proposed methodology is capable of mitigating the effect of these errors on the ensemble averages computed for the simulated system.
 
-In Fig.~\ref{fig:properties}, we show the computed averages of some thermodynamic properties as a function of $\timestep$.
-The simulations were carried out by specifying $T = 298~\text{K}$ and the analyzed quantities correspond to temperature ($T$), potential energy ($U$) and virial ($W$) per molecule, as well as the constant-volume specific heat capacity ($C_V$).
-The lines are a guide for the eyes. 
-Also, the horizontal dotted lines (black) help to visualize the departure of the computed averages with increasing $\timestep$.
-They correspond to the results obtained employing $\timestep = 1~\text{fs}$ in the Refined NHC numerical scheme.
+In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid water system using several NVT-MD methods with different time step sizes.
+All simulations were carried out by specifying $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
+the mean potential energy per molecule $\langle U/N \rangle$,
+the mean internal virial per molecule $\langle W/N \rangle$, and
+the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
+The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
+All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
+The horizontal, dotted lines are meant to help visualizing the departure of the computed averages from their expected values.
+They correspond to the results obtained by employing the Refined NHC method with $\timestep = 1~\text{fs}$.
+Error bars were obtained using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
+
+\begin{figure*}
+	\includegraphics{Figures/thermodynamic_properties.eps}
+	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
+	\label{fig:properties}
+\end{figure*}
+
 We have included the results of applying our method, but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
 Recall that the computed averages for both refined schemes are obtained via reweighting by employing Eq.~\eqref{eq:reweighting}.
-In addition, it should be pointed out that the temperature values in Fig.~\ref{fig:properties}(a) were computed from the simulated kinetic energies assuming equipartition, i.e.
-$2 \langle K \rangle = (6N -3) k_b T$.
 
@@
 For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects the serious disruption observed in the schemes of \citeauthor{Martyna_1996} \cite{Martyna_1996} and  \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
-
-\begin{figure*}
-	\includegraphics{Figures/thermodynamic_properties.eps}
-	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
-	\label{fig:properties}
-\end{figure*}
 

### 2018-06-14 Figures/thermodynamic_properties.py

@@ -13,10 +13,10 @@ fig.set_size_inches(w=7.5,h=4)
 ax[2].set_xlabel('$h$ (fs)')
 ax[3].set_xlabel('$h$ (fs)')
 
-ax[0].set_ylabel(r'T (K)')
+ax[0].set_ylabel(r'$T_\textnormal{est}$ (K)')
 ax[1].set_ylabel(r'{$\langle U/N \rangle$} (kcal/mol)')
 ax[2].set_ylabel(r'{$\langle W/N \rangle$} (kcal/mol)')
-ax[3].set_ylabel(r'$C_V$ (cal/mol K)')
+ax[3].set_ylabel(r'$C_V$ (cal/mol.K)')
 
 
 ax[1].set_xlim(0.5,7.5)

### 2018-06-14 rigid_bodies.tex
Change of style from ACHEMSO to REVTEX

@@
 \documentclass[
-journal=jctcce,
-layout=twocolumn
-]{achemso}
+	aip,
+	jcp,
+%	preprint,
+	reprint,
+]{revtex4-1}
 
-\setkeys{acs}{
-	abbreviations = false,
-	articletitle  = false,
-	keywords      = false,
-	maxauthors    = 10,
-	super         = true
-}
-
-% Comment below before submitting
-\let\titlefont\undefined
-\usepackage[fontsize=11pt]{scrextend}
 \usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
-%\flushbottom
-% Comment above before submitting
-
-\usepackage{xr}
-\externaldocument{rigid_bodies_supp_info}
-\usepackage{booktabs}
-\usepackage{threeparttable}
+\usepackage{natbib}
+\usepackage{natmove}
 \usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage[T1]{fontenc}
-\usepackage[table]{xcolor}
-\definecolor{lightgray}{gray}{0.9}
 \usepackage{tikz}
 \usetikzlibrary{decorations.text}
+
+%\usepackage{xr}
+%\usepackage{booktabs}
+%\usepackage{threeparttable}
+%\usepackage[T1]{fontenc}
+%\usepackage[table]{xcolor}
+%\definecolor{lightgray}{gray}{0.9}
+%\usepackage[exponent-product=\times]{siunitx}
 
@@
 \newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
-\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
 \newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}           % Hamiltonian
@@
 
-\usepackage{array}
-\newcolumntype{L}{>{$}l<{$}}
-\newcolumntype{C}{>{$}c<{$}}
-\newcolumntype{R}{>{$}r<{$}}
-\usepackage{booktabs}    
-\usepackage[exponent-product=\times]{siunitx}
-
 \newenvironment{smallarray}[1]                          % small arrays
-{\null\,\vcenter\bgroup\scriptsize
+{
+	\null\,\vcenter\bgroup\scriptsize
 	\renewcommand{\arraystretch}{1.5}
@@
 	\hbox\bgroup$\left[\array{@{}#1@{}}}
-{\endarray\right]$\egroup\egroup\,\null}
-\listfiles
+	{\endarray\right]$\egroup\egroup\,\null
+}
+
+\begin{document}
 
@@
 
-\begin{document}
-
 %\begin{tocentry}
@@
 \end{abstract}
+
+\maketitle
 
@@
 
-\begin{acknowledgement}
+\begin{acknowledgements}
 	The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).
-\end{acknowledgement}
+\end{acknowledgements}
 
 \appendix
-\numberwithin{equation}{section}
+%\numberwithin{equation}{section}
 

### 2018-06-14 rigid_bodies.tex
Keywords and other small changes

@@
 
-\keywords{aaa, bbb, ccc}
+\keywords{Molecular Dynamics, Thermostat, Shadow Hamiltonian, Backward Error Analysis}
 
@@
 
-%\begin{tocentry}
-%\end{tocentry}
-
 \begin{abstract}
-%Many deterministic and stochastic molecular dynamics methods that aim at reproducing a target temperature have equipartition of kinetic energy as a premise.
-%While this intends to mimic the real world, discretization errors, which are intrinsic to numerical integrators, may distort such assumption.
-%Fortunately, backward error analysis is a powerful tool that allows us to identify the quantity actually subject to equipartition, which is a shadow Hamiltonian that only coincides with the specified Hamiltonian when the time-step size approaches zero.
-%In this paper,  deals with discretization effects in a straightforward way: with a minimal computational overhead, we obtain a refined kinetic energy and a refined Hamiltonian as suitable estimators of their
-%shadow counterparts and then we tune the thermostatting procedure with this refined kinetic energy.
-%The case of rigid water has served for an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
-%By applying our new method to the Nos\'{e}-Hoover chain thermostat, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 5 fs.
+Many deterministic and stochastic molecular dynamics methods that aim at reproducing a target temperature have equipartition of kinetic energy as a premise.
+While this intends to mimic the real world, discretization errors, which are intrinsic to numerical integrators, may distort such assumption.
+Fortunately, backward error analysis is a powerful tool that allows us to identify the quantity actually subject to equipartition, which is a shadow Hamiltonian that only coincides with the specified Hamiltonian when the time-step size approaches zero.
+In this paper,  deals with discretization effects in a straightforward way: with a minimal computational overhead, we obtain a refined kinetic energy and a refined Hamiltonian as suitable estimators of their
+shadow counterparts and then we tune the thermostatting procedure with this refined kinetic energy.
+The case of rigid water has served for an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
+By applying our new method to the Nos\'{e}-Hoover chain thermostat, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 5 fs.
 \end{abstract}
@@
 \appendix
-%\numberwithin{equation}{section}
 
@@
 
-%\suppinfo
+\bibliography{rigid_bodies}
 
-\bibliography{rigid_bodies}
 \end{document}

### 2018-06-14 rigid_bodies.tex

@@
 A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
-This brings into question the NVT-MD integrator developed by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
+This brings into question the NVT-MD integrator developed by Kamberaj \textit{et al}.
+\cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
 In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
-In fact, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
+In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
 Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
@@
 Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
-As \citeauthor{Eastwood_2010} \cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
+As Eastwood \textit{et al}.
+\cite{Eastwood_2010} have shown, Tolman's generalized equipartition theorem \cite{Tolman_1918} holds at least for small steps, when such a \textit{shadow kinetic energy} keeps displaying a quadratic dependency on the momenta.
 The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
@@
 It should be emphasized that this approach is not limited to deterministic thermostats.
-It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of \citeauthor{Bussi_2007} \cite{Bussi_2007} and \citeauthor{Leimkuhler_2009} \cite{Leimkuhler_2009}, for instance.
+It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}.
+\cite{Bussi_2007} and Leimkuhler \textit{et al}.
+\cite{Leimkuhler_2009}, for instance.
 
@@
 Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
-In fact, the approach of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
+In fact, the approach of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
 The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
-In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and a reimplementation of the single-chain method by \citeauthor{Martyna_1996} \cite{Martyna_1996} for rigid bodies.
+In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by Kamberaj \textit{et al}.
+\cite{Kamberaj_2005} and a reimplementation of the single-chain method by Martyna \textit{et al}.
+\cite{Martyna_1996} for rigid bodies.
 
@@
 Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
-In the general case, by taking all these facts into account and applying the analysis of \citeauthor{Tuckerman_2001} \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
+In the general case, by taking all these facts into account and applying the analysis of Tuckerman \textit{et al}.
+\cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
 In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.
@@
 Eq.~\eqref{eq:ODE system for NVT} defines an invariant measure in the extended phase-space \cite{Tuckerman_1999}, and it is possible to devise numerical solvers that preserve such measure \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
-We have done it by adapting the integrator introduced by \citeauthor{Martyna_1996} \cite{Martyna_1996} in light of a criterion explained by \citeauthor{Ezra_2006} \cite{Ezra_2006}.
+We have done it by adapting the integrator introduced by Martyna \textit{et al}. \cite{Martyna_1996} in light of a criterion explained by Ezra \cite{Ezra_2006}.
 It is worth noting, however, that such adaptation required us to alter the integration of the $\eta$ coordinates only.
@@
 A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
-\citeauthor{Omelyan_2008}'s processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
+Omelyan's processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
 Here we choose to take a simpler path, though.
@@
 In contrast, although it is possible to compute $\refined K$ by direct evaluation of Eq.~\eqref{eq:modified kinetic energy}, this is a complex and expensive task.
-Fortunately, we can estimate it rather easily by following \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
+Fortunately, we can estimate it rather easily by following Eastwood \textit{et al}.
+\cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
 Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
@@
 where subscript $\timestep$ points out the dependency of this average on the employed time step size.
-The expression above extends the main result of \citeauthor{Eastwood_2010} \cite{Eastwood_2010} to systems containing rigid bodies.
+The expression above extends the main result of Eastwood \textit{et al}. \cite{Eastwood_2010} to systems containing rigid bodies.
 It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
@@
 
-In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} with our described adaptations.
+In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005} and the method of Martyna \textit{et al}.
+\cite{Martyna_1996} with our described adaptations.
 
@@
 This corresponds to the angular coefficient of such quantity expressed as a function of time and is determined by least-square regression.
-Recall that the symmetric scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
+Recall that the symmetric scheme of Martyna \textit{et al}. \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
 In the case of our proposed method, which we refer to as Refined NHC, the conserved extended energy is $\refined H$, such as defined by Eq.~\eqref{eq:this extended energy}.
@@
 	\includegraphics{Figures/energy_drift.eps}
-	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, down-pointing triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up-pointing triangles) for the method due to \citeauthor{Martyna_1996} \cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
+	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, down-pointing triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up-pointing triangles) for the method due to Martyna \textit{et al}.
+\cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}
@@
 Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time steps, which shows that the thermostat stabilizes the numerical integration of the physical variables.
-This has already been noticed by \citeauthor{Davidchack_2010} \cite{Davidchack_2010} and also by \citeauthor{Bond_2007} \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
+This has already been noticed by Davidchack \cite{Davidchack_2010} and also by Bond and Leimkhuler \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
 
-To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} is employed.
+To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} is employed.
 Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each considered time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
 These results were obtained with our own simulation code and double-checked using LAMMPS \cite{Plimpton_1995} with a custom pair style.
-As already mentioned, the scheme due to \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} couples separate NHC thermostats to the rotational and translational degrees of freedom.
-Considering a different factorization scheme, \citeauthor{Davidchack_2010} \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
-%Nevertheless, this finding did not result in a revision of the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in the literature.
-This is confirmed for the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} in Fig.~\ref{fig:num_stab}(b), where we depict separately the contribution of each thermostat to the extended energy $H$.
+As already mentioned, the scheme due to Kamberaj \textit{et al}. \cite{Kamberaj_2005} couples separate NHC thermostats to the rotational and translational degrees of freedom.
+Considering a different factorization scheme, Davidchack \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
+%Nevertheless, this finding did not result in a revision of the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} in the literature.
+This is confirmed for the method of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005} in Fig.~\ref{fig:num_stab}(b), where we depict separately the contribution of each thermostat to the extended energy $H$.
 In this figure, the negative energy values always correspond to the thermostat coupled to the translational degrees of freedom.
@@
 	\includegraphics{Figures/numerical_stability.eps}
-	\caption{Long-term drift rate in (a) the extended energy and (b) each thermostat energy for the NVT method introduced by \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
+	\caption{Long-term drift rate in (a) the extended energy and (b) each thermostat energy for the NVT method introduced by Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}.
 In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom.}
@@
 
-We have included the results of applying our method, but using the approximate solution of rotation due to \citeauthor{Miller_2002} \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
+We have included the results of applying our method, but using the approximate solution of rotation due to Miller \textit{et al}. \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
 Recall that the computed averages for both refined schemes are obtained via reweighting by employing Eq.~\eqref{eq:reweighting}.
 
-Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}, because that is not the only problematic issue.
+Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}, because that is not the only problematic issue.
 Particularly, from Fig.~\ref{fig:properties}(a) we note that the simulated temperature decreases systematically with $\timestep$.
 This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
-As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}.
+As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al}.
+\cite{Martyna_1996}.
 Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
 As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
-Thus, the results for the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
+Thus, the results for the method of Martyna \textit{et al}.
+\cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
 
@@
 In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water.
-We see that the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
-The results of Fig.~\ref{fig:energy_partition} (a), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
+We see that the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
+The results of Fig.~\ref{fig:energy_partition} (a), which were obtained for the numerical scheme of Martyna \textit{et al}.
+\cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
 Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
 From Fig.~\ref{fig:energy_partition} (d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to 5 fs.
-For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects the serious disruption observed in the schemes of \citeauthor{Martyna_1996} \cite{Martyna_1996} and  \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}.
+For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects the serious disruption observed in the schemes of Martyna \textit{et al}.
+\cite{Martyna_1996} and  Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}.
 

### 2018-06-14 rigid_bodies.tex
Changes in the bibtex file

@@
 This makes it feasible to increase the time step $\timestep$ without ruining the accuracy of the NHC part.
-In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
+In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
 A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
@@
 \label{eq:reweighting}
-%\langle A \rangle_0 = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\refined{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] },
 \langle A \rangle_0 = \frac{\left\langle A \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}{\left\langle \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}.

### 2018-06-16 rigid_bodies.tex
Some small changes

@@
 	jcp,
-%	preprint,
-	reprint,
+	preprint,
+%	reprint,
 ]{revtex4-1}
@@
 
-\subsection{NVT Dynamics with a Single Thermostat Chain: Notation and Measure-Preserving Integration}
+\subsection{NVT Dynamics with a Single Nos\'{e}-Hoover Chain: Notation and Measure-Preserving Integration}
 \label{sec:nvt}
@@
 Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
-As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of 2 fs.
+As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of $2~\text{fs}$.
 Thus, the results for the method of Martyna \textit{et al}.
@@
 From Fig.~\ref{fig:properties} it becomes clear that the refined schemes yield significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
-Note that a small, still clearly influencing systematic error is present, as the computed averages are slightly modified for $\timestep$ > 5.
+Note that a small, still clearly influencing systematic error is present, as the computed averages are slightly modified for $\timestep > 5~\text{fs}$.
 This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
@@
 $\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\eqref{eq:reweighting}.
-As expected, the free energy is only negligible for $h$ = 1 fs.
+As expected, the free energy is only negligible for $h = 1~\text{fs}$.
 
@@
 In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition.
-In Figs.~\ref{fig:energy_partition} (a)-(c) these were calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
-For Fig.~\ref{fig:energy_partition} (d) we substituted $\langle K \rangle$ for $\langle \refined{K} \rangle$ in the right-hand side of the equations above.
+In Figs.~\ref{fig:energy_partition}(a) to \ref{fig:energy_partition}(c) these were calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
+For Fig.~\ref{fig:energy_partition}(d) we substituted $\langle K \rangle$ for $\langle \refined{K} \rangle$ in the right-hand side of the equations above.
 
-In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water.
+In our previous work \cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water.
 We see that the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
-The results of Fig.~\ref{fig:energy_partition} (a), which were obtained for the numerical scheme of Martyna \textit{et al}.
-\cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
+The results of Fig.~\ref{fig:energy_partition}(a), which were obtained for the numerical scheme of Martyna \textit{et al}.
+\cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition}(b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
 Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
-From Fig.~\ref{fig:energy_partition} (d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to 5 fs.
-For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition} (b), it becomes clear that our method corrects the serious disruption observed in the schemes of Martyna \textit{et al}.
+From Fig.~\ref{fig:energy_partition}(d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to $5~\text{fs}$.
+For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition}(b), it becomes clear that our method corrects the serious disruption observed in the schemes of Martyna \textit{et al}.
 \cite{Martyna_1996} and  Kamberaj \textit{et al}.

### 2018-06-16 Figures/Free_Energy.py
add Bussi results

@@ -27,12 +27,12 @@ data = np.genfromtxt('S1.csv', delimiter=',', skip_header=1, names=[' ','T','f',
                      'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
                      'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
 ax.errorbar(timesteps, data['f'][0:7]/(NB*beta),yerr = data['df'][0:7]/(NB*beta),linestyle ='-', color = 'black', marker = 'o',markersize = 4, linewidth = lw)
-#ax2 = ax.twinx()
-#ax2.set_ylim(297.0,299.0)
-#yticks = [297.0,298.0,299.0]
-#ax2.set_yticks(yticks)
-#ax2.set_ylabel('$T$ (K)')
-#ax2.errorbar(timesteps, Ts, yerr=Ts_error,linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw)
+ax2 = ax.twinx()
+ax2.set_ylim(297.0,299.0)
+yticks = [297.0,298.0,299.0]
+ax2.set_yticks(yticks)
+ax2.set_ylabel('$T$ (K)')
+ax2.errorbar(timesteps, Ts, yerr=Ts_error,linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw)
 
 
 #ax.legend(loc='upper center', bbox_to_anchor=(0.5,0.7), ncol=2, fancybox=True, frameon=False, numpoints = 1)

### 2018-06-16 Figures/thermodynamic_properties.py

@@ -29,10 +29,10 @@ lw = 0.5
 NB = 903
 kB = 0.0019872041 #Boltzmann's constant (kcal/mol/K).
 beta = 1./(kB*298.0)
-integrators = ['S1','S1mr','P1','K2']
-colors = ['black','magenta','green','blue']
-markers = ['s','^','o','v']
-labels = ['Refined/Unsplit','Refined/NO-SQUISH', r'Martyna', r'Kamberaj']
+integrators = ['S1','S1mr','B1s','P1','K2']
+colors = ['black','magenta','green','red','blue']
+markers = ['s','^','x','o','v']
+labels = ['Refined/NHC','Refined/NHC/NO-SQUISH','Refined/Bussi', r'Martyna', r'Kamberaj']
 k = 0
 for i in integrators:
     data = np.genfromtxt(i + '.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','temperatura',\
@@ -45,12 +45,12 @@ for i in integrators:
 
     k = k + 1
 
-data = np.genfromtxt('S1wr.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','temperatura',\
-                     'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
-                     'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
-ax[0].errorbar(timesteps, data['temperatura'][0:7],yerr = data['dtemperatura'][0:7],linestyle ='-', color = 'red', marker = 'x',markersize = 4, linewidth = lw, markeredgewidth=0.5,markeredgecolor='red', markerfacecolor='None', label = 'No Reweighting')
-ax[1].errorbar(timesteps, data['P'][0:7],yerr = data['dP'][0:7],linestyle ='-', color = 'red', marker = 'x',markersize = 4, linewidth = lw, markeredgewidth=0.5,markeredgecolor='red', markerfacecolor='None', label = 'No Reweighting')
-ax[3].errorbar(timesteps, data['Cv1'][0:7]/0.903,yerr = data['dCv1'][0:7]/0.903,linestyle ='-', color = 'red', marker = 'x',markersize = 4, linewidth = lw, markeredgewidth=0.5,markeredgecolor='red', markerfacecolor='None', label = 'No Reweighting' )
[…]

### 2018-06-16 rigid_bodies.tex
add Bussi results
Small changes in the figure with thermodynamics properties

@@
 It should be emphasized that this approach is not limited to deterministic thermostats.
-It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}.
+It is straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}.
 \cite{Bussi_2007} and Leimkuhler \textit{et al}.
@@
 In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing full-fledged instability from happening.
-This allows simulations to be executed with large time steps, which could in principle save computer time.
-However, as we will show now, discretization errors can become manifest much earlier than any risk of instability.
+This allows simulations to be executed with large time steps, which could in principle be used to save computer time.
+However, as we will show in this section, discretization errors can become manifest much earlier than does any risk of instability.
 Fortunately, our proposed methodology is capable of mitigating the effect of these errors on the ensemble averages computed for the simulated system.

### 2018-06-16 rigid_bodies.tex
Small changes in the figure with energy partition

@@
 
-In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid water system using several NVT-MD methods with different time step sizes.
-All simulations were carried out by specifying $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
+In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for the liquid water system using several NVT-MD methods with different time step sizes.
+All simulations were carried out by specifying $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
 the mean potential energy per molecule $\langle U/N \rangle$,

### 2018-06-16 Figures/energy_partition.py

@@ -15,6 +15,7 @@ ax[2].set_xlabel('$h$ (fs)')
 ax[3].set_xlabel('$h$ (fs)')
 
 ax[0].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
+ax[1].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
 ax[2].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
 ax[3].set_ylabel(r'{$\langle \widetilde{K} \rangle$} (kcal/mol)')
 ax[0].set_xlim(0.5,7.5)
@@ -37,8 +38,8 @@ data = np.genfromtxt('S1_KEpartition.csv', delimiter=',', skip_header=1, names=[
                      'dKEt','KEr','dKEr','Kt_eq','dKt_eq','Kr_eq','dKr_eq'])
 ax[3].errorbar(timesteps, data['KEt'][0:7],yerr = data['dKEt'][0:7],linestyle ='-', color = 'black', marker = 'v', markersize = 4, linewidth = lw, label = r'{$\langle K_t \rangle$}',capsize=2, markeredgewidth=1)
 ax[3].errorbar(timesteps, data['KEr'][0:7],yerr = data['dKEr'][0:7],linestyle ='-', color = 'black', marker = '^', markersize = 4, linewidth = lw, label = r'{$\langle K_r \rangle$}',capsize=2, markeredgewidth=1)
-ax[3].errorbar(timesteps, data['Kt_eq'][0:7],yerr =data['dKt_eq'][0:7],linestyle ='-', color = 'green', marker = 's', markersize = 4, linewidth = lw, label = r'{$\langle K_t \rangle_{eq}$}', capsize=2,  markeredgewidth=1)
-ax[3].errorbar(timesteps, data['Kr_eq'][0:7],yerr = data['dKr_eq'][0:7],linestyle ='-', color = 'red', marker = 'x', markersize = 4, linewidth = lw, label = r'{$\langle K_r \rangle_{eq}$}', capsize=2, markeredgewidth=1)
+ax[3].errorbar(timesteps, data['Kt_eq'][0:7],yerr =data['dKt_eq'][0:7],linestyle ='-', color = 'green', marker = 's', markersize = 4, linewidth = lw, label = r'{$\langle K_t \rangle_\textnormal{eq}$}', capsize=2,  markeredgewidth=1)
[…]

### 2018-06-18 rigid_bodies.tex
addition of refined stochastic sec

--- 2018-06-16 rigid_bodies.tex
@@
 Fortunately, backward error analysis is a powerful tool that allows us to identify the quantity actually subject to equipartition, which is a shadow Hamiltonian that only coincides with the specified Hamiltonian when the time-step size approaches zero.
-In this paper,  deals with discretization effects in a straightforward way: with a minimal computational overhead, we obtain a refined kinetic energy and a refined Hamiltonian as suitable estimators of their
+This paper deals with discretization effects in a straightforward way: with a minimal computational overhead, we obtain a refined kinetic energy and a refined Hamiltonian as suitable estimators of their
 shadow counterparts and then we tune the thermostatting procedure with this refined kinetic energy.
-The case of rigid water has served for an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
-By applying our new method to the Nos\'{e}-Hoover chain thermostat, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to 5 fs.
+The case of rigid water has served as an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
+By applying our new approach, we have been able to correct the discretization effects in numerically computed averages for time-step sizes up to 5 fs.
 \end{abstract}
@@
 
-It seems then reasonable to argue that a NHC thermostat should regulate the shadow kinetic energy instead of the specified one.
-However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators, whereas NHC thermostats introduce non-Hamiltonian components to the equations of motion.
+It seems then reasonable to argue that any thermostat should regulate the shadow kinetic energy instead of the specified one.
+However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators.
 In this paper, we devise a novel procedure that partially overcomes this limitation.
@@
 Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one.
-It should be emphasized that this approach is not limited to deterministic thermostats.
-It would be straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}.
+Although we will show the detailed derivation for the particular case of the Nos\'{e}-Hoover chain thermostat, it should be emphasized that our new approach is not limited to deterministic thermostats.
+It is straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}.
 \cite{Bussi_2007} and Leimkuhler \textit{et al}.
@@
 Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
-\begin{equation*}
+\begin{equation}
 \label{eq:equipartition}
 L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
-\end{equation*}
+\end{equation}
 where subscript $\timestep$ points out the dependency of this average on the employed time step size.
@@
 Higher-order derivative estimates can also be obtained in this fashion if necessary.
+
+\subsection{Refinement of Stochastic NVT Dynamics}
+\label{sec:refined_bussi}
+
+As an example of application in stochastic thermostats, we consider the  velocity-rescaling scheme introduced by Bussi \textit{et al}. \cite{Bussi_2007}.
+Its refined version is simply obtained by substituting $K$ by $\refined{K}$ in Eq.~(A7) of Ref.~\citenum{Bussi_2007}.
 

### 2018-06-18 rigid_bodies.tex

@@
 In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing full-fledged instability from happening.
-This allows simulations to be executed with large time steps, which could in principle save computer time.
-However, as we will show now, discretization errors can become manifest much earlier than any risk of instability.
+This allows simulations to be executed with large time steps, which could in principle be used to save computer time.
+However, as we will show in this section, discretization errors can become manifest much earlier than does any risk of instability.
 Fortunately, our proposed methodology is capable of mitigating the effect of these errors on the ensemble averages computed for the simulated system.
 
-In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid water system using several NVT-MD methods with different time step sizes.
-All simulations were carried out by specifying $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
+In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for the liquid water system using several NVT-MD methods with different time step sizes.
+All simulations were carried out by specifying $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
 the mean potential energy per molecule $\langle U/N \rangle$,

### 2018-06-19 rigid_bodies.tex
Inclusion of an method-introducing paragraph in the Discretization section

@@
 All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
-The horizontal, dotted lines are meant to help visualizing the departure of the computed averages from their expected values.
-They correspond to the results obtained by employing the Refined NHC method with $\timestep = 1~\text{fs}$.
+The horizontal, dotted lines are meant to help visualizing the departure of the computed ensemble averages from their expected values.
+They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods.
 Error bars were obtained using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
@@
 \end{figure*}
+
+Fig.~\ref{fig:properties} contains results obtained by using five different NVT integrators.
+First, for the sake of comparison, we test our measure-preserving extension to rigid bodies of the method of Martyna \textit{et al}.
+\cite{Martyna_1996} without doing any refinement, such as detailed in Sec.~\ref{sec:nvt}.
+Another method we use with no refinement is that of Kamberaj \textit{et al}. \cite{Kamberaj_2005}.
+We will show that, besides the spurious energy flux between the thermostats, such method exhibits some discretization-related issues....................................
 
@@
 However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement between schemes Refined/Unsplit and Refined/NO-SQUISH.
+
+SAY THAT BUSSI AND MARTYNA, BOTH WITHOUT REFINEMENT, YIELD SIMILAR RESULTS.
 

### 2018-06-20 rigid_bodies.tex
Further revision of the Discretization Effect section

@@
 This justifies its more widespread use.
-However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:modified_h} and Appendix \ref{sec:rigid body shadow hamiltonian}.
+However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.
 
@@
 \subsection{Backward Error Analysis}
-\label{sec:modified_h}
+\label{sec:backward error analysis}
 
@@
 They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods.
-Error bars were obtained using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
+Error bars represent 95\% confidence intervals obtained using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
 
@@
 Fig.~\ref{fig:properties} contains results obtained by using five different NVT integrators.
-First, for the sake of comparison, we test our measure-preserving extension to rigid bodies of the method of Martyna \textit{et al}.
-\cite{Martyna_1996} without doing any refinement, such as detailed in Sec.~\ref{sec:nvt}.
-Another method we use with no refinement is that of Kamberaj \textit{et al}. \cite{Kamberaj_2005}.
-We will show that, besides the spurious energy flux between the thermostats, such method exhibits some discretization-related issues....................................
+Our measure-preserving extension to rigid bodies of the splitting scheme developed by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
+Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
+First, the method is applied in a conventional (that is, refinement-free) way using the unsplit solution of Ref.~\citenum{Silveira_2017} for handling rotations.
+Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
+In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
+Next, we substitute the unsplit rotation by the simpler NO-SQUISH splitting of Miller \textit{et al}.
+\cite{Miller_2002}, denoting the obtained integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
+We note that our refinement procedure is rigorously sustained (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
+However, NO-SQUISH rotations have been shown to approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
+The fourth integrator we test here is a refined version of the velocity-rescaling method of Bussi \textit{et al}.
+\cite{Bussi_2007}, which differs from the other ones for involving a single stochastic (rather than deterministic) thermostat.
+This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
+Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux between thermostats described in the preceding section.
+This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.
 
-We have included the results of applying our method, but using the approximate solution of rotation due to Miller \textit{et al}. \cite{Miller_2002}, which is denoted as ``Refined/NO-SQUISH''.
-Recall that the computed averages for both refined schemes are obtained via reweighting by employing Eq.~\eqref{eq:reweighting}.
+Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated by simple averages taken from the sampled configurations in the case of a conventional method (either Martyna or Kamberaj), whereas reweighting via Eq.~\eqref{eq:reweighting} is carried out in the case of a refined one.
 
-Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al}.
+%Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al}.
 \cite{Kamberaj_2005}, because that is not the only problematic issue.

### 2018-06-20 Figures/thermodynamic_properties.py
Changes in the thermodynamic properties figure

@@ -29,19 +29,19 @@ lw = 0.5
 NB = 903
 kB = 0.0019872041 #Boltzmann's constant (kcal/mol/K).
 beta = 1./(kB*298.0)
-integrators = ['S1','S1mr','B1s','P1','K2']
-colors = ['black','magenta','limegreen','red','blue']
-markers = ['s','^','x','o','v']
-labels = ['Refined/NHC','Refined/NHC/NO-SQUISH','Refined/Bussi', r'Martyna', r'Kamberaj']
+integrators = ['P1','S1','S1mr','B1s','K2']
+colors = ['red','black','magenta','green','blue']
+markers = ['o','s','^','x','v']
+labels = [r'Martyna', 'Refined/NHC','Refined/NHC/Miller','Refined/Bussi', r'Kamberaj']
 k = 0
 for i in integrators:
     data = np.genfromtxt(i + '.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','temperatura',\
                      'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
                      'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
-    ax[0].errorbar(timesteps, data['temperatura'][0:7],yerr = data['dtemperatura'][0:7],linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2,   markeredgewidth=1)
-    ax[1].errorbar(timesteps, data['P'][0:7],yerr = data['dP'][0:7],linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2,  markeredgewidth=1)
-    ax[2].errorbar(timesteps, data['virial'][0:7],yerr = data['dvirial'][0:7],linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2,  markeredgewidth=1)
-    ax[3].errorbar(timesteps, data['Cv1'][0:7]/0.903,yerr = data['dCv1'][0:7]/0.903,linestyle ='-', color = colors[k], marker = markers[k],markersize = 4, linewidth = lw, label = labels[k],capsize=2, markeredgewidth=1)
[…]

### 2018-07-14 rigid_bodies.tex
Changes in Discretization Effects section

@@
 
-In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing full-fledged instability from happening.
+In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing full-fledged instability.
 This allows simulations to be executed with large time steps, which could in principle be used to save computer time.
-However, as we will show in this section, discretization errors can become manifest much earlier than does any risk of instability.
-Fortunately, our proposed methodology is capable of mitigating the effect of these errors on the ensemble averages computed for the simulated system.
+However, as we will show in this section, discretization errors can become manifest much earlier than any risk of instability.
+Fortunately, our proposed methodology is capable of mitigating the effect of these errors on ensemble averages computed for the simulated system.
 
-In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for the liquid water system using several NVT-MD methods with different time step sizes.
-All simulations were carried out by specifying $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
+In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods with various time-step sizes.
+All simulations were carried out for $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
 the mean potential energy per molecule $\langle U/N \rangle$,
@@
 The horizontal, dotted lines are meant to help visualizing the departure of the computed ensemble averages from their expected values.
-They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods.
-Error bars represent 95\% confidence intervals obtained using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
+They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods (details below).
+Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
 
@@
 
-Fig.~\ref{fig:properties} contains results obtained by using five different NVT integrators.
-Our measure-preserving extension to rigid bodies of the splitting scheme developed by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
+Fig.~\ref{fig:properties} contains results of five different NVT integrators.
+Our measure-preserving extension to rigid bodies of the splitting scheme by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
 Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
-First, the method is applied in a conventional (that is, refinement-free) way using the unsplit solution of Ref.~\citenum{Silveira_2017} for handling rotations.
+First, the method is applied in a conventional, refinement-free way, with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
 Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
 In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
-Next, we substitute the unsplit rotation by the simpler NO-SQUISH splitting of Miller \textit{et al}.
-\cite{Miller_2002}, denoting the obtained integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
+Next, we replace the unsplit rotation solution by the simpler NO-SQUISH splitting of Miller \textit{et al}.
+\cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
 We note that our refinement procedure is rigorously sustained (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
-However, NO-SQUISH rotations have been shown to approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
-The fourth integrator we test here is a refined version of the velocity-rescaling method of Bussi \textit{et al}.
-\cite{Bussi_2007}, which differs from the other ones for involving a single stochastic (rather than deterministic) thermostat.
+Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
+The fourth integrator we test is a refined version of the velocity-rescaling method of Bussi \textit{et al}.
+\cite{Bussi_2007}, which differs from the others for involving a single stochastic thermostat.
 This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
@@
 
-Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated by simple averages taken from the sampled configurations in the case of a conventional method (either Martyna or Kamberaj), whereas reweighting via Eq.~\eqref{eq:reweighting} is carried out in the case of a refined one.
+Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
+For the conventional ones, Martyna and Kamberaj, simple averages are taken from the sampled configurations.
+For the refined ones, reweighting is carried out via Eq.~\eqref{eq:reweighting}.
+
+Fig.~\ref{fig:properties}(a)
 

### 2018-07-20 rigid_bodies.tex
small changes

@@
 
-It seems then reasonable to argue that any thermostat should regulate the shadow kinetic energy instead of the specified one.
+It seems reasonable to argue that any thermostat should regulate the shadow kinetic energy instead of the specified one.
 However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators.
 In this paper, we devise a novel procedure that partially overcomes this limitation.
-It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis out of this NVE part only.
+It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and applying backward error analysis only to this part of the integrator.
 This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
@@
 \cite{Bussi_2007} and Leimkuhler \textit{et al}.
-\cite{Leimkuhler_2009}, for instance.
+\cite{Leimkuhler_2009}.
 
-Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors.
-Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
-In fact, the approach of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
+Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time-step sizes large enough to yield pronounced discretization errors.
 The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
@@
 \cite{Martyna_1996} for rigid bodies.
+
+It should be pointed out that rigid-body MD is relevant not only because some small molecules are usually modeled as rigid, such as the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
+In fact, the approach of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
 
@@
 
-In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
-1) derive a new exact solution for torque-free rotations and use it as part of a symplectic integrator for rigid bodies and
-2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} and unveil its equivalence to the earlier method by Dullweber and co-workers \cite{Dullweber_1997}.
+In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to derive a new exact solution for torque-free rotations and use it as part of a symplectic integrator for rigid bodies.
 
@@
 The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
-Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
+Finally, the matrix-valued functions $\mt B$ and $\vt C$, given by
 \begin{equation*}
@@
  q_3 & -q_2 &  q_1
-\end{smallarray}.
+\end{smallarray},
 \end{equation*}
-
-The factorization which played a central role in Ref.~\citenum{Silveira_2017} is ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, where ${\mt A}(\vt q)$ is the rotation matrix that connects the body-fixed and space-fixed frames.
+are related to the rotation matrix ${\mt A}(\vt q)$ through the factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ that was mentioned at the beginning of this section.
 Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
@@
 where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
-In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state of the system immediately before the propagation occurs.
+In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively; where the asterisk denotes the state of the system immediately before the propagation occurs.
 The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
@@
 With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
-If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as a driven variables \cite{Tuckerman_2001}.
+If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as driven variables \cite{Tuckerman_2001}.
 Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
@@
 We have done it by adapting the integrator introduced by Martyna \textit{et al}. \cite{Martyna_1996} in light of a criterion explained by Ezra \cite{Ezra_2006}.
-It is worth noting, however, that such adaptation required us to alter the integration of the $\eta$ coordinates only.
+It is worth noting, however, that such adaptation required us to only alter the integration of the $\eta$ coordinates.
 Since these are driven variables, with no influence on the dynamics of the system, we can conclude that the practical importance of such alteration is small.
@@
 
-The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
-It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
+The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence: it ensures
+the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
 Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
@@
 It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
-We emphasize that such temperature concerns the system that has actually been simulated, rather than the one initially specified.
+We emphasize that the temperature in Eq.~\ref{eq:equipartition} concerns the system that has actually been simulated, rather than the one initially specified.
 
@@
 
-Bringing the attention back to the NVT case, our proposal for dealing with discretization effects goes as follows.
-It consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
+Bringing the attention back to the NVT case, our proposal for dealing with discretization effects consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
 Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
@@
 \end{tikzpicture}
-\caption{Schematic representation}
+\caption{Schematic representation of the executed steps in a single time-step of the refined NVT methods.
+The solid arrow represents the NVE part of the integration, while the dotted arrows emphasize that forward and backward steps are carried out for positions and orientations with the only purpose of approximating their time-derivatives at point $t$.}
 \label{fig:virtual steps}
@@
 Higher-order derivative estimates can also be obtained in this fashion if necessary.
-
-\subsection{Refinement of Stochastic NVT Dynamics}
-\label{sec:refined_bussi}
 
@@
 
-In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing full-fledged instability.
+In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing numerical instability.
 This allows simulations to be executed with large time steps, which could in principle be used to save computer time.
@@
 All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
-The horizontal, dotted lines are meant to help visualizing the departure of the computed ensemble averages from their expected values.
-They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods (details below).
+The horizontal dotted lines are meant to help visualize the departure of the computed ensemble averages from their expected values, which correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods (details below).
 Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
@@
 	\includegraphics{Figures/thermodynamic_properties.eps}
-	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
+	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d)
+the constant-volume specific heat, as obtained for a system of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing different numerical schemes: ``Martyna'' corresponds to our implementation for rigid bodies of the Nos\'{e}-Hoover chain thermostat introduced in Ref.~\citenum{Martyna_1996} (red circles), the ``Refined/NHC'' scheme (black squares) is obtained by refining ``Martyna'', as described in Sec.~\ref{sec:refined_method}, ``Refined/NHC/Miller'' (pink up triangles) differs from this last one in that it employs the approximate solution of free rotations\cite{Miller_2002}, ``Refined/Bussi'' (green x) stands for a refined version of the stochastic method of Bussi \textit{et al}.
+\cite{Bussi_2007}  and, finally, ``Kamberaj'' (blue down triangles) refers to the double-thermostat introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}  .}
 	\label{fig:properties}
@@
 \cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
-We note that our refinement procedure is rigorously sustained (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
+We note that our refinement procedure is rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
 Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
@@
 Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
-For the conventional ones, Martyna and Kamberaj, simple averages are taken from the sampled configurations.
+For the conventional ones, ``Martyna'' and ``Kamberaj'', simple averages are taken from the sampled configurations.
 For the refined ones, reweighting is carried out via Eq.~\eqref{eq:reweighting}.
 
-Fig.~\ref{fig:properties}(a)
+In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for ``Kamberaj'' decreases appreciably with increasing $\timestep$.
+This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, reflecting that particle momenta is weakly modified by the NHC factorization scheme of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}.
+On the other hand, the stringent temperature control achieved in ``Martyna'' leads to a severely degraded accuracy in both $\langle U \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b)-(c).
+According to the results of Fig.~\ref{fig:nve}, discretization errors in the NVE integration become manifest at $2~\text{fs}$.
+Thus, the apparent success of the scheme by Martyna \textit{et al.} \cite{Martyna_1996}, in terms of the kinetic energy control for the entire range of time-step sizes considered, is wiped out by those discretization errors which, if not properly handled, yield considerably degraded configurational properties.
 
-%Despite the non-equilibrium process occurring within the thermostats, it is worth showing the results obtained with the method of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}, because that is not the only problematic issue.
-Particularly, from Fig.~\ref{fig:properties}(a) we note that the simulated temperature decreases systematically with $\timestep$.
-This behavior is observed in NVE simulations\cite{Davidchack_2010, Silveira_2017}, reflecting that the kinetic energy is weakly controlled in that factorization scheme.
-As a result, the potential energy and virial are not severely influenced by $\timestep$ as they are in the scheme of Martyna \textit{et al}.
-\cite{Martyna_1996}.
-Figs.~\ref{fig:properties}(a) and \ref{fig:properties}(b) clearly show that the temperature control described in Sec.~\ref{sec:nvt} leads to a severely degraded accuracy in $\langle U \rangle$, which systematically increases with $\timestep$.
-As demonstrated in Fig.~\ref{fig:nve}, the discretization errors are non-negligible even for a small step size of $2~\text{fs}$.
-Thus, the results for the method of Martyna \textit{et al}.
-\cite{Martyna_1996} can be interpreted as follows: if the kinetic energy is restricted to adopt a certain value, in accordance with the target temperature, the unavoidable discretization errors will accumulate in the remaining degrees of freedom.
+From Fig.~\ref{fig:properties}, it becomes clear that the refined schemes yield significantly improved results: for $\timestep < 5~\text{fs}$ we are able to simultaneously reproduce the specified temperature and the other mechanical properties considered.
+For longer time-step sizes, a small, still clearly influencing systematic error is present, as the computed averages slightly depart from the reference values.
+This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
+As stated above, the accuracy of the approximate solution is very high for the time-step sizes considered here\cite{Silveira_2017} and this explains the good agreement between the ``Refined/NHC'' and ``Refined/NHC/Miller'' results.
+These, in turn, are similar to the results obtained with the ``Refined/Bussi'' method.
+This was expected as, although not shown here, the performances of the unrefined Bussi\cite{Bussi_2007} and Martyna\cite{Martyna_1996} schemes were found to be analogous.
 
-From Fig.~\ref{fig:properties} it becomes clear that the refined schemes yield significantly improved results: we are able to simultaneously reproduce the specified temperature and the other thermodynamic properties.
-Note that a small, still clearly influencing systematic error is present, as the computed averages are slightly modified for $\timestep > 5~\text{fs}$.
-This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
-Mathematically, the derivation of the refined Hamiltonian was facilitated by the fewer number of operators associated with the unsplit solution of rotation.
-However, the accuracy of the approximate solution is very high for the time step sizes considered\cite{Silveira_2017} and this explains the good agreement between schemes Refined/Unsplit and Refined/NO-SQUISH.
+Interestingly, for $\timestep < 5~\text{fs}$ all the NVT schemes yield similar results of $c_\mini{V}$.
+Even though the discretization errors affect $\Ham{}$,  Fig.~\ref{fig:properties}(d) shows that the derivative of $\Ham{}$ with respect $T$ is being successfully predicted by all schemes, as $c_\mini{V} \equiv \partial \Ham{} / \partial T$.
+In addition, the fact that $c_\mini{V}$ does not change much in the range of estimated temperatures depicted in Fig.~\ref{fig:properties}(a)\cite{Yigzawe_2013} explains why the poor performance of the scheme by Kamberaj \textit{et al.} \cite{Kamberaj_2005} has almost no impact on the predicted $c_\mini{V}$.
 
-SAY THAT BUSSI AND MARTYNA, BOTH WITHOUT REFINEMENT, YIELD SIMILAR RESULTS.
-
-In Fig.~\ref{fig:properties}(a) we also depict the refined temperature, calculated by means of the generalized equipartition theorem, given by Eq.~\eqref{eq:equipartition}.
-Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\eqref{eq:thermostat_force} we expect the refined schemes to reproduce the target temperature.
-Indeed, for all the time-step sizes considered the agreement is excellent.
-
-With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for each time-step size the free energy difference per water molecule ($\Delta F/N$, circles) between the sampled and target states.
+With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for the ``Refined/NHC'' scheme the free energy differences per water molecule ($\Delta F/N$, circles) between the sampled and target states as a function of the time-step size.
 $\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\eqref{eq:reweighting}.
-As expected, the free energy is only negligible for $h = 1~\text{fs}$.
+It should be noted that the other refined schemes yield identical results and, as expected, the free energy is only negligible for $h = 1~\text{fs}$.
 
@@
 	\includegraphics{Figures/Free_Energy.eps}
-	\caption{Free Energy}
+	\caption{Difference in free energy per water molecule between the sampled and target states as a function of the time-step size.
+The sampled state was obtained by employing the ``Refined/NHC'' scheme, while the target state is consistent with a distribution proportional to $e^{-\beta \Ham{}}$, where $\Ham{}$ corresponds to the specified Hamiltonian.}
 	\label{fig:free_energy}
@@
 
-We end this section with a discussion of discretization effects on the kinetic energy partition.
+We end this section with a discussion of discretization effects on the kinetic energy partition, for which we present the results corresponding to the methods referred as ``Kamberaj'', ``Martyna'' and ``Refined/NCH''.
 In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition.
-In Figs.~\ref{fig:energy_partition}(a) to \ref{fig:energy_partition}(c) these were calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
-For Fig.~\ref{fig:energy_partition}(d) we substituted $\langle K \rangle$ for $\langle \refined{K} \rangle$ in the right-hand side of the equations above.
+In Figs.~\ref{fig:energy_partition}(a) to \ref{fig:energy_partition}(c), these were calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
+For Fig.~\ref{fig:energy_partition}(d), we substituted $\langle K \rangle$ for $\langle \refined{K} \rangle$ in the right-hand side of these equations.
 
 In our previous work \cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water.
-We see that the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
+From Fig.~\ref{fig:energy_partition}(b), we see that the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
 The results of Fig.~\ref{fig:energy_partition}(a), which were obtained for the numerical scheme of Martyna \textit{et al}.
@@
 From Fig.~\ref{fig:energy_partition}(d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to $5~\text{fs}$.
-For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition}(b), it becomes clear that our method corrects the serious disruption observed in the schemes of Martyna \textit{et al}.
+For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition}(c), it becomes clear that our method corrects the serious disruption observed in the schemes of Martyna \textit{et al}.
 \cite{Martyna_1996} and Kamberaj \textit{et al}.
@@
 	\includegraphics{Figures/energy_partition.eps}
-		\caption{Effect of the time-step size on ... 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
+		\caption{Influence of the time-step size on the kinetic energy partition in NVT MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules for the numerical schemes introduced by (a) Martyna \textit{et al.} \cite{Martyna_1996} and (b) Kamberaj \textit{et al.} \cite{Kamberaj_2005}, as well as for (c) our refined version of the thermostat by Martyna \textit{et al.} \cite{Martyna_1996}, for which we also consider in (d) the results obtained without applying the reweighting procedure.
+For each case, we depict the simulated translational (black down triangles) and rotational (black up triangles) kinetic energies, and show its deviation from the expected translational (green squares) and rotational (red x) values assuming equipartition.}
 	\label{fig:energy_partition}

### 2018-07-20 Figures/Free_Energy.py

@@ -27,12 +27,12 @@ data = np.genfromtxt('S1.csv', delimiter=',', skip_header=1, names=[' ','T','f',
                      'dtemperatura','P','dP','E','dE','virial','dvirial','KEtotal','dKEtotal', \
                      'KEt','dKEt','KEr','dKEr','Cv1','dCv1','dfdT','ddfdT'])
 ax.errorbar(timesteps, data['f'][0:7]/(NB*beta),yerr = data['df'][0:7]/(NB*beta),linestyle ='-', color = 'black', marker = 'o',markersize = 4, linewidth = lw)
-ax2 = ax.twinx()
-ax2.set_ylim(297.0,299.0)
-yticks = [297.0,298.0,299.0]
-ax2.set_yticks(yticks)
-ax2.set_ylabel('$T$ (K)')
-ax2.errorbar(timesteps, Ts, yerr=Ts_error,linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw)
+#ax2 = ax.twinx()
+#ax2.set_ylim(297.0,299.0)
+#yticks = [297.0,298.0,299.0]
+#ax2.set_yticks(yticks)
+#ax2.set_ylabel('$T$ (K)')
+#ax2.errorbar(timesteps, Ts, yerr=Ts_error,linestyle ='-', color = 'red', marker = 's',markersize = 4, linewidth = lw)
 
 
 #ax.legend(loc='upper center', bbox_to_anchor=(0.5,0.7), ncol=2, fancybox=True, frameon=False, numpoints = 1)

### 2018-07-20 rigid_bodies.tex
begin of concluding remarks

@@
 	jcp,
-	preprint,
-%	reprint,
+%	preprint,
+	reprint,
 ]{revtex4-1}
@@
 By virtual and incomplete we mean, respectively, that these steps are never incorporated to the trajectory and that they are aborted once the positions and orientations have been updated.
-By this means, we can obtain estimates $\dot{\vt r}^{[3]}$ and $\dot{\vt q}^{[3]}$ without any additional force evaluation, which results in a small computational overhead.
+By this means, we can obtain the estimates $\dot{\vt r}^{[3]}$ and $\dot{\vt q}^{[3]}$ without any additional force evaluation, which results in a small computational overhead.
 
@@
 \end{tikzpicture}
-\caption{Schematic representation of the executed steps in a single time-step of the refined NVT methods.
+\caption{Schematic representation of the executed steps in a single time-step of our refined NVT dynamics.
 The solid arrow represents the NVE part of the integration, while the dotted arrows emphasize that forward and backward steps are carried out for positions and orientations with the only purpose of approximating their time-derivatives at point $t$.}
@@
 Because such action leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ is done as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.
+
+As an example of application in stochastic thermostats, we consider the  velocity-rescaling scheme introduced by Bussi \textit{et al}. \cite{Bussi_2007}.
+Its refined version is simply obtained by substituting $K$ by $\refined{K}$ in Eq.~(A7) of Ref.~\citenum{Bussi_2007}.
 
@@
 Higher-order derivative estimates can also be obtained in this fashion if necessary.
-
-As an example of application in stochastic thermostats, we consider the  velocity-rescaling scheme introduced by Bussi \textit{et al}. \cite{Bussi_2007}.
-Its refined version is simply obtained by substituting $K$ by $\refined{K}$ in Eq.~(A7) of Ref.~\citenum{Bussi_2007}.
 
@@
 This is an indication that
-1) the employed four-point formulas do not provide derivatives with the required accuracy or
+1) the employed four-point formulas do not provide derivatives with the required accuracy and/or
 2) the neglected high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would contribute notably to the shadow Hamiltonian.
@@
 	\includegraphics{Figures/energy_drift.eps}
-	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, down-pointing triangles) and its refined Hamiltonian ($\refined{\Ham{}}$, squares) part for our NVT method, as well as in the extended energy ($H$, up-pointing triangles) for the method due to Martyna \textit{et al}.
-\cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
+	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, black squares) and its refined Hamiltonian ($\refined{\Ham{}}$, green up triangles) part for our NVT method, as well as in the extended energy ($H$, red x) for the method due to Martyna \textit{et al}.
+\cite{Martyna_1996} and in the refined Hamiltonian ($\refined{\Ham{}}$, blue circles) for the numerical scheme of Eq.~\eqref{eq:splitting NVE} with unsplit solution of rotation.}
 	\label{fig:energy_drift}
@@
 Fig.~\ref{fig:energy_drift} also contains drift rates computed for the refined Hamiltonian $\refined{\Ham{}}$ in an NVE simulation \cite{Silveira_2017}.
-Up to $h = 6~\text{fs}$, these rates practically coincide with those obtained for $\refined H$ with the Refined NHC scheme.
-This indicates that the major causes for the observed drift took place during evaluations of the NVE propagator.
+Up to $h = 6~\text{fs}$, these rates practically coincide with those obtained for $\refined H$ with the Refined NHC scheme, indicating that the major causes for the observed drift in this scheme took place during evaluations of the NVE propagator.
 As the interaction potential we use is very smooth at cutoff, a possible cause is the amplification of round-off errors which can occur in events of close interatomic approaches \cite{Engle_2005}.
@@
 Recall that this part of $\refined H$ represents the total energy of the physical system.
-Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time steps, which shows that the thermostat stabilizes the numerical integration of the physical variables.
+Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest time steps considered, which shows that the thermostat stabilizes the numerical integration of the physical variables.
 This has already been noticed by Davidchack \cite{Davidchack_2010} and also by Bond and Leimkhuler \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
@@
 To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} is employed.
-Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each considered time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
+Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
 These results were obtained with our own simulation code and double-checked using LAMMPS \cite{Plimpton_1995} with a custom pair style.
@@
 Considering a different factorization scheme, Davidchack \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
-%Nevertheless, this finding did not result in a revision of the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} in the literature.
 This is confirmed for the method of Kamberaj \textit{et al}.
@@
 	\includegraphics{Figures/numerical_stability.eps}
-	\caption{Long-term drift rate in (a) the extended energy and (b) each thermostat energy for the NVT method introduced by Kamberaj \textit{et al}.
+	\caption{Time evolution of (a) the extended energy and (b) of each thermostat energy for the NVT method introduced by Kamberaj \textit{et al}.
 \cite{Kamberaj_2005}.
@@
 
-With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for the ``Refined/NHC'' scheme the free energy differences per water molecule ($\Delta F/N$, circles) between the sampled and target states as a function of the time-step size.
+With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for the ``Refined/NHC'' scheme the free energy difference per water molecule ($\Delta F/N$) between the sampled and target states as a function of the time-step size.
 $\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\eqref{eq:reweighting}.
@@
 	\includegraphics{Figures/energy_partition.eps}
-		\caption{Influence of the time-step size on the kinetic energy partition in NVT MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules for the numerical schemes introduced by (a) Martyna \textit{et al.} \cite{Martyna_1996} and (b) Kamberaj \textit{et al.} \cite{Kamberaj_2005}, as well as for (c) our refined version of the thermostat by Martyna \textit{et al.} \cite{Martyna_1996}, for which we also consider in (d) the results obtained without applying the reweighting procedure.
-For each case, we depict the simulated translational (black down triangles) and rotational (black up triangles) kinetic energies, and show its deviation from the expected translational (green squares) and rotational (red x) values assuming equipartition.}
+		\caption{Influence of the time-step size on the kinetic energy partition in NVT MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules for (a) our implementation of the numerical scheme by Martyna \textit{et al.} \cite{Martyna_1996} and (b) the method due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}, as well as for (c) our refined version of the thermostat by Martyna \textit{et al.} \cite{Martyna_1996}, for which we also depict in (d) the results obtained without applying the reweighting procedure.
+For each case, we show the simulated translational (black down triangles) and rotational (black up triangles) kinetic energies, as well as the expected translational (green squares) and rotational (red x) values assuming equipartition.}
 	\label{fig:energy_partition}
@@
 
-To be included.
+In this paper, we have devised an approach that integrates the powerful machinery of backward error analysis into Nos\'{e}-Hoover chains dynamics and global stochastic thermostats.
+A valuable enhancement is achieved with a simple approximation of the shadow Hamiltonian: in the case of TIP3P water, we have been able to overcome the serious disruptions observed in the schemes of Martyna \textit{et al}.
+\cite{Martyna_1996} and Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}, correcting the discretization effects in numerically computed averages for time-step sizes up to 5 fs.
 

### 2018-07-21 rigid_bodies.tex

@@
 \usetikzlibrary{decorations.text}
-
-%\usepackage{xr}
-%\usepackage{booktabs}
-%\usepackage{threeparttable}
-%\usepackage[T1]{fontenc}
-%\usepackage[table]{xcolor}
-%\definecolor{lightgray}{gray}{0.9}
-%\usepackage[exponent-product=\times]{siunitx}
 
@@
 The case of rigid water has served as an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
-By applying our new approach, we have been able to correct the discretization effects in numerically computed averages for time-step sizes up to 5 fs.
+By applying our new approach, we have been able to correct the discretization effects in equilibrium properties for time-step sizes up to 5 fs.
 \end{abstract}
@@
 From Fig.~\ref{fig:energy_partition}(d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to $5~\text{fs}$.
-For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition}(c), it becomes clear that our method corrects the serious disruption observed in the schemes of Martyna \textit{et al}.
+For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition}(c), it becomes clear that our method corrects the artifacts observed in the schemes of Martyna \textit{et al}.
 \cite{Martyna_1996} and Kamberaj \textit{et al}.
@@
 
-In this paper, we have devised an approach that integrates the powerful machinery of backward error analysis into Nos\'{e}-Hoover chains dynamics and global stochastic thermostats.
-A valuable enhancement is achieved with a simple approximation of the shadow Hamiltonian: in the case of TIP3P water, we have been able to overcome the serious disruptions observed in the schemes of Martyna \textit{et al}.
+In this paper, we have devised a general approach that integrates backward error analysis into the Nos\'{e}-Hoover chains and global stochastic thermostats.
+The key rationale behind our proposal came from recognizing that a Shadow kinetic energy should substitute the specified one in the equations dictating the thermostat action.
+This turns out to be crucial for simulations that display considerable discretization errors, as a ``clear definition of simulation temperature''\cite{Eastwood_2010} is obtained by applying Tolman's generalized equipartition theorem to a Shadow kinetic energy.
+
+A significant enhancement was achieved for a system of TIP3P rigid water molecules\cite{Jorgensen_1983} by using an analytically approximation to the Shadow Hamiltonian (here refered to as \textit{refined Hamiltonian}) with a minimal computatioal overhead, and whose derivation was greatly facilitated by the unsplit solution of rotation introduced in our previous paper\cite{Silveira_2017}.
+Particularly, we have been able to remove the serious artifacts observed in the schemes of Martyna \textit{et al}.
 \cite{Martyna_1996} and Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}, correcting the discretization effects in numerically computed averages for time-step sizes up to 5 fs.
+\cite{Kamberaj_2005}, correcting the discretization effects in numerically computed averages for time-step sizes up to $5~\text{fs}$.
+From our rigorous analysis of these schemes, we would not recommend a time-step size larger than $1~\text{fs}$.
+The fact of having overcome that barrier is of paramount importance as allows us to actually exploit the high numerical stability provided by rigid-body MD.
+
+Finally, it should be remarked that our method is not limited to rigid-body MD.
+It could be employed, for instance, in conjuction with other constrained MD schemes, such as SHAKE\cite{Ryckaert_1977} and RATTLE\cite{Andersen_1983}.
+Of course, restrictions arising from these methods must be properly taken into account, as we did for the quaternion time-derivative.
 

### 2018-08-02 rigid_bodies.tex
Revision of the abstract

@@
 \begin{abstract}
-Many deterministic and stochastic molecular dynamics methods that aim at reproducing a target temperature have equipartition of kinetic energy as a premise.
-While this intends to mimic the real world, discretization errors, which are intrinsic to numerical integrators, may distort such assumption.
-Fortunately, backward error analysis is a powerful tool that allows us to identify the quantity actually subject to equipartition, which is a shadow Hamiltonian that only coincides with the specified Hamiltonian when the time-step size approaches zero.
-This paper deals with discretization effects in a straightforward way: with a minimal computational overhead, we obtain a refined kinetic energy and a refined Hamiltonian as suitable estimators of their
-shadow counterparts and then we tune the thermostatting procedure with this refined kinetic energy.
-The case of rigid water has served as an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
-By applying our new approach, we have been able to correct the discretization effects in equilibrium properties for time-step sizes up to 5 fs.
+Kinetic energy equipartition is a premise for many deterministic and stochastic molecular dynamics methods that aim at sampling a canonical ensemble.
+While this is expected for real systems, discretization errors introduced by the numerical integration may distort such assumption.
+Fortunately, backward error analysis allows us to identify the quantity that is actually subject to equipartition.
+This is related to a shadow Hamiltonian, which coincides with the specified Hamiltonian only when the time-step size approaches zero.
+This paper deals with discretization effects in a straightforward way.
+With a small computational overhead, we obtain refined versions of the kinetic and potential energies, whose sum is a suitable estimator of the shadow Hamiltonian.
+Then, we tune the thermostatting procedure by employing the refined kinetic energy instead of the conventional one.
+This procedure is shown to reproduce a canonical ensemble compatible with the refined system, but canonical averages regarding the original system can be easily recovered by reweighting.
+Water modeled as a rigid body is an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
+By applying our new approach, we were able to mitigate discretization effects in equilibrium properties of liquid water for time-step sizes up to 5 fs.
 \end{abstract}

### 2018-09-17 rigid_bodies.tex
Some small changes

@@
 This paper proceeds as follows.
-In Secs.~\ref{sec:nve}-\ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
-This is followed by our developments on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive our new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
-Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}. Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the shadow Hamiltonian approximation.
+In Secs.~\ref{sec:nve} and \ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
+This is followed by our developments on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive a new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
+Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.
+Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the shadow Hamiltonian approximation for a system containing rigid bodies.
 
@@
 As an example of application in stochastic thermostats, we consider the  velocity-rescaling scheme introduced by Bussi \textit{et al}. \cite{Bussi_2007}.
-Its refined version is simply obtained by substituting $K$ by $\refined{K}$ in Eq.~(A7) of Ref.~\citenum{Bussi_2007}.
+Its refined version can be simply obtained by substituting $K$ by $\refined{K}$ in Eq.~(A7) of Ref.~\citenum{Bussi_2007}.
+In the implementation used to obtain some of the results in Sec.~\ref{sec:discretization effects}, we also split the stochastic propagator in order to have a Trotter-Suzuki type of integration (whereas a leap-frog type is used originally).
 
@@
 \subsection{Correction of Discretization Effects on Thermodynamic Properties}
+\label{sec:discretization effects}
 
@@
 In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods with various time-step sizes.
-All simulations were carried out for $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
+All simulations were carried out for $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
 the mean potential energy per molecule $\langle U/N \rangle$,
@@
 All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
-The horizontal dotted lines are meant to help visualize the departure of the computed ensemble averages from their expected values, which correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods (details below).
+The horizontal dotted lines are meant to help visualize the departure of the computed properties from their expected values, which correspond to average results obtained with $\timestep = 1~\text{fs}$ for the three refined methods (details below).
 Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
@@
 First, the method is applied in a conventional, refinement-free way, with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
-Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
+Then, it is modified by introducing the refinement/reweighting procedure proposed here in Sec.~\ref{sec:refined_method}.
 In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
@@
 \cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
-We note that our refinement procedure is rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
+We note that our refinement procedure is only rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations.
 Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
 The fourth integrator we test is a refined version of the velocity-rescaling method of Bussi \textit{et al}.
-\cite{Bussi_2007}, which differs from the others for involving a single stochastic thermostat.
+\cite{Bussi_2007}, which differs from the others for involving a global stochastic thermostat.
 This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
 Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux between thermostats described in the preceding section.
+\cite{Kamberaj_2005} for comparison purposes, despite its spurious energy flux described in the preceding section.
 This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.
+We stress that thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
+For the former, simple averages are taken from the sampled configurations.
+For the latter, reweighting is carried out via Eq.~\eqref{eq:reweighting}.
 
-Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
-For the conventional ones, ``Martyna'' and ``Kamberaj'', simple averages are taken from the sampled configurations.
-For the refined ones, reweighting is carried out via Eq.~\eqref{eq:reweighting}.
-
-In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for ``Kamberaj'' decreases appreciably with increasing $\timestep$.
-This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, reflecting that particle momenta is weakly modified by the NHC factorization scheme of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}.
-On the other hand, the stringent temperature control achieved in ``Martyna'' leads to a severely degraded accuracy in both $\langle U \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b)-(c).
+In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for the thermostat identified as ``Kamberaj'' decreases appreciably as the time step size increases.
+This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, indicating that the effect of the NHC factorization scheme of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005} on the particle momenta is weak.
+On the other hand, the stringent temperature control achieved by the thermostat named ``Martyna'' leads to a severely degraded accuracy in both $\langle U/N \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b) and \ref{fig:properties}(c).
 According to the results of Fig.~\ref{fig:nve}, discretization errors in the NVE integration become manifest at $2~\text{fs}$.

### 2018-09-18 rigid_bodies.tex
Text review: introduction and results

@@
 Then, we tune the thermostatting procedure by employing the refined kinetic energy instead of the conventional one.
-This procedure is shown to reproduce a canonical ensemble compatible with the refined system, but canonical averages regarding the original system can be easily recovered by reweighting.
-Water modeled as a rigid body is an excellent test case for our proposal, because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
+This procedure is shown to reproduce a canonical ensemble compatible with the refined system, as opposed to the original one, but canonical averages regarding the latter can be easily recovered by reweighting.
+Water, modeled as a rigid body, is an excellent test case for our proposal because its numerical stability extends up to time steps large enough to yield pronounced discretization errors in Verlet-type integrators.
 By applying our new approach, we were able to mitigate discretization effects in equilibrium properties of liquid water for time-step sizes up to 5 fs.
@@
 Molecular dynamics (MD) simulation involves several types of approximations in the quest for mimicking real systems.
-Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the force field.
+Naively, we tend to rank simulation results by their proximity to experimental data and attribute any disagreement to a lack of accuracy of the employed force field.
 However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
-For instance, it has been shown \cite{Eastwood_2010} that errors due to the discretization of the equations of motion appear to disrupt the equipartition of kinetic energy among molecules of distinct sizes.
-A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom of rigid bodies \cite{Davidchack_2010, Silveira_2017}.
+For instance, it has been shown \cite{Eastwood_2010} that errors due to the discretization of the equations of motion can disrupt the equipartition of kinetic energy among molecules of distinct sizes.
+A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body systems \cite{Davidchack_2010, Silveira_2017}.
 This brings into question the NVT-MD integrator developed by Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} in which, assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
-In addition, as we will show later on, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
+\cite{Kamberaj_2005} in which, by assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
+In addition, as we will show here, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
 In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
 Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
-This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods routinely used in MD are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.
+This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.
 
@@
 The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
-Therefore, what seems to occur in fact is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
+Therefore, what seems to occur, in fact, is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
 As Eastwood \textit{et al}.
@@
 
-It seems reasonable to argue that any thermostat should regulate the shadow kinetic energy instead of the specified one.
-However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators.
-In this paper, we devise a novel procedure that partially overcomes this limitation.
+It seems reasonable to argue that a thermostat should regulate the shadow kinetic energy instead of the specified one.
+However, the theory underlying the concept and computation of shadow Hamiltonians is only applicable to symplectic equations and symplectic integrators, thus ruling out most NVT-MD methods in use today.
+In this paper, we devise a novel procedure which partially overcomes this limitation.
 It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and applying backward error analysis only to this part of the integrator.
 This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
-By recasting the thermostat equations with this refined kinetic energy, we are able to predict its equipartition and to reproduce a canonical ensemble consistent with the refined Hamiltonian.
-Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one.
-Although we will show the detailed derivation for the particular case of the Nos\'{e}-Hoover chain thermostat, it should be emphasized that our new approach is not limited to deterministic thermostats.
-It is straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}.
-\cite{Bussi_2007} and Leimkuhler \textit{et al}.
+By recasting the thermostat equations with such a refined kinetic energy, we are able to predict its equipartition and reproduce a canonical ensemble consistent with the refined Hamiltonian.
+Finally, a reweighting procedure yields ensemble averages which are, instead, consistent with the specified one.
+Although we restrict our detailed derivation to the case of a Nos\'{e}-Hoover chain thermostat, our new approach is applicable to other deterministic and stochastic thermostats.
+It is straightforward to use it, for instance, in conjunction with global thermostats such as the Stochastic Velocity Rescaling method of Bussi \textit{et al}.
+\cite{Bussi_2007} and the Nos\'{e}-Hoover-Langevin thermostat of Leimkuhler \textit{et al}.
 \cite{Leimkuhler_2009}.
 
-Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time-step sizes large enough to yield pronounced discretization errors.
-The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
-In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} and a reimplementation of the single-chain method by Martyna \textit{et al}.
-\cite{Martyna_1996} for rigid bodies.
-
-It should be pointed out that rigid-body MD is relevant not only because some small molecules are usually modeled as rigid, such as the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
+Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability can extend up to time-step sizes large enough to yield pronounced discretization errors.
+It should be pointed out that rigid-body MD is relevant not only because some small molecules are usually modeled as rigid structures, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be envisioned as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
 In fact, the approach of Kamberaj \textit{et al}.
 \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
+In the case of rigid water, which served as an excellent test case for our proposal, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the double-chain method of Kamberaj \textit{et al}.
+\cite{Kamberaj_2005}, but also a single-chain reimplementation of the method by Martyna \textit{et al}.
+\cite{Martyna_1996} and the stochastic rescaling method \cite{Bussi_2007} applied simultaneously to translational and angular velocities.
 
@@
 The damping of Coulombic interactions was done by a factor $\text{erfc}(-\alpha r)$, with $\alpha = 0.29~\text{\AA}^{-1}$.
-In addition, all thermostated simulations were carried out considering a time-scale constant $t_d = 100~\text{fs}$.
+In addition, all thermostated simulations were carried out considering a time-scale constant $t_d = 100~\text{fs}$ (or $\tau = 100~\text{fs}$, in the nomenclature of Ref.~\citenum{Bussi_2007}).
 
@@
 
-We start by examining the departure of the refined Hamiltonian from the specified one in NVE simulations.
-This is the most direct way of determining the magnitude of the discretization errors \cite{Engle_2005}, which is essential for discussing the NVT case.
+We start by examining how much the refined Hamiltonian departs from the specified one in NVE simulations.
+This is a direct way of determining the magnitude of the discretization errors \cite{Engle_2005} and is essential for discussing the NVT case.
 In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system using different time-step sizes.
@@
 However, this can only be achieved by limiting the order of approximation of numerical derivatives and sacrificing short-term reversibility.
-This poses the question whether these conditions will influence the long-term stability of the method, especially if one tries to make use of large time steps.
+This poses the question whether these conditions will influence the long-term stability of the method or not, especially if one tries to make use of large time steps.
 
-Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in an equilibrated MD simulation.
-This corresponds to the angular coefficient of such quantity expressed as a function of time and is determined by least-square regression.
+Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in a supposedly equilibrated MD simulation.
+It is defined as the angular coefficient of such quantity expressed as a function of time, determined by a least-square linear regression.
 Recall that the symmetric scheme of Martyna \textit{et al}. \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
@@
 Fig.~\ref{fig:energy_drift} also contains drift rates computed for the refined Hamiltonian $\refined{\Ham{}}$ in an NVE simulation \cite{Silveira_2017}.
-Up to $h = 6~\text{fs}$, these rates practically coincide with those obtained for $\refined H$ with the Refined NHC scheme, indicating that the major causes for the observed drift in this scheme took place during evaluations of the NVE propagator.
+Up to $h = 6~\text{fs}$, these rates practically coincide with those obtained for $\refined H$ with the Refined NHC scheme, indicating that the major causes for the observed drift takes place during evaluations of the NVE propagator.
 As the interaction potential we use is very smooth at cutoff, a possible cause is the amplification of round-off errors which can occur in events of close interatomic approaches \cite{Engle_2005}.
 For time steps larger than $6~\text{fs}$, the NVE integration becomes fully unstable.
-In contrast, each NVT integrator manages to remain numerically stable at least up to $h = 10~\text{fs}$, but at the cost of displaying an exceedingly large drift in its corresponding extended energy, as one can see the inset of Fig.~\ref{fig:energy_drift}.
-Trying to look for the cause of this behavior, we also present drift rates computed for the Hamiltonian $\refined{\Ham{}}$ in the simulations carried out with the Refined NHC method.
-Recall that this part of $\refined H$ represents the total energy of the physical system.
-Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest time steps considered, which shows that the thermostat stabilizes the numerical integration of the physical variables.
+In contrast, all NVT integrators manage to remain numerically stable at least up to $h = 10~\text{fs}$, but at the cost of displaying exceedingly large drifts in their corresponding extended energies, as one can see the inset of Fig.~\ref{fig:energy_drift}.
+Trying to look for the cause of this behavior, we also present drift rates computed for the refined Hamiltonian $\refined{\Ham{}}$ in the simulations carried out with the Refined NHC method.
+Recall that this part of $\refined H$ represents the total energy of the system.
+Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time steps, showing that the thermostat stabilizes the numerical integration of the physical variables.
 This has already been noticed by Davidchack \cite{Davidchack_2010} and also by Bond and Leimkhuler \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
@@
 Therefore, energy flows from this thermostat to the one coupled to the rotational degrees.
-Interestingly, Fig.~\ref{fig:num_stab} shows that the rate at which the thermostats approach the steady state decreases with $\timestep$. As far as we know, this behavior has not been reported before.
+Interestingly, Fig.~\ref{fig:num_stab} shows that the rate at which the thermostats approach steady state decreases with $\timestep$. As far as we know, this behavior has not been reported before.
 
@@
 
-\subsection{Correction of Discretization Effects on Thermodynamic Properties}
+\subsection{Correction of Discretization Effects on Computed Thermodynamic Properties}
 \label{sec:discretization effects}
 
-In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing numerical instability.
+In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing numerical instability from happening.
 This allows simulations to be executed with large time steps, which could in principle be used to save computer time.
-However, as we will show in this section, discretization errors can become manifest much earlier than any risk of instability.
+However, as we will show in this section, discretization errors can become manifest before any risk of instability arises.
 Fortunately, our proposed methodology is capable of mitigating the effect of these errors on ensemble averages computed for the simulated system.
 
-In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods with various time-step sizes.
+In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods and time-step sizes.
 All simulations were carried out for $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
@@
 The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
-All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
-The horizontal dotted lines are meant to help visualize the departure of the computed properties from their expected values, which correspond to average results obtained with $\timestep = 1~\text{fs}$ for the three refined methods (details below).
-Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
+All graphical symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
+The horizontal dotted lines are meant to help us visualize the departure of the computed properties from their expected values, which are estimated by averaging the results obtained from the three refined methods (see details below) with $\timestep = 1~\text{fs}$.
+Error bars represent 95\% confidence intervals estimated via overlapping batch means \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.
 
@@
 	\includegraphics{Figures/thermodynamic_properties.eps}
-	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d)
-the constant-volume specific heat, as obtained for a system of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing different numerical schemes: ``Martyna'' corresponds to our implementation for rigid bodies of the Nos\'{e}-Hoover chain thermostat introduced in Ref.~\citenum{Martyna_1996} (red circles), the ``Refined/NHC'' scheme (black squares) is obtained by refining ``Martyna'', as described in Sec.~\ref{sec:refined_method}, ``Refined/NHC/Miller'' (pink up triangles) differs from this last one in that it employs the approximate solution of free rotations\cite{Miller_2002}, ``Refined/Bussi'' (green x) stands for a refined version of the stochastic method of Bussi \textit{et al}.
-\cite{Bussi_2007}  and, finally, ``Kamberaj'' (blue down triangles) refers to the double-thermostat introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}  .}
+	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d) the constant-volume specific heat of water, as obtained from NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules at $T=298~\mathrm{K}$ and $\rho=970~\mathrm{kg/m^3}$ and employing different numerical integration schemes.
+	The ``Martyna'' scheme ($\bullet$) corresponds to our rigid-body implementation of the Nos\'{e}-Hoover chain integrator introduced in Ref.~\citenum{Martyna_1996}, with analytically computed free rotations \cite{Silveira_2017}.
+	The ``Refined/NHC'' scheme ($\blacksquare$) is obtained by refining the ``Martyna'' method such as described in Sec.~\ref{sec:refined_method}.
+	The ``Refined/NHC/Miller'' scheme ($\blacktriangle$) differs from the preceding one only in that an approximate solution is used for free rotations \cite{Miller_2002}.
+	The ``Refined/Bussi'' scheme ($\times$) is a refined version of the stochastic method of Bussi \textit{et al}. \cite{Bussi_2007}.
+	Finally, the ``Kamberaj'' scheme ($\blacktriangledown$) consists in the double-thermostat method introduced in Ref.~\citenum{Kamberaj_2005}.}
 	\label{fig:properties}

### 2018-09-19 rigid_bodies.tex
Reformulation of introduction

@@
 A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body systems \cite{Davidchack_2010, Silveira_2017}.
-This brings into question the NVT-MD integrator developed by Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} in which, by assuming translational-rotational equipartition, separate chains of Nos\'{e}-Hoover thermostats are coupled to the two classes of degrees of freedom.
-In addition, as we will show here, this method fails to reproduce the specified temperature even for a time-step size $\timestep$ typically employed in MD.
+This issue is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
+This brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
+As we will show here, the method fails to reproduce the specified temperature even for time-step sizes typically employed in MD.
 In fact, Davidchack \cite{Davidchack_2010} recommends that the use of separate thermostats for translation and rotation should always be avoided.
-Nevertheless, issues arise even for freely-moving particles or when translation and rotation are coupled to a single thermostat.
-This occurs because Nos\'{e}-Hoover chains (NHC) \cite{Martyna_1992} and other thermostat methods are based on the relation between temperature and kinetic energy that is a corollary of the equipartition theorem.
+Nonetheless, the cited double-thermostat method is available in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995} and has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}, in spite of all the issues observed here.
+
+Debenedetti....
 
 For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
-The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian when $\timestep \to 0$.
+The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian as the time-step size approaches zero.
 Therefore, what seems to occur, in fact, is a failure to identify the quantity that is subject to equipartition in place of the specified kinetic energy.
@@
 
-It seems reasonable to argue that a thermostat should regulate the shadow kinetic energy instead of the specified one.
+Issues arise in NVT-MD because thermostatting methods usually rely on the corollary of the equipartition theorem which relates temperature to kinetic energy.
+Thus, it seems reasonable to argue that a thermostat should regulate the shadow kinetic energy instead of the specified one.
 However, the theory underlying the concept and computation of shadow Hamiltonians is only applicable to symplectic equations and symplectic integrators, thus ruling out most NVT-MD methods in use today.
-In this paper, we devise a novel procedure which partially overcomes this limitation.
-It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and applying backward error analysis only to this part of the integrator.
-This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
+In this paper, we devise a novel procedure that is able to overcome this limitation in the case of global thermostatting methods such as, for instance, the Nos\'{e}-Hoover Chain (NHC) \cite{Martyna_1992}, the Stochastic Velocity Rescaling (SVR) \cite{Bussi_2007}, and the Nos\'{e}-Hoover-Langevin (NHL) \cite{Leimkuhler_2009} methods.
+It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and then applying backward error analysis to that part only.
+This allows one to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
 By recasting the thermostat equations with such a refined kinetic energy, we are able to predict its equipartition and reproduce a canonical ensemble consistent with the refined Hamiltonian.
 Finally, a reweighting procedure yields ensemble averages which are, instead, consistent with the specified one.
-Although we restrict our detailed derivation to the case of a Nos\'{e}-Hoover chain thermostat, our new approach is applicable to other deterministic and stochastic thermostats.
-It is straightforward to use it, for instance, in conjunction with global thermostats such as the Stochastic Velocity Rescaling method of Bussi \textit{et al}.
-\cite{Bussi_2007} and the Nos\'{e}-Hoover-Langevin thermostat of Leimkuhler \textit{et al}.
-\cite{Leimkuhler_2009}.
+For simplicity, we restrict our detailed derivation to the case of a NHC thermostat \cite{Martyna_1992}.
 
-Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability can extend up to time-step sizes large enough to yield pronounced discretization errors.
-It should be pointed out that rigid-body MD is relevant not only because some small molecules are usually modeled as rigid structures, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be envisioned as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
-In fact, the approach of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
-In the case of rigid water, which served as an excellent test case for our proposal, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the double-chain method of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}, but also a single-chain reimplementation of the method by Martyna \textit{et al}.
-\cite{Martyna_1996} and the stochastic rescaling method \cite{Bussi_2007} applied simultaneously to translational and angular velocities.
+The new method is specially appealing for rigid-body MD, primarily because numerical stability can, in this case, extend up to time-step sizes large enough to yield pronounced discretization errors.
+For rigid water, which served as an excellent test case, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the double-thermostat method of Kamberaj, Low, and Neal (KLN) \cite{Kamberaj_2005}, but also a reimplementation for rigid bodies of the NHC integrator developed by Martyna \textit{et al}.
+\cite{Martyna_1996}, as well as the SVR method \cite{Bussi_2007} employed to simultaneously rescale translational and angular velocities.
 
@@
 The damping of Coulombic interactions was done by a factor $\text{erfc}(-\alpha r)$, with $\alpha = 0.29~\text{\AA}^{-1}$.
-In addition, all thermostated simulations were carried out considering a time-scale constant $t_d = 100~\text{fs}$ (or $\tau = 100~\text{fs}$, in the nomenclature of Ref.~\citenum{Bussi_2007}).
+In addition, all thermostated simulations were carried out considering a time-scale constant $t_d = 100~\text{fs}$ (or $\tau = 100~\text{fs}$, in the terminology of Ref.~\citenum{Bussi_2007}).
 

### 2018-09-20 rigid_bodies.tex
Several changes in the Discretization Error Effects

@@
 \usepackage{amssymb}
+\usepackage{float}
 \usepackage{tikz}
@@
 This issue is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
-This brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
+This brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
 As we will show here, the method fails to reproduce the specified temperature even for time-step sizes typically employed in MD.
@@
 However, the theory underlying the concept and computation of shadow Hamiltonians is only applicable to symplectic equations and symplectic integrators, thus ruling out most NVT-MD methods in use today.
-In this paper, we devise a novel procedure that is able to overcome this limitation in the case of global thermostatting methods such as, for instance, the Nos\'{e}-Hoover Chain (NHC) \cite{Martyna_1992}, the Stochastic Velocity Rescaling (SVR) \cite{Bussi_2007}, and the Nos\'{e}-Hoover-Langevin (NHL) \cite{Leimkuhler_2009} methods.
+In this paper, we devise a novel procedure that is able to overcome this limitation in the case of global thermostatting methods such as, for instance, the Nos\'{e}-Hoover Chain (NHC) \cite{Martyna_1992}, the Stochastic Velocity Rescaling \cite{Bussi_2007}, and the Nos\'{e}-Hoover-Langevin \cite{Samoletov_2007, Leimkuhler_2009} methods.
 It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and then applying backward error analysis to that part only.
@@
 The new method is specially appealing for rigid-body MD, primarily because numerical stability can, in this case, extend up to time-step sizes large enough to yield pronounced discretization errors.
-For rigid water, which served as an excellent test case, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the double-thermostat method of Kamberaj, Low, and Neal (KLN) \cite{Kamberaj_2005}, but also a reimplementation for rigid bodies of the NHC integrator developed by Martyna \textit{et al}.
-\cite{Martyna_1996}, as well as the SVR method \cite{Bussi_2007} employed to simultaneously rescale translational and angular velocities.
+For rigid water, which served as an excellent test case, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the KLN method \cite{Kamberaj_2005}, but also a reimplementation for rigid bodies of the NHC integrator introduced by Martyna \textit{et al}.
+\cite{Martyna_1996}, as well as stochastic rescaling \cite{Bussi_2007} applied simultaneously to translational and angular velocities.
 
@@
 \end{equation}
-where subscript $\timestep$ points out the dependency of this average on the employed time step size.
+where subscript $\timestep$ points out that this average depends on the employed time-step size.
 The expression above extends the main result of Eastwood \textit{et al}. \cite{Eastwood_2010} to systems containing rigid bodies.
 It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
-We emphasize that the temperature in Eq.~\ref{eq:equipartition} concerns the system that has actually been simulated, rather than the one initially specified.
+We emphasize that the temperature in Eq.~\eqref{eq:equipartition} concerns the system that has actually been simulated, rather than the one initially specified.
 
@@
 
-In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} and the method of Martyna \textit{et al}.
-\cite{Martyna_1996} with our described adaptations.
+In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) and the NHC integrator of Martyna \textit{et al}.
+\cite{Martyna_1996} with our adaptations for rigid bodies.
 
@@
 \subsection{Numerical Stability}
+\label{sec:numerical stability}
 
@@
 
-To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} is employed.
+To end this section, we report in Fig.~\ref{fig:num_stab} an anomalous behavior observed when the KLN method \cite{Kamberaj_2005} is employed.
 Particularly, one can see in Fig.~\ref{fig:num_stab}(a) that, for each time-step size, the extended energy $H$ drifts upwards until reaching a steady-state value at which the integrator becomes stable.
-These results were obtained with our own simulation code and double-checked using LAMMPS \cite{Plimpton_1995} with a custom pair style.
-As already mentioned, the scheme due to Kamberaj \textit{et al}. \cite{Kamberaj_2005} couples separate NHC thermostats to the rotational and translational degrees of freedom.
+These results were obtained with our own simulation code and double-checked using LAMMPS \cite{Plimpton_1995} with a customized pair style.
+As already mentioned, the KLN scheme \cite{Kamberaj_2005} couples separate NHC thermostats to the rotational and translational degrees of freedom.
 Considering a different factorization scheme, Davidchack \cite{Davidchack_2010} discovered the existence of energy flows between the two thermostats.
-This is confirmed for the method of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} in Fig.~\ref{fig:num_stab}(b), where we depict separately the contribution of each thermostat to the extended energy $H$.
+This is confirmed for the KLN method \cite{Kamberaj_2005} in Fig.~\ref{fig:num_stab}(b), where we depict separately the contribution of each thermostat to the extended energy $H$.
 In this figure, the negative energy values always correspond to the thermostat coupled to the translational degrees of freedom.
@@
 	\includegraphics{Figures/numerical_stability.eps}
-	\caption{Time evolution of (a) the extended energy and (b) of each thermostat energy for the NVT method introduced by Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}.
+	\caption{Time evolution of (a) the extended energy and (b) of each thermostat energy for the NVT method introduced by Kamberaj, Low, and Neal \cite{Kamberaj_2005}.
 In part (b), the negative values correspond to the thermostat coupled to the translational degrees of freedom.}
@@
 In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods and time-step sizes.
-All simulations were carried out for $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
+All simulations were carried out for $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature
+\begin{equation*}
+T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}},
+\end{equation*}
 the mean potential energy per molecule $\langle U/N \rangle$,
 the mean internal virial per molecule $\langle W/N \rangle$, and
-the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
+the constant-volume specific heat capacity
+\begin{equation*}
+c_\mini{V} = \frac{\partial \langle \Ham{}/N \rangle}{\partial T} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}.
+\end{equation*}
+
 The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
@@
 	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d) the constant-volume specific heat of water, as obtained from NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules at $T=298~\mathrm{K}$ and $\rho=970~\mathrm{kg/m^3}$ and employing different numerical integration schemes.
-	The ``Martyna'' scheme ($\bullet$) corresponds to our rigid-body implementation of the Nos\'{e}-Hoover chain integrator introduced in Ref.~\citenum{Martyna_1996}, with analytically computed free rotations \cite{Silveira_2017}.
-	The ``Refined/NHC'' scheme ($\blacksquare$) is obtained by refining the ``Martyna'' method such as described in Sec.~\ref{sec:refined_method}.
-	The ``Refined/NHC/Miller'' scheme ($\blacktriangle$) differs from the preceding one only in that an approximate solution is used for free rotations \cite{Miller_2002}.
-	The ``Refined/Bussi'' scheme ($\times$) is a refined version of the stochastic method of Bussi \textit{et al}. \cite{Bussi_2007}.
-	Finally, the ``Kamberaj'' scheme ($\blacktriangledown$) consists in the double-thermostat method introduced in Ref.~\citenum{Kamberaj_2005}.}
+	The ``NHC'' scheme ($\bullet$) corresponds to the Nos\'{e}-Hoover Chain integrator of Martyna \textit{et al}.
+\cite{Martyna_1996} extended to rigid bodies with analytically computed free rotations \cite{Silveira_2017}.
+	The ``Refined NHC'' scheme ($\blacksquare$) is obtained by subjecting the preceding method to the refinement procedure described in Sec.~\ref{sec:refined_method}.
+	The ``Refined NHC/NSQ'' scheme ($\blacktriangle$) differs from ``Refined NHC'' only in that the approximate NO-SQUISH \cite{Dullweber_1997, Miller_2002} solution is used for free rotations.
+	The ``Refined CSVR'' scheme ($\times$) corresponds to the canonical stochastic velocity rescaling method \cite{Bussi_2007} subject to refinement as well.
+	Finally, the ``KLN'' scheme ($\blacktriangledown$) consists in the double-thermostat method of Kamberaj, Low, and Neal \cite{Kamberaj_2005}.}
 	\label{fig:properties}
@@
 Our measure-preserving extension to rigid bodies of the splitting scheme by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
-Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
-First, the method is applied in a conventional, refinement-free way, with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
-Then, it is modified by introducing the refinement/reweighting procedure proposed here in Sec.~\ref{sec:refined_method}.
-In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
+Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain \cite{Martyna_1992} is attached to all degrees of freedom.
+First, the method is applied in a conventional (i.e. unrefined) way with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
+Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
+In Fig.~\ref{fig:properties}, these two integrators are referred to as ``NHC'' and ``Refined NHC'', respectively.
 Next, we replace the unsplit rotation solution by the simpler NO-SQUISH splitting of Miller \textit{et al}.
-\cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
+\cite{Miller_2002}, denoting the resulting integrator by ``Refined NHC/NSQ'' in Fig.~\ref{fig:properties}.
 We note that our refinement procedure is only rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations.
 Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
-The fourth integrator we test is a refined version of the velocity-rescaling method of Bussi \textit{et al}.
-\cite{Bussi_2007}, which differs from the others for involving a global stochastic thermostat.
-This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
-Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} for comparison purposes, despite its spurious energy flux described in the preceding section.
-This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.
-We stress that thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
-For the former, simple averages are taken from the sampled configurations.
+The fourth integrator we test is a refined version of the velocity rescaling method of Bussi, Donadio, and Parrinello \cite{Bussi_2007}, which differs from the others for involving a global stochastic thermostat.
+This is identified in Fig.~\ref{fig:properties} as ``Refined CSVR'' (acronym for Canonical Stochastic Velocity Rescaling).
+Finally, we also consider the double-thermostat KLN integrator \cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux it generates (see Sec.~\ref{sec:numerical stability}).
+This is to show that the method, denoted by ``KLN'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.
+We stress that thermodynamic properties are estimated in distinct ways for the conventional and the refined integrators.
+For the former class, simple averages are taken from the sampled configurations.
 For the latter, reweighting is carried out via Eq.~\eqref{eq:reweighting}.
 
-In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for the thermostat identified as ``Kamberaj'' decreases appreciably as the time step size increases.
-This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, indicating that the effect of the NHC factorization scheme of Kamberaj \textit{et al}.
-\cite{Kamberaj_2005} on the particle momenta is weak.
-On the other hand, the stringent temperature control achieved by the thermostat named ``Martyna'' leads to a severely degraded accuracy in both $\langle U/N \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b) and \ref{fig:properties}(c).
+In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for the thermostat identified as ``KLN'' decreases appreciably as the time step size increases.
+This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, indicating that the KLN factorization scheme \cite{Kamberaj_2005} produces a relatively weak effect on the particle momenta.
+On the other hand, the stringent temperature control achieved by the ``NHC'' scheme leads to a severely degraded accuracy in both $\langle U/N \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b) and \ref{fig:properties}(c), respectively.
 According to the results of Fig.~\ref{fig:nve}, discretization errors in the NVE integration become manifest at $2~\text{fs}$.
-Thus, the apparent success of the scheme by Martyna \textit{et al.} \cite{Martyna_1996}, in terms of the kinetic energy control for the entire range of time-step sizes considered, is wiped out by those discretization errors which, if not properly handled, yield considerably degraded configurational properties.
+Thus, the apparent success of the unrefined ``NHC'' scheme, in terms of kinetic energy control along the entire range of time-step sizes considered, is wiped out by those discretization errors.
+If not handled properly, they can yield unreliable configurational properties.
 
-From Fig.~\ref{fig:properties}, it becomes clear that the refined schemes yield significantly improved results: for $\timestep < 5~\text{fs}$ we are able to simultaneously reproduce the specified temperature and the other mechanical properties considered.
-For longer time-step sizes, a small, still clearly influencing systematic error is present, as the computed averages slightly depart from the reference values.
-This could, in principle, be remediated by using a higher-order approximation to the shadow Hamiltonian.
-As stated above, the accuracy of the approximate solution is very high for the time-step sizes considered here\cite{Silveira_2017} and this explains the good agreement between the ``Refined/NHC'' and ``Refined/NHC/Miller'' results.
-These, in turn, are similar to the results obtained with the ``Refined/Bussi'' method.
-This was expected as, although not shown here, the performances of the unrefined Bussi\cite{Bussi_2007} and Martyna\cite{Martyna_1996} schemes were found to be analogous.
+From Fig.~\ref{fig:properties}, it becomes clear that the refined schemes produce significantly improved results.
+For $\timestep \leq 5~\mathrm{fs}$, we are able to accurately reproduce the specified temperature and compute other thermodynamic properties considered.
+For longer time steps, a small but clear systematic error is present, as the computed averages slightly depart from the reference values.
+In principle, this could be mitigated by using a higher-order approximation to the shadow Hamiltonian.
+Nonetheless, we cannot anticipate whether the ability to increase the step size would compensate the additional effort required per time step.
 
-Interestingly, for $\timestep < 5~\text{fs}$ all the NVT schemes yield similar results of $c_\mini{V}$.
-Even though the discretization errors affect $\Ham{}$,  Fig.~\ref{fig:properties}(d) shows that the derivative of $\Ham{}$ with respect $T$ is being successfully predicted by all schemes, as $c_\mini{V} \equiv \partial \Ham{} / \partial T$.
-In addition, the fact that $c_\mini{V}$ does not change much in the range of estimated temperatures depicted in Fig.~\ref{fig:properties}(a)\cite{Yigzawe_2013} explains why the poor performance of the scheme by Kamberaj \textit{et al.} \cite{Kamberaj_2005} has almost no impact on the predicted $c_\mini{V}$.
+As stated above, the NO-SQUISH solution \cite{Dullweber_1997, Miller_2002} is very accurate for the system and time-step sizes considered here \cite{Silveira_2017},
+which explains the remarkable agreement between the results shown in Fig.~\ref{fig:properties} for the ``Refined NHC'' and ``Refined NHC/NSQ'' schemes.
+This is an advantage because the NO-SQUISH method is much simpler to implement than the exact rotation solution and is already available in LAMMPS \cite{Plimpton_1995} and other MD packages.
 
-With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for the ``Refined/NHC'' scheme the free energy difference per water molecule ($\Delta F/N$) between the sampled and target states as a function of the time-step size.
+Regarding the refinement of stochastic methods, the results obtained from the ``Refined CSVR'' scheme are similar to those from other refined schemes.
+This was expected because the unrefined versions of the CSVR \cite{Bussi_2007} (stochastic) and the NHC \cite{Martyna_1996} (deterministic) thermostats for rigid bodies produce virtually identical thermodynamic properties of water as functions of the time-step size (results not shown).
+Similar behavior is expected for other global thermostats such as the Nos\'{e}-Hoover-Langevin method \cite{Samoletov_2007, Leimkuhler_2009}.
+Extension of our approach to massive Langevin-type thermostats \cite{Brunger_1984, Davidchack_2009, Davidchack_2015}, in which every degree of freedom is acted upon independently, is beyond the scope of this work.
+Nonetheless, we envision that such an extension would entail using the components of entrywise products between $\vt p$ and $\dot{\vt r}^{[n]}$, and also between ${\vt \pi}$ and $\dot{\vt q}^{[n]}$ in the case of rigid bodies, instead of the scalar products present in Eq.~\eqref{eq:modified kinetic energy estimator}.
+
+Interestingly, all thermostatting schemes analyzed in Fig.~\ref{fig:properties}, including the KLN method, yield similar results for $c_\mini{V}$.
+Even though the average $\langle \Ham{} \rangle$ gets affected by discretization errors, Fig.~\ref{fig:properties}(d) shows that the magnitude of fluctuations in the sampled $\Ham{}$ values does not, and this is successfully predicted by all schemes.
+To put it differently, consider that discretization errors in a simulation at temperature $T$ cause the computed average $\langle \Ham{} \rangle$ to erroneously correspond to a nearby temperature, say, $T^\ast$.
+If the specific heat of the system changes little with temperature in the proximities of $T$, then it is natural to expect little effect of the same discretization errors on the computed $c_\mini{V}$.
+This is the case of liquid water in the vicinities of $298~\mathrm{K}$.
+
+With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for the ``Refined NHC'' scheme the free energy difference per water molecule ($\Delta F/N$) between the sampled and target states as a function of the time-step size.
 $\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\eqref{eq:reweighting}.
@@
 	\caption{Difference in free energy per water molecule between the sampled and target states as a function of the time-step size.
-The sampled state was obtained by employing the ``Refined/NHC'' scheme, while the target state is consistent with a distribution proportional to $e^{-\beta \Ham{}}$, where $\Ham{}$ corresponds to the specified Hamiltonian.}
+The sampled state was obtained by employing the ``Refined NHC'' scheme, while the target state is consistent with a distribution proportional to $e^{-\beta \Ham{}}$, where $\Ham{}$ corresponds to the specified Hamiltonian.}
 	\label{fig:free_energy}
@@
 
-We end this section with a discussion of discretization effects on the kinetic energy partition, for which we present the results corresponding to the methods referred as ``Kamberaj'', ``Martyna'' and ``Refined/NCH''.
+We end this section with a discussion of discretization effects on the kinetic energy partition, for which we present the results corresponding to the methods referred as ``KLN'', ``NHC'' and ``Refined NHC''.
 In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition.

### 2018-09-20 Figures/energy_partition.py

@@ -5,7 +5,7 @@ from matplotlib import rc
 plt.rcParams.update({'font.size':8})
 rc('text', usetex=True)
 
-fig,((ax1,ax2),(ax3,ax4)) = plt.subplots(2,2,sharex=True,sharey=True)  
+fig,((ax1,ax4),(ax3,ax2)) = plt.subplots(2,2,sharex=True,sharey=True)  
 ax = (ax1,ax2,ax3,ax4)
 fig.subplots_adjust(hspace=0.1,wspace=0.1)
 fig.set_size_inches(w=7.5,h=4)
@@ -53,13 +53,13 @@ for i in integrators:
     k = k + 1
 
 
-ax[3].annotate('Refined/NHC (No Reweighting)', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
+ax[3].annotate('Refined NHC \& Refined Kinetic Energy', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
                  textcoords='axes fraction')
-ax[2].annotate('Refined/NHC', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
+ax[2].annotate('Refined NHC \& Reweighted Kinetic Energy', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
                  textcoords='axes fraction')
-ax[0].annotate('Martyna', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
+ax[0].annotate('NHC', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
                  textcoords='axes fraction')
-ax[1].annotate('Kamberaj', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
+ax[1].annotate('KLN', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
   textcoords='axes fraction')
 
 ax[0].annotate('(a)', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.008,0.9),  \

### 2018-09-20 Figures/thermodynamic_properties.py

@@ -32,7 +32,7 @@ beta = 1./(kB*298.0)
 integrators = ['P1','S1','S1mr','B1s','K2']
 colors = ['red','black','magenta','green','blue']
 markers = ['o','s','^','x','v']
-labels = [r'Martyna', 'Refined/NHC','Refined/NHC/Miller','Refined/Bussi', r'Kamberaj']
+labels = [r'NHC', 'Refined NHC','Refined NHC/NSQ','Refined CSVR', r'KLN']
 k = 0
 for i in integrators:
     data = np.genfromtxt(i + '.csv', delimiter=',', skip_header=1, names=[' ','T','f','df','temperatura',\

### 2018-09-20 rigid_bodies.tex
Replacement of \text{\tiny B} by \mini{B}

@@
 \label{eq:nvt extended energy}
-H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
+H = K + U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\mini{B} T\eta_1 + k_\mini{B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j,
 \end{equation}
-where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the target temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
-As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.
+where $k_\mini{B}$ is the Boltzmann constant, $T$ is the target temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter.
+As recommended in Ref.~\citenum{Martyna_1992}, we can make $Q_1 = L k_\mini{B} T t_d^2$ and $Q_j = k_\mini{B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostat chain.
 
@@
 \label{eq:generalized force NVT}
-&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
-&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1.
+&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\mini{B} T \quad \text{and}\\
+&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\mini{B} T  \qquad \text{for} \quad j > 1.
 \end{align}
@@
 
-With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
+With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\mini{B} T$.
 If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as driven variables \cite{Tuckerman_2001}.
@@
 
-By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
+By defining a propagator $e^{\timestep \mathcal{L}_\mini{NVT}}$, where $\mathcal{L}_\mini{NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
 \begin{equation}
 \label{eq:splitting NVT}
-e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
+e^{\timestep \mathcal{L}_\mini{NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\mini{NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\mini{NHC}} + \mathcal{O}(\timestep^3),
 \end{equation}
-where $\Liu{\tiny NVE}$ is the same operator of Sec.~\ref{sec:nve} and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
+where $\Liu{\tiny NVE}$ is the same operator of Sec.~\ref{sec:nve} and $\mathcal{L}_\mini{NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain.
 This one can be split even further as
@@
 \label{eq:equipartition}
-L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
+L k_\mini{B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
 \end{equation}
@@
 Bringing the attention back to the NVT case, our proposal for dealing with discretization effects consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
-Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
+Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\mini{B} T}$.
 Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
@@
 \label{eq:thermostat_force}
-G_1 = 2 \refined K - L k_\text{\tiny B} T.
+G_1 = 2 \refined K - L k_\mini{B} T.
 \end{equation}
@@
 \label{eq:this extended energy}
-{\refined H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
+{\refined H} = \refined K + \refined U + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\mini{B} T\eta_1 + k_\mini{B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
 \end{equation}
@@
 \label{eq:reweighting}
-\langle A \rangle_0 = \frac{\left\langle A \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}{\left\langle \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\text{\tiny B} T}}\right)\right\rangle_\timestep}.
+\langle A \rangle_0 = \frac{\left\langle A \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\mini{B} T}}\right)\right\rangle_\timestep}{\left\langle \exp \left({\frac{\refined{\Ham{}} - \Ham{}}{k_\mini{B} T}}\right)\right\rangle_\timestep}.
 \end{equation}

### 2018-11-10 rigid_bodies.tex
Further changes in the discretization error correction section

@@
 
-\newenvironment{smallarray}[1]                          % small arrays
-{
-	\null\,\vcenter\bgroup\scriptsize
-	\renewcommand{\arraystretch}{1.5}
-	\arraycolsep=.13885em
-	\hbox\bgroup$\left[\array{@{}#1@{}}}
-	{\endarray\right]$\egroup\egroup\,\null
-}
-
 \begin{document}
@@
 
-\title{Refinement of Thermostated Molecular Dynamics via Backward Error Analysis}
+\title{Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis}
 
@@
 
-Debenedetti....
+Debenedetti.... \cite{Palmer_2018} \cite{Limmer_2011, Limmer_2013}
 
@@
 \label{eq:def_B_and_C}
-\mt B(\vt q) = \begin{smallarray}{rrrr}
+\mt B(\vt q) = \left[\begin{array}{rrrr}
 -q_2 & -q_3 & -q_4 \\
@@
 -q_3 &  q_2 &  q_1
-\end{smallarray}
-%
-\; \text{and} \;
-%
-\mt C(\vt q) = \begin{smallarray}{rrrr}
+\end{array}\right]
+\end{equation*}
+and
+\begin{equation*}
+\mt C(\vt q) = \left[\begin{array}{rrrr}
 -q_2 & -q_3 & -q_4 \\
@@
  q_3 & -q_2 &  q_1
-\end{smallarray},
+\end{array}\right]
 \end{equation*}
@@
 
-Regarding the refinement of stochastic methods, the results obtained from the ``Refined CSVR'' scheme are similar to those from other refined schemes.
-This was expected because the unrefined versions of the CSVR \cite{Bussi_2007} (stochastic) and the NHC \cite{Martyna_1996} (deterministic) thermostats for rigid bodies produce virtually identical thermodynamic properties of water as functions of the time-step size (results not shown).
+Regarding the refinement of stochastic methods, the results obtained from the ``Refined CSVR'' scheme are similar to those from other refined ones.
+This was expected because the unrefined versions of the CSVR \cite{Bussi_2007} (stochastic) and the NHC \cite{Martyna_1996} (deterministic) thermostats for rigid bodies produce virtually identical properties as functions of the time-step size (results not shown).
 Similar behavior is expected for other global thermostats such as the Nos\'{e}-Hoover-Langevin method \cite{Samoletov_2007, Leimkuhler_2009}.
 Extension of our approach to massive Langevin-type thermostats \cite{Brunger_1984, Davidchack_2009, Davidchack_2015}, in which every degree of freedom is acted upon independently, is beyond the scope of this work.
-Nonetheless, we envision that such an extension would entail using the components of entrywise products between $\vt p$ and $\dot{\vt r}^{[n]}$, and also between ${\vt \pi}$ and $\dot{\vt q}^{[n]}$ in the case of rigid bodies, instead of the scalar products present in Eq.~\eqref{eq:modified kinetic energy estimator}.
+Nonetheless, we envision that this would entail using the components of entrywise products between $\vt p$ and $\dot{\vt r}^{[n]}$, and also between ${\vt \pi}$ and $\dot{\vt q}^{[n]}$ in the case of rigid bodies, instead of the scalar products present in Eq.~\eqref{eq:modified kinetic energy estimator}.
 
-Interestingly, all thermostatting schemes analyzed in Fig.~\ref{fig:properties}, including the KLN method, yield similar results for $c_\mini{V}$.
-Even though the average $\langle \Ham{} \rangle$ gets affected by discretization errors, Fig.~\ref{fig:properties}(d) shows that the magnitude of fluctuations in the sampled $\Ham{}$ values does not, and this is successfully predicted by all schemes.
-To put it differently, consider that discretization errors in a simulation at temperature $T$ cause the computed average $\langle \Ham{} \rangle$ to erroneously correspond to a nearby temperature, say, $T^\ast$.
-If the specific heat of the system changes little with temperature in the proximities of $T$, then it is natural to expect little effect of the same discretization errors on the computed $c_\mini{V}$.
-This is the case of liquid water in the vicinities of $298~\mathrm{K}$.
+Fig.~\ref{fig:properties}(d) contains results for the constant-volume specific heat capacity.
+Interestingly, these results do not get affected by the discretization errors when $\timestep \leq 5~\mathrm{fs}$, not even if the KLN method is employed.
+Given that the specific heat of water changes little with temperature in the vicinity of $298~\mathrm{K}$, it is not surprising to observe that $c_\mini{V}$ remains unaltered while other properties vary.
 
-With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for the ``Refined NHC'' scheme the free energy difference per water molecule ($\Delta F/N$) between the sampled and target states as a function of the time-step size.
-$\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\eqref{eq:reweighting}.
-It should be noted that the other refined schemes yield identical results and, as expected, the free energy is only negligible for $h = 1~\text{fs}$.
+With the aim of highlighting the importance of the reweighting procedure, in Fig.~\ref{fig:free_energy} we show the free energy difference per water molecule ($\Delta F/N$) between the actually sampled and the target states, obtained with the ``Refined NHC'' scheme as a function of the time-step size.
+Note that $\Delta F$ corresponds to the logarithm of the denominator of Eq.~\eqref{eq:reweighting}.
+It should be mentioned that the other refined schemes yield similar results and, as expected, the free energy difference is non-negligible except for the smallest time steps.
 
@@
 
-A significant enhancement was achieved for a system of TIP3P rigid water molecules\cite{Jorgensen_1983} by using an analytically approximation to the Shadow Hamiltonian (here refered to as \textit{refined Hamiltonian}) with a minimal computatioal overhead, and whose derivation was greatly facilitated by the unsplit solution of rotation introduced in our previous paper\cite{Silveira_2017}.
+A significant enhancement was achieved for a system of TIP3P rigid water molecules\cite{Jorgensen_1983} by using an analytically approximation to the Shadow Hamiltonian (here referred to as \textit{refined Hamiltonian}) with a minimal computational overhead, and whose derivation was greatly facilitated by the unsplit solution of rotation introduced in our previous paper\cite{Silveira_2017}.
 Particularly, we have been able to remove the serious artifacts observed in the schemes of Martyna \textit{et al}.

### 2018-11-11 rigid_bodies.tex
Removal of quotation marks in scheme denominations

@@
 	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d) the constant-volume specific heat of water, as obtained from NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules at $T=298~\mathrm{K}$ and $\rho=970~\mathrm{kg/m^3}$ and employing different numerical integration schemes.
-	The ``NHC'' scheme ($\bullet$) corresponds to the Nos\'{e}-Hoover Chain integrator of Martyna \textit{et al}.
+	The NHC scheme ($\bullet$) corresponds to the Nos\'{e}-Hoover Chain integrator of Martyna \textit{et al}.
 \cite{Martyna_1996} extended to rigid bodies with analytically computed free rotations \cite{Silveira_2017}.
-	The ``Refined NHC'' scheme ($\blacksquare$) is obtained by subjecting the preceding method to the refinement procedure described in Sec.~\ref{sec:refined_method}.
-	The ``Refined NHC/NSQ'' scheme ($\blacktriangle$) differs from ``Refined NHC'' only in that the approximate NO-SQUISH \cite{Dullweber_1997, Miller_2002} solution is used for free rotations.
-	The ``Refined CSVR'' scheme ($\times$) corresponds to the canonical stochastic velocity rescaling method \cite{Bussi_2007} subject to refinement as well.
-	Finally, the ``KLN'' scheme ($\blacktriangledown$) consists in the double-thermostat method of Kamberaj, Low, and Neal \cite{Kamberaj_2005}.}
+	The Refined NHC scheme ($\blacksquare$) is obtained by subjecting the preceding method to the refinement procedure described in Sec.~\ref{sec:refined_method}.
+	The Refined NHC/NSQ scheme ($\blacktriangle$) differs from Refined NHC only in that the approximate NO-SQUISH \cite{Dullweber_1997, Miller_2002} solution is used for free rotations.
+	The Refined CSVR scheme ($\times$) corresponds to the canonical stochastic velocity rescaling method \cite{Bussi_2007} subject to refinement as well.
+	Finally, the KLN scheme ($\blacktriangledown$) consists in the double-thermostat method of Kamberaj, Low, and Neal \cite{Kamberaj_2005}.}
 	\label{fig:properties}
@@
 Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
-In Fig.~\ref{fig:properties}, these two integrators are referred to as ``NHC'' and ``Refined NHC'', respectively.
+In Fig.~\ref{fig:properties}, these two integrators are referred to as NHC and Refined NHC, respectively.
 Next, we replace the unsplit rotation solution by the simpler NO-SQUISH splitting of Miller \textit{et al}.
-\cite{Miller_2002}, denoting the resulting integrator by ``Refined NHC/NSQ'' in Fig.~\ref{fig:properties}.
+\cite{Miller_2002}, denoting the resulting integrator by Refined NHC/NSQ in Fig.~\ref{fig:properties}.
 We note that our refinement procedure is only rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations.
@@
 The fourth integrator we test is a refined version of the velocity rescaling method of Bussi, Donadio, and Parrinello \cite{Bussi_2007}, which differs from the others for involving a global stochastic thermostat.
-This is identified in Fig.~\ref{fig:properties} as ``Refined CSVR'' (acronym for Canonical Stochastic Velocity Rescaling).
+This is identified in Fig.~\ref{fig:properties} as Refined CSVR (acronym for Canonical Stochastic Velocity Rescaling).
 Finally, we also consider the double-thermostat KLN integrator \cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux it generates (see Sec.~\ref{sec:numerical stability}).
-This is to show that the method, denoted by ``KLN'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.
+This is to show that the method, denoted by KLN in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.
 We stress that thermodynamic properties are estimated in distinct ways for the conventional and the refined integrators.
@@
 
-In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for the thermostat identified as ``KLN'' decreases appreciably as the time step size increases.
+In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for the thermostat identified as KLN decreases appreciably as the time step size increases.
 This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, indicating that the KLN factorization scheme \cite{Kamberaj_2005} produces a relatively weak effect on the particle momenta.
-On the other hand, the stringent temperature control achieved by the ``NHC'' scheme leads to a severely degraded accuracy in both $\langle U/N \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b) and \ref{fig:properties}(c), respectively.
+On the other hand, the stringent temperature control achieved by the NHC scheme leads to a severely degraded accuracy in both $\langle U/N \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b) and \ref{fig:properties}(c), respectively.
 According to the results of Fig.~\ref{fig:nve}, discretization errors in the NVE integration become manifest at $2~\text{fs}$.
-Thus, the apparent success of the unrefined ``NHC'' scheme, in terms of kinetic energy control along the entire range of time-step sizes considered, is wiped out by those discretization errors.
+Thus, the apparent success of the unrefined NHC scheme, in terms of kinetic energy control along the entire range of time-step sizes considered, is wiped out by those discretization errors.
 If not handled properly, they can yield unreliable configurational properties.
@@
 As stated above, the NO-SQUISH solution \cite{Dullweber_1997, Miller_2002} is very accurate for the system and time-step sizes considered here \cite{Silveira_2017},
-which explains the remarkable agreement between the results shown in Fig.~\ref{fig:properties} for the ``Refined NHC'' and ``Refined NHC/NSQ'' schemes.
+which explains the remarkable agreement between the results shown in Fig.~\ref{fig:properties} for the Refined NHC and Refined NHC/NSQ schemes.
 This is an advantage because the NO-SQUISH method is much simpler to implement than the exact rotation solution and is already available in LAMMPS \cite{Plimpton_1995} and other MD packages.
 
-Regarding the refinement of stochastic methods, the results obtained from the ``Refined CSVR'' scheme are similar to those from other refined ones.
+Regarding the refinement of stochastic methods, the results obtained from the Refined CSVR scheme are similar to those from other refined ones.
 This was expected because the unrefined versions of the CSVR \cite{Bussi_2007} (stochastic) and the NHC \cite{Martyna_1996} (deterministic) thermostats for rigid bodies produce virtually identical properties as functions of the time-step size (results not shown).
@@
 
-With the aim of highlighting the importance of the reweighting procedure, in Fig.~\ref{fig:free_energy} we show the free energy difference per water molecule ($\Delta F/N$) between the actually sampled and the target states, obtained with the ``Refined NHC'' scheme as a function of the time-step size.
+With the aim of highlighting the importance of the reweighting procedure, in Fig.~\ref{fig:free_energy} we show the free energy difference per water molecule ($\Delta F/N$) between the actually sampled and the target states, obtained with the Refined NHC scheme as a function of the time-step size.
 Note that $\Delta F$ corresponds to the logarithm of the denominator of Eq.~\eqref{eq:reweighting}.
@@
 	\caption{Difference in free energy per water molecule between the sampled and target states as a function of the time-step size.
-The sampled state was obtained by employing the ``Refined NHC'' scheme, while the target state is consistent with a distribution proportional to $e^{-\beta \Ham{}}$, where $\Ham{}$ corresponds to the specified Hamiltonian.}
+The sampled state was obtained by employing the Refined NHC scheme, while the target state is consistent with a distribution proportional to $e^{-\beta \Ham{}}$, where $\Ham{}$ corresponds to the specified Hamiltonian.}
 	\label{fig:free_energy}
@@
 
-We end this section with a discussion of discretization effects on the kinetic energy partition, for which we present the results corresponding to the methods referred as ``KLN'', ``NHC'' and ``Refined NHC''.
+We end this section with a discussion of discretization effects on the kinetic energy partition, for which we present the results corresponding to the methods referred as KLN, NHC and Refined NHC.
 In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition.

### 2018-11-11 rigid_bodies.tex
Discretization effects on kinetic energies

@@
 	\includegraphics{Figures/Free_Energy.eps}
-	\caption{Difference in free energy per water molecule between the sampled and target states as a function of the time-step size.
-The sampled state was obtained by employing the Refined NHC scheme, while the target state is consistent with a distribution proportional to $e^{-\beta \Ham{}}$, where $\Ham{}$ corresponds to the specified Hamiltonian.}
+	\caption{Difference in free energy per water molecule between the actually sampled and the target states as a function of the time-step size.
+The sampling was done by employing the Refined NHC scheme, while the target state is consistent with a distribution proportional to $e^{-\beta \Ham{}}$, where $\Ham{}$ corresponds to the specified Hamiltonian.}
 	\label{fig:free_energy}
@@
 
-We end this section with a discussion of discretization effects on the kinetic energy partition, for which we present the results corresponding to the methods referred as KLN, NHC and Refined NHC.
-In Fig.~\ref{fig:energy_partition} we show the translational ($K_t$) and rotational ($K_r$) kinetic energies as a function of the time-step size, and compare the results with the expected values assuming equipartition.
-In Figs.~\ref{fig:energy_partition}(a) to \ref{fig:energy_partition}(c), these were calculated from the kinetic energy $\langle K \rangle$, as follows: $\langle K_t \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times (3N-3)$ and $\langle K_r \rangle_\text{eq}$ = $\frac{\langle K \rangle}{6N-3} \times 3N$.
-For Fig.~\ref{fig:energy_partition}(d), we substituted $\langle K \rangle$ for $\langle \refined{K} \rangle$ in the right-hand side of these equations.
-
-In our previous work \cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water.
-From Fig.~\ref{fig:energy_partition}(b), we see that the method of Kamberaj \textit{et al}. \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
-The results of Fig.~\ref{fig:energy_partition}(a), which were obtained for the numerical scheme of Martyna \textit{et al}.
-\cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition}(b), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions.
-Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
-From Fig.~\ref{fig:energy_partition}(d) one can note that the refined translational and rotational kinetic energies obey the energy partition predicted by Tolman's theorem \cite{Tolman_1918} for time-step sizes up to $5~\text{fs}$.
-For this range of $\timestep$ and after reweighting, which yields Fig.~\ref{fig:energy_partition}(c), it becomes clear that our method corrects the artifacts observed in the schemes of Martyna \textit{et al}.
-\cite{Martyna_1996} and Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}.
+We end this section with a discussion about discretization effects on the kinetic energy partition amongst translational and rotational degrees of freedom.
+Davidchack \cite{Davidchack_2010} has already performed a thorough analysis of such effects in thermostated rigid-body dynamics.
+Here we complement his study by including the KLN method \cite{Kamberaj_2005} and by showing that the refinement procedure is able to mitigate the observed artifacts.
+In Fig.~\ref{fig:energy_partition}, we present computed averages of the translational ($K_t$) and rotational ($K_r$) kinetic energies as functions of the time-step size, and compare them to the values that would be expected by assuming equipartition.
+In Figs.~\ref{fig:energy_partition}(a) to \ref{fig:energy_partition}(c), these were calculated from the average kinetic energy $\langle K \rangle$ by means of $\langle K_t \rangle_\text{eq}$ = $\frac{3N-3}{6N-3} \langle K \rangle$ and $\langle K_r \rangle_\text{eq}$ = $\frac{3N}{6N-3} \langle K \rangle$.
+For Fig.~\ref{fig:energy_partition}(d), $K$ is simply replaced by $\refined{K}$ in these equations.
+In a previous work \cite{Silveira_2017}, we have shown that the rotational energy is more strongly influenced by the time-step size in the NVE case than the translational energy.
+This occurs because the rotational degrees of freedom display the fastest motion in a rigid water molecule \cite{Silveira_2017}.
+In Fig.~\ref{fig:energy_partition}(a), one can see that the KLN method \cite{Kamberaj_2005} tends, once again, to reproduce the behavior of the NVE integration.
+The results of Fig.~\ref{fig:energy_partition}(b) were obtained by using the NHC scheme of Martyna \textit{et al}. \cite{Martyna_1996} applied to rigid bodies.
+These results are similar to those shown in Fig.~\ref{fig:energy_partition}(a), in the sense that the simulated translational and rotational components deviate upwards and downwards, respectively, from their equipartioned values.
+This observation is in consonance with the analysis of Ref.~\citenum{Davidchack_2010}.
+To some extent, our refinement/reweighting procedure is able to correct these deviations, as one can see in Fig.~\ref{fig:energy_partition}(c).
+Finally, by observing Fig.~\ref{fig:energy_partition}(d), one can note that the refined translational and rotational kinetic energies, being estimates of their ``shadow'' counterparts, comply with Tolman's equipartition theorem \cite{Tolman_1918} for time-step sizes up to $5~\text{fs}$.
 
@@
 	\includegraphics{Figures/energy_partition.eps}
-		\caption{Influence of the time-step size on the kinetic energy partition in NVT MD simulations of 903 TIP3P\cite{Jorgensen_1983} water molecules for (a) our implementation of the numerical scheme by Martyna \textit{et al.} \cite{Martyna_1996} and (b) the method due to Kamberaj \textit{et al.} \cite{Kamberaj_2005}, as well as for (c) our refined version of the thermostat by Martyna \textit{et al.} \cite{Martyna_1996}, for which we also depict in (d) the results obtained without applying the reweighting procedure.
-For each case, we show the simulated translational (black down triangles) and rotational (black up triangles) kinetic energies, as well as the expected translational (green squares) and rotational (red x) values assuming equipartition.}
+		\caption{Influence of the time-step size on the kinetic energy partition in NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules.
+		Part (a) contains results obtained from the KLN method \cite{Kamberaj_2005}.
+		The results in Part (b) were obtained by using the NHC method \cite{Martyna_1996} extended for rigid bodies and without any refinement.
+		In Part (c), they  were obtained from our Refined NHC \cite{Martyna_1996} method, with the kinetic energies computed at the target state via reweighting.
+		Finally, Part (d) depict the refined kinetic energies at every sampled state resulting from the same simulations of Part (c).
+		In all parts, we present the simulated translational ($\blacktriangledown$) and rotational ($\blacktriangle$) kinetic energies, as well as the expected translational ($\blacksquare$) and rotational ($\times$) values assuming equipartition.}
 	\label{fig:energy_partition}

### 2018-11-11 Figures/energy_partition.py

@@ -5,7 +5,7 @@ from matplotlib import rc
 plt.rcParams.update({'font.size':8})
 rc('text', usetex=True)
 
-fig,((ax1,ax4),(ax3,ax2)) = plt.subplots(2,2,sharex=True,sharey=True)  
+fig,((ax2,ax1),(ax3,ax4)) = plt.subplots(2,2,sharex=True,sharey=True)  
 ax = (ax1,ax2,ax3,ax4)
 fig.subplots_adjust(hspace=0.1,wspace=0.1)
 fig.set_size_inches(w=7.5,h=4)
@@ -62,9 +62,9 @@ ax[0].annotate('NHC', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9), 
 ax[1].annotate('KLN', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.08,0.9),  \
   textcoords='axes fraction')
 
-ax[0].annotate('(a)', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.008,0.9),  \
+ax[0].annotate('(b)', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.008,0.9),  \
                  textcoords='axes fraction')
-ax[1].annotate('(b)', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.008,0.9),  \
+ax[1].annotate('(a)', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.008,0.9),  \
                  textcoords='axes fraction')
 ax[2].annotate('(c)', xy=(0, 0) , xycoords= 'axes fraction', xytext=(0.008,0.9),  \
                  textcoords='axes fraction')

### 2018-11-11 rigid_bodies.tex
Further changes in discretization effects on kinetic energies

@@
 To some extent, our refinement/reweighting procedure is able to correct these deviations, as one can see in Fig.~\ref{fig:energy_partition}(c).
-Finally, by observing Fig.~\ref{fig:energy_partition}(d), one can note that the refined translational and rotational kinetic energies, being estimates of their ``shadow'' counterparts, comply with Tolman's equipartition theorem \cite{Tolman_1918} for time-step sizes up to $5~\text{fs}$.
+Finally, by observing Fig.~\ref{fig:energy_partition}(d), one can note that the refined translational and rotational kinetic energies, 
+$\refined K_t = \frac{1}{2} \tr{\vt p} \dot{\vt r}^{[3]}$ and $\refined K_r = \frac{1}{2} \tr{\vt \pi} \dot{\vt q}^{[3]}$, being estimates of their ``shadow'' counterparts, comply with Tolman's equipartition theorem \cite{Tolman_1918} for time-step sizes up to $5~\text{fs}$.
 

### 2018-11-11 Figures/energy_partition.py

@@ -16,7 +16,7 @@ ax[3].set_xlabel('$h$ (fs)')
 
 ax[0].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
 ax[1].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
-ax[2].set_ylabel(r'{$\langle K \rangle$} (kcal/mol)')
+ax[2].set_ylabel(r'{$\langle K \rangle_0$} (kcal/mol)')
 ax[3].set_ylabel(r'{$\langle \widetilde{K} \rangle$} (kcal/mol)')
 ax[0].set_xlim(0.5,7.5)
 ax[0].set_ylim(740,825)

### 2018-11-11 rigid_bodies.tex
Revision of concluding remarks

@@
 
-In this paper, we have devised a general approach that integrates backward error analysis into the Nos\'{e}-Hoover chains and global stochastic thermostats.
-The key rationale behind our proposal came from recognizing that a Shadow kinetic energy should substitute the specified one in the equations dictating the thermostat action.
-This turns out to be crucial for simulations that display considerable discretization errors, as a ``clear definition of simulation temperature''\cite{Eastwood_2010} is obtained by applying Tolman's generalized equipartition theorem to a Shadow kinetic energy.
+In this paper, we have devised a general approach that integrates backward error analysis into Nos\'{e}-Hoover chains and other global thermostats.
+The rationale for our proposal arises from recognizing that a ``shadow kinetic energy'' should substitute the conventional one in the equations that dictate the thermostat action.
+In practice, we employ backward error analysis to compute, with minimal computational overhead, a refined version of the kinetic energy, which serves as an estimate for its shadow analogue.
+This turns out to be crucial for simulations that display considerable discretization errors, in which a clearer definition of the simulation temperature \cite{Eastwood_2010} is obtained by applying Tolman's generalized equipartition theorem to the refined kinetic energy rather than to the conventional one.
 
-A significant enhancement was achieved for a system of TIP3P rigid water molecules\cite{Jorgensen_1983} by using an analytically approximation to the Shadow Hamiltonian (here referred to as \textit{refined Hamiltonian}) with a minimal computational overhead, and whose derivation was greatly facilitated by the unsplit solution of rotation introduced in our previous paper\cite{Silveira_2017}.
-Particularly, we have been able to remove the serious artifacts observed in the schemes of Martyna \textit{et al}.
-\cite{Martyna_1996} and Kamberaj \textit{et al}.
-\cite{Kamberaj_2005}, correcting the discretization effects in numerically computed averages for time-step sizes up to $5~\text{fs}$.
-From our rigorous analysis of these schemes, we would not recommend a time-step size larger than $1~\text{fs}$.
-The fact of having overcome that barrier is of paramount importance as allows us to actually exploit the high numerical stability provided by rigid-body MD.
+A significant enhancement was achieved for a system of rigid TIP3P-water molecules \cite{Jorgensen_1983}.
+Analytical derivation of a refined Hamiltonian formula for rigid bodies was greatly facilitated by considering an unsplit integration of rotations, such as in the method we have recently introduced \cite{Silveira_2017}.
+Particularly, we have been able to remove serious artifacts observed when using the method of Kamberaj, Low, and Neal \cite{Kamberaj_2005} or a rigid-body extension of the standard Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992, Martyna_1996}.
+The refinement/reweighting approach described in Sec.~\ref{sec:refined_method} corrected discretization effects in computed thermodynamic properties of water for time-step sizes up to $5~\text{fs}$.
+Without the refinement, we would not recommend the use of time steps larger than $1~\text{fs}$.
+Having overcome such barrier is of paramount importance as it allows one to actually exploit the high numerical stability and useful coarse-graining provided by rigid-body MD.
 
-Finally, it should be remarked that our method is not limited to rigid-body MD.
-It could be employed, for instance, in conjuction with other constrained MD schemes, such as SHAKE\cite{Ryckaert_1977} and RATTLE\cite{Andersen_1983}.
-Of course, restrictions arising from these methods must be properly taken into account, as we did for the quaternion time-derivative.
+Finally, it is worth remarking that our method is not limited to rigid bodies.
+It can be employed, for instance, in conjunction with other constrained MD schemes, such as SHAKE \cite{Ryckaert_1977} and RATTLE \cite{Andersen_1983}, for instance.
+Of course, restraints arising from these methods must be properly taken into account, as done here for the time-derivatives of orientation quaternions.
 

### 2018-11-12 rigid_bodies.tex
Text revision finished: ready for submission

@@
 \affiliation{Planta Piloto de Ingenier\'ia Qu\'imica, PLAPIQUI, Universidad Nacional del Sur, Camino La Carrindanga Km 7-CC: 717, Bah\'ia Blanca, Argentina}
+\altaffiliation{Current address: Computational \& Systems Biology Program, Memorial Sloan Kettering Cancer Center, New York, NY, USA}
 
@@
 However, systematic errors can possibly be hidden in these deviations, being the cause of non-physical behavior.
-For instance, it has been shown \cite{Eastwood_2010} that errors due to the discretization of the equations of motion can disrupt the equipartition of kinetic energy among molecules of distinct sizes.
-A similar artifact occurs for the partition of kinetic energy between translational and rotational degrees of freedom in rigid-body systems \cite{Davidchack_2010, Silveira_2017}.
-This issue is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
-This brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
+It has been recently shown \cite{Palmer_2018} that erroneous simulation results \cite{Limmer_2011, Limmer_2013} were the cause of a long-standing controversy \cite{Smart_2018} about a liquid-liquid transition believed to exist for supercooled water.
+In this case, hybrid Monte Carlo steps were employed with initial velocities that failed to comply with the equipartition between translational and rotational degrees of freedom \cite{Palmer_2018}.
+As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations as an artifact due to the discretization of the equations of motion.
+This can affect not only the translation/rotation equipartition in rigid-body systems \cite{Davidchack_2010, Silveira_2017}, but also the distribution of energy among molecules of distinct sizes \cite{Eastwood_2010}.
+The issue with rigid-body MD is relevant not only because water \cite{Jorgensen_1983} and other small molecules are usually modeled as rigid bodies, but also because designing proteins, nanoparticles, or other large structures as collections of interconnected bodies \cite{Miller_2002, Knorowski_2012, Patra_2013} can be seen as a useful coarse-graining strategy.
+It also brings into question the NVT-MD method developed by Kamberaj, Low, and Neal \cite{Kamberaj_2005} (KLN) who, by assuming equipartition, coupled two independent thermostats to the translational and rotational degrees of freedom.
 As we will show here, the method fails to reproduce the specified temperature even for time-step sizes typically employed in MD.
@@
 Nonetheless, the cited double-thermostat method is available in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995} and has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}, in spite of all the issues observed here.
-
-Debenedetti.... \cite{Palmer_2018} \cite{Limmer_2011, Limmer_2013}
 
@@
 The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
-Here, we prove that their findings can be extended to the symplectic integration of rigid-body motion.
+As we manage to prove here, their findings can be extended to the symplectic integration of rigid-body motion as well.
 
@@
 
-In this paper, we have devised a general approach that integrates backward error analysis into Nos\'{e}-Hoover chains and other global thermostats.
+In this paper, we have devised a simple, general approach that integrates backward error analysis into Nos\'{e}-Hoover chains and other global thermostats.
 The rationale for our proposal arises from recognizing that a ``shadow kinetic energy'' should substitute the conventional one in the equations that dictate the thermostat action.
-In practice, we employ backward error analysis to compute, with minimal computational overhead, a refined version of the kinetic energy, which serves as an estimate for its shadow analogue.
-This turns out to be crucial for simulations that display considerable discretization errors, in which a clearer definition of the simulation temperature \cite{Eastwood_2010} is obtained by applying Tolman's generalized equipartition theorem to the refined kinetic energy rather than to the conventional one.
+In practice, we employ backward error analysis to compute, with small computational overhead, a refined version of the kinetic energy, which serves as an estimate for its shadow analogue.
+This turns out to be crucial for simulations that display considerable discretization errors, in which a more precise definition of the simulation temperature \cite{Eastwood_2010} is obtained by applying Tolman's generalized equipartition theorem to the refined kinetic energy rather than to the conventional one.
 
-A significant enhancement was achieved for a system of rigid TIP3P-water molecules \cite{Jorgensen_1983}.
+Significant enhancement was achieved for a system of rigid TIP3P-water molecules \cite{Jorgensen_1983}.
 Analytical derivation of a refined Hamiltonian formula for rigid bodies was greatly facilitated by considering an unsplit integration of rotations, such as in the method we have recently introduced \cite{Silveira_2017}.
-Particularly, we have been able to remove serious artifacts observed when using the method of Kamberaj, Low, and Neal \cite{Kamberaj_2005} or a rigid-body extension of the standard Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992, Martyna_1996}.
+In practice, we have been able to remove serious artifacts which take place when the method of Kamberaj, Low, and Neal \cite{Kamberaj_2005} or a rigid-body extension of the standard Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992, Martyna_1996} is used with time steps larger than $1~\text{fs}$.
 The refinement/reweighting approach described in Sec.~\ref{sec:refined_method} corrected discretization effects in computed thermodynamic properties of water for time-step sizes up to $5~\text{fs}$.
-Without the refinement, we would not recommend the use of time steps larger than $1~\text{fs}$.
-Having overcome such barrier is of paramount importance as it allows one to actually exploit the high numerical stability and useful coarse-graining provided by rigid-body MD.
+Having overcome such barrier is of paramount importance as it allows one to actually exploit the high numerical stability and useful coarse-graining strategy provided by rigid-body MD.
 
 Finally, it is worth remarking that our method is not limited to rigid bodies.
-It can be employed, for instance, in conjunction with other constrained MD schemes, such as SHAKE \cite{Ryckaert_1977} and RATTLE \cite{Andersen_1983}, for instance.
-Of course, restraints arising from these methods must be properly taken into account, as done here for the time-derivatives of orientation quaternions.
+It can also be employed in conjunction with other constrained MD schemes, such as SHAKE \cite{Ryckaert_1977} and RATTLE \cite{Andersen_1983}, for instance.
+Of course, constraints must be properly taken into account in the computation of time-derivatives of atomic positions, as was done here for the time-derivatives of quaternions.
 

### 2019-02-20 makediff.sh
Latexdiff support

@@ -1,0 +1,2 @@
+#!/bin/bash
+latexdiff -c ld.cfg rigid_bodies_submitted.tex rigid_bodies.tex > rigid_bodies_diff.tex

### 2019-02-20 response_to_reviewers.tex
Response to reviewers - v. 1

\documentclass[]{article}
\usepackage{setspace}
\usepackage{xcolor}
\usepackage[version=4]{mhchem}
\usepackage{textcomp}
%opening
\title{\normalsize \textbf{Response to the Reviewers' Comments on Manuscript A18.12.0101}}
\author{}
\date{}

\begin{document}
\doublespacing
\maketitle

To: Prof. Carlos Vega

Associate Editor

The Journal of Chemical Physics
\vspace{1cm}

Dear Prof. Vega:

Thank you for your message forwarding the reviewer’s comments on our manuscript entitled “Refinement of Thermostated Molecular Dynamics Using Backward Error Analysis”. We have carefully addressed all comments, which required us to make changes in the text. In order to facilitate your evaluation, we have attached an annotated version in which we crossed out in red all removed text and incorporated new content in blue.

We hope the revised version of our manuscript is suitable for publication in the Journal of Chemical Physics.

\begin{center}
Sincerely yours,
\end{center}
\begin{flushright}
Ana J. Silveira

Memorial Sloan Kettering Cancer Center

\vspace{1cm}

Charlles R. A. Abreu

Federal University of Rio de Janeiro

\end{flushright}

\section{\textbf{Reviewer \#1}}

{\color{blue}{
The authors investigate the influence of finite time step numerical discretization errors on the computation of averages in the molecular system with rigid bodies. They analyze the possibility of compensating for such errors by using 'refined' expressions for the kinetic and potential energies and the re-weighting of computed quantities. While the effects of the numerical integration error and the re-weighting procedures have been studied before, the authors demonstrate how this approach can work in practice, using a model of rigid water molecules as an example.

Besides, the authors investigate the effect of using different thermostatting approaches, highlighting some of the pitfalls of using multiple thermostats, like in the KLN method. Unfortunately, the MD practitioners sometimes don't pay attention to such problems with their thermostats, which may lead to erroneous results.

Therefore, I recommend publication of this manuscript in the JCP, after the authors implement minor revisions listed below. 
}}

\textbf{Answer}:
We thank the reviewer for having done such a careful examination of our manuscript. All comments and suggestions were very meaningful and constructive. We really appreciate it.

{\color{blue}{
- Abstract: ``may distort such assumption". This needs re-wording, since equipartition is not an assumption. Maybe better say ``may lead to deviation from equipartition in measured quantities".
}}

\textbf{Answer}:
We replaced ``may distort such assumption'' by ``may lead to deviations from equipartition”. In order to maintain the specific reference to the kinetic energy, made in the preceding sentence, we did not include the fragment “in measured quantities".

{\color{blue}{
- Abstract: ``quantity that is actually subject to equipartition". I think this need to be stated more precisely, since we don't know the expression for such a quantity. What we can derive is a quantity that deviations from equipartition to higher order in the time step.
}}

\textbf{Answer}:
We replaced “identify” by “obtain a higher-order estimate of”.

{\color{blue}{
- Abstract: ``which coincides with'' better to say ``which converges to".
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.1, middle of right column: ``breakage'' \textrightarrow ``breakdown". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.1, bottom of right column: ``manage to prove here'' - you are not proving, but rather ``show'' or ``demonstrate'' 
}}

\textbf{Answer}:
We were referring to the mathematical derivation in the Appendix. Anyways, we agree that this is not a formal proof and thus replaced “prove” by “demonstrate”.

{\color{blue}{
- p.2, middle of left column: ``specially'' \textrightarrow ``especially'' 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.2, bottom of right column: ``While U(r,q) is an arbitrary function". This may be interpreted too broadly. Better to say something like ``While the form of U(r,q) depends on the specific interaction model". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.2, bottom of right column: ``all bodies piled together'' \textrightarrow ``all bodies combined in a single vector". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.3, top of left column: ``freely moving atoms". I guess the authors talk about a mixed system of rigid bodies and point masses. Freely moving atoms has the meaning of 'non-interacting atoms'. I suggest replacing ``freely moving atoms'' with ``point masses'' or ``interacting point masses". 
}}

\textbf{Answer}:The reviewer is right. The expression ``freely moving atoms'' is not adequate. We decided to use the term “individual point masses” instead.

{\color{blue}{
- p.3, middle of left column: ``is a kick in the linear and quaternion momenta given by", better say ``is a kick that changes the linear and quaternion momenta according to'' 
}}

\textbf{Answer}:Done.

{\color{blue}{
- p.3, before subsection B: ``unplit'' \textrightarrow ``unsplit", although it may be better to say ``exact solution for free rotations'' instead. 
}}

\textbf{Answer}:
We used the second option, that is, “exact solution for free rotations”.

{\color{blue}{
- same place: ``might become clear'' \textrightarrow ``will become clear'' 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.4, middle of left column: ``for entailing the computation (or numerical estimation) of", replace with ``due to the need to evaluate (or approximate numerically) the". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- same place: ``a simpler path, though". Not sure it is simpler, so better say ``different path'' and delete ``though". 
}}

\textbf{Answer}:
We did the suggested replacement.

{\color{blue}{
- same place: delete ``implied", since not sure what this word changes in the meaning of ``we attempt to quantify the discretization errors".
}}

\textbf{Answer}:
Done.

{\color{blue}{
- same place: ``average'' \textrightarrow ``averages'' 
}}

\textbf{Answer}:
Done.
{\color{blue}{
- Start of subsection C: ``our proposal'' \textrightarrow ``our approach'' 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.4, after Eq. (10): ``but keep being independent'' \textrightarrow ``but, like U, independent". 
 }}

\textbf{Answer}:
Done.
{\color{blue}{
- p.4, after Eq. (11): ``stacked vector [p pi]". I think this should be [r, q]. Also, not sure why it is necessary to say ``stacked vector", maybe just enough to say ``with respect to [r q]'' 
}}

\textbf{Answer}:
The reviewer is right. We made the correction and removed the expression “stacked vector”.

{\color{blue}{
- p.4, bottom of right column: introduced ``n-th order estimator", but variable n was used earlier to denote sub-steps in the splitting after Eq. (9) and in the middle of right column on p.5. Suggest to use a different letter to avoid confusion. 
}}

\textbf{Answer}:
We thank the reviewer for pointing this out. We now use “k” instead of “n”. We have also replaced “k” by “j” as the index of a few summation signs in the Appendix.

{\color{blue}{
- p.5, middle of right column: ``By this means'' \textrightarrow ``Thus'' 
}}

\textbf{Answer}:
Done.
{\color{blue}{
- the next paragraph: ``intact'' \textrightarrow ``unchanged'' 
}}

\textbf{Answer}:
Done.
{\color{blue}{
- p.6, middle of right column: Change ``poses the question whether'' to ``introduces the question of whether'' and delete ``or not". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- the next paragraph: what do you mean by ``the angular coefficient"? Is this the slope of the regression line? If yes, then use this term instead of the angular coefficient. 
}}

\textbf{Answer}:
We replaced the original sentence by “It is defined as the slope of a least-square regression line expressing such quantity as a function of time.”

{\color{blue}{
- p.7, middle of left column: ``Leimkhuler'' \textrightarrow ``Leimkuhler".
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.8, middle of left column: ``which differs from the others for involving a global stochastic thermostat", should be ``by involving". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.8, right column, paragraph starting with ``Fig. 5(d) contains...". The conclusion in this paragraph seems dubious. The authors appear to confuse numerical artefacts, due to finite time step, with the physical phenomenon of heat capacity dependence on temperature. A much more plausible explanation is that statistical error in the measurement of $c_v$ (which is related to the variance of the total energy) is much larger (for the same simulation sample) than that in other quantities (like kinetic or potential energy). So, the statistical error masks the systematic error due to finite time step. In other words, the dependence of measured $c_v$ on time step size is present, but would require more simulation time in order to measure with sufficient precision. 
}}

\textbf{Answer}:
Our (unfulfilled) intention here was to suggest that the effect caused by the numerical artifact is similar to that of a small variation in temperature, in which the average $\langle H \rangle$ changes while $c_v$ remains practically unaltered. We have rewritten the paragraph so as to make this intention clearer. We also employed the expression “are not noticeably influenced” in place of “does not get affected”, in order to acknowledge the statistical nature of these measurements.

{\color{blue}{
- Fig. 5 end of caption: replace ``consists in'' with ``is", as in ``Finally, the KLN scheme (triangle) is the double-thermostat method...". 
}}

\textbf{Answer}:
Done.

{\color{blue}{
- p.10, bottom of left column: ``Having overcome such barrier ...'' I'm not sure what ``barrier'' the authors have in mind. What they propose allows to increase the accuracy of the results (maybe double it, if we are talking about increasing the time step from 1fs to 4fs), or get the same accuracy with 25% of the computational resource (again from 1fs to 4fs), but this is not overcoming a ``barrier". So, I suggest either deleting this sentence as unnecessary, or modifying it to reflect more accurately the contribution of this work. 
}}

\textbf{Answer}:

\section{\textbf{Reviewer \#2}}

{\color{blue}{
The manuscript presents results from a computational study examining the influence of integration errors on kinetic energy equipartitioning in canonical ensemble MD simulations. The authors argue that an auxiliary shadow kinetic energy should be used instead of the conventional definition when formulating thermostats and analyzing equipartitioning behavior. They show that use of this shadow kinetic energy leads to improved numerical stability and subsequently demonstrate that accurate estimates of properties in the target ensemble can be obtained by reweighting. This reformulation approach is shown to work successfully for several widely-used thermostats.

This is a nicely presented article that is likely to be of interest the readership of JCP. Overall, it is technically sound, but I have a couple of comments that the authors should consider. 
}}

\textbf{Answer}:

{\color{blue}{
(1) There is a recent paper by Smit and co-workers on equipartitioning in thermostatted MD simulations [Smit and co-workers J. Chem. Theory Comput., 2018, 14 (10), pp 5262-5272]. There is no direct overlap with the work presented here, but the authors may wish to add this reference for interested readers and discuss the results in the context of the current study. 
}}

\textbf{Answer}:

{\color{blue}{
(2) The authors assess the magnitude of sampling errors by examining the dependence of average thermodynamics properties on time step. This assumes that correct canonical sampling is recovered when using a small time step, but this may not be true in practice for a number of reasons (e.g., errors in thermostat implementation). The authors should consider analyzing the sample distributions rather than just the averages. To this end, they may wish to consider implementing some of the numerical tests described in J. Chem. Theory Comput., 2013, 9 (2), pp 909-926. 
}}

\textbf{Answer}:

%\bibliographystyle{ieeetr}
%\bibliography{rigid_bodies.bib}
\end{document}

### 2019-02-25 response_to_reviewers.tex
add comments on check ensemble results

@@
 \usepackage{textcomp}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{booktabs}
+\usepackage{threeparttable}
+
+\usepackage{array}
+\newcolumntype{L}{>{$}l<{$}}
+\newcolumntype{C}{>{$}c<{$}}
+\newcolumntype{R}{>{$}r<{$}}
+\usepackage{booktabs}    
+\usepackage[exponent-product=\times]{siunitx}
+\newcommand{\timestep}{h}
+
 %opening
@@
 Done.
+
 {\color{blue}{
@@
 The conclusion in this paragraph seems dubious.
-The authors appear to confuse numerical artefacts, due to finite time step, with the physical phenomenon of heat capacity dependence on temperature.
+The authors appear to confuse numerical artifacts, due to finite time step, with the physical phenomenon of heat capacity dependence on temperature.
 A much more plausible explanation is that statistical error in the measurement of $c_v$ (which is related to the variance of the total energy) is much larger (for the same simulation sample) than that in other quantities (like kinetic or potential energy).
@@
 \textbf{Answer}:
-Our (unfulfilled) intention here was to suggest that the effect caused by the numerical artifact is similar to that of a small variation in temperature, in which the average $\langle H \rangle$ changes while $c_v$ remains practically unaltered.
-We have rewritten the paragraph so as to make this intention clearer.
-We also employed the expression “are not noticeably influenced” in place of “does not get affected”, in order to acknowledge the statistical nature of these measurements.
+Our (unfulfilled) intention here was to suggest that the effect caused by numerical artifact is similar to that associated with small temperature variations in experiments, where the average energy changes while $c_v$ remains practically unaltered.
+However, as pointed out by the reviewer,  $c_v$ is related to the variance of the total energy, and not to the center of its distribution, which is being affected by discretization errors.
+To emphasize this point, we have rewritten the paragraph avoiding the analogy with the physical phenomena.
+We have also employed the expression “are not noticeably influenced” in place of “does not get affected”, in order to acknowledge the statistical nature of these measurements.
 
@@
 - p.10, bottom of left column: ``Having overcome such barrier ...'' I'm not sure what ``barrier'' the authors have in mind.
-What they propose allows to increase the accuracy of the results (maybe double it, if we are talking about increasing the time step from 1fs to 4fs), or get the same accuracy with 25% of the computational resource (again from 1fs to 4fs), but this is not overcoming a ``barrier".
+What they propose allows to increase the accuracy of the results (maybe double it, if we are talking about increasing the time step from 1fs to 4fs), or get the same accuracy with 25\% of the computational resource (again from 1fs to 4fs), but this is not overcoming a ``barrier".
 So, I suggest either deleting this sentence as unnecessary, or modifying it to reflect more accurately the contribution of this work.
@@
 \textbf{Answer}:
+We were referring to the improvement in accuracy for time-step sizes up to 5 fs, so we deleted the sentence ``Having overcome such barrier'' and modified that paragraph to emphasize our contribution in that terms.
 
@@
 \textbf{Answer}:
+ We sincerely thank the reviewer for the kind comments and suggestions. They have all been considered and are answered below.
 
@@
 \textbf{Answer}:
+The reviewer is totally right.
+We have applied the numerical tests introduced in the suggested reference and verified that the sampled energies are in accordance with the canonical distribution.
+We have included this information on page 6 of the manuscript.
+For time-step sizes with considerable discretization errors, however, those numerical tests are not suitable for analyzing the potential and kinetic energies, since they assume that the Hamiltonian is separable into exclusively momentum- and position-dependent parts, which is no longer true.
+Thus, we applied the maximum likelihood approach on the total energies only.
+The results for the NVT-MD methods are reported in the table below.
+Note that the deviations from the true slope are generally 1$\sigma$ or less for time-step sizes up to 5 fs.
+This is intimately related to the effect observed when examining the constant-volume heat capacity, where the variance in total energy is not influenced by discretization errors as much as the mean of the distribution.
+We consider that this result is analogous to the ensemble validation test and, that is why we decided to not include the table below in the revised manuscript.
 
-%\bibliographystyle{ieeetr}
-%\bibliography{rigid_bodies.bib}
+\begin{table*}
+	\caption{Ensemble validation of the NVT-MD methods using the maximum likelihood approach for different time-step sizes}
+	\label{table:check_ens}
+		\centering
+		\resizebox{\columnwidth}{!}{%
+	\begin{tabular}{r c c c c c c}
+		{$\timestep$} & {estimated $\Delta T$} & {$\sigma$ deviations}  & {estimated $\Delta T$}  & {$\sigma$ deviations}  &    {estimated $\Delta T$}  & {$\sigma$ deviations} \\
+		{(fs)} & {(K)} &   & {(K)}  &  & {(K)}  & \\
+		\hline
+		&\multicolumn{2}{c}{Refined NHC} & \multicolumn{2}{c}{NHC }   &	\multicolumn{2}{c}{KLN}\\
+		1.0	& 3.02(2) & 1.07 & 2.98(2) & 0.66 &  2.97(2) & 1.16 \\
+		2.0 & 3.02(2) & 0.81 & 2.98(2) & 0.73 &  3.00(2) & 0.02 \\
+		3.0 & 2.98(2) & 0.73 & 2.95(2) & 2.33  & 2.99(4) & 0.18 \\
+		4.0 & 3.01(2) & 0.41 & 3.01(2) & 0.65 & 2.92(8) & 0.99 \\
+		5.0 & 3.02(2) & 0.79 & 2.93(2) & 3.05  & 3.02(5) & 0.40 \\
+		6.0 & 3.03(3) & 1.23  & 2.92(2) & 3.87  & 3.14(5) & 2.69  \\
+		7.0 & 2.98(2) & 0.74 & 2.89(2) & 5.35 & 3.03(4) & 0.73 \\
+		\hline
+	\end{tabular}
+}
+\end{table*}
+
 \end{document}

### 2019-02-25 rigid_bodies.tex
Symbolic links for EPS files in main directory
Version submitted to arXiv
add comments on check ensemble results

@@
 
-\usepackage[hidelinks,colorlinks,citecolor=blue]{hyperref}
+\usepackage[hyperindex,breaklinks,hidelinks,colorlinks,citecolor=blue]{hyperref}
 \usepackage{natbib}
@@
 Kinetic energy equipartition is a premise for many deterministic and stochastic molecular dynamics methods that aim at sampling a canonical ensemble.
-While this is expected for real systems, discretization errors introduced by the numerical integration may distort such assumption.
-Fortunately, backward error analysis allows us to identify the quantity that is actually subject to equipartition.
-This is related to a shadow Hamiltonian, which coincides with the specified Hamiltonian only when the time-step size approaches zero.
+While this is expected for real systems, discretization errors introduced by the numerical integration may lead to deviations from equipartition.
+Fortunately, backward error analysis allows us to obtain a higher-order estimate of the quantity that is actually subject to equipartition.
+This is related to a shadow Hamiltonian, which converges to the specified Hamiltonian only when the time-step size approaches zero.
 This paper deals with discretization effects in a straightforward way.
@@
 
-For a symplectic integration of Hamiltonian equations of motion, the breakage of equipartition is arguably only apparent \cite{Eastwood_2010}.
+For a symplectic integration of Hamiltonian equations of motion, the breakdown of equipartition is arguably only apparent \cite{Eastwood_2010}.
 The produced trajectory actually follows the dynamics dictated by a \textit{shadow Hamiltonian} \cite{Tuckerman_2010}, which is no longer separable into exclusively momentum- and position-dependent parts and only coincides with the specified Hamiltonian as the time-step size approaches zero.
@@
 The authors then employed backward error analysis to successfully eliminate the artifacts mentioned above.
-As we manage to prove here, their findings can be extended to the symplectic integration of rigid-body motion as well.
+As we manage to demonstrate here, their findings can be extended to the symplectic integration of rigid-body motion as well.
 
@@
 
-The new method is specially appealing for rigid-body MD, primarily because numerical stability can, in this case, extend up to time-step sizes large enough to yield pronounced discretization errors.
+The new method is especially appealing for rigid-body MD, primarily because numerical stability can, in this case, extend up to time-step sizes large enough to yield pronounced discretization errors.
 For rigid water, which served as an excellent test case, we were able to correct discretization errors in ensemble averages computed with time-step sizes up to $5~\text{fs}$, thus outperforming not only the KLN method \cite{Kamberaj_2005}, but also a reimplementation for rigid bodies of the NHC integrator introduced by Martyna \textit{et al}.
@@
 where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
-While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
+While the form of $U(\vt r, \vt q)$ depends on the specific interaction model, the kinetic energy is given by
 \begin{equation}
@@
 
-Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies piled together.
+Eqs.~\eqref{eq:ODE system for NVE} to \eqref{eq:kinetic energy} can also represent a system of $N$ interacting rigid bodies if we consider that vectors $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$ contain the corresponding properties of all bodies combined in a single vector.
 The size of diagonal matrices $\mt M$ and $\mt I$ also increases to $3N \times 3N$ in this case.
 In addition, $\mt B(\vt q)$ and $\mt C(\vt q)$ become block-diagonal matrices with $4N$ rows and $3N$ columns.
-Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further so as to accommodate a number of freely moving atoms.
+Yet, the notation can be made even more general if $\vt r$, $\vt p$, and $\mt M$ are enlarged further so as to accommodate a number of individual point masses.
 
@@
 where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
-In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively; where the asterisk denotes the state of the system immediately before the propagation occurs.
+In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick that changes the linear and quaternion momenta according to ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively; where the asterisk denotes the state of the system immediately before the propagation occurs.
 The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
@@
 This justifies its more widespread use.
-However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.
+However, having an exact solution for free rotations at hand is crucial for the feasibility of backward error analysis, as it will become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.
 
@@
 In contrast, evaluating the NVE part is expensive for involving force/torque computations and, in addition, its accuracy goes down quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}.
-A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients.
+A high-order integration scheme\cite{Omelyan_2007, Van_zon_2008} could possibly admit larger time steps, but it would raise both cost and complexity due to the need to evaluate (or approximate numerically) the force/torque gradients.
 Omelyan's processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.
-Here we choose to take a simpler path, though.
-Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the implied discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble average are computed.
+Here we choose to take a different path.
+Instead of trying to increase accuracy in the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we attempt to quantify the discretization errors, with which we can both tune the thermostatting procedure and properly weight the sampled configurations when ensemble averages are computed.
 
@@
 
-In order to explain our proposal, we turn the attention again to the NVE case.
+In order to explain our approach, we turn the attention again to the NVE case.
 A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are Hamiltonian as well.
@@
 
-Note that $\refined U$ becomes dependent on $\timestep^2$, but keeps being independent of $\vt p$ and $\vt \pi$.
+Note that $\refined U$ becomes dependent on $\timestep^2$ but, like $U$, it is independent of $\vt p$ and $\vt \pi$.
 In turn, the \textit{refined kinetic energy} can be expressed as
@@
 where $\refined{\mathbf \Omega}$ is a matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$ given by $\tilde{\mt \Omega} = {\mt \Omega} + ({\timestep^2}/{6}) {\mt \Omega} {\mt \Xi} \tr{\mt \Omega}$.
-In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to the stacked vector $[\substack{\vt p \\ \vt \pi}]$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
+In this definition, $\mt \Xi$ is the Hessian matrix of the potential energy with respect to $[\substack{\vt r \\ \vt q}]$, while ${\mt \Omega}$ is a block-diagonal matrix defined so that $K = \frac{1}{2} \tr{[\substack{\vt p \\ \vt \pi}]} {\mt \Omega} [\substack{\vt p \\ \vt \pi}]$.
 Although $\refined{\mathbf \Omega}$ is symmetric like $\mt \Omega$, its structure is not block-diagonal due to the inter-body interactions accounted for in $U(\vt r, \vt q)$ and to the dependency of forces and torques on both $\vt r$ and $\vt q$.
@@
 \cite{Eastwood_2010}, who employed numerical differentiation to estimate the time-derivatives $\dot{\vt r}$ and $\dot{\vt q}$ directly from the obtained NVE trajectory.
-Considering $n$-th order estimators $\dot{\vt r}^{[n]}$ and $\dot{\vt q}^{[n]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
+Considering $k$-th order estimators $\dot{\vt r}^{[k]}$ and $\dot{\vt q}^{[k]}$, we substitute Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:modified kinetic energy} to find out that
 \begin{equation}
 \label{eq:modified kinetic energy estimator}
-\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[n]} + \tr{\vt \pi} \dot{\vt q}^{[n]} \right) + \mathcal{O}(\timestep^n).
+\refined K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}^{[k]} + \tr{\vt \pi} \dot{\vt q}^{[k]} \right) + \mathcal{O}(\timestep^k).
 \end{equation}
@@
 By virtual and incomplete we mean, respectively, that these steps are never incorporated to the trajectory and that they are aborted once the positions and orientations have been updated.
-By this means, we can obtain the estimates $\dot{\vt r}^{[3]}$ and $\dot{\vt q}^{[3]}$ without any additional force evaluation, which results in a small computational overhead.
+Thus, we can obtain the estimates $\dot{\vt r}^{[3]}$ and $\dot{\vt q}^{[3]}$ without any additional force evaluation, which results in a small computational overhead.
 
@@
 In addition, $\refined K$ must be reevaluated after the action of propagator $e^{\frac{\timestep}{2n} {\mathcal L}_0}$, which scales both $\vt p$ and $\vt \pi$ by a factor $e^{-\alpha_1^\ast \frac{\timestep}{2n}}$.
-Because such action leaves the positions and quaternions intact and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ is done as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.
+Because such action leaves the positions and quaternions unchanged and, as a consequence, the matrix $\refined{\mathbf \Omega}(\vt r, \vt q, \timestep^2)$ in Eq.~\eqref{eq:modified kinetic energy} remains unaltered, the reevaluation of $\refined K$ is done as simply as ${\refined K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\refined K}^\ast$.
 
@@
 
+Although the results are not shown here, we have applied a method, introduced and implemented by Shirts ~\cite{Shirts_2013}, that serves for checking the consistency between the sampled energy distributions and the theoretical ones.
+For the NVT-MD methods investigated, we have verified that, generally, the total energy distributions deviate 1$\sigma$ or less from the theoretical ones for time-step sizes up to 5 fs.
+
 \subsection{Refined Hamiltonian Calculation in NVE Simulations}
@@
 \begin{figure}
-	\includegraphics{Figures/NVE.eps}
+	\includegraphics{NVE.eps}
 	\caption{Specified Hamiltonian $\Ham{}$ (solid lines) and refined Hamiltonian $\refined{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\eqref{eq:splitting NVE} with the unsplit rotation method \cite{Silveira_2017}.}
@@
 However, this can only be achieved by limiting the order of approximation of numerical derivatives and sacrificing short-term reversibility.
-This poses the question whether these conditions will influence the long-term stability of the method or not, especially if one tries to make use of large time steps.
+This introduces the question of whether these conditions will influence the long-term stability of the method, especially if one tries to make use of large time steps.
 
 Here we define a drift rate, denoted by $R$, as the long-term rate of change of a quantity in a supposedly equilibrated MD simulation.
-It is defined as the angular coefficient of such quantity expressed as a function of time, determined by a least-square linear regression.
+It is defined as the slope of a least-square regression line expressing such quantity as a function of time.
 Recall that the symmetric scheme of Martyna \textit{et al}. \cite{Martyna_1996} is supposed to preserve the extended energy $H$ given by Eq.~\eqref{eq:nvt extended energy}.
@@
 \begin{figure}
-	\includegraphics{Figures/energy_drift.eps}
+	\includegraphics{energy_drift.eps}
 	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, black squares) and its refined Hamiltonian ($\refined{\Ham{}}$, green up triangles) part for our NVT method, as well as in the extended energy ($H$, red x) for the method due to Martyna \textit{et al}.
@@
 Note that no drift occurs for $\refined{\Ham{}}$, not even for the largest considered time steps, showing that the thermostat stabilizes the numerical integration of the physical variables.
-This has already been noticed by Davidchack \cite{Davidchack_2010} and also by Bond and Leimkhuler \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
+This has already been noticed by Davidchack \cite{Davidchack_2010} and also by Bond and Leimkuhler \cite{Bond_2007}, who in this respect stated that ``the thermostat acts as a sort of reservoir for numerical errors''.
 
@@
 \begin{figure}
-	\includegraphics{Figures/numerical_stability.eps}
+	\includegraphics{numerical_stability.eps}
 	\caption{Time evolution of (a) the extended energy and (b) of each thermostat energy for the NVT method introduced by Kamberaj, Low, and Neal \cite{Kamberaj_2005}.
@@
 \begin{figure*}
-	\includegraphics{Figures/thermodynamic_properties.eps}
+	\includegraphics{thermodynamic_properties.eps}
 	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d) the constant-volume specific heat of water, as obtained from NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules at $T=298~\mathrm{K}$ and $\rho=970~\mathrm{kg/m^3}$ and employing different numerical integration schemes.
@@
 	The Refined CSVR scheme ($\times$) corresponds to the canonical stochastic velocity rescaling method \cite{Bussi_2007} subject to refinement as well.
-	Finally, the KLN scheme ($\blacktriangledown$) consists in the double-thermostat method of Kamberaj, Low, and Neal \cite{Kamberaj_2005}.}
+	Finally, the KLN scheme ($\blacktriangledown$) is the double-thermostat method of Kamberaj, Low, and Neal \cite{Kamberaj_2005}.}
 	\label{fig:properties}
@@
 Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
-The fourth integrator we test is a refined version of the velocity rescaling method of Bussi, Donadio, and Parrinello \cite{Bussi_2007}, which differs from the others for involving a global stochastic thermostat.
+The fourth integrator we test is a refined version of the velocity rescaling method of Bussi, Donadio, and Parrinello \cite{Bussi_2007}, which differs from the others by involving a global stochastic thermostat.
 This is identified in Fig.~\ref{fig:properties} as Refined CSVR (acronym for Canonical Stochastic Velocity Rescaling).
@@
 
-Fig.~\ref{fig:properties}(d) contains results for the constant-volume specific heat capacity.
-Interestingly, these results do not get affected by the discretization errors when $\timestep \leq 5~\mathrm{fs}$, not even if the KLN method is employed.
-Given that the specific heat of water changes little with temperature in the vicinity of $298~\mathrm{K}$, it is not surprising to observe that $c_\mini{V}$ remains unaltered while other properties vary.
+Fig.~\ref{fig:properties}(d) contains results for the constant-volume specific heat capacity, which
+are not noticeably influenced by the discretization errors when $\timestep \leq 5~\mathrm{fs}$, not even if the KLN method is employed.
+This means that the numerical artifacts do not affect the variance in total energy, which determines the $c_v$, as much as they affect the center of the corresponding distribution.
 
@@
 \begin{figure}
-	\includegraphics{Figures/Free_Energy.eps}
+	\includegraphics{Free_Energy.eps}
 	\caption{Difference in free energy per water molecule between the actually sampled and the target states as a function of the time-step size.
@@
 \begin{figure*}
-	\includegraphics{Figures/energy_partition.eps}
+	\includegraphics{energy_partition.eps}
 		\caption{Influence of the time-step size on the kinetic energy partition in NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules.
@@
 The refinement/reweighting approach described in Sec.~\ref{sec:refined_method} corrected discretization effects in computed thermodynamic properties of water for time-step sizes up to $5~\text{fs}$.
-Having overcome such barrier is of paramount importance as it allows one to actually exploit the high numerical stability and useful coarse-graining strategy provided by rigid-body MD.
+This is of paramount importance as it allows one to increase the time-step size without sacrificing accuracy and, thus, to actually exploit the useful coarse-graining strategy provided by rigid-body MD.
 
@@
 The term above is identically null, as it corresponds to the quadratic form built with a skew-symmetric matrix.
-Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{k=1}^3 \omega_k \hat{\mt B}_k ) \vt q$, where each $\hat{\mt B}_k$ is a skew-symmetric permutation matrix (i.e.
-$\tr{\hat{\mt B}}_k = -\hat{\mt B}_k$).
+Another helpful identity taken from Appendix B of Ref.~\citenum{Silveira_2017} is ${\mt B}(\vt q){\vt \omega} = ( \sum_{j=1}^3 \omega_j \hat{\mt B}_j ) \vt q$, where each $\hat{\mt B}_j$ is a skew-symmetric permutation matrix (i.e.
+$\tr{\hat{\mt B}}_j = -\hat{\mt B}_j$).
 This allows us to obtain
 \begin{align*}
-K_{\vt \pi \vt q} &= \frac{1}{2} \left(\diff{\tr{\vt \omega}}{\vt q}\right) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{k=1}^3 \omega_k \tr{\hat{\mt B}}_k = \\
-&= -\frac{1}{4} {\mt B}(\vt \pi) {\mt I}^{-1} \tr{\mt B}(\vt q) - \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k.
+K_{\vt \pi \vt q} &= \frac{1}{2} \left(\diff{\tr{\vt \omega}}{\vt q}\right) \tr{\mt B}(\vt q) + \frac{1}{2} \sum_{j=1}^3 \omega_j \tr{\hat{\mt B}}_j = \\
+&= -\frac{1}{4} {\mt B}(\vt \pi) {\mt I}^{-1} \tr{\mt B}(\vt q) - \frac{1}{2} \sum_{j=1}^3 \omega_j \hat{\mt B}_j.
 \end{align*}

### 2019-02-25 rigid_bodies.tex
include braun reference

@@
 In this case, hybrid Monte Carlo steps were employed with initial velocities that failed to comply with the equipartition between translational and rotational degrees of freedom \cite{Palmer_2018}.
-As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations as an artifact due to the discretization of the equations of motion.
+As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations coupled with noncanonical schemes \cite{Braun_2018}, or as an artifact due to the discretization of the equations of motion.
 This can affect not only the translation/rotation equipartition in rigid-body systems \cite{Davidchack_2010, Silveira_2017}, but also the distribution of energy among molecules of distinct sizes \cite{Eastwood_2010}.

### 2019-02-26 response_to_reviewers.tex
include braun reference
Small changes

@@
 Our (unfulfilled) intention here was to suggest that the effect caused by numerical artifact is similar to that associated with small temperature variations in experiments, where the average energy changes while $c_v$ remains practically unaltered.
-However, as pointed out by the reviewer,  $c_v$ is related to the variance of the total energy, and not to the center of its distribution, which is being affected by discretization errors.
-To emphasize this point, we have rewritten the paragraph avoiding the analogy with the physical phenomena.
-We have also employed the expression “are not noticeably influenced” in place of “does not get affected”, in order to acknowledge the statistical nature of these measurements.
+As the reviewer pointed out, $c_v$ is related to the variance of the total energy distribution.
+However, the results show that such variance is not being affected by discretization errors in the same extent as the average is.
+To emphasize this point, we have rewritten the paragraph while avoiding the analogy with the physical phenomenon.
+We have also employed the expression ``are not noticeably influenced'' in place of ``does not get affected'', in order to acknowledge the statistical nature of these measurements.
 
@@
 \textbf{Answer}:
-We were referring to the improvement in accuracy for time-step sizes up to 5 fs, so we deleted the sentence ``Having overcome such barrier'' and modified that paragraph to emphasize our contribution in that terms.
+%We did not have anything unusual in mind.
+This was simply an unfortunate choice of words.
+We deleted the sentence ``Having overcome such barrier'' and modified the paragraph so as to describe our contribution in a simpler way.
 
@@
 \textbf{Answer}:
- We sincerely thank the reviewer for the kind comments and suggestions. They have all been considered and are answered below.
+ We sincerely thank the reviewer for the kind comments and helpful suggestions. They have been considered in full.
 

### 2019-02-26 rigid_bodies.tex
Small changes

--- 2019-02-25 rigid_bodies.tex
@@
 are not noticeably influenced by the discretization errors when $\timestep \leq 5~\mathrm{fs}$, not even if the KLN method is employed.
-This means that the numerical artifacts do not affect the variance in total energy, which determines the $c_v$, as much as they affect the center of the corresponding distribution.
+This means that the numerical artifacts do not affect the variance as much as they affect the mean of the total energy distribution.
 
@@
 The refinement/reweighting approach described in Sec.~\ref{sec:refined_method} corrected discretization effects in computed thermodynamic properties of water for time-step sizes up to $5~\text{fs}$.
-This is of paramount importance as it allows one to increase the time-step size without sacrificing accuracy and, thus, to actually exploit the useful coarse-graining strategy provided by rigid-body MD.
+This is of paramount importance as it allows us to actually exploit the useful coarse-graining strategy provided by rigid-body MD by increasing the time-step size without sacrificing accuracy.
 

### 2019-02-26 response_to_reviewers.tex
Resubmitted (revised) version

@@
 \textbf{Answer}:
- We sincerely thank the reviewer for the kind comments and helpful suggestions. They have been considered in full.
+ We sincerely thank the reviewer for the kind comments and helpful suggestions. They have been considered in full. Please find our answers below.
 
@@
 \textbf{Answer}:
+We thank the reviewer for pointing out this reference.
+A citation has been included in the introductory section (on page 1 of the manuscript).
 
@@
 \textbf{Answer}:
-The reviewer is totally right.
-We have applied the numerical tests introduced in the suggested reference and verified that the sampled energies are in accordance with the canonical distribution.
+The reviewer is right.
+We have applied the numerical tests introduced in the suggested reference (Shirts, 2013) and verified that the sampled energies are in accordance with the canonical distribution.
 We have included this information on page 6 of the manuscript.
-For time-step sizes with considerable discretization errors, however, those numerical tests are not suitable for analyzing the potential and kinetic energies, since they assume that the Hamiltonian is separable into exclusively momentum- and position-dependent parts, which is no longer true.
-Thus, we applied the maximum likelihood approach on the total energies only.
+It is important to note that, for time-step sizes with noticeable discretization errors, the numerical tests are not suitable for analyzing the potential and kinetic energies separately.
+They assume that the Hamiltonian can be separated into exclusively momentum- and position-dependent parts, which is not true in these cases.
+Therefore, we applied the maximum likelihood approach only to the total energies.
 The results for the NVT-MD methods are reported in the table below.
-Note that the deviations from the true slope are generally 1$\sigma$ or less for time-step sizes up to 5 fs.
-This is intimately related to the effect observed when examining the constant-volume heat capacity, where the variance in total energy is not influenced by discretization errors as much as the mean of the distribution.
-We consider that this result is analogous to the ensemble validation test and, that is why we decided to not include the table below in the revised manuscript.
+Note that the deviations from the true slope are always smaller than 3$\sigma$ for time-step sizes up to 5 fs.
+Higher deviations, according to Shirts (2013), would indicate inconsistency with the canonical ensemble (e.g.
+the author found values above 40$\sigma$ for the knowingly problematic Berendsen thermostat).
+Because the numerical values in the table below would not be essential to the discussions in the manuscript, we decided to leave them out.
 
-\begin{table*}
+\begin{table*}[h!]
 	\caption{Ensemble validation of the NVT-MD methods using the maximum likelihood approach for different time-step sizes}

### 2019-02-26 rigid_bodies.tex
Resubmitted (revised) version

--- 2019-02-26 rigid_bodies.tex
@@
 In this case, hybrid Monte Carlo steps were employed with initial velocities that failed to comply with the equipartition between translational and rotational degrees of freedom \cite{Palmer_2018}.
-As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations as an artifact due to the discretization of the equations of motion.
+As a matter of fact, similar ill-partitioning of kinetic energy can arise in MD simulations, either associated with noncanonical rescaling of velocities \cite{Braun_2018} or as an artifact due to the discretization of the equations of motion.
 This can affect not only the translation/rotation equipartition in rigid-body systems \cite{Davidchack_2010, Silveira_2017}, but also the distribution of energy among molecules of distinct sizes \cite{Eastwood_2010}.
@@
 
-Although the results are not shown here, we have applied a method, introduced and implemented by Shirts ~\cite{Shirts_2013}, that serves for checking the consistency between the sampled energy distributions and the theoretical ones.
-For the NVT-MD methods investigated, we have verified that, generally, the total energy distributions deviate 1$\sigma$ or less from the theoretical ones for time-step sizes up to 5 fs.
+In order to verify if the total energies sampled in the simulations were consistent with the canonical ensemble, we employed the testing procedure developed by Shirts in Ref.~\citenum{Shirts_2013}.
+All the NVT-MD methods investigated here succeeded in the test for time-step sizes up to 5 fs (results not shown).
 

### 2019-02-26 rigid_bodies.tex
Dedication to Affonso Telles

@@
 \begin{acknowledgements}
+This work is dedicated to the memory of Prof. Affonso da Silva Telles, a true scientist who played a major role in laying the foundation of Chemical Engineering research in Brazil.
+
 The authors acknowledge the financial support provided by Petrobras (project code CENPES 16113).

### 2019-02-26 rigid_bodies.tex
change tikz figure for eps

@@
 \usepackage{float}
-\usepackage{tikz}
-\usetikzlibrary{decorations.text}
+\usepackage{graphicx}
+\usepackage{epstopdf}
 
@@
 \begin{figure}
-\begin{tikzpicture}
-\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{${\vt r}(t)$};
-\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
-\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
-\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
-\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
-\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
-\node at (0.0,-0.2)[below]{${\vt r}_{t-2\timestep}$};
-\node at (2.2,-0.2)[below]{${\vt r}_{t-\timestep}$};
-\node at (4.4,-0.2)[below]{${\vt r}_{t}$};
-\node at (6.6,-0.2)[below]{${\vt r}_{t+\timestep}$};
-\path[->,thick] (1) edge [bend left=45] (2);
-\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
-\path[dashed,->,thick] (1) edge [bend right=45] (0);
-\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
-\path[dashed,->,thick] (2) edge [bend left=45] (3);
-\end{tikzpicture}
+	\includegraphics{scheme}
 \caption{Schematic representation of the executed steps in a single time-step of our refined NVT dynamics.
@@
 \begin{figure}
-	\includegraphics{NVE.eps}
+	\includegraphics{NVE}
 	\caption{Specified Hamiltonian $\Ham{}$ (solid lines) and refined Hamiltonian $\refined{\Ham{}}$ (dotted lines) for different time-step sizes ($\timestep$) obtained from NVE MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules using the numerical scheme of Eq.~\eqref{eq:splitting NVE} with the unsplit rotation method \cite{Silveira_2017}.}
@@
 \begin{figure}
-	\includegraphics{energy_drift.eps}
+	\includegraphics{energy_drift}
 	\caption{Influence of the time-step size ($\timestep$) on the long-term drift rate ($R$) in: the extended energy ($\refined{H}$, black squares) and its refined Hamiltonian ($\refined{\Ham{}}$, green up triangles) part for our NVT method, as well as in the extended energy ($H$, red x) for the method due to Martyna \textit{et al}.
@@
 \begin{figure}
-	\includegraphics{numerical_stability.eps}
+	\includegraphics{numerical_stability}
 	\caption{Time evolution of (a) the extended energy and (b) of each thermostat energy for the NVT method introduced by Kamberaj, Low, and Neal \cite{Kamberaj_2005}.
@@
 \begin{figure*}
-	\includegraphics{thermodynamic_properties.eps}
+	\includegraphics{thermodynamic_properties}
 	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d) the constant-volume specific heat of water, as obtained from NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules at $T=298~\mathrm{K}$ and $\rho=970~\mathrm{kg/m^3}$ and employing different numerical integration schemes.
@@
 \begin{figure}
-	\includegraphics{Free_Energy.eps}
+	\includegraphics{Free_Energy}
 	\caption{Difference in free energy per water molecule between the actually sampled and the target states as a function of the time-step size.
@@
 \begin{figure*}
-	\includegraphics{energy_partition.eps}
+	\includegraphics{energy_partition}
 		\caption{Influence of the time-step size on the kinetic energy partition in NVT-MD simulations of 903 TIP3P \cite{Jorgensen_1983} water molecules.
